---
title: "How wrong is the naive cumulative-proportion reformat? (Drew's question)"
date: 2026-09-26
format:
  html:
    embed-resources: true
    toc: true
---

## The question

Drew (UW) clarified his design: constant temperatures, the same cohort checked
repeatedly, individual times of death. He proposes the "trivial reformat" —
compute cumulative survival of the cohort at every checkpoint and feed those
as if they were independent temperature × duration cells into the current
package — and himself suspects this "could produce overconfident intervals".
He also suggests adding an individual random effect to compensate.

This note quantifies the reformat against the correct likelihood (see the
2026-09-25 feasibility note for the theory), including whether the package's
**default beta-binomial** family accidentally compensates.

## Design

20 replicates. Each: 4 temperatures (37–43 °C) × 40 individuals, true
log₁₀ t₅₀ linear in T with **z = 4**, log-logistic shape k = 8, daily checks
to day 14, right-censoring after. Three fits per replicate, identical 4PL
mean structure, identical mean-parameter priors, same seeds:

- **A** naive **binomial**: cumulative alive at each checkpoint out of the
  initial 40 (Drew's reformat, with the plain binomial);
- **B** naive **beta-binomial**: same cells, the package's default family
  (φ estimated);
- **C** correct **person-period** conditional Bernoulli
  ($p_j = 1 - S(t_j)/S(t_{j-1})$).

Script: [`2026-09-26-naive-vs-conditional-sim.R`](2026-09-26-naive-vs-conditional-sim.R)
(ported from the session scratchpad; regenerates the CSV). Results are read
from the committed CSV and summarised live below.

## Results

```{r summary}
d <- read.csv("2026-09-26-naive-vs-conditional-results.csv")
stopifnot(nrow(d) == 60, all(d$fit_ok == 1), all(d$n_div == 0))
lab <- c(A = "A naive binomial", B = "B naive beta-binomial (pkg default)",
         C = "C correct person-period")
out <- do.call(rbind, lapply(c("A", "B", "C"), function(m) {
  x <- d[d$method == m, ]
  data.frame(method = lab[m], z_mean_of_medians = round(mean(x$z_med), 3),
             mean_CrI_width = round(mean(x$width), 3),
             coverage = paste0(sum(x$covers), "/", nrow(x)),
             max_Rhat = round(max(x$max_rhat), 3))
}))
knitr::kable(out, row.names = FALSE)
wr <- function(m) d$width[d$method == m] / d$width[d$method == "C"]
cat(sprintf("paired width ratio vs C:  A/C mean %.2f (range %.2f-%.2f);  B/C mean %.2f (range %.2f-%.2f)\n",
            mean(wr("A")), min(wr("A")), max(wr("A")),
            mean(wr("B")), min(wr("B")), max(wr("B"))))
cat(sprintf("beta-binomial phi, median across replicates: %.0f\n",
            median(d$phi_med[d$method == "B"], na.rm = TRUE)))
```

## Reading, stated flat

- Point estimates of z are essentially unbiased under all three methods —
  the naive reformat's mean structure is correct, which is exactly why it
  *looks* fine.
- **Naive binomial**: intervals ~28% narrower than the correct method;
  empirical coverage 16/20. The predicted overconfidence, confirmed in
  direction; with R = 20 the coverage estimate carries ~±5 pp Monte-Carlo
  error, so this resolves gross under-coverage, not its precise size.
- **Naive beta-binomial (the package default) largely compensated here**:
  widths within ~4% of correct, same empirical coverage. Mechanism: the
  cohort's realised survival trajectory wanders around S(t) as one correlated
  path, which the per-cell dispersion parameter absorbs.
- Individual random effects are **not** the fix Drew hopes: after aggregation
  to cohort proportions there is no individual row to attach them to, and the
  problem is likelihood structure, not a missing variance component.

## What this does NOT license

One design point only (daily checks, moderate censoring, n = 40, generator
matching the fitted model with **no real overdispersion**). With real data,
φ has to absorb genuine extra-binomial variation *and* the dependence — it
cannot do both, and there is no way to tell how it splits. Sparser checks,
heavier censoring, smaller n, or interest in quantities other than z were not
tested. The beta-binomial shortcut is a tolerable stopgap in designs like
this one; the conditional construction is the defensible method — and its
reformat (per-interval deaths over at-risk counts, instead of cumulative
proportions over the initial count) is no harder to produce.

## Daniel's proposed formulation (added same day)

Daniel proposed: individual × check-time rows with 0/1 dead status, Bernoulli,
the 4PL as the mean (direct CTmax/z parameterisation), plus `(1 | individual)`
and possibly `(1 + log t | individual)`. Tested on the same seeds
([`2026-09-26-daniel-formulation-sim.R`](2026-09-26-daniel-formulation-sim.R)):

```{r daniel}
d <- read.csv("2026-09-26-daniel-formulation-results.csv")
lab <- c(D1 = "D1 Daniel rows, no RE", D2 = "D2 + (1|id) on mid",
         D3 = "D3 + (1 + lt|id), 3 reps")
out <- do.call(rbind, lapply(c("D1","D2","D3"), function(m) {
  x <- d[d$method == m & d$fit_ok == 1, ]
  data.frame(method = lab[m], n = nrow(x),
             z_mean_of_medians = round(mean(x$z_med), 3),
             mean_CrI_width = round(mean(x$width), 3),
             coverage = paste0(sum(x$covers), "/", nrow(x)),
             total_divergences = sum(x$n_div),
             max_Rhat = round(max(x$max_rhat), 3),
             sd_id_median = round(median(x$sd_mid_med), 3))
}))
knitr::kable(out, row.names = FALSE)
```

Reference (same seeds): naive binomial 0.520 / 16-20; naive beta-binomial
0.690 / 18-20; correct person-period 0.722 / 18-20.

Reading, flat:

- **D1 = the naive binomial**, as predicted by aggregation: per-replicate
  differences are Monte-Carlo noise (mean |Δz median| ≈ 0.01). Cumulative 0/1
  rows with the 4PL mean are Drew's reformat, disaggregated.
- **D2: the `(1|id)` patch largely worked for z intervals** in this design —
  widths ~97% of correct, coverage matching. An honest correction: the prior
  analytical expectation here was that the RE could not repair the
  dependence; for interval *width* on z, in this design, it did (same
  compensation family as the beta-binomial's φ).
- **But the estimated "frailty" is an artifact.** The generator contains
  **zero** individual heterogeneity, yet sd(1|id) posteriors sat at 0.19–0.27
  across all 20 replicates. The RE absorbs pseudo-replication and reports it
  as biology. Drew's stated plan — "add a random effect for individual and
  see what variance is explained" — would read dependence as individual
  variation every time.
- **Sampler health degrades**: divergences appear (24 across the D2 reps,
  max R̂ 1.23 in one), fits ~15× slower than the correct construction. D3's
  random time-slope is not identifiable from one event per animal: 2–17%
  divergent transitions per fit and sd(lt) pulled to ~0.06 — the term adds
  geometry problems and no information. (Time-varying death probability is
  already the 4PL curve's job; the slope RE would model individual
  differences in it, which single-spell data cannot support.)
- Same scope limit as before: one benign design; with real heterogeneity
  *and* dependence present, one variance component cannot represent both, and
  there is no way to tell how it splits.

Verdict: Daniel's formulation is two substitutions from correct — drop rows
after death (at-risk rows only) and replace the cumulative 4PL mean with the
conditional interval probability $1 - S(t_j)/S(t_{j-1})$. Then it *is* the
verified construction, the geometry is clean, and `(1|id)` becomes a real
frailty estimating real heterogeneity (with the single-spell identifiability
caveat from the 2026-09-25 note).

## Data layouts (Daniel's question, same day; no slopes anywhere)

Both layouts derive by one-line transforms from the same minimal raw sheet —
one row per animal: `id, temp, last_alive, first_dead (NA = censored)`,
optional `tank`, `enrol` (enrolment time), or an exact `t_death` for
biosensor animals.

**(a) Conditional-interval (correct).** One row per animal per check-interval
while alive at the interval start; rows STOP at death. Columns: `id`,
`temp_c`, `first` (first-interval flag), `logd_lo`, `logd_hi` (interval
endpoints, log10 scale), `death` (1 only on a final observed-death row).
Bernoulli mean `1 - S(logd_hi)/S(logd_lo)`, with `first = 1` rows using
S(start of exposure) as the denominator (no log10(0) stand-in needed).
IMPORTANT READING NOTE: `death` is the CONDITIONAL indicator for that one
interval — "given alive at logd_lo, did it die by logd_hi?" — NOT cumulative
status. Each animal therefore has at most one 1, always on its final row;
censored animals end with all zeros and simply run out of rows. Rows are
keyed by (animal, interval); an animal owns as many rows as intervals it
started alive. Irregular per-animal schedules fine; late enrolment = first interval
starts at `enrol` (the left-truncation fix). Example, animal dead at the
day-4 check: 4 rows, death = 0,0,0,1.

**(b) Daniel's cumulative-status Bernoulli.** One row per animal per check,
single `logd` column, cumulative `dead` status, mean `1 - S(logd)`, plus
`(1|id)` on `mid`. Same animal: 14 rows, death = 0,0,0,1,1,...,1 — the ten
post-death rows are the pseudo-data quantified above, and the reason (b)
needs the RE crutch that then misreports frailty (sd ~0.2 with truth 0).

Notes common to both: same grain, same 0/1 response, `temp_c` and log10 time
exactly as the model already carries them (`logd`); no random slopes —
time-varying death probability is the 4PL curve itself, and `(1 + lt | id)`
was shown unidentifiable above. In (a), `(1|id)` is optional biology (true
frailty; weakly identified single-spell — tank/batch REs preferred), not a
validity requirement. Exact-time biosensor animals skip row expansion and
enter via the log-logistic density term. The planned `standardize_data()`
extension takes the RAW sheet and builds layout (a) internally.

### Worked example (runnable)

```{r layout-demo}
## RAW sheet — what Drew records. One row per animal. Checks at days 1..6.
raw <- data.frame(
  id         = 1:4,
  temp       = c(39, 41, 39, 41),
  enrol      = c(0,  0,  0,  2),           # animal 4 enrolled late (day 2)
  last_alive = c(2,  0,  6,  4),           # 0 = never seen alive at a check
  first_dead = c(3,  1,  NA, 5)            # NA = right-censored at day 6
)
cat("== RAW (one row per animal) ==\n"); print(raw)

checks <- 0:6
## (a) CONDITIONAL-INTERVAL expansion: rows only while at risk; stops at death.
pp <- do.call(rbind, lapply(seq_len(nrow(raw)), function(r) {
  a   <- raw[r, ]
  cps <- checks[checks >= a$enrol]                      # this animal's schedule
  end <- if (is.na(a$first_dead)) max(cps) else a$first_dead
  lo  <- cps[cps < end]; hi <- cps[cps > a$enrol & cps <= end]
  data.frame(id = a$id, temp_c = a$temp - 40,
             first   = as.integer(lo == 0),             # S(lo) = up exactly
             logd_lo = ifelse(lo == 0, NA, round(log10(lo), 3)),
             logd_hi = round(log10(hi), 3),
             death   = as.integer(!is.na(a$first_dead) & hi == a$first_dead))
}))
cat("\n== (a) CONDITIONAL-INTERVAL rows (mean = 1 - S(hi)/S(lo); S(lo)=up when first) ==\n")
print(pp, row.names = FALSE)

## (b) Daniel's cumulative-status layout for the same animals.
cum <- do.call(rbind, lapply(seq_len(nrow(raw)), function(r) {
  a  <- raw[r, ]; d <- checks[checks > a$enrol]
  data.frame(id = a$id, temp_c = a$temp - 40, logd = round(log10(d), 3),
             dead = as.integer(!is.na(a$first_dead) & d >= a$first_dead))
}))
cat("\n== (b) CUMULATIVE-STATUS rows (mean = 1 - S(logd); needs (1|id)) ==\n")
print(cum, row.names = FALSE)
cat(sprintf("\nrow counts: conditional = %d, cumulative = %d (the difference is post-death padding)\n",
            nrow(pp), nrow(cum)))
```

## Updated draft reply to Drew (supersedes the 2026-09-25 draft)

> Hi Drew — thanks, that clarifies things nicely, and your statistical
> instincts are good on every point. Three things.
>
> First, your suspicion about the cumulative reformat is correct. We
> simulated your exact situation (daily checks of one cohort, four constant
> temperatures): feeding cumulative proportions into a plain binomial gave
> intervals for z about 30% too narrow (empirical coverage ~80% at nominal
> 95%). Interestingly, with our default beta-binomial family the estimated
> overdispersion largely absorbed the dependence in that benign setting
> (widths within a few percent of the correct likelihood) — but we would not
> rely on that with real data, where φ also has to soak up genuine
> overdispersion. An individual random effect can't attach to the
> aggregated version at all (no individual rows left), and on
> individual-level cumulative status rows it does something sneakier: in our
> simulations it recovered honest-looking interval widths, but the estimated
> individual variance was pure artifact — the generator had zero individual
> heterogeneity, yet sd(individual) came out ~0.2 every time. So "see what
> variance is explained" would read pseudo-replication as biology.
>
> Second, the correct construction is just as easy to format as the one you
> described: instead of cumulative survival out of the initial n, record
> per-interval deaths out of the number still alive at the interval start.
> Modelled with conditional death probability 1 − S(t_j)/S(t_{j−1}) — the
> standard discrete-time survival construction — this is *algebraically
> identical* to the interval-censored log-logistic likelihood you were
> considering, and it uses exactly the parameters bayesTLS already estimates
> (our 4PL in log₁₀ duration *is* the log-logistic survivor function; z is
> its temperature-acceleration coefficient). We've prototyped this in brms
> and our post-fit machinery (`tls()`, which works on hand-written brms fits)
> extracts z and CTmax from it unchanged — including absolute thresholds —
> provided the model is written with the package's four named sub-parameters;
> the ready-made template script we'd send you does exactly that.
> Your continuously monitored oysters/mussels slot into the same model via
> the exact-time log-logistic density (that removes the *interval* censoring;
> survivors at the end are still right-censored, which is handled).
>
> Third, the coral case fits more neatly than you might expect: define the
> event as "fragment reaches 10% colony death" and each fragment contributes
> one (interval-censored) event time — the polyp-level dependence then lives
> inside the measurement of the endpoint, not the likelihood. Four
> temperatures is thin, but that is what priors are for.
>
> Please do open the GitHub issue — native support (a time-to-event input in
> `standardize_data()`, same model underneath) is a planned extension, and
> your mussel experiments would make an ideal test case. One caveat to keep
> us honest: all of this assumes constant temperature per individual, which
> you've confirmed; ramping designs are a genuinely open modelling problem.

## Open questions for Daniel

1. Happy with the reply as drafted (esp. sharing the brms recipe)?
2. Should the eventual docs explicitly warn against the cumulative reformat,
   given the beta-binomial default makes it *look* acceptable?
