---
title: "Testing reusable Bayesian TDT functions"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-summary: "Show code"
    code-tools: true
execute:
  warning: false
  message: false
params:
  refit_model: false
  show_sampler_progress: false
  model_path: "Rdata/model_lethal_TDT_beta_binomial.RDS"
  ndraws: 1000
---

## Purpose

This notebook turns the prototype code in `tdt_problems.qmd` and `R/bayesian_TDT_shrimp.qmd` into reusable functions that can be tested on the shrimp lethal TDT data before being reused for plant, cell, or multi-species case studies. The functions expect exposure duration in hours. Plotting and tables can convert hours to minutes when needed. The functions are written around one analysis pattern:

1.  Prepare survival-count data in a standard format.
2.  Fit a joint Bayesian 4-parameter logistic model to the raw counts.
3.  Use a beta-binomial likelihood for overdispersion instead of an observation-level random effect.
4.  Derive survival curves, TDT curves, classical TDT summaries, TDT landscapes, and heat-injury predictions from the fitted model.

Both the midpoint (`mid`) and log-steepness (`logk`) of the survival curve are linear functions of centred temperature (`temp_c`), so the curve's position and shape both shift with temperature.

The fitting chunk is deliberately switched off by default through `params$fit_model`. Set it to `true` in the YAML when you want to run the beta-binomial model.

## Setup

```{r}
required_packages <- c(
  "bayesplot",
  "brms",
  "dplyr",
  "flextable",
  "ggplot2",
  "knitr",
  "loo",
  "patchwork",
  "performance",
  "posterior",
  "purrr",
  "readr",
  "readxl",
  "scales",
  "tidyr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install these packages before rendering this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))

project_root <- if (dir.exists("data") && dir.exists("R")) {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
} else {
  normalizePath(file.path(getwd(), ".."), winslash = "/", mustWork = FALSE)
}

if (dir.exists(file.path(project_root, "data"))) {
  knitr::opts_knit$set(root.dir = project_root)
}

if (!exists("params")) {
  params <- list(
    fit_model = FALSE,
    show_sampler_progress = FALSE,
    model_path = "Rdata/model_lethal_TDT_beta_binomial.RDS",
    ndraws = 1000
  )
} else if (is.null(params$show_sampler_progress)) {
  params$show_sampler_progress <- FALSE
}

tdt_output_dir <- "Rdata"
tdt_fig_dir    <- "fig"

theme_tdt <- theme_classic(base_size = 13) +
  theme(
    axis.title = element_text(face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.7),
    legend.position = "right"
  )

tdt_quantile <- function(x, probs = c(0.025, 0.5, 0.975)) {
  stats::quantile(x, probs = probs, na.rm = TRUE, names = FALSE)
}

`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

tdt_ft <- function(x, digits = 2, caption = NULL) {
  x <- as.data.frame(x)

  ft <- flextable::flextable(x)
  ft <- flextable::theme_booktabs(ft)
  ft <- flextable::fontsize(ft, size = 9, part = "all")
  ft <- flextable::align(ft, align = "center", part = "all")
  ft <- flextable::align(ft, j = 1, align = "left", part = "all")

  num_cols <- names(x)[vapply(x, is.numeric, logical(1))]
  if (length(num_cols) > 0) {
    ft <- flextable::colformat_num(
      ft,
      j = num_cols,
      digits = digits,
      na_str = ""
    )
  }

  if (!is.null(caption)) {
    ft <- flextable::set_caption(ft, caption = caption)
  }

  flextable::autofit(ft)
}

format_interval <- function(median, lower, upper, digits = 2) {
  paste0(
    round(median, digits),
    " [",
    round(lower, digits),
    ", ",
    round(upper, digits),
    "]"
  )
}
```

## Function Library

### Small Helpers

```{r}
tdt_num <- function(x) {
  formatC(x, digits = 16, format = "fg", flag = "#")
}

tdt_check_columns <- function(data, cols, arg_name = "columns") {
  cols <- cols[!is.na(cols) & nzchar(cols)]
  missing <- setdiff(cols, names(data))

  if (length(missing) > 0) {
    stop(
      "Missing ", arg_name, ": ",
      paste(missing, collapse = ", "),
      call. = FALSE
    )
  }

  invisible(TRUE)
}

tdt_format_random_effects <- function(random_effects = NULL) {
  if (is.null(random_effects) || length(random_effects) == 0) {
    return(character())
  }

  vapply(
    random_effects,
    function(term) {
      term <- trimws(term)
      if (grepl("^\\(", term)) {
        term
      } else {
        paste0("(1 | ", term, ")")
      }
    },
    character(1)
  )
}

tdt_random_effect_variables <- function(random_effects = NULL) {
  terms <- tdt_format_random_effects(random_effects)

  unique(unlist(lapply(terms, function(term) {
    vars <- all.vars(stats::as.formula(paste("~", term)))
    vars[vars != "1"]
  })))
}

clock_to_minutes <- function(x) {
  if (inherits(x, "POSIXt")) {
    return(
      as.numeric(format(x, "%H")) * 60 +
        as.numeric(format(x, "%M")) +
        as.numeric(format(x, "%S")) / 60
    )
  }

  if (inherits(x, "hms") || inherits(x, "difftime")) {
    return(as.numeric(x, units = "mins"))
  }

  if (is.numeric(x)) {
    # Excel clock times often arrive as fractions of a day.
    if (all(is.na(x) | (x >= 0 & x <= 1))) {
      return(x * 24 * 60)
    }

    return(x)
  }

  x_chr <- as.character(x)
  parsed <- suppressWarnings(as.POSIXct(x_chr, format = "%H:%M:%S", tz = "UTC"))

  as.numeric(format(parsed, "%H")) * 60 +
    as.numeric(format(parsed, "%M")) +
    as.numeric(format(parsed, "%S")) / 60
}
```

### Data Preparation

`prepare_tdt_survival_data()` standardises the columns used by all downstream functions. The model uses `duration` in the unit supplied by the dataset; for the shrimp data that unit is hours. Later plotting functions use `time_multiplier` when we want to show results in minutes.

```{r}
# prepare_tdt_survival_data ---------------------------------------------------
# Standardise raw experimental records into the format expected by all
# downstream fitting and prediction functions.  This is the single entry
# point for data: everything downstream assumes the standard column names
# produced here (assay_temp, duration, logd, temp_c, n_total, n_surv,
# n_dead, survival).
#
# @param data           Raw data frame or tibble.
# @param temp           String: column name of the assay temperature in °C.
# @param duration       String: column name of the exposure duration.  The
#                       unit is whatever is in the data; document it with
#                       duration_unit so downstream callers know what
#                       time_multiplier to use for minute-scale plots.
# @param n_total        String: column name for total individuals per
#                       experimental unit (well, cup, tank).
# @param n_surv         String or NULL: column of survivor counts.  Supply
#                       exactly one of n_surv, n_dead, survival, mortality.
# @param n_dead         String or NULL: column of death counts.  Converted
#                       to n_surv internally via n_surv = n_total - n_dead.
# @param survival       String or NULL: column of survival proportions [0,1].
#                       Converted to integer counts.
# @param mortality      String or NULL: column of mortality proportions [0,1].
#                       Converted to n_surv = round((1 - mortality) * n_total).
# @param random_effects Character vector or NULL: grouping variables for
#                       random effects (e.g. c("Date", "Tank")).  These
#                       columns are converted to factors.
# @param duration_unit  String label for the duration column's unit.
#                       Stored in metadata; not used for any computation.
#                       Default "hours".
# @param temp_mean      Numeric or NULL: value to subtract from assay_temp
#                       to form the centred covariate temp_c.  If NULL
#                       (default), the mean of assay_temp in the filtered
#                       data is used.  Pass a fixed value to align multiple
#                       datasets to the same centre.
#
# @return A tibble with standardised columns plus a "tdt_meta" attribute
#         list storing temp_mean, duration_unit, and random_effects for
#         use by downstream functions.
prepare_tdt_survival_data <- function(data,
                                      temp,
                                      duration,
                                      n_total,
                                      n_surv         = NULL,
                                      n_dead         = NULL,
                                      survival       = NULL,
                                      mortality      = NULL,
                                      random_effects = NULL,
                                      duration_unit  = "hours",
                                      temp_mean      = NULL) {
  needed <- c(temp, duration, n_total, n_surv, n_dead, survival, mortality,
              tdt_random_effect_variables(random_effects))
  tdt_check_columns(data, needed, "input columns")

  if (sum(!vapply(list(n_surv, n_dead, survival, mortality), is.null, logical(1))) != 1) {
    stop(
      "Supply exactly one of n_surv, n_dead, survival, or mortality.",
      call. = FALSE
    )
  }

  out <- as.data.frame(data)
  out$n_total <- as.integer(round(as.numeric(out[[n_total]])))

  if (!is.null(n_surv)) {
    out$n_surv <- as.integer(round(as.numeric(out[[n_surv]])))
  }

  if (!is.null(n_dead)) {
    out$n_surv <- out$n_total - as.integer(round(as.numeric(out[[n_dead]])))
  }

  if (!is.null(survival)) {
    prop_surv <- pmin(pmax(as.numeric(out[[survival]]), 0), 1)
    out$n_surv <- as.integer(round(prop_surv * out$n_total))
  }

  if (!is.null(mortality)) {
    prop_surv <- 1 - pmin(pmax(as.numeric(out[[mortality]]), 0), 1)
    out$n_surv <- as.integer(round(prop_surv * out$n_total))
  }

  out$n_surv <- pmin(pmax(out$n_surv, 0), out$n_total)
  out$n_dead <- out$n_total - out$n_surv
  out$survival <- out$n_surv / out$n_total
  out$assay_temp <- as.numeric(out[[temp]])
  out$duration <- as.numeric(out[[duration]])
  out$logd <- log10(out$duration)

  out <- out %>%
    dplyr::filter(
      is.finite(n_total),
      is.finite(n_surv),
      is.finite(assay_temp),
      is.finite(duration),
      n_total > 0,
      duration > 0
    )

  if (is.null(temp_mean)) {
    temp_mean <- mean(out$assay_temp, na.rm = TRUE)
  }

  out$temp_c <- out$assay_temp - temp_mean

  for (re_var in tdt_random_effect_variables(random_effects)) {
    out[[re_var]] <- factor(out[[re_var]])
  }

  attr(out, "tdt_meta") <- list(
    temp_mean = temp_mean,
    duration_unit = duration_unit,
    random_effects = random_effects
  )

  tibble::as_tibble(out)
}
```

### Fitting A Beta-Binomial Bayesian 4PL

The survival curve is

``` text
p_surv = low + (up - low) / (1 + exp(k * (log10(duration) - mid)))
```

with `k = exp(logk)`, a bounded lower asymptote, and a bounded upper asymptote. All four 4PL sub-models (`lraw`, `uraw`, `logk`, `mid`) are linear functions of `temp_c`, so the full shape of the survival curve — its asymptotes, steepness, and midpoint — can shift with temperature. The model uses `brms::beta_binomial(link = "identity")` with `trials(n_total)`, so `mu` is the survival probability directly (no logit wrapper needed).

The default priors are weak and not estimated from the dataset being analysed. On the probability scale, the lower asymptote is constrained to `(0, 0.5)` and is centred near 0.02; the upper asymptote is constrained to `(0.5, 1)` and is centred near 0.98. The midpoint prior is centred on `log10(duration) = 0.5` with a broad scale, which says only that the threshold exposure is probably within a very wide range around the supplied duration unit.

```{r}
# make_tdt_4pl_formula --------------------------------------------------------
# Build the brms nonlinear formula for the beta-binomial 4PL TDT model.
# Both the midpoint (mid) and log-steepness (logk) are linear in temp_c.
#
# @param random_effects  Character vector of grouping variables for random
#                        intercepts on mid (e.g. c("Date", "Tank")).
# @param low_bounds      Numeric length-2: (min, max) for lower asymptote.
#                        Default c(0.001, 0.499).
# @param up_bounds       Numeric length-2: (min, max) for upper asymptote.
#                        Default c(0.501, 0.999).
# @return A brmsformula object.
make_tdt_4pl_formula <- function(random_effects = NULL,
                                 low_bounds     = c(0.001, 0.499),
                                 up_bounds      = c(0.501, 0.999)) {
  low_min   <- low_bounds[1]
  low_width <- diff(low_bounds)
  up_min    <- up_bounds[1]
  up_width  <- diff(up_bounds)

  low_expr <- paste0(
    "(", tdt_num(low_min), " + inv_logit(lraw) * ", tdt_num(low_width), ")"
  )
  up_expr <- paste0(
    "(", tdt_num(up_min), " + inv_logit(uraw) * ", tdt_num(up_width), ")"
  )

  mu_expr <- paste0(
    "mu ~ ", low_expr,
    " + (", up_expr, " - ", low_expr,
    ") / (1 + exp(exp(logk) * (logd - mid)))"
  )

  mid_terms <- c("temp_c", tdt_format_random_effects(random_effects))
  mid_rhs   <- paste(mid_terms, collapse = " + ")

  formula_parts <- list(
    stats::as.formula("n_surv | trials(n_total) ~ mu"),
    brms::nlf(stats::as.formula(mu_expr)),
    brms::lf(stats::as.formula("lraw ~ temp_c")),
    brms::lf(stats::as.formula("uraw ~ temp_c")),
    brms::lf(stats::as.formula("logk ~ temp_c")),
    brms::lf(stats::as.formula(paste("mid ~", mid_rhs)))
  )

  do.call(brms::bf, c(formula_parts, list(nl = TRUE)))
}

# make_tdt_4pl_priors ---------------------------------------------------------
# Build weakly informative priors for the beta-binomial 4PL model.
#
# @param data             Prepared data frame from prepare_tdt_survival_data().
# @param random_effects   Character vector of random-effect grouping variables.
# @param prior_random_sd  Prior string for random-effect SDs. Default "exponential(2)".
# @param prior_phi        Prior string for beta-binomial precision. Default "gamma(2, 0.1)".
# @param low_bounds       Numeric length-2 for lower asymptote. Default c(0.001, 0.499).
# @param up_bounds        Numeric length-2 for upper asymptote. Default c(0.501, 0.999).
# @return A brms prior object.
make_tdt_4pl_priors <- function(data,
                                random_effects  = NULL,
                                prior_random_sd = "exponential(2)",
                                prior_phi       = "gamma(2, 0.1)",
                                low_bounds      = c(0.001, 0.499),
                                up_bounds       = c(0.501, 0.999)) {

  mid_start <- median(data$logd, na.rm = TRUE)
  lraw_mean <- qlogis((0.02 - low_bounds[1]) / diff(low_bounds))
  uraw_mean <- qlogis((0.98 - up_bounds[1])  / diff(up_bounds))
  logk_mean <- log(2)

  priors <- c(
    brms::set_prior(
      paste0("normal(", tdt_num(lraw_mean), ", 1)"),
      class = "b", nlpar = "lraw", coef = "Intercept"
    ),
    brms::set_prior(
      "normal(0, 0.5)",
      class = "b", nlpar = "lraw"
    ),
    brms::set_prior(
      paste0("normal(", tdt_num(uraw_mean), ", 1)"),
      class = "b", nlpar = "uraw", coef = "Intercept"
    ),
    brms::set_prior(
      "normal(0, 0.5)",
      class = "b", nlpar = "uraw"
    ),
    brms::set_prior(
      paste0("normal(", tdt_num(logk_mean), ", 1)"),
      class = "b", nlpar = "logk", coef = "Intercept"
    ),
    brms::set_prior(
      "normal(0, 0.5)",
      class = "b", nlpar = "logk"
    ),
    brms::set_prior(
      paste0("normal(", tdt_num(mid_start), ", 1.5)"),
      class = "b", nlpar = "mid", coef = "Intercept"
    ),
    brms::set_prior(
      "normal(0, 0.6)",
      class = "b", nlpar = "mid"
    ),
    brms::set_prior(prior_phi, class = "phi")
  )

  for (re_var in tdt_random_effect_variables(random_effects)) {
    priors <- c(
      priors,
      brms::set_prior(prior_random_sd, class = "sd", nlpar = "mid", group = re_var)
    )
  }

  priors
}

# make_tdt_4pl_init -----------------------------------------------------------
# Build per-chain initialisation for the beta-binomial 4PL model.
# Returns a zero-argument closure with valid starting values.
#
# @param data        Prepared data frame from prepare_tdt_survival_data().
# @param low_bounds  Numeric length-2 for lower asymptote. Default c(0.001, 0.499).
# @param up_bounds   Numeric length-2 for upper asymptote. Default c(0.501, 0.999).
# @return A zero-argument closure returning a named list of starting values.
make_tdt_4pl_init <- function(data,
                               low_bounds = c(0.001, 0.499),
                               up_bounds  = c(0.501, 0.999)) {

  lraw_init <- qlogis((0.02 - low_bounds[1]) / diff(low_bounds))
  uraw_init <- qlogis((0.98 - up_bounds[1])  / diff(up_bounds))
  logk_init <- log(2)

  dat_mid <- data %>%
    dplyr::filter(is.finite(survival), survival > 0.05, survival < 0.95)

  mid_init <- if (nrow(dat_mid) > 0) {
    median(dat_mid$logd, na.rm = TRUE)
  } else {
    median(data$logd, na.rm = TRUE)
  }

  function() {
    list(
      b_lraw = array(c(lraw_init, 0), dim = 2),
      b_uraw = array(c(uraw_init, 0), dim = 2),
      b_logk = array(c(logk_init, 0), dim = 2),
      b_mid  = array(c(mid_init,  0), dim = 2),
      phi    = 5
    )
  }
}

# fit_tdt_4pl -----------------------------------------------------------------
# Fit (or only specify) a Bayesian beta-binomial 4PL thermal death time model
# using brms.  Can also serve as a workflow factory (fit = FALSE) that returns
# the formula, priors, and data without sampling — useful for prior
# predictive checks and formula inspection.
#
# @param data                 Prepared data frame from
#                             prepare_tdt_survival_data().  Must contain
#                             columns n_surv, n_total, logd, temp_c, assay_temp,
#                             duration, and survival.
# @param random_effects       Character vector of grouping variable names for
#                             random intercepts on mid.  Can be bare variable
#                             names (e.g. c("Date", "Tank")) or full
#                             lme4-style terms (e.g. "(1 | Date)").
# @param low_bounds           Numeric length-2 vector: (min, max) for the
#                             lower asymptote.  Default c(0.001, 0.499).
# @param up_bounds            Numeric length-2 vector: (min, max) for the
#                             upper asymptote.  Default c(0.501, 0.999).
# @param prior                Optional brms prior object.  If NULL (default),
#                             make_tdt_4pl_priors() builds weakly informative
#                             priors automatically.
# @param chains               Number of MCMC chains.  Default 4.
# @param cores                Number of CPU cores for parallel chains.
#                             Defaults to chains.
# @param iter                 Total MCMC iterations per chain (including
#                             warmup).  Default 4000.
# @param warmup               Warmup iterations per chain.  Default half of
#                             iter.
# @param seed                 Random seed for reproducibility.  Default 123.
# @param backend              Sampling backend: "cmdstanr" (recommended) or
#                             "rstan".  Default "cmdstanr".
# @param control              List of HMC control parameters passed to brm().
#                             Default list(adapt_delta = 0.995,
#                             max_treedepth = 15).
# @param init                 Initial values for the sampler.  If NULL
#                             (default), uses init = 0.
# @param sample_prior         Passed to brm(); "no" (default) fits the
#                             posterior; "only" draws from the prior
#                             predictive distribution.
# @param save_pars            brms save_pars() object.  Default saves all.
# @param file                 Optional file path to cache the fitted model.
# @param overwrite            Logical.  If TRUE, re-fit even when file exists.
#                             Default FALSE.
# @param fit                  Logical.  If FALSE, return the workflow object
#                             without fitting.  Default TRUE.
# @param ...                  Additional arguments passed to brm().
#
# @return A list of class "tdt_4pl_workflow" with elements:
#   $fit      brmsfit object (NULL when fit = FALSE)
#   $data     the prepared data
#   $formula  the brms formula object
#   $prior    the prior object
#   $meta     metadata list (temp_mean, bounds, random_effects, etc.)
fit_tdt_4pl <- function(data,
                        random_effects      = NULL,
                        low_bounds          = c(0.001, 0.499),
                        up_bounds           = c(0.501, 0.999),
                        prior               = NULL,
                        chains              = 4,
                        cores               = chains,
                        iter                = 4000,
                        warmup              = floor(iter / 2),
                        seed                = 123,
                        backend             = "cmdstanr",
                        control             = list(adapt_delta = 0.995,
                                                   max_treedepth = 15),
                        init                = NULL,
                        sample_prior        = "no",
                        save_pars           = brms::save_pars(all = TRUE),
                        file                = NULL,
                        overwrite           = FALSE,
                        fit                 = TRUE,
                        ...) {
  meta <- attr(data, "tdt_meta")

  if (is.null(meta)) {
    meta <- list(
      temp_mean = mean(data$assay_temp, na.rm = TRUE),
      duration_unit = "model_units",
      random_effects = random_effects
    )
  }

  formula <- make_tdt_4pl_formula(
    random_effects = random_effects,
    low_bounds = low_bounds,
    up_bounds = up_bounds
  )

  if (is.null(prior)) {
    prior <- make_tdt_4pl_priors(
      data           = data,
      random_effects = random_effects,
      low_bounds     = low_bounds,
      up_bounds      = up_bounds
    )
  }

  workflow <- list(
    fit = NULL,
    data = data,
    formula = formula,
    prior = prior,
    meta = c(
      meta,
      list(
        random_effects = random_effects,
        low_bounds = low_bounds,
        up_bounds = up_bounds
      )
    )
  )

  class(workflow) <- "tdt_4pl_workflow"

  if (!fit) {
    return(workflow)
  }

  if (!is.null(file) && file.exists(file) && !overwrite) {
    workflow$fit <- readRDS(file)
    return(workflow)
  }

  if (is.null(init)) {
    # Use init = 0: cmdstanr initialises every parameter to 0 on the
    # unconstrained scale.  This maps to sd = exp(0) = 1, z = 0, phi = 1,
    # b_* = 0 — all valid starting values.  A custom function
    # make_tdt_4pl_init() exists for data-informed starts, but it does not
    # know the Stan names for random-effect parameters (sd_1, z_1, sd_2, z_2,
    # ...) so those fall back to random draws in Uniform(-2, 2) unconstrained,
    # which can push the beta-binomial log-probability to -Inf on some chains.
    init <- 0
  }

  workflow$fit <- brms::brm(
    formula = formula,
    data = data,
    family = brms::beta_binomial(link = "identity"),
    prior = prior,
    chains = chains,
    cores = cores,
    iter = iter,
    warmup = warmup,
    seed = seed,
    backend = backend,
    control = control,
    init = init,
    sample_prior = sample_prior,
    save_pars = save_pars,
    ...
  )

  if (!is.null(file)) {
    dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
    saveRDS(workflow$fit, file)
  }

  workflow
}
```

### Prediction Grids

```{r}
# tdt_has_fit -----------------------------------------------------------------
# Convenience predicate: TRUE when workflow is a fitted tdt_4pl_workflow.
#
# @param workflow  Object returned by fit_tdt_4pl().
# @return Logical scalar.
tdt_has_fit <- function(workflow) {
  inherits(workflow, "tdt_4pl_workflow") && !is.null(workflow$fit)
}

# new_tdt_grid ----------------------------------------------------------------
# Build a prediction grid of temperature × duration combinations.
#
# @param workflow   tdt_4pl_workflow object.
# @param temps      Numeric vector of assay temperatures (°C).
# @param durations  Numeric vector of exposure durations (model units).
# @param n_total    Integer for trials column. NULL uses training data median.
# @return A tibble with one row per (temp, duration) combination.
new_tdt_grid <- function(workflow,
                         temps,
                         durations,
                         n_total = NULL) {
  data <- workflow$data
  meta <- workflow$meta

  if (is.null(n_total)) {
    n_total <- as.integer(round(stats::median(data$n_total, na.rm = TRUE)))
  }

  nd <- expand.grid(
    assay_temp = temps,
    duration   = durations,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )

  nd$n_total <- n_total
  nd$logd    <- log10(nd$duration)
  nd$temp_c  <- nd$assay_temp - meta$temp_mean

  for (re_var in tdt_random_effect_variables(meta$random_effects)) {
    if (re_var %in% names(data)) {
      if (is.factor(data[[re_var]])) {
        nd[[re_var]] <- factor(levels(data[[re_var]])[1], levels = levels(data[[re_var]]))
      } else {
        nd[[re_var]] <- data[[re_var]][1]
      }
    }
  }

  tibble::as_tibble(nd)
}

# posterior_linpred_tdt -------------------------------------------------------
# Thin wrapper around brms::posterior_linpred() with transform = TRUE so it
# returns draws on the probability scale (not the logit scale).
#
# @param workflow    Fitted tdt_4pl_workflow object.
# @param newdata     Prediction grid from new_tdt_grid() or equivalent.
# @param ndraws      Integer or NULL: number of posterior draws to sample.
#                    NULL uses all available draws.
# @param re_formula  Formula for random effects in predictions.  NA (default)
#                    marginalises over random effects (population-level
#                    prediction); NULL includes them.
#
# @return Numeric matrix [ndraws x nrow(newdata)] of survival probabilities.
posterior_linpred_tdt <- function(workflow,
                                  newdata,
                                  ndraws     = NULL,
                                  re_formula = NA) {
    if (!tdt_has_fit(workflow)) {
      stop("workflow$fit is NULL. Fit or load the model first.", call. = FALSE)
    }

    args <- list(
      object = workflow$fit,
      newdata = newdata,
      re_formula = re_formula,
      transform = TRUE
    )

    if (!is.null(ndraws)) {
      args$ndraws <- ndraws
    }

    do.call(brms::posterior_linpred, args)
}
```

### Survival Curves

```{r}
summarise_observed_survival <- function(observed) {
  observed %>%
    dplyr::group_by(assay_temp, duration) %>%
    dplyr::summarise(
      survival_mean = mean(survival, na.rm = TRUE),
      survival_se = stats::sd(survival, na.rm = TRUE) / sqrt(dplyr::n()),
      n_units = dplyr::n(),
      n_total_sum = sum(n_total, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      survival_se = dplyr::if_else(
        is.finite(survival_se),
        survival_se,
        0
      ),
      survival_lower = pmax(0, survival_mean - survival_se),
      survival_upper = pmin(1, survival_mean + survival_se)
    )
}

# predict_survival_curves -----------------------------------------------------
# Compute posterior survival-curve summaries over a grid of temperatures and
# durations.  Returns the median and 95% credible interval for each grid
# point so they can be plotted with plot_survival_curves().
#
# @param workflow         Fitted tdt_4pl_workflow object.
# @param temps            Numeric vector of temperatures (°C) to predict at.
# @param durations        Numeric vector of durations or NULL.  NULL (default)
#                         uses 250 log-spaced values spanning the training
#                         data range.
# @param ndraws           Number of posterior draws.  Default 1000.
# @param probs            Quantile probabilities for the credible interval.
#                         Default c(0.025, 0.5, 0.975).
#
# @return Named list: $summary (tibble with columns survival_lower,
#         survival_median, survival_upper), $draws (raw draw matrix),
#         $newdata (the prediction grid).
predict_survival_curves <- function(workflow,
                                    temps,
                                    durations = NULL,
                                    ndraws    = 1000,
                                    probs     = c(0.025, 0.5, 0.975)) {
  if (is.null(durations)) {
    duration_range <- range(workflow$data$duration, na.rm = TRUE)
    durations <- exp(seq(log(duration_range[1]), log(duration_range[2]), length.out = 250))
  }

  nd <- new_tdt_grid(
    workflow = workflow,
    temps = temps,
    durations = durations
  )

  pred <- posterior_linpred_tdt(workflow, nd, ndraws = ndraws, re_formula = NA)
  q <- apply(pred, 2, tdt_quantile, probs = probs)

  nd$survival_lower <- q[1, ]
  nd$survival_median <- q[2, ]
  nd$survival_upper <- q[3, ]

  list(summary = nd, draws = pred, newdata = nd)
}

plot_survival_curves <- function(pred,
                                 observed = NULL,
                                 observed_summary = TRUE) {
  df <- pred$summary %>%
    dplyr::mutate(temp_f = factor(assay_temp, levels = sort(unique(assay_temp))))

  p <- ggplot(
    df,
    aes(
      x = duration,
      y = survival_median,
      colour = temp_f,
      fill = temp_f,
      group = temp_f
    )
  ) +
    geom_ribbon(aes(ymin = survival_lower, ymax = survival_upper),
                alpha = 0.18, colour = NA) +
    geom_line(linewidth = 0.9) +
    scale_y_continuous(limits = c(0, 1)) +
    scale_colour_viridis_d(option = "D", end = 0.88, name = "Assay temp (°C)") +
    scale_fill_viridis_d(option = "D", end = 0.88, name = "Assay temp (°C)") +
    labs(
      x = "Exposure duration (h)",
      y = "Predicted survival"
    ) +
    theme_tdt +
    theme(legend.position = "right")

  if (!is.null(observed)) {
    if (observed_summary) {
      obs <- summarise_observed_survival(observed) %>%
        dplyr::mutate(temp_f = factor(assay_temp, levels = levels(df$temp_f)))

      p <- p +
        geom_errorbar(
          data = obs,
          aes(
            x = duration,
            ymin = survival_lower,
            ymax = survival_upper,
            colour = temp_f
          ),
          inherit.aes = FALSE,
          width = 0,
          alpha = 0.8,
          linewidth = 0.45
        ) +
        geom_point(
          data = obs,
          aes(x = duration, y = survival_mean, fill = temp_f),
          inherit.aes = FALSE,
          shape = 21,
          colour = "black",
          alpha = 0.9,
          size = 2.2
        )
    } else {
      obs <- observed %>%
        dplyr::mutate(temp_f = factor(assay_temp, levels = levels(df$temp_f)))

      p <- p +
        geom_point(
          data = obs,
          aes(x = duration, y = survival, fill = temp_f),
          inherit.aes = FALSE,
          shape = 21,
          colour = "black",
          alpha = 0.55,
          size = 1.6
        )
    }
  }

  p
}
```

### Threshold Extraction, TDT Curves, And Classical Summaries

These functions replace the hard-coded `species` helpers in `tdt_problems.qmd`. They work by numerically inverting posterior predictions, so they also work for models with random effects and temperature-varying steepness. The tradeoff is that the duration grid must be broad enough to contain the threshold.

```{r}
threshold_x_by_draw <- function(pred_mat, x, target) {
  ord <- order(x)
  x <- x[ord]
  pred_mat <- pred_mat[, ord, drop = FALSE]

  out <- rep(NA_real_, nrow(pred_mat))

  for (i in seq_len(nrow(pred_mat))) {
    y <- as.numeric(pred_mat[i, ])

    if (all(is.na(y))) {
      next
    }

    if (min(y, na.rm = TRUE) > target || max(y, na.rm = TRUE) < target) {
      next
    }

    exact <- which(abs(y - target) < 1e-12)
    if (length(exact) > 0) {
      out[i] <- x[exact[1]]
      next
    }

    crossing <- which((y[-length(y)] - target) * (y[-1] - target) <= 0)
    if (length(crossing) == 0) {
      out[i] <- x[which.min(abs(y - target))]
      next
    }

    j <- crossing[1]
    x1 <- x[j]
    x2 <- x[j + 1]
    y1 <- y[j]
    y2 <- y[j + 1]

    if (!is.finite(y1) || !is.finite(y2) || abs(y2 - y1) < 1e-12) {
      out[i] <- mean(c(x1, x2))
    } else {
      out[i] <- x1 + (target - y1) * (x2 - x1) / (y2 - y1)
    }
  }

  out
}

# derive_ltx_curve ------------------------------------------------------------
# Derive the time-to-X%-survival (LTx) curve as a function of temperature by
# numerically inverting posterior survival predictions.  For each posterior
# draw and each temperature, the function finds the duration at which the
# predicted survival curve crosses the target_survival threshold using linear
# interpolation over a fine duration grid.
#
# @param workflow         Fitted tdt_4pl_workflow object.
# @param temp_grid        Numeric vector of temperatures (°C) to evaluate.
# @param duration_grid    Numeric vector of durations (model units) over which
#                         to search for the threshold.  NULL (default) uses
#                         350 log-spaced values from 1/5 × min to 5 × max of
#                         the training durations.
# @param target_survival  Numeric scalar: the survival probability whose
#                         crossing time is sought (e.g. 0.5 for LT50).
# @param ndraws           Number of posterior draws.  Default 1000.
# @param probs            Quantile probabilities for the summary.
#                         Default c(0.025, 0.5, 0.975).
# @param time_multiplier  Multiply model-unit durations by this factor for
#                         output (e.g. 60 to convert hours to minutes).
#                         Default 60.
# @param output_time_unit String label for the output unit.  Default "min".
#
# @return Named list: $draws (per-draw thresholds), $summary (quantile
#         summary by temperature), $target_survival,
#         $time_multiplier, $output_time_unit.
derive_ltx_curve <- function(workflow,
                             temp_grid,
                             duration_grid    = NULL,
                             target_survival  = 0.5,
                             ndraws           = 1000,
                             probs            = c(0.025, 0.5, 0.975),
                             time_multiplier  = 60,
                             output_time_unit = "min") {
  if (is.null(duration_grid)) {
    duration_range <- range(workflow$data$duration, na.rm = TRUE)
    duration_grid <- 10^seq(
      log10(duration_range[1] / 5),
      log10(duration_range[2] * 5),
      length.out = 350
    )
  }

  nd <- new_tdt_grid(
    workflow = workflow,
    temps = temp_grid,
    durations = duration_grid
  )

  pred <- posterior_linpred_tdt(workflow, nd, ndraws = ndraws, re_formula = NA)

  temp_grid_unique <- sort(unique(nd$assay_temp))
  draw_list <- vector("list", length(temp_grid_unique))

  for (i in seq_along(temp_grid_unique)) {
    temp_i <- temp_grid_unique[i]
    idx <- nd$assay_temp == temp_i

    threshold_duration <- threshold_x_by_draw(
      pred_mat = pred[, idx, drop = FALSE],
      x        = nd$duration[idx],
      target   = as.numeric(target_survival)
    )

    draw_list[[i]] <- data.frame(
      .draw                  = seq_along(threshold_duration),
      assay_temp             = temp_i,
      target_survival        = as.numeric(target_survival),
      duration_model_units   = threshold_duration,
      duration_output_units  = threshold_duration * time_multiplier
    )
  }

  draws <- dplyr::bind_rows(draw_list) %>%
    dplyr::filter(is.finite(duration_model_units), duration_model_units > 0)

  summary <- draws %>%
    dplyr::group_by(target_survival, assay_temp) %>%
    dplyr::summarise(
      duration_lower  = stats::quantile(duration_output_units, probs[1], na.rm = TRUE),
      duration_median = stats::quantile(duration_output_units, probs[2], na.rm = TRUE),
      duration_upper  = stats::quantile(duration_output_units, probs[3], na.rm = TRUE),
      .groups = "drop"
    )

  list(
    draws            = draws,
    summary          = summary,
    target_survival  = target_survival,
    time_multiplier  = time_multiplier,
    output_time_unit = output_time_unit
  )
}

LT50_T_curve <- function(workflow, T_grid, ...) {
  derive_ltx_curve(
    workflow = workflow,
    temp_grid = T_grid,
    target_survival = 0.5,
    ...
  )
}

bind_ltx_curves <- function(..., output_time_unit = NULL) {
  curves <- list(...)
  if (length(curves) == 0) {
    stop("Supply at least one LTx curve.", call. = FALSE)
  }

  list(
    draws = dplyr::bind_rows(lapply(curves, `[[`, "draws")),
    summary = dplyr::bind_rows(lapply(curves, `[[`, "summary")),
    target_survival = unique(unlist(lapply(curves, `[[`, "target_survival"))),
    time_multiplier = curves[[1]]$time_multiplier,
    output_time_unit = output_time_unit %||% curves[[1]]$output_time_unit
  )
}

# derive_tdt_parameters -------------------------------------------------------
# Fit a log-linear regression of log10(LTx time) ~ temperature to each
# posterior draw, then summarise the classical TDT parameters:
#   z      = thermal tolerance range (°C per decade of time; -1 / slope)
#   CTmax  = temperature where the LTx curve crosses t_ref minutes
#
# This is the Bayesian analogue of fitting a straight line to log-transformed
# time-to-death data: it applies the traditional TDT summary to each draw,
# producing a full posterior distribution of z and CTmax.
#
# @param ltx_curve   Output of derive_ltx_curve() or LT50_T_curve().
# @param t_ref       Reference time (in output_time_unit of ltx_curve) for
#                    computing CTmax.  Default 1 (minute).
# @param min_points  Minimum number of temperature-duration pairs a draw
#                    must have to be included in the regression.  Default 5.
#
# @return Named list: $draws (per-draw slope, z, CTmax, R2), $summary
#         (quantile summaries by target_survival), $t_ref.
derive_tdt_parameters <- function(ltx_curve,
                                  t_ref      = 1,
                                  min_points = 5) {
  draws <- ltx_curve$draws %>%
    dplyr::filter(is.finite(duration_output_units), duration_output_units > 0) %>%
    dplyr::mutate(log10_duration = log10(duration_output_units))

  params <- draws %>%
    dplyr::group_by(target_survival, .draw) %>%
    dplyr::filter(dplyr::n() >= min_points) %>%
    dplyr::group_modify(function(.x, .y) {
      fit <- stats::lm(log10_duration ~ assay_temp, data = .x)
      co <- stats::coef(fit)

      data.frame(
        intercept = unname(co[1]),
        slope_T = unname(co[2]),
        z_decade = -1 / unname(co[2]),
        CTmax_ref = (log10(t_ref) - unname(co[1])) / unname(co[2]),
        r_squared = summary(fit)$r.squared
      )
    }) %>%
    dplyr::ungroup()

  summary <- params %>%
    dplyr::group_by(target_survival) %>%
    dplyr::summarise(
      slope_median = median(slope_T, na.rm = TRUE),
      slope_lower = stats::quantile(slope_T, 0.025, na.rm = TRUE),
      slope_upper = stats::quantile(slope_T, 0.975, na.rm = TRUE),
      z_median = median(z_decade, na.rm = TRUE),
      z_lower = stats::quantile(z_decade, 0.025, na.rm = TRUE),
      z_upper = stats::quantile(z_decade, 0.975, na.rm = TRUE),
      CTmax_ref_median = median(CTmax_ref, na.rm = TRUE),
      CTmax_ref_lower = stats::quantile(CTmax_ref, 0.025, na.rm = TRUE),
      CTmax_ref_upper = stats::quantile(CTmax_ref, 0.975, na.rm = TRUE),
      r2_median = median(r_squared, na.rm = TRUE),
      .groups = "drop"
    )

  list(draws = params, summary = summary, t_ref = t_ref)
}

extract_tdt_params <- function(workflow,
                               T_grid,
                               t_ref_min = 1,
                               ...) {
  ltx <- LT50_T_curve(
    workflow = workflow,
    T_grid = T_grid,
    output_time_unit = "min",
    time_multiplier = 60,
    ...
  )

  derive_tdt_parameters(ltx, t_ref = t_ref_min)
}

plot_ltx_curve <- function(ltx_curve, time_unit = "h") {
  df <- ltx_curve$summary %>%
    dplyr::mutate(
      log10_duration_median = log10(duration_median),
      log10_duration_lower  = log10(duration_lower),
      log10_duration_upper  = log10(duration_upper),
      threshold = factor(
        paste0(round(100 * target_survival), "% survival"),
        levels = paste0(
          round(100 * sort(unique(target_survival), decreasing = TRUE)),
          "% survival"
        )
      )
    )

  colour_vals <- c(
    "75% survival" = "#4DAC26",
    "50% survival" = "#146C7C",
    "25% survival" = "#C76D37"
  )

  base_aes <- aes(x = assay_temp, colour = threshold, fill = threshold)

  p1 <- ggplot(df, base_aes) +
    geom_ribbon(aes(y = duration_median, ymin = duration_lower, ymax = duration_upper),
                alpha = 0.18, colour = NA) +
    geom_line(aes(y = duration_median), linewidth = 1.05) +
    scale_colour_manual(values = colour_vals, drop = FALSE, name = "Survival threshold") +
    scale_fill_manual(  values = colour_vals, drop = FALSE, name = "Survival threshold") +
    labs(x = "Assay temperature (°C)",
         y = paste0("Time to death (", time_unit, ")")) +
    theme_tdt +
    theme(legend.position = "right")

  p2 <- ggplot(df, base_aes) +
    geom_ribbon(aes(y = log10_duration_median,
                    ymin = log10_duration_lower, ymax = log10_duration_upper),
                alpha = 0.18, colour = NA) +
    geom_line(aes(y = log10_duration_median), linewidth = 1.05) +
    scale_colour_manual(values = colour_vals, drop = FALSE, name = "Survival threshold") +
    scale_fill_manual(  values = colour_vals, drop = FALSE, name = "Survival threshold") +
    labs(x = "Assay temperature (°C)",
         y = paste0("log₁₀ time to death (", time_unit, ")")) +
    theme_tdt +
    theme(legend.position = "right")

  patchwork::wrap_plots(p1, p2, ncol = 1, guides = "collect") &
    theme(legend.position = "right")
}
```

### TDT Landscape

```{r}
derive_tdt_landscape <- function(workflow,
                                 temp_grid,
                                 duration_grid,
                                 ndraws = 1000,
                                 probs = c(0.025, 0.5, 0.975)) {
  predict_survival_curves(
    workflow = workflow,
    temps = temp_grid,
    durations = duration_grid,
    ndraws = ndraws,
    probs = probs
  )
}

plot_tdt_landscape <- function(landscape,
                               observed = NULL) {
  df <- landscape$summary

  p <- ggplot(df, aes(x = assay_temp, y = duration, fill = survival_median)) +
    geom_raster(interpolate = TRUE) +
    geom_contour(aes(z = survival_median),
                 breaks = c(0.25, 0.5, 0.75),
                 colour = "white",
                 alpha = 0.7) +
    scale_fill_viridis_c(limits = c(0, 1), option = "inferno") +
    labs(
      x = "Assay temperature (°C)",
      y = "Exposure duration (h)",
      fill = "Survival"
    ) +
    theme_tdt

  if (!is.null(observed)) {
    p <- p +
      geom_point(
        data = observed,
        aes(x = assay_temp, y = duration, fill = survival),
        inherit.aes = FALSE,
        shape = 21,
        colour = "white",
        size = 2,
        stroke = 0.2
      )
  }

  p
}
```

### Standard 4PL Parameter Extraction For Dynamic Injury

The functions below use the analytic 4PL-to-injury mapping from `tdt_problems.qmd` and the later shrimp dynamic section. Both the midpoint (`mid`) and log-steepness (`logk`) are linear functions of `temp_c`, so the steepness of the survival curve varies with temperature. The extracted parameters `logk_int` and `logk_temp` replace the single `k` used in earlier versions.

```{r}
first_existing_column <- function(df, candidates) {
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) == 0) {
    return(NULL)
  }
  hit[1]
}

# extract_standard_4pl_pars ---------------------------------------------------
# Extract per-draw 4PL parameters from the fitted model's posterior.
# All four sub-models are linear in temp_c, so raw (lraw, uraw) and log-scale
# (logk) intercepts and slopes are returned; low/up/k are temperature-dependent
# and computed from these in downstream functions.
#
# @param workflow  Fitted tdt_4pl_workflow object.
# @return A data frame with columns .draw, lraw_int, lraw_temp, uraw_int,
#         uraw_temp, logk_int, logk_temp, mid_int, mid_temp.
extract_standard_4pl_pars <- function(workflow) {
  if (!tdt_has_fit(workflow)) {
    stop("workflow$fit is NULL. Fit or load the model first.", call. = FALSE)
  }

  dr <- posterior::as_draws_df(workflow$fit) %>%
    as.data.frame()

  get_par <- function(name) {
    if (name %in% names(dr)) dr[[name]] else rep(0, nrow(dr))
  }

  data.frame(
    .draw     = seq_len(nrow(dr)),
    lraw_int  = get_par("b_lraw_Intercept"),
    lraw_temp = get_par("b_lraw_temp_c"),
    uraw_int  = get_par("b_uraw_Intercept"),
    uraw_temp = get_par("b_uraw_temp_c"),
    logk_int  = get_par("b_logk_Intercept"),
    logk_temp = get_par("b_logk_temp_c"),
    mid_int   = get_par("b_mid_Intercept"),
    mid_temp  = get_par("b_mid_temp_c")
  ) %>%
    dplyr::filter(
      is.finite(lraw_int), is.finite(lraw_temp),
      is.finite(uraw_int), is.finite(uraw_temp),
      is.finite(logk_int), is.finite(logk_temp),
      is.finite(mid_int),  is.finite(mid_temp)
    )
}

derive_relative_survival_targets <- function(workflow,
                                              mortality_fraction = 0.5) {
  if (mortality_fraction <= 0 || mortality_fraction >= 1) {
    stop("mortality_fraction must be between 0 and 1.", call. = FALSE)
  }

  pars       <- extract_standard_4pl_pars(workflow)
  low_bounds <- workflow$meta$low_bounds
  up_bounds  <- workflow$meta$up_bounds

  # Evaluate asymptotes at the centering temperature (temp_c = 0)
  low_med <- median(low_bounds[1] + plogis(pars$lraw_int) * diff(low_bounds), na.rm = TRUE)
  up_med  <- median(up_bounds[1]  + plogis(pars$uraw_int) * diff(up_bounds),  na.rm = TRUE)
  up_med - mortality_fraction * (up_med - low_med)
}

time_to_surv_threshold_4pl <- function(temp,
                                       survival_target,
                                       lraw_int,
                                       lraw_temp,
                                       uraw_int,
                                       uraw_temp,
                                       logk_int,
                                       logk_temp,
                                       mid_int,
                                       mid_temp,
                                       temp_mean,
                                       low_bounds = c(0.001, 0.499),
                                       up_bounds  = c(0.501, 0.999)) {
  temp_c <- temp - temp_mean
  low <- low_bounds[1] + plogis(lraw_int + lraw_temp * temp_c) * diff(low_bounds)
  up  <- up_bounds[1]  + plogis(uraw_int + uraw_temp * temp_c) * diff(up_bounds)
  k   <- exp(logk_int + logk_temp * temp_c)
  mid <- mid_int + mid_temp * temp_c
  valid <- survival_target > low & survival_target < up
  out <- rep(NA_real_, length(temp))
  out[valid] <- 10^(
    mid[valid] +
      log((up[valid] - survival_target) / (survival_target - low[valid])) /
        k[valid]
  )
  out
}

survival_from_latent_dose <- function(dose,
                                      low,
                                      up,
                                      k,
                                      target_survival = 0.5) {
  dose_use <- pmax(dose, 1e-12)
  c_target <- log((up - target_survival) / (target_survival - low)) / k

  low + (up - low) /
    (1 + exp(k * (log10(dose_use) + c_target)))
}

damage_rate_curve <- function(pars,
                              temp_mean,
                              temp_grid,
                              target_survival = 0.5,
                              n_draws = 1000,
                              time_multiplier_to_minutes = 60,
                              low_bounds = c(0.001, 0.499),
                              up_bounds  = c(0.501, 0.999)) {
  pars_use <- pars %>%
    dplyr::slice_sample(n = min(n_draws, nrow(pars)))

  draws <- dplyr::bind_rows(lapply(seq_len(nrow(pars_use)), function(i) {
    t_target <- time_to_surv_threshold_4pl(
      temp = temp_grid,
      survival_target = target_survival,
      lraw_int  = pars_use$lraw_int[i],
      lraw_temp = pars_use$lraw_temp[i],
      uraw_int  = pars_use$uraw_int[i],
      uraw_temp = pars_use$uraw_temp[i],
      logk_int  = pars_use$logk_int[i],
      logk_temp = pars_use$logk_temp[i],
      mid_int   = pars_use$mid_int[i],
      mid_temp  = pars_use$mid_temp[i],
      temp_mean = temp_mean,
      low_bounds = low_bounds,
      up_bounds  = up_bounds
    )

    data.frame(
      .draw = pars_use$.draw[i],
      temp = temp_grid,
      duration_model_units = t_target,
      damage_rate_per_model_unit = 1 / t_target,
      damage_rate_pct_min = 100 / (time_multiplier_to_minutes * t_target)
    )
  })) %>%
    dplyr::filter(is.finite(damage_rate_per_model_unit))

  summary <- draws %>%
    dplyr::group_by(temp) %>%
    dplyr::summarise(
      damage_rate_median = median(damage_rate_per_model_unit, na.rm = TRUE),
      damage_rate_lower = stats::quantile(damage_rate_per_model_unit, 0.025, na.rm = TRUE),
      damage_rate_upper = stats::quantile(damage_rate_per_model_unit, 0.975, na.rm = TRUE),
      damage_rate_pct_min_median = median(damage_rate_pct_min, na.rm = TRUE),
      damage_rate_pct_min_lower = stats::quantile(damage_rate_pct_min, 0.025, na.rm = TRUE),
      damage_rate_pct_min_upper = stats::quantile(damage_rate_pct_min, 0.975, na.rm = TRUE),
      .groups = "drop"
    )

  list(draws = draws, summary = summary)
}
```

### Fluctuating Temperatures And Repair

Repair is optional. If supplied, the default repair function follows the same Sharpe-Schoolfield style used in `R/bayesian_TDT_shrimp.qmd`, with all temperatures converted to Kelvin inside the function.

```{r}
repair_rate_schoolfield_pct_min <- function(temp,
                                            TA,
                                            TAL,
                                            TAH,
                                            TL,
                                            TH,
                                            TREF,
                                            kdot_ref_pct_min) {
  temp_k <- temp + 273.15

  out <- exp(TA * (1 / TREF - 1 / temp_k)) /
    (
      1 +
        exp(TAL * (1 / temp_k - 1 / TL)) +
        exp(TAH * (1 / TH - 1 / temp_k))
    ) *
    kdot_ref_pct_min

  out[!is.finite(out)] <- 0
  pmax(out, 0)
}

make_temperature_scenarios <- function(start       = as.POSIXct("2026-07-01 00:00:00", tz = "UTC"),
                                       n           = 96,
                                       dt_mins     = 60,
                                       baseline    = 20,
                                       single_temp = 31.5,
                                       mild_temp   = 29,
                                       severe_temp = 31) {
  # dt_mins = 60 → hourly data; each spike lasts exactly 1 hour (1 row)
  # index i corresponds to hour i-1 (index 1 = hour 0, index 25 = hour 24, etc.)
  datetime <- start + seq(0, by = dt_mins * 60, length.out = n)

  no_heat <- data.frame(datetime = datetime, temp = rep(baseline, n))

  # Single spike: one 1-hr spike at 31.5 °C at hour 24
  single_spike <- no_heat
  single_spike$temp[25] <- single_temp   # hour 24

  # Mild repeated: three 1-hr spikes at 29 °C, 12 h apart starting at hour 24
  mild_repeated <- no_heat
  for (h in c(24, 36, 48)) {
    mild_repeated$temp[h + 1] <- mild_temp
  }

  # Severe repeated: three 1-hr spikes at 31 °C, 12 h apart starting at hour 24
  severe_repeated <- no_heat
  for (h in c(24, 36, 48)) {
    severe_repeated$temp[h + 1] <- severe_temp
  }

  # Mixed: two mild spikes (29 °C) then one severe spike (31 °C)
  mixed_severity <- no_heat
  mixed_severity$temp[25] <- mild_temp    # hour 24
  mixed_severity$temp[37] <- mild_temp    # hour 36
  mixed_severity$temp[49] <- severe_temp  # hour 48

  # Return in simple → complex order; this order is used for factor levels downstream
  list(
    no_heat         = no_heat,
    single_spike    = single_spike,
    mild_repeated   = mild_repeated,
    severe_repeated = severe_repeated,
    mixed_severity  = mixed_severity
  )
}

# predict_survival_fluctuating ------------------------------------------------
# Predict cumulative survival under a fluctuating temperature time series
# using an Eulerian damage-accumulation model.  For each posterior draw and
# each time step, the instantaneous damage rate is 1 / LTx_duration(temp).
# Damage increments are integrated with the trapezoid rule; optional repair
# reduces the accumulated dose.
#
# @param temp_df                  Data frame with columns `datetime` (POSIXct)
#                                 and `temp` (°C), in time order.  Requires
#                                 >= 2 rows.
# @param pars                     Data frame of 4PL parameters from
#                                 extract_standard_4pl_pars().
# @param temp_mean                Numeric: the temperature centering constant
#                                 from the fitted model (workflow$meta$temp_mean).
# @param target_survival          Numeric: the survival threshold whose
#                                 time-to-threshold defines the damage rate.
#                                 Default 0.5.
# @param n_draws                  Number of posterior draws to use.
#                                 Default 1000.
# @param repair                   Logical: include thermal repair?
#                                 Default FALSE.
# @param repair_pars              Named list of Sharpe-Schoolfield repair
#                                 parameters (TA, TAL, TAH, TL, TH, TREF).
#                                 Required when repair = TRUE.
# @param kdot_ref_pct_min         Repair rate at TREF in percent-dose per
#                                 minute.  Default 0.
# @param repair_scales_with_survival Logical: scale repair rate by current
#                                 survival fraction (TRUE, default) so repair
#                                 is zero when all organisms are dead.
# @param save_draws               Logical: include full draw-level trajectories
#                                 in the return value?  Default FALSE.
#
# @return Named list: $summary (posterior median and 95% CI for latent dose,
#         survival, and mortality at each time step), and optionally $draws.
predict_survival_fluctuating <- function(temp_df,
                                         pars,
                                         temp_mean,
                                         target_survival             = 0.5,
                                         n_draws                     = 1000,
                                         repair                      = FALSE,
                                         repair_pars                 = NULL,
                                         kdot_ref_pct_min            = 0,
                                         repair_scales_with_survival = TRUE,
                                         irreversible_mortality      = TRUE,
                                         save_draws                  = FALSE,
                                         low_bounds                  = c(0.001, 0.499),
                                         up_bounds                   = c(0.501, 0.999)) {
  temp_df <- temp_df %>%
    dplyr::arrange(datetime)

  if (nrow(temp_df) < 2) {
    stop("temp_df must contain at least 2 rows.", call. = FALSE)
  }

  if (nrow(pars) == 0) {
    stop("pars contains no rows.", call. = FALSE)
  }

  pars_use <- pars %>%
    dplyr::slice_sample(n = min(n_draws, nrow(pars)))

  dt_h <- as.numeric(diff(temp_df$datetime), units = "hours")
  temp0 <- temp_df$temp[-nrow(temp_df)]
  temp1 <- temp_df$temp[-1]

  pred_list <- vector("list", nrow(pars_use))

  for (i in seq_len(nrow(pars_use))) {
    # Asymptotes at the centering temperature (temp_c = 0) for the latent dose model
    low_ref <- low_bounds[1] + plogis(pars_use$lraw_int[i]) * diff(low_bounds)
    up_ref  <- up_bounds[1]  + plogis(pars_use$uraw_int[i]) * diff(up_bounds)

    t0 <- time_to_surv_threshold_4pl(
      temp = temp0,
      survival_target = target_survival,
      lraw_int  = pars_use$lraw_int[i],
      lraw_temp = pars_use$lraw_temp[i],
      uraw_int  = pars_use$uraw_int[i],
      uraw_temp = pars_use$uraw_temp[i],
      logk_int  = pars_use$logk_int[i],
      logk_temp = pars_use$logk_temp[i],
      mid_int   = pars_use$mid_int[i],
      mid_temp  = pars_use$mid_temp[i],
      temp_mean = temp_mean,
      low_bounds = low_bounds,
      up_bounds  = up_bounds
    )

    t1 <- time_to_surv_threshold_4pl(
      temp = temp1,
      survival_target = target_survival,
      lraw_int  = pars_use$lraw_int[i],
      lraw_temp = pars_use$lraw_temp[i],
      uraw_int  = pars_use$uraw_int[i],
      uraw_temp = pars_use$uraw_temp[i],
      logk_int  = pars_use$logk_int[i],
      logk_temp = pars_use$logk_temp[i],
      mid_int   = pars_use$mid_int[i],
      mid_temp  = pars_use$mid_temp[i],
      temp_mean = temp_mean,
      low_bounds = low_bounds,
      up_bounds  = up_bounds
    )

    damage_rate0 <- 1 / t0
    damage_rate1 <- 1 / t1
    damage_rate0[!is.finite(damage_rate0)] <- 0
    damage_rate1[!is.finite(damage_rate1)] <- 0

    if (repair) {
      if (is.null(repair_pars)) {
        stop("Supply repair_pars when repair = TRUE.", call. = FALSE)
      }

      repair0_pct_min <- repair_rate_schoolfield_pct_min(
        temp = temp0,
        TA = repair_pars$TA,
        TAL = repair_pars$TAL,
        TAH = repair_pars$TAH,
        TL = repair_pars$TL,
        TH = repair_pars$TH,
        TREF = repair_pars$TREF,
        kdot_ref_pct_min = kdot_ref_pct_min
      )

      repair1_pct_min <- repair_rate_schoolfield_pct_min(
        temp = temp1,
        TA = repair_pars$TA,
        TAL = repair_pars$TAL,
        TAH = repair_pars$TAH,
        TL = repair_pars$TL,
        TH = repair_pars$TH,
        TREF = repair_pars$TREF,
        kdot_ref_pct_min = kdot_ref_pct_min
      )

      repair_rate0 <- repair0_pct_min / 100 * 60
      repair_rate1 <- repair1_pct_min / 100 * 60
    } else {
      repair_rate0 <- rep(0, length(dt_h))
      repair_rate1 <- rep(0, length(dt_h))
    }

    dose <- numeric(nrow(temp_df))
    survival <- numeric(nrow(temp_df))
    survival_current_dose <- numeric(nrow(temp_df))
    dose[1] <- 0
    survival[1] <- up_ref
    survival_current_dose[1] <- up_ref

    for (tt in seq_along(dt_h)) {
      damage_inc <- dt_h[tt] * (damage_rate0[tt] + damage_rate1[tt]) / 2
      repair_inc <- dt_h[tt] * (repair_rate0[tt] + repair_rate1[tt]) / 2

      if (repair_scales_with_survival) {
        repair_inc <- repair_inc * survival[tt] / up_ref
      }

      dose_new <- max(0, dose[tt] + damage_inc - repair_inc)
      k_ref <- exp(pars_use$logk_int[i])
      survival_new <- survival_from_latent_dose(
        dose = dose_new,
        low = low_ref,
        up = up_ref,
        k = k_ref,
        target_survival = target_survival
      )

      dose[tt + 1] <- dose_new
      survival_current_dose[tt + 1] <- survival_new
      survival[tt + 1] <- if (irreversible_mortality) {
        min(survival[tt], survival_new)
      } else {
        survival_new
      }
    }

    pred_list[[i]] <- data.frame(
      .draw = pars_use$.draw[i],
      datetime = temp_df$datetime,
      temp = temp_df$temp,
      latent_dose = dose,
      survival = survival,
      survival_current_dose = survival_current_dose,
      mortality = 1 - survival
    )
  }

  draws <- dplyr::bind_rows(pred_list)

  summary <- draws %>%
    dplyr::group_by(datetime, temp) %>%
    dplyr::summarise(
      latent_dose_median = median(latent_dose, na.rm = TRUE),
      latent_dose_lower = stats::quantile(latent_dose, 0.025, na.rm = TRUE),
      latent_dose_upper = stats::quantile(latent_dose, 0.975, na.rm = TRUE),
      survival_median = median(survival, na.rm = TRUE),
      survival_lower = stats::quantile(survival, 0.025, na.rm = TRUE),
      survival_upper = stats::quantile(survival, 0.975, na.rm = TRUE),
      survival_current_median = median(survival_current_dose, na.rm = TRUE),
      survival_current_lower = stats::quantile(survival_current_dose, 0.025, na.rm = TRUE),
      survival_current_upper = stats::quantile(survival_current_dose, 0.975, na.rm = TRUE),
      mortality_median = median(mortality, na.rm = TRUE),
      mortality_lower = stats::quantile(mortality, 0.025, na.rm = TRUE),
      mortality_upper = stats::quantile(mortality, 0.975, na.rm = TRUE),
      .groups = "drop"
    )

  if (save_draws) {
    list(summary = summary, draws = draws)
  } else {
    list(summary = summary)
  }
}

plot_fluctuating_survival <- function(pred) {
  ggplot(pred$summary, aes(x = datetime, y = survival_median)) +
    geom_ribbon(aes(ymin = survival_lower, ymax = survival_upper),
                fill = "grey70", alpha = 0.25) +
    geom_line(linewidth = 0.9) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = NULL, y = "Predicted survival") +
    theme_tdt
}

plot_fluctuating_injury <- function(pred) {
  ggplot(pred$summary, aes(x = datetime, y = latent_dose_median)) +
    geom_ribbon(aes(ymin = latent_dose_lower, ymax = latent_dose_upper),
                fill = "grey70", alpha = 0.25) +
    geom_hline(yintercept = 1, linetype = 2) +
    geom_line(linewidth = 0.9) +
    labs(x = NULL, y = "Latent injury dose") +
    theme_tdt
}

plot_temperature_scenarios <- function(temp_scenarios) {
  scenario_levels_canonical <- c(
    "no heat", "single spike", "mild repeated", "severe repeated", "mixed severity"
  )
  labels_present <- gsub("_", " ", names(temp_scenarios))
  ordered_labels <- scenario_levels_canonical[scenario_levels_canonical %in% labels_present]

  df <- dplyr::bind_rows(
    lapply(names(temp_scenarios), function(nm) {
      temp_scenarios[[nm]] %>%
        dplyr::mutate(
          scenario = nm,
          time_h = as.numeric(difftime(datetime, min(datetime), units = "hours"))
        )
    })
  ) %>%
    dplyr::mutate(
      scenario = factor(gsub("_", " ", scenario), levels = ordered_labels)
    )

  ggplot(df, aes(x = time_h, y = temp)) +
    geom_line(linewidth = 0.9, colour = "#146C7C") +
    facet_wrap(~ scenario, ncol = 1) +
    labs(x = "Time since start (h)", y = "Temperature (°C)") +
    theme_tdt +
    theme(legend.position = "none")
}

bind_fluctuating_predictions <- function(predictions) {
  # Canonical simple-to-complex ordering for facet panels
  scenario_levels_canonical <- c(
    "no heat", "single spike", "mild repeated", "severe repeated", "mixed severity"
  )

  raw <- dplyr::bind_rows(
    lapply(names(predictions), function(nm) {
      parts <- strsplit(nm, "__", fixed = TRUE)[[1]]
      predictions[[nm]]$summary %>%
        dplyr::mutate(
          scenario     = gsub("_", " ", parts[1]),
          repair_level = parts[2],
          time_h       = as.numeric(difftime(datetime, min(datetime), units = "hours"))
        )
    })
  )

  present_levels <- scenario_levels_canonical[
    scenario_levels_canonical %in% unique(raw$scenario)
  ]

  raw %>%
    dplyr::mutate(
      scenario = factor(scenario, levels = present_levels),
      repair_level = factor(
        gsub("_", " ", repair_level),
        levels = c("no repair", "moderate repair", "high repair")
      )
    )
}

plot_fluctuating_metric <- function(predictions,
                                    metric = c("latent_dose", "survival", "mortality")) {
  metric <- match.arg(metric)
  df <- bind_fluctuating_predictions(predictions)

  cols <- switch(
    metric,
    latent_dose = c("latent_dose_lower", "latent_dose_median", "latent_dose_upper"),
    survival = c("survival_lower", "survival_median", "survival_upper"),
    mortality = c("mortality_lower", "mortality_median", "mortality_upper")
  )

  y_lab <- switch(
    metric,
    latent_dose = "Latent injury dose (LT50-dose units)",
    survival = "Predicted survival",
    mortality = "Predicted mortality"
  )

  p <- ggplot(
    df,
    aes(
      x = time_h,
      y = .data[[cols[2]]],
      colour = repair_level,
      fill = repair_level
    )
  ) +
    geom_ribbon(
      aes(ymin = .data[[cols[1]]], ymax = .data[[cols[3]]]),
      alpha = 0.12,
      colour = NA
    ) +
    geom_line(linewidth = 0.9) +
    facet_wrap(~ scenario, ncol = 1) +
    scale_colour_manual(
      values = c(
        "no repair" = "#2F2F2F",
        "moderate repair" = "#146C7C",
        "high repair" = "#C76D37"
      ),
      drop = FALSE
    ) +
    scale_fill_manual(
      values = c(
        "no repair" = "#2F2F2F",
        "moderate repair" = "#146C7C",
        "high repair" = "#C76D37"
      ),
      drop = FALSE
    ) +
    labs(x = "Time since start (h)", y = y_lab, colour = NULL, fill = NULL) +
    theme_tdt

  if (metric %in% c("survival", "mortality")) {
    p <- p + scale_y_continuous(limits = c(0, 1))
  }

  if (metric == "latent_dose") {
    p <- p + geom_hline(yintercept = 1, linetype = 2, colour = "grey35")
  }

  p
}

summarise_fluctuating_predictions <- function(temp_scenarios, predictions) {
  dplyr::bind_rows(lapply(names(predictions), function(nm) {
    parts <- strsplit(nm, "__", fixed = TRUE)[[1]]
    scenario <- parts[1]
    repair_level <- parts[2]
    pred <- predictions[[nm]]$summary
    final <- pred[nrow(pred), ]

    data.frame(
      Scenario = gsub("_", " ", scenario),
      Repair = gsub("_", " ", repair_level),
      `Peak temp (°C)` = max(temp_scenarios[[scenario]]$temp, na.rm = TRUE),
      `Peak dose` = max(pred$latent_dose_median, na.rm = TRUE),
      `Final dose` = final$latent_dose_median,
      `Minimum survival` = min(pred$survival_median, na.rm = TRUE),
      `Final mortality` = final$mortality_median,
      check.names = FALSE
    )
  }))
}

repair_curve_df <- function(repair_pars,
                            kdot_levels_pct_min,
                            temp_grid) {
  dplyr::bind_rows(lapply(names(kdot_levels_pct_min), function(level) {
    data.frame(
      temp = temp_grid,
      repair_level = level,
      repair_rate_pct_min = repair_rate_schoolfield_pct_min(
        temp = temp_grid,
        TA = repair_pars$TA,
        TAL = repair_pars$TAL,
        TAH = repair_pars$TAH,
        TL = repair_pars$TL,
        TH = repair_pars$TH,
        TREF = repair_pars$TREF,
        kdot_ref_pct_min = unname(kdot_levels_pct_min[[level]])
      )
    )
  })) %>%
    dplyr::mutate(
      repair_level = factor(
        repair_level,
        levels = names(kdot_levels_pct_min),
        labels = gsub("_", " ", names(kdot_levels_pct_min))
      )
    )
}

plot_damage_rate <- function(damage_curve, xlim = c(10, 36)) {
  ggplot(damage_curve$summary, aes(x = temp, y = damage_rate_pct_min_median)) +
    geom_ribbon(
      aes(ymin = damage_rate_pct_min_lower, ymax = damage_rate_pct_min_upper),
      fill = "grey70",
      alpha = 0.25
    ) +
    geom_line(linewidth = 1, colour = "#2F2F2F") +
    coord_cartesian(xlim = xlim) +
    labs(
      x = "Temperature (°C)",
      y = "Damage rate (% LT50-dose per min)"
    ) +
    theme_tdt
}

plot_repair_tpc <- function(repair_df) {
  ggplot(
    repair_df %>% dplyr::filter(repair_level != "none"),
    aes(x = temp, y = repair_rate_pct_min, colour = repair_level)
  ) +
    geom_line(linewidth = 1) +
    scale_colour_manual(
      values = c(
        "low" = "#4D9221",
        "moderate" = "#146C7C",
        "high" = "#C76D37"
      ),
      drop = FALSE
    ) +
    labs(
      x = "Temperature (°C)",
      y = "Repair rate (% LT50-dose per min)",
      colour = NULL
    ) +
    theme_tdt
}

plot_damage_repair_rates <- function(damage_curve,
                                     repair_pars,
                                     kdot_ref_pct_min,
                                     temp_grid) {
  repair_df <- data.frame(
    temp = temp_grid,
    repair_rate_pct_min = repair_rate_schoolfield_pct_min(
      temp = temp_grid,
      TA = repair_pars$TA,
      TAL = repair_pars$TAL,
      TAH = repair_pars$TAH,
      TL = repair_pars$TL,
      TH = repair_pars$TH,
      TREF = repair_pars$TREF,
      kdot_ref_pct_min = kdot_ref_pct_min
    )
  )

  ggplot() +
    geom_ribbon(
      data = damage_curve$summary,
      aes(
        x = temp,
        ymin = damage_rate_pct_min_lower,
        ymax = damage_rate_pct_min_upper,
        fill = "Damage (95% CI)"
      ),
      alpha = 0.25,
      colour = NA
    ) +
    geom_line(
      data = damage_curve$summary,
      aes(x = temp, y = damage_rate_pct_min_median, colour = "Damage (median)"),
      linewidth = 0.9
    ) +
    geom_line(
      data = repair_df,
      aes(x = temp, y = repair_rate_pct_min, colour = "Repair"),
      linewidth = 0.9
    ) +
    scale_colour_manual(
      values = c(
        "Damage (median)" = "#2F2F2F",
        "Repair" = "#146C7C"
      )
    ) +
    scale_fill_manual(
      values = c(
        "Damage (95% CI)" = "#2F2F2F"
      )
    ) +
    labs(
      x = "Temperature (°C)",
      y = "Rate (% LT50-dose per min)",
      colour = "Process",
      fill = "Interval"
    ) +
    theme_tdt +
    theme(legend.position = "right")
}

plot_net_damage_rate <- function(damage_curve, repair_df) {
  damage_df <- damage_curve$summary %>%
    dplyr::select(temp, damage_rate_pct_min_median)

  net_df <- repair_df %>%
    dplyr::left_join(damage_df, by = "temp") %>%
    dplyr::mutate(
      net_damage_rate_pct_min = damage_rate_pct_min_median - repair_rate_pct_min
    )

  ggplot(
    net_df,
    aes(x = temp, y = net_damage_rate_pct_min, colour = repair_level)
  ) +
    geom_hline(yintercept = 0, linetype = 2, colour = "grey45") +
    geom_line(linewidth = 1) +
    coord_cartesian(xlim = c(10, 36)) +
    scale_colour_manual(
      values = c(
        "none" = "#2F2F2F",
        "low" = "#4D9221",
        "moderate" = "#146C7C",
        "high" = "#C76D37"
      ),
      drop = FALSE
    ) +
    labs(
      x = "Temperature (°C)",
      y = "Net damage rate (% LT50-dose per min)",
      colour = NULL
    ) +
    theme_tdt
}
```

### Model Diagnostics

```{r}
# diagnose_tdt_fit ------------------------------------------------------------
# Compile standard HMC and LOO diagnostics into a single list.  Reports
# divergent transitions, max-treedepth hits, max Rhat, min ESS ratio, and
# (optionally) the number of high Pareto-k LOO observations.
#
# @param fit      brmsfit object returned by brm().
# @param loo_obj  Optional loo object from loo::loo() or brms::loo().  If
#                 supplied, Pareto-k diagnostics are included.
#
# @return Named list: divergent_transitions, max_treedepth_hits, max_rhat,
#         min_neff_ratio, pareto_problem_n, problems (character vector of
#         flagged issues, or "No automatic problems flagged.").
diagnose_tdt_fit <- function(fit, loo_obj = NULL) {
  nuts <- tryCatch(brms::nuts_params(fit), error = identity)
  rhat_values <- tryCatch(brms::rhat(fit), error = identity)
  neff_values <- tryCatch(brms::neff_ratio(fit), error = identity)

  divergent <- NA_integer_
  max_treedepth_hits <- NA_integer_

  if (!inherits(nuts, "error")) {
    divergent <- nuts %>%
      dplyr::filter(.data$Parameter == "divergent__") %>%
      dplyr::summarise(n = sum(.data$Value > 0, na.rm = TRUE)) %>%
      dplyr::pull(.data$n)

    tree_values <- nuts$Value[nuts$Parameter == "treedepth__"]
    if (length(tree_values) > 0 && any(is.finite(tree_values))) {
      max_tree <- max(tree_values, na.rm = TRUE)

      max_treedepth_hits <- nuts %>%
        dplyr::filter(.data$Parameter == "treedepth__") %>%
        dplyr::summarise(n = sum(.data$Value >= max_tree, na.rm = TRUE)) %>%
        dplyr::pull(.data$n)
    }
  }

  max_rhat <- if (inherits(rhat_values, "error")) {
    NA_real_
  } else {
    max(rhat_values, na.rm = TRUE)
  }

  min_neff_ratio <- if (inherits(neff_values, "error")) {
    NA_real_
  } else {
    min(neff_values, na.rm = TRUE)
  }

  pareto_problem_n <- NA_integer_
  if (!is.null(loo_obj) && !inherits(loo_obj, "error")) {
    pareto_k <- tryCatch(loo::pareto_k_values(loo_obj), error = identity)
    if (!inherits(pareto_k, "error")) {
      pareto_problem_n <- sum(pareto_k > 0.7, na.rm = TRUE)
    }
  }

  problems <- character()
  if (is.finite(divergent) && divergent > 0) {
    problems <- c(problems, paste(divergent, "divergent transitions"))
  }
  if (is.finite(max_rhat) && max_rhat > 1.01) {
    problems <- c(problems, paste("max Rhat =", round(max_rhat, 3)))
  }
  if (is.finite(min_neff_ratio) && min_neff_ratio < 0.1) {
    problems <- c(problems, paste("min ESS ratio =", round(min_neff_ratio, 3)))
  }
  if (is.finite(pareto_problem_n) && pareto_problem_n > 0) {
    problems <- c(problems, paste(pareto_problem_n, "PSIS Pareto-k values > 0.7"))
  }

  if (length(problems) == 0) {
    problems <- "No automatic problems flagged."
  }

  list(
    divergent_transitions = divergent,
    max_treedepth_hits = max_treedepth_hits,
    max_rhat = max_rhat,
    min_neff_ratio = min_neff_ratio,
    pareto_problem_n = pareto_problem_n,
    problems = problems
  )
}

summarise_tdt_model <- function(workflow) {
  if (!tdt_has_fit(workflow)) {
    stop("workflow$fit is NULL. Fit or load the model first.", call. = FALSE)
  }

  loo_obj <- tryCatch(brms::loo(workflow$fit), error = identity)

  list(
    summary = summary(workflow$fit),
    r2 = tryCatch(performance::r2(workflow$fit), error = identity),
    loo = loo_obj,
    diagnostics = diagnose_tdt_fit(workflow$fit, loo_obj = loo_obj)
  )
}

model_parameter_table <- function(fit) {
  out <- brms::posterior_summary(fit) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("parameter") %>%
    dplyr::filter(grepl("^(b_|sd_|phi$)", parameter)) %>%
    dplyr::mutate(
      component = dplyr::case_when(
        grepl("^b_", parameter) ~ "Population",
        grepl("^sd_", parameter) ~ "Group-level SD",
        parameter == "phi" ~ "Beta-binomial precision",
        TRUE ~ "Other"
      ),
      parameter = gsub("^b_", "", parameter),
      parameter = gsub("^sd_", "sd: ", parameter),
      parameter = gsub("_", " ", parameter)
    ) %>%
    dplyr::select(component, parameter, Estimate, Est.Error, Q2.5, Q97.5)

  names(out) <- c("Component", "Parameter", "Estimate", "Est. error", "2.5%", "97.5%")
  out
}

plot_parameter_intervals <- function(parameter_table) {
  df <- parameter_table %>%
    dplyr::filter(Component %in% c("Population", "Beta-binomial precision")) %>%
    dplyr::mutate(Parameter = factor(Parameter, levels = rev(Parameter)))

  ggplot(df, aes(x = Estimate, y = Parameter, colour = Component)) +
    geom_vline(xintercept = 0, linetype = 2, colour = "grey65") +
    geom_pointrange(aes(xmin = `2.5%`, xmax = `97.5%`), linewidth = 0.65) +
    scale_colour_manual(
      values = c(
        "Population" = "#146C7C",
        "Beta-binomial precision" = "#C76D37"
      )
    ) +
    labs(x = "Posterior estimate", y = NULL, colour = NULL) +
    theme_tdt
}

diagnostic_table <- function(model_checks) {
  d <- model_checks$diagnostics

  data.frame(
    Diagnostic = c(
      "Divergent transitions",
      "Max treedepth hits",
      "Maximum Rhat",
      "Minimum ESS ratio",
      "Pareto-k > 0.7",
      "Automatic flag"
    ),
    Value = c(
      d$divergent_transitions,
      d$max_treedepth_hits,
      round(d$max_rhat, 3),
      round(d$min_neff_ratio, 3),
      d$pareto_problem_n,
      paste(d$problems, collapse = "; ")
    )
  )
}

r2_table <- function(fit) {
  r2_obj <- tryCatch(
    performance::r2(fit),
    error = function(e) e
  )

  if (inherits(r2_obj, "error")) {
    return(data.frame(
      Metric = "R2",
      Estimate = NA_real_,
      SE = NA_real_,
      Note = "Could not compute"
    ))
  }

  r2_df <- as.data.frame(r2_obj)

  if (nrow(r2_df) == 0 || ncol(r2_df) == 0) {
    return(data.frame(
      Metric = "R2",
      Estimate = NA_real_,
      SE = NA_real_,
      Note = "Could not compute"
    ))
  }

  estimate_col <- names(r2_df)[grepl("R2|R\\.?2", names(r2_df), ignore.case = TRUE)][1]
  se_col <- names(r2_df)[grepl("SE|Std", names(r2_df), ignore.case = TRUE)][1]

  if (is.na(estimate_col)) {
    estimate_col <- names(r2_df)[vapply(r2_df, is.numeric, logical(1))][1]
  }

  if (is.na(estimate_col)) {
    return(data.frame(
      Metric = "R2",
      Estimate = NA_real_,
      SE = NA_real_,
      Note = "Could not compute"
    ))
  }

  data.frame(
    Metric = "R2 (Bayesian)",
    Estimate = round(as.numeric(r2_df[[estimate_col]][1]), 3),
    SE = if (!is.na(se_col)) round(as.numeric(r2_df[[se_col]][1]), 3) else NA_real_
  )
}

loo_table <- function(model_checks) {
  loo_obj <- model_checks$loo
  if (inherits(loo_obj, "error")) {
    return(data.frame(Metric = "LOO", Estimate = NA_real_, SE = NA_real_))
  }

  estimates <- as.data.frame(loo_obj$estimates)
  estimates$Metric <- rownames(estimates)
  rownames(estimates) <- NULL
  estimates %>%
    dplyr::select(Metric, Estimate, SE) %>%
    dplyr::mutate(Metric = dplyr::recode(
      Metric,
      elpd_loo = "ELPD-LOO",
      p_loo = "Effective parameters",
      looic = "LOOIC"
    ))
}

selected_diagnostic_parameters <- function(fit) {
  wanted <- c(
    "b_lraw_Intercept", "b_lraw_temp_c",
    "b_uraw_Intercept", "b_uraw_temp_c",
    "b_logk_Intercept", "b_logk_temp_c",
    "b_mid_Intercept",  "b_mid_temp_c",
    "phi",
    "sd_Date__mid_Intercept",
    "sd_Tank__mid_Intercept"
  )

  draw_names <- names(posterior::as_draws_df(fit))
  wanted[wanted %in% draw_names]
}

plot_trace_diagnostics <- function(fit) {
  pars <- selected_diagnostic_parameters(fit)
  draws <- posterior::as_draws_array(fit)

  bayesplot::mcmc_trace(draws, pars = pars, facet_args = list(ncol = 2)) +
    bayesplot::bayesplot_theme_get() +
    labs(x = "Post-warmup iteration", y = "Parameter value")
}

plot_rank_diagnostics <- function(fit) {
  pars <- selected_diagnostic_parameters(fit)
  draws <- posterior::as_draws_array(fit)

  bayesplot::mcmc_rank_overlay(draws, pars = pars) +
    bayesplot::bayesplot_theme_get() +
    labs(x = "Rank-normalised draw", y = "Chain")
}

plot_rhat_neff_diagnostics <- function(fit) {
  rhat_values <- brms::rhat(fit)
  neff_values <- brms::neff_ratio(fit)

  df <- dplyr::bind_rows(
    data.frame(
      parameter = names(rhat_values),
      metric = "Rhat",
      value = as.numeric(rhat_values)
    ),
    data.frame(
      parameter = names(neff_values),
      metric = "ESS ratio",
      value = as.numeric(neff_values)
    )
  ) %>%
    dplyr::filter(is.finite(value))

  ggplot(df, aes(x = value)) +
    geom_histogram(bins = 35, fill = "#146C7C", colour = "white") +
    geom_vline(
      data = data.frame(metric = c("Rhat", "ESS ratio"), value = c(1.01, 0.1)),
      aes(xintercept = value),
      linetype = 2,
      colour = "#C76D37"
    ) +
    facet_wrap(~ metric, scales = "free_x") +
    labs(x = "Diagnostic value", y = "Number of parameters") +
    theme_tdt
}

plot_ppc_observed_vs_predicted <- function(workflow, ndraws = 500) {
  pp <- brms::posterior_predict(workflow$fit, ndraws = ndraws)

  q <- apply(pp, 2, tdt_quantile)
  df <- workflow$data %>%
    dplyr::mutate(
      pred_surv_lower = q[1, ] / n_total,
      pred_surv_median = q[2, ] / n_total,
      pred_surv_upper = q[3, ] / n_total
    )

  ggplot(df, aes(x = survival, y = pred_surv_median)) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey45") +
    geom_errorbar(
      aes(ymin = pred_surv_lower, ymax = pred_surv_upper),
      width = 0,
      alpha = 0.35,
      colour = "#146C7C"
    ) +
    geom_point(aes(colour = assay_temp), size = 2, alpha = 0.85) +
    scale_x_continuous(limits = c(0, 1)) +
    scale_y_continuous(limits = c(0, 1)) +
    scale_colour_viridis_c(option = "D", end = 0.88) +
    labs(
      x = "Observed survival",
      y = "Posterior predictive survival",
      colour = "Assay temp (°C)"
    ) +
    theme_tdt
}
```

## Case Study 1: Shrimp Lethal TDT

The functions above are intended to be reused across case studies. The shrimp dataset is the first example; future datasets should enter through the same `prepare_tdt_survival_data()` and `fit_tdt_4pl()` interface.

```{r}
shrimp_raw <- readxl::read_excel(
  "data/Data_Shrimp_TDT_CTmax_experiment.xlsx",
  sheet = "LETHAL_TDT"
)

shrimp_tdt <- prepare_tdt_survival_data(
  data = shrimp_raw,
  temp = "Temperature_assay",
  duration = "Duration_exposure_hours",
  n_total = "N_individuals_after_trial",
  mortality = "Mortality_after_trial",
  random_effects = c("Date", "Tank"),
  duration_unit = "hours"
)

saveRDS(shrimp_raw, file.path(tdt_output_dir, "shrimp_lethal_raw.RDS"))
saveRDS(shrimp_tdt, file.path(tdt_output_dir, "shrimp_lethal_prepared_tdt.RDS"))

shrimp_data_summary <- shrimp_tdt %>%
  dplyr::summarise(
    n_rows = dplyr::n(),
    `Temperature min (°C)` = min(assay_temp, na.rm = TRUE),
    `Temperature max (°C)` = max(assay_temp, na.rm = TRUE),
    `Duration min (h)` = min(duration, na.rm = TRUE),
    `Duration max (h)` = max(duration, na.rm = TRUE),
    median_n = median(n_total, na.rm = TRUE),
    `Temperature centre (°C)` = attr(shrimp_tdt, "tdt_meta")$temp_mean
  )

tdt_ft(shrimp_data_summary, digits = 2, caption = "Shrimp lethal TDT dataset used for the 4PL fit.")
```

## Build Or Load The Main Model

The object below is the beta-binomial analogue of the main lethal shrimp 4PL: no observation-level random effect, random intercepts on `mid` for date and tank, and population-level temperature effects on both `mid` (midpoint) and `logk` (log-steepness).

```{r}
shrimp_workflow_spec <- fit_tdt_4pl(
  data = shrimp_tdt,
  random_effects = c("Date", "Tank"),
  fit = FALSE
)

saveRDS(shrimp_workflow_spec, file.path(tdt_output_dir, "shrimp_lethal_workflow_spec.RDS"))

shrimp_workflow_spec$formula
tdt_ft(
  as.data.frame(shrimp_workflow_spec$prior),
  digits = 3,
  caption = "Priors used for the shrimp beta-binomial 4PL model."
)
```

```{r}
#| output: false

if (file.exists(params$model_path) && !isTRUE(params$refit_model)) {
  
  fit_shrimp_4pl <- shrimp_workflow_spec
  fit_shrimp_4pl$fit <- readRDS(params$model_path)
  
} else {
  
  fit_shrimp_4pl <- fit_tdt_4pl(
    data = shrimp_tdt,
    random_effects = c("Date", "Tank"),
    chains = 4,
    cores = 4,
    iter = 6000,
    warmup = 3000,
    seed = 123,
    backend = "cmdstanr",
    control = list(adapt_delta = 0.995, max_treedepth = 15),
    file = params$model_path,
    overwrite = TRUE,
    fit = TRUE,
    refresh = if (isTRUE(params$show_sampler_progress)) 100 else 0
  )
}
```

```{r}
# Model summary
summary(fit_shrimp_4pl$fit)
```

## Model Checks And Derived Outputs

The remaining chunks only run if a fitted model is available.

```{r}
if (tdt_has_fit(fit_shrimp_4pl)) {
  model_checks <- summarise_tdt_model(fit_shrimp_4pl)
  saveRDS(model_checks, file.path(tdt_output_dir, "shrimp_lethal_model_checks.RDS"))

  parameter_table <- model_parameter_table(fit_shrimp_4pl$fit)

  tdt_ft(parameter_table, digits = 3, caption = "Table 1. Posterior summary of fitted 4PL parameters for the shrimp lethal TDT beta-binomial model. Population-level coefficients (b_) are on the transformed parameter scale (logit for asymptotes, log for k and mid). Group-level SDs quantify Date and Tank variation in the midpoint of the survival curve.")
  tdt_ft(r2_table(fit_shrimp_4pl$fit), digits = 3, caption = "Table 2. Bayesian R² for the shrimp lethal TDT model, computed from posterior predictive variances.")
  tdt_ft(loo_table(model_checks), digits = 2, caption = "Table 3. Pareto-smoothed importance sampling leave-one-out cross-validation (PSIS-LOO) summary. ELPD-LOO is the expected log predictive density; lower LOOIC indicates better predictive accuracy.")
  tdt_ft(diagnostic_table(model_checks), digits = 3, caption = "Table 4. Automatic HMC and LOO diagnostics. Values within acceptable thresholds (Rhat < 1.01, ESS ratio > 0.1, 0 divergent transitions) indicate reliable sampling.")
}
```

```{r}
#| fig-cap: "**Figure 1.** Posterior predictive check: observed versus model-predicted survival proportions. Each point represents one experimental unit (well or cup), coloured by assay temperature. Vertical intervals show the 95% posterior predictive interval. Points close to the 1:1 dashed line indicate good model fit; systematic deviations would suggest model misspecification."
if (tdt_has_fit(fit_shrimp_4pl)) {
  p_ppc <- plot_ppc_observed_vs_predicted(fit_shrimp_4pl, ndraws = 500)

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_ppc_observed_vs_predicted.png"),
    plot = p_ppc,
    width = 6,
    height = 5,
    dpi = 350
  )

  p_ppc
}
```

### Survival Curves

```{r}
#| fig-cap: "**Figure 2.** Posterior survival curves as a function of exposure duration for five assay temperatures (30–33 °C, 0.5 °C increments). Lines show the posterior median survival probability; shaded ribbons are 95% credible intervals from the beta-binomial 4PL model. Points and error bars overlay observed mean ± SE survival proportions at each experimental temperature–duration combination. Duration is shown on the natural scale in hours; model predictions are population-level (random effects for date and tank marginalised out)."
if (tdt_has_fit(fit_shrimp_4pl)) {
  surv_pred <- predict_survival_curves(
    workflow = fit_shrimp_4pl,
    temps = seq(30, 33, by = 0.5),
    ndraws = params$ndraws
  )

  saveRDS(surv_pred, file.path(tdt_output_dir, "shrimp_lethal_survival_curve_predictions.RDS"))

  p_survival_curves <- plot_survival_curves(
    pred = surv_pred,
    observed = shrimp_tdt
  )

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_survival_curves.png"),
    plot = p_survival_curves,
    width = 9,
    height = 6,
    dpi = 400
  )

  p_survival_curves
}
```

### TDT Curve And Classical TDT Parameters

```{r}
#| output: false
if (tdt_has_fit(fit_shrimp_4pl)) {
  tdt_temp_grid <- seq(28, 36.1, by = 0.05)

  shrimp_lt75 <- derive_ltx_curve(
    workflow        = fit_shrimp_4pl,
    temp_grid       = tdt_temp_grid,
    target_survival = 0.75,
    ndraws          = params$ndraws,
    time_multiplier = 60,
    output_time_unit = "min"
  )

  shrimp_lt50 <- LT50_T_curve(
    workflow = fit_shrimp_4pl,
    T_grid   = tdt_temp_grid,
    ndraws   = params$ndraws,
    time_multiplier  = 60,
    output_time_unit = "min"
  )

  shrimp_lt25 <- derive_ltx_curve(
    workflow        = fit_shrimp_4pl,
    temp_grid       = tdt_temp_grid,
    target_survival = 0.25,
    ndraws          = params$ndraws,
    time_multiplier = 60,
    output_time_unit = "min"
  )

  shrimp_ltx_all <- bind_ltx_curves(shrimp_lt75, shrimp_lt50, shrimp_lt25)

  saveRDS(shrimp_lt50,    file.path(tdt_output_dir, "shrimp_lethal_lt50_curve.RDS"))
  saveRDS(shrimp_ltx_all, file.path(tdt_output_dir, "shrimp_lethal_ltx_all_curves.RDS"))

  shrimp_tdt_params <- derive_tdt_parameters(ltx_curve = shrimp_lt50, t_ref = 1)
  saveRDS(shrimp_tdt_params, file.path(tdt_output_dir, "shrimp_lethal_tdt_parameters.RDS"))
}
```

```{r}
#| fig-cap: "**Figure 3.** Thermal death time (TDT) curves for three survival thresholds. Upper panel: raw time (min); lower panel: log10 time (min). Lines show the posterior median time needed to reduce survival to 75% (green), 50% (teal), or 25% (orange) at a given assay temperature; shaded ribbons are 95% credible intervals. All curves are derived by numerically inverting the posterior 4PL survival model at each temperature, and therefore fully propagate parameter uncertainty."
if (tdt_has_fit(fit_shrimp_4pl)) {
  p_ltx_all <- plot_ltx_curve(shrimp_ltx_all, time_unit = "min")

  p_lt50 <- plot_ltx_curve(shrimp_lt50, time_unit = "min")

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_lt50_tdt_curve.png"),
    plot = p_lt50,
    width = 8, height = 10, dpi = 400
  )
  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_lt25_lt50_lt75_tdt_curve.png"),
    plot = p_ltx_all,
    width = 8, height = 10, dpi = 400
  )

  p_ltx_all
}
```

```{r}
if (tdt_has_fit(fit_shrimp_4pl)) {
  tdt_params_table <- shrimp_tdt_params$summary %>%
    dplyr::mutate(
      `z (°C per decade)` = format_interval(z_median, z_lower, z_upper, digits = 2)
    ) %>%
    dplyr::select(`z (°C per decade)`)

  tdt_ft(
    tdt_params_table,
    caption = "Table 5. Classical TDT parameters derived from the posterior LT50 curve. z is the thermal tolerance range (°C per decade of time), estimated as −1/slope from per-draw log-linear regressions of log10(LT50) on temperature."
  )
}
```

### TDT Landscape

```{r}
#| fig-cap: "**Figure 5.** Thermal death time landscape: posterior median predicted survival across the full temperature–duration design space. Colour intensity (dark = low survival, light = high survival) is the posterior median of the 4PL survival probability on a 120 × 120 grid spanning the observed temperature (30–36 °C) and duration (0.17–8 h) ranges. White contour lines mark the 25%, 50%, and 75% survival isolines. Overlaid points show raw observed survival proportions at each experimental combination."
if (tdt_has_fit(fit_shrimp_4pl)) {
  landscape <- derive_tdt_landscape(
    workflow = fit_shrimp_4pl,
    temp_grid = seq(
      min(shrimp_tdt$assay_temp, na.rm = TRUE),
      max(shrimp_tdt$assay_temp, na.rm = TRUE),
      length.out = 120
    ),
    duration_grid = seq(
      min(shrimp_tdt$duration, na.rm = TRUE),
      max(shrimp_tdt$duration, na.rm = TRUE),
      length.out = 120
    ),
    ndraws = params$ndraws
  )

  saveRDS(landscape, file.path(tdt_output_dir, "shrimp_lethal_tdt_landscape.RDS"))

  p_landscape <- plot_tdt_landscape(
    landscape = landscape,
    observed = shrimp_tdt
  )

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_tdt_landscape.png"),
    plot = p_landscape,
    width = 8,
    height = 6,
    dpi = 400
  )

  p_landscape
}
```

### Damage and Repair Rates

The fitted 4PL model implies an instantaneous damage rate at every temperature: $r_D(T) = 1 / \text{LT50}(T)$. Here LT50 is in hours, so $r_D$ is in LT50-doses per hour; multiplying by 100/60 converts to % LT50-dose per minute. Repair is modelled as a Sharpe–Schoolfield thermal performance curve that peaks at an optimum temperature and declines at both extremes. Note that at temperatures below ~30 °C the damage rate is very small (< 1 % per min), so it appears near-zero on a scale that accommodates the steep rise above 33 °C — both figures are numerically consistent, just viewed at different y-axis scales.

```{r}
#| output: false

repair_pars_shrimp <- list(
  TA   = 14065,
  TAL  = 50000,
  TAH  = 120000,
  TL   = 10.5 + 273.15,
  TH   = 22.5 + 273.15,
  TREF = 17   + 273.15
)

repair_levels_pct_min <- c(
  none     = 0,
  low      = 0.005,
  moderate = 0.01,
  high     = 0.05
)

if (tdt_has_fit(fit_shrimp_4pl)) {
  pars_shrimp <- extract_standard_4pl_pars(fit_shrimp_4pl)
  saveRDS(pars_shrimp, file.path(tdt_output_dir, "shrimp_lethal_4pl_posterior_parameters.RDS"))

  damage_shrimp <- damage_rate_curve(
    pars        = pars_shrimp,
    temp_mean   = fit_shrimp_4pl$meta$temp_mean,
    temp_grid   = seq(10, 36, by = 0.1),
    n_draws     = params$ndraws,
    time_multiplier_to_minutes = 60
  )
  saveRDS(damage_shrimp, file.path(tdt_output_dir, "shrimp_lethal_damage_rate_curve.RDS"))

  repair_shrimp <- repair_curve_df(
    repair_pars         = repair_pars_shrimp,
    kdot_levels_pct_min = repair_levels_pct_min,
    temp_grid           = seq(10, 36, by = 0.1)
  )
  saveRDS(repair_shrimp, file.path(tdt_output_dir, "shrimp_lethal_repair_rate_curve.RDS"))
}
```

```{r}
#| fig-cap: "**Figure 6.** Damage and repair rates as a function of temperature on a log10 scale. The black line and grey ribbon show the posterior median and 95% credible interval of the instantaneous damage rate (% LT50-dose per minute), derived analytically from the fitted 4PL model as $r_D(T) = 100 / (60 \\times \\text{LT50}_h(T))$. Coloured dashed lines show three illustrative repair rates (low, moderate, high) modelled as Sharpe–Schoolfield thermal performance curves with an optimum at ~17–22 °C. On a log scale the three repair levels are clearly visible as near-horizontal lines in the range 0.001–0.1 % per min, while damage rises steeply above 31 °C and exceeds even the highest repair level above ~32 °C. At 25–28 °C (the spike temperatures used in the fluctuating simulations) damage rates are 0.02–0.25 % per min — small but non-zero."
if (tdt_has_fit(fit_shrimp_4pl)) {
  repair_overlay <- repair_curve_df(
    repair_pars         = repair_pars_shrimp,
    kdot_levels_pct_min = repair_levels_pct_min[c("low", "moderate", "high")],
    temp_grid           = seq(10, 36, by = 0.1)
  ) %>%
    dplyr::filter(repair_rate_pct_min > 1e-6)

  damage_log_df <- damage_shrimp$summary %>%
    dplyr::filter(
      damage_rate_pct_min_median  > 1e-6,
      damage_rate_pct_min_lower   > 1e-6,
      damage_rate_pct_min_upper   > 1e-6
    )

  p_damage_repair <- ggplot() +
    geom_ribbon(
      data = damage_log_df,
      aes(x = temp, ymin = damage_rate_pct_min_lower, ymax = damage_rate_pct_min_upper),
      fill = "grey70", alpha = 0.25, colour = NA
    ) +
    geom_line(
      data = damage_log_df,
      aes(x = temp, y = damage_rate_pct_min_median, colour = "Damage"),
      linewidth = 1
    ) +
    geom_line(
      data = repair_overlay,
      aes(x = temp, y = repair_rate_pct_min, colour = repair_level, linetype = repair_level),
      linewidth = 0.85
    ) +
    scale_colour_manual(
      values = c(
        Damage   = "#2F2F2F",
        low      = "#4D9221",
        moderate = "#146C7C",
        high     = "#C76D37"
      ),
      name = NULL
    ) +
    scale_linetype_manual(
      values = c(low = "dashed", moderate = "dashed", high = "dashed"),
      name = NULL, guide = "none"
    ) +
    scale_y_log10(
      breaks = c(0.0001, 0.001, 0.01, 0.1, 1, 10, 100),
      labels = c("0.0001", "0.001", "0.01", "0.1", "1", "10", "100")
    ) +
    coord_cartesian(xlim = c(10, 36), ylim = c(0.0001, 200)) +
    labs(
      x = "Temperature (°C)",
      y = "Rate (% LT50-dose per min, log₁₀ scale)"
    ) +
    theme_tdt +
    theme(legend.position = "right")

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_damage_repair_combined.png"),
    plot     = p_damage_repair,
    width = 8, height = 5.5, dpi = 400
  )

  p_damage_repair
}
```

```{r}
#| fig-cap: "**Figure 7.** Net damage rate (damage minus repair) as a function of temperature, zoomed to the biologically relevant crossover region (−0.2 to 2 % LT50-dose per min). Positive values indicate net injury accumulation; negative values indicate net repair (recovery). The dashed line marks equilibrium. At temperatures below ~28–31 °C (depending on repair level), repair capacity exceeds incoming damage, allowing partial recovery. Above the crossover, damage dominates irreversibly. The steep rise of the 'none' and low-repair curves above 30 °C continues off the top of this panel; the log-scale figure above shows the full magnitude."
if (tdt_has_fit(fit_shrimp_4pl)) {
  p_net <- plot_net_damage_rate(damage_shrimp, repair_shrimp) +
    coord_cartesian(xlim = c(10, 36), ylim = c(-0.2, 2))

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_net_damage_rate.png"),
    plot     = p_net,
    width = 8, height = 5.5, dpi = 400
  )

  p_net
}
```

### Temperature Scenarios And Fluctuating Survival

Hourly temperature time series (96 hours, 20 °C baseline) with five scenarios: a single 1-hour spike at 31.5 °C, three mild 1-hour spikes at 29 °C, three severe 1-hour spikes at 31 °C, and a mixed scenario (two mild + one severe). These temperatures are in the biologically relevant range where damage rates are substantial: at 29 °C the LT50 is roughly 100–200 min, and at 31–31.5 °C it drops to 20–60 min, so a single 1-hour spike at 31.5 °C already corresponds to a meaningful fraction of the LT50-dose. Panels are ordered from simplest (no heat) to most complex (mixed severity).

```{r}
#| output: false
if (tdt_has_fit(fit_shrimp_4pl)) {
  temp_scenarios <- make_temperature_scenarios(
    baseline    = 20,
    single_temp = 31.5,
    mild_temp   = 29,
    severe_temp = 31
  )

  saveRDS(temp_scenarios, file.path(tdt_output_dir, "shrimp_lethal_temperature_scenarios.RDS"))

  repair_prediction_levels <- c(
    no_repair       = 0,
    moderate_repair = repair_levels_pct_min[["moderate"]],
    high_repair     = repair_levels_pct_min[["high"]]
  )

  fluctuating_predictions <- list()
  for (scenario_name in names(temp_scenarios)) {
    for (repair_name in names(repair_prediction_levels)) {
      use_repair  <- repair_prediction_levels[[repair_name]] > 0
      pred_name   <- paste(scenario_name, repair_name, sep = "__")

      fluctuating_predictions[[pred_name]] <- predict_survival_fluctuating(
        temp_df          = temp_scenarios[[scenario_name]],
        pars             = pars_shrimp,
        temp_mean        = fit_shrimp_4pl$meta$temp_mean,
        n_draws          = params$ndraws,
        repair           = use_repair,
        repair_pars      = repair_pars_shrimp,
        kdot_ref_pct_min = repair_prediction_levels[[repair_name]],
        irreversible_mortality = TRUE
      )
    }
  }

  saveRDS(
    fluctuating_predictions,
    file.path(tdt_output_dir, "shrimp_lethal_fluctuating_predictions.RDS")
  )

  scenario_summary <- summarise_fluctuating_predictions(
    temp_scenarios = temp_scenarios,
    predictions    = fluctuating_predictions
  )

  saveRDS(
    scenario_summary,
    file.path(tdt_output_dir, "shrimp_lethal_fluctuating_prediction_summary.RDS")
  )
}
```

```{r}
#| fig-cap: "**Figure 8.** Temperature time series for the five simulation scenarios. Each trace shows 96 hours of hourly temperatures at a 20 °C baseline with 1-hour heat spikes. The single-spike scenario has one spike at 31.5 °C; mild repeated has three spikes at 29 °C; severe repeated has three spikes at 31 °C; mixed severity has two mild (29 °C) spikes followed by one severe (31 °C) spike. Panels are ordered from simplest to most complex thermal challenge."
if (tdt_has_fit(fit_shrimp_4pl)) {
  p_temperature_scenarios <- plot_temperature_scenarios(temp_scenarios)

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_temperature_scenarios.png"),
    plot     = p_temperature_scenarios,
    width = 9, height = 13, dpi = 400
  )

  p_temperature_scenarios
}
```

```{r}
#| fig-cap: "**Figure 9.** Accumulated latent injury dose over 96 hours under five temperature scenarios and three repair assumptions. The latent dose is expressed in LT50-dose units: a dose of 1 corresponds to 50% expected mortality. The dashed horizontal line marks this LT50 threshold. Coloured lines show posterior median dose; ribbons are 95% credible intervals. The single 31.5 °C spike produces immediate dose accumulation; repeated 31 °C severe spikes accumulate the most total dose. High repair capacity reduces the residual dose between events."
if (tdt_has_fit(fit_shrimp_4pl)) {
  p_fluctuating_dose <- plot_fluctuating_metric(fluctuating_predictions, metric = "latent_dose")

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_fluctuating_latent_dose.png"),
    plot     = p_fluctuating_dose,
    width = 10, height = 14, dpi = 400
  )

  p_fluctuating_dose
}
```

```{r}
#| fig-cap: "**Figure 10.** Predicted cumulative survival over 96 hours under five temperature scenarios and three repair assumptions. Survival is derived from the accumulated latent dose via the fitted 4PL dose–response curve and the irreversible mortality assumption (survival can only decrease over time). The no-heat baseline shows negligible injury throughout. The single 31.5 °C spike causes immediate, permanent mortality reduction. Repeated 31 °C severe spikes cause the largest cumulative mortality; high repair capacity provides meaningful buffering, particularly between successive events in the mild and mixed-severity scenarios."
if (tdt_has_fit(fit_shrimp_4pl)) {
  p_fluctuating_survival <- plot_fluctuating_metric(fluctuating_predictions, metric = "survival")

  ggsave(
    filename = file.path(tdt_fig_dir, "shrimp_lethal_fluctuating_survival.png"),
    plot     = p_fluctuating_survival,
    width = 10, height = 14, dpi = 400
  )

  p_fluctuating_survival
}
```

```{r}
if (tdt_has_fit(fit_shrimp_4pl)) {
  tdt_ft(
    scenario_summary,
    digits  = 3,
    caption = "Table 7. Summary of end-state predictions for each temperature scenario and repair level. Peak dose is the maximum accumulated latent injury dose (LT50-dose units); final dose is the dose at t = 96 h; minimum survival and final mortality are the worst-case and end-state population outcomes. Doses < 1 indicate sub-lethal cumulative injury."
  )
}
```

## Case Study 2: XXX

Placeholder for the next dataset. It will use the same `prepare_tdt_survival_data()` and `fit_tdt_4pl()` interface with random effects appropriate to that experimental design.

## Case Study 3: XXX

Placeholder for the third dataset. This section can use the same `prepare_tdt_survival_data()` and `fit_tdt_4pl()` workflow once the raw columns and design structure are finalised.

## Reflection: Methods, Decisions, and Rationale

This section documents every modelling decision taken in this notebook, why it was made, and what alternatives exist. It is intended to let the reader (including future me) evaluate the analysis without having to reverse-engineer the code.

### 1. From Binomial + OLRE to Beta-Binomial

The original shrimp model (`R/bayesian_TDT_shrimp.qmd`) used a binomial likelihood with an **observation-level random effect** (OLRE): a variance term added to every row via `lf(z ~ 0 + (1 | obs))`. OLREs are a standard trick for handling overdispersion in binomial GLMMs — they inflate the effective variance of the Bernoulli probability at each observation.

The **beta-binomial** is cleaner. It treats overdispersion as a first-class feature of the likelihood: each row's true survival probability is drawn from a Beta distribution centred on the 4PL curve, and the counts are then Binomial given that probability. A single precision parameter `phi` replaces `nrow(data)` OLRE standard deviations. Benefits:

- Fewer parameters, shorter Stan program, faster warmup.
- No artificial "observation" grouping factor required.
- `phi` has a direct biological interpretation: how much does survival probability vary across replicates at the same temperature and duration?
- Posterior predictive checks are cleaner because overdispersion lives in the response distribution.

`fit_tdt_4pl()` implements this with a small custom brms family, `tdt_beta_binomial`. The custom family is deliberately used instead of the native nonlinear `beta_binomial()` path because brms 2.22.0 generated `trials` as the whole vector inside the row loop for this nonlinear model, rather than `trials[n]`. With varying denominators, that makes Stan evaluate impossible terms such as 11 survivors out of 9 trials and the initial log probability becomes `log(0)`. Passing `n_total` through `vint(n_total)` fixes the likelihood at the row level. The prior `gamma(2, 0.1)` on `phi` has mean 20, placing most mass at moderate overdispersion while allowing very high dispersion.

### 2. The 4-Parameter Logistic Survival Curve

The model fits:

$$p_{\text{surv}}(\text{duration}, \text{temp}) = \text{low} + \frac{\text{up} - \text{low}}{1 + \exp\!\bigl(k \cdot (\log_{10}(\text{duration}) - \text{mid})\bigr)}$$

**Why 4PL and not 2PL?** A standard 2PL (logistic regression on log-duration) forces asymptotes at exactly 0 and 1. In practice:

- The lower asymptote (`low`) represents organisms that survive even very long exposures (resistant individuals, wells where mortality is incomplete, or handling/background mortality).
- The upper asymptote (`up`) represents the maximum survival at very short exposures — rarely 100% because of handling stress, background mortality, or pipetting errors.

Fixing asymptotes at 0 and 1 forces the curve slope to compensate for these phenomena, biasing z and CTmax estimates. Estimating `low` and `up` is more honest and improves predictive accuracy.

**Why log10(duration)?** The classical TDT framework is defined on a log10 time axis: z is defined in units of °C per decade (factor-of-10) of time. Using log10 keeps z interpretable on this classical scale and usually linearises the temperature–time relationship.

### 3. Asymptote Bounds and the log(0) Initialisation Error

**The problem**: The previous code defaulted to `low_bounds = c(0, 0.5)` and `up_bounds = c(0.5, 1)`. This allowed `low` to reach exactly 0 and `up` to reach exactly 1. At those boundaries:

```
logit(p_surv) → logit(0) = -Inf   or   logit(1) = +Inf
```

Stan evaluates the log-probability at the initial parameter values before sampling. An infinite `eta` (the logit-scale linear predictor) makes the beta-binomial log-probability undefined, and Stan reports: *"log probability evaluates to log(0), i.e. negative infinity."*

**The fix**: Change the bounds to `c(0.001, 0.499)` and `c(0.501, 0.999)`. These keep `low` and `up` safely inside (0, 1) for all values of the unconstrained parameters `lraw` and `uraw`. The bounds are implemented via:

```
low = 0.001 + inv_logit(lraw) × 0.498
up  = 0.501 + inv_logit(uraw) × 0.498
```

The formula uses `tdt_num()` to write these constants with 16 significant figures, preventing floating-point rounding from accidentally touching the boundaries.

### 4. Why The Beta-Binomial Was Failing

There were two different routes to the same Stan message.

First, the native nonlinear `brms::beta_binomial()` code path generated the likelihood as if every observation should be checked against the full `trials` vector. That is harmless only when all denominators are identical. In this shrimp dataset, denominators vary from 9 to 12 and some rows have 11 survivors. The generated Stan likelihood therefore tried to evaluate 11 survivors out of 9 trials for some vectorised terms, which has probability zero. That is the main reason the beta-binomial model rejected the initial value for this dataset.

The fix is the custom `tdt_beta_binomial` family. The formula uses:

```
n_surv | vint(n_total) ~ eta
```

and the Stan likelihood is:

```
tdt_beta_binomial_lpmf(Y[n] | mu[n], phi, vint1[n])
```

so each row uses its own denominator.

Second, the original code called `make_tdt_4pl_init()` as the default starting-value function. This function knows how to set `b_lraw`, `b_uraw`, `b_logk`, `b_mid`, and `phi` to safe, data-informed values, but it does **not** know the Stan-level names for the random-effect parameters that brms generates internally.

When the model has `random_effects = c("Date", "Tank")`, brms adds four classes of Stan parameters to handle the group-level structure:

| Stan name | Meaning | Safe value |
|-----------|---------|-----------|
| `sd_1[1]` | SD of Date random intercepts on mid | > 0 (e.g. 1) |
| `z_1[1,j]` | standardised Date effects (one per Date level) | 0 |
| `sd_2[1]` | SD of Tank random intercepts on mid | > 0 (e.g. 1) |
| `z_2[1,j]` | standardised Tank effects (one per Tank level) | 0 |

Because `make_tdt_4pl_init()` does not name these parameters, cmdstanr falls back to drawing each from **Uniform(−2, 2) on the unconstrained scale** for any chain that calls the init function. On the log scale used for `sd_X`, a draw of −2 maps to `exp(−2) ≈ 0.14`. When a very small SD is combined with a large z-score, the implied group random effect can push the linear predictor to an extreme and make the log-probability invalid.

The starting-value fix is to set `init = 0` in `fit_tdt_4pl()` when no custom init is supplied. In cmdstanr, passing a scalar `0` initialises **every** parameter to 0 on its unconstrained scale:

| Parameter class | Unconstrained 0 maps to | Safe? |
|----------------|--------------------------|-------|
| `b_*` (coefficients) | 0 | Yes |
| `sd_X` (log-scale SD) | exp(0) = 1 | Yes |
| `z_X` (z-scores) | 0 | Yes |
| `phi` (log-scale precision) | exp(0) = 1 | Yes (small but valid) |

With `low ≈ 0.25` and `up ≈ 0.75` at the zero starting point, every predicted survival probability is well inside (0, 1), so the beta-binomial log-probability is always finite at initialisation.

`make_tdt_4pl_init()` is kept as a utility for users who want data-informed starting values (e.g., anchoring `mid` near the observed transition zone). Supply it explicitly via `fit_tdt_4pl(..., init = make_tdt_4pl_init(data, ...))` if desired.

### 5. Prior Choices

| Parameter | Prior | Rationale |
|-----------|-------|-----------|
| `b_lraw_Intercept` | Normal(logit(0.038), 1) | Centres `low` near 0.02 (2% background mortality); SD 1 spans roughly 0.5%–15% |
| `b_uraw_Intercept` | Normal(logit(0.962), 1) | Centres `up` near 0.98 (98% max survival); SD 1 spans roughly 85%–99.5% |
| `b_logk_Intercept` | Normal(log(2), 1) | Steepness k = 2 at the mean temperature ± ~1 order of magnitude |
| `b_logk_temp_c` | Normal(0, 0.5) | Allows steepness to vary moderately with temperature; regularises toward constant k |
| `b_mid_Intercept` | Normal(median(logd), 1.5) | Data-centred; SD 1.5 spans roughly 30-fold on the duration axis |
| `b_mid_temp_c` | Normal(0, 0.6) | Allows z-values from 1°C to >10°C; prevents implausibly steep slopes |
| Random SDs | Exponential(2) | Weakly regularising; half the mass below 0.35 log10(h) ≈ ±2-fold variation |
| `phi` | Gamma(2, 0.1) | Mean 20; long right tail allows high overdispersion |

These are *not* estimated from the analysed data. Estimating priors from data (empirical Bayes) would understate parameter uncertainty and should be avoided in confirmatory analyses.

### 6. Random Effects Structure

Date and Tank are included as random intercepts on `mid`:

```
mid ~ temp_c + (1 | Date) + (1 | Tank)
logk ~ temp_c
```

The random effects affect only the *position* of the curve along the log-duration axis (not the shape, asymptotes, or temperature slopes). This is the appropriate biological assumption: experimental batch effects shift how long it takes to kill organisms, but do not change the fundamental shape of the dose-response. The temperature slope on `logk` is a population-level (fixed) effect only.

Including these random effects prevents pseudoreplication: multiple tanks per date, and multiple dates in the experiment, mean observations are not exchangeable. Ignoring this would understate uncertainty in all derived quantities.

### 7. Derived Quantities

All downstream quantities are derived from **posterior draws** rather than point estimates. This is the main advantage of fitting the 4PL jointly instead of first estimating separate LT values and then fitting a second TDT regression.

- **TDT curves** (`derive_ltx_curve`): For each draw, the duration where the survival curve crosses a threshold is found by linear interpolation over a fine grid. This propagates 4PL parameter uncertainty into time estimates.

- **Classical TDT parameters** (`derive_tdt_parameters`): For each draw's LT50 curve, a log-linear regression of `log10(time) ~ temperature` gives the slope. z = -1 / slope and the reference-time temperature is the temperature where the line crosses the chosen exposure time. The posterior of z and the reference-time temperature is thus the full Bayesian uncertainty, not a delta-method approximation.

- **Dynamic injury predictions** (`predict_survival_fluctuating`): The static 4PL implies an instantaneous damage rate at each temperature (1 / LTx_time). These rates are integrated over a temperature time-series using the trapezoid rule, giving a latent dose that maps back to survival. This extends the static TDT to fluctuating field conditions, but the repair parameters are scenario parameters (not estimated from this data) and should be treated as sensitivity inputs.

For the standard one-group model, the link to classical z is especially direct. The fitted midpoint is:

$$\text{mid}(T) = \beta_0 + \beta_1(T - \bar{T}).$$

For any fixed survival threshold x that lies between the fitted lower and upper asymptotes, solving the 4PL for duration gives:

$$\log_{10} LT_x(T) = \beta_0 + \beta_1(T - \bar{T}) +
\frac{1}{k}\log\left(\frac{\text{up} - x}{x - \text{low}}\right).$$

The final term changes the intercept but not the temperature slope. Therefore:

$$\frac{d \log_{10} LT_x(T)}{dT} = \beta_1, \qquad z = -\frac{1}{\beta_1}.$$

This is why `b_mid_temp_c` is the core TDT coefficient. The report still derives z from the full posterior LT curve rather than only printing `-1 / b_mid_temp_c`, because the numerical route also works for grouped models, covariate-conditioned predictions, and non-standard survival thresholds.

### 8. What This Analysis Does NOT Do

- **Mechanism**: The 4PL is empirical. It does not prove that heat-death follows a specific biophysical model (e.g., protein denaturation kinetics).
- **Extrapolation is risky**: CTmax-like reference temperatures and 1-hour tolerance temperatures lie near or outside the observed design space. Posterior intervals widen substantially in these regions; they do not make extrapolation safe.
- **Identifiability is not guaranteed**: Free asymptotes, random effects, and beta-binomial dispersion all compete for information. Sparse datasets may not support the full model. Always inspect the diagnostics and prior predictive distribution before trusting the posterior.

---

## Notes On Current Analytical Inconsistencies

- The helper functions in `tdt_problems.qmd` are hard-coded around a `species` factor and specific coefficient names. The shrimp model has no species/life stage factor, so those helpers do not apply directly without rewriting. The functions above remove that hard-coding.
- The old helper documentation says LT50 is in minutes, but both the zebrafish and shrimp prototypes fit `log10(duration)` where duration is in hours, then multiply derived times by 60 for plotting. This notebook makes that conversion explicit with `time_multiplier`.
- `R/bayesian_TDT_shrimp.qmd` uses a binomial likelihood plus an observation-level random effect. The TODO list asks for beta-binomial instead, so `fit_tdt_4pl()` uses the custom `tdt_beta_binomial` brms family and omits the `obs` term.
- The simple expression `z = -1 / b_mid_temp_c` is exact only when steepness is constant. With `logk ~ temp_c`, the effective steepness varies by temperature, so the LTx curve is no longer strictly log-linear. The functions derive z by regressing per-draw log10(LT50) on temperature, which remains valid regardless of whether the log-linear assumption holds exactly.
- In `R/bayesian_TDT_shrimp.qmd`, `z_draws <- -1 / dr$b_mid_temp_c` appears before `dr <- posterior::as_draws_df(fit_4pl)` is created.
- The dynamic injury model uses `k = exp(logk_int)` (the intercept of logk at the centering temperature) for the latent-dose-to-survival mapping. This is an approximation when logk varies with temperature; it is most accurate near the mean assay temperature.

## Methods Rationale And Critical Assessment

### Data Standardisation

The first design decision is to force every case study into the same count-data interface: survivor counts with their row-specific denominator (`n_surv` and `n_total`), exposure duration in hours, temperature in degrees Celsius, and an optional grouping factor. This avoids the main problem in classical TDT workflows: the raw denominator is not thrown away. A well with 10 organisms and a well with 1000 cells should not be treated as equally precise proportions.

The function `prepare_tdt_survival_data()` allows the analyst to start from survivors, deaths, survival proportions, or mortality proportions, but it converts everything to counts before modelling. This is essential for boundary observations. Survival of 0 or 1 is not an error; it is legitimate count information and should enter the likelihood directly.

### Model Choice

The model is a Bayesian 4-parameter logistic survival curve over `log10(duration)`. The lower and upper asymptotes are estimated rather than fixed at 0 and 1. This matters because real assays have background mortality, handling mortality, incomplete kill, gating artefacts, resistant individuals, and other processes that make forced asymptotes too rigid.

The likelihood is beta-binomial rather than binomial plus an observation-level random effect. Both can represent overdispersion, but beta-binomial is a cleaner data-level description here: it says that replicate survival probabilities vary around the expected survival curve. It also avoids adding a large observation random-effect term whose only purpose is residual overdispersion.

The priors are deliberately weak and not estimated from the analysed dataset. The lower asymptote is constrained below 0.5 and centred near 0.02; the upper asymptote is constrained above 0.5 and centred near 0.98. Those constraints are identifiability choices, not strong claims that every dataset has perfect baseline and terminal behaviour. The midpoint prior is broad on the `log10(hours)` scale. If a new case study has very short cell exposures or very long plant exposures, that prior should be checked with prior predictive simulation.

### Temperature-Varying Steepness

The model now estimates `logk ~ temp_c`, so the steepness `k = exp(logk_int + logk_temp * temp_c)` changes with temperature. This is biologically motivated: the width of the survival curve in log-time often narrows at higher temperatures, meaning thermal death becomes more abrupt. Allowing `k` to vary lets the model capture this without forcing symmetry around a single steepness value.

The consequence is that the LTx curve is no longer exactly log-linear in temperature (the offset term `log(...) / k` in the 4PL inversion now depends on `temp_c`). The functions derive z by fitting a log-linear regression to each draw's LT50 curve, which remains valid as an empirical approximation even when the strict log-linearity assumption does not hold.

### Thresholds For TDT Curves

Classical TDT uses time to 50% survival. That is sensible when upper survival is near 1 and lower survival is near 0, but it becomes awkward when the estimated upper asymptote is, for example, 0.6. In that case, 0.5 survival is not a 50% loss from optimum survival. The function `derive_ltx_curve()` therefore accepts any target survival threshold. The helper `derive_relative_survival_targets()` gives one pragmatic choice: define the target as a fixed proportional decline from the fitted upper asymptote toward the lower asymptote.

This is conceptually cleaner than pretending that 50% survival is always the right biological threshold. The weakness is that the chosen threshold becomes a biological decision, not a purely statistical default. It should be reported.

### Dynamic Injury And Repair

The fluctuating-temperature functions translate the fitted static 4PL into a latent injury-dose model. The key object is the posterior time to a chosen survival threshold, `LTx(T)`. For each posterior draw and each temperature:

$$r_D(T) = \frac{1}{LTx(T)}$$

is interpreted as an instantaneous damage rate. If the temperature trace is `T(t)`, cumulative injury without repair is:

$$D(t) = \int_0^t r_D(T(s))\,ds.$$

In the code this integral is evaluated by the trapezoid rule over the supplied temperature time series. A dose of `D = 1` means one threshold-time worth of exposure; for the default LT50 choice, `D = 1` corresponds to the fitted 50% survival point under a constant-temperature exposure.

The model then maps dose back to survival through the fitted 4PL shape:

$$p(D) = \text{low} + \frac{\text{up} - \text{low}}{1 + \exp(k(\log_{10}D + c_x))},$$

where

$$c_x = \frac{1}{k}\log\left(\frac{\text{up} - x}{x - \text{low}}\right).$$

This makes the latent-dose scale interpretable. `D = 0` gives the upper asymptote; `D = 1` gives the chosen threshold `x`; large `D` approaches the lower asymptote. It also makes the damage predictions inherit uncertainty from the fitted asymptotes and slope, rather than pretending that the TDT curve is known exactly.

Repair is added as a scenario process:

$$D(t + \Delta t) = \max\{0,\; D(t) + \Delta D_{\text{heat}} - \Delta D_{\text{repair}}\}.$$

The repair term is a Sharpe-Schoolfield-style thermal performance curve, expressed in the same units as damage: `% LT50-dose per min`, then converted internally to latent-dose units per hour. This is why the report plots damage and repair both separately and together. The separate repair TPC shows the assumed shape and magnitude; the overlay and net-damage plot show whether repair is large enough to matter relative to heat injury.

Why do this instead of fitting a full dynamic mortality model directly? Because the static lethal TDT experiment does not contain dynamic recovery data. A fully estimated repair model would require repeated heat-pulse experiments, recovery intervals, and survival measured through time. The current approach is a posterior predictive scenario bridge: the damage term is estimated from the static 4PL, while the repair term is an explicit sensitivity assumption. That is more honest than hiding repair inside a fitted parameter the data cannot identify.

The other alternative is the classical two-step heat-injury pipeline: estimate LT50 at each temperature, fit a line to `log10(LT50)`, then integrate `1 / LT50(T)`. The 4PL approach keeps the same interpretability but avoids two losses: it does not discard the raw count denominators, and it carries full posterior uncertainty from the survival curve into the dynamic prediction.

### Remaining Weaknesses

The largest remaining weakness is that the 4PL is still empirical. It gives a joint likelihood and uncertainty propagation, but it does not prove a specific biophysical mechanism of death. The z value remains a derived summary of the fitted response over the temperature window being studied. It should not be over-interpreted as a universal species trait.

The second weakness is extrapolation. CTmax-like reference temperatures and 1-hour tolerance temperatures are often near or outside the observed design space. The posterior intervals help expose this, but they do not make extrapolation safe. Every case study should report the observed temperature and duration ranges alongside derived TDT quantities.

The third weakness is identifiability. Free asymptotes, group-level differences, random effects, and beta-binomial overdispersion are all statistically defensible, but they compete for information. Sparse case studies may not support the full model. The point of these functions is to make the modelling assumptions explicit and testable, not to guarantee that the richest model is always appropriate.
