---
title: "Drosophila suzukii sublethal endpoints: knockdown and productivity TDTs"
date: 2026-06-13
format: html
execute:
  warning: false
  message: false
---

## Question

Case Study 4 (*D. suzukii*, Ørsted et al. 2024) already analyses **mortality** by
sex. The long-format `dsuzukii` dataset also carries two sublethal endpoints —
**time to heat coma** (`t_coma`) and **productivity** (`prod`, offspring / female
/ day). This note validates how we estimate thermal-death-time quantities ($z$,
$CT_{max_{1hr}}$) for both, by sex, in a joint model, and how those compare to the
two-stage `drc` values Ørsted report (their `TDT_parameters_R3.csv`):

| trait  | sex | z    | sCTmax1h |
|--------|-----|------|----------|
| coma   | M   | 2.33 | 36.31    |
| coma   | F   | 2.23 | 36.36    |
| fecund | M   | 3.46 | 36.66    |
| fecund | F   | 3.27 | 36.27    |

The analysis itself lives in `ms/supplement.qmd` (§ *Analysing sublethal
measures*, `#sec-dros-coma` and `#sec-dros-productivity`). This note records the
decisions and the validation, loading the cached fits rather than re-fitting.

```{r setup}
library(bayesTLS); library(dplyr); library(brms); library(posterior); library(here)
models_dir <- here::here("output", "models")
data(dsuzukii)
TBAR <- mean(dsuzukii$temp)
```

## Knockdown (time to coma)

**Key assumption — right-censoring.** A fly was scored awake/in-coma at the end
of its assigned exposure (`time` = a % of the temperature-specific median coma
time), and `t_coma` was recorded only for flies that comatosed *during* the
exposure. Flies still awake have `t_coma = NA`. So `t_coma` is **right-censored
at `time`** (≈ `r round(100*mean(is.na(dsuzukii$t_coma)))`% of flies). Dropping
the NA rows would select for early-coma individuals and bias the slope. We use
two censoring-aware routes:

1. **Censored linear** log-time regression (`log10(coma) | cens(...) ~ temp_c *
   sex + (1|cell)`, Gaussian) — keeps every fly via the censoring bound.
2. **4PL on proportion awake** — `in_coma = !is.na(t_coma)`; the proportion awake
   per (temp × level × sex) cell is a survival-type curve, fit with `sex` on the
   midpoint and read with `tls()`. This is exactly Ørsted's `awake ~ time`
   dose-response and is censoring-safe by construction.

The proportion-awake reconstruction reproduces Ørsted's published cell
proportions for 91/94 cells (the 3 misses are the denominator discrepancies
documented in `?dsuzukii`).

```{r coma-validate}
coma_tbar <- mean(dsuzukii$temp)
fit_coma_lin <- readRDS(file.path(models_dir, "fit_dsuzukii_coma_linear.rds"))
fit_coma_4pl <- readRDS(file.path(models_dir, "fit_dsuzukii_coma_4pl.rds"))

p  <- as_draws_df(fit_coma_lin)
sF <- p$b_temp_c; sM <- p$b_temp_c + p[["b_temp_c:sexM"]]
iF <- p$b_Intercept; iM <- p$b_Intercept + p$b_sexM
lin <- data.frame(
  sex = c("F","M"),
  z   = c(median(-1/sF), median(-1/sM)),
  ct1 = c(median(coma_tbar + (log10(60)-iF)/sF), median(coma_tbar + (log10(60)-iM)/sM)))

coma_panel <- dsuzukii |> group_by(temp, lvl, sex) |>
  summarise(n_total = n(), n_awake = sum(is.na(t_coma)),
            duration = first(time), .groups = "drop") |> filter(duration > 0)
std_coma <- standardize_data(coma_panel, temp = "temp", duration = "duration",
              n_total = "n_total", n_surv = "n_awake", duration_unit = "minutes")
set.seed(123)
tls4 <- tls(fit_coma_4pl, by = "sex", mode = "relative", t_ref = 60,
            time_multiplier = 1,
            temp_mean = mean(std_coma$temp - std_coma$temp_c))$summary

knitr::kable(lin, digits = 2, caption = "Censored linear: z, CTmax1h")
knitr::kable(subset(tls4, quantity %in% c("z","CTmax")), digits = 2,
             caption = "4PL (proportion awake): z, CTmax1h")
```

**Result.** Both routes give $z \approx 2.3$–2.4 and $CT_{max_{1hr}} \approx
36.3$–36.5 for both sexes, matching Ørsted's coma values (2.23 / 2.33) within the
credible intervals. Sexes are statistically indistinguishable.

## Productivity (reproductive output)

**Family.** `prod` is continuous with a large zero-mass (≈55%). brms has no
Tweedie, so we use a **hurdle** model and pick the positive part by LOO:

```{r prod-loo}
ga <- add_criterion(readRDS(file.path(models_dir,"fit_dsuzukii_prod_hurdle_gamma.rds")), "loo")
ln <- add_criterion(readRDS(file.path(models_dir,"fit_dsuzukii_prod_hurdle_lognormal.rds")), "loo")
loo_compare(ga, ln)
```

hurdle_**gamma** wins decisively (≈48 elpd, ~7 SE), so the original "log-normal"
preference does not hold for these data — we use gamma.

**Two components, two stories.** We read each hurdle component's TDT from its
posterior by extracting the exposure that halves the component ($ED_{50}$) at
each temperature, then regressing $\log_{10} ED_{50}$ on temperature (per draw).
The helper (`tdt_from_epred()`, defined in the supplement) generalises
`tls()`/`extract_tdt()` to an arbitrary brms response curve.

```{r prod-tdt}
prod_dat <- dsuzukii |> mutate(logd = log10(time), temp_c = temp - mean(temp), sex = factor(sex))
tdt_from_epred <- function(fit, data, predict_fn, t_ref = 60, n_grid = 80) {
  temps <- sort(unique(data$temp)); tbar <- mean(data$temp)
  grid <- do.call(rbind, lapply(temps, function(tt){ rng <- range(data$time[data$temp==tt])
    data.frame(temp=tt, temp_c=tt-tbar, time=10^seq(log10(rng[1]),log10(rng[2]),length.out=n_grid))}))
  grid$logd <- log10(grid$time)
  lapply(setNames(levels(data$sex), levels(data$sex)), function(sx){
    nd <- grid; nd$sex <- factor(sx, levels=levels(data$sex)); M <- predict_fn(nd)
    led <- matrix(NA_real_, nrow(M), length(temps))
    for (j in seq_along(temps)){ cl <- which(nd$temp==temps[j]); lt <- log10(nd$time[cl]); sub <- M[,cl,drop=FALSE]
      thr <- (sub[,1]+sub[,ncol(sub)])/2
      for (d in seq_len(nrow(sub))){ v <- sub[d,]; k <- which(v<=thr[d])[1]
        if(!is.na(k)&&k>1) led[d,j] <- approx(c(v[k-1],v[k]), c(lt[k-1],lt[k]), thr[d])$y }}
    w <- temps-mean(temps); slope <- as.vector((led%*%w)/sum(w^2)); inter <- rowMeans(led)
    list(z=-1/slope, ct=mean(temps)+(log10(t_ref)-inter)/slope, frac_defined=mean(rowSums(is.na(led))==0))})
}
inc <- tdt_from_epred(ga, prod_dat, function(nd) 1 - posterior_epred(ga, newdata=nd, dpar="hu", re_formula=NA))
mag <- tdt_from_epred(ga, prod_dat, function(nd)     posterior_epred(ga, newdata=nd, dpar="mu", re_formula=NA))

data.frame(component = c("incidence (F)","incidence (M)","magnitude (F)","magnitude (M)"),
           z   = c(median(inc$F$z),  median(inc$M$z),  median(mag$F$z),  median(mag$M$z)),
           ct1 = c(median(inc$F$ct), median(inc$M$ct), median(mag$F$ct), median(mag$M$ct)),
           ED50_defined = c(inc$F$frac_defined, inc$M$frac_defined,
                            mag$F$frac_defined, mag$M$frac_defined)) |>
  knitr::kable(digits = 2)
```

**Result / biological finding.**

- **Incidence** (whether a fly reproduces at all): a clean TDT, $z \approx 3.0$ /
  2.8, $CT_{max_{1hr}} \approx 36.2$ / 36.4, $ED_{50}$ defined in 100% of draws.
  Close to Ørsted's fecundity values (3.27 / 3.46). This is the *more*
  heat-sensitive endpoint (lower $z$ than mortality's ~3.0–3.3).
- **Magnitude** (offspring among reproducers): $ED_{50}$ identified in only ~45%
  (F) / 67% (M) of draws — the conditional clutch halves only at the most
  stressful temperatures (36–38 °C) and not at 34–35 °C (it even rises at 34).
  So a magnitude TDT slope is **not identified**.

Reproductive failure under heat is therefore about **whether** a fly reproduces,
not **how much** a surviving reproducer produces. Reporting the components
separately makes this visible; **pooling** them (e.g. ED50 of overall expected
productivity) is artifactual — the per-temperature baseline differs because each
temperature's shortest exposure differs, and there are no `lvl = 0` controls in
the long file to anchor a common baseline (`?dsuzukii`). Decision (with D.N.,
2026-06-13): report incidence + magnitude separately, no pooled number.

**Caveat.** Per-individual `prod` is faithful to the raw counts but is *not*
Ørsted's undocumented normalised cell average (`?dsuzukii`), so exact matches to
their fecundity point estimates are not expected.

## Open questions

- `tdt_from_epred()` is currently a supplement-local helper. It is a candidate
  for promotion to `R/` (a general "TLS for arbitrary brms response models",
  complementing `tls()`/`extract_tdt()` which target the disjoint-bounds 4PL),
  with roxygen + tests — pending a decision on the package API.
