---
title: "Plan: helper functions to apply the 4PL correction factor and derive z / CTmax from arbitrary custom models"
author: "Daniel W.A. Noble (planning dialogue with Claude)"
date: 2026-06-04
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

## Goal

Make it **very easy** for users who fit their own 4PL models — arbitrary
moderators (sex, life stage, clone, …) on *any* sub-parameter, plus random
effects for quantitative-genetic / variance-decomposition work — to:

1. apply the asymmetry **correction factor** and get the absolute
   $\log_{10}\mathrm{LT}_p(T)$, $z$ and $CT_{max}$ as posterior draws; and
2. decompose those derived quantities (e.g. heritability / repeatability of
   absolute $z$ or $CT_{max}$ across clones).

The by-hand demo in `ms/supplement.qmd` works but is model-specific (it parses
coefficient names like `b_mid_sexM:temp_c`). That does not generalise. This note
plans a small, versatile helper set.

## The maths it must wrap

4PL sub-parameters: asymptotes `low`, `up`; steepness `k`; midpoint `mid`. The
package reparameterises to keep them identifiable:
`low = lower + inv_logit(lowraw)·w_low`, `up = up_min + inv_logit(upraw)·w_up`,
`k = exp(logk)`, `mid` on the natural ($\log_{10}$-time) scale.

- **Correction factor (asymmetry term):**
  $$C = \tfrac{1}{k}\,\ln\!\frac{up - p}{p - low}.$$
- **Absolute LT$_p$ curve:** $\log_{10}\mathrm{LT}_p(T) = \mathrm{mid}(T) + C(T)$.
- **Thermal sensitivity:** $z(T) = -1/\big(\mathrm{d}\log_{10}\mathrm{LT}_p/\mathrm{d}T\big)$.
  Relative threshold ($p$ = midpoint) $\Rightarrow C=0 \Rightarrow z = -1/\beta_{1,\text{mid}}$
  exactly; absolute $\Rightarrow$ local slope of the full curve (see
  [[project_z_ctmax_derivation]], [[2026-05-31-delta-method-two-stage]]).
- **CTmax:** the $T$ solving $\log_{10}\mathrm{LT}_p(T) = \log_{10}t_{ref}$.

## The versatility problem — and the clean solution

Users may put moderators on **any** of `low`, `up`, `k`, `mid`, and add random
effects (clone, batch). Reading off `b_mid_sexM:temp_c`-style coefficients is
fragile and model-specific, and it is exactly the thing that makes the supplement
demo non-reusable.

**Solution:** never parse coefficients. Evaluate each sub-parameter at a
user-supplied **moderator grid** with brms's own machinery,
`posterior_linpred(fit, nlpar = , newdata = , re_formula = )`
[@burkner_brms_2017]. brms evaluates the *full* linear predictor for that
sub-parameter — every moderator and the chosen random-effect structure —
automatically. The helper then applies the reparam transform and the closed
forms above. This handles "sex on `up`/`low`/`k` too" with zero extra code: the
user just sets those columns in `newdata`.

### Proof of concept (validated)

On the custom joint-sex model (`mid ~ sex * temp_c`, asymptotes/steepness shared)
this recovers the by-hand absolute $z$ and $CT_{max}$ exactly — generically.

```{r poc}
#| message: false
#| warning: false
library(brms)
fit <- readRDS(here::here("output", "models", "fit_dsuzukii_mortality_sex_joint.rds"))

# Moderator grid: each sex at temp_c = -0.5, 0, +0.5 (dummy logd/n_total for the
# main formula). No coefficient names anywhere.
nd <- expand.grid(temp_c = c(-0.5, 0, 0.5), sex = c("F", "M"), logd = 1, n_total = 1)

lp <- function(np) posterior_linpred(fit, nlpar = np, newdata = nd, re_formula = NA)
low <- 0.001 + plogis(lp("lowraw")) * 0.498
up  <- 0.501 + plogis(lp("upraw"))  * 0.498
k   <- exp(lp("logk"))
mid <- lp("mid")

logLT <- mid + log((up - 0.5) / (0.5 - low)) / k   # absolute LT50 (p = 0.5)
zF <- -1 / (logLT[, 3] - logLT[, 1]);  zM <- -1 / (logLT[, 6] - logLT[, 4])
T_bar <- 36.106
ctF <- T_bar + (log10(240) - logLT[, 2]) / (logLT[, 3] - logLT[, 1])
ctM <- T_bar + (log10(240) - logLT[, 5]) / (logLT[, 6] - logLT[, 4])
round(c(z_F = median(zF), z_M = median(zM),
        CTmax4h_F = median(ctF), CTmax4h_M = median(ctM)), 2)
```

(By-hand earlier: $z$ = 3.04 / 3.18; $CT_{max,4h}$ = 35.14 / 35.17.)

## How the `newdata` approach works (step by step)

The one concept to get: **`newdata` is a grid whose rows are the
(moderator-level $\times$ temperature) combinations at which you want the
LT$_p$ curve evaluated.** Three steps:

1. **Pick the groups** you want separate $z$ / $CT_{max}$ for — the moderator
   combinations (e.g. F, M; or each clone). These are columns in `newdata`.
2. **Each group needs its LT$_p$ curve *across temperature*** — because $z$ is
   the curve's slope and $CT_{max}$ is where it crosses $\log_{10}t_{ref}$. A
   single temperature gives no slope, so `newdata` also spans a short
   **temperature axis**.
3. So `newdata` = (group levels) $\times$ (temperature points).
   `posterior_linpred(nlpar=)` returns each sub-parameter at every row; we form
   $\log_{10}\mathrm{LT}_p$ at every row, then **within each group** take the
   slope (→ $z$) or the crossing (→ $CT_{max}$) along the temperature axis.

Every `newdata` column therefore plays one of three roles:

| role | example columns | used for |
|---|---|---|
| **temperature axis** | `temp_c` | x-axis for the slope ($z$) / crossing ($CT_{max}$) |
| **moderators (the groups)** | `sex`, `life_stage`, `clone` | one $z$/$CT_{max}$ per unique combination |
| **inert** | `logd`, `n_total` | satisfy the formula only; dummy values, ignored by the nlpar predictors |

Worked example — per-sex absolute $z$, fully generic (no coefficient names, no
hard-coded sexes; verified to give F 3.04, M 3.18):

```{r newdata-logic}
#| eval: false
# 1. groups (sex) x a short temperature axis
grid <- expand.grid(sex = c("F","M"), temp_c = seq(-1, 1, by = 0.5),
                    logd = 1, n_total = 1)

# 2. sub-parameters at every row (draws x rows), then absolute log10 LT50
lp  <- function(np) posterior_linpred(fit, nlpar = np, newdata = grid, re_formula = NA)
low <- 0.001 + plogis(lp("lowraw")) * 0.498
up  <- 0.501 + plogis(lp("upraw"))  * 0.498
k   <- exp(lp("logk")); mid <- lp("mid")
logLT <- mid + log((up - 0.5) / (0.5 - low)) / k

# 3. within each group, z = -1 / slope of logLT over temp_c (per-draw LS slope)
z_by_sex <- sapply(unique(grid$sex), function(s) {
  cols <- grid$sex == s; w <- grid$temp_c[cols] - mean(grid$temp_c[cols])
  -1 / as.vector(logLT[, cols] %*% w / sum(w^2))
})                                            # draws x groups; summarise columns
```

`tls_z()` / `tls_ctmax()` just wrap steps 2–3: hand them `newdata`, say which
column is the **temperature axis** (`temp = "temp_c"`) and which are the
**grouping moderators** (`by = "sex"`), and they return per-group posterior
draws. Random effects (e.g. clone) enter by adding the clone column to `newdata`
and `by` and choosing `re_formula` to include the clone REs — same machinery.

## Proposed architecture — three layers

**Layer 1 — atomic math helpers** (operate on numeric posterior draws;
completely model-agnostic — the "very simple" functions Daniel asked for):

```r
tls_correction(low, up, k, p = 0.5)        # (1/k) * log((up - p)/(p - low))
tls_logLT(mid, low, up, k, p = 0.5)        # mid + correction
# plus draw-level curve -> z and curve -> CTmax given a temperature grid
```

A user with an exotic model can extract draws of `low/up/k/mid` however they like
and call these directly. Maximum flexibility, no assumptions.

**Layer 2 — evaluate sub-parameters at a moderator grid** (the versatile
primitive, built on `posterior_linpred(nlpar=)`):

```r
tls_pars(fit, newdata, re_formula = NA, lower = 0, upper = 1, nlpars = c(...))
#   -> tidy draws of (low, up, k, mid) for every row of `newdata`
tls_logLT(fit, newdata, p = 0.5)                       # Layer 2 -> Layer 1
tls_z(fit, newdata, temp = "temp_c", by = NULL,        # group-wise z
      p = 0.5, mode = c("absolute","relative"))        #   default: absolute
tls_ctmax(fit, newdata, temp = "temp_c", by = NULL,    # group-wise CTmax
          t_ref = 60, p = 0.5)
#   temp = temperature-axis column; by = grouping moderator column(s),
#   default = every non-temp, non-inert column. re_formula passes through.
```

For a `bayes_tls` workflow, the bounds (`lower/upper`), nlpar names and the
temperature/centring come from `tdt_meta`. For a **bring-your-own brms fit**, the
user passes them (defaults = the package reparam), so custom models like the
joint-sex one work unchanged.

**Layer 3 — variance decomposition / quant-gen — deferred (future development).**
Not built now (decision 2026-06-04). Recorded for later: Layer 2 already returns
per-group draws (one row per `clone`, `re_formula` including the clone REs), so a
future `tls_var_decomp()` could turn those into a posterior of among-group
variance / repeatability / $H^2$ of absolute $z$ or $CT_{max}$. Two constraints to
respect when it is built: (i) $z$ and $CT_{max}$ are *non-linear* transforms, so
decompose the **derived quantity at the draw level**, not by summing raw-nlpar
variance components; (ii) a `clone` effect gives broad-sense repeatability/$H^2$
— narrow-sense $h^2$ needs an animal-model relatedness structure
[@nakagawa_repeatability_2010; @nakagawa_rsquared_2013].

## Top-level `tls()` wrapper

One call that returns the classical TDT table — $z$, $CT_{max}$, and (for lethal
endpoints) $T_{crit}$ — **per moderator group**, built on the Layer-2 functions.

```r
tls(fit, newdata,
    temp   = "temp_c",
    by     = NULL,             # grouping moderators (default: all non-temp/inert cols)
    params = "all",            # "all" | "z" | "ctmax" | c("z","ctmax") | ...
    mode   = "absolute",       # absolute (default) or relative
    p      = 0.5, t_ref = 60,
    lethal = FALSE,            # T_crit only meaningful for lethal endpoints
    TC_rate_range = c(0.1, 1),
    re_formula = NA, lower = 0, upper = 1, ...)
```

**`params` semantics**

- `"all"` (default) → $z$ + $CT_{max}$ always, **plus $T_{crit}$ iff
  `lethal = TRUE`** ("possibly tcrit if relevant"). With `lethal = FALSE`
  (sublethal — productivity, coma, PSII) `"all"` returns just $z$ + $CT_{max}$ and
  notes $T_{crit}$ was skipped.
- a subset (`c("z","ctmax")`) or a single value (`"z"`) → exactly those.
- requesting `"tcrit"` with `lethal = FALSE` → error: the rate-multiplier
  $T_{crit}$ is undefined for performance endpoints (see [[project_tcrit_definition]]).

**Under the hood — one consistent posterior.** `tls()` evaluates the four
sub-parameters once at `newdata` (Layer 2), forms $\log_{10}LT_p$, then derives
$z$, $CT_{max}$ and $T_{crit}$ from the **same draws** — so $z$ and $CT_{max}$
share draws and $T_{crit} = CT_{max,1h} + z\cdot\log_{10}(r^\*)$ pairs them per
draw (genuinely joint, exactly as `extract_tdt()`). $T_{crit}$ is always anchored
to $CT_{max}$ at 1 h regardless of the `t_ref` shown for $CT_{max}$.

**Returns** a `tls` object: a tidy per-group summary (group × quantity × median +
95% CrI) plus the per-group draws (so contrasts and, later, variance
decomposition run on them). Accessors mirror `get_*_draws()` / `get_*_summary()`.

**Relationship to `extract_tdt()`.** `tls()` is the moderator-aware
generalisation; `extract_tdt()` is the special case (standard formula, `by = NULL`).
`tls_z()` / `tls_ctmax()` / `tls_tcrit()` stay callable for one quantity at a time;
`tls()` bundles them.

Validated prototype (joint-sex fit) — one call yields the per-group TDT table:

```{r tls-wrapper-proto}
#| message: false
#| warning: false
set.seed(123)
tls_proto <- function(fit, by, params = "all", mode = "absolute", p = 0.5,
                      t_ref = 60, lethal = FALSE, TC_rate_range = c(0.1, 1),
                      temps = seq(-1, 1, by = 0.5), T_bar = 36.106, re_formula = NA) {
  if (identical(params, "all")) params <- c("z", "ctmax", if (lethal) "tcrit")
  nd <- do.call(expand.grid, c(by, list(temp_c = temps, logd = 1, n_total = 1)))
  lp <- function(np) brms::posterior_linpred(fit, nlpar = np, newdata = nd, re_formula = re_formula)
  low <- 0.001 + plogis(lp("lowraw")) * 0.498; up <- 0.501 + plogis(lp("upraw")) * 0.498
  logLT <- lp("mid") + if (mode == "absolute") log((up - p) / (p - low)) / exp(lp("logk")) else 0
  g <- interaction(nd[names(by)], drop = TRUE)
  do.call(rbind, lapply(levels(g), function(lev) {
    cols <- which(g == lev); tc <- nd$temp_c[cols]; w <- tc - mean(tc)
    b <- as.vector(logLT[, cols] %*% w / sum(w^2))               # per-draw slope
    a <- rowMeans(logLT[, cols]) - b * mean(tc)                  # intercept at temp_c = 0
    qs <- function(x) round(quantile(x, c(.5, .025, .975)), 2)
    out <- NULL
    if ("z"     %in% params) out <- rbind(out, c(group = lev, quantity = "z",     setNames(qs(-1 / b), c("med","lo","hi"))))
    if ("ctmax" %in% params) out <- rbind(out, c(group = lev, quantity = "CTmax", setNames(qs(T_bar + (log10(t_ref) - a) / b), c("med","lo","hi"))))
    if ("tcrit" %in% params) {
      ct1 <- T_bar + (log10(60) - a) / b
      u   <- runif(length(b), log10(TC_rate_range[1] / 100), log10(TC_rate_range[2] / 100))
      out <- rbind(out, c(group = lev, quantity = "Tcrit", setNames(qs(ct1 + (-1 / b) * u), c("med","lo","hi"))))
    }
    out
  }))
}
tls_proto(fit, by = list(sex = c("F", "M")), params = "all", lethal = TRUE, t_ref = 60 * 4)
```

## Use-case walkthroughs

1. **Moderator on `mid` only (sex):** `newdata` = sex × temperature grid,
   `re_formula = NA` → per-sex $z$, $CT_{max}$. (The supplement case, now generic.)
2. **Moderators on `up`/`low`/`k` too:** identical call — `posterior_linpred`
   evaluates the sex-specific `up`, `low`, `k`, so the correction factor is the
   correctly-moderated one. No extra work; this is the key win.
3. **Quant-gen / heritability (future):** fit e.g.
   `mid ~ temp_c + (1 + temp_c | clone)`; `newdata` one row per clone with
   `re_formula` including the clone REs gives per-clone $z$/$CT_{max}$ draws
   *today* (Layer 2). Turning those into among-clone variance / $H^2$ is the
   deferred Layer 3.

## Integration into the general workflow

`standardize_data()` → `fit_4pl()` **or the user's own brms 4PL** → these helpers
→ contrasts / variance decomposition. `extract_tdt()` becomes a thin wrapper over
Layer 2 for the standard case, so there is one derivation code path. The current
`derive_tdt_curve()` / `derive_temperature_for_duration()` / `derive_z()` are the
standard-formula special case; Layer 2 generalises them to arbitrary moderators
via `newdata`.

## Decisions (2026-06-04) and open questions

**Settled with Daniel (2026-06-04):**

- **Naming — RESOLVED:** new `tls_*` family for the generic layer; keep
  `derive_*` as today.
- **z default — RESOLVED: absolute.** The full-curve (absolute) threshold is the
  field standard, so `tls_z()` / `tls_ctmax()` default to it; relative is
  available via `mode = "relative"`.
- **Heritability — RESOLVED: deferred** to future development (Layer 3 above).

**Still open:**

1. **Bring-your-own-fit contract:** require a `bayes_tls` workflow (bounds/nlpars
   known) or also accept a raw `brmsfit` + a small spec (nlpar names,
   `lower/upper`, temperature variable)? Defaults = package reparam. (Lean: accept
   raw fits — that is the whole point for custom models like the joint-sex one.)
2. **`extract_tdt()` default — needs a decision.** It currently defaults to the
   *relative* threshold (`target_surv`), whereas the field standard (and the new
   `tls_*` default) is *absolute*. Flipping `extract_tdt()` to absolute would
   change existing case-study numbers (shrimp, zebrafish, leaf all call it without
   `target_surv`) and require re-caching/re-checking those. Options: (a) flip the
   default to absolute for consistency and re-verify the case studies, or (b) keep
   `extract_tdt()` relative and make only the new `tls_*` layer default absolute.
3. **Temperature handling:** auto from `tdt_meta` for workflows; explicit `temp =`
   for raw fits (reflected in the signatures above).
4. **Convenience:** ship a `tls_newgrid()` builder for the moderator ×
   temperature grid?

## Verification status

- The `posterior_linpred(nlpar=)` mechanism is validated above on a custom model
  (recovers by-hand $z$ and $CT_{max}$ to the reported precision). **No helper
  functions implemented yet** — this is the design. Related:
  [[project_function_architecture]], [[project_z_ctmax_derivation]],
  [[2026-06-01-productivity-tdt-analysis-plan]].
