---
title: "Joint Bayesian 4PL vs two-stage TDT estimators — bias, coverage, and sensitivity"
date: 2026-05-16
author: "Daniel Noble (with Claude assistance)"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    fig-cap-location: bottom
    code-fold: true
    df-print: kable
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(patchwork)
library(scales)

results_dir <- here::here("output", "sim_twostage")

# ---- Method order and colours used throughout --------------------------
method_levels <- c("two_stage_bin", "two_stage_bb", "joint_4pl")
method_labels <- c("Two-stage (binomial)", "Two-stage (beta-binomial)", "Joint 4PL")
method_cols   <- setNames(c("#7570b3", "#d95f02", "#1b9e77"), method_labels)

# ---- Quantity labels with units ---------------------------------------
quantity_labeller <- as_labeller(c(
  z          = "z (°C)",
  CTmax_1hr  = "CT[max,1h] (°C)"
))

# ---- Helpers -----------------------------------------------------------
load_summaries <- function(prefix) {
  files <- list.files(results_dir,
                      pattern = paste0("^summary_", prefix, "_.*[.]rds$"),
                      full.names = TRUE)
  files <- files[!grepl("summary_conv", files)]
  dplyr::bind_rows(lapply(files, readRDS))
}

load_diagnostics <- function(prefix) {
  files <- list.files(results_dir,
                      pattern = paste0("^diag_", prefix, "_.*[.]rds$"),
                      full.names = TRUE)
  dplyr::bind_rows(lapply(files, readRDS))
}

prep_summary <- function(df) {
  df %>%
    mutate(
      method   = factor(method, levels = method_levels, labels = method_labels),
      quantity = factor(quantity, levels = c("CTmax_1hr", "z")),
      bias_lo  = mean_bias - 1.96 * mcse_bias,
      bias_hi  = mean_bias + 1.96 * mcse_bias,
      cov_lo   = coverage  - 1.96 * mcse_cov,
      cov_hi   = coverage  + 1.96 * mcse_cov
    )
}

theme_proj <- theme_classic(base_size = 11) +
  theme(strip.background = element_blank(),
        strip.text       = element_text(face = "bold"),
        legend.position  = "bottom",
        legend.title     = element_blank(),
        panel.spacing    = unit(0.8, "lines"))

# ---- Per-scenario forest figure (used for Scen 1-5) -------------------
plot_scen_forest <- function(df) {
  d <- prep_summary(df)

  d <- d %>%
    mutate(panel_n = paste0("n[reps] = ",
                            sub(".*_n([0-9]+)$", "\\1", scenario)))

  p_bias <- ggplot(d, aes(x = mean_bias, y = method, colour = method)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbarh(aes(xmin = bias_lo, xmax = bias_hi), height = 0.25) +
    geom_point(size = 2.6) +
    facet_grid(panel_n ~ quantity, scales = "free_x",
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    labs(x = "Mean bias (estimate − truth)", y = NULL,
         title = "(a) Bias") +
    theme_proj + theme(legend.position = "none")

  p_cov <- ggplot(d, aes(x = coverage, y = method, colour = method)) +
    geom_vline(xintercept = 0.95, linetype = "dashed", colour = "grey50") +
    geom_errorbarh(aes(xmin = cov_lo, xmax = cov_hi), height = 0.25) +
    geom_point(size = 2.6) +
    facet_grid(panel_n ~ quantity, scales = "free_x",
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
    labs(x = "Empirical 95% CI coverage", y = NULL,
         title = "(b) Coverage") +
    theme_proj + theme(legend.position = "none")

  p_bias / p_cov + plot_layout(heights = c(1, 1))
}

# ---- Per-cell summary table (compact) ---------------------------------
make_table <- function(df) {
  df %>%
    mutate(
      method   = factor(method, levels = method_levels, labels = method_labels),
      quantity = factor(quantity, levels = c("z", "CTmax_1hr"),
                        labels = c("z", "CTmax_1hr"))
    ) %>%
    arrange(scenario, quantity, method) %>%
    transmute(
      Cell      = scenario,
      Method    = method,
      Quantity  = quantity,
      n_ok      = n,
      Bias      = sprintf("%+.3f (%.3f)", mean_bias, mcse_bias),
      RMSE      = sprintf("%.3f", rmse),
      Coverage  = sprintf("%.1f%%", 100 * coverage),
      Width     = sprintf("%.2f", med_width)
    )
}

# ---- Sweep figure: bias / coverage as a function of swept parameter ---
plot_sweep <- function(df, sweep_x, x_label, x_breaks = waiver()) {
  d <- prep_summary(df)

  p_bias <- ggplot(d, aes(x = .data[[sweep_x]], y = mean_bias, colour = method)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = bias_lo, ymax = bias_hi), width = 0) +
    geom_point(size = 2.4) +
    geom_line() +
    facet_wrap(~ quantity, scales = "free_y",
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    scale_x_continuous(breaks = x_breaks) +
    labs(x = x_label, y = "Mean bias",
         title = "(a) Bias") +
    theme_proj

  p_cov <- ggplot(d, aes(x = .data[[sweep_x]], y = coverage, colour = method)) +
    geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = cov_lo, ymax = cov_hi), width = 0) +
    geom_point(size = 2.4) +
    geom_line() +
    facet_wrap(~ quantity,
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    scale_x_continuous(breaks = x_breaks) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
    labs(x = x_label, y = "Empirical 95% coverage",
         title = "(b) Coverage") +
    theme_proj

  p_bias / p_cov + plot_layout(guides = "collect", heights = c(1, 1)) &
    theme(legend.position = "bottom")
}
```

::: {.callout-note}
This note supersedes
[`2026-05-14-two-stage-bias-sim-results.qmd`](2026-05-14-two-stage-bias-sim-results.html)
and
[`2026-05-15-sensitivity-sweeps.qmd`](2026-05-15-sensitivity-sweeps.html),
both of which were written against pre-bug-fix simulation output and are
no longer reliable. The current writeup is built from the on-disk RDS
files in `output/sim_twostage/` produced by the corrected driver
(2026-05-15 / 2026-05-16). Methods are documented separately in
[`2026-05-13-two-stage-bias-sim-methods.qmd`](2026-05-13-two-stage-bias-sim-methods.qmd);
this note refers back rather than restating.
:::

# Aim

The thermal-death-time (TDT) framework is overwhelmingly fitted with a
two-stage pipeline: per-temperature binomial GLMs estimate
$\log_{10}\text{LT}_{50}$, then those five (or so) point estimates are
regressed against temperature with unweighted OLS to extract the
classical quantities $z$ and $CT_{\max,1\text{h}}$. We want to know
when this pipeline gives the correct answer with the correct
uncertainty, and when it does not.

The joint Bayesian 4PL fits the full dose-response surface in one go.
Algebraically it returns the same $z$ and $CT_{\max,1\text{h}}$ when
the strict-equivalence assumptions hold (constant asymptotes near 0/1,
no overdispersion, common slope across temperature). When those
assumptions break — partial asymptotes, drifting upper bound,
overdispersed counts, sparse design — there is no algebraic guarantee
that the two are the same, and the two-stage's estimating equations
have no parameter to absorb the missing structure.

The simulation study compares three estimators side-by-side over eight
controlled scenarios. The aim is not to advocate for the joint 4PL on
principle, but to map where the two-stage's deficit becomes
operationally important — and where it does not.

# Common methods

::: {.callout-tip}
Full details, including the analytical truth construction, the
parameter feasibility check that runs before any compute starts, and
the diagnostic-criterion table, are in
[`2026-05-13-two-stage-bias-sim-methods.qmd`](2026-05-13-two-stage-bias-sim-methods.qmd).
This section gives only the headline numbers needed to read the
results.
:::

**Three estimators per dataset.** (i) *Two-stage (binomial)* — the
field default: per-temperature binomial GLM Stage 1, unweighted OLS
Stage 2. (ii) *Two-stage (beta-binomial)* — same architecture but
Stage 1 swaps in `glmmTMB` with a beta-binomial likelihood, isolating
the *likelihood* component of the field default's deficit. (iii)
*Joint 4PL* — a `brms`-fit non-linear `beta_binomial(link = "identity")`
model in which all four 4PL sub-parameters carry a centred-temperature
slope; $z$ and $CT_{\max,1\text{h}}$ are extracted from the posterior
with `bayesTLS::extract_tdt()`.

**Design grid.** Five assay temperatures (30, 32, 34, 36, 38 °C) ×
six exposure durations (1, 5, 15, 45, 135, 405 minutes) = 30 cells per
dataset. Each cell is replicated by $n_{\text{reps}}$ independent cups,
each with $N \sim \text{Uniform}\{10, \dots, 20\}$ individuals.
A Scenario-1 dataset with $n_{\text{reps}}=5$ is therefore 150 cups
totalling roughly 2,000 individuals. Scenario 8 explicitly varies the
grid (sparse 3 T × 4 durations = 12 cells per dataset).

**Truth.** For every cell, $z$ and $CT_{\max,1\text{h}}$ are computed
analytically from the data-generating curve by evaluating the true
$\log_{10}\text{LT}_{50}$ at every design temperature and fitting a
straight line — the same population target both estimators converge to
under their estimating equations.

**Sample sizes.** Scenarios 1–5 use $N_{\text{sim}} = 1000$ datasets per
cell; Scenarios 6–8 (sensitivity sweeps) use $N_{\text{sim}} = 500$.
Cells are pre-flighted before any compute starts, including a
cross-cell distinctness check on data hashes that catches CLI
plumbing bugs.

**Performance metrics.** For each estimator and each output quantity
($z$, $CT_{\max,1\text{h}}$): mean signed bias and its Monte-Carlo SE,
RMSE, empirical 95% coverage and its MC SE (binomial proportion), and
median 95% CI width. All computed across the per-simulation rows in the
cell.

**Diagnostics.** Joint 4PL chains: 2 chains × 2,000 iterations
(half warm-up), `adapt_delta = 0.9`, `max_treedepth = 12`. R̂ < 1.01,
ESS > 400, zero divergent transitions, BFMI > 0.3. Per-cell pass rates
are reported alongside the bias / coverage tables.

# Scenario 1 — Strict equivalence baseline

**Question.** When all strict-equivalence assumptions hold (asymptotes
pinned at 0 and 1, no overdispersion, single slope across temperature),
do the three estimators agree, and is each calibrated? This is the
floor scenario — divergence here would signal a bug.

**Prediction.** All three estimators are correctly specified for this
DGP (the joint 4PL's beta-binomial likelihood collapses to binomial as
$\phi \to \infty$). Bias should be within MC SE and coverage near 0.95
for all three. We expected the joint 4PL to deliver the narrowest CIs
(joint fit pools information across temperature), with the two two-stage
variants near-identical (no overdispersion to exploit).

**Design.** DGP: `low = 0.001`, `up = 0.999`, `k = 8`, midpoint linear in
$T$ with slope $-0.15$, binomial likelihood. True $z = 6.667$, true
$CT_{\max,1\text{h}} = 32.146\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen1-table}
scen1 <- load_summaries("scen1")
make_table(scen1)
```

```{r}
#| label: fig-scen1
#| fig-cap: "Scenario 1 (strict equivalence baseline). (a) Mean bias with 95% MC-SE intervals; dashed line at zero. (b) Empirical 95% CI coverage with 95% binomial intervals; dashed line at the nominal 0.95. Top row: $n_{reps}=3$; bottom row: $n_{reps}=5$. Both quantities ($z$ and $CT_{\\max,1h}$) are within MC SE of zero bias and within MC SE of nominal coverage for all three estimators, as predicted."
#| fig-width: 8
#| fig-height: 5.5
plot_scen_forest(scen1)
```

**Read.** The bias prediction held for all three estimators: signed
bias is within MC SE of zero on both quantities in both cells. The
coverage prediction *did not* hold for the two-stage methods: the
binomial two-stage covers at 84.2–85.1% on both $z$ and
$CT_{\max,1\text{h}}$; the beta-binomial two-stage at 82.0–86.3%.
Joint 4PL coverage is 96.9–98.5% (slightly over nominal but within MC
SE of 0.95). The CI-width ordering matches prediction (two-stage
variants narrowest, joint 4PL wider) — but those narrower two-stage
intervals are the source of the under-coverage: Stage-2 OLS on five
LT50 points is propagating less variance than the underlying
sampling distribution actually has. **The two-stage's coverage gap
is structural and shows up even when its mean structure is correctly
specified.** A separate concern: the BB Stage-1 variant has 2.1–2.5%
of fits producing degenerate Stage-1 GLM solutions (n_ok = 975 / 979
vs 1000), inflating the unconditional RMSE on $z$ to 0.89–3.40 and on
$CT_{\max,1\text{h}}$ to 1.06–1.72. Stripping those fits is open issue 6
below.

# Scenario 2 — Likelihood misspecification only

**Question.** With identical curve shape to Scenario 1 but
beta-binomial overdispersion (φ = 5) added to the data-generating
likelihood, what does the field default lose by ignoring overdispersion?

**Prediction.** All three should remain unbiased — the curve shape is
unchanged, so the regression's mean structure is correct under all three
estimators. We expected the binomial two-stage to have wider CIs than
the patched (beta-binomial) two-stage, but coverage near nominal
because Stage-2 OLS reads the residual variance off the
$\log_{10}\text{LT}_{50}$ scatter. The patched two-stage should give
narrower, calibrated CIs. The joint 4PL — narrowest, calibrated.

**Design.** Same shape as Scenario 1 (boundary asymptotes), but DGP
likelihood is beta-binomial with φ = 5. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen2-table}
scen2 <- load_summaries("scen2")
make_table(scen2)
```

```{r}
#| label: fig-scen2
#| fig-cap: "Scenario 2 (likelihood misspecification only, beta-binomial DGP). All three estimators stay within MC SE of zero bias on both quantities. Coverage of the two two-stage variants drops to 84–90%, regardless of Stage-1 family; joint 4PL coverage stays near nominal 0.95."
#| fig-width: 8
#| fig-height: 5.5
plot_scen_forest(scen2)
```

**Read.** Bias predictions held: all three estimators are within MC SE
of zero on $z$ and $CT_{\max,1\text{h}}$. Coverage predictions did not
hold: both two-stage variants drop to 84–90%, regardless of whether
Stage 1 is binomial or beta-binomial. The patched Stage 1 does not
rescue the architecture under overdispersion — the Stage-2 OLS already
reads dispersion off the per-T point spread, but evidently
under-propagates it once five (or three) point estimates feed the
regression. The joint 4PL keeps coverage at 93–96%.

# Scenario 3 — Heat lowers max survival (asymmetric u drift)

**Question.** When the upper asymptote drops with temperature
(intrinsic heat-induced mortality reduces the proportion of the cohort
ever susceptible to the timed exposure) but neither two-stage estimator
has an asymptote parameter, does the missing structure get absorbed
into Stage 1's effective midpoint, biasing $z$ and $CT_{\max,1\text{h}}$?

**Prediction.** Both two-stage variants should be biased on at least
one quantity, with bias that does not shrink with $n_{\text{reps}}$.
Joint 4PL should be unbiased and calibrated.

**Design.** DGP: `up` drops 1 pp/°C from baseline 0.92, `low` constant
at 0.05, `k = 8`, midpoint linear in $T$ with slope $-0.15$,
beta-binomial likelihood (φ = 5). True $z = 6.537$, true
$CT_{\max,1\text{h}} = 32.124\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen3-table}
scen3 <- load_summaries("scen3")
make_table(scen3)
```

```{r}
#| label: fig-scen3
#| fig-cap: "Scenario 3 (asymmetric u drift, beta-binomial). Two-stage CTmax bias is systematically negative (~ −0.27 to −0.36 °C) and does not shrink with $n_{reps}$; coverage on the BB CTmax falls to 75% at $n_{reps}=5$. Joint 4PL bias is within MC SE on CTmax in both cells; small positive z bias persists (+0.06 to +0.10) but coverage stays at 94–96%."
#| fig-width: 8
#| fig-height: 5.5
plot_scen_forest(scen3)
```

**Read.** The two-stage CTmax bias is systematically negative across
both cells (−0.27 °C binomial, −0.36 °C beta-binomial) and does not
shrink with $n_{\text{reps}}$. Coverage on the BB CTmax drops to 75% at
$n_{\text{reps}}=5$. Mechanism: at high temperature the true `up` is
lower, so the per-T binomial GLM (which assumes the curve approaches 1)
fits a midpoint that is too late, dragging the Stage-2 intercept upward
and CTmax downward. The joint 4PL absorbs the T-varying `up` and stays
near-nominal on coverage; a small positive $z$ bias of +0.06 to +0.10
is the only residual deviation.

# Scenario 4 — Asymptotes compress (symmetric drift)

**Question.** When *both* asymptotes drift toward each other
symmetrically (`up` drops, `low` rises), the logit-difference is
nearly preserved but the curve range shrinks. How does each estimator
handle this in $z$ and $CT_{\max,1\text{h}}$?

**Prediction.** True $z$ barely changes (the slope around the midpoint
is preserved). Two-stage methods should show smaller bias than
Scenario 3 because the symmetric perturbation partially cancels at the
midpoint. Joint 4PL near-nominal.

**Design.** `up` drops 1 pp/°C from baseline 0.92; `low` rises 1 pp/°C
from baseline 0.05; `k = 8`; midpoint linear in $T$ with slope $-0.15$;
beta-binomial likelihood (φ = 5). True $z = 6.658$, true
$CT_{\max,1\text{h}} = 32.090\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen4-table}
scen4 <- load_summaries("scen4")
make_table(scen4)
```

```{r}
#| label: fig-scen4
#| fig-cap: "Scenario 4 (asymptotes compress). Two-stage z bias *flips sign relative to Scenario 3*: now positive (+0.39 binomial, +0.55 beta-binomial). CTmax bias remains negative and grows (~ −0.38 to −0.58 °C). Coverage on the BB CTmax falls to 64% at $n_{reps}=5$. Joint 4PL coverage holds at 94–95%."
#| fig-width: 8
#| fig-height: 5.5
plot_scen_forest(scen4)
```

**Read.** Two-stage CTmax bias is now larger than in Scenario 3
(−0.38 °C binomial, −0.58 °C beta-binomial) and the $z$ bias *flips
sign*: where Scenario 3 had two-stage $z$ biased low (−0.08 to −0.17),
Scenario 4 has it biased high (+0.39 to +0.56). Both biases are
insensitive to $n_{\text{reps}}$ — they are model misspecification, not
sampling noise. Coverage on the BB CTmax drops to 64.2% at $n_{\text{reps}}=5$,
the worst seen in the main scenarios. The patched Stage-1 (BB) is
consistently *worse* than the binomial Stage-1 here, contradicting the
prediction that the patch would help. Joint 4PL holds coverage at
94–95% with bias within MC SE on both quantities at $n_{\text{reps}}=5$.

# Scenario 5 — Heat sharpens curve (varying k)

**Question.** When the slope $k$ varies with temperature (curves get
sharper as it gets hotter), the per-T binomial GLM still recovers a
midpoint, but the Stage-2 OLS regression of $\log_{10}\text{LT}_{50}$
on $T$ assumes a linear relationship. How robust is the two-stage to
T-varying $k$?

**Prediction.** Smallest bias of the shape-perturbing scenarios. Two
stage variants: slight bias, coverage near nominal. Joint 4PL: bias
within MC SE, calibrated.

**Design.** `up = 0.92`, `low = 0.05`, $k$ linear in $T$ (8 at $T=34$,
slope +0.25 — so k = 7 at T=30, k = 9 at T=38). Midpoint linear in $T$
with slope $-0.15$. Beta-binomial likelihood (φ = 5). True $z = 6.679$,
true $CT_{\max,1\text{h}} = 32.084\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen5-table}
scen5 <- load_summaries("scen5")
make_table(scen5)
```

```{r}
#| label: fig-scen5
#| fig-cap: "Scenario 5 (varying k). Two-stage z biased high (+0.45 to +0.55), CTmax biased low (−0.44 to −0.62 °C); coverage on BB CTmax falls to 62% at $n_{reps}=5$. Joint 4PL bias within MC SE on z; small positive CTmax bias (+0.03, +0.06); coverage holds at 93–95%."
#| fig-width: 8
#| fig-height: 5.5
plot_scen_forest(scen5)
```

**Read.** The prediction was wrong: $k$ varying with $T$ produces
bias as large as Scenario 4, not the smallest of the shape scenarios.
Two-stage $z$ is biased high (+0.45 to +0.55) and CTmax biased low
(−0.44 to −0.62 °C); BB CTmax coverage falls to 61.6% at $n_{\text{reps}}=5$,
the worst across all scenarios. Mechanism: the Stage-1 single-shape
logistic is pulled towards the steepest curves at high temperature
(where data drops from `up` to `low` quickly), so the high-T
$\log_{10}\text{LT}_{50}$ estimates are pulled earlier and the Stage-2
slope is flattened. Joint 4PL absorbs the $k(T)$ drift; bias within
MC SE on $z$, +0.03–0.06 °C on CTmax, coverage 93–95%.

# Scenario 6 — Sweep: strength of u drift

**Question.** How does the two-stage bias scale with the strength of
the upper-asymptote drift? Is there a threshold of $|\beta_u|$ below
which the bias is negligible in practice?

**Prediction.** Bias monotonically increasing in $|\beta_u|$; coverage
correspondingly worse. Joint 4PL: bias within MC SE throughout.

**Design.** Same shape as Scenario 3 but vary the upper-asymptote
slope: $\beta_u \in \{-0.005, -0.010, -0.015, -0.019\}$ pp/°C.
$n_{\text{reps}} = 5$, $N_{\text{sim}} = 500$ per cell. Four cells.
True $z$ ranges 6.42–6.60; true $CT_{\max,1\text{h}}$ ranges 32.11–32.15.

```{r}
#| label: fig-scen6
#| fig-cap: "Scenario 6 sweep over $\\beta_u$ (strength of upper-asymptote drift). (a) Bias as a function of $\\beta_u$, with 95% MC-SE intervals; dashed line at zero. (b) Empirical 95% coverage; dashed line at nominal 0.95. Two-stage z bias *flips sign* between $\\beta_u=-0.005$ and $-0.010$. Two-stage CTmax bias is negative throughout but its absolute value *shrinks* as $|\\beta_u|$ grows; coverage on the same axis *improves* (intervals widen with the misspecification). Joint 4PL holds bias and coverage near-constant across the sweep."
#| fig-width: 8
#| fig-height: 6
scen6 <- load_summaries("scen6") %>%
  mutate(u_beta1 = -as.numeric(sub("scen6_ub_m", "", scenario)) / 1000)
plot_sweep(scen6, "u_beta1", expression(beta[u]~"(pp / °C)"),
           x_breaks = c(-0.005, -0.010, -0.015, -0.019))
```

**Read.** The monotonicity prediction was partly wrong. Two-stage $z$
*flips sign* between $\beta_u = -0.005$ (positive bias of +0.16 to +0.19)
and $\beta_u = -0.010$ (negative bias of −0.07 to −0.14), then becomes
increasingly negative. Two-stage CTmax bias stays negative throughout
but its absolute magnitude *shrinks* as $|\beta_u|$ grows. CI widths
inflate at stronger drift, which means coverage on CTmax *improves*
with $|\beta_u|$ even while bias remains. Coverage on $z$ does the
opposite: it *worsens* at stronger drift (down to 58% at $\beta_u=-0.019$
for the BB variant). The joint 4PL holds coverage at 94–96% across the
sweep.

# Scenario 7 — Sweep: level of upper asymptote

**Question.** Sublethal endpoints (PSII signal, fertility, knockdown)
rarely reach 1 even at $T \to T_{\min}$. How does each estimator's
performance scale with the baseline level of `up`?

**Prediction.** Two-stage $z$ bias should stay small (a constant
asymptote bias across temperature cancels in the Stage-2 slope).
$CT_{\max,1\text{h}}$ bias should grow as $u_0$ drops below 1 because
the per-T midpoint estimates do not coincide with the absolute
50%-survival time when `up` is below 1. Joint 4PL: bias within MC SE
throughout.

**Design.** `up` constant in $T$ at $u_0 \in \{0.65, 0.75, 0.85, 0.95, 0.99\}$;
`low = 0.05`, `k = 8`, midpoint linear in $T$ with slope $-0.15$.
Beta-binomial likelihood (φ = 5). $n_{\text{reps}} = 5$, $N_{\text{sim}} = 500$
per cell. Five cells. True $z = 6.667$ (constant by construction);
true $CT_{\max,1\text{h}}$ ranges 31.23 (at $u_0=0.65$) to 32.22 (at
$u_0=0.99$).

```{r}
#| label: fig-scen7
#| fig-cap: "Scenario 7 sweep over $u_0$ (baseline upper asymptote). (a) Bias and (b) coverage as a function of $u_0$. Two-stage $CT_{\\max,1h}$ bias explodes from +0.26 °C at $u_0=0.99$ to −6.33 °C at $u_0=0.65$, with coverage collapsing to 15%. Two-stage $z$ bias also grows, opposite to prediction. Joint 4PL coverage on $z$ stays at 94–95% across the sweep but its $CT_{\\max,1h}$ bias also grows (+0.67 °C at $u_0=0.65$); see open issues for the likely cause (relative vs absolute CTmax definition mismatch)."
#| fig-width: 8
#| fig-height: 6
scen7 <- load_summaries("scen7") %>%
  mutate(u0 = as.numeric(sub("scen7_u0_", "", scenario)) / 100)
plot_sweep(scen7, "u0", expression(u[0]),
           x_breaks = c(0.65, 0.75, 0.85, 0.95, 0.99))
```

**Read.** Two-stage $CT_{\max,1\text{h}}$ catastrophically biased at low
$u_0$: −6.33 °C at $u_0=0.65$, with coverage of 15%. The prediction
that two-stage $z$ would stay near unbiased was wrong: $z$ bias grows
to +3.1 (BB) at $u_0=0.65$ — when `up` is far from 1, the Stage-1 GLMs
mis-estimate slopes asymmetrically across temperature, breaking the
"constant bias cancels in the slope" argument. Joint 4PL $z$ holds at
near-zero bias across the sweep with coverage 94–95%, but its CTmax
bias grows from −0.03 (at $u_0=0.99$) to +0.67 (at $u_0=0.65$) with
coverage falling to 70%. **This last finding is most likely a
*definition* mismatch, not a model deficiency**: the analytical truth
function returns absolute-50%-survival CTmax, but the joint 4PL
extraction now returns the midpoint-based (relative-to-asymptotes)
CTmax by default. When `up` is well below 1 the two definitions
diverge by the asymmetry correction, which is exactly the gap reported
here. A re-run of `extract_tdt(threshold = 0.5, relative = FALSE)`
against the saved draws is the cleanest test (open issue 1 below).

# Scenario 8 — Sweep: design intensity

**Question.** When does each estimator's overhead pay off as the
sampling design gets sparser and replication drops?

**Prediction.** All methods: wider CIs as data sparsens. Joint 4PL has
smallest inflation (information shared across temperature). Two-stage
may fail or carry severe bias at the sparsest design.

**Design.** Two grids: full (5 T × 6 durations × $n_{\text{reps}}$
cups) or sparse (3 T × 4 durations × $n_{\text{reps}}$ cups). Three
$n_{\text{reps}}$ levels: $\{1, 3, 5\}$. Six cells. DGP: `up = 0.92`,
`low = 0.05`, `k = 8`, midpoint linear in $T$ with slope $-0.15$,
beta-binomial likelihood (φ = 5). True $z = 6.667$, true
$CT_{\max,1\text{h}} = 32.088\,^\circ\text{C}$. $N_{\text{sim}} = 500$ per
cell.

```{r}
#| label: fig-scen8
#| fig-cap: "Scenario 8 sweep over design intensity. Six cells: sparse vs full grid × $n_{reps} \\in \\{1, 3, 5\\}$. (a) Bias by cell. (b) Coverage by cell. Two-stage at the sparsest cell (sparse $\\times$ 1) gives nonsense (z RMSE > 60); joint 4PL has wide intervals at the same cell but coverage of 99% on $z$ and 95% on $CT_{\\max,1h}$. Two-stage CTmax coverage stays below 70% even at the most generous (full × 5) cell."
#| fig-width: 9
#| fig-height: 6
scen8 <- load_summaries("scen8") %>%
  mutate(
    design = ifelse(grepl("full", scenario), "full", "sparse"),
    n_reps = as.integer(sub(".*_n([0-9]+)$", "\\1", scenario)),
    cell   = factor(sprintf("%s × %d", design, n_reps),
                    levels = c("sparse × 1", "full × 1",
                               "sparse × 3", "full × 3",
                               "sparse × 5", "full × 5"))
  )

d <- prep_summary(scen8)
d <- d %>%
  mutate(cell = factor(sprintf("%s × %d",
                               ifelse(grepl("full", scenario), "full", "sparse"),
                               as.integer(sub(".*_n([0-9]+)$", "\\1", scenario))),
                       levels = c("sparse × 1", "full × 1",
                                  "sparse × 3", "full × 3",
                                  "sparse × 5", "full × 5")))

p_bias <- ggplot(d, aes(x = mean_bias, y = cell, colour = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = bias_lo, xmax = bias_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  labs(x = "Mean bias", y = NULL, title = "(a) Bias") +
  theme_proj

p_cov <- ggplot(d, aes(x = coverage, y = cell, colour = method)) +
  geom_vline(xintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = cov_lo, xmax = cov_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
  labs(x = "Empirical 95% coverage", y = NULL, title = "(b) Coverage") +
  theme_proj

p_bias / p_cov + plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "bottom")
```

**Read.** At sparse × 1 (12 cups, 1 replicate each), two-stage produces
nonsense — z RMSE 23–62 (binomial / BB), CTmax RMSE 12–30. Joint 4PL
returns wide but coherent intervals at the same cell ($z$ RMSE 2.3,
CTmax RMSE 0.9; coverage 99% / 95%). At the most generous cell
(full × 5), the two-stage CTmax coverage is still 59% (BB) — the design
budget cannot rescue the architecture under T-varying `up`. Joint 4PL
coverage holds across all six cells.

# Cross-cutting summary

```{r}
#| label: fig-summary
#| fig-cap: "Bias (top) and coverage (bottom) summaries for the five main scenarios at $n_{reps}=5$. Each point shows mean bias / empirical coverage with MC-SE intervals; dashed reference lines at zero (bias) and 0.95 (coverage). Joint 4PL is within MC SE of zero on bias and at-or-above 0.93 on coverage in every scenario. Two-stage methods accumulate substantial bias and lose coverage as soon as the DGP departs from strict equivalence (Scenarios 3–5)."
#| fig-width: 9
#| fig-height: 6.5

main <- bind_rows(
  load_summaries("scen1"),
  load_summaries("scen2"),
  load_summaries("scen3"),
  load_summaries("scen4"),
  load_summaries("scen5")
) %>%
  filter(grepl("_n5$", scenario)) %>%
  mutate(
    scenario_short = recode(sub("_n5$", "", scenario),
      scen1_strict_eq      = "1. Strict eq.",
      scen2_lik_misspec    = "2. Lik. misspec.",
      scen3_heat_lowers_u  = "3. Asym. u drift",
      scen4_compress       = "4. u + ell drift",
      scen5_sharpen        = "5. Varying k"
    )
  )

d <- prep_summary(main)

p_b <- ggplot(d, aes(x = mean_bias, y = scenario_short, colour = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = bias_lo, xmax = bias_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  labs(x = "Mean bias", y = NULL, title = "(a) Bias  (n_reps = 5)") +
  theme_proj

p_c <- ggplot(d, aes(x = coverage, y = scenario_short, colour = method)) +
  geom_vline(xintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = cov_lo, xmax = cov_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
  labs(x = "Empirical 95% coverage", y = NULL, title = "(b) Coverage  (n_reps = 5)") +
  theme_proj

p_b / p_c + plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "bottom")
```

# Joint 4PL convergence across scenarios

```{r diag-table}
diag <- bind_rows(
  load_diagnostics("scen1"),
  load_diagnostics("scen2"),
  load_diagnostics("scen3"),
  load_diagnostics("scen4"),
  load_diagnostics("scen5"),
  load_diagnostics("scen6"),
  load_diagnostics("scen7"),
  load_diagnostics("scen8")
)

diag %>%
  transmute(
    Cell                = scenario,
    `n sims`            = n_total,
    `n joint OK`        = n_joint_ok,
    `R̂ pass %`          = sprintf("%.1f%%", 100 * rhat_pass_rate),
    `ESS pass %`        = sprintf("%.1f%%", 100 * ess_pass_rate),
    `Diverg. pass %`    = sprintf("%.1f%%", 100 * divergence_pass_rate),
    `Treedepth pass %`  = sprintf("%.1f%%", 100 * treedepth_pass_rate),
    `BFMI pass %`       = sprintf("%.1f%%", 100 * bfmi_pass_rate)
  )
```

R̂ and ESS pass rates are above 96% in every cell; divergent transitions
are rare (≤ 154 across 1000–500 sims per cell). The conspicuous failure
is **treedepth saturation on essentially every joint-4PL fit** across
all 23 cells — the chain hits the `max_treedepth = 12` ceiling at every
post-warmup iteration. Coverage holds at near-nominal in 22 of 23 cells
despite this, but the issue is unresolved (open issue 2).

# Open issues

1. **Definition mismatch on joint-4PL CTmax in Scenario 7 (and likely
   contributing in Scenarios 3–5 and 8).** The analytical truth
   function evaluates absolute-50%-survival LT50 at every design
   temperature and fits OLS through those points; the joint 4PL
   `extract_tdt()` defaults to a midpoint-based (relative-to-asymptotes)
   CTmax. When `up` deviates from 1, these are not the same number, and
   the gap is exactly the asymmetry correction
   $\frac{1}{\beta_1 k}\ln\!\frac{\text{up}-0.5}{0.5-\text{low}}$.
   Diagnostic: re-extract from the saved draws with
   `extract_tdt(threshold = 0.5, relative = FALSE)` and recompute
   the bias against the truth. If the +0.67 °C bias at $u_0 = 0.65$
   collapses, the issue is labelling, not modelling. If it does not,
   the joint 4PL is genuinely deficient on absolute CTmax under
   constant low `up`.
2. **Treedepth saturation on every joint-4PL fit.** All 23 cells.
   Chains converge by R̂ / ESS / divergence criteria but hit
   `max_treedepth = 12`. A small re-run with `max_treedepth = 15` on
   one well-behaved cell (Scenario 1 n5) and one degenerate cell
   (Scenario 7 $u_0 = 0.65$) would tell us whether the saturation is
   masking interval underestimation.
3. **Bias conditional on full convergence.** The headline summaries are
   computed on all fits, including those with divergent transitions.
   The same summaries restricted to "all-pass" fits have not been
   computed yet. For Scenarios 3 (n3: 48 sims with divergences) and
   5 (n3: 59 sims), this is worth checking — whether the small positive
   $z$ bias is driven by a tail of pathological fits.
4. **Patched two-stage was *not* a patch.** Across Scenarios 2–5, the
   beta-binomial Stage-1 variant performs *worse* on coverage than the
   binomial Stage-1 variant, contradicting the prediction that
   correcting the likelihood would help. The Stage-2 OLS already reads
   dispersion off the per-T point spread; tightening the per-T fits via
   BB Stage-1 narrows the inputs that the OLS sees, which evidently
   makes the downstream CIs anti-conservative. Worth a separate note.
5. **Prior sensitivity** of the joint 4PL has not been varied. The
   small persistent positive $z$ bias of +0.04–0.10 across Scenarios
   1–5 might shift under wider priors on `mid_T_slope`.
6. **Beta-binomial Stage-1 outlier fits in Scenario 1.** 2.1–2.5% of
   sims (n_ok = 975 / 979 vs 1000) deliver degenerate Stage-1 GLM
   solutions whose Stage-2 OLS produces extreme estimates. RMSE on $z$
   inflates to 3.40 at $n_{\text{reps}}=5$ purely from this tail.
   Stripping those fits (or screening on Stage-1 GLM convergence) would
   give a cleaner picture of how the BB Stage-1 variant behaves on
   typical, well-converged data — and would let us separate the
   architecture-coverage gap from the GLM-failure gap.

# Bottom line

- When the strict-equivalence assumptions hold (Scenario 1), all three
  estimators are unbiased, but the two-stage variants already
  under-cover at 82–86% (vs nominal 95%) — the Stage-2 delta-method CIs
  are structurally too tight even when the mean structure is correctly
  specified. Joint 4PL covers at 97–99%.
- Adding overdispersion alone (Scenario 2) leaves bias untouched and
  shifts two-stage coverage another 2–6 percentage points; neither
  the binomial nor the beta-binomial Stage-1 variant restores nominal
  coverage. Joint 4PL coverage stays at 93–96%.
- Any structural drift in the asymptotes or slope (Scenarios 3–5)
  produces two-stage biases of 0.27–0.62 °C on $CT_{\max,1\text{h}}$
  and / or 0.4–0.6 on $z$, with coverage as low as 62%. The biases do
  not shrink with replication.
- The failure modes scale with the magnitude of the misspecification
  (Scenario 6) and the *baseline* level of partial mortality
  (Scenario 7) — the latter is the regime where most sublethal
  endpoints sit, and the two-stage $CT_{\max,1\text{h}}$ bias reaches
  6 °C with coverage of 15%.
- At the sparsest design × replication budgets (Scenario 8 sparse × 1),
  the two-stage produces estimates with no operational meaning. The
  joint 4PL returns intervals of appropriate width.
- One outstanding issue (definition mismatch on absolute vs relative
  CTmax) needs to be settled before the Scenario 7 joint-4PL CTmax
  numbers can be quoted in the manuscript.
