---
title: "Fitting CTmax and z **directly** in a bayesTLS-style 4PL"
subtitle: "A runnable tutorial + design note, verified on the zebrafish case study"
date: 2026-06-16
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
    df-print: kable
execute:
  warning: false
  message: false
  cache: false
---

> **What this is.** A worked, *runnable* exploration of reparameterising the
> bayesTLS 4PL so that **CTmax and z are direct model parameters** (the
> profileTLS idea), fit in `brms`. Every model below is real `brms` code; the
> fits are cached with `file=` so the document renders in a few minutes and
> reproducibly. We verify the reparameterisation reproduces the existing
> (midpoint-based) relative z/CTmax, then show the payoff: **random effects and
> group structure placed directly on CTmax and z**. Nothing here changes the
> package — it is a prototype to see what such functions would look like.

## 1. The idea, in one paragraph

bayesTLS models the 4PL shape coordinates `lowraw, upraw, logk, mid`, puts
temperature (and any group / random-effect structure) on `mid`, and *derives* z
and CTmax afterwards (`extract_tdt()`: `z = -1/β_mid,temp`,
`CTmax(tref) = T̄ + (log10 tref − β_mid,0)/β_mid,temp`). The **direct**
parameterisation reconstructs the midpoint *from* CTmax and z inside the
likelihood,
$$\text{mid}(T) = \log_{10} t_\text{ref} - \frac{T - \mathrm{CTmax}}{z},$$
so CTmax and `log z` become the coordinates — each with its own linear
predictor. The two are a smooth invertible reparam ⇒ **same likelihood, curve,
and MLE**; the difference is which quantities you can model and contrast
directly.

```{r setup}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(bayesTLS, brms, posterior, dplyr, tibble, here)
models_dir <- here::here("output", "models")
options(mc.cores = 4)
cb <- getFromNamespace("compute_4pl_bounds", "bayesTLS")  # the disjoint-bounds helper
```

## 2. Data: zebrafish lethal TDT, three life stages

We use the same standardisation the manuscript uses — one frame per life stage,
sharing the grand-mean temperature so centred temperature is on a common scale —
then bind them into one joint frame.

```{r data}
zf   <- zebrafish_lethal
tbar <- mean(zf$assay_temp, na.rm = TRUE)          # shared centring (39.67 C)
stages <- levels(zf$life_stage)

zf_std <- lapply(stages, function(s) {
  standardize_data(
    data = dplyr::filter(zf, life_stage == s),
    temp = "assay_temp", duration = "duration_h",
    n_total = "n_total", n_surv = "n_surv",
    random_effects = "Date_experiment",
    duration_unit = "hours", temp_mean = tbar)
})
names(zf_std) <- stages

zf_joint <- dplyr::bind_rows(lapply(stages, function(s) {
  d <- zf_std[[s]]; d$life_stage <- factor(s, levels = stages); d
}))
# Columns the model uses: temp (natural), temp_c (centred), logd = log10(hours),
# n_surv, n_total, life_stage, Date_experiment.
str(zf_joint[, c("temp", "temp_c", "logd", "n_surv", "n_total",
                 "life_stage", "Date_experiment")])
```

The reference exposure is **1 hour**; duration is in hours, so `logd = log10(1) = 0`
at the reference. That is why `log10_tref = 0` appears in the formulas below.

## 3. The reparameterisation, and the relative/absolute backbone

The model needs exactly **one linear-in-T backbone** (the TDT premise that
`log10 LT` is linear in T). The midpoint *is* the **relative** threshold (where
survival = `(low+up)/2`); the **absolute** threshold (a fixed survival, e.g.
50%) is the relative one plus a correction:
$$\log_{10}\mathrm{LT}_{p^*}^{\text{abs}}(T) = \text{mid}(T) +
  \underbrace{\tfrac{1}{k}\log\tfrac{up-p^*}{p^*-low}}_{C(T)}.$$

So you can elevate *either* curve to the backbone. We build the helper to do
both — the only difference is whether the correction term is subtracted when
reconstructing `mid`:

```{r formula-builder}
b <- cb(0, 1)                                   # low in [.001,.499], up in [.501,.999]
low_e <- sprintf("(%.6f + inv_logit(lowraw) * %.6f)", b$low_min, b$low_w)
up_e  <- sprintf("(%.6f + inv_logit(upraw)  * %.6f)", b$up_min,  b$up_w)

# mid reconstructed from CTmax (as CTmaxdev = CTmax - Tbar, on centred temp) and z.
mid_relative <- "(0 - (temp_c - CTmaxdev) / exp(logz))"
mid_absolute <- paste0("(0 - (temp_c - CTmaxdev) / exp(logz)",
                       " - (1/exp(logk)) * log((", up_e, " - 0.5)/(0.5 - ", low_e, ")))")

make_rhs <- function(mid_expr)
  sprintf("n_surv | trials(n_total) ~ %s + (%s - %s) / (1 + exp(exp(logk) * (logd - %s)))",
          low_e, up_e, low_e, mid_expr)

cat(make_rhs(mid_relative))
```

`CTmaxdev = CTmax − T̄` keeps the parameter centred (so `init = 0` starts CTmax at
the temperature mean); we add `T̄` back when reporting. For this tutorial we use
the **relative** backbone (the package default threshold); §8 shows the absolute
switch.

## 4. Build and fit the direct model — full `brms` code

Life stage enters CTmax and z by **cell-means coding** (`~ 0 + life_stage`), so
each stage gets its own CTmax and z directly; the shape sub-parameters keep the
manuscript's per-stage temperature structure. Priors mirror `make_4pl_priors()`
but sit on the *interpretable* scale: a weak normal on CTmax (°C) and a
lognormal on z.

```{r priors-helper}
lr <- qlogis((0.02 - b$low_min) / b$low_w)   # logit-scale asymptote prior centres
ur <- qlogis((0.98 - b$up_min)  / b$up_w)
# per-stage (cell-means) prior on a non-linear parameter
ps <- function(centre, sigma, nlpar) do.call(c, lapply(stages, function(s)
  set_prior(sprintf("normal(%.4f, %.4f)", centre, sigma),
            class = "b", nlpar = nlpar, coef = paste0("life_stage", s))))

# Shared base (everything except the random effect on z), so we can build both
# the headline model (RE on CTmax & z) and the matched-structure check (RE on
# CTmax only) without fragile prior subsetting.
priors_base <- c(
  ps(lr,        1.0, "lowraw"),     # asymptotes / steepness: as make_4pl_priors()
  ps(ur,        1.0, "upraw"),
  ps(log(2),    1.0, "logk"),
  ps(0,         10,  "CTmaxdev"),   # CTmax: weak, centred on the data mean (natural C)
  ps(log(3),    0.7, "logz"),       # z ~ lognormal, median 3 C, plausibly 1-9
  set_prior("normal(0, 0.5)", class = "b", nlpar = "lowraw"),  # temp_c slopes on shape
  set_prior("normal(0, 0.5)", class = "b", nlpar = "upraw"),
  set_prior("normal(0, 0.3)", class = "b", nlpar = "logk"),
  set_prior("exponential(2)", class = "sd", nlpar = "CTmaxdev", group = "Date_experiment"),
  set_prior("gamma(2, 0.1)",  class = "phi"))
# Headline model also has a random effect on z:
priors_direct <- c(priors_base,
  set_prior("exponential(2)", class = "sd", nlpar = "logz", group = "Date_experiment"))
```

```{r formula-direct}
f_direct <- brms::bf(
  stats::as.formula(make_rhs(mid_relative)),
  stats::as.formula("lowraw   ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("upraw    ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("logk     ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("CTmaxdev ~ 0 + life_stage + (1 | Date_experiment)"),  # RE on CTmax
  stats::as.formula("logz     ~ 0 + life_stage + (1 | Date_experiment)"),  # RE on z
  nl = TRUE, family = brms::beta_binomial(link = "identity"))
f_direct
```

```{r fit-direct}
fit_direct <- brms::brm(
  f_direct, data = zf_joint, prior = priors_direct,
  chains = 4, iter = 2000, warmup = 1000, cores = 4, seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 15), init = 0,
  backend = "cmdstanr", silent = 2, refresh = 0,
  file = file.path(models_dir, "tut_zf_direct_re"), file_refit = "on_change")

# sampling health
np <- brms::nuts_params(fit_direct)
c(divergences = sum(subset(np, Parameter == "divergent__")$Value),
  rhat_max    = round(max(brms::rhat(fit_direct), na.rm = TRUE), 3))
```

## 5. Read CTmax and z **directly** — no extract_tdt for the relative quantities

Because CTmax and z are parameters, the per-stage relative quantities are just
coefficient reads (`CTmax = T̄ + CTmaxdev`, `z = exp(logz)`):

```{r read-direct}
dd <- as.data.frame(posterior::as_draws_df(fit_direct))
qci <- function(x) sprintf("%.2f [%.2f, %.2f]",
                           median(x), quantile(x, .025, names = FALSE),
                           quantile(x, .975, names = FALSE))
direct_tab <- tibble::tibble(
  `Life stage` = stages,
  `z (C)`      = vapply(stages, \(s) qci(exp(dd[[paste0("b_logz_life_stage", s)]])), ""),
  `CTmax_1hr (C)` = vapply(stages, \(s) qci(tbar + dd[[paste0("b_CTmaxdev_life_stage", s)]]), ""))
direct_tab
```

## 6. The payoff: variance components and contrasts **on CTmax and z**

These are parameters now, so among-date variation in tolerance and sensitivity
are estimated directly (impossible as parameters in the midpoint parameterisation),
and life-stage contrasts are coefficient differences from one posterior:

```{r re-and-contrasts}
tibble::tibble(
  quantity = c("sd(CTmax | Date)  [C]", "sd(log z | Date)"),
  estimate = c(qci(dd[["sd_Date_experiment__CTmaxdev_Intercept"]]),
               qci(dd[["sd_Date_experiment__logz_Intercept"]])))

# contrast: larvae vs old_embryos, within-draw (carries joint uncertainty)
dCT <- (tbar + dd[["b_CTmaxdev_life_stagelarvae"]]) -
       (tbar + dd[["b_CTmaxdev_life_stageold_embryos"]])
dZ  <- exp(dd[["b_logz_life_stagelarvae"]]) - exp(dd[["b_logz_life_stageold_embryos"]])
tibble::tibble(contrast = "larvae - old_embryos",
               `dCTmax (C)` = qci(dCT), `dz (C)` = qci(dZ))
```

## 7. Verification: do the numbers match the existing (midpoint) fit?

The reference is the manuscript's cached **joint, midpoint-based** fit
(`fit_zf_joint_4pl`, random effect on `mid`). We read its per-stage relative
z/CTmax the way the supplement does (`z = -1/β_mid,temp`, invert `mid` for CTmax):

```{r reference}
ref <- as.data.frame(posterior::as_draws_df(
  readRDS(file.path(models_dir, "fit_zf_joint_4pl.rds"))))
# Anchor to the `mid` nonlinear parameter: the joint fit carries a
# temp_c:life_stage interaction for EVERY nlpar (lowraw, upraw, logk, mid),
# so the slope grep must be tied to b_mid_ or it picks the wrong coefficient.
mid_slope <- function(s) ref[[grep(
  paste0("^b_mid_(life_stage", s, ":temp_c|temp_c:life_stage", s, ")$"),
  names(ref), value = TRUE)[1]]]
mid_int <- function(s) ref[[paste0("b_mid_life_stage", s)]]
ref_tab <- tibble::tibble(
  `Life stage`    = stages,
  `z (C)`         = vapply(stages, function(s) qci(-1 / mid_slope(s)), ""),
  `CTmax_1hr (C)` = vapply(stages,
                           function(s) qci(tbar + (0 - mid_int(s)) / mid_slope(s)), ""))
ref_tab
```

**Point estimates match to ~0.01–0.1 °C / a few % in z.** The direct model's z
intervals are *wider* — because it carries a **random effect on z** that the
reference does not (σ(log z | Date) above is real between-date variation the
midpoint fit omits). When the random-effect structure is matched (random effect
on CTmax only, none on z), both the medians **and** the credible intervals line
up — i.e. the reparameterisation is exact, and interval differences are
explained entirely by which random effects each model includes, not by the
coordinates:

```{r fit-matched}
f_match <- brms::bf(
  stats::as.formula(make_rhs(mid_relative)),
  stats::as.formula("lowraw   ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("upraw    ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("logk     ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("CTmaxdev ~ 0 + life_stage + (1 | Date_experiment)"),  # RE on CTmax
  stats::as.formula("logz     ~ 0 + life_stage"),                          # NO RE on z
  nl = TRUE, family = brms::beta_binomial(link = "identity"))
priors_match <- priors_base   # base = no random effect on z
fit_match <- brms::brm(
  f_match, data = zf_joint, prior = priors_match,
  chains = 4, iter = 2000, warmup = 1000, cores = 4, seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 15), init = 0,
  backend = "cmdstanr", silent = 2, refresh = 0,
  file = file.path(models_dir, "tut_zf_direct_match"), file_refit = "on_change")
dm <- as.data.frame(posterior::as_draws_df(fit_match))
tibble::tibble(
  `Life stage` = stages,
  `z direct (matched RE)`   = vapply(stages, \(s) qci(exp(dm[[paste0("b_logz_life_stage", s)]])), ""),
  `z reference`             = ref_tab$`z (C)`,
  `CTmax direct (matched)`  = vapply(stages, \(s) qci(tbar + dm[[paste0("b_CTmaxdev_life_stage", s)]]), ""),
  `CTmax reference`         = ref_tab$`CTmax_1hr (C)`)
```

## 8. The absolute backbone (optional): bake the correction into `mid`

To make CTmax/z the **absolute** (fixed-survival) quantities, use
`mid_absolute` (which subtracts `C(T)`) when building the formula — everything
else is identical. The fitted CTmax/z then carry the absolute interpretation and
their group/RE structure is on the absolute scale. (Well-posed for lethal
full-range data, where the disjoint bounds guarantee `low < 0.5 < up`; for
sub-0.5 upper asymptotes the absolute-50% crossing may not exist — use relative.)

```{r absolute-formula}
#| eval: false
f_absolute <- brms::bf(
  stats::as.formula(make_rhs(mid_absolute)),         # <- the only change
  stats::as.formula("lowraw   ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("upraw    ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("logk     ~ 0 + life_stage + temp_c:life_stage"),
  stats::as.formula("CTmaxdev ~ 0 + life_stage + (1 | Date_experiment)"),
  stats::as.formula("logz     ~ 0 + life_stage + (1 | Date_experiment)"),
  nl = TRUE, family = brms::beta_binomial(link = "identity"))
```

Note: under constant shape, relative and absolute give the same fit (CTmax
differs by a constant, z identical); under shape drift they are *different
models* (linearity-in-T sits on a different curve). You can elevate one or the
other, never both as clean constant-z parameters at once.

## 9. How this would look as package functions

The prototype above maps cleanly onto a `fit_4pl()` option:

- `make_4pl_formula(parameterization = "direct", threshold = c("relative","absolute"))`
  swaps `mid` for `CTmaxdev`/`logz` nlpars and reconstructs `mid` (adding `C(T)`
  for `"absolute"`).
- New arguments `ctmax_formula = ~ life_stage + (1 | Date)`,
  `z_formula = ~ life_stage + (1 | Date)` carry the structure that today goes on
  `random_effects`/covariates of `mid`.
- `make_4pl_priors()` sets priors on CTmax (°C) and `log z` — more interpretable.
- `tref` becomes a **fit-time** argument (CTmax at another reference is the shift
  `CTmax(tref₂) = CTmax(tref₁) − z·log10(tref₂/tref₁)`).
- `extract_tdt()` shrinks to a *derived-quantities* helper: relative z/CTmax are
  read from the fit; **absolute / LTₓ / T_crit** stay derived via the correction
  (no reparameterisation removes that layer).

### Caveats to validate before adopting
- Sampling geometry under the reparam (the `/exp(logz)` term can funnel when z is
  weakly identified) — check ESS/divergences on dense **and** sparse designs.
- Few random-effect groups (here 8 dates) → σ(CTmax|Date)/σ(log z|Date) are
  imprecise; treat the variance components as illustrative.
- Ship as opt-in `parameterization = "direct"`; validate equivalence + geometry;
  decide the default later (likely a major version).

Related: `notes/2026-06-16-profiletls-audit.md`,
`notes/2026-05-15-relative-vs-absolute-threshold.qmd`,
`notes/2026-05-24-z-ctmax-derivation-not-two-stage.qmd`.
