---
title: "Prior sensitivity of the direct CTmax/z 4PL: the lower-asymptote funnel"
subtitle: "Why the asymptote priors can't be naively flattened"
date: 2026-06-21
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: show
---

## Question

The direct CTmax/z parameterisation
(`notes/2026-06-16-direct-ctmax-z-parameterisation.qmd`) carries a fairly long
prior block — a per-life-stage (cell-means) prior on each 4PL sub-parameter plus
separate priors for the temperature slopes. Daniel asked whether this can be
collapsed to a single generic weakly-informative prior per parameter.

**Short answer: no — not for the asymptotes.** Flattening them turns a clean fit
(0 divergences) into a badly-behaved one (≈150–190 divergences) because the
lower asymptote is *identified only at its boundary*, and the prior is the only
thing supplying curvature there. Increasing `adapt_delta` does **not** fix it,
confirming the problem is the posterior geometry (the prior), not the sampler
step size.

All numbers below are read live from the cached zebrafish fits.

```{r setup}
#| message: false
#| warning: false
pacman::p_load(brms, posterior, dplyr, tibble, purrr, here)
models_dir <- here::here("output", "models")
inv_logit  <- function(x) 1 / (1 + exp(-x))
low_of     <- function(raw) 0.001 + inv_logit(raw) * 0.498   # disjoint-bounds lower asymptote
stages     <- c("young_embryos", "old_embryos", "larvae")
```

## The two prior regimes

The model is identical in all fits (relative direct CTmax/z 4PL, beta-binomial,
random effects on CTmax and z; see the 2026-06-16 note). Only the priors on the
**asymptotes / steepness** differ.

```{r priors-shown}
#| eval: false
# CENTRED (package-style) — asymptotes pinned near survival 0/1:
prior(normal(-3.23, 1), class = b, nlpar = lowraw, coef = life_stage<each>)  # low ~ 0.02
prior(normal( 3.23, 1), class = b, nlpar = upraw,  coef = life_stage<each>)  # up  ~ 0.98
prior(normal(log(2), 1), class = b, nlpar = logk,  coef = life_stage<each>)
prior(normal(0, 0.5), class = b, nlpar = lowraw)   # temp slopes
prior(normal(0, 0.5), class = b, nlpar = upraw)
prior(normal(0, 0.3), class = b, nlpar = logk)

# GENERIC — one weakly-informative prior per parameter, centred at survival 0.5:
prior(normal(0, 1.5), class = b, nlpar = lowraw)   # prior-median low = 0.25
prior(normal(0, 1.5), class = b, nlpar = upraw)    # prior-median up  = 0.75
prior(normal(0, 1),   class = b, nlpar = logk)
# (CTmaxdev ~ normal(0,10), logz ~ normal(log 3, 0.7), RE-SDs, phi: identical in both)
```

The `normal(0, 1.5)` on `lowraw` expects the survival curve to bottom out at
**25 %**, not at ~0. For lethal full-range data (everything dies at long
exposure × high temperature) that is the wrong centre.

## Three fits, compared

```{r diagnostics}
#| message: false
#| warning: false
fits <- c(
  "centred N(±3.23,1), ad=0.99"  = "tut_zf_direct_re",
  "generic N(0,1.5), ad=0.99"    = "tut_zf_direct_simplepriors",
  "generic N(0,1.5), ad=0.999"   = "tut_zf_direct_simplepriors_ad999")

diag_tbl <- purrr::imap_dfr(fits, function(file, lab) {
  f  <- readRDS(file.path(models_dir, paste0(file, ".rds")))
  np <- brms::nuts_params(f)
  s  <- posterior::summarise_draws(posterior::as_draws_df(f), "rhat", "ess_bulk")
  tibble::tibble(
    `prior / sampler` = lab,
    divergences  = sum(subset(np, Parameter == "divergent__")$Value),
    `max R-hat`  = round(max(s$rhat, na.rm = TRUE), 3),
    `min ESS`    = round(min(s$ess_bulk, na.rm = TRUE)))
})
diag_tbl
```

The centred priors give **0 divergences**. The generic priors give ~190
divergences at `adapt_delta = 0.99`, and still ~140 at `adapt_delta = 0.999`
with a deeper tree — a small reduction, nowhere near zero. (The centred fit's
*min ESS* is actually lower, but that is its random-effect SD `sd(CTmax | Date)`,
which is inherently imprecise with only 8 date groups — a benign kind of
"slow", not divergent exploration.)

## Why: a boundary funnel, not a step-size problem

On the disjoint-bounds scale `low = 0.001 + inv_logit(lowraw) · 0.498`. The data
want `low ≈ 0`, which means `lowraw → −∞`, where `inv_logit` **saturates** and
the likelihood becomes flat in `lowraw` — the data cannot distinguish
`low = 0.005` from `low = 0.001`. That is a classic funnel: a direction the
likelihood does not constrain, where only the prior supplies curvature.

- A **tight centred prior** `N(−3.23, 1)` provides that curvature → the sampler
  stays in the identified region → 0 divergences.
- A **wide generic prior** `N(0, 1.5)` is mis-centred (pulls toward `low = 0.25`)
  *and* too diffuse to regularise the saturated tail → it funnels.

Because the funnel is a property of the posterior geometry, shrinking the step
size (`adapt_delta`) can only chip at it — hence 191 → 143, not 191 → 0.

```{r asymptote-comparison}
#| message: false
#| warning: false
asym <- purrr::imap_dfr(fits[c(1, 2)], function(file, lab) {
  d <- as.data.frame(posterior::as_draws_df(readRDS(file.path(models_dir, paste0(file, ".rds")))))
  purrr::map_dfr(stages, function(s) {
    raw <- d[[paste0("b_lowraw_life_stage", s)]]
    tibble::tibble(prior = lab, stage = s,
      `low (median)` = round(median(low_of(raw)), 3),
      `low 97.5%`    = round(quantile(low_of(raw), .975, names = FALSE), 3))
  })
})
tidyr::pivot_wider(asym, names_from = prior,
                   values_from = c(`low (median)`, `low 97.5%`))
```

Under the generic prior the estimated survival floor drifts **up** and widens
(e.g. old_embryos' upper credible bound reaches ~0.17 vs ~0.03 under the centred
prior), and that asymptote uncertainty propagates into a slightly inflated,
wider `z`.

```{r z-comparison}
#| message: false
#| warning: false
zt <- purrr::imap_dfr(fits, function(file, lab) {
  d <- as.data.frame(posterior::as_draws_df(readRDS(file.path(models_dir, paste0(file, ".rds")))))
  purrr::map_dfr(stages, function(s) {
    z <- exp(d[[paste0("b_logz_life_stage", s)]])
    tibble::tibble(prior = lab, stage = s,
      z = sprintf("%.2f [%.2f, %.2f]", median(z),
                  quantile(z, .025, names = FALSE), quantile(z, .975, names = FALSE)))
  })
})
tidyr::pivot_wider(zt, names_from = prior, values_from = z)
```

## Conclusion and recommendation

1. The asymptote priors are **regularising a boundary-identified parameter**, not
   injecting biological opinion. They are load-bearing.
2. The funnel is **structural** (geometry/prior), confirmed by the near-null
   effect of raising `adapt_delta` to 0.999. Tuning the sampler is the wrong
   lever; the prior is the right one.
3. The only genuine verbosity is the **per-stage repetition forced by cell-means
   coding** (`~ 0 + life_stage`). `CTmaxdev` and `logz` (no temperature slope)
   collapse safely to one no-`coef` prior each. The asymptote/steepness blocks
   cannot, unless the *model* drops temperature dependence on those parameters
   (then each is a single coefficient — still wanting a centred prior).
4. Practical guidance for the package / tutorial: keep centred asymptote priors;
   do **not** offer a flat `N(0, 1.5)`-style default. If a "simple priors" mode
   is ever exposed, it must still centre `low` near 0 and `up` near 1.

Related: `notes/2026-06-16-direct-ctmax-z-parameterisation.qmd` (the model and
the package-style priors).
