---
title: "Why you can't just multiply per-interval survival probabilities — and what Rezende and Jørgensen actually do"
date: 2026-05-07
author: "Daniel W.A. Noble (clarification dialogue with Claude)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
---

## Question

Given a dose-response model that returns $S(T, \tau)$ — survival probability at temperature $T$ for duration $\tau$ — the natural intuition is: for a temperature time series, compute the per-interval survival $p_i = S(T_i, \Delta i)$ at each step and multiply: $S_{\text{total}} = \prod_i p_i$. Is this what `@rezende_drosophila_mortality_2020` and `@jorgensen_unifying_2021` do? Can the joint Bayesian 4PL be used in this way?

## Short answer

**No — and the reason matters.** The naive product overestimates survival because it pretends the organism resets to a fresh state at the start of every interval. The dose-response $S(T, \tau)$ is a *marginal* probability conditional on starting at $\tau = 0$; multiplying marginals across consecutive intervals double-counts the protective effect of being fresh.

Rezende 2020 and Jørgensen 2021 instead track a single running quantity — the **effective cumulative exposure time at a reference temperature** — and evaluate the dose-response *once* at the end. This handles the damage history correctly.

## Why naive multiplication fails

### The mathematical setup

The 4PL gives $S(T, \tau)$, the probability of surviving from a fresh start at temperature $T$ for duration $\tau$. By definition, $S(T, 0) = 1$ (no time elapsed, no damage), $S(T, \infty) = 0$ at temperatures where prolonged exposure is lethal, and $S(T, LT_{50}(T)) = 0.5$.

The naive proposal is

$$
S_{\text{naive}}(t) \;\stackrel{?}{=}\; \prod_i S(T_i, \Delta i)
$$

across discretised intervals.

### A concrete example showing the error

Take a constant exposure at $T = 40\,°\text{C}$ where $LT_{50} = 1$ h, so by definition $S(40, 1\text{ h}) = 0.5$. With a 4PL of moderate steepness, $S(40, 30\,\text{min}) \approx 0.85$ (more time elapsed at lethal $T$ → less alive, but at half the LT50 we're still well above the inflection).

Splitting the 1-hour exposure into two 30-minute intervals and multiplying marginals:

$$
S_{\text{naive}}(60\,\text{min}) \;=\; S(40, 30) \times S(40, 30) \;=\; 0.85 \times 0.85 \;=\; 0.72.
$$

But the truth, by definition of $LT_{50}$, is $S(40, 60) = 0.50$. The naive product overestimates survival by $\approx 0.22$ in absolute terms — a 44 % relative overestimate.

The discrepancy widens further when more intervals are taken (60 × 1-min intervals at 40 °C would give an even bigger naive over-estimate).

### Why this happens

The product $\prod_i S(T_i, \Delta i)$ implicitly assumes each interval starts on a fresh, fully-healthy organism. But after interval 1 the organism has accumulated damage; its conditional survival in interval 2 is *lower* than the fresh-start marginal $S(T_2, \Delta i_2)$ would suggest.

The correct conditional, by chain rule, is

$$
P(\text{survive interval } i \mid \text{alive at start}) \;=\; \frac{S(T_i, \tau_{\text{cumulative}}(i) + \Delta i)}{S(T_i, \tau_{\text{cumulative}}(i))},
$$

where $\tau_{\text{cumulative}}(i)$ is the dose-equivalent exposure already accumulated by the start of interval $i$ at the current temperature $T_i$. The numerator is the marginal survival up through the *end* of interval $i$ at $T_i$; the denominator is the marginal at the *start*. Multiplying these conditional ratios across all intervals telescopes correctly back to the true marginal at the end.

For our example:

$$
\frac{S(40, 60)}{S(40, 30)} \;=\; \frac{0.50}{0.85} \;\approx\; 0.59 \quad\text{(conditional survival of interval 2)}
$$

$$
0.85 \times 0.59 \;=\; 0.50 \quad\text{✓ (matches the truth)}
$$

So the chain-of-conditionals approach is correct. But it's awkward to implement directly under variable $T(t)$, because at each step you need to know what fresh-start time at $T_i$ is equivalent to your current damaged state.

## What Rezende 2020 and Jørgensen 2021 actually do

Both papers use the **survival-probability shift** identity (Rezende 2020 Eq. 2):

$$
S_{T_1}(\tau) \;=\; S_{T_2}\!\big(10^{(T_2 - T_1)/z}\,\tau\big).
$$

In words: surviving $\tau$ at $T_1$ is *equivalent in damage* to surviving $10^{(T_2 - T_1)/z}\,\tau$ at $T_2$. Hot temperatures translate into longer effective exposure at a reference $T_2$; cool temperatures, into shorter effective exposure.

Using this identity, you can pick any reference temperature $T_{\text{ref}}$ and accumulate the **effective cumulative exposure time** under any temperature trajectory:

$$
\tau_{\text{eff}}(t) \;=\; \int_0^t 10^{\,(T(i)\,-\,T_{\text{ref}})/z}\,di.
$$ {#eq-tau-eff}

Each instant $di$ at temperature $T(i)$ contributes $10^{(T(i) - T_{\text{ref}})/z}\,di$ to the running effective-time accumulator. Hot moments add a lot; cool moments add a little; moments below $T_c$ add nothing.

Survival at any time $t$ is then read off the dose-response surface *once*:

$$
S_{\text{pred}}(t) \;=\; S\!\big(T_{\text{ref}},\, \tau_{\text{eff}}(t)\big).
$$ {#eq-S-pred}

This is the formulation the joint Bayesian 4PL plugs into directly: $\tau_{\text{eff}}(t)$ accumulates from the temperature trace and the posterior draw of $z$; $S(T_{\text{ref}}, \tau_{\text{eff}}(t))$ comes from the posterior draw of $(\ell, u, k, m(T_{\text{ref}}))$. Each posterior draw produces a complete predicted survival trajectory under the field exposure.

### Connection to the canonical HI integral

The HI integral `@eq-hi` is a one-summary-statistic version of `@eq-tau-eff` and `@eq-S-pred` together. It measures *cumulative dose* in units of one $LT_{50}$ at a 1-hour reference. When $HI(t) = 100\,\%$, $\tau_{\text{eff}}(t)$ has reached one $LT_{50,\,1\text{h}}$, and $S_{\text{pred}}(t) = 0.5$ regardless of how the trajectory got there. HI compresses the entire survival trajectory into a single threshold-crossing time; the survival-surface simulation gives the entire curve.

Both use the dose-response model. Both use the same $z$ and the same TDT machinery. The HI formulation is the *summary-statistic* form; the survival-surface formulation is the *full-trajectory* form.

## Why this matters for the manuscript's framing

Daniel's "multiply marginals" intuition is useful pedagogically because it frames the framework's substantive contribution: **the joint Bayesian 4PL gives you the full $S(T, \tau)$ surface, but using it correctly under variable $T(t)$ requires effective-time tracking, not naive multiplication.** This is exactly the kind of subtlety that the two-stage pipeline (which collapses each $T$ to a single $LT_{50}$) cannot represent — but using the 4PL "wrong" (naive multiplication) gives a worse answer than even the two-stage canonical HI, which at least handles the damage-additivity correctly within its scalar summary.

A short Discussion paragraph or Box could set this up explicitly: the 4PL supports richer survival-trajectory predictions via @eq-tau-eff and @eq-S-pred, but the naive "multiply per-interval $S$" shortcut is incorrect — and seeing why is one of the most useful didactic exercises in the framework.

## Yes — you *can* use the predict function directly, two equivalent ways

The intuition "use the model's predict function on each (temperature, duration) pair to get probabilities, then combine them while keeping track of cumulative survival impact" is right — but the *combine* step is what matters, and naive multiplication is not it. Two equivalent correct approaches both use the predict function (or the underlying $S(T, \tau)$ closed form), and both give the same answer.

### Approach A — effective-time accumulator, predict once

For each interval $i$ with temperature $T_i$ and duration $\Delta i$, contribute an *effective-time increment* at a reference temperature $T_{\text{ref}}$:

$$
\Delta \tau_{\text{eff},\,i} \;=\; \Delta i \cdot 10^{\,(T_i - T_{\text{ref}})/z}.
$$

Sum to get the running $\tau_{\text{eff}}(t) = \sum_{i \leq t} \Delta\tau_{\text{eff},\,i}$. Then predict survival *once*:

$$
S_{\text{pred}}(t) \;=\; S\!\big(T_{\text{ref}},\, \tau_{\text{eff}}(t)\big).
$$

This is the form @rezende_drosophila_mortality_2020 and @jorgensen_unifying_2021 use directly.

### Approach B — track current $S$, invert-and-re-predict

Start with $S = 1$ at $t = 0$ (fresh). For each interval $i$ with temperature $T_i$ and duration $\Delta i$:

1. **Invert** the 4PL: find the fresh-start time at $T_i$ that would have produced the current $S$. Algebraically, $\tau_{\text{at } T_i} = 10^{m(T_i)\,+\,(1/k)\log_{10}((u-S)/(S-\ell))}$.
2. **Add** the interval duration: $\tau_{\text{new}} = \tau_{\text{at } T_i} + \Delta i$.
3. **Predict** new survival at $(T_i, \tau_{\text{new}})$: $S \leftarrow S(T_i, \tau_{\text{new}})$.

This is the most direct "predict-function-only" recipe — it never references $z$ explicitly, only the 4PL surface.

### Why A and B are the same

Approaches A and B are mathematically identical. The cross-temperature scaling needed in B's step 1 ($\tau_{\text{at } T_i}$ at the current damage state) falls out of the relationship between $m(T_i)$ at consecutive temperatures, which equals $-(T_i - T_{\text{ref}})/z$ by construction of the linear midpoint regression. So B doesn't avoid $z$ — it's just baked into $m(T)$ rather than explicit in the integrand.

Use whichever is easier to code. Approach A needs one cumulative-sum and one predict call. Approach B needs an invert-predict-update loop but only ever uses temperature-current 4PL evaluations.

## Simulation: naive product vs three correct approaches that all give the same answer

The simulation below uses a hypothetical organism with 4PL parameters $\ell = 0$, $u = 1$, $k = 4$, and $\log_{10} LT_{50}(T) = \log_{10}(60) - (T - 36)/z$ with $z = 4$ °C — i.e., $LT_{50} = 60$ min at $T = 36$ °C, and $LT_{50}$ shifts by a factor of 10 for every 4 °C. So $CT_{max_{1\text{hr}}} = 36$ °C in this organism. A 5-day sinusoidal hourly temperature trace runs between 28 °C and 44 °C.

We compute survival three different correct ways, plus the naive marginal product, and check that all three correct approaches give numerically identical answers.

```{r}
#| label: predict-vs-naive-setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)

# 4PL survival surface
ell  <- 0
uu   <- 1
kk   <- 4
z    <- 4
T_eq <- 36                                          # CTmax_1hr in this organism
m_T  <- function(T_val) log10(60) - (T_val - T_eq) / z   # midpoint as fn of T
S_4PL <- function(T_val, tau_min) {
  ell + (uu - ell) / (1 + exp(kk * (log10(pmax(tau_min, .Machine$double.eps)) - m_T(T_val))))
}

# Inverse: given survival S at temperature T, return tau (min) on the fresh-start curve at T.
# Note: the 4PL inversion uses natural log, not log10 — solving
#   S = ell + (u-ell)/(1 + exp(k*(log10(tau) - m(T))))
# for tau gives  log10(tau) = m(T) + (1/k) * ln((u-S)/(S-ell))
tau_at_T <- function(S, T_val) {
  S <- pmin(pmax(S, ell + 1e-12), uu - 1e-12)
  10 ^ (m_T(T_val) + (1 / kk) * log((uu - S) / (S - ell)))
}

# Temperature trajectory: 5 days, hourly, sinusoidal between 28 and 44 °C
hours    <- 0:(24 * 5 - 1)
T_traj   <- 36 + 8 * sin(2 * pi * (hours - 6) / 24)
T_c      <- 30                                     # damage threshold
delta_t  <- 60                                     # minutes per hourly interval

T_ref      <- T_eq                                 # reference for τ_eff
CT_max_1hr <- T_eq                                 # = T_ref by construction
t_ref_min  <- 60                                   # 1 hour reference duration in minutes
```

### Approach 1 — canonical HI integral (the manuscript's `@eq-hi`), then convert HI back to $S$

```{r}
#| label: approach-1-HI

# HI integral in canonical TDT form, time in hours
hi_per_hour <- ifelse(T_traj > T_c,
                      100 * 10 ^ ((T_traj - CT_max_1hr) / z),
                      0)
HI_canonical <- cumsum(hi_per_hour * 1)            # Δi = 1 hour

# Convert HI back to predicted survival via the 4PL surface at CTmax_1hr.
# When HI = 100, cumulative dose = 1 LT50 at CTmax_1hr → S = 0.5.
# Generally:  τ_eff at CTmax_1hr (min) = HI × t_ref_min / 100
tau_eff_from_HI <- HI_canonical * t_ref_min / 100
S_from_HI       <- S_4PL(CT_max_1hr, pmax(tau_eff_from_HI, .Machine$double.eps))
```

### Approach 2 — effective-time accumulator at $T_{\text{ref}}$, then predict once

```{r}
#| label: approach-2-tau-eff

inc        <- ifelse(T_traj > T_c, delta_t * 10^((T_traj - T_ref) / z), 0)
tau_eff    <- cumsum(inc)                          # cumulative effective minutes at T_ref
S_tau_eff  <- S_4PL(T_ref, pmax(tau_eff, .Machine$double.eps))
```

### Approach 3 — track current $S$, invert the 4PL at each step, advance, re-predict

```{r}
#| label: approach-3-invert-predict

S_invert  <- numeric(length(hours))
S_current <- uu                                    # fresh start
for (i in seq_along(hours)) {
  if (T_traj[i] > T_c) {
    tau_now   <- tau_at_T(S_current, T_traj[i])
    S_current <- S_4PL(T_traj[i], tau_now + delta_t)
  }
  S_invert[i] <- S_current
}
```

### Naive multiplication — the wrong approach, included for contrast

```{r}
#| label: naive-product

s_per_hour <- ifelse(T_traj > T_c, S_4PL(T_traj, delta_t), 1)
S_naive    <- cumprod(s_per_hour)
```

### Numerical comparison: all three correct approaches must agree

```{r}
#| label: agreement-check

sprintf("Max |S_from_HI − S_tau_eff|:   %.3e", max(abs(S_from_HI  - S_tau_eff)))
sprintf("Max |S_from_HI − S_invert|:    %.3e", max(abs(S_from_HI  - S_invert)))
sprintf("Max |S_tau_eff − S_invert|:    %.3e", max(abs(S_tau_eff  - S_invert)))
sprintf("Max |S_from_HI − S_naive|:     %.3e", max(abs(S_from_HI  - S_naive)))
```

```{r}
#| label: fig-all-approaches
#| fig-cap: "Predicted survival fraction over a 5-day hourly temperature trajectory under four approaches. The three correct approaches — HI-then-S, effective-time accumulator, and invert-and-re-predict — overlap to floating-point precision. The naive marginal product diverges because it pretends the organism resets to fresh state at every interval."
#| fig-height: 3.5

df <- tibble(
  hour                                                  = hours,
  `1: HI integral (eq-hi) → S`                          = S_from_HI,
  `2: effective-time accumulator → S`                   = S_tau_eff,
  `3: invert-and-re-predict each hour`                  = S_invert,
  `Naive: multiply per-interval marginals`              = S_naive
) |>
  pivot_longer(-hour, names_to = "Approach", values_to = "S")

ggplot(df, aes(hour, S,
               colour    = Approach,
               linetype  = Approach)) +
  geom_line(linewidth = 1.5, alpha = 0.65) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "longdash")) +
  labs(x = "Time (h)", y = "Predicted survival fraction") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.direction = "vertical")
```

### What the simulation shows

- **All three correct approaches agree to numerical precision.** Approaches 1 (HI integral converted to $S$) and 2 (effective-time accumulator) agree to floating-point round-off ($\approx 10^{-16}$). Approach 3 (invert-and-re-predict) agrees with the other two to $\approx 10^{-8}$ — slightly larger because each hourly inversion picks up a tiny round-trip error that accumulates over 120 hours, but still far below any biologically meaningful magnitude. The three are mathematically equivalent; the small spread is purely numerical, not conceptual.
- **The naive product diverges substantially** from the three correct approaches (max difference $\approx 0.088$ in absolute survival fraction). Each hour, it pretends the organism is fresh again, so the per-interval marginals overstate cumulative survival. The bias compounds monotonically across the trace.
- **Each correct approach uses the 4PL surface differently.**
  - Approach 1 uses only $z$ and $CT_{max_{1\text{hr}}}$ to accumulate HI, then plugs into the full 4PL surface at $CT_{max_{1\text{hr}}}$ once at the end to convert HI back to $S$.
  - Approach 2 uses $z$ to accumulate effective time at $T_{\text{ref}}$, then the full 4PL surface once at the end.
  - Approach 3 uses the full 4PL surface at every step (with $z$ baked in via $m(T)$ during the invert step).
  All three agree because they describe the same TDT line and the same survival surface — just labelled and computed differently.
- **For the manuscript**: this confirms numerically that the canonical `@eq-hi` form (HI tracking) is *fully consistent* with predicting from the 4PL model directly (Approaches 2 and 3). HI tracking and predict-from-4PL are not competing methods — they are alternative views of the same underlying inference, with HI as the threshold-crossing summary statistic and Approaches 2/3 yielding the full survival curve.

## Verification status

- **Mathematical claim** (naive product overestimates survival): direct chain-rule argument; verified by the worked numerical example. The discrepancy generalises: any partition into $\geq 2$ intervals with the marginal-product approach gives an upper bound on true survival, with the gap widening as the partition refines.
- **`@rezende_drosophila_mortality_2020`** Eq. 2 (survival-probability shift) and Eq. 3 (finite-difference update under variable $T$): **[verified]** against the paper text on 2026-05-07 in the parent literature note. The paper explicitly says "the survival rate at any given time $\tau'$ can be calculated by summing up the infinitesimal small changes in the time interval $[0, \tau']$" — i.e. running-clock accumulation, not marginal-product multiplication.
- **`@jorgensen_unifying_2021`** Eqs. 1, 3, 7 (damage integral, equivalent to @eq-tau-eff under reparameterisation): **[verified]** against the paper text. Their R-scripts at https://github.com/MOersted/Thermal-tolerances implement exactly the running-clock approach.
- **No code chunk in this note yet** — a short worked simulation comparing naive product to the running-clock formulation for a 40 °C constant trajectory would demonstrate the discrepancy concretely. Flagged for a future session.
