---
title: "Drosophila suzukii sublethal endpoints: knockdown and productivity TDTs"
date: 2026-06-13
format: html
execute:
  warning: false
  message: false
---

## Question

Case Study 4 (*D. suzukii*, Ørsted et al. 2024) already analyses **mortality** by
sex. The long-format `dsuzukii` dataset also carries two sublethal endpoints —
**time to heat coma** (`t_coma`) and **productivity** (`prod`, offspring / female
/ day). This note validates how we estimate thermal-death-time quantities ($z$,
$CT_{max_{1hr}}$) for both, by sex, in a joint model, and how those compare to the
two-stage `drc` values Ørsted report (their `TDT_parameters_R3.csv`):

| trait  | sex | z    | sCTmax1h |
|--------|-----|------|----------|
| coma   | M   | 2.33 | 36.31    |
| coma   | F   | 2.23 | 36.36    |
| fecund | M   | 3.46 | 36.66    |
| fecund | F   | 3.27 | 36.27    |

The analysis itself lives in `ms/supplement.qmd` (§ *Analysing sublethal
measures*, `#sec-dros-coma` and `#sec-dros-productivity`). This note records the
decisions and the validation, loading the cached fits rather than re-fitting.

```{r setup}
library(bayesTLS); library(dplyr); library(brms); library(posterior); library(here)
models_dir <- here::here("output", "models")
data(dsuzukii)
TBAR <- mean(dsuzukii$temp)
```

## Knockdown (time to coma)

**Key assumption — right-censoring.** A fly was scored awake/in-coma at the end
of its assigned exposure (`time` = a % of the temperature-specific median coma
time), and `t_coma` was recorded only for flies that comatosed *during* the
exposure. Flies still awake have `t_coma = NA`. So `t_coma` is **right-censored
at `time`** (≈ `r round(100*mean(is.na(dsuzukii$t_coma)))`% of flies). Dropping
the NA rows would select for early-coma individuals and bias the slope. We use
two censoring-aware routes:

1. **Censored linear** log-time regression (`log10(coma) | cens(...) ~ temp_c *
   sex + (1|cell)`, Gaussian) — keeps every fly via the censoring bound.
2. **4PL on proportion awake** — `in_coma = !is.na(t_coma)`; the proportion awake
   per (temp × level × sex) cell is a survival-type curve, fit with `sex` on the
   midpoint and read with `tls()`. This is exactly Ørsted's `awake ~ time`
   dose-response and is censoring-safe by construction.

The proportion-awake reconstruction reproduces Ørsted's published cell
proportions for 91/94 cells (the 3 misses are the denominator discrepancies
documented in `?dsuzukii`).

```{r coma-validate}
coma_tbar <- mean(dsuzukii$temp)
fit_coma_lin <- readRDS(file.path(models_dir, "fit_dsuzukii_coma_linear.rds"))
fit_coma_4pl <- readRDS(file.path(models_dir, "fit_dsuzukii_coma_4pl.rds"))

p  <- as_draws_df(fit_coma_lin)
sF <- p$b_temp_c; sM <- p$b_temp_c + p[["b_temp_c:sexM"]]
iF <- p$b_Intercept; iM <- p$b_Intercept + p$b_sexM
lin <- data.frame(
  sex = c("F","M"),
  z   = c(median(-1/sF), median(-1/sM)),
  ct1 = c(median(coma_tbar + (log10(60)-iF)/sF), median(coma_tbar + (log10(60)-iM)/sM)))

coma_panel <- dsuzukii |> group_by(temp, lvl, sex) |>
  summarise(n_total = n(), n_awake = sum(is.na(t_coma)),
            duration = first(time), .groups = "drop") |> filter(duration > 0)
std_coma <- standardize_data(coma_panel, temp = "temp", duration = "duration",
              n_total = "n_total", n_surv = "n_awake", duration_unit = "minutes")
set.seed(123)
tls4 <- tls(fit_coma_4pl, by = "sex", mode = "relative", t_ref = 60,
            time_multiplier = 1,
            temp_mean = mean(std_coma$temp - std_coma$temp_c))$summary

knitr::kable(lin, digits = 2, caption = "Censored linear: z, CTmax1h")
knitr::kable(subset(tls4, quantity %in% c("z","CTmax")), digits = 2,
             caption = "4PL (proportion awake): z, CTmax1h")
```

**Result.** Both routes give $z \approx 2.3$–2.4 and $CT_{max_{1hr}} \approx
36.3$–36.5 for both sexes, matching Ørsted's coma values (2.23 / 2.33) within the
credible intervals. Sexes are statistically indistinguishable.

## Productivity (reproductive output)

**Family.** `prod` is continuous with a large zero-mass (≈55%). brms has no
Tweedie, so we use a **hurdle** model and pick the positive part by LOO:

```{r prod-loo}
ga <- add_criterion(readRDS(file.path(models_dir,"fit_dsuzukii_prod_hurdle_gamma.rds")), "loo")
ln <- add_criterion(readRDS(file.path(models_dir,"fit_dsuzukii_prod_hurdle_lognormal.rds")), "loo")
loo_compare(ga, ln)
```

hurdle_**gamma** wins decisively (≈48 elpd, ~7 SE), so the original "log-normal"
preference does not hold for these data — we use gamma.

**Two components, read from the single model.** Critically (D.N., 2026-06-13),
we do **not** extract an $ED_{50}$ at each temperature and regress them — that is
the two-stage structure this paper avoids, and it breaks when a component has no
50% crossing. Instead we read each component's sensitivity **analytically from
the joint posterior**, the way the 4PL gives $z = -1/\beta_1$. Each linear
predictor is linear in log-duration and temperature, so the duration to a
threshold and its temperature slope are functions of the fitted coefficients.

```{r prod-tdt}
prod_dat <- dsuzukii |> mutate(logd = log10(time), temp_c = temp - mean(temp), sex = factor(sex))
p <- as_draws_df(ga); tbar <- mean(dsuzukii$temp)
temps <- sort(unique(dsuzukii$temp)); tc <- temps - tbar
# effective (Intercept, temp, logd, logd:temp) coefs for a dpar prefix and sex
eff <- function(pre, male) { pk <- function(s){ nm<-paste0(pre,s); if(nm%in%names(p)) p[[nm]] else 0 }
  list(I=pk("Intercept")+(if(male)pk("sexM")else 0), T=pk("temp_c")+(if(male)pk("temp_c:sexM")else 0),
       D=pk("logd")+(if(male)pk("logd:sexM")else 0), DT=pk("logd:temp_c")+(if(male)pk("logd:temp_c:sexM")else 0)) }
# INCIDENCE: intrinsic 50% threshold (hu=0.5); logd50(T) = -(I+T*tc)/(D+DT*tc)
inc_tdt <- function(male){ cc<-eff("b_hu_",male); ld<-function(t)-(cc$I+cc$T*t)/(cc$D+cc$DT*t); e<-1e-3
  z<-rowMeans(vapply(tc,function(t)-1/((ld(t+e)-ld(t-e))/(2*e)),numeric(nrow(p))))
  g<-seq(min(tc)-2,max(tc)+2,length.out=400); ldm<-vapply(g,ld,numeric(nrow(p)))
  ct<-apply(ldm,1,function(r){k<-which(diff(sign(r-log10(60)))!=0)[1]; if(is.na(k))NA else tbar+approx(r[k:(k+1)],g[k:(k+1)],log10(60))$y})
  list(z=z,ct=ct) }
# MAGNITUDE: unbounded log-linear clutch -> no intrinsic threshold; report the
# duration-sensitivity b(T)=dlog10(clutch)/dlog10(exposure)=(D+DT*tc)/log(10)
mag_b <- function(male){ cc<-eff("b_",male); vapply(tc,function(t)(cc$D+cc$DT*t)/log(10),numeric(nrow(p))) }
iF<-inc_tdt(FALSE); iM<-inc_tdt(TRUE); bF<-mag_b(FALSE); bM<-mag_b(TRUE)
data.frame(quantity=c("incidence z (F)","incidence z (M)","incidence CTmax1h (F)","incidence CTmax1h (M)"),
           value=c(median(iF$z),median(iM$z),median(iF$ct),median(iM$ct))) |> knitr::kable(digits=2)
data.frame(temp=temps, b_F=apply(bF,2,median), Pdecl_F=apply(bF,2,function(b)mean(b<0)),
           b_M=apply(bM,2,median), Pdecl_M=apply(bM,2,function(b)mean(b<0))) |> knitr::kable(digits=2)
```

**Result / biological finding.**

- **Incidence** (whether a fly reproduces): a clean single-model TDT,
  $z \approx 3.2$ (F) / 2.9 (M), $CT_{max_{1hr}} \approx 36.1$ / 36.5 °C, close to
  Ørsted's fecundity values (3.27 / 3.46; 36.27 / 36.66). Reproducing-or-not is
  bounded with an intrinsic 50% threshold, so its $z$ is well defined (the local
  $z$ rises gently across the range; we report the mean).
- **Magnitude** (offspring among reproducers): the conditional clutch is an
  **unbounded log-linear decline with no intrinsic threshold**, so a single $z$
  is not well defined — its implied value swings through a singularity where the
  duration-sensitivity crosses zero. The identifiable quantity is $b(T)$, which
  is **temperature-gated**: $\approx 0$ at 34 °C ($P(\text{decline})\approx 0.45$)
  and steepening to $\approx -0.6$ at 38 °C ($P(\text{decline})=1$).

So heat suppresses reproduction at **every** temperature by stopping flies
reproducing at all (incidence), but only **shrinks the clutch** of those that
still reproduce once it is hot ($\gtrsim 36$ °C). Decision (with D.N.,
2026-06-13): report incidence ($z$) + magnitude ($b(T)$) separately, no pooled
number (pooling mixes a bounded threshold-defined effect with an unbounded
threshold-less one).

**Caveat.** Per-individual `prod` is faithful to the raw counts but is *not*
Ørsted's undocumented normalised cell average (`?dsuzukii`), so exact matches to
their fecundity point estimates are not expected.

## Open questions

- The analytic single-model TDT reading for non-4PL brms response models (the
  `eff()`/`inc_tdt()` logic above) is currently inline in the supplement. It is a
  candidate for promotion to `R/` (a general "TLS from a joint brms model",
  complementing `tls()`/`extract_tdt()` which target the disjoint-bounds 4PL),
  with roxygen + tests — pending a decision on the package API.
