---
title: "Why does rate-multiplier $T_{crit}$ disagree between lethal and sublethal endpoints?"
date: 2026-05-13
author: "Patrice Pottier + Claude (Sonnet 4.6)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-cap-location: bottom
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| echo: false

# Snapshot produced by notes/.tcrit_compare_v2.R using the cached
# brown-shrimp fits in output/models/. The script saves a list with
# tdt_lines (data for the comparison plot) and posteriors (per-draw
# vectors of z, CTmax, T_crit for each pipeline). Re-run the .R script
# if any fit cache is refreshed.
library(ggplot2)
library(dplyr)
library(tidyr)
dat <- readRDS(here::here("notes", "2026-05-13-tcrit-compare-data.rds"))
posteriors <- dat$posteriors
tdt_lines  <- dat$tdt_lines
qd <- function(x, d = 2) round(quantile(x, c(0.5, 0.025, 0.975),
                                         na.rm = TRUE, names = FALSE), d)

# Small helper for inline reporting.
post_med  <- function(x, d = 2) round(median(x, na.rm = TRUE), d)
post_lo   <- function(x, d = 2) round(quantile(x, 0.025, na.rm = TRUE, names = FALSE), d)
post_hi   <- function(x, d = 2) round(quantile(x, 0.975, na.rm = TRUE, names = FALSE), d)
```

## The observation

On the brown shrimp dataset, the rate-multiplier $T_{crit}$ posteriors from
the lethal and sublethal pipelines disagree ŌĆö their 95 % credible
intervals do not overlap.

```{r tbl-numbers}
#| label: tbl-numbers
#| tbl-cap: "Posterior medians and 95% credible intervals for the three pipelines, computed from the cached `output/models/*.rds` fits. The lethal fit uses `time_multiplier = 60` (model time in hours, t_ref = 60 min). The sublethal 4PL uses `time_multiplier = 1` (model time already in minutes), matching the supplement's `extract_tdt()` call. The linear sublethal Tcrit is computed analytically from the linear-model posteriors using the same $r^*$ range."

tibble::tribble(
  ~Endpoint,                              ~`z (┬░C)`,                                 ~`CTmax_1hr (┬░C)`,                              ~`T_crit (┬░C)`,
  "Lethal 4PL (counts)",                  paste0(qd(posteriors$lethal$z), collapse = ", "),
                                          paste0(qd(posteriors$lethal$CTmax), collapse = ", "),
                                          paste0(qd(posteriors$lethal$Tcrit), collapse = ", "),
  "Sublethal 4PL (knocked-down)",         paste0(qd(posteriors$sublethal_4pl$z), collapse = ", "),
                                          paste0(qd(posteriors$sublethal_4pl$CTmax), collapse = ", "),
                                          paste0(qd(posteriors$sublethal_4pl$Tcrit), collapse = ", "),
  "Sublethal linear (log time)",          paste0(qd(posteriors$sublethal_linear$z), collapse = ", "),
                                          paste0(qd(posteriors$sublethal_linear$CTmax), collapse = ", "),
                                          paste0(qd(posteriors$sublethal_linear$Tcrit), collapse = ", ")
) |>
  knitr::kable(align = "lccc")
```

Two things stand out.

- **$CT_{max_{1hr}}$ for the three pipelines is essentially the same**
  (lethal `r post_med(posteriors$lethal$CTmax)` ┬░C, sublethal 4PL
  `r post_med(posteriors$sublethal_4pl$CTmax)` ┬░C, sublethal linear
  `r post_med(posteriors$sublethal_linear$CTmax)` ┬░C). The 1-hour-50%
  temperature for "knocked down" and for "dead" sits at roughly the
  same place on the temperature axis on this panel.
- **$z$ is more than halved going from lethal to sublethal**
  (`r post_med(posteriors$lethal$z)` ┬░C lethal ŌåÆ `r post_med(posteriors$sublethal_4pl$z)` ┬░C
  sublethal). The two sublethal pipelines agree closely on $z$ and
  $CT_{max_{1hr}}$, so this is not an artefact of the 4PL aggregation;
  the sublethal log-time data really do have a steeper slope on
  $\log_{10}(\text{time})$ vs. $T$ than the lethal data do.

Plug those into $T_{crit} \approx CT_{max_{1hr}} - 2.5\,z$ (the
posterior median of $T_{crit} = CT_{max_{1hr}} + z \cdot \log_{10}(r^*/100)$
with $r^*$ uniform on $\log_{10}$ across $[0.1, 1]$ % HI per hour)
[@arnold_thermal_2025; @jorgensen_unifying_2021; **havro_faber_2023**]
and you get the observed gap:

- Lethal:     `r post_med(posteriors$lethal$CTmax)` ŌłÆ 2.5 ├Ś `r post_med(posteriors$lethal$z)` Ōēł `r round(post_med(posteriors$lethal$CTmax) - 2.5 * post_med(posteriors$lethal$z), 2)` ┬░C
- Sublethal:  `r post_med(posteriors$sublethal_4pl$CTmax)` ŌłÆ 2.5 ├Ś `r post_med(posteriors$sublethal_4pl$z)` Ōēł `r round(post_med(posteriors$sublethal_4pl$CTmax) - 2.5 * post_med(posteriors$sublethal_4pl$z), 2)` ┬░C

So the gap is entirely driven by the difference in $z$, magnified
2.5├Ś by the rate-multiplier formula.

## The TDT lines tell the same story

The cleanest way to see what is happening is to plot the predicted
1-hour-50% temperature across a temperature grid for each pipeline ŌĆö
i.e., the TDT line $\log_{10}(\text{time to endpoint}) = f(T)$. If both
endpoints reflected the same underlying damage process with different
manifestation thresholds, the two lines would be **parallel**, offset
by a constant on the $\log_{10}$ axis. They are not.

```{r fig-tdt-lines}
#| label: fig-tdt-lines
#| fig-cap: "Posterior $\\log_{10}$ (time to 50% endpoint) across temperature for the three shrimp pipelines. The horizontal dashed line is $\\log_{10}(60)$ minutes = 1 hour; the vertical reference is the posterior median $CT_{max_{1hr}}$ from the lethal fit. Where each TDT line crosses the dashed line is its $CT_{max_{1hr}}$. The two sublethal pipelines agree across the panel temperature range, and both cross the lethal line at $\\approx CT_{max_{1hr}}$. At lower temperatures the steeper sublethal line predicts *longer* times to knockdown than the lethal line predicts to death ŌĆö biologically impossible ŌĆö so the linearity assumption holds only over a narrow window near $CT_{max_{1hr}}$ and the two lines should not be extrapolated past the data."
#| fig-width: 7
#| fig-height: 5

ct_let_med <- post_med(posteriors$lethal$CTmax)

ggplot(tdt_lines,
        aes(x = temp, y = log10_time_med, colour = method, fill = method)) +
  geom_ribbon(aes(ymin = log10_time_lo, ymax = log10_time_hi),
              alpha = 0.18, colour = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = log10(60), linetype = "dashed",
             colour = "grey40") +
  geom_vline(xintercept = ct_let_med, linetype = "dotted",
             colour = "grey40") +
  annotate("text", x = min(tdt_lines$temp), y = log10(60),
           label = "log10(60 min) = 1 hour",
           hjust = 0, vjust = -0.3, size = 3, colour = "grey40") +
  annotate("text", x = ct_let_med, y = max(tdt_lines$log10_time_hi),
           label = "CTmax_1hr (lethal)",
           hjust = -0.05, vjust = 1, size = 3, colour = "grey40") +
  scale_colour_manual(values = c("Lethal 4PL (counts)" = "#b2182b",
                                  "Sublethal 4PL (knocked-down)" = "#2166ac",
                                  "Sublethal linear (log time-to-knockdown)" = "#1b7837")) +
  scale_fill_manual(values   = c("Lethal 4PL (counts)" = "#b2182b",
                                  "Sublethal 4PL (knocked-down)" = "#2166ac",
                                  "Sublethal linear (log time-to-knockdown)" = "#1b7837")) +
  labs(x = "Temperature (┬░C)",
       y = expression(log[10]("time to 50% endpoint, min")),
       colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "top",
        legend.direction = "vertical",
        legend.text = element_text(size = 9))
```

Three things to read off @fig-tdt-lines:

1. **The two sublethal pipelines agree closely.** Both the joint 4PL
   on proportion-knocked-down and the classical linear regression on
   $\log_{10}$ times give essentially the same TDT line. The
   discrepancy with the lethal line is therefore a real feature of
   the data, not a 4PL-vs-linear artefact.
2. **The lethal and sublethal lines cross at $\approx CT_{max_{1hr}}$.**
   The posterior median lines intersect near the 1-hour duration
   right at the lethal $CT_{max_{1hr}}$ of `r round(ct_let_med, 1)` ┬░C.
   At higher temperatures the lethal line predicts longer times than
   the sublethal ŌĆö which is biologically sensible (knockdown precedes
   death). At lower temperatures the sublethal line predicts *longer*
   times to knockdown than the lethal line predicts to death ŌĆö which
   is **biologically impossible**: an animal cannot die before it
   loses response to touch.
3. **Linearity breaks down outside the assay window.** The
   sublethal panel runs 31ŌĆō33 ┬░C; below 31 ┬░C the line is pure
   extrapolation, and the extrapolation crosses through the lethal
   line and into the impossible region almost immediately. This is
   the classic warning against extrapolating a TDT line outside its
   assay range [@jorgensen_unifying_2021; @arnold_thermal_2025].

This is the underlying reason for the $T_{crit}$ disagreement.
$T_{crit}$ in the rate-multiplier definition is computed by
extrapolating each TDT line down to a low rate ŌĆö but the sublethal
TDT line is wrong in exactly the temperature range that
extrapolation lands in. The lethal $T_{crit}$ at ~25 ┬░C is itself an
extrapolation, but at least one that stays in the region where
ramping experiments (acclimation/recovery regimes) and the literature
[**havro_faber_2023**] suggest the breakpoint sits. The sublethal
$T_{crit}$ at ~28.7 ┬░C sits on a TDT line that has *already crossed
into the impossible region*.

## What does this say about $T_{crit}$?

Two related but distinct quantities are being conflated when we call
both numbers "$T_{crit}$":

(a) **$T_{crit}$ as an extrapolated TDT-line quantity** ŌĆö what
    `extract_tdt()` returns. It is the temperature at which the
    endpoint's TDT line crosses a chosen rate floor $r^*$. The
    sublethal TDT line is steeper than the lethal one on this panel,
    so its $T_{crit}$ sits higher (closer to $CT_{max_{1hr}}$). This
    is mathematically correct given the input $(\hat z, \hat{CT}_{max})$
    but the line being extrapolated is *demonstrably wrong* below the
    assay range (it predicts knockdown taking longer than death).

(b) **$T_{crit}$ as a biological breakpoint** ŌĆö Faber's, J├Ėrgensen's
    definition. The temperature at which the absolute damage-rate
    function $r(T)$ stops following the $Q_{10} \sim 2$ permissive
    regime and starts the $Q_{10} \sim 100\text{--}25{,}000$ stressful
    regime [**havro_faber_2023**; @jorgensen_unifying_2021]. This is
    endpoint-agnostic: $r(T)$ is the same physiological process whether
    measured by knockdown or by death. It is detected empirically as a
    piecewise breakpoint in $\log(\text{rate})$ vs. $T$.

The current `extract_tdt()` computes (a), not (b), and labels it
$T_{crit}$. **That is the source of the confusion**: (a) is *not* a
good proxy for (b) unless the TDT line is approximately linear over the
range $[T_{crit}^{(b)}, CT_{max_{1hr}}]$. For the lethal shrimp panel
the line is well-behaved over that range and (a) lands in the right
neighbourhood. For the sublethal shrimp panel the line is steeper at
high $T$ and shallower at low $T$ (the data only see the high-$T$
limb), so extrapolating linearly produces a $T_{crit}$ that is biased
upward ŌĆö the line steepens just before $CT_{max_{1hr}}$ and the
linear extrapolation continues that steep slope downward, missing the
true (shallower) low-$T$ regime entirely.

## What is the "right" way to estimate $T_{crit}$?

Three options, in increasing order of how much they fix:

### 1. Stop pretending it is endpoint-agnostic

Keep the rate-multiplier $T_{crit}$ as the *operational* summary
returned by `extract_tdt()`, but call it $T_{crit}^{(\text{endpoint},
r^*)}$ in the manuscript and never cross-compare lethal and sublethal
values as if they were the same physiological threshold. This is the
status quo plus honest labelling. It is the easiest change but it
does not give us a single biologically defensible $T_{crit}$ for the
species. We may want to make this clearer in the manuscript.

### 2. Compute Faber's empirical breakpoint per posterior draw

For each posterior draw, take the fitted $\log_{10}\text{LT50}(T)$
curve over a *wide* temperature grid that spans both the permissive
and stressful regimes. Fit a piecewise model (two segments meeting at
a breakpoint, e.g. via `segmented::segmented()` or a brms
piecewise-linear formula). The breakpoint location is the per-draw
estimate of biological $T_c$. Pool across draws to get a posterior of
$T_c$ that does **not** depend on the endpoint ŌĆö the breakpoint of
$\log(r(T))$ vs $T$ is the same physiological feature whether $r$ is
"die-rate" or "knockdown-rate."

There is one fundamental data requirement: the model has to
have some support in the permissive regime ($T \lesssim T_c$,
$Q_{10} \sim 2$). The shrimp panels do not: the lethal panel is
30ŌĆō33 ┬░C, all clearly above the expected $T_c \approx 25$ ┬░C. So
on this dataset the breakpoint would be entirely prior-driven for
the lethal fit, and pure extrapolation for the sublethal fit. For
future studies that aim to estimate $T_c$ specifically (rather than
$CT_{max}$ at a reference duration), the design implication is
clear: include at least one assay temperature below the expected
$T_c$ and run it for long enough to see whether any death/knockdown
accumulates.

### 3. Reframe $T_{crit}$ in absolute-damage units

If we had a calibration between the sublethal endpoint dose and the
lethal endpoint dose (e.g., "50% knockdown corresponds to $\alpha$
fraction of the LT50 dose"), we could convert $r^*_{\text{sub}} =
\alpha \cdot r^*_{\text{leth}}$ and bring both $T_{crit}$ values onto
a common absolute-rate axis. This is biologically motivated but
species- and endpoint-specific and beyond what we can do with the
shrimp data alone. Worth a sentence in the discussion as a future
direction.

## Recommendation for the current manuscript

The supplement now contains an explicit caveat
(*A caution on cross-endpoint comparison of $T_{crit}$* in the shrimp
sublethal section) that the rate-multiplier $T_{crit}$
is endpoint-conditional and should not be cross-compared between
lethal and sublethal fits without that adjustment. We should keep
that caveat. In the discussion of the manuscript we should:

- Acknowledge that the rate-multiplier $T_{crit}$ as implemented is
  the extrapolated-TDT-line variant (a), not the empirical breakpoint
  (b), and that the two only coincide when the TDT line is roughly
  linear over $[T_c, CT_{max_{1hr}}]$.
- Recommend that future TDT designs aimed at $T_c$ include at least
  one permissive-range assay temperature.
- Flag option 2 (piecewise breakpoint per draw) as the natural extension
  for datasets that do have permissive-range data.

## Open follow-ups

- Implement `derive_breakpoint_tcrit(workflow, temp_grid)` that uses
  `segmented::segmented()` per posterior draw on the fitted
  $\log_{10}\text{LT50}(T)$ curve. Returns the breakpoint posterior;
  `NA` per draw when `segmented` fails. Endpoint-agnostic by
  construction. **Not** usable on the current shrimp panels (no
  permissive-range data); useful for any dataset whose assay range
  brackets the suspected $T_c$.
- Add a $T_{crit}(r^*)$ sensitivity figure to the supplement ŌĆö one
  panel each for lethal and sublethal, with $T_{crit}$ on the y-axis
  and $r^*$ on a log-scale x-axis from 10Ōü╗Ōü┤ to 10Ōü╗┬╣ % per hour. Should
  show that the lethal and sublethal curves are simply offset by the
  endpoint dose ratio in log-space, and that no choice of $r^*$ in
  $[0.1, 1]$ brings the two into agreement.

See also [[reference_faber_2023_tcrit]] and the earlier note
`2026-05-12-tcrit-reconsidered-faber-et-al.qmd`.
