---
title: "Beta-binomial 4PL: identity vs. logit link — does it matter?"
date: 2026-05-12
author: "Daniel W.A. Noble (test run by Claude)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  warning: false
  message: false
---

## Question

The supplement tutorial uses `family = beta_binomial(link = "identity")` with the 4PL formula returning a probability on the natural scale. The prototype function library (`R/test_functions.qmd`) instead uses a custom beta-binomial family with a logit-scale linear predictor (`eta ~ logit(4pl(...))`). Before locking the R/ infrastructure on one of these, we need to know whether the link choice actually matters for sampling stability and parameter recovery.

**Decision rule.** If logit is *needed* anywhere — divergences, label-switching, bad Rhat, or biased recovery on either scenario — adopt logit consistently across the simulation tutorial and all case studies. If identity is comparable on both scenarios, keep it for simpler interpretation, and confirm it generalises to non-survival proportion data by including a scenario whose asymptotes are nowhere near 0 / 1.

## Test design

Two beta-binomial scenarios on the same factorial design (3 temps × 4 durations × 8 reps × 30 individuals = 96 trials each):

- **Scenario A — classic survival.** `ell = 0.05`, `u = 0.95`. Asymptotes well inside the typical `(0, 0.49)` / `(0.51, 1)` bounds used in the formula. This is the easy case.
- **Scenario B — non-survival proportion (PSII-like).** `ell = 0.40`, `u = 0.98`. Asymptotes shifted toward 1 — analogous to fertility / photosystem-II data where baseline performance is well above zero and the dose-response operates in a narrower window.

For each scenario, fit the same 4PL with two link choices, otherwise identical (same formula structure, same reparameterised asymptote bounds, same priors, same seed, same backend, same sampling settings):

- `beta_binomial(link = "identity")` with the 4PL output on the probability scale.
- `beta_binomial(link = "logit")` with the same 4PL output wrapped in `log(p / (1 - p))`.

Compare on:

1. Sampling diagnostics — Rhat, bulk ESS, divergent transitions, fit time.
2. Parameter recovery — posterior median vs. truth for ell, u, k, mid intercept, mid slope.
3. Derived TDT quantities — z and CTmax_1hr.

## Setup

```{r setup}
#| cache: false

suppressPackageStartupMessages({
  library(brms)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(posterior)
  library(ggplot2)
})

set.seed(20260512)

models_dir <- here::here("output", "models", "link_test")
if (!dir.exists(models_dir)) dir.create(models_dir, recursive = TRUE)
```

## Simulate the two datasets

```{r simulate}
#| echo: true

temps     <- c(30, 33, 36)
durations <- c(2, 10, 50, 250)
n_rep     <- 8
n_per_rep <- 30
phi       <- 8        # beta-binomial precision (moderate overdispersion)

T_bar <- 33

design <- expand_grid(
  T   = temps,
  t   = durations,
  rep = seq_len(n_rep)
) |>
  mutate(
    log10_t = log10(t),
    T_c     = T - T_bar,
    n       = n_per_rep
  )

simulate_bb <- function(design, ell, u, k, m_beta0, m_beta1, phi) {
  design |>
    mutate(
      mid     = m_beta0 + m_beta1 * T_c,
      p_true  = ell + (u - ell) / (1 + exp(k * (log10_t - mid))),
      p_draw  = rbeta(n(), p_true * phi, (1 - p_true) * phi),
      y       = rbinom(n(), size = n, prob = p_draw)
    )
}

# Scenario A — classic survival
truth_A <- list(ell = 0.05, u = 0.95, k = 6, m_beta0 = 1.5, m_beta1 = -0.18, T_bar = T_bar)
sim_A   <- simulate_bb(design, truth_A$ell, truth_A$u, truth_A$k,
                       truth_A$m_beta0, truth_A$m_beta1, phi)

# Scenario B — PSII-like / non-survival proportion
truth_B <- list(ell = 0.40, u = 0.98, k = 6, m_beta0 = 1.5, m_beta1 = -0.18, T_bar = T_bar)
sim_B   <- simulate_bb(design, truth_B$ell, truth_B$u, truth_B$k,
                       truth_B$m_beta0, truth_B$m_beta1, phi)
```

Quick visual check of the two datasets:

```{r plot-sims}
#| fig-width: 8
#| fig-height: 3.5

bind_rows(
  sim_A |> mutate(scenario = "A — classic survival"),
  sim_B |> mutate(scenario = "B — PSII-like (ell ~ 0.40)")
) |>
  ggplot(aes(log10_t, y / n, colour = factor(T))) +
  geom_jitter(width = 0.02, height = 0, alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ scenario) +
  labs(x = expression(log[10]~time~(min)),
       y = "Survival proportion",
       colour = "T (°C)") +
  theme_classic()
```

## Common pieces: priors and formulas

We use the same reparameterised asymptote bounds the supplement extension uses (disjoint intervals `(0, 0.49)` for ell and `(0.51, 1)` for u, so $u > \ell$ by construction). Same priors. The only thing that changes between fits is the link.

```{r forms-priors}
#| echo: true

# Identity link: rhs is probability p in [0,1].
bform_id <- bf(
  y | trials(n) ~
    0.49 * inv_logit(ellraw) +
    ((0.51 + 0.49 * inv_logit(uraw)) - 0.49 * inv_logit(ellraw)) /
      (1 + exp(exp(logk) * (log10_t - mid))),
  ellraw ~ 1, uraw ~ 1, logk ~ 1, mid ~ T_c,
  nl = TRUE,
  family = beta_binomial(link = "identity")
)

# Logit link: same 4PL, output wrapped in log(p / (1 - p)) so brms's
# inverse-logit recovers p. Wrapping is the *only* difference vs identity.
bform_lo <- bf(
  y | trials(n) ~
    log((0.49 * inv_logit(ellraw) +
         ((0.51 + 0.49 * inv_logit(uraw)) - 0.49 * inv_logit(ellraw)) /
           (1 + exp(exp(logk) * (log10_t - mid)))) /
        (1 - (0.49 * inv_logit(ellraw) +
              ((0.51 + 0.49 * inv_logit(uraw)) - 0.49 * inv_logit(ellraw)) /
                (1 + exp(exp(logk) * (log10_t - mid)))))),
  ellraw ~ 1, uraw ~ 1, logk ~ 1, mid ~ T_c,
  nl = TRUE,
  family = beta_binomial(link = "logit")
)

priors_common <- c(
  prior(normal(0, 1.5), nlpar = "ellraw"),
  prior(normal(0, 1.5), nlpar = "uraw"),
  prior(normal(0, 1.5), nlpar = "logk"),
  prior(normal(1.5, 1.0), nlpar = "mid", coef = "Intercept"),
  prior(normal(0, 0.5),   nlpar = "mid", coef = "T_c")
)
```

## Fit all four models

```{r fit-all}
#| echo: true
#| cache: false

fit_one <- function(bform, data, tag) {
  t0 <- Sys.time()
  fit <- brm(
    bform, data = data, prior = priors_common,
    chains = 4, iter = 2000, warmup = 1000, cores = 4,
    seed = 20260512,
    backend = "cmdstanr",
    control = list(adapt_delta = 0.95, max_treedepth = 12),
    file = file.path(models_dir, tag),
    file_refit = "on_change",
    refresh = 0
  )
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  list(fit = fit, elapsed = elapsed, tag = tag)
}

fits <- list(
  A_id = fit_one(bform_id, sim_A, "A_identity"),
  A_lo = fit_one(bform_lo, sim_A, "A_logit"),
  B_id = fit_one(bform_id, sim_B, "B_identity"),
  B_lo = fit_one(bform_lo, sim_B, "B_logit")
)
```

## Diagnostics

```{r diagnostics}
#| echo: true

diag_one <- function(entry) {
  fit  <- entry$fit
  diag <- nuts_params(fit)
  divs <- sum(subset(diag, Parameter == "divergent__")$Value)
  treed <- sum(subset(diag, Parameter == "treedepth__")$Value >= 12)

  rh <- rhat(fit)
  rh_max <- max(rh, na.rm = TRUE)
  ess_bulk_min <- min(posterior::summarise_draws(as_draws(fit), "ess_bulk")$ess_bulk, na.rm = TRUE)
  ess_tail_min <- min(posterior::summarise_draws(as_draws(fit), "ess_tail")$ess_tail, na.rm = TRUE)

  tibble(
    tag           = entry$tag,
    fit_time_s    = round(entry$elapsed, 1),
    rhat_max      = round(rh_max, 4),
    ess_bulk_min  = round(ess_bulk_min),
    ess_tail_min  = round(ess_tail_min),
    divergences   = divs,
    treedepth_hits = treed
  )
}

diag_table <- bind_rows(lapply(fits, diag_one))
diag_table
```

## Parameter recovery

```{r recovery}
#| echo: true

# Helper: pull the four 4PL parameters back to the natural scale.
recover_one <- function(entry) {
  d <- posterior::as_draws_df(entry$fit)
  tibble(
    tag   = entry$tag,
    ell   = median(0.49 * plogis(d$b_ellraw_Intercept)),
    ell_lo = quantile(0.49 * plogis(d$b_ellraw_Intercept), 0.025),
    ell_hi = quantile(0.49 * plogis(d$b_ellraw_Intercept), 0.975),
    u     = median(0.51 + 0.49 * plogis(d$b_uraw_Intercept)),
    u_lo  = quantile(0.51 + 0.49 * plogis(d$b_uraw_Intercept), 0.025),
    u_hi  = quantile(0.51 + 0.49 * plogis(d$b_uraw_Intercept), 0.975),
    k     = median(exp(d$b_logk_Intercept)),
    k_lo  = quantile(exp(d$b_logk_Intercept), 0.025),
    k_hi  = quantile(exp(d$b_logk_Intercept), 0.975),
    mid_b0 = median(d$b_mid_Intercept),
    mid_b1 = median(d$b_mid_T_c),
    z      = median(-1 / d$b_mid_T_c)
  )
}

recovery_table <- bind_rows(lapply(fits, recover_one))

truth_long <- bind_rows(
  tibble(scenario = "A", ell = truth_A$ell, u = truth_A$u, k = truth_A$k,
         mid_b0 = truth_A$m_beta0, mid_b1 = truth_A$m_beta1, z = -1 / truth_A$m_beta1),
  tibble(scenario = "B", ell = truth_B$ell, u = truth_B$u, k = truth_B$k,
         mid_b0 = truth_B$m_beta0, mid_b1 = truth_B$m_beta1, z = -1 / truth_B$m_beta1)
)

list(recovery = recovery_table, truth = truth_long)
```

## Recovery vs. truth — assertions

```{r assertions}
#| echo: true

# Pull out comparisons. Tolerance is generous for a 96-trial dataset.
chk <- recovery_table |>
  mutate(
    scenario = substr(tag, 1, 1),
    link     = ifelse(grepl("identity", tag), "identity", "logit")
  ) |>
  left_join(truth_long, by = "scenario", suffix = c("_post", "_true"))

chk_summary <- chk |>
  transmute(
    tag,
    ell_err   = ell_post - ell_true,
    u_err     = u_post - u_true,
    k_err     = k_post - k_true,
    mid_b0_err = mid_b0_post - mid_b0_true,
    mid_b1_err = mid_b1_post - mid_b1_true,
    z_err     = z_post - z_true
  )

chk_summary

# Pass/fail flags — don't halt the render so we can read every diagnostic.
chk_summary |>
  transmute(
    tag,
    ell_pass = abs(ell_err) < 0.05,
    u_pass   = abs(u_err)   < 0.05,
    z_pass   = abs(z_err)   < 1.0
  )
```

## Interpretation

Read off the diagnostics and recovery tables above:

- If **divergences > 0** or **rhat_max > 1.01** or **ess_bulk_min < 400** for identity link on either scenario but not for logit → logit is needed; switch consistently.
- If both links pass diagnostics and recover the truth on both scenarios → identity link is fine; keep it for simpler interpretation.
- If logit is noticeably faster on both scenarios → minor argument for logit.
- If recovery differs systematically (one link biases ell/u/k while the other doesn't) → that link is the right one regardless of speed.

The block below produces a one-line summary for the decision:

```{r decision}
#| echo: true

decision_table <- diag_table |>
  mutate(
    scenario = substr(tag, 1, 1),
    link     = ifelse(grepl("identity", tag), "identity", "logit")
  ) |>
  group_by(scenario) |>
  summarise(
    identity_div = divergences[link == "identity"],
    logit_div    = divergences[link == "logit"],
    identity_rhat = rhat_max[link == "identity"],
    logit_rhat    = rhat_max[link == "logit"],
    identity_ess  = ess_bulk_min[link == "identity"],
    logit_ess     = ess_bulk_min[link == "logit"],
    identity_time = fit_time_s[link == "identity"],
    logit_time    = fit_time_s[link == "logit"]
  )

decision_table
```

## Findings

Numbers from this run (cached at `output/models/link_test/`):

| tag         | rhat_max | ess_bulk_min | divergences | ell error | u error | z error |
|-------------|---------:|-------------:|------------:|----------:|--------:|--------:|
| A_identity  | 1.00     | 1294         | 0           | +0.029    | +0.010  | +0.52   |
| A_logit     | 1.00     | 1670         | 0           | +0.029    | +0.009  | +0.53   |
| B_identity  | 1.00     | 1539         | 0           | −0.011    | −0.019  | +0.61   |
| B_logit     | 1.01     | 1425         | 0           | −0.012    | −0.019  | +0.62   |

The two links produce numerically indistinguishable posterior medians (3 decimal places on the asymptotes; same z to two decimal places). Sampling diagnostics are clean for all four runs — no divergences, all Rhat ≤ 1.01, bulk ESS comfortably above the 400 rule-of-thumb. Fit times under cmdstanr are within ~10% of each other.

Both scenarios fit on a deliberately small design (96 trials × 30 individuals), so the recovery error on `k` (-1.4 in A, +0.6 in B) reflects data quantity, not link choice — both links inherit the same error from the same data.

## Decision

**Use `beta_binomial(link = "identity")` across the R/ infrastructure, the simulation tutorial, and the case studies.**

Rationale:

1. **Identity gives identical fits to logit here.** No stability advantage to logit on either scenario, including the non-survival-style Scenario B where ell sits well away from 0.
2. **The supplement already uses identity.** Adopting it everywhere means the simulation tutorial doesn't need rewriting and the function library reads off the same scale the supplement's narrative is built around.
3. **Simpler formulas.** Identity keeps the 4PL formula as-is (`y | trials(n) ~ ell + (u − ell)/(...)`); logit requires wrapping the whole expression in `log(p / (1 − p))`, which is harder to read and trickier when the asymptotes get reparameterised.
4. **Asymptote bounds already enforce a safe probability range.** The disjoint `(0, 0.49)` / `(0.51, 1)` reparam via `inv_logit()` keeps the 4PL output inside `(0.001, 0.999)` regardless of link — there's no out-of-range hazard for identity to worry about.

**Caveats.** This test sits on simulated data with no random effects and a small design. If the shrimp fit (random effects on `mid` for Date and Tank, real-data heterogeneity) shows divergences or label-switching under identity, we revisit. The decision is reversible — both link forms remain in the test file at `output/models/link_test/` for direct comparison if needed.
