---
title: "Test: does brms::beta_binomial() + trials() work in a nonlinear model?"
date: 2026-05-11
format:
  html:
    embed-resources: true
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
  eval: false
---

## Background

`test_functions.qmd` uses a custom `tdt_beta_binomial` family with `vint(n_total)` to work
around a brms 2.22.0 code-generation bug: in nonlinear models, `trials()` produced
unindexed Stan code (`vn_total` instead of `vn_total[n]`), causing impossible likelihood
evaluations for rows with varying denominators.

This note tests whether the current brms version has fixed the bug.  If the check below
passes we can drop the custom family.

## Step 1 — Update brms

Run in your R console **before rendering this note**:

```{r update-brms}
#| eval: false
install.packages("brms")
packageVersion("brms")  # should be >= 2.22.0; note the exact version here
```

## Step 2 — Synthetic data with intentionally varying trial counts

The critical stress test is **varying `n_total` per row** — that is precisely the
condition that triggers the indexing bug.

```{r setup}
library(brms)
library(dplyr)

set.seed(42)

temps     <- c(28, 30, 32, 34, 36)
durations <- c(0.25, 0.5, 1, 2, 4)   # hours
temp_mean <- mean(temps)

# Deliberately vary n_total (5–12) so the bug manifests if present
sim_dat <- expand.grid(assay_temp = temps, duration = durations) |>
  mutate(
    n_total  = sample(5:12, n(), replace = TRUE),
    temp_c   = assay_temp - temp_mean,
    logd     = log10(duration),
    # True 4PL: low=0.02, up=0.97, k=3, mid = -0.3 - 0.15*temp_c
    mid_true = -0.3 - 0.15 * temp_c,
    p_true   = 0.02 + (0.97 - 0.02) /
                 (1 + exp(3 * (logd - mid_true))),
    n_surv   = rbinom(n(), size = n_total, prob = p_true)
  )

glimpse(sim_dat)
```

## Step 3 — Inspect the generated Stan code WITHOUT fitting

`sample = FALSE` tells brms to build the Stan model but not sample.  We search the
likelihood block for `vn_total` to check whether it appears as a scalar (bug) or
indexed (fixed).

```{r inspect-stan-code}
formula_test <- bf(
  n_surv | trials(n_total) ~ eta,
  nlf(eta ~ logit(
    0.001 + inv_logit(lraw) * 0.498 +
    (0.501 + inv_logit(uraw) * 0.498 - (0.001 + inv_logit(lraw) * 0.498)) /
    (1 + exp(exp(logk) * (logd - mid)))
  )),
  lf(lraw ~ 1),
  lf(uraw ~ 1),
  lf(logk ~ 1),
  lf(mid ~ temp_c),
  nl = TRUE
)

priors_test <- c(
  set_prior("normal(-3, 1)",  class = "b", nlpar = "lraw"),
  set_prior("normal(3, 1)",   class = "b", nlpar = "uraw"),
  set_prior("normal(0.7, 1)", class = "b", nlpar = "logk"),
  set_prior("normal(0, 1.5)", class = "b", nlpar = "mid", coef = "Intercept"),
  set_prior("normal(0, 0.6)", class = "b", nlpar = "mid"),
  set_prior("gamma(2, 0.1)",  class = "phi")
)

sc <- make_stancode(
  formula  = formula_test,
  data     = sim_dat,
  family   = beta_binomial(),
  prior    = priors_test
)

# --- Key check ---
# Search for how n_total appears in the likelihood line.
# GOOD: "vn_total[n]" — row-indexed scalar (bug is fixed)
# BAD:  "vn_total,"   — whole vector passed (bug still present)
lines_with_n_total <- grep("vn_total", strsplit(sc, "\n")[[1]], value = TRUE)
cat("Lines containing 'vn_total' in the Stan likelihood:\n")
cat(lines_with_n_total, sep = "\n")
```

**Interpret the output:**

- If you see `vn_total[n]` → the bug is fixed; safe to replace the custom family.
- If you see `vn_total` without `[n]` → bug still present; keep the custom family.

## Step 4 — Fit the model (only run if Step 3 passes)

```{r fit-test}
#| eval: false
fit_test <- brm(
  formula   = formula_test,
  data      = sim_dat,
  family    = beta_binomial(),
  prior     = priors_test,
  chains    = 2,
  cores     = 2,
  iter      = 1000,
  warmup    = 500,
  seed      = 42,
  init      = 0,
  backend   = "cmdstanr"
)

summary(fit_test)

# Sanity check: posterior medians should be in the right ballpark
# mid_Intercept ~ -0.3, mid_temp_c ~ -0.15, exp(logk) ~ 3
```

## Step 5 — Repeat the Stan-code check with the real shrimp data

The synthetic test (Steps 2–4) has no random effects.  This section applies the same
model structure — following the parameterisation in ms.qmd (eq-reparam and eq-linpred)
— to the shrimp lethal TDT data, adding random intercepts on `mid` for `Date` and
`Tank`.  No helper functions are used.

```{r shrimp-setup}
library(brms)
library(dplyr)

shrimp_tdt <- readRDS("Rdata/shrimp_lethal_prepared_tdt.RDS")

# Columns required: n_surv, n_total, logd, temp_c, Date, Tank
# temp_c = assay_temp - grand mean temperature (already in the prepared data)
temp_mean <- attr(shrimp_tdt, "tdt_meta")$temp_mean
mid_start <- median(shrimp_tdt$logd, na.rm = TRUE)

cat("temp_mean:", temp_mean, "\n")
cat("mid_start (median logd):", mid_start, "\n")
cat("n_total range:", range(shrimp_tdt$n_total), "\n")
```

```{r shrimp-stan-check}
# Formula follows ms.qmd eq-reparam exactly:
#   low = 0.49 * inv_logit(lraw)          -> low in [0, 0.49]
#   up  = 0.51 + 0.49 * inv_logit(uraw)   -> up  in [0.51, 1]
#   k   = exp(logk)
#   mid = b0 + b1 * temp_c + (1|Date) + (1|Tank)   (eq-linpred)
formula_shrimp <- bf(
  n_surv | trials(n_total) ~ eta,
  nlf(eta ~ logit(
    0.49 * inv_logit(lraw) +
    (0.51 + 0.49 * inv_logit(uraw) - 0.49 * inv_logit(lraw)) /
    (1 + exp(exp(logk) * (logd - mid)))
  )),
  lf(lraw ~ 1),
  lf(uraw ~ 1),
  lf(logk ~ 1),
  lf(mid ~ temp_c + (1 | Date) + (1 | Tank)),
  nl = TRUE
)

# Priors: same structure as the synthetic test in Steps 2-4, with
# mid_start anchored to the data and random-effect SD priors added.
# lraw/uraw centres target low ~ 0.02 and up ~ 0.98 under the ms bounds.
priors_shrimp <- c(
  set_prior(paste0("normal(", round(qlogis(0.02 / 0.49),   2), ", 1)"),
            class = "b", nlpar = "lraw"),
  set_prior(paste0("normal(", round(qlogis((0.98 - 0.51) / 0.49), 2), ", 1)"),
            class = "b", nlpar = "uraw"),
  set_prior(paste0("normal(", round(log(2), 2), ", 1)"),
            class = "b", nlpar = "logk"),
  set_prior(paste0("normal(", round(mid_start, 2), ", 1.5)"),
            class = "b", nlpar = "mid", coef = "Intercept"),
  set_prior("normal(0, 0.6)",  class = "b",  nlpar = "mid"),
  set_prior("exponential(2)",  class = "sd", nlpar = "mid", group = "Date"),
  set_prior("exponential(2)",  class = "sd", nlpar = "mid", group = "Tank"),
  set_prior("gamma(2, 0.1)",   class = "phi")
)

# Stan code check — same key question as before
sc_shrimp <- make_stancode(
  formula = formula_shrimp,
  data    = shrimp_tdt,
  family  = beta_binomial(link = "logit"),
  prior   = priors_shrimp
)

lines_with_n_total <- grep("vn_total", strsplit(sc_shrimp, "\n")[[1]], value = TRUE)
cat("Lines containing 'vn_total' in the shrimp Stan code:\n")
cat(lines_with_n_total, sep = "\n")
```

**Interpret:**

- `vn_total[n]` everywhere → bug is fixed; safe to use `beta_binomial()` + `trials()`.
- `vn_total` without `[n]` → bug still present; keep the custom family.

## Step 6 — Quick shrimp fit (only if Step 5 shows `vn_total[n]`)

```{r shrimp-fit-test}
#| eval: false
fit_shrimp_test <- brm(
  formula  = formula_shrimp,
  data     = shrimp_tdt,
  family   = beta_binomial(),
  prior    = priors_shrimp,
  chains   = 2,
  cores    = 2,
  iter     = 1000,
  warmup   = 500,
  seed     = 123,
  init     = 0,
  backend  = "cmdstanr",
  control  = list(adapt_delta = 0.995, max_treedepth = 15)
)

summary(fit_shrimp_test)
# Expected:  b_mid_temp_c < 0,  exp(b_logk_Intercept) > 0
```

## Decision

| Step 3/5 result | Step 4/6 result | Action |
|---|---|---|
| `vn_total[n]` (both) | Samples cleanly | Replace custom family in `test_functions.qmd` |
| `vn_total[n]` (both) | Divergences / log(0) | Investigate init or prior, not the family |
| `vn_total` (no `[n]`) | — | Keep custom family; note brms version |
