---
title: "Feasibility: repeated-measures time-to-death data in bayesTLS (UW postdoc query)"
date: 2026-09-25
format:
  html:
    embed-resources: true
    toc: true
---

## The question

A UW postdoc monitors the *same* mollusc individuals through time (daily checks,
or near-continuously with biosensors) and records individual times of death. He
asks: bayesTLS assumes independent temperature × duration trials — could it
support non-independent, repeated-measures data? He was considering a Bayesian
log-logistic likelihood.

## Answer, in one paragraph

Yes — and it is an extension of our model, not a different one. Our 4PL curve
in log₁₀(duration) with asymptotes 0/1 **is** the survivor function of a
log-logistic time-to-death distribution (his proposed likelihood): the median
death time is $10^{mid}$ and the shape is $k/\ln 10$ (Check 1 below, machine
precision). Because $mid$ is linear in temperature, the joint model is a
log-logistic accelerated-failure-time model in which **z is literally the
acceleration coefficient**: each +1 °C multiplies median survival time by
$10^{-1/z}$. For repeated observation, the correct likelihood is the classic
discrete-time survival construction: one row per individual per check-interval
while alive, with death probability $p_j = 1 - S(t_j)/S(t_{j-1})$ —
**identical** to the interval-censored log-logistic likelihood (Check 2,
agreement ~1e-13; the identity even survives our low > 0 "immortal fraction"
and up < 1 cases). Our current binomial model is the single-interval special
case, so the extension nests everything we already do. A working brms
prototype fit this way recovered a known z to within 2% with clean
diagnostics, and the entire post-fit machinery (`tls()`, `derive_z()`, the
whole extraction engine) ran **unchanged** on that fit because the parameters
are the same.

Daniel's instinct was right in an important way: this *is* a hazards model —
a discrete-time hazard, the standard survival-analysis treatment of
interval-checked data. The precise point is that he does not need a **Cox
proportional-hazards** model, which would discard the z/CTmax
parameterisation; the log-logistic AFT form *keeps* our estimands intact.
(One condition: the clean "AFT" label holds when $k$ is constant across
temperatures; with $k \sim T$, the model is log-logistic at each temperature
and the z mapping is unaffected, but there is no single acceleration factor.)

## Evidence (all independently reproduced by adversarial re-runs)

- **Math**: exact 4PL ↔ log-logistic mapping (error ≤ 3e-16); person-period ↔
  interval-censored likelihood identity (≤ 2e-13) including up < 1 (death mass
  at t = 0) and low > 0 (defective, "cure fraction") cases; exact-time density
  for biosensor data verified — interval and exact-time individuals can share
  one likelihood.
- **Prototype**: simulated 4 temperatures × 40 individuals, daily checks,
  day-14 censoring; brms nonlinear Bernoulli person-period fit sampled in
  ~1 s; z posterior median 3.92 vs truth 4.00 (95% CrI [3.52, 4.32]);
  0 divergences, max R̂ 1.018 (one chain, one seed — an existence proof, not a
  calibration study). A `(1|tank)` cluster effect compiles into the same
  machinery.
- **Package audit** (verified against installed v1.0.0 on a real
  person-period fit): touched — `standardize_data()` (person-period builder),
  a response branch in `make_4pl_formula()`, `validate_4pl_family()`, a
  column-fill fix in `new_tdt_grid()`, and survival-vs-hazard semantics in
  `predict_survival_curves()`. **Untouched — the whole extraction engine**
  (`tls()`, `derive_z()`, `tls_engine.R`), which recovered z from the
  prototype as-is. Zero new `fit_4pl()` arguments needed (the Beta-family
  precedent).
- **Literature**: the original TDT (Bigelow 1921) *was* time-to-death;
  Kingsolver & Umbanhowar 2018 fit individual knockdown times with a hazard
  framework recovering log-linear TDT; ecotoxicology's GUTS models use exactly
  this conditional-binomial likelihood; Prentice & Gloeckler 1978 is the
  canonical identity. We found **no existing joint Bayesian individual-level
  TDT implementation** (targeted searches, 2024–2026) — Vajedsamiei is
  Bayesian but two-stage; Bullard 2026 frequentist and per-temperature. This
  is an open niche. *(Bigelow 1921, Prentice & Gloeckler 1978, and the GUTS
  papers are not yet in `bib/` — add via Zotero if this proceeds.)*

## Real limits (ranked)

1. **Ramping / fluctuating temperature under biosensor monitoring is
   unresolved.** Our likelihood assumes constant T per individual (as does
   the current package). `predict_heat_injury()` is post-fit dose
   *integration*, not a likelihood — a time-varying-T death-time likelihood
   is genuinely new research (GUTS/TKTD territory), not this extension. If
   the postdoc's biosensor work is under ramps, that part is future work.
2. **Left truncation**: individuals enrolled late must have their first
   interval anchored at enrolment — the naive 0-anchor is provably biased
   (demonstrated numerically in the stress runs). Implementation must carry
   entry times.
3. **Frailty identifiability**: `(1|id)` compiles, but with one death per
   individual it is weakly identified and prior-sensitive; tank/batch cluster
   effects are the honest default.
4. **No overdispersion parameter**: beta-binomial φ has no Bernoulli
   analogue; heterogeneity must be *modelled* (cluster REs), and this should
   be documented loudly.
5. Numerics: the first-interval log₁₀(0) stand-in needs an explicit
   indicator (2.5% error at very shallow k); model-comparison metrics
   (loo/R²) must group by individual, not by person-period row.
6. Biology: monotone death assumed; valve-gape "death" can misclassify —
   worth a sentence in any reply.

## Scope if we build it

A contained third response type ("time-to-event") beside counts and
proportions: most work in `standardize_data()` + one formula branch; the
estimation and extraction core untouched. Bigger than issue #12 (phylo), so
sequence after it — candidate for 1.2.0, and plausibly a small methods paper
of its own given the open niche. A proper simulation study (coverage,
calibration, frailty sensitivity) is mandatory before any release claim.

## Checks (runnable)

### Check 1 — 4PL is a log-logistic survivor function

```{r check1}
S4pl <- function(t, k, mid) 1 / (1 + exp(k * (log10(t) - mid)))   # low=0, up=1
F_ll <- function(t, scale, shape) 1 / (1 + (t / scale)^(-shape))  # log-logistic CDF
t <- 10^seq(-3, 3, length.out = 400)
for (par in list(c(4, 1.2), c(8, 0.5), c(1.5, -0.3))) {
  k <- par[1]; mid <- par[2]
  stopifnot(max(abs((1 - S4pl(t, k, mid)) -
                    F_ll(t, scale = 10^mid, shape = k / log(10)))) < 1e-12)
}
"mapping verified: scale = 10^mid (t50), shape = k/ln(10)"
```

### Check 2 — person-period Bernoulli = interval-censored likelihood

```{r check2}
set.seed(1)
k <- 6; mid <- 0.9; checks <- 0:14                       # daily checks
td  <- 10^(mid + rlogis(60) * log(10) / k)               # true death times
int <- findInterval(td, checks)                           # death interval index
obs_die  <- td <= max(checks)

# (i) interval-censored loglik
Fll <- function(t) ifelse(t <= 0, 0, 1 - 1/(1 + exp(k * (log10(t) - mid))))
ll_ic <- sum(log(Fll(checks[int + 1]) - Fll(checks[int]))[obs_die]) +
         sum(!obs_die) * log(1 - Fll(max(checks)))

# (ii) person-period conditional Bernoulli
ll_pp <- 0
for (i in seq_along(td)) {
  last <- if (obs_die[i]) int[i] + 1 else length(checks)
  for (j in 2:last) {
    S1 <- 1 - Fll(checks[j - 1]); S2 <- 1 - Fll(checks[j])
    p  <- 1 - S2 / S1
    died <- obs_die[i] && j == last
    ll_pp <- ll_pp + ifelse(died, log(p), log(1 - p))
  }
}
c(interval_censored = ll_ic, person_period = ll_pp, diff = ll_ic - ll_pp)
stopifnot(abs(ll_ic - ll_pp) < 1e-10)
"identity verified: repeated measures are fully absorbed by conditioning"
```

### Check 3 — the working prototype (not run at render; recorded results)

Full script: session scratchpad `tte_prototype.R` (person-period simulate →
brms nonlinear Bernoulli, `p = 1 - S(t_hi)/S(t_lo)`, `mid ~ temp_c`,
cmdstanr, 1 chain / 800 iter). Recorded: z median **3.923** vs truth 4.000
(CrI [3.522, 4.320]); mid slope −0.2549 vs −0.25; 0 divergences; R̂ ≤ 1.018.
The installed package's `tls()` then extracted z from this fit **without any
modification**.

```{r check3, eval=FALSE}
# person-period fit skeleton (see scratchpad tte_prototype.R for the full script)
bf_pp <- brms::bf(
  death ~ 1 - (1 + exp(exp(logk) * (lt_lo - mid))) /
              (1 + exp(exp(logk) * (lt_hi - mid))),
  logk ~ 1, mid ~ temp_c, nl = TRUE)
fit <- brms::brm(bf_pp, data = person_period, family = brms::bernoulli("identity"),
                 backend = "cmdstanr", chains = 1, iter = 800)
```

## Addendum (2026-09-26)

Drew clarified: constant temperatures (no ramping), cohort checks or
continuous monitoring, plus a coral time-to-10%-mortality dataset. His
proposed "trivial reformat" (cumulative proportions into the current model)
is quantified in `2026-09-26-naive-cumulative-refit-overconfidence.qmd`:
naive binomial ~28% too-narrow CrIs (coverage 16/20), the beta-binomial
default largely compensated in that one benign design, and the correct
conditional reformat is equally trivial. The draft reply below is
SUPERSEDED by the updated draft in that note.

## Draft reply to the postdoc (SUPERSEDED — see 2026-09-26 note)

> Thanks — you've read the framework exactly right: the released bayesTLS
> likelihood assumes independent temperature × duration batches, so repeatedly
> observed cohorts violate it. The good news is that your instinct about the
> log-logistic is more right than you may realise: our 4PL in log₁₀(duration)
> *is* the log-logistic survivor function (median death time $10^{mid}$, shape
> $k/\ln 10$), so the thermal sensitivity z and CTmax you'd want are exactly
> preserved under a proper time-to-event treatment. For interval checks (daily
> scoring) the correct likelihood is the standard discrete-time survival
> construction — person-period records with conditional death probability
> $1 - S(t_j)/S(t_{j-1})$ — which is algebraically identical to the
> interval-censored log-logistic likelihood; for biosensor-resolution data the
> exact-time density applies, and both can be mixed in one model. We have
> prototyped this in brms with the same parameterisation and recovered z
> cleanly, with our post-fit machinery working unchanged, so yes: it is on our
> radar as a planned extension rather than a redesign. Two honest caveats:
> individual-level frailty is weakly identified with one death per animal
> (tank/batch effects are the robust default), and if your biosensor work is
> under ramping rather than constant temperatures, that requires the
> damage-integral machinery and is a genuinely open modelling problem we're
> thinking about. We'd be happy to hear more about your design — a worked
> example with real mollusc data would make an excellent test case.

## Open questions for Daniel

1. Reply + open a GitHub issue to track it (visible roadmap), or reply only?
2. Sequence: after #12 (phylo) as 1.2.0? The niche looks open enough that
   this could be a small stand-alone methods paper with the postdoc as a test
   case / collaborator — worth considering.
3. Is his biosensor work constant-T or ramping? The answer decides whether we
   promise an extension or a research collaboration.
