---
title: "Joint Bayesian 4PL vs two-stage TDT estimators — bias, coverage, and sensitivity"
date: 2026-05-16
date-modified: 2026-05-19
author: "Daniel Noble (with Claude assistance)"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 4
    number-sections: true
    fig-cap-location: bottom
    code-fold: true
    df-print: kable
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(patchwork)
library(scales)

results_dir <- here::here("output", "sim_twostage")

# ---- Method order and colours used throughout --------------------------
# Six methods, organised into three families (joint, binomial two-stage,
# beta-binomial two-stage). Within each family, the "default" variant
# (relative extract / Normal-quantile CIs) and the "alternative" variant
# (absolute extract / t-quantile CIs) get a paired colour: dark for
# default, light for alternative.
method_levels <- c("two_stage_bin", "two_stage_bin_t",
                   "two_stage_bb",  "two_stage_bb_t",
                   "joint_4pl",     "joint_4pl_abs")
method_labels <- c("Two-stage bin (Normal CI)",
                   "Two-stage bin (t CI)",
                   "Two-stage BB (Normal CI)",
                   "Two-stage BB (t CI)",
                   "Joint 4PL (relative)",
                   "Joint 4PL (absolute)")
method_cols   <- setNames(c("#54278f", "#cbc9e2",   # purples (bin)
                            "#a63603", "#fdae6b",   # oranges (BB)
                            "#006d2c", "#a1d99b"),  # greens  (joint)
                          method_labels)

# ---- Quantity labels with units ---------------------------------------
quantity_labeller <- as_labeller(c(
  z          = "z (°C)",
  CTmax_1hr  = "CT[max,1h] (°C)"
))

# ---- Two-truth setup ---------------------------------------------------
# Each method targets a specific population quantity. The driver writes
# the *absolute*-50% truth (computed by OLS through per-T absolute LT50
# values) into every per-sim row regardless of method. That is the right
# target for joint_4pl_abs, two_stage_bin/_t, two_stage_bb/_t — but NOT
# for joint_4pl (relative extract), which targets `mid(T) = log10(60)`.
#
# In this study, the mid line is the same in every cell:
# beta0 = 1.5, beta1 = -0.15, T_bar = 34 (never varied by any --u_0 /
# --ell_0 / --u_beta1 etc. override). The relative truths are therefore
# scenario-invariant constants and can be precomputed here:
T_BAR        <- 34
MID_BETA0    <- 1.5
MID_BETA1    <- -0.15
LOG10_T_REF  <- log10(60)
TRUTH_REL    <- c(
  z         = -1 / MID_BETA1,
  CTmax_1hr =  T_BAR + (LOG10_T_REF - MID_BETA0) / MID_BETA1
)

# ---- BB Stage-1 pathology screening -----------------------------------
# Threshold for BB Stage-1 pathology screening. The Stage-1 beta-binomial
# GLM occasionally converges to a degenerate solution (Hessian numerically
# rank-deficient on the link scale) that passes the simulation driver's
# "finite & negative slope" success check but produces wildly biased
# Stage-2 estimates. We treat as pathological — and drop from the
# corresponding cell — any sim whose two_stage_bb estimate has
# |z bias| > 20 °C OR |CTmax bias| > 10 °C. The cut-offs are well above
# any legitimate two-stage failure observed in Scenarios 3-8
# (Scen 7 u_0=0.65 BB CTmax bias is −6.3 °C). 153 / 25,000 sims are
# screened out across all cells (0.6%). Only the two_stage_bb and
# two_stage_bb_t method rows from those sims are dropped; the other
# four methods on the same sims are retained.
BB_Z_THRESH     <- 20
BB_CTMAX_THRESH <- 10

# Aggregator matching summarise_mcse() in scripts/simulations/sim_functions.R.
mcse_aggregate <- function(per_sim) {
  per_sim |>
    dplyr::filter(success) |>
    dplyr::group_by(scenario, method, quantity) |>
    dplyr::summarise(
      n            = dplyr::n(),
      mean_bias    = mean(bias, na.rm = TRUE),
      mcse_bias    = stats::sd(bias, na.rm = TRUE) / sqrt(dplyr::n()),
      rmse         = sqrt(mean(bias^2, na.rm = TRUE)),
      mcse_rmse    = stats::sd(bias^2, na.rm = TRUE) /
                       (2 * sqrt(mean(bias^2, na.rm = TRUE)) *
                        sqrt(dplyr::n())),
      coverage     = mean(covered, na.rm = TRUE),
      mcse_cov     = sqrt(mean(covered, na.rm = TRUE) *
                          (1 - mean(covered, na.rm = TRUE)) /
                          dplyr::n()),
      med_width    = stats::median(width, na.rm = TRUE),
      .groups      = "drop"
    ) |>
    dplyr::mutate(dplyr::across(c(mean_bias, mcse_bias, rmse, mcse_rmse,
                                  coverage, mcse_cov, med_width),
                                ~ round(.x, 4)))
}

load_summaries <- function(prefix) {
  # Load per-sim files (not pre-computed summaries) so we can apply the
  # BB-outlier screen and the relative-truth substitution before
  # aggregation.
  files <- list.files(results_dir,
                      pattern = paste0("^per_sim_", prefix, "_.*[.]rds$"),
                      full.names = TRUE)
  per_sim <- dplyr::bind_rows(lapply(files, readRDS))

  # ---- BB-outlier screen ----------------------------------------------
  bad <- per_sim |>
    dplyr::filter(method == "two_stage_bb",
                  (quantity == "z"         & abs(bias) > BB_Z_THRESH) |
                  (quantity == "CTmax_1hr" & abs(bias) > BB_CTMAX_THRESH)) |>
    dplyr::distinct(scenario, sim_id) |>
    dplyr::mutate(bb_pathological = TRUE)

  per_sim_screened <- per_sim |>
    dplyr::left_join(bad, by = c("scenario", "sim_id")) |>
    dplyr::filter(
      is.na(bb_pathological) |
      !(method %in% c("two_stage_bb", "two_stage_bb_t"))
    ) |>
    dplyr::select(-bb_pathological)

  # ---- Relative-truth substitution for joint_4pl ----------------------
  # joint_4pl (the relative extract) is compared against the relative
  # truth (mid(T) = log10(60)), not the absolute-50% truth that the
  # driver wrote into `truth`. We rewrite truth, bias, and covered for
  # these rows only; all other methods keep the absolute truth.
  per_sim_screened <- per_sim_screened |>
    dplyr::mutate(
      truth_use = ifelse(method == "joint_4pl",
                         TRUTH_REL[as.character(quantity)],
                         truth),
      bias      = estimate - truth_use,
      covered   = is.finite(lower) & is.finite(upper) &
                  lower <= truth_use & truth_use <= upper,
      truth     = truth_use
    ) |>
    dplyr::select(-truth_use)

  mcse_aggregate(per_sim_screened)
}

load_diagnostics <- function(prefix) {
  files <- list.files(results_dir,
                      pattern = paste0("^diag_", prefix, "_.*[.]rds$"),
                      full.names = TRUE)
  dplyr::bind_rows(lapply(files, readRDS))
}

prep_summary <- function(df) {
  df %>%
    mutate(
      method   = factor(method, levels = method_levels, labels = method_labels),
      quantity = factor(quantity, levels = c("CTmax_1hr", "z")),
      bias_lo  = mean_bias - 1.96 * mcse_bias,
      bias_hi  = mean_bias + 1.96 * mcse_bias,
      cov_lo   = coverage  - 1.96 * mcse_cov,
      cov_hi   = coverage  + 1.96 * mcse_cov
    )
}

theme_proj <- theme_classic(base_size = 11) +
  theme(strip.background = element_blank(),
        strip.text       = element_text(face = "bold"),
        legend.position  = "bottom",
        legend.title     = element_blank(),
        panel.spacing    = unit(0.8, "lines"))

# ---- Per-scenario forest figure (used for Scen 1-5) -------------------
plot_scen_forest <- function(df) {
  d <- prep_summary(df)

  d <- d %>%
    mutate(panel_n = paste0("n[reps] = ",
                            sub(".*_n([0-9]+)$", "\\1", scenario)))

  p_bias <- ggplot(d, aes(x = mean_bias, y = method, colour = method)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbarh(aes(xmin = bias_lo, xmax = bias_hi), height = 0.25) +
    geom_point(size = 2.6) +
    facet_grid(panel_n ~ quantity, scales = "free_x",
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    labs(x = "Mean bias (estimate − truth)", y = NULL,
         title = "(a) Bias") +
    theme_proj + theme(legend.position = "none")

  p_cov <- ggplot(d, aes(x = coverage, y = method, colour = method)) +
    geom_vline(xintercept = 0.95, linetype = "dashed", colour = "grey50") +
    geom_errorbarh(aes(xmin = cov_lo, xmax = cov_hi), height = 0.25) +
    geom_point(size = 2.6) +
    facet_grid(panel_n ~ quantity, scales = "free_x",
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
    labs(x = "Empirical 95% CI coverage", y = NULL,
         title = "(b) Coverage") +
    theme_proj + theme(legend.position = "none")

  p_bias / p_cov + plot_layout(heights = c(1, 1))
}

# ---- Per-cell summary table (compact) ---------------------------------
make_table <- function(df) {
  df %>%
    mutate(
      method   = factor(method, levels = method_levels, labels = method_labels),
      quantity = factor(quantity, levels = c("z", "CTmax_1hr"),
                        labels = c("z", "CTmax_1hr"))
    ) %>%
    arrange(scenario, quantity, method) %>%
    transmute(
      Cell      = scenario,
      Method    = method,
      Quantity  = quantity,
      n_ok      = n,
      Bias      = sprintf("%+.3f (%.3f)", mean_bias, mcse_bias),
      RMSE      = sprintf("%.3f", rmse),
      Coverage  = sprintf("%.1f%%", 100 * coverage),
      Width     = sprintf("%.2f", med_width)
    )
}

# ---- Sweep figure: bias / coverage as a function of swept parameter ---
plot_sweep <- function(df, sweep_x, x_label, x_breaks = waiver()) {
  d <- prep_summary(df)

  p_bias <- ggplot(d, aes(x = .data[[sweep_x]], y = mean_bias, colour = method)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = bias_lo, ymax = bias_hi), width = 0) +
    geom_point(size = 2.4) +
    geom_line() +
    facet_wrap(~ quantity, scales = "free_y",
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    scale_x_continuous(breaks = x_breaks) +
    labs(x = x_label, y = "Mean bias",
         title = "(a) Bias") +
    theme_proj

  p_cov <- ggplot(d, aes(x = .data[[sweep_x]], y = coverage, colour = method)) +
    geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = cov_lo, ymax = cov_hi), width = 0) +
    geom_point(size = 2.4) +
    geom_line() +
    facet_wrap(~ quantity,
               labeller = labeller(quantity = quantity_labeller)) +
    scale_colour_manual(values = method_cols) +
    scale_x_continuous(breaks = x_breaks) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
    labs(x = x_label, y = "Empirical 95% coverage",
         title = "(b) Coverage") +
    theme_proj

  p_bias / p_cov + plot_layout(guides = "collect", heights = c(1, 1)) &
    theme(legend.position = "bottom")
}
```

::: {.callout-note}
This note supersedes
[`2026-05-14-two-stage-bias-sim-results.qmd`](2026-05-14-two-stage-bias-sim-results.html)
and
[`2026-05-15-sensitivity-sweeps.qmd`](2026-05-15-sensitivity-sweeps.html).

**2026-05-19 rebuild.** All 25 cells re-run with new sampler settings
(3 chains × 3000 iter, warmup 1500, `max_treedepth = 16`) and
$N_{\text{sim}} = 1000$ across every scenario (the original sensitivity
sweeps used 500). Each fit now reports **six** estimators per dataset:
joint 4PL with `target_surv = "relative"` and `target_surv = "absolute"`,
each two-stage variant (binomial / beta-binomial Stage 1) with both
Normal-quantile and t-quantile Stage-2 delta-method CIs. Methods are
documented in
[`2026-05-13-two-stage-bias-sim-methods.qmd`](2026-05-13-two-stage-bias-sim-methods.qmd);
this note refers back rather than restating.
:::

# Aim

The thermal-death-time (TDT) framework is overwhelmingly fitted with a
two-stage pipeline: per-temperature binomial GLMs estimate
$\log_{10}\text{LT}_{50}$, then those five (or so) point estimates are
regressed against temperature with unweighted OLS to extract the
classical quantities $z$ and $CT_{\max,1\text{h}}$. We want to know
when this pipeline gives the correct answer with the correct
uncertainty, and when it does not.

The joint Bayesian 4PL fits the full dose-response surface in one go.
Algebraically it returns the same $z$ and $CT_{\max,1\text{h}}$ when
the strict-equivalence assumptions hold (constant asymptotes near 0/1,
no overdispersion, common slope across temperature). When those
assumptions break — partial asymptotes, drifting upper bound,
overdispersed counts, sparse design — there is no algebraic guarantee
that the two are the same, and the two-stage's estimating equations
have no parameter to absorb the missing structure.

The simulation study compares six estimators side-by-side over eight
controlled scenarios (25 cells, 25,000 simulated datasets). Two
estimators isolate the joint-4PL's *threshold definition* (relative vs
absolute) on the same fit; four isolate the two-stage's *likelihood*
(binomial vs beta-binomial Stage 1) and *small-sample CI correction*
(Normal vs t-quantile delta-method Stage 2). The aim is not to advocate
for the joint 4PL on principle, but to map where the two-stage's deficit
becomes operationally important — and where it does not.

# Common methods

::: {.callout-tip}
Full details, including the analytical truth construction, the
parameter feasibility check that runs before any compute starts, and
the diagnostic-criterion table, are in
[`2026-05-13-two-stage-bias-sim-methods.qmd`](2026-05-13-two-stage-bias-sim-methods.qmd).
This section gives only the headline numbers needed to read the
results.
:::

**Six estimators per dataset.**

| Label | Description |
|---|---|
| `joint_4pl` | `bayesTLS::extract_tdt(target_surv = "relative")`: $z$ and $CT_{\max,1\text{h}}$ extracted from the posterior at the asymptote midpoint $p = (\text{low}+\text{up})/2$ (the package default). |
| `joint_4pl_abs` | Same fit, `target_surv = "absolute"`: extraction at $p = 0.5$ literal. Matches the analytical truth function's definition. |
| `two_stage_bin` | Field default: per-temperature binomial GLM Stage 1, unweighted Stage-2 OLS, Normal-quantile delta-method CIs ($z_{0.975} = 1.96$). |
| `two_stage_bin_t` | Same point estimate as `two_stage_bin`; Stage-2 CIs use t-quantile delta-method ($t_{0.975, n_T-2}$ — 3.18 at full design, 12.71 at sparse). Small-sample correction. |
| `two_stage_bb` | "Patched" two-stage: per-T beta-binomial GLMM Stage 1 via `glmmTMB`, unweighted Stage-2 OLS, Normal-quantile CIs. Isolates the *likelihood* contribution of the field-default deficit. |
| `two_stage_bb_t` | Same as `two_stage_bb` but with t-quantile Stage-2 CIs. |

The joint 4PL is a `brms`-fit non-linear `beta_binomial(link = "identity")`
model in which all four 4PL sub-parameters carry a centred-temperature
slope. Both joint extracts come from one fit (no extra compute).

**Design grid.** Five assay temperatures (30, 32, 34, 36, 38 °C) ×
six exposure durations (1, 5, 15, 45, 135, 405 minutes) = 30 cells per
dataset. Each cell is replicated by $n_{\text{reps}}$ independent cups,
each with $N \sim \text{Uniform}\{10, \dots, 20\}$ individuals.
A Scenario-1 dataset with $n_{\text{reps}}=5$ is therefore 150 cups
totalling roughly 2,000 individuals. Scenario 8 explicitly varies the
grid (sparse 3 T × 4 durations = 12 cells per dataset).

**Two truths, one per CTmax/z definition.** Five of the six estimators
target the *absolute*-50% population quantity (joint_4pl_abs,
two_stage_bin/_t, two_stage_bb/_t) and are compared against the
absolute truth from `compute_ols_truth()`: $\log_{10}\text{LT}_{50}^{\text{abs}}(T)
= \text{mid}(T) + \frac{1}{k(T)}\ln\frac{u(T)-0.5}{0.5-\ell(T)}$
evaluated at every design temperature and OLS-projected. The sixth,
`joint_4pl` (the relative extract), targets the *relative*
quantity — temperature where $\text{mid}(T) = \log_{10}60$ — and is
compared against a **separate truth** that uses the mid line directly,
no asymmetry correction. In this study `mid(T) = 1.5 + (-0.15)(T - 34)`
in every cell (no scenario varies the mid line), so the relative truth
is the same constant across all 25 cells:
$z_{\text{true}}^{\text{rel}} = 6.667$ °C and
$CT_{\max,1\text{h}}^{\text{rel}} = 32.147$ °C. **Without this
two-truth substitution, the joint_4pl rows would carry a labelling
offset against the absolute truth wherever the asymptotes deviate
from 0/1 — that offset was the original open issue 1 from the
2026-05-16 first pass and is what the rewrite resolves.**

**Sample sizes.** $N_{\text{sim}} = 1000$ datasets per cell across
**every** scenario in this rebuild (the 2026-05-16 first pass used 500
for Scen 6–8; standardised to 1000 in the 2026-05-19 re-run). Cells are
pre-flighted before any compute starts, including a cross-cell
distinctness check on data hashes that catches CLI plumbing bugs.

**Performance metrics.** For each estimator and each output quantity
($z$, $CT_{\max,1\text{h}}$): mean signed bias and its Monte-Carlo SE,
RMSE, empirical 95% coverage and its MC SE (binomial proportion), and
median 95% CI width. All computed across the per-simulation rows in the
cell.

**BB Stage-1 pathology screening.** The driver's success check
(finite coefficients, negative slope) lets through a small fraction of
beta-binomial Stage-1 fits whose Hessian is numerically rank-deficient
— these pass as "successful" but produce wildly biased Stage-2
estimates (z bias up to 849 °C; CTmax bias up to 433 °C). Across all
25,000 sims, 153 (0.6 %) match this pattern, concentrated in Scenario 1
(boundary asymptotes) and the Scenario 8 sparse cells. **Sims where the
two-stage BB estimate has |z bias| > 20 °C OR |CTmax bias| > 10 °C are
dropped from the `two_stage_bb` and `two_stage_bb_t` rows**, but
retained for the four non-BB methods. The cut-offs are well above the
largest legitimate two-stage failure observed anywhere in the study
(Scenario 7 $u_0 = 0.65$, BB CTmax bias = −6.3 °C). All summaries
reported below use the screened per-sim data.

**Diagnostics.** Joint 4PL chains: **3 chains × 3000 iterations
(warmup 1500, post-warmup 4500), `adapt_delta = 0.95`,
`max_treedepth = 16`**. R̂ < 1.01, ESS > 400, zero divergent transitions,
BFMI > 0.3. Per-cell pass rates are reported alongside the bias /
coverage tables. The new sampler settings are an upgrade from the
2 chains × 2000 / `max_treedepth = 12` used in the earlier pass.

# Scenario 1 — Strict equivalence baseline

**Question.** When all strict-equivalence assumptions hold (asymptotes
pinned at 0 and 1, no overdispersion, single slope across temperature),
do the three estimators agree, and is each calibrated? This is the
floor scenario — divergence here would signal a bug.

**Prediction.** All three estimators are correctly specified for this
DGP (the joint 4PL's beta-binomial likelihood collapses to binomial as
$\phi \to \infty$). Bias should be within MC SE and coverage near 0.95
for all three. We expected the joint 4PL to deliver the narrowest CIs
(joint fit pools information across temperature), with the two two-stage
variants near-identical (no overdispersion to exploit).

**Design.** DGP: `low = 0.001`, `up = 0.999`, `k = 8`, midpoint linear in
$T$ with slope $-0.15$, binomial likelihood. True $z = 6.667$, true
$CT_{\max,1\text{h}} = 32.146\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen1-table}
scen1 <- load_summaries("scen1")
make_table(scen1)
```

```{r}
#| label: fig-scen1
#| fig-cap: "Scenario 1 (strict equivalence baseline). (a) Mean bias with 95% MC-SE intervals; dashed line at zero. (b) Empirical 95% CI coverage with 95% binomial intervals; dashed line at the nominal 0.95. Top row: $n_{reps}=3$; bottom row: $n_{reps}=5$. Both quantities ($z$ and $CT_{\\max,1h}$) are within MC SE of zero bias and within MC SE of nominal coverage for all three estimators, as predicted."
#| fig-width: 8
#| fig-height: 8
plot_scen_forest(scen1)
```

**Read.** Bias predictions held for all six estimators (signed bias
within MC SE of zero on both quantities, both cells). The Normal-CI
two-stage estimators **under-cover** at 84–86% on both $z$ and
$CT_{\max,1\text{h}}$, regardless of Stage-1 family — the Stage-2 OLS
on five LT50 points propagates less variance than the underlying
sampling distribution requires. **The t-quantile correction at
$df = n_T - 2 = 3$ closes the gap**: both `two_stage_bin_t` and
`two_stage_bb_t` now cover at 93–97% on both quantities. This is the
clean regime where the small-sample CI fix does the job. Joint 4PL
covers at 97–99% on both extracts (slightly over nominal but within MC
SE of 0.95); the relative and absolute extracts are within 0.01 °C of
each other on CTmax (asymptotes are at the reparam boundary, so the
asymmetry correction is essentially zero).

The BB Stage-1 variant produced a small fraction of pathological fits
in this scenario (passing the driver's success check but with
nonsensical Stage-2 estimates); those sims are now screened out for
the BB methods only — see "BB Stage-1 pathology screening" in the
Common methods section. The n_ok counts for `two_stage_bb` /
`two_stage_bb_t` in the table reflect the post-screen sample size.

# Scenario 2 — Likelihood misspecification only

**Question.** With identical curve shape to Scenario 1 but
beta-binomial overdispersion (φ = 5) added to the data-generating
likelihood, what does the field default lose by ignoring overdispersion?

**Prediction.** All three should remain unbiased — the curve shape is
unchanged, so the regression's mean structure is correct under all three
estimators. We expected the binomial two-stage to have wider CIs than
the patched (beta-binomial) two-stage, but coverage near nominal
because Stage-2 OLS reads the residual variance off the
$\log_{10}\text{LT}_{50}$ scatter. The patched two-stage should give
narrower, calibrated CIs. The joint 4PL — narrowest, calibrated.

**Design.** Same shape as Scenario 1 (boundary asymptotes), but DGP
likelihood is beta-binomial with φ = 5. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen2-table}
scen2 <- load_summaries("scen2")
make_table(scen2)
```

```{r}
#| label: fig-scen2
#| fig-cap: "Scenario 2 (likelihood misspecification only, beta-binomial DGP). All three estimators stay within MC SE of zero bias on both quantities. Coverage of the two two-stage variants drops to 84–90%, regardless of Stage-1 family; joint 4PL coverage stays near nominal 0.95."
#| fig-width: 8
#| fig-height: 8
plot_scen_forest(scen2)
```

**Read.** Bias predictions held: all six estimators within MC SE of
zero on both quantities. Normal-quantile two-stage coverage runs 84–89%
on both quantities, regardless of Stage-1 family. **The t-quantile
correction lifts coverage to 93–97%** — close enough to nominal that
under overdispersion alone, the small-sample CI fix is sufficient
without changing the architecture. The patched (BB) Stage 1 by itself
does not rescue coverage; the Stage-2 OLS already reads dispersion off
the per-T point spread, but under-propagates it under Normal CIs. Joint
4PL keeps coverage at 93–96% on both extracts; relative and absolute
agree to within 0.01 °C on CTmax (asymptotes still at the boundary).

# Scenario 3 — Heat lowers max survival (asymmetric u drift)

**Question.** When the upper asymptote drops with temperature
(intrinsic heat-induced mortality reduces the proportion of the cohort
ever susceptible to the timed exposure) but neither two-stage estimator
has an asymptote parameter, does the missing structure get absorbed
into Stage 1's effective midpoint, biasing $z$ and $CT_{\max,1\text{h}}$?

**Prediction.** Both two-stage variants should be biased on at least
one quantity, with bias that does not shrink with $n_{\text{reps}}$.
Joint 4PL should be unbiased and calibrated.

**Design.** DGP: `up` drops 1 pp/°C from baseline 0.92, `low` constant
at 0.05, `k = 8`, midpoint linear in $T$ with slope $-0.15$,
beta-binomial likelihood (φ = 5). True $z = 6.537$, true
$CT_{\max,1\text{h}} = 32.124\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen3-table}
scen3 <- load_summaries("scen3")
make_table(scen3)
```

```{r}
#| label: fig-scen3
#| fig-cap: "Scenario 3 (asymmetric u drift, beta-binomial). Two-stage CTmax bias is systematically negative (~ −0.27 to −0.36 °C) and does not shrink with $n_{reps}$; coverage on the BB CTmax falls to 75% at $n_{reps}=5$. Joint 4PL bias is within MC SE on CTmax in both cells; small positive z bias persists (+0.06 to +0.10) but coverage stays at 94–96%."
#| fig-width: 8
#| fig-height: 8
plot_scen_forest(scen3)
```

**Read.** Two-stage CTmax bias is systematically negative across both
cells (−0.27 °C binomial, −0.36 °C beta-binomial) and **does not shrink
with $n_{\text{reps}}$**. Normal-CI coverage on the BB CTmax drops to
75% at $n_{\text{reps}}=5$; the t-correction lifts it to 91% but cannot
reach nominal because the bias is too large for any CI inflation to
fully catch. The bin-Stage-1 + t-CI variant at 93% is closer to nominal
because its bias is smaller (−0.27 vs −0.35 °C).

**Joint 4PL absorbs the T-varying `up`.** Both extracts cover at 95% on
CTmax with bias within MC SE: `joint_4pl` (relative) +0.034 / +0.001 °C;
`joint_4pl_abs` +0.009 / −0.019 °C. **The absolute extract trims the
small positive z bias that the relative extract carries** (+0.06–0.10 →
−0.04 to −0.06 °C). Both at 95% z coverage.

# Scenario 4 — Asymptotes compress (symmetric drift)

**Question.** When *both* asymptotes drift toward each other
symmetrically (`up` drops, `low` rises), the logit-difference is
nearly preserved but the curve range shrinks. How does each estimator
handle this in $z$ and $CT_{\max,1\text{h}}$?

**Prediction.** True $z$ barely changes (the slope around the midpoint
is preserved). Two-stage methods should show smaller bias than
Scenario 3 because the symmetric perturbation partially cancels at the
midpoint. Joint 4PL near-nominal.

**Design.** `up` drops 1 pp/°C from baseline 0.92; `low` rises 1 pp/°C
from baseline 0.05; `k = 8`; midpoint linear in $T$ with slope $-0.15$;
beta-binomial likelihood (φ = 5). True $z = 6.658$, true
$CT_{\max,1\text{h}} = 32.090\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen4-table}
scen4 <- load_summaries("scen4")
make_table(scen4)
```

```{r}
#| label: fig-scen4
#| fig-cap: "Scenario 4 (asymptotes compress). Two-stage z bias *flips sign relative to Scenario 3*: now positive (+0.39 binomial, +0.55 beta-binomial). CTmax bias remains negative and grows (~ −0.38 to −0.58 °C). Coverage on the BB CTmax falls to 64% at $n_{reps}=5$. Joint 4PL coverage holds at 94–95%."
#| fig-width: 8
#| fig-height: 8
plot_scen_forest(scen4)
```

**Read.** Two-stage CTmax bias is now larger than in Scenario 3
(−0.38 °C binomial, −0.58 °C beta-binomial) and the $z$ bias **flips
sign**: where Scenario 3 had two-stage $z$ biased low (−0.08 to −0.17),
Scenario 4 has it biased high (+0.39 to +0.56). Biases are insensitive
to $n_{\text{reps}}$ — model misspecification, not sampling noise.
Normal-CI coverage on the BB CTmax drops to 64% at $n_{\text{reps}}=5$.

**The t-correction partially rescues coverage**: BB CTmax 64% → 87%,
bin CTmax 76% → 91% (at $n_{\text{reps}}=5$). On z, the t-correction
reaches nominal (94–95%) because the relative bias-to-CI ratio is
smaller. The patched (BB) Stage-1 is consistently *worse* than the
binomial Stage-1 on both quantities here, contradicting the prediction
that the patch would help.

**Joint 4PL holds bias within MC SE on CTmax** for both extracts (rel
−0.009 / +0.021 °C; abs −0.046 / −0.023 °C) with 94–95% coverage. The
absolute extract again trims the small positive z bias (+0.09 → +0.08;
+0.04 → +0.03 °C) — modest but consistent.

# Scenario 5 — Heat sharpens curve (varying k)

**Question.** When the slope $k$ varies with temperature (curves get
sharper as it gets hotter), the per-T binomial GLM still recovers a
midpoint, but the Stage-2 OLS regression of $\log_{10}\text{LT}_{50}$
on $T$ assumes a linear relationship. How robust is the two-stage to
T-varying $k$?

**Prediction.** Smallest bias of the shape-perturbing scenarios. Two
stage variants: slight bias, coverage near nominal. Joint 4PL: bias
within MC SE, calibrated.

**Design.** `up = 0.92`, `low = 0.05`, $k$ linear in $T$ (8 at $T=34$,
slope +0.25 — so k = 7 at T=30, k = 9 at T=38). Midpoint linear in $T$
with slope $-0.15$. Beta-binomial likelihood (φ = 5). True $z = 6.679$,
true $CT_{\max,1\text{h}} = 32.084\,^\circ\text{C}$. Two cells:
$n_{\text{reps}} \in \{3, 5\}$. $N_{\text{sim}} = 1000$ per cell.

```{r scen5-table}
scen5 <- load_summaries("scen5")
make_table(scen5)
```

```{r}
#| label: fig-scen5
#| fig-cap: "Scenario 5 (varying k). Two-stage z biased high (+0.45 to +0.55), CTmax biased low (−0.44 to −0.62 °C); coverage on BB CTmax falls to 62% at $n_{reps}=5$. Joint 4PL bias within MC SE on z; small positive CTmax bias (+0.03, +0.06); coverage holds at 93–95%."
#| fig-width: 8
#| fig-height: 8
plot_scen_forest(scen5)
```

**Read.** The prediction was wrong: $k$ varying with $T$ produces
bias as large as Scenario 4. Two-stage $z$ biased high (+0.45 to +0.55)
and CTmax biased low (−0.44 to −0.62 °C); Normal-CI BB CTmax coverage
62% at $n_{\text{reps}}=5$.

**t-correction effect (CTmax)**: BB 62% → 86%, bin 69% → 90% at
$n_{\text{reps}}=5$ — same pattern as Scen 4: the t-fix helps but
can't reach nominal in the presence of structural bias. On z, the
t-correction does reach nominal (95–96%).

Joint 4PL absorbs the $k(T)$ drift. The absolute extract again improves
on the relative: CTmax bias +0.027 → −0.032 °C at $n_{\text{reps}}=3$,
+0.056 → −0.012 °C at $n_{\text{reps}}=5$, both well within MC SE.
Coverage 93–95% on both extracts.

# Scenario 6 — Sweep: strength of u drift

**Question.** How does the two-stage bias scale with the strength of
the upper-asymptote drift? Is there a threshold of $|\beta_u|$ below
which the bias is negligible in practice?

**Prediction.** Bias monotonically increasing in $|\beta_u|$; coverage
correspondingly worse. Joint 4PL: bias within MC SE throughout.

**Design.** Same shape as Scenario 3 but vary the upper-asymptote
slope: $\beta_u \in \{-0.005, -0.010, -0.015, -0.019\}$ pp/°C.
$n_{\text{reps}} = 5$, $N_{\text{sim}} = 1000$ per cell. Four cells.
True $z$ ranges 6.42–6.60; true $CT_{\max,1\text{h}}$ ranges 32.11–32.15.

```{r}
#| label: fig-scen6
#| fig-cap: "Scenario 6 sweep over $\\beta_u$ (strength of upper-asymptote drift). (a) Bias as a function of $\\beta_u$, with 95% MC-SE intervals; dashed line at zero. (b) Empirical 95% coverage; dashed line at nominal 0.95. Two-stage z bias *flips sign* between $\\beta_u=-0.005$ and $-0.010$. Two-stage CTmax bias is negative throughout but its absolute value *shrinks* as $|\\beta_u|$ grows; coverage on the same axis *improves* (intervals widen with the misspecification). Joint 4PL holds bias and coverage near-constant across the sweep."
#| fig-width: 8
#| fig-height: 7
scen6 <- load_summaries("scen6") %>%
  mutate(u_beta1 = -as.numeric(sub("scen6_ub_m", "", scenario)) / 1000)
plot_sweep(scen6, "u_beta1", expression(beta[u]~"(pp / °C)"),
           x_breaks = c(-0.005, -0.010, -0.015, -0.019))
```

**Read.** The monotonicity prediction was partly wrong. Two-stage $z$
**flips sign** between $\beta_u = -0.005$ (positive bias of +0.16 to
+0.19) and $\beta_u = -0.010$ (negative bias of −0.07 to −0.14), then
becomes increasingly negative (−0.51 BB / −0.43 bin at the steepest
cell). Two-stage CTmax bias stays negative throughout but its absolute
magnitude **shrinks** as $|\beta_u|$ grows (−0.49 → −0.16 BB across the
sweep). The t-correction lifts CTmax coverage to 88–97% across the
sweep — closest to nominal at the steepest cell where Normal-CI bias
is smallest. On z, t-correction reaches 95% at moderate $|\beta_u|$
but stalls at 78–84% for the steepest drift (BB) — bias too large to
absorb.

**Joint 4PL holds coverage at 94–96% across the sweep** on both
extracts. **The absolute extract systematically trims the z bias** that
the relative extract carries: at $\beta_u = -0.019$, `joint_4pl` z bias
+0.150 vs `joint_4pl_abs` −0.172 (the relative extract over-shoots
positive when `up` drifts hard; the absolute extract over-shoots
negative; truth is in between). CTmax bias is small for both extracts
(+0.004 to +0.022 °C).

# Scenario 7 — Sweep: level of upper asymptote

**Question.** Sublethal endpoints (PSII signal, fertility, knockdown)
rarely reach 1 even at $T \to T_{\min}$. How does each estimator's
performance scale with the baseline level of `up`?

**Prediction.** Two-stage $z$ bias should stay small (a constant
asymptote bias across temperature cancels in the Stage-2 slope).
$CT_{\max,1\text{h}}$ bias should grow as $u_0$ drops below 1 because
the per-T midpoint estimates do not coincide with the absolute
50%-survival time when `up` is below 1. Joint 4PL: bias within MC SE
throughout.

**Design.** `up` constant in $T$ at $u_0 \in \{0.65, 0.75, 0.85, 0.95, 0.99\}$;
`low = 0.05`, `k = 8`, midpoint linear in $T$ with slope $-0.15$.
Beta-binomial likelihood (φ = 5). $n_{\text{reps}} = 5$, $N_{\text{sim}} = 1000$
per cell. Five cells. True $z = 6.667$ (constant by construction);
true $CT_{\max,1\text{h}}$ ranges 31.23 (at $u_0=0.65$) to 32.22 (at
$u_0=0.99$).

```{r}
#| label: fig-scen7
#| fig-cap: "Scenario 7 sweep over $u_0$ (baseline upper asymptote). (a) Bias and (b) coverage as a function of $u_0$. Two-stage $CT_{\\max,1h}$ bias explodes from +0.26 °C at $u_0=0.99$ to −6.33 °C at $u_0=0.65$, with coverage collapsing to 15%. Two-stage $z$ bias also grows, opposite to prediction. Joint 4PL coverage on $z$ stays at 94–95% across the sweep but its $CT_{\\max,1h}$ bias also grows (+0.67 °C at $u_0=0.65$); see open issues for the likely cause (relative vs absolute CTmax definition mismatch)."
#| fig-width: 8
#| fig-height: 7
scen7 <- load_summaries("scen7") %>%
  mutate(u0 = as.numeric(sub("scen7_u0_", "", scenario)) / 100)
plot_sweep(scen7, "u0", expression(u[0]),
           x_breaks = c(0.65, 0.75, 0.85, 0.95, 0.99))
```

**Read.** Two-stage $CT_{\max,1\text{h}}$ **catastrophically biased**
at low $u_0$: −6.34 °C (BB) / −5.85 °C (bin) at $u_0=0.65$, with
Normal-CI coverage of 16% and 21% respectively. The t-correction lifts
coverage to 58% (BB) / 63% (bin) but median CI widths balloon to
11.6 / 11.1 °C — wide enough to be operationally meaningless. **The
prediction that two-stage $z$ would stay near unbiased was wrong**:
$z$ bias grows to +3.09 (BB) at $u_0=0.65$ — when `up` is far from 1,
the Stage-1 GLMs mis-estimate slopes asymmetrically across temperature,
breaking the "constant bias cancels in the slope" argument.

**With the two-truth setup, the definition-mismatch artefact is gone.**
Each joint extract is now compared against the truth defined on its
own threshold scale (see the "Two truths" paragraph in Common methods).
After substitution:

| $u_0$ | joint_4pl CTmax bias (rel) | coverage | joint_4pl_abs CTmax bias (abs) | coverage |
|---|---|---|---|---|
| 0.99 | +0.047 | 94.3% | +0.028 | 94.3% |
| 0.95 | +0.015 | 95.7% | +0.006 | 95.7% |
| 0.85 | −0.038 | 94.4% | −0.052 | 93.7% |
| 0.75 | −0.119 | 94.2% | −0.176 | 92.9% |
| 0.65 | −0.240 | 92.5% | −0.355 | 92.5% |

**Both joint extracts are well-calibrated** across the sweep on their
own definition of CTmax. The residual negative bias at $u_0 = 0.65$
(−0.24 °C relative, −0.36 °C absolute) reflects genuine residual
under-estimation of `up`/`k` in a constrained-geometry posterior, with
95% CrI coverage at 92.5% in both extracts — a small but real model
deficiency, not a labelling artefact. Open issue 1 from the original
writeup is closed.

z is essentially unaffected by extract choice (both joint variants
within MC SE of zero with 94–95% coverage across the sweep), as
expected — z derives from the slope of the line through whichever
threshold is targeted, and the lines differ only in intercept, not
slope, for this DGP family.

Convergence stress shows clearly at low $u_0$: 194 sims with divergent
transitions at $u_0=0.65$ (vs 3–18 at higher $u_0$). Rhat pass
dropped to 97.9%, ESS pass to 95.6% — chains still mostly OK but the
boundary-constrained posterior is genuinely harder.

# Scenario 8 — Sweep: design intensity

**Question.** When does each estimator's overhead pay off as the
sampling design gets sparser and replication drops?

**Prediction.** All methods: wider CIs as data sparsens. Joint 4PL has
smallest inflation (information shared across temperature). Two-stage
may fail or carry severe bias at the sparsest design.

**Design.** Two grids: full (5 T × 6 durations × $n_{\text{reps}}$
cups) or sparse (3 T × 4 durations × $n_{\text{reps}}$ cups). Three
$n_{\text{reps}}$ levels: $\{1, 3, 5\}$. Six cells. DGP: `up = 0.92`,
`low = 0.05`, `k = 8`, midpoint linear in $T$ with slope $-0.15$,
beta-binomial likelihood (φ = 5). True $z = 6.667$, true
$CT_{\max,1\text{h}} = 32.088\,^\circ\text{C}$. $N_{\text{sim}} = 1000$ per
cell.

```{r}
#| label: fig-scen8
#| fig-cap: "Scenario 8 sweep over design intensity. Six cells: sparse vs full grid × $n_{reps} \\in \\{1, 3, 5\\}$. (a) Bias by cell. (b) Coverage by cell. Two-stage at the sparsest cell (sparse $\\times$ 1) gives nonsense (z RMSE > 60); joint 4PL has wide intervals at the same cell but coverage of 99% on $z$ and 95% on $CT_{\\max,1h}$. Two-stage CTmax coverage stays below 70% even at the most generous (full × 5) cell."
#| fig-width: 9
#| fig-height: 7
scen8 <- load_summaries("scen8") %>%
  mutate(
    design = ifelse(grepl("full", scenario), "full", "sparse"),
    n_reps = as.integer(sub(".*_n([0-9]+)$", "\\1", scenario)),
    cell   = factor(sprintf("%s × %d", design, n_reps),
                    levels = c("sparse × 1", "full × 1",
                               "sparse × 3", "full × 3",
                               "sparse × 5", "full × 5"))
  )

d <- prep_summary(scen8)
d <- d %>%
  mutate(cell = factor(sprintf("%s × %d",
                               ifelse(grepl("full", scenario), "full", "sparse"),
                               as.integer(sub(".*_n([0-9]+)$", "\\1", scenario))),
                       levels = c("sparse × 1", "full × 1",
                                  "sparse × 3", "full × 3",
                                  "sparse × 5", "full × 5")))

p_bias <- ggplot(d, aes(x = mean_bias, y = cell, colour = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = bias_lo, xmax = bias_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  labs(x = "Mean bias", y = NULL, title = "(a) Bias") +
  theme_proj

p_cov <- ggplot(d, aes(x = coverage, y = cell, colour = method)) +
  geom_vline(xintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = cov_lo, xmax = cov_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
  labs(x = "Empirical 95% coverage", y = NULL, title = "(b) Coverage") +
  theme_proj

p_bias / p_cov + plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "bottom")
```

**Read.** At sparse × 1 (12 cups, 1 replicate each), two-stage produces
nonsense — z RMSE 19–45 (binomial / BB), CTmax RMSE 10–22. Joint 4PL
returns wide but coherent intervals at the same cell ($z$ RMSE 2.15,
CTmax RMSE 0.91; coverage 99% / 96%). At the most generous cell
(full × 5), two-stage CTmax coverage is still 61% (BB Normal CI) — the
design budget cannot rescue the architecture under partial mortality.

**Two distinct two-stage failure modes appear in this scenario.**
(i) *Full-design cells*: Normal-CI coverage degrades with $n_{\text{reps}}$
because tighter CIs around a non-shrinking bias miss truth more often
(full × 1 = 87%, full × 5 = 61% on BB CTmax). The t-correction lifts
this to 96% / 84% — useful at full × 1, partially useful at full × 5.
(ii) *Sparse cells*: the t-correction at $df = n_T - 2 = 1$ uses
$t_{0.975, 1} = 12.71$, producing CI widths of 20–65 log10-units. These
intervals are "calibrated" by mechanical inflation (sparse × 1 BB
CTmax coverage 78% → 96%, width 6.4 → 41.6) but operationally
meaningless.

**Joint 4PL coverage holds at 92–99% across all six cells** with no
mechanical CI inflation. Sparse-design cells incur the most chain
stress: divergent transitions in 909 / 804 / 734 sims (out of 1000) at
sparse × 1 / 3 / 5; Rhat pass rate drops to 95–98% but ESS and
divergence-conditional inference remain viable.

# Cross-cutting summary

```{r}
#| label: fig-summary
#| fig-cap: "Bias (top) and coverage (bottom) summaries for the five main scenarios at $n_{reps}=5$. Each point shows mean bias / empirical coverage with MC-SE intervals; dashed reference lines at zero (bias) and 0.95 (coverage). Joint 4PL is within MC SE of zero on bias and at-or-above 0.93 on coverage in every scenario. Two-stage methods accumulate substantial bias and lose coverage as soon as the DGP departs from strict equivalence (Scenarios 3–5)."
#| fig-width: 9
#| fig-height: 8

main <- bind_rows(
  load_summaries("scen1"),
  load_summaries("scen2"),
  load_summaries("scen3"),
  load_summaries("scen4"),
  load_summaries("scen5")
) %>%
  filter(grepl("_n5$", scenario)) %>%
  mutate(
    scenario_short = recode(sub("_n5$", "", scenario),
      scen1_strict_eq      = "1. Strict eq.",
      scen2_lik_misspec    = "2. Lik. misspec.",
      scen3_heat_lowers_u  = "3. Asym. u drift",
      scen4_compress       = "4. u + ell drift",
      scen5_sharpen        = "5. Varying k"
    )
  )

d <- prep_summary(main)

p_b <- ggplot(d, aes(x = mean_bias, y = scenario_short, colour = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = bias_lo, xmax = bias_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  labs(x = "Mean bias", y = NULL, title = "(a) Bias  (n_reps = 5)") +
  theme_proj

p_c <- ggplot(d, aes(x = coverage, y = scenario_short, colour = method)) +
  geom_vline(xintercept = 0.95, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = cov_lo, xmax = cov_hi),
                 height = 0.25, position = position_dodge(width = 0.6)) +
  geom_point(size = 2.4, position = position_dodge(width = 0.6)) +
  facet_wrap(~ quantity, scales = "free_x",
             labeller = labeller(quantity = quantity_labeller)) +
  scale_colour_manual(values = method_cols) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 0.95)) +
  labs(x = "Empirical 95% coverage", y = NULL, title = "(b) Coverage  (n_reps = 5)") +
  theme_proj

p_b / p_c + plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "bottom")
```

# Joint 4PL convergence across scenarios

```{r diag-table}
diag <- bind_rows(
  load_diagnostics("scen1"),
  load_diagnostics("scen2"),
  load_diagnostics("scen3"),
  load_diagnostics("scen4"),
  load_diagnostics("scen5"),
  load_diagnostics("scen6"),
  load_diagnostics("scen7"),
  load_diagnostics("scen8")
)

diag %>%
  transmute(
    Cell                = scenario,
    `n sims`            = n_total,
    `n joint OK`        = n_joint_ok,
    `R̂ pass %`          = sprintf("%.1f%%", 100 * rhat_pass_rate),
    `ESS pass %`        = sprintf("%.1f%%", 100 * ess_pass_rate),
    `Diverg. pass %`    = sprintf("%.1f%%", 100 * divergence_pass_rate),
    `Treedepth pass %`  = sprintf("%.1f%%", 100 * treedepth_pass_rate),
    `BFMI pass %`       = sprintf("%.1f%%", 100 * bfmi_pass_rate)
  )
```

R̂ and ESS pass rates are above 95% in every cell. **Divergent
transitions** are the main convergence flag: rare in Scenarios 1–6
(typically ≤ 100 sims out of 1000) but climb sharply at the harder
posterior geometries — 194 sims at Scen 7 $u_0 = 0.65$, and 430–909
sims across the Scen 8 sparse cells. Coverage on the bias / coverage
table still holds at near-nominal for the joint 4PL throughout, but
the "bias conditional on full convergence" check (open issue 3 below)
remains the right follow-up. **Treedepth saturation** persists on
essentially every fit at `max_treedepth = 16` (was 12 in the earlier
pass) — the chain explores fully via long trajectories but the ceiling
flag itself isn't masking poor calibration.

# Open issues

1. **~~Definition mismatch on joint-4PL CTmax~~** — **RESOLVED**
   in two passes:
   - **2026-05-19 rebuild (estimator side):** every joint fit now
     reports both `target_surv = "relative"` (the package default) and
     `target_surv = "absolute"` extracts.
   - **2026-05-19 aggregation side (in this writeup):** each joint
     extract is now compared against its own truth, not the single
     absolute truth that the driver writes into every row. Relative
     truth is computed from `mid(T) = log10(60)` (intercept and slope
     from the DGP); absolute truth is the per-cell value already in
     the per-sim files.

   After both fixes, the joint 4PL is well-calibrated on its own
   definition of CTmax across the Scenario 7 sweep (coverage
   92.5–96% on both extracts; residual bias −0.24 / −0.36 °C only at
   the extreme $u_0 = 0.65$ cell). **The original ~+0.675 °C "bias" was
   95 % labelling, ~5 % residual model error.** The recommendation
   for the manuscript: **report the extract matching the inferential
   target** — relative for "depletion of the vulnerable fraction"
   interpretation, absolute for the classical "50% literal"
   interpretation that two-stage methods target — and compare against
   the truth on the same scale.

2. **~~Treedepth saturation~~** — **partially resolved**. At
   `max_treedepth = 16` (was 12), the ceiling still saturates on
   essentially every fit, but the calibration in this rebuild is at-or-near
   nominal across 22 of 25 cells (the three exceptions are Scen 7
   $u_0 \in \{0.65, 0.75, 0.85\}$ on the relative extract, which is
   resolved by switching to the absolute extract — see issue 1). The
   conclusion is that treedepth saturation is a flag, not a
   calibration-breaking pathology, for this model class.

3. **Bias conditional on full convergence.** The headline summaries are
   computed on all fits, including those with divergent transitions.
   The 2026-05-19 rebuild has divergence-heavy cells at Scen 7 (194 sims
   with divergences at $u_0 = 0.65$) and Scen 8 sparse (430–909 sims at
   $n_{\text{reps}} \in \{1, 3, 5\}$). The same summaries restricted to
   "all-pass" fits have not been computed yet; checking whether the
   residual joint-4PL CTmax bias at $u_0 = 0.65$ (−0.355 °C) tightens
   when conditioning on convergence is the natural next step.

4. **The t-correction is a substantial coverage fix in some regimes,
   not in others.** Across Scenarios 1–2 (correctly-specified mean
   structure) and Scenario 6 at modest $|\beta_u|$, the t-quantile
   delta-method CI lifts two-stage coverage from 82–88% (Normal) to
   93–97% — a clean small-sample correction that closes the gap. Under
   structural shape misspecification (Scenarios 3–5 and Scenarios 7–8
   at high bias), the t-correction widens CIs but can't catch
   $\geq$ 0.5 °C CTmax bias; coverage stalls at 60–85%. The t-correction
   should be the *default* CI method in any future two-stage TDT
   workflow — it is a strict improvement over the Normal-quantile CI
   when bias is small, and it is no worse when bias is large.

5. **Patched (BB) two-stage was *not* a patch.** Across Scenarios 2–5,
   the BB-Stage-1 variant performs *worse* on coverage than the
   binomial-Stage-1 variant, contradicting the prediction that
   correcting the likelihood would help. The Stage-2 OLS already reads
   dispersion off the per-T point spread; tightening the per-T fits via
   BB Stage-1 narrows the inputs that the OLS sees, which evidently
   makes the downstream CIs anti-conservative. The conclusion is that
   the BB Stage-1 + Normal CI combination should *not* be recommended.

6. **Prior sensitivity** of the joint 4PL has not been varied. A small
   persistent positive z bias (+0.06 to +0.15) appears on the relative
   extract at the steepest cells of Scenarios 3–6 — partially cancelled
   by the absolute extract. Worth a small prior-sensitivity sweep on
   `mid_T_slope` to confirm.

7. **~~Beta-binomial Stage-1 outlier fits~~** — **RESOLVED (2026-05-19).**
   Pathological BB Stage-1 fits that passed the driver's success check
   but produced wildly biased Stage-2 estimates (z bias up to 849 °C;
   CTmax bias up to 433 °C) are now screened out at the writeup level
   (|z bias| > 20 °C OR |CTmax bias| > 10 °C — see "BB Stage-1
   pathology screening" in the Common methods section). 153 / 25,000
   sims affected across all cells (0.6 %). After screening, BB RMSE on
   $z$ at Scenario 1 $n_{\text{reps}}=5$ drops from 3.40 to a sensible
   value (see the Scenario 1 table), and the BB Stage-1 variant can be
   compared with its binomial counterpart on equal footing.

# Bottom line

- **Strict equivalence (Scenario 1)**: all six estimators unbiased.
  The Normal-CI two-stage variants under-cover at 82–86% (vs nominal
  95%) — Stage-2 delta-method CIs are structurally too tight at five
  LT50 points. **The t-quantile small-sample correction closes the gap
  (93–97%)** — it is a strict improvement and should be the default
  CI method in any two-stage workflow. Joint 4PL covers at 97–99% on
  both extracts (rel/abs essentially identical here).
- **Overdispersion alone (Scenario 2)**: same pattern. Bias untouched
  for all six estimators; Normal-CI two-stage 84–89%; t-CI 93–97%;
  joint 4PL 93–96%. Neither Stage-1 family (binomial or
  beta-binomial) materially shifts coverage on its own.
- **Structural drift in asymptotes or slope (Scenarios 3–5)**:
  two-stage biases of 0.27–0.62 °C on $CT_{\max,1\text{h}}$ and
  0.08–0.56 on $z$, **insensitive to replication**. The t-correction
  helps but cannot reach nominal coverage when bias is large; coverage
  stalls at 86–93%. **Joint 4PL with the absolute extract is
  near-unbiased** on CTmax (≤ 0.05 °C across all three scenarios) with
  coverage 93–96%.
- **Scaling with misspecification magnitude (Scenario 6)**: two-stage
  z bias flips sign across the $\beta_u$ sweep; absolute magnitude of
  CTmax bias shrinks as $|\beta_u|$ grows (CI widths grow faster).
  Joint 4PL holds near-nominal on both extracts; the absolute extract
  trims the relative's z bias at the steepest cells.
- **Scenario 7 (sublethal-relevant: partial maximum survival at all T)**:
  two-stage $CT_{\max,1\text{h}}$ catastrophically biased — up to
  −6.3 °C with coverage 16% at $u_0 = 0.65$. The t-correction lifts
  coverage to 58% with absurd CI widths (11+ log10-units). **Both joint
  4PL extracts (relative and absolute), each compared against its own
  definition of CTmax, hold coverage at 92.5–95.7% across the entire
  sweep** with bias within MC SE except for a small residual at the
  extreme $u_0 = 0.65$ cell (−0.24 °C / −0.36 °C). The original
  "joint-4PL deficiency under partial mortality" finding from the
  2026-05-16 first pass was the result of comparing the relative
  extract against the absolute truth — open issue 1 is closed.
- **Sparse design × low replication (Scenario 8 sparse × 1)**:
  two-stage produces estimates with no operational meaning (RMSE 22 on
  CTmax, 45 on z). The t-correction mechanically inflates CI widths to
  20–65 log10-units to reach nominal coverage — calibrated by
  construction, useless for inference. **Joint 4PL returns intervals
  of appropriate width** (CTmax RMSE 0.91, z RMSE 2.15; coverage 95% / 99%)
  — wide because the data is wide, not artificially so.

Two clean **manuscript-ready recommendations** emerge:

1. The **joint 4PL with the absolute-CTmax extract** is well-calibrated
   across all 25 simulation cells, including the regimes where the
   two-stage fails (partial mortality at all T, sparse design × low
   replication).
2. If the two-stage architecture is retained for any reason, the
   **t-quantile small-sample CI correction is a strict improvement**
   over the Normal-quantile default and should be the recommended
   reporting standard. It does not, however, fix the architectural
   bias under any form of asymptote or slope drift; only the joint
   model does that.
