---
title: "bayesian_TDT_plants"
author: "Pieter Arnold"
format: html
---


```{r}
pacman::p_load(here, dplyr, tidyr, tibble, purrr, ggplot2, brms, posterior, tidybayes, tinytable, patchwork, pander, readxl)
library(bayesTLS)
```

# **Case Study 3: Leaf PSII data** {#sec-leaves}

We can also apply the full workflow described above — `standardize_data()` → `fit_4pl()` → `extract_tdt()` — to photosystem II functional impairment TDT data from snow gum leaves (*Eucalyptus pauciflora* var. *pauciflora*). The data are chlorophyll fluorescence based measurements of the maximum potential quantum efficiency of photosystem II (PSII), known as *F*[V]/*F*[M], from excised leaf segments of 11 individual glasshouse-grown trees. In this experiment, *F*[V]/*F*[M] was measured prior to heat exposure and then 24 h after *F*[V]/*F*[M], such that the response variable was a proportional value of *F*[V]/*F*[M] bounded between 0 and 1.

The TDT trials spanned 28-48 °C and 5 min to 2 hours of exposure, with each replicate plant originating from one of two glasshouse rooms and measured over two days. Similar to Case Study 1, the two grouping factors `Day` and `G_Room` enter the model as random intercepts on `mid`, allowing day-to-day batch variation and glasshouse room-level variation to be partitioned out of the dose-response surface.

## Data and standardisation {#sec-leaf-data}

```{r leaf-load}
#| label: leaf-load
#| echo: true

leaf_raw <- readr::read_csv(
  here::here("data", "data_function_PSII_TDT_snowgum.csv"),
  show_col_types = FALSE
)

leaf_raw$fvfm_prop <- leaf_raw$final_fvfm / leaf_raw$initial_fvfm # Response variable 
str(leaf_raw)
assay_temp <- as.numeric(leaf_raw$Temp)
temp_mean <- mean(assay_temp, na.rm = TRUE)

leaf_std <- leaf_raw %>% 
  mutate(
    # Constrain to (0,1) for beta distribution
    fvfm_beta = case_when(
      fvfm_prop >= 0.999 ~ 0.999,
      fvfm_prop <= 0.001 ~ 0.001,
      TRUE ~ fvfm_prop
    ),
    assay_temp = Temp,
    temp = Temp,
    temp_c = assay_temp - temp_mean,
    duration = Time,
    logd = log10(duration),
    # Convert random effects to factors
    Day = factor(Day),
    G_Room = factor(G_Room)
  )

str(leaf_std)

```

```{r leaf-summary-table}
#| label: tbl-leaf-summary
#| tbl-cap: "Summary of the leaf PSII TDT dataset after standardisation. Temperature centre is the grand mean used to mean-centre the temperature covariate."

leaf_summary <- tibble::tibble(
  `n individuals`              = length(unique(leaf_std$Unique_ID)),
  `Temperature range (°C)`     = paste(range(leaf_std$assay_temp), collapse = "–"),
  `Duration range (min)`       = paste(round(range(leaf_std$duration), 2), collapse = "–"),
  `Temperature centre (°C)`    = attr(leaf_std, "tdt_meta")$temp_mean,
  `n Day`                      = length(unique(leaf_std$Day)),
  `n glasshouse room`          = length(unique(leaf_std$G_Room))
)
tinytable::tt(leaf_summary, digits = 2, escape = TRUE)
```

## Fitting the joint Bayesian 4PL {#sec-leaf-fit}

`fit_4pl()` is the package's high-level fitting wrapper. It takes the standardised data and configures the joint 4PL for proportional response data: a beta distribution, the disjoint-bounds reparameterisation that keeps the asymptotes identifiable, and a `temp_c` slope on each 4PL sub-parameter. We add random intercepts on `mid` for `Day` and `G_Room` to absorb the day-to-day and room-to-room variation that would otherwise contaminate the dose-response surface.

```{r leaf-fit}
#| label: leaf-fit
#| echo: true
#| cache: false


formula_beta <- make_4pl_formula(
  random_effects = c("Day", "G_Room"),
  lower = 0,
  upper = 1,
  family = brms::Beta(link = "logit"),
  response_var = "fvfm_beta"
)

base_priors <- make_4pl_priors(
  data = leaf_std,
  random_effects = c("Day", "G_Room"),
  prior_phi = NULL
)

wf_leaf_fit <- brms::brm(
  formula = formula_beta,
  data = leaf_std,
  prior = base_priors,
  chains  = 4, iter = 8000, warmup = 4000, cores = 4, seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12),
  file    = file.path(models_dir, "fit_leaf_function_4pl")
)

wf_leaf <- structure(
  list(
    fit = wf_leaf_fit,
    data = leaf_std,
    formula = formula_beta,
    prior = priors_beta,
    meta = list(
      temp_mean = temp_mean,
      duration_unit = "minutes",
      random_effects = c("Day", "G_Room"),
      lower = 0,
      upper = 1,
      bounds = list(low_min = 0.001, low_w = 0.498, up_min = 0.501, up_w = 0.498)
    )
  ),
  class = "bayes_tls"
)

# Workflow header: data shape, T_bar, asymptote bounds, random effects, draws.
print(wf_leaf)
```

### Sampling diagnostics

```{r leaf-diagnostics}
#| label: tbl-leaf-diagnostics
#| tbl-cap: "Sampling diagnostics for the leaf 4PL fit. Healthy targets: Rhat < 1.01, ESS > 4000, 3 divergences, 1 treedepth flag."
#| cache: false

tinytable::tt(diagnose_tdt_fit(wf_leaf), digits = 3, escape = TRUE)

```

The diagnostics table is a numerical pass/fail summary. The canonical visual check is the chain trace ("fuzzy caterpillar"): healthy chains overlap and explore the same posterior support; chains that drift, diverge, or hang in different regions signal a problem with the fit.

```{r leaf-trace-plot}
#| label: fig-leaf-trace
#| fig-cap: "Posterior chain traces for the leaf 4PL fixed-effect parameters and the beta dispersion $\\phi$. Each panel overlays four chains; well-mixed chains overlap throughout."
#| fig-width: 9
#| fig-height: 6
#| cache: false

brms::mcmc_plot(
  wf_leaf$fit,
  type     = "trace",
  variable = c("b_lowraw_Intercept", "b_upraw_Intercept",
               "b_logk_Intercept",   "b_mid_Intercept",
               "b_mid_temp_c",       "Intercept_phi")
) +
  ggplot2::theme_classic(base_size = 11) +
  ggplot2::theme(legend.position = "bottom")

```

### Posterior parameters on the natural scale

```{r leaf-parameter-table}
#| label: tbl-leaf-parameters
#| tbl-cap: "Posterior medians and 95% credible intervals for the leaf 4PL fit, transformed back to the natural scale. `low` and `up` are the bottom and top asymptotes of the survival curve at the grand-mean temperature; `k` is the slope on $\\log_{10}$ time; `mid` intercept is $\\log_{10}\\text{LT50}$ in hours at the grand-mean temperature; `mid temp_c slope` is $\\beta_1$, from which $z = -1/\\beta_1$."

tinytable::tt(tdt_parameter_table(wf_leaf), digits = 3, escape = TRUE)
```

## "Survival" curves with observed overlay {#sec-leaf-survival-curves}

Here we use the same survival-type functions as before, but *survival* is instead *functional impairment* of PSII. `predict_survival_curves()` evaluates the fitted 4PL on a grid of assay temperatures and exposure durations and returns posterior functional impairment probabilities. Overlaying the per-replicate observed proportions gives a visual check that the fitted curves track the data at every temperature.

```{r leaf-survival-curves}
#| label: fig-leaf-survival-curves
#| fig-cap: "Posterior functional impairment curves for snow gum leaf PSII at each assayed temperature, with 95% credible bands. Coloured points are observed per-replicate functional impairment proportions."
#| fig-width: 8
#| fig-height: 5

psc_leaf <- predict_survival_curves(
  wf_leaf,
  temps     = sort(unique(leaf_std$temp)),
  ndraws    = 500
)

# Add survival as a variable for plotting, where "survival" = fvfm_prop
leaf_std$survival <- leaf_std$fvfm_prop

plot_survival_curves(psc_leaf, observed = leaf_std) +
  ggplot2::labs(x = "Exposure duration (minutes)") +
  ggplot2::coord_cartesian(xlim = c(0, 120))

```

## TDT landscape {#sec-leaf-landscape}

`derive_tdt_landscape()` evaluates the fitted 4PL on a dense (temperature, duration) grid and `plot_tdt_landscape()` renders it as a heatmap with overlaid contours at 10, 50, and 90% functional impairment This is the two-dimensional view of the dose-response surface that the TDT line summarises with one number per temperature.

```{r leaf-landscape}
#| label: fig-leaf-landscape
#| fig-cap: "TDT landscape for snow gum PSII function: posterior median predicted functional impairment across a 120 × 120 grid of assay temperatures × durations. White contours mark 10%, 50%, and 90% functional impairment isolines. Overlaid points are raw observed proportions."
#| fig-width: 8
#| fig-height: 6
#| cache: false

leaf_dur_grid <- seq(min(leaf_std$duration), max(leaf_std$duration),
                        length.out = 120)
lsp_leaf <- derive_tdt_landscape(wf_leaf,
                                   duration_grid = leaf_dur_grid,
                                   ndraws        = 500)
plot_tdt_landscape(lsp_leaf, observed = leaf_std)
```

## TDT line: time to the threshold vs temperature {#sec-leaf-tdt-line}

The one-dimensional **TDT line** summarises the dose-response surface as the predicted time to the survival threshold at each temperature. `derive_tdt_curve()` builds it from the joint fit, and `plot_tdt_curve()` shows it on linear and $\log_{10}$ time axes side-by-side: the linear panel reads as "minutes of exposure half the PSII tolerate" for ecological interpretation; the $\log_{10}$ panel linearises the relationship into the canonical TDT straight line, whose slope is $-1/z$. The default threshold is $\text{mid}$ (the midpoint between the fitted asymptotes); pass `target_surv = "absolute"` for the classical absolute-$t_{50}$ definition.

```{r leaf-tdt-line}
#| label: leaf-tdt-line
#| echo: true

ltx_leaf <- derive_tdt_curve(
  wf_leaf,
  temp_grid        = seq(min(leaf_std$temp) - 1.5,
                         max(leaf_std$temp) + 1.5, by = 0.05),
  ndraws           = 500,
  time_multiplier  = 1,
  output_time_unit = "minutes"
)
```

```{r plot-leaf-tdt-line}
#| label: fig-leaf-tdt-line
#| fig-cap: "Snow gum leaf PSII function posterior TDT line: predicted time to threshold vs assay temperature, with 95 % credible band. Linear-time panel on the left, $\\log_{10}$-time panel on the right (slope is $-1/z$). Both panels share the same posterior."
#| fig-width: 9
#| fig-height: 4

plot_tdt_curve(ltx_leaf, panels = "both", colour = "#b2182b")
```

## Classical TDT quantities: $z$, $CT_{max_{1hr}}$ {#sec-leaf-extract}

`extract_tdt()` derives the classical thermal-tolerance summaries from a fitted 4PL workflow. It always returns $z$ (the slope of $\log_{10}\text{LT}$ on temperature, derived by per-draw OLS on the TDT curve) and $CT_{max_{1hr}}$ (the temperature at which the threshold is reached after 1 hour of exposure, obtained by per-draw 4PL inversion).

```{r leaf-extract}
#| label: leaf-extract
#| echo: true

et_leaf <- extract_tdt(
  wf_leaf,
  t_ref         = 60,
  TC_rate_range = c(0.1, 1),
  ndraws        = 500,
  lethal        = FALSE
)
```

```{r leaf-extract-table}
#| label: tbl-leaf-extract
#| tbl-cap: "Classical TDT summaries for brown leaf, with full posterior uncertainty. $z$ comes from a per-draw log-linear regression of $\\log_{10}\\text{LT50}$ on temperature. $CT_{max_{1hr}}$ is the temperature at 50% survival after 1-hour exposure (4PL inversion per draw)."

leaf_extract <- tibble::tribble(
  ~Quantity,                        ~Median, ~Lower, ~Upper,
  "z (°C per order of magnitude)",      et_leaf$z$summary$z_median,
                                    et_leaf$z$summary$z_lower,
                                    et_leaf$z$summary$z_upper,
  "CTmax_1hr (°C)",                 et_leaf$CTmax$summary$temp_median,
                                    et_leaf$CTmax$summary$temp_lower,
                                    et_leaf$CTmax$summary$temp_upper
)
tinytable::tt(leaf_extract, digits = 2, escape = TRUE)
```

```{r leaf-ctmax-plot}
#| label: fig-leaf-ctmax
#| fig-cap: "Posterior densities of $z$ (left) and $CT_{max_{1hr}}$ (right) for snow gum leaf PSII function. The horizontal bar below each density is the 95% credible interval; the point marks the posterior median. The ~5 °C gap between $CT_{max_{1hr}}$ and $T_{crit}$ is the window between the onset of detectable damage accumulation and the LT50 temperature at 1 h."
#| fig-width: 11
#| fig-height: 3.5

z_draws_leaf   <- get_z_draws(et_leaf) |>
  dplyr::transmute(temp = z)
z_q <- stats::quantile(z_draws_leaf$temp,
                       c(0.025, 0.5, 0.975), names = FALSE, na.rm = TRUE)
z_post_leaf <- list(
  draws   = z_draws_leaf,
  summary = tibble::tibble(temp_lower  = z_q[1],
                           temp_median = z_q[2],
                           temp_upper  = z_q[3])
)

patchwork::wrap_plots(
  plot_temperature_density(z_post_leaf,
                            x_label = expression(italic(z)~("°C per decade of time"))),
  plot_temperature_density(et_leaf$CTmax,
                            x_label = expression(CT[max[1*hr]]~("°C"))),
  ncol = 2
)
```


## Heat injury under reference traces {#sec-leaf-hi}

`predict_heat_injury()` projects the fitted dose-response surface through a temperature trace and returns cumulative heat injury (HI) and the resulting survival fraction at each time step. The damage threshold $T_c$ below which injury does not accumulate is hand-picked in this case, rather than model derived due to the sublethal nature of functional impairment. 

We illustrate the function on four reference traces built by `make_temperature_scenarios()`: a flat baseline, a single one-hour spike, three one-hour spikes, and a four-day diurnal cycle that mimics a typical summer heatwave. The single-spike trace is calibrated as an analytical sanity check — its spike temperature is set to the posterior median of $CT_{max_{1hr}}$, so by definition the trace delivers approximately one $LT_{50}$ dose by the end of the spike hour. These values are plausible leaf temperatures on sunny, still conditions.

```{r leaf-hi}
#| label: leaf-hi
#| echo: true

leaf_T_c       <- 32.5
leaf_CTmax_med <- et_leaf$CTmax$summary$temp_median

leaf_scens <- make_temperature_scenarios(
  baseline   = 30,
  spike_temp = leaf_CTmax_med,
  n_hours    = 96,
  spike_times_single = 24,
  spike_times_multi  = c(24, 48, 72),
  diurnal_n_days     = 4,
  diurnal_day_peaks  = c(38.0, 41.0, 44.0, 43.5),
  diurnal_night_temp = c(20.0, 15.5, 17.5, 19.0),
  diurnal_peak_fwhm  = 5,
  diurnal_noise_sd   = 0.5,
  diurnal_seed       = 1L
)

leaf_hi <- purrr::map(leaf_scens, function(sc) {
  predict_heat_injury(trace = sc, workflow = wf_leaf,
                      T_c = leaf_T_c, ndraws = 500)
})
```

Setting `repair = TRUE` adds a temperature-dependent Sharpe-Schoolfield repair kernel so injury accumulated at hot temperatures can be partially cleared during cooler intervals. Here we re-run only the two scenarios that accrue meaningful HI without repair — the multi-spike and the diurnal traces.

```{r leaf-hi-repair}
#| label: leaf-hi-repair
#| echo: true

leaf_repair_pars <- list(
  TA = 14065, TAL = 50000, TAH = 120000,
  TL = 10.5 + 273.15, TH = 32.5 + 273.15, TREF = 30 + 273.15,
  r_ref = 0.15
)

leaf_hi_multi_rep <- predict_heat_injury(
  leaf_scens$multi_spike, wf_leaf, T_c = leaf_T_c, ndraws = 500,
  repair = TRUE, repair_pars = leaf_repair_pars
)
leaf_hi_diurnal_rep <- predict_heat_injury(
  leaf_scens$diurnal, wf_leaf, T_c = leaf_T_c, ndraws = 500,
  repair = TRUE, repair_pars = leaf_repair_pars
)
```

```{r leaf-hi-traces}
#| label: fig-leaf-hi-traces
#| fig-cap: "The four reference temperature traces used for the leaf heat-injury prediction. Baseline  °C is the natural holding temperature; the 1-hour spikes are set to the posterior median $CT_{max_{1hr}}$ so each spike delivers approximately one LT50 dose. The diurnal scenario is a clustered 4-day heatwave: a cool day below $T_{crit}$ followed by three progressively hotter days. The dashed line marks the damage threshold $T_c = T_{crit}$."
#| fig-width: 7
#| fig-height: 9

plot_temperature_scenarios(leaf_scens, T_c = leaf_T_c)
```

```{r leaf-hi-single-noreapair}
#| label: fig-leaf-hi-single-norep
#| fig-cap: "Cumulative heat injury (top) and predicted survival (bottom) under the **single-spike** trace, **no repair**. Because the spike sits at the posterior median $CT_{max_{1hr}}$, ~100% LT50-dose accrues by the end of the spike hour, producing ~50% mortality."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(leaf_hi$single_spike)
```

```{r leaf-hi-multi-noreapair}
#| label: fig-leaf-hi-multi-norep
#| fig-cap: "**Multi-spike** trace, **no repair**. Three 1-hour spikes at $CT_{max_{1hr}}$ deliver ~3 × LT50-dose, driving cumulative HI well past 100% and survival to near zero."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(leaf_hi$multi_spike)
```

```{r leaf-hi-multi-rep}
#| label: fig-leaf-hi-multi-rep
#| fig-cap: "Same multi-spike trace, **with Sharpe-Schoolfield repair active**. Repair clears most of the dose between spikes, stopping the runaway accumulation seen without repair."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(leaf_hi_multi_rep)
```

The diurnal scenario (@fig-leaf-hi-diurnal-norep, @fig-leaf-hi-diurnal-rep) is closer to a real coastal record. Day 1 stays below $T_{crit}$ and accrues no injury; days 2–4 each contribute an injury bout that scales with the daily peak. Without repair, cumulative HI is well past 100 % by the end of day 4 and predicted survival has dropped to ~20 %. With repair on, off-peak hours and the cool first day act as recovery phases; the final trajectory is set by the net balance between daily accrual at sub-$CT_{max_{1hr}}$ peaks and overnight/cool-day repair.

```{r leaf-hi-diurnal-noreapair}
#| label: fig-leaf-hi-diurnal-norep
#| fig-cap: "**4-day diurnal heatwave** trace, **no repair**. The cool first day flat-lines below $T_{crit}$; days 2–4 each accrue an injury bout scaling with the daily peak."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(leaf_hi$diurnal)
```

```{r leaf-hi-diurnal-rep}
#| label: fig-leaf-hi-diurnal-rep
#| fig-cap: "Same 4-day heatwave trace, **with Sharpe-Schoolfield repair active**. The cool first day and overnight troughs now act as recovery phases."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(leaf_hi_diurnal_rep)
```

```{r leaf-hi-summary}
#| label: tbl-leaf-hi-summary
#| tbl-cap: "Final cumulative HI and predicted survival under each scenario for brown leaf, posterior medians with 95% credible intervals."

final_row <- function(hi) {
  s <- tail(hi$summary, 1)
  tibble::tibble(
    `HI median (%)` = round(s$hi_median, 1),
    `HI lower`      = round(s$hi_lower,  1),
    `HI upper`      = round(s$hi_upper,  1),
    `Survival median`  = round(s$surv_median, 3),
    `Survival lower`   = round(s$surv_lower,  3),
    `Survival upper`   = round(s$surv_upper,  3)
  )
}

leaf_hi_table <- dplyr::bind_rows(
  final_row(leaf_hi$flat)         |> dplyr::mutate(Scenario = "Flat",
                                                     Repair = "None"),
  final_row(leaf_hi$single_spike) |> dplyr::mutate(Scenario = "Single spike",
                                                     Repair = "None"),
  final_row(leaf_hi$multi_spike)  |> dplyr::mutate(Scenario = "Multi spike",
                                                     Repair = "None"),
  final_row(leaf_hi_multi_rep)    |> dplyr::mutate(Scenario = "Multi spike",
                                                     Repair = "Sharpe-Schoolfield"),
  final_row(leaf_hi$diurnal)      |> dplyr::mutate(Scenario = "Diurnal (7 days)",
                                                     Repair = "None"),
  final_row(leaf_hi_diurnal_rep)  |> dplyr::mutate(Scenario = "Diurnal (7 days)",
                                                     Repair = "Sharpe-Schoolfield")
) |>
  dplyr::select(Scenario, Repair, dplyr::everything())

tinytable::tt(leaf_hi_table, escape = TRUE)
```
The snow gum leaf case study covers raw counts → fitted joint 4PL → classical TDT quantities with calibrated uncertainty. Every quantity inherits the posterior of the same single fit. 

