---
title: "Standard binomial logistic regression is a 2PL — `mid` is implicit, not free"
date: 2026-05-07
author: "Daniel W.A. Noble (clarification dialogue with Claude)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
---

## Question

A standard binomial regression (logistic model) `glm(cbind(dead, alive) ~ log10(time), family = binomial)` only fits a slope and intercept. So is that a 2PL? And if so, where does the curve midpoint `mid` live, given it's not estimated as a separate parameter?

## Short answer

**Yes — a binomial GLM with logit link and one continuous predictor is a 2PL** (2-parameter logistic). The asymptotes $\ell=0$ and $u=1$ are pinned by the link function, leaving only the steepness $k$ and midpoint $\text{mid}$ free. They are encoded in the slope and intercept of the linear predictor as

$$
k \;=\; -\beta_1, \qquad \mathrm{mid} \;=\; -\beta_0 / \beta_1.
$$ {#eq-2pl-mapping}

So `mid` is *not* a separate free parameter in the GLM — it's the ratio of intercept to negative slope, hiding in the same fit.

## Algebra: 4PL ↔ logit GLM

Start with the 4PL with asymptotes pinned at $(\ell, u) = (0, 1)$:

$$
S(\tau) \;=\; \frac{1}{1 + \exp\!\big(k\,(\log_{10}\tau - \mathrm{mid})\big)}.
$$

Apply the logit transformation:

$$
\mathrm{logit}(S) \;=\; \log\!\frac{S}{1 - S} \;=\; -k\,(\log_{10}\tau - \mathrm{mid}) \;=\; k\,\mathrm{mid} \;-\; k\,\log_{10}\tau.
$$

This is linear in $\log_{10}\tau$. Matching to a logit-binomial GLM,

$$
\mathrm{logit}(S) \;=\; \beta_0 \;+\; \beta_1 \log_{10}\tau,
$$

gives the 2PL ↔ GLM mapping:

$$
\beta_0 \;=\; k\,\mathrm{mid}, \qquad \beta_1 \;=\; -k.
$$

Inverting:

$$
k \;=\; -\beta_1, \qquad \mathrm{mid} \;=\; -\beta_0/\beta_1.
$$

The midpoint $\mathrm{mid}$ — i.e. $\log_{10}(LT_{50})$ — falls out as the *intercept-to-slope ratio*, with a sign flip. Setting $S = 0.5$ confirms this directly: $\mathrm{logit}(0.5) = 0$, so $0 = \beta_0 + \beta_1 \cdot \mathrm{mid}$, giving $\mathrm{mid} = -\beta_0/\beta_1$.

## Why GLM users don't usually think of `mid` as a parameter

A GLM is fit on the linear-predictor scale: you estimate $(\beta_0, \beta_1)$, those are what `summary(fit)` reports, and `mid` doesn't appear in the output. But the curve's midpoint is fully determined the moment $(\beta_0, \beta_1)$ are estimated — it's just a derived quantity. To get $LT_{50}$ from a GLM:

```r
fit  <- glm(cbind(dead, alive) ~ log10(time), family = binomial(link = "logit"))
b0   <- coef(fit)[1]
b1   <- coef(fit)[2]
mid  <- -b0 / b1                # log10(LT50)
LT50 <- 10 ^ mid                # back to the time scale
```

Equivalently, the `MASS::dose.p()` function in R computes $LT_{50}$ (or any $LT_x$) from a binomial GLM fit by exactly this inversion.

## When the 2PL is enough vs when you actually need the 4PL

Use the **2PL (logit-binomial GLM)** when:

- Baseline mortality is effectively zero — i.e. essentially every organism is alive at very short exposures.
- No irreducible-survivor plateau — i.e. essentially all organisms eventually die under prolonged extreme heat in the assay.
- Data span enough of the dose-response that asymptote behaviour is well-pinned by the data (you're not extrapolating into uncovered regions).

Use the **full 4PL** when:

- Some baseline mortality exists (handling stress, controls aren't perfect) → $u < 1$.
- A fraction of the population is irreducibly resistant or the assay can't kill everyone in a tractable timeframe → $\ell > 0$.
- Data are sparse near the asymptotes and the asymptote values matter for downstream quantities like $LT_{50}$ and $z$.

In the manuscript's joint Bayesian 4PL, $\ell$ and $u$ are estimated parameters with priors that constrain them away from the disjoint intervals $[0, 0.49]$ and $[0.51, 1]$ respectively (`@eq-reparam`), so the model fits the asymptotes rather than pinning them. This handles real-world data where the GLM logit-binomial assumption ($\ell=0, u=1$) is too strong.

## Worked simulation: GLM and 4PL recover the same $k$ and `mid`

```{r}
#| label: glm-vs-4pl-setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)

set.seed(20260507)

# True 4PL parameters (with asymptotes 0 and 1 so the GLM is well-specified)
ell_true <- 0
u_true   <- 1
k_true   <- 4
mid_true <- log10(60)

# Generate binomial data at a range of times
time_min <- c(5, 10, 15, 20, 30, 45, 60, 90, 120, 180, 240, 360)
n_per    <- 50
S_true   <- ell_true + (u_true - ell_true) / (1 + exp(k_true * (log10(time_min) - mid_true)))
dead     <- rbinom(length(time_min), size = n_per, prob = 1 - S_true)
alive    <- n_per - dead

# Standard binomial GLM (the 2PL fit)
fit_glm <- glm(cbind(alive, dead) ~ log10(time_min), family = binomial(link = "logit"))
b0      <- coef(fit_glm)[1]
b1      <- coef(fit_glm)[2]
k_hat   <- -b1
mid_hat <- -b0 / b1

sprintf("True parameters:    k = %.3f, mid = %.3f (LT50 = %.1f min)",
        k_true, mid_true, 10^mid_true)
sprintf("GLM-recovered:      k = %.3f, mid = %.3f (LT50 = %.1f min)",
        k_hat, mid_hat, 10^mid_hat)
sprintf("β0 = %.3f, β1 = %.3f", b0, b1)
```

```{r}
#| label: fig-glm-vs-truth
#| fig-cap: "Simulated binomial-survival data (points) at 12 time points, with the true 4PL curve (solid) and the GLM 2PL fit (dashed). The two curves are visually indistinguishable when the asymptotes are 0 and 1 — confirming the 2PL is the GLM with $\\ell, u$ pinned, and `mid` recoverable from $-\\beta_0/\\beta_1$."
#| fig-height: 3.0

t_grid <- 10 ^ seq(log10(2), log10(500), length.out = 200)
S_truth <- ell_true + (u_true - ell_true) / (1 + exp(k_true * (log10(t_grid) - mid_true)))
S_glm   <- 1 / (1 + exp(-(b0 + b1 * log10(t_grid))))

df_curves <- tibble(
  time_min = rep(t_grid, 2),
  S        = c(S_truth, S_glm),
  Curve    = rep(c("Truth (4PL with l=0, u=1)", "GLM 2PL fit"), each = length(t_grid))
)
df_data <- tibble(time_min = time_min,
                  S = alive / (alive + dead))

ggplot() +
  geom_line(data = df_curves,
            aes(time_min, S, colour = Curve, linetype = Curve),
            linewidth = 1.2, alpha = 0.7) +
  geom_point(data = df_data, aes(time_min, S),
             colour = "black", size = 2) +
  scale_x_log10() +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "Time (min, log scale)", y = "Survival fraction") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The simulation confirms the algebraic mapping: the 2PL GLM recovers the true $k$ and `mid` (and hence $LT_{50}$) of a 4PL with $\ell=0, u=1$ to within sampling noise.

## Implications for the manuscript

- The two-stage TDT pipeline as practised in the literature [@rezende_landscapes_2014; @jorgensen_drosophila_2019] typically uses a 2PL (logit binomial GLM) at Stage 1 to extract $LT_{50}$ at each temperature, then OLS regresses $\log_{10}(LT_{50})$ on $T$ at Stage 2 to extract $z$ and $CT_{max}$. The 2PL is hardcoded into Stage 1 — there's no asymptote estimation; the model assumes $\ell = 0, u = 1$.
- The joint Bayesian 4PL relaxes this assumption by estimating $\ell$ and $u$ as parameters with informative priors. When the data are consistent with $\ell \approx 0, u \approx 1$, the 4PL collapses to a 2PL and the two methods give identical $k$, `mid`, $LT_{50}$. When the data show baseline mortality or irreducible survivors, the 4PL captures it; the 2PL would force the curve to pass through 0 and 1 and produce biased $LT_{50}$ estimates.
- This is one specific instance of the general "the joint Bayesian 4PL collapses to the two-stage approach in benign data, but adds robustness in non-benign data" framing that runs through the manuscript.

## Verification status

- **Algebra (`@eq-2pl-mapping`)**: standard derivation, verified by direct substitution in both directions.
- **Simulation**: built fresh in this note; recovers true parameters within sampling noise. Reproducible from the code chunk.
- **`MASS::dose.p()` claim**: standard R function for $LT_x$ extraction from a binomial GLM; widely documented (e.g. Venables & Ripley *Modern Applied Statistics with S*).
- **Manuscript framing**: the two-stage Stage 1 typically uses the 2PL or a Probit equivalent — not formally cited yet here, but the framing is consistent with @jorgensen_drosophila_2019's description of their Stage 1 fits and standard practice in the TDT literature.
