---
title: "Plan: phylogenetic models in fit_4pl() (issue #12)"
date: 2026-09-25
format:
  html:
    embed-resources: true
    toc: true
---

## What Pieter wants

One joint model comparing $CT_{max}$ and $z$ across many species, with related
species allowed to be similar (phylogenetic covariance). In `brms` this is
written:

```r
fit_4pl(std_phy,
        ctmax = ~ 1 + (1 | gr(species, cov = A)),
        z     = ~ 1 + (1 | gr(species, cov = A)),
        t_ref = 60, data2 = list(A = A))
```

`A` is the phylogenetic correlation matrix from `ape::vcv(phy, corr = TRUE)`.
Right now this call fails at the `gr()` part. Target: **bayesTLS 1.1.0**.

## Why it breaks — three separate problems

**1. Our formula handling treats formulas as text.** Two helpers
(`rhs_fixed_vars()` at `R/fit_4pl.R` ~662 and `normalize_cellmeans()` ~693)
strip out random-effect terms using text patterns. The patterns cannot cope
with brackets inside brackets — `(1 | gr(species, cov = A))` — so they leave a
stray `)` behind and the formula step aborts. This is the error Pieter saw
(Snippet 1 below reproduces it).

**2. We misread which variables are data columns.** The helper that pulls
grouping variables out of random-effect terms (`R/utils.R:133`) would return
both `species` **and** `A` — but `A` is a matrix passed separately, not a data
column, so a column check would wrongly demand a column called `A`.

**3. Estimates per species would silently be wrong.** All post-fit quantities
come from one engine, and it currently ignores random effects
(`re_formula = NA`, `R/tls_engine.R:76`). With species included only as a
random effect, `tls(by = "species")` would return the *same* number for every
species — the overall mean — with no warning. Today this is harmless because
all our models put groups in as fixed effects, but for Pieter it would be a
silent trap.

Good news: the sampler itself needs no work (`fit_4pl()` already passes extra
arguments through to `brms::brm()`, so `data2` arrives), and the prior code
already copes when the grouping factor is not a fixed effect (`R/priors.R:186`).

## What we should do, and why

**Fix the formula handling so users write plain `brms` syntax, add `data2` as
a proper documented argument, and teach the extraction engine to include
random effects when asked for them.**

Why this option:

- No new modelling interface to learn or maintain — anything `gr()` can do in
  `brms` works for free (e.g. `gr(species, by = region, cov = A)`).
- The fix is small: parse formulas as R code instead of text (~12 lines,
  Snippet 2), name and validate `data2`, and add one careful default to the
  engine.
- It is what the issue itself asks for.

Considered and rejected: a convenience wrapper like
`phylo = list(group, A)` (nice sugar, can add later; not the foundation), and
depending on `lme4` just for its formula parser (heavy dependency for 12
lines).

## API impact (decided 2026-09-25: keep changes minimal)

Exactly **one new argument in the whole package**: `data2 = NULL` on
`fit_4pl()` (the `brms` name users already know; it already reaches
`brms::brm()` through `...`, so naming it buys documentation, autocomplete,
and a home for validation). Everything else — the formula parser, the
grouping-variable reader, the extraction engine's random-effect rule — is
internal. No argument anywhere is renamed, removed, or re-defaulted, and every
existing call stays bit-identical (test T13).

To be precise about `make_4pl_formula()`, since it is where formulas are
built: its **behaviour changes** (today it aborts on `gr()`; afterwards it
emits the term verbatim), but the **edited code is not in its definition**.
It splices each sub-parameter's right-hand side into the `brms` formula as-is
and never dissects random-effect terms itself — all dissection happens in four
small internal helpers (`rhs_fixed_vars`, `normalize_cellmeans`,
`direct_group_vars`, `tdt_random_effect_variables`), which are what get
rewritten. In Pieter's direct-mode call the crash happens in `fit_4pl()`'s
body (`normalize_cellmeans`, `R/fit_4pl.R:417`) before `make_4pl_formula()`
is even reached; in midpoint mode it happens inside `make_4pl_formula()` via
the helper it calls at line 192. Caveat: if the new splitter changes what that
line-192 guard consumes, a 1–2 line touch there may be needed — test T4
(emitted formulas contain `gr()` verbatim, both parameterisations) settles
it.

## The plan, in order

1. **Write the failing tests first** (Pieter's call, at the formula level —
   no Stan needed).
2. **Replace the text patterns** with a small proper parser (Snippet 2). Both
   broken helpers get rebuilt on it.
3. **Read grouping variables correctly**: inside `gr()`, everything except
   `cov =` and `dist =` names data columns; `cov =` names a `data2` matrix.
   `mm()` (multi-membership) gets a clear "not supported" error.
4. **Add `data2` as a named argument** with checks: every `cov =` matrix must
   be present in `data2`, symmetric, with row/column names covering the
   species levels; warn if it looks like a covariance rather than a
   correlation matrix (diagonal not 1).
5. **Extraction**: no new argument (decided 2026-09-25 — keep function
   signatures minimal). The engine gets an *internal* rule: if a `by` variable
   is only a random effect, include exactly that random-effect term; otherwise
   behave exactly as today. Never blanket-include all random effects — the
   internal prediction grid pads unrelated grouping columns with arbitrary
   values, which would inject phantom effects. An explicit `re_formula`
   override is deferred to a follow-up issue if anyone needs it.
6. **Docs and release**: document `data2` with an `ape` example, README line,
   NEWS, version 1.1.0. `ape` stays out of our dependencies — tests use a
   small committed correlation matrix (with a `data-raw/` script to
   regenerate it).

Statistical notes, briefly: the species effect on $z$ acts on the log scale
(where the model's predictor lives — natural for a positive quantity); users
with repeated measures per species can add a plain `(1 | species)` next to the
`gr()` term to separate phylogenetic from other species differences; and
per-species estimates are shrinkage estimates — that is the point of the
joint model.

## Tests

Fast tests that run in every check (no Stan):

| # | What it checks |
|---|----------------|
| T1 | The new parser on: plain `(1 | g)`, `gr(cov=)`, nested `(1 | a/b)`, slopes `(temp_c | g)`, several terms mixed with fixed effects |
| T2 | Grouping variables: `gr(species, cov = A)` → `species` only, never `A`; `gr(g, by = b)` → both; `mm()` → clear error |
| T3 | `normalize_cellmeans()` passes `gr()` formulas through untouched |
| T4 | `make_4pl_formula()` puts the `gr()` term, verbatim, into the right sub-parameter formula (both parameterisations); `fit_4pl(fit = FALSE)` builds the spec |
| T5 | Error quality: `gr(cov = A)` with no `data2`, missing names, asymmetry, level mismatch → clear messages; covariance-instead-of-correlation → warning |
| T6 | `brms::make_stancode()` accepts our built formula + `data2` (generates the model code without compiling; Snippet 3) |
| T7 | `standardize_data()` keeps the species column when it appears only inside `gr()` |
| T8 | The `re_formula` default rule as a pure function: fixed-effect `by` → unchanged behaviour; random-effect-only `by` → exactly that term |

Slower tests behind `RUN_BRMS_TESTS=true` (cached fits, as the suite already
does):

| # | What it checks |
|---|----------------|
| T9 | New fixture: ~10 simulated species with known phylogenetic effects on $CT_{max}$ and $z$, small cached fit |
| T10 | Recovery: per-species estimates track the simulated truth; overall values inside their credible intervals |
| T11 | Parity: `gr(species, cov = identity)` vs plain `(1 | species)` on the same data agree within sampling noise |
| T12 | Isolation: adding an unrelated `(1 | batch)` term does not move the per-species estimates (guards step 5's grid subtlety) |
| T13 | No drift: the entire existing suite stays green and bit-identical — the new default must not change any current result |
| T14 | `derive_tdt_curve()` and `predict_survival_curves()` run on the phylo fixture and return per-species output |

## Checks you can run now

### Snippet 1 — reproduce the bug

The current helper, applied to Pieter's formula:

```{r repro}
rhs_fixed_vars_current <- function(rhs) {
  rhs <- gsub("\\+?\\s*\\([^|]*\\|[^)]*\\)", "", rhs)
  rhs <- gsub("\\+\\s*\\+", "+", rhs)
  rhs <- trimws(gsub("\\+\\s*$|^\\s*\\+", "", rhs))
  if (rhs %in% c("", "1", "0")) return(character(0))
  setdiff(all.vars(stats::as.formula(paste("~", rhs))), "temp_c")
}

ok    <- rhs_fixed_vars_current("1 + (1 | species)")            # legacy: fine
broke <- tryCatch(
  rhs_fixed_vars_current("1 + (1 | gr(species, cov = A))"),
  error = function(e) conditionMessage(e))
ok
broke
stopifnot(identical(ok, character(0)), is.character(broke), length(broke) == 1)
```

The plain form works; the `gr()` form errors — the reported break.

### Snippet 2 — the proposed parser handles every case

```{r splitter}
# Split a formula right-hand side into fixed terms and (x | g) terms by
# walking the parsed expression — no text patterns.
tdt_split_bars <- function(rhs) {
  expr <- str2lang(paste("~", rhs))[[2]]
  bars <- list()
  walk <- function(e) {
    if (is.call(e)) {
      if (identical(e[[1]], as.name("(")) && is.call(e[[2]]) &&
          identical(e[[2]][[1]], as.name("|"))) {
        bars[[length(bars) + 1]] <<- e[[2]]
        return(NULL)
      }
      if (identical(e[[1]], as.name("+"))) {
        l <- walk(e[[2]]); r <- if (length(e) > 2) walk(e[[3]]) else NULL
        if (is.null(l)) return(r); if (is.null(r)) return(l)
        return(call("+", l, r))
      }
    }
    e
  }
  fixed <- walk(expr)
  list(fixed = if (is.null(fixed)) "1" else deparse(fixed), bars = bars)
}

# Data columns named by a random-effect term: inside gr(), everything except
# cov=/dist=.
bar_group_vars <- function(bar) {
  g <- bar[[3]]
  if (is.call(g) && identical(g[[1]], as.name("mm")))
    stop("mm() grouping is not yet supported in bayesTLS formulas.")
  if (is.call(g) && identical(g[[1]], as.name("gr"))) {
    keep <- setdiff(seq_along(g)[-1], which(names(g) %in% c("cov", "dist")))
    return(unique(unlist(lapply(as.list(g)[keep], all.vars))))
  }
  all.vars(g)
}

cases <- list(
  "1 + (1 | gr(species, cov = A))",
  "0 + region + (1 | gr(species, cov = A)) + (1 | batch)",
  "temp_c + (temp_c | site) + (1 | a/b)")
res <- lapply(cases, tdt_split_bars)
str(lapply(res, function(r)
  list(fixed = r$fixed, groups = lapply(r$bars, bar_group_vars))))

stopifnot(
  identical(res[[1]]$fixed, "1"),
  identical(bar_group_vars(res[[1]]$bars[[1]]), "species"),      # never "A"
  identical(res[[2]]$fixed, "0 + region"),
  identical(bar_group_vars(res[[2]]$bars[[2]]), "batch"),
  identical(res[[3]]$fixed, "temp_c"),
  identical(bar_group_vars(res[[3]]$bars[[2]]), c("a", "b")))
```

### Snippet 3 — brms accepts the target model (no compiling)

```{r stancode, message=FALSE, warning=FALSE}
library(brms)
set.seed(1)
sp  <- paste0("s", 1:4)
A   <- diag(4); A[A == 0] <- 0.3; dimnames(A) <- list(sp, sp)
toy <- data.frame(species = rep(sp, each = 3),
                  temp_c  = rep(c(-1, 0, 1), 4),
                  n = 20, y = rbinom(12, 20, 0.6))

bf_phy <- bf(y | trials(n) ~ a, nl = TRUE,
             a ~ 1 + (1 | gr(species, cov = A)))
code <- make_stancode(bf_phy, data = toy, data2 = list(A = A),
                      family = binomial())
grep("cholesky|Lcov", strsplit(as.character(code), "\n")[[1]],
     value = TRUE, ignore.case = TRUE)[1:3]
stopifnot(any(grepl("Lcov|cholesky", as.character(code), ignore.case = TRUE)))
```

The generated model code contains the covariance term — so once our formula
handling stops mangling `gr()`, `brms` takes it from there.

## Open questions for you

1. Want a phylogenetic-signal summary in `tls()` (share of species variance
   attributable to phylogeny, lambda-like)? Useful for Pieter, but I suggest a
   follow-up issue so 1.1.0 stays small.
2. Does Pieter need `gr(species, by = region, ...)` (different phylo variance
   per region)? It will parse, but I would only test it properly if he needs
   it.
