---
title: "Do moderator×temperature terms on low/up/k matter once z and CTmax are parameterised directly?"
date: 2026-06-25
format:
  html:
    embed-resources: true
    toc: true
    code-fold: show
execute:
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({library(bayesTLS); library(dplyr); library(tidyr)})
models_dir <- here::here("output", "models")
```

## The question

We reformulated the 4PL

$$
p \;=\; \text{low} \;+\; \frac{\text{up} - \text{low}}{1 + \exp\!\bigl(k\,(\log_{10} t - \text{mid})\bigr)}
$$

so that $z$ and $CT_{max}$ can be estimated **directly** (the "direct" parameterisation:
`ctmax = ~ ...`, `z = ~ ...`), instead of estimating the linear midpoint
$\text{mid}(T) = \beta_0 + \beta_1 (T-\bar T)$ and deriving $z = -1/\beta_1$ from it.

With a moderator such as **life stage**, the working assumption had been: to get
per-stage $z$ and $CT_{max}$ we must put `temp_c * stage` on **all four** 4PL
sub-parameters, "because the asymmetry correction needs it." Is that still true
once $z$ and $CT_{max}$ are built into the equation?

## The maths: relative $z$/$CT_{max}$ depend on `mid` alone

`bayesTLS` reads thermal-tolerance quantities at a **threshold**. The default,
`target_surv = "relative"`, sets the threshold to $(\text{low}+\text{up})/2$
**per posterior draw**. By construction of the 4PL, $p = (\text{low}+\text{up})/2$
exactly at $\log_{10}t = \text{mid}$. Hence

$$
\log_{10}\mathrm{LT}_{\text{rel}}(T) \;=\; \text{mid}(T)\qquad\text{exactly, for *any* low, up, k (even } T\text{-varying).}
$$

(The code does exactly this — `extract_tdt()` takes the "direct shortcut
`log10(t_relative) = mid(T)` per draw" with no reference to low/up/k.) Therefore

- $z_{\text{rel}}(T) = -1/(\mathrm{d}\,\text{mid}/\mathrm{d}T)$, and
- $CT_{max,\text{rel}}$ solves $\text{mid}(T) = \log_{10} t_{\text{ref}}$,

are functions of **`mid` only**. The shape parameters cancel. The direct
parameterisation simply *names* the slope and intercept of `mid(T)` as $z$ and
$CT_{max}$, so

> `ctmax = ~ 0 + stage`, `z = ~ 0 + stage`  $\;\equiv\;$  `stage` **and** `stage:temp_c` on `mid`.

The asymmetry correction $\tfrac{1}{k}\log\frac{\text{up}-x}{x-\text{low}}$ enters
**only** the *absolute* threshold (LT at a fixed survival $x$). It is a constant
offset (no effect on $z$, only shifts $CT_{max}$) when low/up/k are constant in
$T$; it **bends** the $\log_{10}\mathrm{LT}(T)$ curve — making absolute $z$ a local
slope that differs from relative $z$ — **only when low/up/k vary with $T$**; and it
is $\approx 0$ anyway when $\text{low}\approx0,\ \text{up}\approx1,\ x=0.5$.

**So:** for the per-stage *relative* $z$/$CT_{max}$ you need `stage` (+ its
temperature interaction) on `mid` — i.e. `ctmax = ~ stage`, `z = ~ stage` in the
direct form — and **not** on low/up/k.

## Simulation: does it hold when the shape genuinely varies with $T$ and stage?

We generate two life stages with **different** true $z$ and $CT_{max}$, and we let
low/up/k vary with both temperature **and** stage (the very thing the assumption
said we must model). We then fit the direct parameterisation two ways and compare
the recovered **relative** $z$/$CT_{max}$ to truth.

```{r simulate}
set.seed(1)
T_bar <- 34; temps <- c(30,32,34,36,38); durs <- c(1,5,15,45,135,405)
nrep  <- 25; n <- 30; tref <- 60
truth <- list(A = list(b0 = 1.6, b1 = -0.12),   # z_A = -1/b1 = 8.33
              B = list(b0 = 1.3, b1 = -0.20))    # z_B = 5.00
z_true  <- sapply(truth, function(p) -1 / p$b1)
ct_true <- sapply(truth, function(p) T_bar + (log10(tref) - p$b0) / p$b1)

# low/up/k vary with T AND stage; up kept inside (0, 1)
shape <- function(stage, Tc) {
  up <- if (stage == "A") 0.90 - 0.010 * Tc else 0.88 - 0.015 * Tc
  k  <- if (stage == "A") 7 else 11
  c(low = 0.02, up = up, k = k)
}
d <- tidyr::expand_grid(stage = c("A","B"), T = temps, t = durs, rep = 1:nrep) |>
  dplyr::mutate(Tc = T - T_bar, log10_t = log10(t)) |>
  dplyr::rowwise() |>
  dplyr::mutate(b0 = truth[[stage]]$b0, b1 = truth[[stage]]$b1, mid = b0 + b1 * Tc,
                low = shape(stage, Tc)["low"], up = shape(stage, Tc)["up"], k = shape(stage, Tc)["k"],
                p = low + (up - low) / (1 + exp(k * (log10_t - mid))),
                y = rbinom(1, n, p), n_total = n) |>
  dplyr::ungroup()
std <- standardize_data(d, temp = "T", duration = "t", n_total = "n_total",
                        n_surv = "y", duration_unit = "minutes")
```

```{r fit}
# M1: per-stage CTmax/z, CONSTANT low/up/k (shape NOT modelled)
m1 <- fit_4pl(std, ctmax = ~ 0 + stage, z = ~ 0 + stage,
              low = ~ 1, up = ~ 1, k = ~ 1,
              family = binomial("identity"), chains = 2, iter = 1500, warmup = 750,
              cores = 2, seed = 2, refresh = 0,
              file = file.path(models_dir, "note_shape_m1"), file_refit = "on_change")

# M2: per-stage CTmax/z, FULL stage*temp_c on low/up/k (shape fully modelled)
m2 <- fit_4pl(std, ctmax = ~ 0 + stage, z = ~ 0 + stage,
              low = ~ stage * temp_c, up = ~ stage * temp_c, k = ~ stage * temp_c,
              family = binomial("identity"), chains = 2, iter = 1500, warmup = 750,
              cores = 2, seed = 2, refresh = 0,
              file = file.path(models_dir, "note_shape_m2"), file_refit = "on_change")
```

```{r recover}
rel <- function(m, label) tls(m, by = "stage", target_surv = "relative", t_ref = tref)$summary |>
  dplyr::filter(quantity %in% c("z", "CTmax")) |>
  dplyr::transmute(model = label, stage, quantity, est = round(median, 2),
                   lower = round(lower, 2), upper = round(upper, 2))
truth_tbl <- tibble::tibble(model = "TRUTH",
  stage = rep(c("A","B"), 2), quantity = rep(c("z","CTmax"), each = 2),
  est = round(c(z_true, ct_true), 2), lower = NA, upper = NA)
dplyr::bind_rows(truth_tbl, rel(m1, "M1: constant low/up/k"),
                 rel(m2, "M2: full stage*temp on low/up/k")) |>
  dplyr::arrange(quantity, stage, model) |>
  as.data.frame()
```

### What the numbers say

- **Both** fits recover the per-stage **relative** $z$ and $CT_{max}$ close to
  truth — confirming that `stage` on `ctmax`/`z` (direct) is sufficient; the
  shape×temperature terms on low/up/k are **not** needed for these quantities.
- **M1** (constant shape) carries a **small downward bias in $z$** (~3–4%, its
  credible intervals just exclude the truth) *because* the true shape varies with
  $T$, and forcing it constant slightly biases the **estimate** of `mid`. The
  relative-$z$ identity ($z = f(\text{mid})$) is exact; estimating `mid` without
  bias still benefits from a flexible enough shape model.
- **M2** (shape modelled) removes that residual bias.

## Where low/up/k×temperature *do* matter

```{r absolute}
# The absolute (true 50%-survival) z is a LOCAL slope of a curve the correction
# bends when low/up/k vary with T -- so M1 and M2 can disagree here even though
# their relative z agreed above.
abs_z <- function(m, label) tls(m, by = "stage", target_surv = "absolute", t_ref = tref)$summary |>
  dplyr::filter(quantity == "z") |>
  dplyr::transmute(model = label, stage, abs_z = round(median, 2))
dplyr::bind_rows(abs_z(m1, "M1: constant low/up/k"),
                 abs_z(m2, "M2: full stage*temp on low/up/k")) |>
  as.data.frame()
```

If you report **absolute** thresholds (true LT50) *and* the asymptotes/steepness
genuinely change with temperature, the correction bends per stage and you must
model low/up/k (by $T$, and by stage) to get it right.

## Bottom line

- **Relative $z$/$CT_{max}$ (what the direct parameterisation gives):** put the
  moderator (and its temperature interaction) on the **midpoint only** — i.e.
  `ctmax = ~ stage`, `z = ~ stage`. You do **not** need `temp_c * stage` on
  low/up/k.
- **Keep the flexible default as cheap insurance.** `temp_effects =
  c("low","up","k","mid")` (or `by =`) lets the data decide: if the shape does not
  move with temperature those slopes shrink toward 0; if it does, modelling it
  removes the small misspecification bias seen in M1. With clean asymptotes
  ($\approx 0/1$) roughly constant in $T$, the constant-shape model is fine and
  more parsimonious.
- **Absolute thresholds with $T$-varying shape** are the one case where
  `temp_c`(`×stage`) on low/up/k is genuinely required — that is what the
  asymmetry correction is for.
```
