---
title: "Direct CTmax/z parameterisation: prototype + validation"
subtitle: "Design mock-up (NOT yet in the package), validated on zebrafish and Drosophila"
date: 2026-06-22
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: show
---

```{r setup}
#| message: false
#| warning: false
pacman::p_load(brms, posterior, dplyr, tibble, here)
devtools::load_all(here::here(), quiet = TRUE)   # for standardize_data, tls, compute_4pl_bounds
models_dir <- here::here("output", "models")
options(mc.cores = 4)
cb  <- getFromNamespace("compute_4pl_bounds", "bayesTLS")
qci <- function(x) sprintf("%.2f [%.2f, %.2f]", median(x),
                           quantile(x, .025, names = FALSE), quantile(x, .975, names = FALSE))
```

## 1. What this designs (and the decisions behind it)

A **mock-up** of the direct CTmax/z parameterisation, prototyped in *this note
only* — `R/` and the supplement are untouched. Design decisions (Daniel,
2026-06-22):

- **No `parameterization` argument.** Supplying a `ctmax` and/or `z` formula *is*
  the switch to direct mode. Don't overcomplicate.
- **Arguments are formulas, no `_formula` suffix:** `ctmax`, `z`, `up`, `low`,
  `k` (the docs/examples make clear they take formulas); the asymptote bounds
  collapse to a single `bounds = c(0, 1)` so `low`/`up` don't clash with the old
  `lower`/`upper`.
- **`threshold` selects the fitted backbone.** `"relative"` (default) puts the
  linear-in-T backbone on the midpoint; `"absolute"` bakes the up/low/k
  temperature effects into the asymmetry correction so the fitted CTmax and z
  **are** the absolute (fixed-survival) quantities.
- **Per-parameter formulas for `low`/`up`/`k` too** (symmetry with CTmax/z);
  when omitted they inherit the CTmax/z fixed effects crossed with `temp_c` (RE
  not inherited), else `~ temp_c` (see *Formula resolution / inheritance*).
- Estimated nonlinear parameters change from `(lowraw, upraw, logk, mid)` to
  `(lowraw, upraw, logk, CTmaxdev, logz)`; `low/up/k/mid` become `nlf`-derived.

Full design + the function-by-function change list:
`notes/2026-06-22-direct-ctmax-z-package-design.md`. Priors follow the
load-bearing-asymptote finding in
`notes/2026-06-21-prior-sensitivity-direct-ctmax-z.qmd`.

## 2. Core function changes (the prototype)

This is what `make_4pl_formula()` becomes. `low/up/k/mid` are defined with
`brms::nlf()`; the estimated nlpars carry the user formulas.

```{r proto-formula}
rhs <- function(f, default) if (is.null(f)) default else paste(deparse(f[[2]]), collapse = " ")

# Resolution rule: an explicit shape formula wins; otherwise the shape inherits
# the CTmax/z FIXED effects crossed with temp_c (RE is NOT inherited — it stays
# on CTmax/z); intercept-only CTmax/z -> shape ~ temp_c.
resolve_shape <- function(shape_f, inherit_rhs) {
  if (!is.null(shape_f)) return(rhs(shape_f, "temp_c"))
  fe <- trimws(gsub("\\+?\\s*\\([^|]*\\|[^)]*\\)", "", inherit_rhs))   # drop (x | g) RE terms
  fe <- trimws(gsub("\\+\\s*$|^\\s*\\+\\s*", "", fe))
  if (fe %in% c("", "1")) return("temp_c")
  if (grepl("^0\\s*\\+", fe)) {                                        # cell-means: 0 + G
    core <- trimws(sub("^0\\s*\\+\\s*", "", fe))
    if (grepl("\\+", core)) paste0("0 + ", core, " + temp_c:(", core, ")")
    else                    paste0("0 + ", core, " + temp_c:", core)
  } else paste0("(", fe, ") * temp_c")                                 # treatment: G * temp_c
}

# Final API: ctmax / z / up / low / k are FORMULAS (no `_formula` suffix);
# `bounds = c(lower, upper)` is the disjoint-bounds asymptote range.
proto_4pl_formula <- function(ctmax = NULL, z = NULL, up = NULL, low = NULL, k = NULL,
                              threshold = c("relative", "absolute"),
                              log10_tref = 0, p = 0.5, bounds = c(0, 1),
                              family = brms::beta_binomial(link = "identity")) {
  threshold <- match.arg(threshold)
  b     <- cb(bounds[1], bounds[2])
  low_e <- sprintf("(%.6f + inv_logit(lowraw) * %.6f)", b$low_min, b$low_w)
  up_e  <- sprintf("(%.6f + inv_logit(upraw)  * %.6f)", b$up_min,  b$up_w)
  # one linear-in-T backbone; mid reconstructed from CTmax (centred) and z
  mid_rel <- sprintf("(%g - (temp_c - CTmaxdev) / exp(logz))", log10_tref)
  mid_e   <- if (threshold == "relative") mid_rel else
    sprintf("(%s - (1/exp(logk)) * log((up - %g)/(%g - low)))", mid_rel, p, p)  # absolute: + C(T)
  inherit <- if (!is.null(ctmax)) rhs(ctmax, "1") else if (!is.null(z)) rhs(z, "1") else "1"
  brms::bf(
    stats::as.formula("n_surv | trials(n_total) ~ low + (up - low) / (1 + exp(exp(logk) * (logd - mid)))"),
    brms::nlf(stats::as.formula(paste("low ~", low_e))),
    brms::nlf(stats::as.formula(paste("up  ~", up_e))),
    brms::nlf(stats::as.formula(paste("mid ~", mid_e))),
    stats::as.formula(paste("lowraw   ~", resolve_shape(low, inherit))),  # inherit FE x temp_c, else temp_c
    stats::as.formula(paste("upraw    ~", resolve_shape(up,  inherit))),
    stats::as.formula(paste("logk     ~", resolve_shape(k,   inherit))),
    stats::as.formula(paste("CTmaxdev ~", rhs(ctmax, "1"))),               # CTmax structure (FE + RE)
    stats::as.formula(paste("logz     ~", rhs(z,     "1"))),               # z structure (FE + RE)
    nl = TRUE, family = family)
}
```

And `make_4pl_priors()`: keep the **centred** asymptote priors (load-bearing —
flattening them gave 191 divergences in the prior-sensitivity note); CTmax and z
are well identified, so they take simple priors. Cell-means terms (`0 + factor`)
get one centred prior per level.

```{r proto-priors}
proto_priors <- function(cells = NULL, bounds = c(0, 1), re_ctmax = NULL, re_z = NULL) {
  b  <- cb(bounds[1], bounds[2])
  lr <- qlogis((bounds[1] + 0.02 * (bounds[2] - bounds[1]) - b$low_min) / b$low_w)
  ur <- qlogis((bounds[1] + 0.98 * (bounds[2] - bounds[1]) - b$up_min)  / b$up_w)
  per <- function(centre, sd, nlpar) if (is.null(cells))
      set_prior(sprintf("normal(%.4f,%.1f)", centre, sd), class = "b", nlpar = nlpar, coef = "Intercept")
    else do.call(c, lapply(cells, \(cf) set_prior(sprintf("normal(%.4f,%.1f)", centre, sd),
                                                  class = "b", nlpar = nlpar, coef = cf)))
  pr <- c(per(lr, 1, "lowraw"),  set_prior("normal(0,0.5)", class = "b", nlpar = "lowraw"),
          per(ur, 1, "upraw"),   set_prior("normal(0,0.5)", class = "b", nlpar = "upraw"),
          per(log(2), 1, "logk"),set_prior("normal(0,0.3)", class = "b", nlpar = "logk"),
          set_prior("normal(0,10)",       class = "b", nlpar = "CTmaxdev"),
          set_prior("normal(1.0986,0.7)", class = "b", nlpar = "logz"),
          set_prior("gamma(2,0.1)", class = "phi"))
  if (!is.null(re_ctmax)) pr <- c(pr, set_prior("exponential(2)", class = "sd", nlpar = "CTmaxdev", group = re_ctmax))
  if (!is.null(re_z))     pr <- c(pr, set_prior("exponential(2)", class = "sd", nlpar = "logz",     group = re_z))
  pr
}
```

### Formula resolution / inheritance

When `up`/`low`/`k` are omitted they are resolved from the CTmax/z structure:
fixed effects are inherited **crossed with `temp_c`** (so the absolute correction
is correct per group), random effects are **not** inherited (they stay on
CTmax/z), and intercept-only CTmax/z gives `shape ~ temp_c`:

```{r resolution-demo}
inh  <- function(ctmax) if (is.null(ctmax)) "1" else rhs(ctmax, "1")
demo <- function(ctmax) data.frame(`ctmax / z` = inh(ctmax),
                                   `shape (resolved)` = resolve_shape(NULL, inh(ctmax)),
                                   check.names = FALSE)
do.call(rbind, list(demo(~ 1),
                    demo(~ 0 + life_stage + (1 | Date_experiment)),
                    demo(~ 0 + sex),
                    demo(~ life_stage)))
```

The grouped default (`shape ~ 0 + life_stage + temp_c:life_stage`, RE on CTmax/z
only) is exactly the `proto_zf_direct_rel` structure validated below — **0
divergences**, matches the reference. The alternative of inheriting RE onto the
shape too was tested (`proto_zf_inherit`) and gave **4 divergences** with no
better estimates, so RE is not inherited.

## 3. The generated model (zebrafish)

```{r show-formula}
zf <- zebrafish_lethal; tbar <- mean(zf$assay_temp, na.rm = TRUE)
stages <- levels(zf$life_stage)
zf_joint <- dplyr::bind_rows(lapply(stages, function(s) {
  d <- standardize_data(dplyr::filter(zf, life_stage == s), temp = "assay_temp", duration = "duration_h",
        n_total = "n_total", n_surv = "n_surv", random_effects = "Date_experiment",
        duration_unit = "hours", temp_mean = tbar)
  d$life_stage <- factor(s, levels = stages); d }))

f_zf <- proto_4pl_formula(
  ctmax = ~ 0 + life_stage + (1 | Date_experiment),
  z     = ~ 0 + life_stage + (1 | Date_experiment),
  low   = ~ 0 + life_stage + temp_c:life_stage,   # explicit here to match the reference structure
  up    = ~ 0 + life_stage + temp_c:life_stage,   # (omitting these would inherit the same by the rule)
  k     = ~ 0 + life_stage + temp_c:life_stage,
  threshold = "relative")
f_zf
```

## 4. O1 — does `posterior_linpred()` reach the `nlf`-derived `mid`?

This matters because `tls()`/`extract_tdt()` derive z/CTmax by evaluating
sub-parameters via `posterior_linpred(nlpar=)`. If `mid` is reachable, those
extractors need **almost no change**.

```{r fit-zf}
#| message: false
#| warning: false
fit_zf <- brm(f_zf, data = zf_joint, prior = proto_priors(cells = paste0("life_stage", stages),
                                                           re_ctmax = "Date_experiment", re_z = "Date_experiment"),
              chains = 4, iter = 2000, warmup = 1000, cores = 4, seed = 123,
              control = list(adapt_delta = 0.99, max_treedepth = 15), init = 0,
              backend = "cmdstanr", silent = 2, refresh = 0,
              file = file.path(models_dir, "proto_zf_direct_rel"), file_refit = "on_change")

nd <- data.frame(temp_c = 0, logd = 0, life_stage = factor("larvae", levels = stages),
                 Date_experiment = zf_joint$Date_experiment[1], n_total = 10)
o1_mid  <- tryCatch({posterior_linpred(fit_zf, newdata = nd, nlpar = "mid",  re_formula = NA); "OK"},
                    error = function(e) paste("FAILED:", conditionMessage(e)))
o1_logz <- tryCatch({posterior_linpred(fit_zf, newdata = nd, nlpar = "logz", re_formula = NA); "OK"},
                    error = function(e) paste("FAILED:", conditionMessage(e)))
tibble::tibble(nlpar = c("mid (nlf-derived)", "logz (estimated)"), posterior_linpred = c(o1_mid, o1_logz))
```

**Both work.** So `tls()` can keep its grid-slope path on `mid` *and* read
CTmax/z straight off `CTmaxdev`/`logz`. The two agree exactly:

```{r tls-path-check}
tg <- seq(-2, 2, length.out = 11)
ndg <- data.frame(temp_c = tg, logd = 0, life_stage = factor("larvae", levels = stages),
                  Date_experiment = zf_joint$Date_experiment[1], n_total = 10)
midmat <- posterior_linpred(fit_zf, newdata = ndg, nlpar = "mid", re_formula = NA)
w <- tg - mean(tg); slope <- as.vector(midmat %*% w / sum(w^2))
dzf <- as.data.frame(as_draws_df(fit_zf))
tibble::tibble(
  larvae_z = c("posterior_linpred(mid) grid-slope", "direct coef read exp(b_logz)"),
  value    = c(qci(-1 / slope), qci(exp(dzf[["b_logz_life_stagelarvae"]]))))
```

## 5. Zebrafish equivalence vs the midpoint reference

Reference = the manuscript's midpoint joint fit, summarised with the package's
own `tls()` (robust to coding). CTmax read at the 1-hour reference.

```{r zf-equiv}
#| message: false
#| warning: false
ref_zf <- readRDS(file.path(models_dir, "fit_zf_joint_4pl.rds"))
ref_t  <- tls(ref_zf, by = "life_stage", params = c("z", "ctmax"),
              t_ref = 1, time_multiplier = 1, temp_mean = tbar)$summary
refz  <- ref_t[ref_t$quantity == "z", ];      refc <- ref_t[ref_t$quantity == "CTmax", ]
fmt   <- function(df, s) { r <- df[df$life_stage == s, ]; sprintf("%.2f [%.2f, %.2f]", r$median, r$lower, r$upper) }
tibble::tibble(
  life_stage    = stages,
  `z proto`     = vapply(stages, \(s) qci(exp(dzf[[paste0("b_logz_life_stage", s)]])), ""),
  `z ref (tls)` = vapply(stages, \(s) fmt(refz, s), ""),
  `CTmax proto` = vapply(stages, \(s) qci(tbar + dzf[[paste0("b_CTmaxdev_life_stage", s)]]), ""),
  `CTmax ref`   = vapply(stages, \(s) fmt(refc, s), ""))
```

CTmax matches to ~0.05 °C; z medians agree. The proto z **intervals are wider**
because this model carries a random effect on z (`logz ~ ... + (1|Date)`) that
the midpoint reference cannot — exactly as the 2026-06-16 note documented.

## 6. Drosophila equivalence (no random effects → CIs match too)

```{r dros-equiv}
#| message: false
#| warning: false
ref_d <- readRDS(file.path(models_dir, "fit_dsuzukii_mortality_sex_joint.rds"))
dat_d <- ref_d$data
fit_d <- brm(proto_4pl_formula(ctmax = ~ 0 + sex, z = ~ 0 + sex,
                               low = ~ 0 + sex + temp_c:sex, up = ~ 0 + sex + temp_c:sex,
                               k = ~ 0 + sex + temp_c:sex, threshold = "relative"),
             data = dat_d, prior = proto_priors(cells = c("sexF", "sexM")),
             chains = 4, iter = 2000, warmup = 1000, cores = 4, seed = 123,
             control = list(adapt_delta = 0.99, max_treedepth = 15), init = 0,
             backend = "cmdstanr", silent = 2, refresh = 0,
             file = file.path(models_dir, "proto_dros_direct_rel"), file_refit = "on_change")
ddir  <- as.data.frame(as_draws_df(fit_d))
refzd <- tls(ref_d, by = "sex", params = "z")$summary
fmtz  <- function(s) { r <- refzd[refzd$sex == s, ]; sprintf("%.2f [%.2f, %.2f]", r$median, r$lower, r$upper) }
tibble::tibble(
  sex           = c("F", "M"),
  `z proto`     = c(qci(exp(ddir[["b_logz_sexF"]])), qci(exp(ddir[["b_logz_sexM"]]))),
  `z ref (tls)` = c(fmtz("F"), fmtz("M")))
```

Near-identical (the Ørsted-reported values are 3.03 F / 3.28 M). Both fits have
**0 divergences**.

## 7. Flexibility: model CTmax by group but pool z

The point of the layered formulas — here CTmax varies by sex while z is shared
(`z` omitted ⇒ `logz ~ 1`). Shape parameters keep their per-sex
temperature structure.

```{r flex}
#| message: false
#| warning: false
fit_flex <- brm(proto_4pl_formula(ctmax = ~ 0 + sex,         # CTmax per sex
                                  low = ~ 0 + sex + temp_c:sex, up = ~ 0 + sex + temp_c:sex,
                                  k = ~ 0 + sex + temp_c:sex),  # z omitted -> logz ~ 1 (pooled)
                data = dat_d, prior = proto_priors(cells = c("sexF", "sexM")),
                chains = 4, iter = 2000, warmup = 1000, cores = 4, seed = 123,
                control = list(adapt_delta = 0.99, max_treedepth = 15), init = 0,
                backend = "cmdstanr", silent = 2, refresh = 0,
                file = file.path(models_dir, "proto_dros_ctmaxonly"), file_refit = "on_change")
dff <- as.data.frame(as_draws_df(fit_flex))
tibble::tibble(
  quantity = c("z (pooled across sex)", "CTmaxdev sex F", "CTmaxdev sex M"),
  estimate = c(qci(exp(dff[["b_logz_Intercept"]])), qci(dff[["b_CTmaxdev_sexF"]]), qci(dff[["b_CTmaxdev_sexM"]])))
```

This single model has **one** `logz` coefficient and **two** `CTmaxdev`
coefficients — the formula layer does the work, no special-casing.

## 8. Absolute threshold — verified

`threshold = "absolute"` changes just the `mid` line: it subtracts the asymmetry
correction `C(T) = (1/k)·log((up−p)/(p−low))`, folding the up/low/k temperature
effects in so the fitted CTmax/z are the **absolute** (p-survival) quantities.
Drosophila is the right test — its upper asymptote is ≈0.85 (not 1), so the curve
is *not* symmetric about 0.5 and the correction is genuinely non-zero (for
full-range lethal data with low≈0, up≈1 the curve is symmetric and `C(T) → 0`,
which is why relative ≈ absolute there — the 2026-06-16 finding).

```{r abs-fit}
#| message: false
#| warning: false
fit_abs <- brm(proto_4pl_formula(ctmax = ~ 0 + sex, z = ~ 0 + sex,
                                 low = ~ 0 + sex + temp_c:sex, up = ~ 0 + sex + temp_c:sex,
                                 k = ~ 0 + sex + temp_c:sex, threshold = "absolute"),
               data = dat_d, prior = proto_priors(cells = c("sexF", "sexM")),
               chains = 4, iter = 2000, warmup = 1000, cores = 4, seed = 123,
               control = list(adapt_delta = 0.99, max_treedepth = 15), init = 0,
               backend = "cmdstanr", silent = 2, refresh = 0,
               file = file.path(models_dir, "proto_dros_direct_abs"), file_refit = "on_change")
fit_abs$formula$pforms$mid    # the absolute mid line: relative backbone minus C(T)
```

**Internal consistency** — the coef-read absolute z/CTmax equals the
grid-derived absolute LT50 (via `tls(mode = "absolute")`), proving the fitted
backbone *is* the absolute curve and `C(T)` is wired correctly (CTmax in centred
units, `temp_mean = 0`):

```{r abs-consistency}
#| message: false
#| warning: false
da    <- as.data.frame(as_draws_df(fit_abs))
t_abs <- tls(fit_abs, by = "sex", params = c("z", "ctmax"), mode = "absolute",
             t_ref = 1, time_multiplier = 1, temp_mean = 0)$summary
fmt2 <- function(df, q, s) { r <- df[df$quantity == q & df$sex == s, ]
                             sprintf("%.2f [%.2f, %.2f]", r$median, r$lower, r$upper) }
tibble::tibble(
  sex               = c("F", "M"),
  `z coef-read`     = c(qci(exp(da[["b_logz_sexF"]])), qci(exp(da[["b_logz_sexM"]]))),
  `z tls(absolute)` = c(fmt2(t_abs, "z", "F"), fmt2(t_abs, "z", "M")),
  `CTdev coef`      = c(qci(da[["b_CTmaxdev_sexF"]]), qci(da[["b_CTmaxdev_sexM"]])),
  `CTdev tls(abs)`  = c(fmt2(t_abs, "CTmax", "F"), fmt2(t_abs, "CTmax", "M")))
```

They match exactly. And the threshold argument genuinely changes the quantity —
the *same fit's* relative quantities (via `tls(mode = "relative")`) differ from
the absolute ones by the correction:

```{r abs-vs-rel}
#| message: false
#| warning: false
t_rel <- tls(fit_abs, by = "sex", params = "ctmax", mode = "relative",
             t_ref = 1, time_multiplier = 1, temp_mean = 0)$summary
inv_logit <- function(x) 1 / (1 + exp(-x))
upF  <- 0.501 + inv_logit(da[["b_upraw_sexF"]]) * 0.498
lowF <- 0.001 + inv_logit(da[["b_lowraw_sexF"]]) * 0.498
Cf   <- (1 / exp(da[["b_logk_sexF"]])) * log((upF - 0.5) / (0.5 - lowF))
tibble::tibble(
  sex              = c("F", "M"),
  `CTdev absolute` = c(qci(da[["b_CTmaxdev_sexF"]]), qci(da[["b_CTmaxdev_sexM"]])),
  `CTdev relative` = c(fmt2(t_rel, "CTmax", "F"), fmt2(t_rel, "CTmax", "M")))
```

For sex F the fitted up ≈ `r round(median(upF), 2)`, low ≈ `r round(median(lowF), 3)`,
giving a correction `C(T)` = `r qci(Cf)` log₁₀-time units at the mean temperature
— small but non-zero, which is why absolute and relative differ slightly here and
would diverge further as the curve becomes more asymmetric. Absolute is well-posed
only when `low < p < up`; sublethal `up < p` needs a guard (warn / fall back to
relative).

## 9. Helper-function compatibility (audited on the demo)

A reparameterisation is only safe if the downstream helpers survive it. Helpers
fall into two camps: those that read the 4PL via `posterior_linpred(nlpar=)`
(parameterisation-agnostic — and §4 confirmed `mid` is reachable) versus those
that **parse coefficient names** (`b_mid_temp_c`, `b_mid_Intercept`) and assume
single-group Intercept coding — these get `b_CTmaxdev_*`/`b_logz_*` instead and
fail. Audited live on a single-group direct fit (shrimp, CTmax/z ~ 1):

```{r helper-audit}
#| message: false
#| warning: false
fit_sg <- readRDS(file.path(models_dir, "proto_shrimp_direct.rds"))
ss <- standardize_data(shrimp_lethal, temp = "Temperature_assay", duration = "Duration_exposure_hours",
  n_total = "N_individuals_after_trial", mortality = "Mortality_after_trial", duration_unit = "hours")
wf_sg <- structure(list(fit = fit_sg, data = ss, formula = fit_sg$formula, prior = NULL,
  meta = utils::modifyList(attr(ss, "tdt_meta"),
    list(bounds = cb(0, 1), lower = 0, upper = 1, response_type = "count"))), class = "bayes_tls")

st <- function(expr) tryCatch({ v <- suppressMessages(force(expr)); "works" },
                              error = function(e) paste0("ERROR: ", conditionMessage(e)))
et_na <- { e <- tryCatch(suppressMessages(extract_tdt(wf_sg, lethal = TRUE)), error = function(e) NULL)
           if (is.null(e)) "error" else if (is.na(e$z$summary$z_median)) "SILENT NA" else "works" }
trace <- data.frame(time = 0:48, temp = 20 + 12 * sin(seq(0, 6, length.out = 49)))

tibble::tibble(
  helper = c("tls() / tls_z/ctmax/tcrit", "predict_survival_curves()", "derive_tdt_landscape()",
             "diagnose_tdt_fit()", "extract_tdt()", "predict_heat_injury()", "tdt_parameter_table()"),
  status = c(st(tls(wf_sg, lethal = TRUE)), st(predict_survival_curves(wf_sg)),
             st(derive_tdt_landscape(wf_sg)), st(diagnose_tdt_fit(wf_sg)),
             et_na, st(predict_heat_injury(trace, wf_sg, ndraws = 30)),
             st(tdt_parameter_table(wf_sg))),
  reads  = c("posterior_linpred(nlpar=)", "posterior_linpred(transform=TRUE)", "posterior_linpred(mid)",
             "rhat/ess/divergences (param-free)", "b_mid_temp_c / b_mid_Intercept", "b_mid_Intercept/temp_c, b_lowraw_Intercept",
             "b_mid_temp_c / b_mid_Intercept"))
```

**Verdict:**

- **Work unchanged** (verified correct values, not just "no error"): `tls()` and
  `tls_z/ctmax/tcrit`, `predict_survival_curves()`, `derive_tdt_landscape()`,
  `diagnose_tdt_fit()`, and the `get_*` accessors (they only read the *output* of
  the producers below). `tls()` is the linchpin — the general
  `posterior_linpred`-based extractor is fully parameterisation-agnostic.
- **Must change** (3 functions): `extract_tdt()` (**silently returns all NA** —
  the dangerous one), `predict_heat_injury()` (hard error `mid_int not found`),
  `tdt_parameter_table()` (hard error, z column size 0). All three parse
  `b_mid_*` coefficient names and assume single-group Intercept coding.
- **Cosmetic:** add `b_CTmaxdev_*`/`b_logz_*` to `R/bayesTLS-package.R`'s
  `globalVariables()` to silence R CMD check NSE notes.

**The fix is the same for all three, and is an improvement:** route them through
`posterior_linpred(nlpar=)` exactly as `tls()` already does (or have
`extract_tdt()` simply delegate to `tls()`). That removes their hard-wired
coefficient-name parsing — which is *also* why they only ever worked on
single-group midpoint fits today (the grouped case studies already had to use
`tls()`). So fixing them for the direct parameterisation simultaneously
generalises them to grouped/cell-means fits. No helper needs the direct logic
duplicated; they all inherit it from the `posterior_linpred` path.

### Fixing `extract_tdt()` is trivial in direct mode — CTmax/z are coefficients

The cleanest fix is *not* to re-derive a slope: in direct mode CTmax and z **are**
model coefficients — `z = exp(b_logz_<intercept|level>)`,
`CTmax = temp_mean + b_CTmaxdev_<intercept|level>`. `extract_tdt()` reads them off
(via `posterior_linpred(nlpar = "logz"/"CTmaxdev")` so intercept, factor levels,
or covariates are all handled uniformly), then derives T_crit / absolute / LTx
from those same per-draw values through the existing correction layer. The
coefficient read equals `posterior_linpred` *and* `tls()` at the matching
threshold:

```{r extract-direct-read}
#| message: false
#| warning: false
ds <- as.data.frame(as_draws_df(wf_sg$fit)); tbar_sg <- wf_sg$meta$temp_mean
tr <- tls(wf_sg, mode = "relative", t_ref = 1, time_multiplier = 1)$summary
ta <- tls(wf_sg, mode = "absolute", t_ref = 1, time_multiplier = 1)$summary
g  <- function(df, q) sprintf("%.2f", df$median[df$quantity == q])
tibble::tibble(
  quantity               = c("z", "CTmax"),
  `coef read (relative)` = c(qci(exp(ds[["b_logz_Intercept"]])), qci(tbar_sg + ds[["b_CTmaxdev_Intercept"]])),
  `tls(relative)`        = c(g(tr, "z"), g(tr, "CTmax")),
  `tls(absolute)`        = c(g(ta, "z"), g(ta, "CTmax")))
```

The coefficient read matches `tls(relative)` exactly. `tls()`'s default
`mode = "absolute"` differs here (shrimp z 2.31 vs relative 2.44) only because
shrimp's shape varies with temperature, so the two thresholds genuinely differ —
exactly the distinction the `threshold` argument controls.

**Implication — `extract_tdt()` can return *either* threshold from *any* direct
fit.** Through its existing `target_surv` argument:

- the **fitted** threshold is a direct coefficient read (cheap, exact);
- the **opposite** threshold is the conversion via the asymmetry correction
  `C(T)` — i.e. exactly what `tls(mode = <other>)` computes. Both directions are
  validated here: a relative fit → absolute (§9 above, `tls(absolute)`), and an
  absolute fit → relative (§8, `tls(relative)`).

So a relative-threshold fit can still report absolute CTmax/z and vice versa; the
correction layer never disappears, it just runs in whichever direction is asked.
The one caveat: the **fitted** threshold is the quantity forced linear in
temperature (the well-constrained backbone); the converted threshold inherits
whatever bend the temperature-varying asymptotes impose. So fit the threshold you
care most about and treat the converted one as a derived quantity. This is both
simpler than today's slope path and removes the single-group limitation.

## 10. Conclusions → implications for the real package

- The `nlf` formulation **works and samples cleanly** (0 divergences on both
  datasets with the centred asymptote priors), and reproduces the midpoint
  reference (CTmax ~0.05 °C; z matches, CI differences fully explained by RE
  structure).
- **O1 resolved:** `posterior_linpred()` reaches the derived `mid`, so the whole
  `tls()` family, `predict_survival_curves()`, `derive_tdt_landscape()` and
  `diagnose_tdt_fit()` work **unchanged** (§9, verified).
- **Only 3 helpers break** (§9): `extract_tdt()` (silent NA), `predict_heat_injury()`
  and `tdt_parameter_table()` — all because they parse `b_mid_*` coefficient names.
  The fix (route through `posterior_linpred`, or delegate to `tls()`) also lifts
  their existing single-group-only limitation.
- The layered formulas give exactly the requested flexibility (model CTmax and/or
  z by group, with shape effects retained by default).

When we move this into `R/` (see the design note for the file-by-file plan): fold
`proto_4pl_formula` into `make_4pl_formula()` (trigger on the CTmax/z formulas),
`proto_priors` logic into `make_4pl_priors()`, store `threshold`/`log10_tref` in
`meta`, **rewire `extract_tdt()`/`predict_heat_injury()`/`tdt_parameter_table()`
onto the `posterior_linpred` path**, add `b_CTmaxdev_*`/`b_logz_*` to
`globalVariables()`, and add equivalence tests.
```
