---
title: "Direct CTmax/z parameterisation — Tier-3 scripted validation gate"
author: "bayesTLS"
date: 2026-06-22
format:
  html:
    toc: true
    embed-resources: true
    code-fold: false
execute:
  warning: false
  message: false
---

```{r setup}
library(bayesTLS)
library(posterior)
here <- here::here
MODELS <- here("output", "models")
```

## Purpose and scope

This is the **Tier-3 pre-merge gate** for the direct CTmax/z parameterisation
(test plan: `notes/2026-06-22-test-design-direct-ctmax-z.md`, §5). It holds the
invariants that genuinely need *real fits* and so cannot live in the fast or
gated `testthat` tiers:

1. **Grouped ABSOLUTE per-group truth (the inheritance-bias catch).** The only
   test of the rule-2 rationale — that shape sub-parameters inherit the CTmax/z
   group structure (`temp_c:G`) by default — so per-group **absolute** CTmax is
   unbiased when groups differ in curve asymmetry. All relative-threshold checks
   are provably insensitive to this, so it has no `testthat` analogue.
2. **Divergence contrast (centred vs flat asymptote priors).** A qualitative
   reproduction of the funnel that motivated the centred default priors. The
   *reproducible* guard for this is the byte-identical-prior test
   (`tests/testthat/test-direct-priors.R`); the sampling contrast here is
   indicative (its magnitude is sim-/platform-/Stan-version-dependent).

Each assertion is a hard `stopifnot()`, so the **render fails** if an invariant
drifts. Fits are cached to `output/models/gate_*.rds` (delete to force a refit).

**Scoped out (deliberate):** case-study golden numbers (e.g. the embryo/sex
studies) are *not* asserted here while the case studies are being restructured
(`ms/case_studies_new.qmd`) — pinning `z`/CTmax to studies in flux would be
throwaway. Direct↔midpoint equivalence meanwhile is covered by the gated fixture
(`tests/testthat/test-direct-fixture.R`) and the prototype-validation note
(`notes/2026-06-22-direct-ctmax-z-prototype-validation.qmd`). Re-point a
case-study golden section here once the new studies settle.

```{r helpers}
# Grouped + per-group-asymmetry simulator. Dose axis matches simulate_tdt():
# logd = log10(t_hours * 60); standardise with duration_unit = "hours".
simulate_grouped <- function(truth, temps = c(30, 33, 36, 39),
                             durations_hours = c(0.1, 0.5, 2, 10),
                             n_rep = 6, n_per_rep = 30, phi = 8, T_bar = 34,
                             sparse_hot = FALSE, seed = 1) {
  set.seed(seed)
  g <- expand.grid(grp = names(truth), T = temps, t_hours = durations_hours,
                   rep = seq_len(n_rep), stringsAsFactors = FALSE)
  if (sparse_hot) g <- g[!(g$T == max(temps) & g$rep > 1), ]   # weakly identified hot asymptote
  g$ld <- log10(g$t_hours * 60)
  g$p  <- vapply(seq_len(nrow(g)), function(i) {
    tt <- truth[[g$grp[i]]]; mid <- tt$a0 - (1 / tt$z) * (g$T[i] - T_bar)
    tt$ell + (tt$u - tt$ell) / (1 + exp(tt$k * (g$ld[i] - mid)))
  }, numeric(1))
  g$n_trials <- n_per_rep
  g$y_alive  <- rbinom(nrow(g), g$n_trials, rbeta(nrow(g), g$p * phi, (1 - g$p) * phi))
  g$grp <- factor(g$grp)
  attr(g, "T_bar") <- T_bar
  g
}

# Per-group analytic truth (relative z and absolute p=0.5 CTmax at 1 h), the
# per-group form of tests/testthat truth_summary().
group_truth <- function(truth, T_bar = 34) {
  do.call(rbind, lapply(names(truth), function(g) {
    tt <- truth[[g]]
    alpha  <- tt$a0 + (1 / tt$k) * log((tt$u - 0.5) / (0.5 - tt$ell)) - (-1 / tt$z) * T_bar
    data.frame(grp = g, z = tt$z, CTmax_abs_1h = (log10(60) - alpha) / (-1 / tt$z))
  }))
}

divergences <- function(wf)
  sum(brms::nuts_params(get_brmsfit(wf), pars = "divergent__")$Value)

CTRL <- list(adapt_delta = 0.95, max_treedepth = 12)
```

## 1. Grouped absolute per-group truth — the inheritance-bias catch

Two groups with **deliberately different asymmetry**: `A` near-symmetric, `B`
strongly asymmetric (high lower asymptote, low upper asymptote). Their curve
asymmetry correction `C(T)` differs, so their **absolute** (p = 0.5) CTmax differ
from their relative midpoints by different amounts. A fit whose shape inherits
the group structure can represent both; a fit forced to share one shape across
groups cannot.

```{r inherit-sim}
truth_inh <- list(
  A = list(ell = 0.03, u = 0.97, k = 6, z = 3.0, a0 = 1.4),   # near-symmetric
  B = list(ell = 0.25, u = 0.78, k = 3, z = 2.0, a0 = 1.1)    # strongly asymmetric
)
raw_inh <- simulate_grouped(truth_inh, seed = 20260622)
T_bar   <- attr(raw_inh, "T_bar")
std_inh <- standardize_data(raw_inh, temp = "T", duration = "t_hours",
                            n_total = "n_trials", n_surv = "y_alive",
                            duration_unit = "hours")
gt <- group_truth(truth_inh, T_bar)
gt
```

```{r inherit-fits}
# INHERITED (default): shape inherits 0 + grp + temp_c:grp -> per-group asymmetry.
fit_inh <- fit_4pl(std_inh, ctmax = ~ 0 + grp, z = ~ 0 + grp, threshold = "absolute",
                   chains = 2, iter = 1200, warmup = 600, cores = 2, seed = 1,
                   control = CTRL, refresh = 0,
                   file = file.path(MODELS, "gate_inherit_inh"))
# SHARED: force one pooled shape for both groups.
fit_shr <- fit_4pl(std_inh, ctmax = ~ 0 + grp, z = ~ 0 + grp,
                   up = ~ temp_c, low = ~ temp_c, k = ~ temp_c, threshold = "absolute",
                   chains = 2, iter = 1200, warmup = 600, cores = 2, seed = 1,
                   control = CTRL, refresh = 0,
                   file = file.path(MODELS, "gate_inherit_shr"))
```

```{r inherit-check}
ti <- tls(fit_inh, by = "grp", mode = "absolute", params = "ctmax",
          temp_mean = T_bar, seed = 1)$summary
ts <- tls(fit_shr, by = "grp", mode = "absolute", params = "ctmax",
          temp_mean = T_bar, seed = 1)$summary

cmp <- merge(merge(gt, ti[, c("grp", "median")], by = "grp"),
             ts[, c("grp", "median")], by = "grp", suffixes = c(".inh", ".shr"))
cmp$err_inh    <- cmp$median.inh - cmp$CTmax_abs_1h
cmp$err_shr    <- cmp$median.shr - cmp$CTmax_abs_1h
cmp$inh_vs_shr <- cmp$median.inh - cmp$median.shr
knitr::kable(cmp, digits = 3,
             caption = "Per-group absolute CTmax (1 h): inherited recovers truth; shared is biased for the asymmetric group.")

B <- cmp[cmp$grp == "B", ]
stopifnot(
  # (a) inherited recovers per-group truth within the ~1 C small-fit tolerance
  all(abs(cmp$err_inh) < 1.0),
  # (b) for the strongly asymmetric group, the shared-shape fit is materially biased,
  #     and inherited differs from shared (the inheritance default matters)
  abs(B$inh_vs_shr) > 0.5,
  abs(B$err_shr) > abs(B$err_inh)
)
cat("PASS §1: inherited recovers per-group absolute CTmax;",
    sprintf("shared biases group B by %.2f °C (inherited err %.2f).\n",
            B$err_shr, B$err_inh))
```

## 2. Divergence contrast — centred vs flat asymptote priors

A funnel-prone grouped sim (sparse data at the hottest temperature, so the upper
asymptote is weakly identified there). With **flat** asymptote priors the raw
asymptote parameters wander (the funnel) and the sampler diverges; the **centred**
default priors tame it. Magnitude is indicative only — the reproducible guard is
the byte-identical-prior test in `test-direct-priors.R`.

```{r diverge}
truth_fun <- list(A = list(ell = 0.03, u = 0.97, k = 6, z = 3.0, a0 = 1.4),
                  B = list(ell = 0.05, u = 0.95, k = 5, z = 2.3, a0 = 1.2))
std_fun <- standardize_data(
  simulate_grouped(truth_fun, sparse_hot = TRUE, seed = 20260622),
  temp = "T", duration = "t_hours", n_total = "n_trials", n_surv = "y_alive",
  duration_unit = "hours")

fit_centred <- fit_4pl(std_fun, ctmax = ~ 0 + grp, z = ~ 0 + grp,
                       chains = 2, iter = 1000, warmup = 500, cores = 2, seed = 1,
                       control = list(adapt_delta = 0.9, max_treedepth = 11),
                       refresh = 0, file = file.path(MODELS, "gate_funnel_centred"))

pr_flat <- make_4pl_priors(std_fun, ctmax = ~ 0 + grp, z = ~ 0 + grp)
pr_flat$prior[pr_flat$nlpar %in% c("lowraw", "upraw") & pr_flat$class == "b"] <-
  "normal(0, 10)"
fit_flat <- fit_4pl(std_fun, ctmax = ~ 0 + grp, z = ~ 0 + grp, prior = pr_flat,
                    chains = 2, iter = 1000, warmup = 500, cores = 2, seed = 1,
                    control = list(adapt_delta = 0.9, max_treedepth = 11),
                    refresh = 0, file = file.path(MODELS, "gate_funnel_flat"))

dv_c <- divergences(fit_centred); dv_f <- divergences(fit_flat)
cat(sprintf("divergent transitions: centred = %d, flat = %d\n", dv_c, dv_f))
stopifnot(dv_f > dv_c)   # qualitative: flat asymptote priors induce more funnel divergences
cat("PASS §2: flat asymptote priors induce more divergences than the centred default.\n")
```

## 3. Where the rest of the validation lives

| Invariant | Where |
|---|---|
| Silent-NA / C(T)-oracle / threshold cells / unit aliases / seed determinism / print | fast `testthat` (`test-direct-extract.R`, `test-direct-priors.R`, `test-bayes_tls_methods.R`) — every CI run |
| Truth-recovery, direct↔midpoint equivalence, grouped-redirect guards, `derive_*`/`predict_heat_injury` wiring | gated `testthat` (`test-direct-fixture.R`, `RUN_BRMS_TESTS=true`) |
| Centred-prior funnel guard (reproducible) | `test-direct-priors.R` byte-identical asymptote-prior snapshot |
| Grouped absolute per-group truth; divergence contrast (real fits) | **this gate** |

Run before merge with `quarto render notes/2026-06-22-direct-ctmax-z-validation-gate.qmd`;
a non-zero exit (failed `stopifnot`) blocks the merge.
