---
title: "bayesian_TDT_shrimp"
format: html
editor: visual
---

```{r}
pacman::p_load(
  tidyverse,
  ggdist,
  broom,
  MASS,
  patchwork,
  readxl,
  lme4,
  brms,
  drc,
  mgcv,
  medrc, #remotes::install_github("DoseResponse/medrc")
  plotly,
  viridisLite,
  posterior,
  scales,
  glmmTMB,
  MuMIn,
  survminer,
  survival,
  flextable,
  performance,
  future,
  future.apply)
```



# **Lethal TDT**

## **Bayesian 4PL model**
```{r}
base_theme <- theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 24, face = "bold"),
    axis.text  = element_text(size = 20),
    legend.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),
    axis.line = element_line(linewidth = 1),
    axis.ticks = element_line(linewidth = 1),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 2)
  )

ybreaks_left  <- c(1, 60, 120, 240, 480, 960)
ybreaks_right <- c(1, 10, 30, 60, 120, 240, 480)
lab_number <- function(x) scales::number(x, big.mark = "", accuracy = 1)


# =========================
# 1) Load + process raw data
# =========================
d <- read_excel("data/Data_Shrimp_TDT_CTmax_experiment.xlsx",
                sheet = "LETHAL_TDT")

# Transform data to get counts of individuals that survived/died
d <- d %>% mutate(survival = 1 - Mortality_after_trial,
                   n_total = N_individuals_after_trial, 
                   n_surv  = round(survival * n_total),
                   n_dead  = n_total - n_surv,
                   assay_temp = Temperature_assay,
                   duration = Duration_exposure_hours,
                   logd = log10(duration),
                   date_experiment = as.factor(Date),
                   tank_ID = as.factor(Tank),
                   body_mass = Total_Sample_body_mass_g,
                   start_min = as.numeric(format(Starting_time, "%H")) * 60 +
                               as.numeric(format(Starting_time, "%M")) +
                               as.numeric(format(Starting_time, "%S")) / 60)

d$obs <- factor(seq_len(nrow(d)))

temp_mean <- mean(d$assay_temp, na.rm = TRUE)
start_min_mean <- mean(d$start_min, na.rm = TRUE)

d <- d %>%
  mutate(
    temp_c = assay_temp - temp_mean,
    start_min_c = start_min - start_min_mean,
    body_mass_c = body_mass - mean(body_mass) # center
  )

# =========================
# 2) Numeric constants for priors / init
# =========================
mid_start <- median(d$logd, na.rm = TRUE)
lraw_mean <- qlogis(0.02 / 0.49)
uraw_mean <- qlogis((0.98 - 0.51) / 0.49)
logk_mean <- log(2)

num_to_string <- function(x) {
  formatC(x, digits = 16, format = "fg", flag = "#")
}

mid_start_s <- num_to_string(mid_start)
lraw_mean_s <- num_to_string(lraw_mean)
uraw_mean_s <- num_to_string(uraw_mean)
logk_mean_s <- num_to_string(logk_mean)

# =========================
# 3) 4-parameter logistic model over duration
# =========================
# Probability-scale curve:
#
# p = low + (up - low) / (1 + exp(exp(logk) * (logd - mid)))
#
# low = inv_logit(lraw) * 0.49
# up  = 0.51 + inv_logit(uraw) * 0.49
#
# mid varies with temperature and life stage:
# mid ~ temp_c * life_stage + (1 | date_experiment)
#
# z is an obs-level random effect

f_tdt_4pl <- bf(
  n_surv | trials(n_total) ~ eta,
  nlf(
    eta ~ logit(
      (inv_logit(lraw) * 0.49) +
        (
          (0.51 + inv_logit(uraw) * 0.49) -
          (inv_logit(lraw) * 0.49)
        ) / (1 + exp(exp(logk) * (logd - mid)))
    ) + z
  ),
  lf(lraw ~ 1),
  lf(uraw ~ 1),
  lf(logk ~ 1),
  lf(mid ~ temp_c + (1 | date_experiment) + (1 | tank_ID)),
  lf(z ~ 0 + (1 | obs)),
  nl = TRUE
)

# =========================
# 4) Priors
# =========================
priors_tdt_4pl <- c(
  set_prior(
    paste0("normal(", lraw_mean_s, ", 1)"),
    class = "b", nlpar = "lraw"
  ),
  set_prior(
    paste0("normal(", uraw_mean_s, ", 1)"),
    class = "b", nlpar = "uraw"
  ),
  set_prior(
    paste0("normal(", logk_mean_s, ", 1)"),
    class = "b", nlpar = "logk"
  ),
  set_prior(
    paste0("normal(", mid_start_s, ", 1.5)"),
    class = "b", nlpar = "mid", coef = "Intercept"
  ),
  set_prior(
    "normal(0, 0.6)",
    class = "b", nlpar = "mid", coef = "temp_c"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "date_experiment"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "tank_ID"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "z", group = "obs"
  )
)

# =========================
# 5) Initial values
# =========================
init_tdt_4pl <- function() {
  list(
    b_lraw_Intercept = lraw_mean,
    b_uraw_Intercept = uraw_mean,
    b_logk_Intercept = logk_mean,
    b_mid_Intercept = mid_start,
    b_mid_temp_c = 0
  )
}


# =========================
# 6) Fit model
# =========================
fit_4pl <- brm(
  formula = f_tdt_4pl,
  data = d,
  family = binomial(link = "logit"),
  prior = priors_tdt_4pl,
  save_pars = save_pars(all = TRUE),
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = init_tdt_4pl,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)

saveRDS(fit_4pl, file = "Rdata/model_lethal_TDT.RDS")


print(fit_4pl)
r2(fit_4pl)

z_draws <- -1 / dr$b_mid_temp_c

quantile(z_draws, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
mean(z_draws, na.rm = TRUE) # z = 2.55 

# =========================
# 8) Survival curves
# =========================
temps_plot <- seq(30, 33, by = 0.5)

dur_min <- min(d$duration, na.rm = TRUE)
dur_max <- max(d$duration, na.rm = TRUE)
dur_grid <- exp(seq(log(dur_min), log(dur_max), length.out = 250))

nd_surv <- expand.grid(
  assay_temp = temps_plot,
  duration = dur_grid,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

nd_surv <- nd_surv %>%
  mutate(
    logd = log10(duration),
    temp_c = assay_temp - temp_mean,
    n_total = as.integer(round(median(d$n_total, na.rm = TRUE))),
    date_experiment = levels(d$date_experiment)[1],
    obs = levels(d$obs)[1],
    temp_f = factor(assay_temp, levels = sort(unique(assay_temp)))
  )

pr_surv <- posterior_linpred(
  fit_4pl,
  newdata = nd_surv,
  re_formula = NA,
  transform = TRUE
)

nd_surv$pred_mean <- apply(pr_surv, 2, median, na.rm = TRUE)
nd_surv$pred_lo   <- apply(pr_surv, 2, quantile, probs = 0.025, na.rm = TRUE)
nd_surv$pred_hi   <- apply(pr_surv, 2, quantile, probs = 0.975, na.rm = TRUE)


d_plot <- d %>% 
    mutate(
    temp_f = factor(assay_temp, levels = sort(unique(nd_surv$assay_temp))))

p_surv_curves <- ggplot(
  nd_surv,
  aes(x = duration, y = pred_mean, colour = temp_f, fill = temp_f, group = temp_f)
) +
  geom_ribbon(
    aes(ymin = pred_lo, ymax = pred_hi),
    alpha = 0.15,
    colour = NA,
    show.legend = FALSE
  ) +
  geom_line(linewidth = 1.1) +
  geom_jitter(
    data = d_plot,
    aes(x = duration, y = survival, fill = temp_f),
    width = 0.01, 
    inherit.aes = FALSE,
    size = 3,
    alpha = 0.9,
    shape = 21, 
    stroke = 0.5,
    colour = "white"
  ) +
  scale_colour_viridis_d(option = "plasma", direction = 1, begin = 0.05, end = 0.95) +
  scale_fill_viridis_d(option = "plasma", direction = 1, begin = 0.05, end = 0.95) +
labs(
  x = "Exposure duration (hours)",
  y = "Predicted survival",
  colour = NULL
) +
base_theme +
  coord_cartesian(ylim = c(0, 1.05)) +
theme(
  legend.position = c(0.95, 0.98),   
  legend.justification = c(1, 1),
  legend.direction = "horizontal",
  legend.title = element_blank(),
  legend.text = element_text(size = 12),
  legend.background = element_blank()
) +
guides(
  colour = guide_legend(nrow = 1),
  fill = "none"
)

print(p_surv_curves)

ggsave(file = "fig/survival_curves.png", dpi = 500, width = 9, height = 7)


# =========================
# 4) Posterior draws helper functions
#    for 4PL TDT model
# =========================
dr <- posterior::as_draws_df(fit_4pl)

get_par <- function(nm) {
  if (nm %in% names(dr)) {
    dr[[nm]]
  } else {
    rep(0, nrow(dr))
  }
}

get_4pl_draws <- function(temp_c_value) {
  data.frame(
    .draw = dr$.draw,
    temp_c = temp_c_value,
    low = plogis(get_par("b_lraw_Intercept")) * 0.49,
    up = 0.51 + plogis(get_par("b_uraw_Intercept")) * 0.49,
    k = exp(get_par("b_logk_Intercept")),
    mid = get_par("b_mid_Intercept") +
      get_par("b_mid_temp_c") * temp_c_value
  )
}

# =========================
# 5) TDT curve
# =========================
temps_df <- data.frame(
  assay_temp = seq(28, 36.1, by = 0.01)
) %>%
  mutate(
    temp_c = assay_temp - temp_mean
  )

tdt_draws_avg <- bind_rows(
  lapply(seq_len(nrow(temps_df)), function(i) {
    out <- get_4pl_draws(
      temp_c_value = temps_df$temp_c[i]
    )
    out$assay_temp <- temps_df$assay_temp[i]
    out
  })
) %>%
  mutate(
    log10_t_h = mid + log((up - 0.5) / (0.5 - low)) / k,
    t50_h = 10^log10_t_h,
    t50_min = 60 * t50_h
  ) %>%
  filter(
    is.finite(t50_min),
    t50_min > 0
  )

pred_df_tdt_avg <- tdt_draws_avg %>%
  group_by(assay_temp) %>%
  summarise(
    time_to_50 = median(t50_min, na.rm = TRUE),
    lower = quantile(t50_min, 0.025, na.rm = TRUE),
    upper = quantile(t50_min, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

p_tdt_avg_lin <- ggplot(pred_df_tdt_avg) +
  geom_ribbon(
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "grey70"
  ) +
  geom_line(
    aes(x = assay_temp, y = time_to_50),
    linewidth = 1.6
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_continuous(breaks = ybreaks_left, labels = lab_number) +
  coord_cartesian(ylim = c(0.9, 960)) +
  labs(
    x = "Assay temperature (°C)",
    y = "Time to 50% survival (min)"
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

p_tdt_avg_log <- ggplot(pred_df_tdt_avg) +
  geom_ribbon(
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "grey70"
  ) +
  geom_line(
    aes(x = assay_temp, y = time_to_50),
    linewidth = 1.6
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(breaks = ybreaks_right, labels = lab_number) +
  coord_cartesian(ylim = c(0.9, 960)) +
  labs(
    x = "Assay temperature (°C)",
    y = "Time to 50% mortality (min)",
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

print(p_tdt_avg_lin + p_tdt_avg_log + patchwork::plot_layout(ncol = 2))

ggsave(
  file = "fig/fig_tdt_lethal.png",
  dpi = 500,
  height = 6,
  width = 13
)

# =========================
# 6) Predicted survival for 1-hour exposure
#    from the 4PL model
# =========================
temps_1h <- seq(29, 34, by = 0.05)
dur_1h <- 1

nd_1h <- expand.grid(
  assay_temp = temps_1h,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

nd_1h <- nd_1h %>%
  mutate(
    duration = dur_1h,
    logd = log10(duration),
    temp_c = assay_temp - temp_mean,
    n_total = as.integer(round(median(d$n_total, na.rm = TRUE))),
    date_experiment = factor(levels(d$date_experiment)[1], levels = levels(d$date_experiment)),
    obs = factor(levels(d$obs)[1], levels = levels(d$obs))
  )

pr_1h <- posterior_linpred(
  fit_4pl,
  newdata = nd_1h,
  re_formula = NA,
  transform = TRUE
)

nd_1h$pred_mean <- apply(pr_1h, 2, median, na.rm = TRUE)
nd_1h$pred_lo   <- apply(pr_1h, 2, quantile, probs = 0.025, na.rm = TRUE)
nd_1h$pred_hi   <- apply(pr_1h, 2, quantile, probs = 0.975, na.rm = TRUE)

p_1h <- ggplot(
  nd_1h,
  aes(x = assay_temp, y = pred_mean)
) +
  geom_ribbon(
    aes(ymin = pred_lo, ymax = pred_hi),
    alpha = 0.20,
    fill = "grey70"
  ) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Assay temperature (°C)",
    y = "Predicted survival (1-hour exposure)"
  ) +
  base_theme

print(p_1h)

# =========================
# 7) Temperature causing 50% mortality after 1 hour
#    numerical extraction from posterior predictions
# =========================
get_temp50_draws <- function(pred_mat, temp_vec) {
  pred_stage <- pred_mat
  temp_stage <- temp_vec

  ord <- order(temp_stage)
  pred_stage <- pred_stage[, ord, drop = FALSE]
  temp_stage <- temp_stage[ord]

  res_list <- vector("list", nrow(pred_stage))

  for (i in seq_len(nrow(pred_stage))) {
    y <- as.numeric(pred_stage[i, ])
    sgn <- y - 0.5

    if (all(is.na(y)) || min(y, na.rm = TRUE) > 0.5 || max(y, na.rm = TRUE) < 0.5) {
      res_list[[i]] <- data.frame(
        .draw = i,
        assay_temp_50_1h = NA_real_
      )
      next
    }

    idx <- which(diff(sign(sgn)) != 0)

    if (length(idx) == 0) {
      j <- which.min(abs(sgn))
      temp_50 <- temp_stage[j]
    } else {
      j <- idx[1]
      x1 <- temp_stage[j]
      x2 <- temp_stage[j + 1]
      y1 <- y[j]
      y2 <- y[j + 1]

      if (abs(y2 - y1) < 1e-12) {
        temp_50 <- mean(c(x1, x2))
      } else {
        temp_50 <- x1 + (0.5 - y1) * (x2 - x1) / (y2 - y1)
      }
    }

    res_list[[i]] <- data.frame(
      .draw = i,
      assay_temp_50_1h = temp_50
    )
  }

  bind_rows(res_list)
}

temp_1h_draws <- get_temp50_draws(
  pred_mat = pr_1h,
  temp_vec = nd_1h$assay_temp
) %>%
  filter(is.finite(assay_temp_50_1h))

temp_1h_summary <- temp_1h_draws %>%
  summarise(
    mean_temp = mean(assay_temp_50_1h, na.rm = TRUE),
    median_temp = median(assay_temp_50_1h, na.rm = TRUE),
    lower = quantile(assay_temp_50_1h, 0.025, na.rm = TRUE),
    upper = quantile(assay_temp_50_1h, 0.975, na.rm = TRUE)
  )

temp_1h_summary


# =========================
# 9) 2D landscape plot
# =========================
temps_surface <- seq(
  min(d$assay_temp, na.rm = TRUE),
  max(d$assay_temp, na.rm = TRUE),
  length.out = 120
)

dur_surface <- seq(
  min(d$duration, na.rm = TRUE),
  max(d$duration, na.rm = TRUE),
  length.out = 120
)

make_heatmap_df <- function() {
  nds <- expand.grid(
    assay_temp = temps_surface,
    duration = dur_surface,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )

  nds <- nds %>%
    mutate(
      logd = log10(duration),
      temp_c = assay_temp - temp_mean,
      n_total = as.integer(round(median(d$n_total, na.rm = TRUE))),
      date_experiment = factor(levels(d$date_experiment)[1], levels = levels(d$date_experiment)),
      obs = factor(levels(d$obs)[1], levels = levels(d$obs))
    )

  pr_s <- posterior_linpred(
    fit_4pl,
    newdata = nds,
    re_formula = NA,
    transform = TRUE
  )

  nds$pred_mean <- apply(pr_s, 2, median, na.rm = TRUE)

  nds
}

heat_df <- make_heatmap_df()

d_points <- d %>%
  group_by(assay_temp, duration) %>%
  summarise(
    survival_mean = mean(survival, na.rm = TRUE),
    .groups = "drop"
  )

p_landscape_2d <- ggplot(
  heat_df,
  aes(x = assay_temp, y = duration, fill = pred_mean)
) +
  geom_raster(interpolate = TRUE) +
  geom_point(
    data = d_points,
    aes(x = assay_temp, y = duration, fill = survival_mean),
    inherit.aes = FALSE,
    size = 3.2,
    alpha = 0.9,
    shape = 21,
    color = "white",
    stroke = 0.1,
  ) +
  geom_contour(
    aes(z = pred_mean),
    breaks = c(0.25, 0.5, 0.75),
    colour = "white",
    alpha = 0.55,
    linewidth = 0.5
  ) +
metR::geom_text_contour(
    aes(
      z = pred_mean,
      label = after_stat(level)
    ),
    breaks = c(0.25, 0.5, 0.75),
    colour = "white",
    size = 3,
    stroke = 0,
    nudge_y = 0.05,
    nudge_x = 0.1,
    label.placer = metR::label_placer_n(2)
  ) +
  scale_fill_viridis_c(
    limits = c(0, 1),
    option = "inferno",
    direction = 1
  ) +
  labs(
    x = "Assay temperature (°C)",
    y = "Exposure duration (h)",
    fill = "Predicted survival"
  ) +
  base_theme +
  theme(
    legend.position = "right"
  )

print(p_landscape_2d)

ggsave(
  file = "fig/thermal_landscape.png",
  dpi = 500,
  height = 6,
  width = 9
)

```

## **Bayesian 4PL model with interaction**

```{r}
f_tdt_4pl_int <- bf(
  n_surv | trials(n_total) ~ eta,
  nlf(
    eta ~ logit(
      (inv_logit(lraw) * 0.49) +
        (
          (0.51 + inv_logit(uraw) * 0.49) -
          (inv_logit(lraw) * 0.49)
        ) / (1 + exp(exp(logk) * (logd - mid) + int * temp_c * logd))
    ) + z
  ),
  lf(lraw ~ 1),
  lf(uraw ~ 1),
  lf(logk ~ 1),
  lf(mid ~ temp_c + (1 | date_experiment) + (1 | tank_ID)),
  lf(int ~ 1),
  lf(z ~ 0 + (1 | obs)),
  nl = TRUE
)

priors_tdt_4pl_int <- c(
  set_prior(
    paste0("normal(", lraw_mean_s, ", 1)"),
    class = "b", nlpar = "lraw"
  ),
  set_prior(
    paste0("normal(", uraw_mean_s, ", 1)"),
    class = "b", nlpar = "uraw"
  ),
  set_prior(
    paste0("normal(", logk_mean_s, ", 1)"),
    class = "b", nlpar = "logk"
  ),
  set_prior(
    paste0("normal(", mid_start_s, ", 1.5)"),
    class = "b", nlpar = "mid", coef = "Intercept"
  ),
  set_prior(
    "normal(0, 0.6)",
    class = "b", nlpar = "mid", coef = "temp_c"
  ),
  set_prior(
    "normal(0, 0.6)",
    class = "b", nlpar = "int"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "date_experiment"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "tank_ID"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "z", group = "obs"
  )
)

init_tdt_4pl_int <- function() {
  list(
    b_lraw_Intercept = lraw_mean,
    b_uraw_Intercept = uraw_mean,
    b_logk_Intercept = logk_mean,
    b_mid_Intercept = mid_start,
    b_mid_temp_c = 0,
    b_int_Intercept = 0
  )
}

fit_4pl_int <- brm(
  formula = f_tdt_4pl_int,
  data = d,
  family = binomial(link = "logit"),
  prior = priors_tdt_4pl_int,
  save_pars = save_pars(all = TRUE),
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = init_tdt_4pl_int,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)

saveRDS(fit_4pl_int, file = "Rdata/model_lethal_TDT_interaction.RDS")

print(fit_4pl_int)
r2(fit_4pl_int)

################
loo_4pl <- loo(fit_4pl)
loo_4pl_int <- loo(fit_4pl_int)

loo_compare(loo_4pl, loo_4pl_int)

###################


dr <- posterior::as_draws_df(fit_4pl_int)

z_draws <- -1 / dr$b_mid_temp_c

quantile(z_draws, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
mean(z_draws, na.rm = TRUE)

# =========================
# 8) Survival curves
# =========================
temps_plot <- seq(30, 33, by = 0.5)

dur_min <- min(d$duration, na.rm = TRUE)
dur_max <- max(d$duration, na.rm = TRUE)
dur_grid <- exp(seq(log(dur_min), log(dur_max), length.out = 250))

nd_surv_int <- expand.grid(
  assay_temp = temps_plot,
  duration = dur_grid,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

nd_surv_int <- nd_surv_int %>%
  mutate(
    logd = log10(duration),
    temp_c = assay_temp - temp_mean,
    n_total = as.integer(round(median(d$n_total, na.rm = TRUE))),
    date_experiment = levels(d$date_experiment)[1],
    obs = levels(d$obs)[1],
    temp_f = factor(assay_temp, levels = sort(unique(assay_temp)))
  )

pr_surv_int <- posterior_linpred(
  fit_4pl_int,
  newdata = nd_surv_int,
  re_formula = NA,
  transform = TRUE
)

nd_surv_int$pred_mean <- apply(pr_surv_int, 2, median, na.rm = TRUE)
nd_surv_int$pred_lo   <- apply(pr_surv_int, 2, quantile, probs = 0.025, na.rm = TRUE)
nd_surv_int$pred_hi   <- apply(pr_surv_int, 2, quantile, probs = 0.975, na.rm = TRUE)

d_plot_int <- d %>% 
  mutate(
    temp_f = factor(assay_temp, levels = sort(unique(nd_surv_int$assay_temp)))
  )

p_surv_curves_int <- ggplot(
  nd_surv_int,
  aes(x = duration, y = pred_mean, colour = temp_f, fill = temp_f, group = temp_f)
) +
  geom_ribbon(
    aes(ymin = pred_lo, ymax = pred_hi),
    alpha = 0.15,
    colour = NA,
    show.legend = FALSE
  ) +
  geom_line(linewidth = 1.1) +
  geom_jitter(
    data = d_plot_int,
    aes(x = duration, y = survival, fill = temp_f),
    width = 0.01, 
    inherit.aes = FALSE,
    size = 3,
    alpha = 0.9,
    shape = 21, 
    stroke = 0.5,
    colour = "white"
  ) +
  scale_colour_viridis_d(option = "plasma", direction = 1, begin = 0.05, end = 0.95) +
  scale_fill_viridis_d(option = "plasma", direction = 1, begin = 0.05, end = 0.95) +
  labs(
    x = "Exposure duration (hours)",
    y = "Predicted survival",
    colour = NULL
  ) +
  base_theme +
  coord_cartesian(ylim = c(0, 1.05)) +
  theme(
    legend.position = c(0.95, 0.98),   
    legend.justification = c(1, 1),
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.background = element_blank()
  ) +
  guides(
    colour = guide_legend(nrow = 1),
    fill = "none"
  )

print(p_surv_curves_int)

ggsave(file = "fig/survival_curves_int.png", dpi = 500, width = 9, height = 7)


# =========================
# 4) Posterior draws helper functions
#    for 4PL TDT model
# =========================
get_par_int <- function(nm) {
  if (nm %in% names(dr)) {
    dr[[nm]]
  } else {
    rep(0, nrow(dr))
  }
}

get_4pl_draws_int <- function(temp_c_value) {
  data.frame(
    .draw = dr$.draw,
    temp_c = temp_c_value,
    low = plogis(get_par_int("b_lraw_Intercept")) * 0.49,
    up = 0.51 + plogis(get_par_int("b_uraw_Intercept")) * 0.49,
    k = exp(get_par_int("b_logk_Intercept")),
    mid = get_par_int("b_mid_Intercept") +
      get_par_int("b_mid_temp_c") * temp_c_value,
    int = get_par_int("b_int_Intercept")
  )
}

# =========================
# 5) TDT curve
# =========================
temps_df_int <- data.frame(
  assay_temp = seq(28, 36.1, by = 0.01)
) %>%
  mutate(
    temp_c = assay_temp - temp_mean
  )

tdt_draws_avg_int <- bind_rows(
  lapply(seq_len(nrow(temps_df_int)), function(i) {
    out <- get_4pl_draws_int(
      temp_c_value = temps_df_int$temp_c[i]
    )
    out$assay_temp <- temps_df_int$assay_temp[i]
    out
  })
) %>%
  mutate(
    rhs = log((up - 0.5) / (0.5 - low)),
    log10_t_h = (k * mid + rhs) / (k + int * temp_c),
    t50_h = 10^log10_t_h,
    t50_min = 60 * t50_h
  ) %>%
  filter(
    is.finite(t50_min),
    t50_min > 0
  )

pred_df_tdt_avg_int <- tdt_draws_avg_int %>%
  group_by(assay_temp) %>%
  summarise(
    time_to_50 = median(t50_min, na.rm = TRUE),
    lower = quantile(t50_min, 0.025, na.rm = TRUE),
    upper = quantile(t50_min, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

p_tdt_avg_lin_int <- ggplot(pred_df_tdt_avg_int) +
  geom_ribbon(
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "grey70"
  ) +
  geom_line(
    aes(x = assay_temp, y = time_to_50),
    linewidth = 1.6
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_continuous(breaks = ybreaks_left, labels = lab_number) +
  coord_cartesian(ylim = c(0.9, 960)) +
  labs(
    x = "Assay temperature (°C)",
    y = "Time to 50% survival (min)"
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

p_tdt_avg_log_int <- ggplot(pred_df_tdt_avg_int) +
  geom_ribbon(
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "grey70"
  ) +
  geom_line(
    aes(x = assay_temp, y = time_to_50),
    linewidth = 1.6
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(breaks = ybreaks_right, labels = lab_number) +
  coord_cartesian(ylim = c(0.9, 960)) +
  labs(
    x = "Assay temperature (°C)",
    y = "Time to 50% mortality (min)"
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

print(p_tdt_avg_lin_int + p_tdt_avg_log_int + patchwork::plot_layout(ncol = 2))

ggsave(
  file = "fig/fig_tdt_lethal_int.png",
  dpi = 500,
  height = 6,
  width = 13
)

# =========================
# 6) Predicted survival for 1-hour exposure
#    from the 4PL model
# =========================
temps_1h <- seq(29, 34, by = 0.05)
dur_1h <- 1

nd_1h_int <- expand.grid(
  assay_temp = temps_1h,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

nd_1h_int <- nd_1h_int %>%
  mutate(
    duration = dur_1h,
    logd = log10(duration),
    temp_c = assay_temp - temp_mean,
    n_total = as.integer(round(median(d$n_total, na.rm = TRUE))),
    date_experiment = factor(levels(d$date_experiment)[1], levels = levels(d$date_experiment)),
    obs = factor(levels(d$obs)[1], levels = levels(d$obs))
  )

pr_1h_int <- posterior_linpred(
  fit_4pl_int,
  newdata = nd_1h_int,
  re_formula = NA,
  transform = TRUE
)

nd_1h_int$pred_mean <- apply(pr_1h_int, 2, median, na.rm = TRUE)
nd_1h_int$pred_lo   <- apply(pr_1h_int, 2, quantile, probs = 0.025, na.rm = TRUE)
nd_1h_int$pred_hi   <- apply(pr_1h_int, 2, quantile, probs = 0.975, na.rm = TRUE)

p_1h_int <- ggplot(
  nd_1h_int,
  aes(x = assay_temp, y = pred_mean)
) +
  geom_ribbon(
    aes(ymin = pred_lo, ymax = pred_hi),
    alpha = 0.20,
    fill = "grey70"
  ) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Assay temperature (°C)",
    y = "Predicted survival (1-hour exposure)"
  ) +
  base_theme

print(p_1h_int)

# =========================
# 7) Temperature causing 50% mortality after 1 hour
#    numerical extraction from posterior predictions
# =========================
get_temp50_draws_int <- function(pred_mat, temp_vec) {
  pred_stage <- pred_mat
  temp_stage <- temp_vec

  ord <- order(temp_stage)
  pred_stage <- pred_stage[, ord, drop = FALSE]
  temp_stage <- temp_stage[ord]

  res_list <- vector("list", nrow(pred_stage))

  for (i in seq_len(nrow(pred_stage))) {
    y <- as.numeric(pred_stage[i, ])
    sgn <- y - 0.5

    if (all(is.na(y)) || min(y, na.rm = TRUE) > 0.5 || max(y, na.rm = TRUE) < 0.5) {
      res_list[[i]] <- data.frame(
        .draw = i,
        assay_temp_50_1h = NA_real_
      )
      next
    }

    idx <- which(diff(sign(sgn)) != 0)

    if (length(idx) == 0) {
      j <- which.min(abs(sgn))
      temp_50 <- temp_stage[j]
    } else {
      j <- idx[1]
      x1 <- temp_stage[j]
      x2 <- temp_stage[j + 1]
      y1 <- y[j]
      y2 <- y[j + 1]

      if (abs(y2 - y1) < 1e-12) {
        temp_50 <- mean(c(x1, x2))
      } else {
        temp_50 <- x1 + (0.5 - y1) * (x2 - x1) / (y2 - y1)
      }
    }

    res_list[[i]] <- data.frame(
      .draw = i,
      assay_temp_50_1h = temp_50
    )
  }

  bind_rows(res_list)
}

temp_1h_draws_int <- get_temp50_draws_int(
  pred_mat = pr_1h_int,
  temp_vec = nd_1h_int$assay_temp
) %>%
  filter(is.finite(assay_temp_50_1h))

temp_1h_summary_int <- temp_1h_draws_int %>%
  summarise(
    mean_temp = mean(assay_temp_50_1h, na.rm = TRUE),
    median_temp = median(assay_temp_50_1h, na.rm = TRUE),
    lower = quantile(assay_temp_50_1h, 0.025, na.rm = TRUE),
    upper = quantile(assay_temp_50_1h, 0.975, na.rm = TRUE)
  )

temp_1h_summary_int


# =========================
# 9) 2D landscape plot
# =========================
temps_surface <- seq(
  min(d$assay_temp, na.rm = TRUE),
  max(d$assay_temp, na.rm = TRUE),
  length.out = 120
)

dur_surface <- seq(
  min(d$duration, na.rm = TRUE),
  max(d$duration, na.rm = TRUE),
  length.out = 120
)

make_heatmap_df_int <- function() {
  nds <- expand.grid(
    assay_temp = temps_surface,
    duration = dur_surface,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )

  nds <- nds %>%
    mutate(
      logd = log10(duration),
      temp_c = assay_temp - temp_mean,
      n_total = as.integer(round(median(d$n_total, na.rm = TRUE))),
      date_experiment = factor(levels(d$date_experiment)[1], levels = levels(d$date_experiment)),
      obs = factor(levels(d$obs)[1], levels = levels(d$obs))
    )

  pr_s <- posterior_linpred(
    fit_4pl_int,
    newdata = nds,
    re_formula = NA,
    transform = TRUE
  )

  nds$pred_mean <- apply(pr_s, 2, median, na.rm = TRUE)

  nds
}

heat_df_int <- make_heatmap_df_int()

d_points_int <- d %>%
  group_by(assay_temp, duration) %>%
  summarise(
    survival_mean = mean(survival, na.rm = TRUE),
    .groups = "drop"
  )

p_landscape_2d_int <- ggplot(
  heat_df_int,
  aes(x = assay_temp, y = duration, fill = pred_mean)
) +
  geom_raster(interpolate = TRUE) +
  geom_point(
    data = d_points_int,
    aes(x = assay_temp, y = duration, fill = survival_mean),
    inherit.aes = FALSE,
    size = 3.2,
    alpha = 0.9,
    shape = 21,
    color = "white",
    stroke = 0.1
  ) +
  geom_contour(
    aes(z = pred_mean),
    breaks = c(0.25, 0.5, 0.75),
    colour = "white",
    alpha = 0.55,
    linewidth = 0.5
  ) +
  metR::geom_text_contour(
    aes(
      z = pred_mean,
      label = after_stat(level)
    ),
    breaks = c(0.25, 0.5, 0.75),
    colour = "white",
    size = 3,
    stroke = 0,
    nudge_y = 0.05,
    nudge_x = 0.1,
    label.placer = metR::label_placer_n(2)
  ) +
  scale_fill_viridis_c(
    limits = c(0, 1),
    option = "inferno",
    direction = 1
  ) +
  labs(
    x = "Assay temperature (°C)",
    y = "Exposure duration (h)",
    fill = "Predicted survival"
  ) +
  base_theme +
  theme(
    legend.position = "right"
  )

print(p_landscape_2d_int)

ggsave(
  file = "fig/thermal_landscape_int.png",
  dpi = 500,
  height = 6,
  width = 9
)
```



### **Model comparisons**

```{r}
# =========================
# Check colinearity with starting time
# =========================
cor(d$temp_c, d$start_min_c) # r = 0.36

# =========================
# 4PL model with body mass on mid
# =========================
f_tdt_4pl_mass <- bf(
  n_surv | trials(n_total) ~ eta,
  nlf(
    eta ~ logit(
      (inv_logit(lraw) * 0.49) +
        (
          (0.51 + inv_logit(uraw) * 0.49) -
          (inv_logit(lraw) * 0.49)
        ) / (1 + exp(exp(logk) * (logd - mid)))
    ) + z
  ),
  lf(lraw ~ 1),
  lf(uraw ~ 1),
  lf(logk ~ 1),
  lf(mid ~ temp_c + body_mass_c + (1 | date_experiment) + (1 | tank_ID)),
  lf(z ~ 0 + (1 | obs)),
  nl = TRUE
)

# =========================
# Priors
# =========================
priors_tdt_4pl <- c(
  set_prior(
    paste0("normal(", lraw_mean_s, ", 1)"),
    class = "b", nlpar = "lraw"
  ),
  set_prior(
    paste0("normal(", uraw_mean_s, ", 1)"),
    class = "b", nlpar = "uraw"
  ),
  set_prior(
    paste0("normal(", logk_mean_s, ", 1)"),
    class = "b", nlpar = "logk"
  ),
  set_prior(
    paste0("normal(", mid_start_s, ", 1.5)"),
    class = "b", nlpar = "mid", coef = "Intercept"
  ),
  set_prior(
    "normal(0, 0.6)",
    class = "b", nlpar = "mid", coef = "temp_c"
  ),
  set_prior(
    "normal(0, 0.6)",
    class = "b", nlpar = "mid", coef = "body_mass_c"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "date_experiment"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "tank_ID"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "z", group = "obs"
  )
)

# =========================
# Initial values
# =========================
init_tdt_4pl <- function() {
  list(
    b_lraw_Intercept = lraw_mean,
    b_uraw_Intercept = uraw_mean,
    b_logk_Intercept = logk_mean,
    b_mid_Intercept = mid_start,
    b_mid_temp_c = 0
  )
}

init_tdt_4pl_mass <- function() {
  list(
    b_lraw_Intercept = lraw_mean,
    b_uraw_Intercept = uraw_mean,
    b_logk_Intercept = logk_mean,
    b_mid_Intercept = mid_start,
    b_mid_temp_c = 0,
    b_mid_body_mass_c = 0
  )
}

# =========================
# Fit body-mass model
# =========================
fit_4pl_mass <- brm(
  formula = f_tdt_4pl_mass,
  data = d,
  family = binomial(link = "logit"),
  prior = priors_tdt_4pl,
  save_pars = save_pars(all = TRUE),
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = init_tdt_4pl_mass,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)

saveRDS(fit_4pl_mass, file = "Rdata/model_lethal_TDT_body_mass.RDS")


# =========================
# 4PL model with starting time on mid
# =========================
f_tdt_4pl_time <- bf(
  n_surv | trials(n_total) ~ eta,
  nlf(
    eta ~ logit(
      (inv_logit(lraw) * 0.49) +
        (
          (0.51 + inv_logit(uraw) * 0.49) -
          (inv_logit(lraw) * 0.49)
        ) / (1 + exp(exp(logk) * (logd - mid)))
    ) + z
  ),
  lf(lraw ~ 1),
  lf(uraw ~ 1),
  lf(logk ~ 1),
  lf(mid ~ temp_c + start_min_c + (1 | date_experiment) + (1 | tank_ID)),
  lf(z ~ 0 + (1 | obs)),
  nl = TRUE
)

# =========================
# Priors
# =========================
priors_tdt_4pl <- c(
  set_prior(
    paste0("normal(", lraw_mean_s, ", 1)"),
    class = "b", nlpar = "lraw"
  ),
  set_prior(
    paste0("normal(", uraw_mean_s, ", 1)"),
    class = "b", nlpar = "uraw"
  ),
  set_prior(
    paste0("normal(", logk_mean_s, ", 1)"),
    class = "b", nlpar = "logk"
  ),
  set_prior(
    paste0("normal(", mid_start_s, ", 1.5)"),
    class = "b", nlpar = "mid", coef = "Intercept"
  ),
  set_prior(
    "normal(0, 0.6)",
    class = "b", nlpar = "mid", coef = "temp_c"
  ),
  set_prior(
    "normal(0, 0.6)",
    class = "b", nlpar = "mid", coef = "start_min_c"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "date_experiment"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "mid", group = "tank_ID"
  ),
  set_prior(
    "exponential(2)",
    class = "sd", nlpar = "z", group = "obs"
  )
)

# =========================
# Initial values
# =========================
init_tdt_4pl <- function() {
  list(
    b_lraw_Intercept = lraw_mean,
    b_uraw_Intercept = uraw_mean,
    b_logk_Intercept = logk_mean,
    b_mid_Intercept = mid_start,
    b_mid_temp_c = 0
  )
}

init_tdt_4pl_time <- function() {
  list(
    b_lraw_Intercept = lraw_mean,
    b_uraw_Intercept = uraw_mean,
    b_logk_Intercept = logk_mean,
    b_mid_Intercept = mid_start,
    b_mid_temp_c = 0,
    b_mid_start_time_c = 0
  )
}

# =========================
# Fit starting time model
# =========================
fit_4pl_time <- brm(
  formula = f_tdt_4pl_time,
  data = d,
  family = binomial(link = "logit"),
  prior = priors_tdt_4pl,
  save_pars = save_pars(all = TRUE),
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = init_tdt_4pl_time,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)

saveRDS(fit_4pl_time, file = "Rdata/model_lethal_TDT_start_time.RDS")

# =========================
# Summaries
# =========================
print(fit_4pl)
print(fit_4pl_mass)
print(fit_4pl_time)


r2(fit_4pl)
r2(fit_4pl_mass)
r2(fit_4pl_time)

# =========================
# LOO + RELOO
# =========================
loo_4pl <- loo(fit_4pl)
loo_4pl_mass <- loo(fit_4pl_mass)
loo_4pl_time <- loo(fit_4pl_time)
loo_4pl_int <- loo(fit_4pl_int)

saveRDS(loo_4pl, file = "Rdata/loo/loo_lethal_TDT.RDS")
saveRDS(loo_4pl_mass, file = "Rdata/loo/loo_lethal_TDT_mass.RDS")
saveRDS(loo_4pl_time, file = "Rdata/loo/loo_lethal_TDT_time.RDS")
saveRDS(loo_4pl_int, file = "Rdata/loo/loo_lethal_TDT_interaction.RDS")

# RELOO since pareto k values are high
reloo_4pl <- reloo(fit_4pl, loo = loo_4pl, cores = 4)
reloo_4pl_mass <- reloo(fit_4pl_mass, loo = loo_4pl_mass, cores = 4)
reloo_4pl_time <- reloo(fit_4pl_time, loo = loo_4pl_time, cores = 4)
reloo_4pl_int <- reloo(fit_4pl_int, loo = loo_4pl_int, cores = 4)

saveRDS(reloo_4pl, file = "Rdata/loo/reloo_lethal_TDT.RDS")
saveRDS(reloo_4pl_mass, file = "Rdata/loo/reloo_lethal_TDT_mass.RDS")
saveRDS(reloo_4pl_time, file = "Rdata/loo/reloo_lethal_TDT_time.RDS")
saveRDS(reloo_4pl_int, file = "Rdata/loo/reloo_lethal_TDT_interaction.RDS")

##############

reloo_4pl <- readRDS("Rdata/loo/reloo_lethal_TDT.RDS")
reloo_4pl_mass <- readRDS("Rdata/loo/reloo_lethal_TDT_mass.RDS")
reloo_4pl_time <- readRDS("Rdata/loo/reloo_lethal_TDT_time.RDS")
reloo_4pl_int <- readRDS("Rdata/loo/reloo_lethal_TDT_interaction.RDS")


loo_compare(reloo_4pl, reloo_4pl_mass) # Adding body mass does not improve fit significantly
loo_compare(reloo_4pl, reloo_4pl_time) # Adding starting time does not improve fit significantly
loo_compare(reloo_4pl, reloo_4pl_int) # Adding an interaction term does not improve fit significantly

```



# Sublethal data (Bayesian approach)

```{r}
tdt_sub <- read_excel("data/Data_Shrimp_TDT_CTmax_experiment.xlsx", sheet = "SUBLETHAL_TDT")
trial_data <- read_excel("data/Data_Shrimp_TDT_CTmax_experiment.xlsx", sheet = "SUBLETHAL_TDT_trial_data")


tdt_sub <- tdt_sub %>%
    filter(Exclude == "no") %>%
    mutate(
        time_to_death = as.duration(hms(tdt$Time_to_death) - hms(tdt$Starting_time)) / dminutes(1),
        assay_temp = as.numeric(Assay_temperature),
        date_experiment = factor(Date),
        tank_ID = factor(Tank),
        cup_ID = factor(paste0(Trial_ID, Sample)),
        temp_c = assay_temp - temp_mean,
        log10_time_to_death = log10(time_to_death),
        start_min = as.numeric(format(Starting_time, "%H")) * 60 +
      as.numeric(format(Starting_time, "%M")) +
      as.numeric(format(Starting_time, "%S")) / 60 
    ) %>%
  dplyr::mutate(
    start_min_c = as.numeric(scale(start_min, center = TRUE, scale = TRUE))
  ) %>% 
    left_join(
        trial_data %>%
            transmute(
                Trial_ID,
                Assay_temperature,
                Sample,
                ind_body_mass = Sample_body_mass_g / N_individuals_body_mass,
                mortality_rate = Mortality_the_next_day,
                n_ind = N_individuals_mortality
            ),
        by = c("Trial_ID", "Assay_temperature", "Sample") # Merge trial data to get body mass and mortality rate after the trial. 
    )

tdt_sub <- tdt_sub %>% 
  mutate(n_dead = n_ind * mortality_rate,
         n_surv = n_ind * (1 - mortality_rate))

# =========================
# Bayesian model
# =========================
f_sub <- bf(
  log10_time_to_death ~ temp_c + (1 | date_experiment) + (1 | tank_ID) + (1 | cup_ID)
)

priors_sub <- c(
  prior(normal(2, 1.5), class = "Intercept"),
  prior(normal(-1, 1), class = "b"),
  prior(exponential(2), class = "sd"),
  prior(exponential(2), class = "sigma")
)

fit_sublethal_tdt <- brm(
  formula = f_sub,
  data = tdt_sub,
  prior = priors_sub,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = 0,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)

print(fit_sublethal_tdt) # z = 1.10
r2(fit_sublethal_tdt)

# =========================
# Posterior-derived sublethal TDT line
# population-level mean line + 95% credible interval
# analytically parallel to the lethal Bayesian workflow
# =========================
temps_sub_df <- data.frame(
  assay_temp = seq(30.5, 34, by = 0.01)
) %>%
  mutate(
    temp_c = assay_temp - temp_mean,
    date_experiment = factor(levels(tdt_sub$date_experiment)[1], levels = levels(tdt_sub$date_experiment)),
    tank_ID = factor(levels(tdt_sub$tank_ID)[1], levels = levels(tdt_sub$tank_ID)),
    cup_ID = factor(levels(tdt_sub$cup_ID)[1], levels = levels(tdt_sub$cup_ID))
  )

lp_sub <- posterior_linpred(
  fit_sublethal_tdt,
  newdata = temps_sub_df,
  re_formula = NA
)

# lp_sub is on the log10(minutes) scale
time_sub_draws <- 10^lp_sub

pred_df_tdt_sub <- temps_sub_df %>%
  mutate(
    time_to_death = apply(time_sub_draws, 2, median, na.rm = TRUE),
    lower = apply(time_sub_draws, 2, quantile, probs = 0.025, na.rm = TRUE),
    upper = apply(time_sub_draws, 2, quantile, probs = 0.975, na.rm = TRUE)
  ) %>%
  select(assay_temp, time_to_death, lower, upper)

# =========================
# Sublethal-only plot
# =========================
p_sublethal_lin <- ggplot(pred_df_tdt_sub) +
  geom_ribbon(
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "black",
    show.legend = FALSE
  ) +
  geom_line(
    aes(x = assay_temp, y = time_to_death),
    linewidth = 1.6,
    colour = "black"
  ) +
  geom_jitter(
    data = tdt_sub,
    aes(x = assay_temp, y = time_to_death),
    width = 0.05,
    height = 0,
    size = 4,
    alpha = 0.9,
    shape = 21,
    col = "white",
    fill = "black"
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_continuous(breaks = ybreaks_left, labels = lab_number) +
  coord_cartesian(
    xlim = c(30.5, 34),
    ylim = c(0.9, 480)
  ) +
  labs(
    x = "Assay temperature (°C)",
    y = "Time to death (min)"
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

p_sublethal_log <- ggplot(pred_df_tdt_sub) +
  geom_ribbon(
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "black",
    show.legend = FALSE
  ) +
  geom_line(
    aes(x = assay_temp, y = time_to_death),
    linewidth = 1.6,
    colour = "black"
  ) +
  geom_jitter(
    data = tdt_sub,
    aes(x = assay_temp, y = time_to_death),
    width = 0.05,
    height = 0,
    size = 4,
    alpha = 0.9,
    shape = 21,
    col = "white",
    fill = "black",
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(breaks = ybreaks_right, labels = lab_number) +
  coord_cartesian(
    xlim = c(30.5, 34),
    ylim = c(0.2, 480)
  ) +
  labs(
    x = "Assay temperature (°C)",
    y = expression(log[10] * "(Time to death, min)")
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

print(p_sublethal_lin + p_sublethal_log + patchwork::plot_layout(ncol = 2))

ggsave(
  file = "fig/fig_tdt_sublethal.png",
  dpi = 500,
  height = 6,
  width = 13
)
```

## **Visualisation with median time to death**
```{r}
# =========================
# Summary data
# =========================
tdt_sub_sum <- tdt_sub %>%
    dplyr::group_by(assay_temp) %>%
    dplyr::summarise(
        median_time_to_death = median(time_to_death, na.rm = TRUE),
        lower = quantile(time_to_death, 0.25, na.rm = TRUE),
        upper = quantile(time_to_death, 0.75, na.rm = TRUE), # IQR range
        .groups = "drop"
    )

# =========================
# Sublethal-only plot
# =========================
p_sublethal_lin <- ggplot(pred_df_tdt_sub) +
    geom_jitter(
        data = tdt_sub,
        aes(x = assay_temp, y = time_to_death),
        width = 0.05,
        height = 0,
        size = 3.5,
        alpha = 0.28,
        shape = 21,
        colour = "white",
        fill = "black"
    ) +
    geom_errorbar(
        data = tdt_sub_sum,
        aes(
            x = assay_temp,
            ymin = lower,
            ymax = upper,
        ),
        width = 0.08,
        linewidth = 0.8,
        colour = "black"
    ) +
    geom_point(
        data = tdt_sub_sum,
        aes(x = assay_temp, y = median_time_to_death),
        size = 3.2,
        shape = 21,
        stroke = 0.8,
        colour = "white",
        fill = "black"
    ) +
    geom_ribbon(
        aes(x = assay_temp, ymin = lower, ymax = upper),
        alpha = 0.20,
        fill = "black",
        show.legend = FALSE
    ) +
    geom_line(
        aes(x = assay_temp, y = time_to_death),
        linewidth = 1.6,
        colour = "black"
    ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    scale_y_continuous(breaks = ybreaks_left, labels = lab_number) +
    coord_cartesian(
        xlim = c(30.5, 34),
        ylim = c(0.9, 480)
    ) +
    labs(
        x = "Assay temperature (°C)",
        y = "Time to death (min)"
    ) +
    base_theme +
    theme(
        legend.position = "none"
    )

p_sublethal_log <- ggplot(pred_df_tdt_sub) +
    geom_jitter(
        data = tdt_sub,
        aes(x = assay_temp, y = time_to_death),
        width = 0.05,
        height = 0,
        size = 3.5,
        alpha = 0.28,
        shape = 21,
        colour = "white",
        fill = "black"
    ) +
    geom_errorbar(
        data = tdt_sub_sum,
        aes(
            x = assay_temp,
            ymin = lower,
            ymax = upper
        ),
        width = 0.08,
        linewidth = 0.8,
        colour = "black"
    ) +
    geom_point(
        data = tdt_sub_sum,
        aes(x = assay_temp, y = median_time_to_death),
        size = 3.2,
        shape = 21,
        stroke = 0.8,
        colour = "white",
        fill = "black"
    ) +
    geom_ribbon(
        aes(x = assay_temp, ymin = lower, ymax = upper),
        alpha = 0.20,
        fill = "black",
        show.legend = FALSE
    ) +
    geom_line(
        aes(x = assay_temp, y = time_to_death),
        linewidth = 1.6,
        colour = "black"
    ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    scale_y_log10(breaks = ybreaks_right, labels = lab_number) +
    coord_cartesian(
        xlim = c(30.5, 34),
        ylim = c(0.2, 480)
    ) +
    labs(
        x = "Assay temperature (°C)",
        y = expression(log[10] * "(Time to death, min)")
    ) +
    base_theme +
    theme(
        legend.position = "none"
    )

print(p_sublethal_lin + p_sublethal_log + patchwork::plot_layout(ncol = 2))
```

## **Without the 32 degrees temperature** 

```{r}

## =========================================================
## Dataset WITHOUT 32°C
## =========================================================
tdt_sub_no32 <- tdt_sub %>%
  dplyr::filter(assay_temp != 32)

fit_sublethal_no32 <- brm(
  formula = f_sub,
  data = tdt_sub_no32,
  prior = priors_sub,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = 0,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)


print(fit_sublethal_no32) # z = 1.06
r2(fit_sublethal_no32)


# =========================
# Prediction grid
# =========================
temps_sub_df <- data.frame(
  assay_temp = seq(30.5, 34, by = 0.01)
) %>%
  mutate(
    temp_c = assay_temp - temp_mean,
    date_experiment = factor(levels(tdt_sub$date_experiment)[1], levels = levels(tdt_sub$date_experiment)),
    tank_ID = factor(levels(tdt_sub$tank_ID)[1], levels = levels(tdt_sub$tank_ID)),
    cup_ID = factor(levels(tdt_sub$cup_ID)[1], levels = levels(tdt_sub$cup_ID))
  )

# =========================
# Predictions: full model
# =========================
lp_sub_full <- posterior_linpred(
  fit_sublethal_tdt,
  newdata = temps_sub_df,
  re_formula = NA
)

time_sub_draws_full <- 10^lp_sub_full

pred_df_tdt_sub_full <- temps_sub_df %>%
  mutate(
    time_to_death = apply(time_sub_draws_full, 2, median, na.rm = TRUE),
    lower = apply(time_sub_draws_full, 2, quantile, probs = 0.025, na.rm = TRUE),
    upper = apply(time_sub_draws_full, 2, quantile, probs = 0.975, na.rm = TRUE)
  ) %>%
  dplyr::select(assay_temp, time_to_death, lower, upper)

# =========================
# Predictions: no-32 model
# =========================
lp_sub_no32 <- posterior_linpred(
  fit_sublethal_no32,
  newdata = temps_sub_df,
  re_formula = NA
)

time_sub_draws_no32 <- 10^lp_sub_no32

pred_df_tdt_sub_no32 <- temps_sub_df %>%
  mutate(
    time_to_death = apply(time_sub_draws_no32, 2, median, na.rm = TRUE),
    lower = apply(time_sub_draws_no32, 2, quantile, probs = 0.025, na.rm = TRUE),
    upper = apply(time_sub_draws_no32, 2, quantile, probs = 0.975, na.rm = TRUE)
  ) %>%
  dplyr::select(assay_temp, time_to_death, lower, upper)

# =========================
# Plot: linear scale
# =========================
p_sublethal_lin_compare <- ggplot() +
  geom_jitter(
    data = tdt_sub,
    aes(x = assay_temp, y = time_to_death),
    width = 0.05,
    height = 0,
    size = 3.5,
    alpha = 0.22,
    shape = 21,
    colour = "white",
    fill = "black"
  ) +
  geom_errorbar(
    data = tdt_sub_sum,
    aes(
      x = assay_temp,
      ymin = lower,
      ymax = upper,
    ),
    width = 0.08,
    linewidth = 0.8,
    colour = "black"
  ) +
  geom_point(
    data = tdt_sub_sum,
    aes(x = assay_temp, y = median_time_to_death),
    size = 3.2,
    shape = 21,
    stroke = 0.8,
    colour = "white",
    fill = "black"
  ) +
  geom_ribbon(
    data = pred_df_tdt_sub_full,
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "black",
    show.legend = FALSE
  ) +
  geom_ribbon(
    data = pred_df_tdt_sub_no32,
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "black",
    show.legend = FALSE
  ) +
  geom_line(
    data = pred_df_tdt_sub_full,
    aes(x = assay_temp, y = time_to_death),
    linewidth = 1.6,
    colour = "black"
  ) +
  geom_line(
    data = pred_df_tdt_sub_no32,
    aes(x = assay_temp, y = time_to_death),
    linewidth = 1.2,
    colour = "black",
    linetype = "dashed"
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_continuous(breaks = ybreaks_left, labels = lab_number) +
  coord_cartesian(
    xlim = c(30.5, 34),
    ylim = c(0.9, 480)
  ) +
  labs(
    x = "Assay temperature (°C)",
    y = "Time to death (min)"
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

# =========================
# Plot: log scale
# =========================
p_sublethal_log_compare <- ggplot() +
  geom_jitter(
    data = tdt_sub,
    aes(x = assay_temp, y = time_to_death),
    width = 0.05,
    height = 0,
    size = 3.5,
    alpha = 0.22,
    shape = 21,
    colour = "white",
    fill = "black"
  ) +
  geom_errorbar(
    data = tdt_sub_sum,
    aes(
      x = assay_temp,
      ymin = lower,
      ymax = upper
    ),
    width = 0.08,
    linewidth = 0.8,
    colour = "black"
  ) +
  geom_point(
    data = tdt_sub_sum,
    aes(x = assay_temp, y = median_time_to_death),
    size = 3.2,
    shape = 21,
    stroke = 0.8,
    colour = "white",
    fill = "black"
  ) +
  geom_ribbon(
    data = pred_df_tdt_sub_full,
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "black",
    show.legend = FALSE
  ) +
  geom_ribbon(
    data = pred_df_tdt_sub_no32,
    aes(x = assay_temp, ymin = lower, ymax = upper),
    alpha = 0.20,
    fill = "black",
    show.legend = FALSE
  ) +
  geom_line(
    data = pred_df_tdt_sub_full,
    aes(x = assay_temp, y = time_to_death),
    linewidth = 1.6,
    colour = "black"
  ) +
  geom_line(
    data = pred_df_tdt_sub_no32,
    aes(x = assay_temp, y = time_to_death),
    linewidth = 1.2,
    colour = "black",
    linetype = "dashed"
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(breaks = ybreaks_right, labels = lab_number) +
  coord_cartesian(
    xlim = c(30.5, 34),
    ylim = c(0.2, 480)
  ) +
  labs(
    x = "Assay temperature (°C)",
    y = expression(log[10] * "(Time to death, min)")
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

print(p_sublethal_lin_compare + p_sublethal_log_compare + patchwork::plot_layout(ncol = 2))
```



## **Model comparisons**
```{r}
tdt_sub <- tdt_sub %>% 
  mutate(ind_body_mass_c = scale(ind_body_mass, scale = F, center = T))

# Starting time and temperature are highly colinear! 
cor(tdt_sub$temp_c, tdt_sub$start_min_c)
# r = 0.71

## =========================================================
## Formulas
## =========================================================
f_sub_base <- bf(
  log10_time_to_death ~ temp_c + (1 | date_experiment) + (1 | tank_ID) + (1 | cup_ID)
)

f_sub_mort <- bf(
  log10_time_to_death ~ temp_c + mortality_rate + (1 | date_experiment) + (1 | tank_ID) + (1 | cup_ID)
)

f_sub_mass <- bf(
  log10_time_to_death ~ temp_c + ind_body_mass_c + (1 | date_experiment) + (1 | tank_ID) + (1 | cup_ID)
)

## =========================================================
## Priors
## =========================================================
priors_sub <- c(
  prior(normal(2, 1.5), class = "Intercept"),
  prior(normal(-1, 1), class = "b", coef = "temp_c"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(2), class = "sd"),
  prior(exponential(2), class = "sigma")
)

## =========================================================
## Fit models
## =========================================================
fit_sub_base <- brm(
  formula = f_sub_base,
  data = tdt_sub,
  prior = priors_sub,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = 0,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)
saveRDS(fit_sub_base, file = "Rdata/model_sublethal_TDT.RDS")

## Mortality rate
fit_sub_mort <- brm(
  formula = f_sub_mort,
  data = tdt_sub,
  prior = priors_sub,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = 0,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)
saveRDS(fit_sub_mort, file = "Rdata/model_sublethal_TDT_mortality_rate.RDS")

## Body mass
fit_sub_mass <- brm(
  formula = f_sub_mass,
  data = tdt_sub,
  prior = priors_sub,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = 0,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)
saveRDS(fit_sub_mass, file = "Rdata/model_sublethal_TDT_body_mass.RDS")

## =========================================================
## LOO
## =========================================================
loo_sub_base <- loo(fit_sub_base)
loo_sub_mort <- loo(fit_sub_mort)
loo_sub_mass <- loo(fit_sub_mass)

loo_compare(
  loo_sub_base,
  loo_sub_mort) # Does not improve fit

loo_compare(
  loo_sub_base,
  loo_sub_mass) # Does not improve fit

# The simple model is the best, and most parsimonious fit to the data. 

```

## **Temperature-driven mortality**
```{r}
tdt_trial <- read_excel("data/Data_Shrimp_TDT_CTmax_experiment.xlsx", sheet = "SUBLETHAL_TDT_trial_data")

tdt_trial <- tdt_trial %>%
    filter(Exclude == "no") %>%
    mutate(
        assay_temp = as.numeric(Assay_temperature),
        trial_ID = factor(Trial_ID), # Same levels as date_experiment
        tank_ID = factor(Tank),
        cup_ID = factor(paste0(Trial_ID, Sample)),
        temp_c = assay_temp - temp_mean,
        mortality_rate = Mortality_the_next_day,
        n_ind = N_individuals_mortality,
        n_dead = as.integer(n_ind * mortality_rate),
        n_surv = n_ind - n_dead,
        survival_rate = n_surv / n_ind) 


f_sub_survival <- bf(n_surv | trials(n_ind) ~ temp_c + (1 | trial_ID) + (1 | tank_ID) + (1 | cup_ID)
)

priors_sub_survival <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(-1, 1), class = "b", coef = "temp_c"),
  prior(exponential(2), class = "sd")
)

fit_sub_survival  <- brm(
  formula = f_sub_survival ,
  data = tdt_trial,
  prior = priors_sub_survival,
  family = binomial(link = "logit"),
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  init = 0,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.999, max_treedepth = 15),
  seed = 123
)

saveRDS(fit_sub_survival, file = "Rdata/model_survival_sublethal_TDT.RDS")


print(fit_sub_survival) 
r2(fit_sub_survival)


## =========================================================
## Prediction grid on temp_c scale, then back-transform
## =========================================================
nd_surv <- data.frame(
    temp_c = seq(
        from = min(tdt_trial$temp_c, na.rm = TRUE),
        to = max(tdt_trial$temp_c, na.rm = TRUE),
        by = 0.001
    )
) %>%
    mutate(
        assay_temp = temp_c + temp_mean,
        n_ind = as.integer(round(median(tdt_trial$n_ind, na.rm = TRUE))),
        trial_ID = factor(levels(tdt_trial$trial_ID)[1], levels = levels(tdt_trial$trial_ID)),
        tank_ID = factor(levels(tdt_trial$tank_ID)[1], levels = levels(tdt_trial$tank_ID)),
        cup_ID = factor(levels(tdt_trial$cup_ID)[1], levels = levels(tdt_trial$cup_ID))
    )

## =========================================================
## Posterior survival probability
## =========================================================
pr_surv <- posterior_linpred(
    fit_sub_survival,
    newdata = nd_surv,
    re_formula = NA,
    transform = TRUE
)

nd_surv$pred_mean <- apply(pr_surv, 2, median, na.rm = TRUE)
nd_surv$pred_lo   <- apply(pr_surv, 2, quantile, probs = 0.025, na.rm = TRUE)
nd_surv$pred_hi   <- apply(pr_surv, 2, quantile, probs = 0.975, na.rm = TRUE)

## Optional sanity checks
print(range(nd_surv$pred_mean, na.rm = TRUE))
print(range(nd_surv$pred_lo, na.rm = TRUE))
print(range(nd_surv$pred_hi, na.rm = TRUE))

## =========================================================
## Plot
## =========================================================
p_sub_survival <- ggplot(
    nd_surv,
    aes(x = assay_temp, y = pred_mean)
) +
    geom_ribbon(
        aes(ymin = pred_lo, ymax = pred_hi),
        alpha = 0.15,
        colour = NA
    ) +
    geom_line(linewidth = 1.2) +
    geom_jitter(
        data = tdt_trial,
        aes(x = assay_temp, y = survival_rate),
        inherit.aes = FALSE,
        width = 0.025,
        height = 0.02,
        size = 2.5,
        alpha = 0.8
    ) +
    labs(
        x = "Assay temperature (°C)",
        y = "Predicted survival"
    ) +
    coord_cartesian(
        xlim = range(tdt_trial$assay_temp, na.rm = TRUE),
        ylim = c(0, 1.02)
    ) +
    theme_classic(base_size = 16)

p_sub_survival
```



# **CTmax**

## **Ramping rate** 

```{r}
# ========================
# Fast ramping trials
# ========================
ctmax_trial <- read_excel(
  "data/Data_Shrimp_TDT_CTmax_experiment.xlsx",
  sheet = "CTMAX_ramping_data"
)

ctmax_trial <- ctmax_trial %>%
  dplyr::mutate(
    Time_min = as.numeric(Time_min),
    Temperature = as.numeric(Temperature)
  ) %>%
  dplyr::filter(!is.na(Time_min), !is.na(Temperature))

# Keep the per-trial ramping-rate calculation
ramping_summary <- ctmax_trial %>%
  dplyr::group_by(Trial_ID) %>%
  dplyr::group_modify(~{
    mod <- lm(Temperature ~ Time_min, data = .x)
    coefs <- summary(mod)$coefficients
    data.frame(
      ramping_rate = coefs["Time_min", "Estimate"],
      ramping_rate_se = coefs["Time_min", "Std. Error"],
      r2 = summary(mod)$r.squared
    )
  }) %>%
  dplyr::ungroup() %>%
  dplyr::filter(ramping_rate > 0)

ctmax_trial$group <- ifelse(
  ctmax_trial$Trial_ID %in% c("CTmax_001", "CTmax_002", "CTmax_007"),
  "~0.3 °C/min",
  "~0.5 °C/min"
)

ctmax_trial_plot <- ctmax_trial %>%
  dplyr::select(Trial_ID, group, Time_min, Temperature)

# ========================
# Slow warming rate data
# ========================
prog1 <- read_excel("data/Shrimp_slow_CTM_1_2025-06-23_18_29_12_CEST.xlsx")
prog2 <- read_excel("data/Shrimp_slow_CTM_3&4_2025-06-25_17_44_50_CEST.xlsx")

prog1 <- prog1[, -1]
prog2 <- prog2[, -1]

names(prog1)[1:2] <- c("datetime", "temp")
names(prog2)[1:2] <- c("datetime", "temp")

prog1 <- prog1 %>%
  dplyr::mutate(
    datetime = as.POSIXct(datetime, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"),
    temp = as.numeric(temp)
  )

prog2 <- prog2 %>%
  dplyr::mutate(
    datetime = as.POSIXct(datetime, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"),
    temp = as.numeric(temp)
  )

start_cutoff_prog1 <- as.POSIXct("2025-06-22 23:12:02", tz = "UTC")
end_cutoff_prog1   <- as.POSIXct("2025-06-23 18:15:02", tz = "UTC")

start_cutoff_prog2 <- as.POSIXct("2025-06-24 22:31:18", tz = "UTC")
end_cutoff_prog2   <- as.POSIXct("2025-06-25 15:26:18", tz = "UTC")

prog1 <- prog1 %>%
  dplyr::filter(datetime >= start_cutoff_prog1,
                datetime <= end_cutoff_prog1) %>%
  dplyr::select(datetime, temp)

prog2 <- prog2 %>%
  dplyr::filter(datetime >= start_cutoff_prog2,
                datetime <= end_cutoff_prog2) %>%
  dplyr::select(datetime, temp)

prog1$hours <- as.numeric(difftime(prog1$datetime, min(prog1$datetime), units = "hours"))
prog2$hours <- as.numeric(difftime(prog2$datetime, min(prog2$datetime), units = "hours"))

# Keep the slow-ramping model fits too
mod_slow1 <- lm(temp ~ hours, data = prog1)
mod_slow2 <- lm(temp ~ hours, data = prog2)

slow_summary <- data.frame(
  Trial_ID = c("CTmax_005 + CTmax_006", "CTmax_009 + CTmax_010"),
  ramping_rate = c(coef(mod_slow1)[["hours"]], coef(mod_slow2)[["hours"]]),
  ramping_rate_se = c(summary(mod_slow1)$coefficients["hours", "Std. Error"],
                      summary(mod_slow2)$coefficients["hours", "Std. Error"]),
  r2 = c(summary(mod_slow1)$r.squared, summary(mod_slow2)$r.squared)
)

# Convert hours to minutes for plotting alongside fast trials
prog1_plot <- prog1 %>%
  dplyr::mutate(
    Trial_ID = "CTmax_005 + CTmax_006",
    group = "~0.02 °C/min",
    Time_min = hours * 60,
    Temperature = temp
  ) %>%
  dplyr::select(Trial_ID, group, Time_min, Temperature)

prog2_plot <- prog2 %>%
  dplyr::mutate(
    Trial_ID = "CTmax_009 + CTmax_010",
    group = "~0.02 °C/min",
    Time_min = hours * 60,
    Temperature = temp
  ) %>%
  dplyr::select(Trial_ID, group, Time_min, Temperature)

all_plot <- dplyr::bind_rows(
  prog1_plot,
  prog2_plot,
  ctmax_trial_plot
)

all_plot$group <- factor(
  all_plot$group,
  levels = c("~0.02 °C/min", "~0.3 °C/min", "~0.5 °C/min")
)

ggplot(all_plot, aes(x = Time_min, y = Temperature)) +
  geom_line(aes(group = Trial_ID), linewidth = 1.2, colour = "black", alpha = 0.9) +
  facet_wrap(~ group, scales = "free_x", nrow = 1) +
  labs(
    x = "Time (min)",
    y = expression(Temperature~(degree*C))
  ) +
  theme_classic(base_size = 16)
```

## **Data processing** 
```{r}
ctmax <- read_excel("data/Data_Shrimp_TDT_CTmax_experiment.xlsx",
                sheet = "CTMAX")

ctmax$Ramping_rate[ctmax$Ramping_rate < 0.1] <- 0.02


ctmax <- ctmax %>%
  dplyr::mutate(
    Date_experiment = as.factor(as.Date(Date)),
    Trial_ID = as.factor(Trial_ID),
    Ramping = as.factor(Ramping_rate),
    start_min = as.numeric(format(Starting_time, "%H")) * 60 +
      as.numeric(format(Starting_time, "%M")) +
      as.numeric(format(Starting_time, "%S")) / 60
  ) %>%
  dplyr::mutate(
    start_min_c = as.numeric(scale(start_min, center = TRUE, scale = FALSE),
    start_temp_c = as.numeric(scale(Starting_temperature, center = TRUE, scale = FALSE)))
  )

str(ctmax)

# Combine with ramping rate data
# Fast ramping rates from CTMAX_ramping_data
ramping_summary_fast <- ctmax_trial %>%
  dplyr::group_by(Trial_ID) %>%
  dplyr::group_modify(~{
    mod <- lm(Temperature ~ Time_min, data = .x)
    coefs <- summary(mod)$coefficients
    data.frame(
      ramping_rate = coefs["Time_min", "Estimate"],
      ramping_rate_se = coefs["Time_min", "Std. Error"]
    )
  }) %>%
  dplyr::ungroup() %>%
  dplyr::filter(ramping_rate > 0)

# Slow ramping rates from prog1 / prog2
ramping_summary_slow <- data.frame(
  Trial_ID = c("CTmax_005", "CTmax_006", "CTmax_009", "CTmax_010"),
  ramping_rate = c(
    coef(mod_slow1)[["hours"]] / 60,
    coef(mod_slow1)[["hours"]] / 60,
    coef(mod_slow2)[["hours"]] / 60,
    coef(mod_slow2)[["hours"]] / 60
  ),
  ramping_rate_se = c(
    summary(mod_slow1)$coefficients["hours", "Std. Error"] / 60,
    summary(mod_slow1)$coefficients["hours", "Std. Error"] / 60,
    summary(mod_slow2)$coefficients["hours", "Std. Error"] / 60,
    summary(mod_slow2)$coefficients["hours", "Std. Error"] / 60
  )
)

# Combine fast + slow
ramping_summary_all <- dplyr::bind_rows(
  ramping_summary_fast,
  ramping_summary_slow
)

# Join to ctmax
ctmax$Trial_ID <- as.character(ctmax$Trial_ID)
ramping_summary_all$Trial_ID <- as.character(ramping_summary_all$Trial_ID)

ctmax <- ctmax %>%
  dplyr::left_join(ramping_summary_all, by = "Trial_ID")

# Quick check
ctmax %>%
  dplyr::group_by(Ramping) %>%
  dplyr::summarise(
    mean_ctmax = mean(CTmax, na.rm = TRUE),
    sd_ctmax = sd(CTmax, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

# Ramping rate variation among trials
ctmax %>%
  dplyr::group_by(Ramping, Trial_ID) %>%
  dplyr::summarise(
    ramping_rate = mean(ramping_rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::group_by(Ramping) %>%
  dplyr::summarise(
    mean_ramp = mean(ramping_rate, na.rm = TRUE),
    sd_ramp = sd(ramping_rate, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

# Starting temperature variation among trials
ctmax %>%
  dplyr::group_by(Ramping, Trial_ID) %>%
  dplyr::summarise(
    start_temp = mean(Starting_temperature, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::group_by(Ramping) %>%
  dplyr::summarise(
    mean_ramp = mean(start_temp, na.rm = TRUE),
    sd_ramp = sd(start_temp, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

# =========================
# Combine body mass and mortality data 
# =========================
ctmax_trial  <- read_excel("data/Data_Shrimp_TDT_CTmax_experiment.xlsx",
                sheet = "CTMAX_trial_data") %>% 
  dplyr::select(-Ramping_rate)

ctmax <- left_join(ctmax, ctmax_trial, by = "Trial_ID")


# =========================
# Process other columns
# =========================
ctmax <- ctmax %>%
  dplyr::mutate(
    body_mass = Average_body_mass_individuals,
    body_mass_c = scale(body_mass, scale = F),
    mortality_rate = Mortality_the_next_day, 
  )

# Calculate deviation from the average ramping rate among replicates
trial_ramping_dev <- ctmax %>%
  dplyr::distinct(Ramping, Trial_ID, ramping_rate) %>%
  dplyr::group_by(Ramping) %>%
  dplyr::mutate(
    mean_ramp_group = mean(ramping_rate, na.rm = TRUE),
    ramping_dev = ramping_rate - mean_ramp_group
  ) %>%
  dplyr::ungroup()

ctmax <- ctmax %>%
  dplyr::left_join(
    trial_ramping_dev %>%
      dplyr::select(Trial_ID, mean_ramp_group, ramping_dev),
    by = "Trial_ID"
  )

# ========================
# Body mass summary
# =======================

ctmax %>% 
  group_by(Ramping) %>% 
  summarise(mean_body_mass = mean(body_mass),
            sd_body_mass = sd(body_mass),
            n = n())
```


## **Bayesian linear model**

### **Model comparisons**

```{r}
# =========================
# 2) Priors
# =========================
priors_ctmax <- c(
  prior(normal(33, 3), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sigma")
)

# =========================
# 3) Models
# =========================

# Base model
fit_ctmax <- brm(
  formula = CTmax ~ Ramping + (1 | Trial_ID) + (1 | Date_experiment),
  data = ctmax,
  prior = priors_ctmax,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  seed = 123,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

saveRDS(fit_ctmax, file = "Rdata/model_ctmax.RDS")


# Model with start time
fit_ctmax_time <- brm(
  formula = CTmax ~ Ramping + start_min_c + (1 | Trial_ID) + (1 | Date_experiment),
  data = ctmax,
  prior = priors_ctmax,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  seed = 123,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

saveRDS(fit_ctmax_time, file = "Rdata/model_ctmax_start_time.RDS")

# Model with body mass
fit_ctmax_mass <- brm(
  formula = CTmax ~ Ramping + body_mass_c + (1 | Trial_ID) + (1 | Date_experiment),
  data = ctmax,
  prior = priors_ctmax,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  seed = 123,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

saveRDS(fit_ctmax_mass, file = "Rdata/model_ctmax_body_mass.RDS")

# Model with variation in ramping rate
fit_ctmax_ramp <- brm(
  formula = CTmax ~ Ramping + ramping_dev + (1 | Trial_ID) + (1 | Date_experiment),
  data = ctmax,
  prior = priors_ctmax,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  seed = 123,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

saveRDS(fit_ctmax_ramp, file = "Rdata/model_ctmax_ramping.RDS")


# Model with mortality rate
fit_ctmax_mortality <- brm(
  formula = CTmax ~ Ramping + mortality_rate + (1 | Trial_ID) + (1 | Date_experiment),
  data = ctmax,
  prior = priors_ctmax,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  seed = 123,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

saveRDS(fit_ctmax_mortality, file = "Rdata/model_ctmax_mortality.RDS")

# Model with observer
fit_ctmax_observer <- brm(
  formula = CTmax ~ Ramping + Observer + (1 | Trial_ID) + (1 | Date_experiment),
  data = ctmax,
  prior = priors_ctmax,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  seed = 123,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

saveRDS(fit_ctmax_observer, file = "Rdata/model_ctmax_observer.RDS")


# Model with starting temperature
fit_ctmax_start_temp <- brm(
  formula = CTmax ~ Ramping + start_temp_c + (1 | Trial_ID) + (1 | Date_experiment),
  data = ctmax,
  prior = priors_ctmax,
  chains = 4,
  cores = 4,
  iter = 10000,
  warmup = 5000,
  seed = 123,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)

saveRDS(fit_ctmax_start_temp, file = "Rdata/model_ctmax_start_temp.RDS")

# =========================
# 4) LOO
# =========================
loo_ctmax <- loo(fit_ctmax)
loo_ctmax_time <- loo(fit_ctmax_time)
loo_ctmax_mass <- loo(fit_ctmax_mass)
loo_ctmax_ramp <- loo(fit_ctmax_ramp)
loo_ctmax_mortality <- loo(fit_ctmax_mortality)
loo_ctmax_observer <- loo(fit_ctmax_observer)
loo_ctmax_start_temp <- loo(fit_ctmax_start_temp)


loo_compare(loo_ctmax, loo_ctmax_time) # Base model is better 
loo_compare(loo_ctmax, loo_ctmax_mass) # Model with body mass is slightly better, but not a big improvement to model fit --> keep parsimonious model
loo_compare(loo_ctmax, loo_ctmax_ramp) # Base model is  better
loo_compare(loo_ctmax, loo_ctmax_mortality) # Base model is better
loo_compare(loo_ctmax, loo_ctmax_observer) # Base model is better
loo_compare(loo_ctmax, loo_ctmax_start_temp) # Base model is better
```

### **Final model** 

```{r}
# =========================
# 5) Summaries
# =========================
summary(fit_ctmax)

posterior_summary(fit_ctmax, pars = c("^b_", "^sd_", "^sigma"))


# =========================
# 6) Newdata for posterior estimates
# =========================
nd_ctmax <- data.frame(
  Ramping = factor(c("0.02", "0.3", "0.5"),
                   levels = levels(ctmax$Ramping))
)

# posterior expected mean CTmax for each ramping rate
post_ctmax <- posterior_epred(
  fit_ctmax,
  newdata = nd_ctmax,
  re_formula = NA
)

# =========================
# 7) Summary table of posterior mean and 95% CI
# =========================
post_summary <- data.frame(
  Ramping = c("0.02", "0.3", "0.5"),
  mean = apply(post_ctmax, 2, mean),
  lwr = apply(post_ctmax, 2, quantile, probs = 0.025),
  upr = apply(post_ctmax, 2, quantile, probs = 0.975)
)

post_summary$Ramping_num <- as.numeric(post_summary$Ramping)

post_summary_print <- post_summary
post_summary_print$mean <- round(post_summary_print$mean, 2)
post_summary_print$lwr  <- round(post_summary_print$lwr, 2)
post_summary_print$upr  <- round(post_summary_print$upr, 2)


ft_ctmax <- flextable::flextable(
  post_summary_print[, c("Ramping", "mean", "lwr", "upr")]
)

ft_ctmax <- flextable::set_header_labels(
  ft_ctmax,
  Ramping = "Ramping rate (°C/min)",
  mean = "Posterior mean",
  lwr = "Lower 95% CI",
  upr = "Upper 95% CI"
)

flextable::autofit(ft_ctmax)


# =========================
# 7) Long dataframe of posterior draws for ggdist
# =========================
post_long <- data.frame(
  Ramping = rep(c("0.02", "0.3", "0.5"), each = nrow(post_ctmax)),
  CTmax_post = c(
    post_ctmax[, 1],
    post_ctmax[, 2],
    post_ctmax[, 3]
  )
)

post_long$Ramping_num <- as.numeric(post_long$Ramping)

# =========================
# 8) Plot
# =========================
ggplot() +

  ggdist::stat_halfeye(
    data = post_long,
    aes(
      x = Ramping_num,
      y = CTmax_post,
      group = Ramping,
      fill = Ramping,
      colour = Ramping
    ),
    point_interval = ggdist::mean_qi,
    .width = 0.95,
    adjust = 0.9,
    width = 0.04,
    size = 5,
    slab_alpha = 0.3,
    point_size = 4,
    slab_colour = NA,
    justification = -0.01,
    position = position_nudge(x = 0.023)
  ) +

  geom_point(
    data = ctmax,
    aes(
      x = as.numeric(as.character(Ramping)),
      y = CTmax,
      colour = factor(Ramping)
    ),
    position = position_jitter(width = 0.0075, height = 0),
    alpha = 0.5,
    size = 4,
    shape = 1,
    stroke = 1.2,
    fill="white"
  ) +

  coord_cartesian(ylim = c(28, 35)) +

  scale_x_continuous(
    breaks = c(0.02, 0.3, 0.5),
    labels = c("0.02", "0.3", "0.5")
  ) +

  scale_fill_manual(
    values = c("0.02" = "#7e03a8",
               "0.3" = "#cc4778",
               "0.5" = "#f89540"),
    guide = "none"
  ) +

  scale_colour_manual(
    values = c("0.02" = "#7e03a8",
               "0.3" = "#cc4778",
               "0.5" = "#f89540"),
    guide = "none"
  ) +

  labs(
    x = "Ramping rate (°C/min)",
    y = "Critical thermal maximum (°C)"
  ) +
  base_theme +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(fill = NA, colour = "black", linewidth = 2.5)
  )

ggsave(file = "fig/CTmax.png", width = 8, height = 6, dpi = 500)
```


## Test CTmax assumptions

```{r}
# =========================================================
# CTmax-TDT conversion using posterior draws from:
# 1) fit_sublethal_tdt -> z uncertainty
# 2) fit_ctmax   -> CTmax uncertainty at each ramping rate

# =========================================================

# =========================================================
# 1) Settings
# =========================================================
To <- 17.5 # Starting temperature
Tc <- 26 # Assume damage starts accumulate above 26 degrees
tLs <- 60 # 1 hour
n_iter <- 1000
set.seed(123)

# =========================================================
# 2) Helper function
# =========================================================
pred_dCTmax_fun <- function(ramprate, tLs, sCTmax, z, To, Tc)
{
  inside <- (log(10) * ramprate * tLs / z) *
    exp((log(10) / z) * (sCTmax - To)) +
    exp((log(10) / z) * (Tc - To))

  out <- rep(NA_real_, length(ramprate))
  ok <- is.finite(inside) & inside > 0

  out[ok] <- To + (z / log(10)) * log(inside[ok])

  out
}

# objective function for one posterior draw
# uses posterior-estimated CTmax values at the 4 ramping rates
obj_sCTmax_fun <- function(sCTmax, obs_ctmax, ramprates, tLs, z, To, Tc)
{
  pred <- pred_dCTmax_fun(
    ramprate = ramprates,
    tLs = tLs,
    sCTmax = sCTmax,
    z = z,
    To = To,
    Tc = Tc
  )

  if(any(!is.finite(pred)))
  {
    return(1e12)
  }

  sum((obs_ctmax - pred)^2)
}

# =========================================================
# 3) Extract posterior draws of z from fit_sublethal_tdt
# z = -1 / slope
# =========================================================
draws_sub <- as.data.frame(posterior::as_draws_df(fit_sublethal_tdt))

z_draws_all <- draws_sub %>%
  dplyr::transmute(
    z = -1 / b_temp_c
  ) %>%
  dplyr::filter(
    is.finite(z),
    z > 0
  )

if(nrow(z_draws_all) == 0)
{
  stop("No valid posterior draws of z were found.")
}

# =========================================================
# 4) Posterior draws of CTmax from fit_ctmax
# population-level fitted values at the 3 ramping rates
# =========================================================
target_ramps_num <- c(0.02, 0.3, 0.5)
target_ramps_chr <- c("0.02", "0.3", "0.5")

nd_ctmax <- data.frame(
  Ramping = factor(target_ramps_chr, levels = levels(ctmax$Ramping)),
  start_min_c = median(ctmax$start_min_c, na.rm = TRUE),
  Trial_ID = factor(
    levels(ctmax$Trial_ID)[1],
    levels = levels(ctmax$Trial_ID)
  )
)

# posterior expected CTmax for each ramping rate
# dimensions: draws x 4
post_ctmax <- posterior_epred(
  fit_ctmax,
  newdata = nd_ctmax,
  re_formula = NA
)

# =========================================================
# 5) Build long posterior data for half-eye plot
# =========================================================
post_long <- data.frame(
  Ramping = rep(target_ramps_chr, each = nrow(post_ctmax)),
  CTmax_post = c(
    post_ctmax[, 1],
    post_ctmax[, 2],
    post_ctmax[, 3]  
    )
)

post_long$Ramping_num <- as.numeric(post_long$Ramping)

# =========================================================
# 6) Sample matched posterior iterations
# one z draw + one CTmax posterior draw per iteration
# =========================================================
n_iter_use <- min(n_iter, nrow(z_draws_all), nrow(post_ctmax))

z_idx <- sample(seq_len(nrow(z_draws_all)), size = n_iter_use, replace = FALSE)
ctmax_idx <- sample(seq_len(nrow(post_ctmax)), size = n_iter_use, replace = FALSE)

iter_df <- data.frame(
  iter = seq_len(n_iter_use),
  z = z_draws_all$z[z_idx],
  ctmax_002 = post_ctmax[ctmax_idx, 1],
  ctmax_03 = post_ctmax[ctmax_idx, 2],
  ctmax_05 = post_ctmax[ctmax_idx, 3]
)

# =========================================================
# 7) For each iteration, estimate sCTmax
# from that iteration's posterior CTmax values
# =========================================================
ramprates_fit <- c(0.02, 0.3, 0.5)

fit_list <- vector("list", nrow(iter_df))

for(i in seq_len(nrow(iter_df)))
{
  if(i %% 100 == 0) cat("Processed", i, "of", nrow(iter_df), "iterations\n")

  obs_ctmax_i <- c(
    iter_df$ctmax_002[i],
    iter_df$ctmax_03[i],
    iter_df$ctmax_05[i]
  )

  fit_try <- try(
    optimize(
      f = obj_sCTmax_fun,
      interval = c(30, 50),
      obs_ctmax = obs_ctmax_i,
      ramprates = ramprates_fit,
      tLs = tLs,
      z = iter_df$z[i],
      To = To,
      Tc = Tc
    ),
    silent = TRUE
  )

  if(inherits(fit_try, "try-error"))
  {
    fit_list[[i]] <- data.frame(
      iter = i,
      z = iter_df$z[i],
      sCTmax = NA_real_,
      converged = FALSE
    )
  } else {
    fit_list[[i]] <- data.frame(
      iter = i,
      z = iter_df$z[i],
      sCTmax = fit_try$minimum,
      converged = TRUE
    )
  }
}

scenario_fits <- do.call(rbind, fit_list) %>%
  dplyr::filter(converged, is.finite(sCTmax), is.finite(z))

if(nrow(scenario_fits) == 0)
{
  stop("All sCTmax optimisation attempts failed.")
}

# =========================================================
# 8) Summaries of propagated parameters
# =========================================================
fit_summary <- data.frame(
  t_coma = tLs,
  To = To,
  Tc = Tc,
  n_iter = nrow(scenario_fits),
  z_median = median(scenario_fits$z, na.rm = TRUE),
  z_lower = quantile(scenario_fits$z, 0.025, na.rm = TRUE),
  z_upper = quantile(scenario_fits$z, 0.975, na.rm = TRUE),
  sCTmax_median = median(scenario_fits$sCTmax, na.rm = TRUE),
  sCTmax_lower = quantile(scenario_fits$sCTmax, 0.025, na.rm = TRUE),
  sCTmax_upper = quantile(scenario_fits$sCTmax, 0.975, na.rm = TRUE)
)

print(fit_summary)

# =========================================================
# 9) Build posterior predictive CTmax curves
# =========================================================
ramp_grid <- data.frame(
  ramprate = seq(0.001, 0.70, length.out = 500)
)

pred_draw_list <- vector("list", nrow(scenario_fits))

for(i in seq_len(nrow(scenario_fits)))
{
  pred_draw_list[[i]] <- data.frame(
    iter = scenario_fits$iter[i],
    ramprate = ramp_grid$ramprate,
    CTmax_pred = pred_dCTmax_fun(
      ramprate = ramp_grid$ramprate,
      tLs = tLs,
      sCTmax = scenario_fits$sCTmax[i],
      z = scenario_fits$z[i],
      To = To,
      Tc = Tc
    )
  )
}

pred_curves_draws <- do.call(rbind, pred_draw_list)

pred_curves_sum <- pred_curves_draws %>%
  dplyr::group_by(ramprate) %>%
  dplyr::summarise(
    median = median(CTmax_pred, na.rm = TRUE),
    lower = quantile(CTmax_pred, 0.025, na.rm = TRUE),
    upper = quantile(CTmax_pred, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

# =========================================================
# 10) Raw CTmax data for plotting
# =========================================================
ctmax_plot <- ctmax %>%
  dplyr::filter(!is.na(CTmax), !is.na(Ramping)) %>%
  dplyr::mutate(
    Ramping_num = as.numeric(as.character(Ramping))
  )

# =========================================================
# 11) Plot
# =========================================================
p_tdt_compare <- ggplot() +
  
  geom_ribbon(
    data = pred_curves_sum,
    aes(
      x = ramprate,
      ymin = lower,
      ymax = upper
    ),
    inherit.aes = FALSE,
    fill = "black",
    alpha = 0.14
  ) +

  geom_line(
    data = pred_curves_sum,
    aes(
      x = ramprate,
      y = median
    ),
    inherit.aes = FALSE,
    linewidth = 1.4,
    colour = "black"
  ) +

  ggdist::stat_halfeye(
    data = post_long,
    aes(
      x = Ramping_num,
      y = CTmax_post,
      group = Ramping,
      fill = Ramping,
      colour = Ramping
    ),
    point_interval = ggdist::mean_qi,
    .width = 0.95,
    adjust = 0.9,
    width = 0.04,
    size = 5,
    slab_alpha = 0.3,
    point_size = 4,
    slab_colour = NA,
    justification = -0.01,
    position = position_nudge(x = 0.023)
  ) +

  geom_point(
    data = ctmax_plot,
    aes(
      x = Ramping_num,
      y = CTmax,
      fill= factor(Ramping)
    ),
    position = position_jitter(width = 0.008, height = 0.001),
    alpha = 0.9,
    size = 4,
    stroke = 1,
    col = "white",
    shape = 21,
  ) +

  coord_cartesian(
    ylim = c(28, 35),
    xlim = c(0, 0.60)
  ) +

  scale_x_continuous(
    breaks = c(0.02, 0.3, 0.5),
    labels = c("0.02", "0.3", "0.5")
  ) +

  scale_fill_manual(
    values = c(
      "0.02" = "#7e03a8",
      "0.3" = "#cc4778",
      "0.5" = "#f89540"
    ),
    guide = "none"
  ) +

  scale_colour_manual(
    values = c(
      "0.02" = "#7e03a8",
      "0.3" = "#cc4778",
      "0.5" = "#f89540"
    ),
    guide = "none"
  ) +

  labs(
    x = "Ramping rate (°C/min)",
    y = "Critical thermal maximum (°C)"
  ) +
  base_theme +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(fill = NA, colour = "black", linewidth = 2.5)
  )

print(p_tdt_compare)

ggsave(
  file = "fig/CTmax_vs_TDT_predictions_with_propagated_uncertainty.png",
  plot = p_tdt_compare,
  width = 9,
  height = 6,
  dpi = 500
)
```


### **Estimate z from ramping rate data** 

#### Using code from Jorgensen et al.
```{r}

# =========================================================
# 1) Inputs
# =========================================================
t_coma <- c(1, 10, 60)
extra_sCTmax <- c(31, 32, 33, 34)
extra_ramprates <- c(0.001, 0.01, 0.1, 1)

Tc <- 26
To <- 17.5

# =========================================================
# 2) Build summary dynamic data
# =========================================================
dynamic <- ctmax %>%
  dplyr::group_by(Ramping) %>%
  dplyr::summarise(
    dCTmax = mean(CTmax, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    ramprate = as.numeric(as.character(Ramping)),
    group = "all"
  )

# =========================================================
# 3) Core prediction function
# =========================================================
pred_dCTmax_fun <- function(ramprate, tLs, sCTmax, z, To, Tc)
{
  To + (z / log(10)) * log(
    (log(10) * ramprate * tLs / z) *
      exp((log(10) / z) * (sCTmax - To)) +
      exp((log(10) / z) * (Tc - To))
  )
}

# =========================================================
# 4) Original nls() check
# =========================================================

dtmp <- dynamic
ramprates <- unique(dtmp$ramprate)

nls_check_results <- data.frame(
  t_coma = t_coma,
  sCTmax_pred = NA_real_,
  z_pred = NA_real_,
  status = NA_character_
)

for(j in seq_along(t_coma))
{
  tLs <- t_coma[j]

  fit_try <- try(
    nls(
      dCTmax ~ To + (z / log(10)) * log(
        (log(10) * ramprate * tLs / z) *
          exp((log(10) / z) * (sCTmax - To)) +
          exp((log(10) / z) * (Tc - To))
      ),
      start = list(sCTmax = 31.5, z = 1.06),
      data = dtmp
    ),
    silent = TRUE
  )

  if(inherits(fit_try, "try-error"))
  {
    err_msg <- as.character(fit_try)
    nls_check_results$status[j] <- err_msg
    cat("\nOriginal nls failed for t_coma =", tLs, "\n")
    cat(err_msg, "\n")
  } else {
    nls_check_results$sCTmax_pred[j] <- as.numeric(fit_try$m$getPars()[1])
    nls_check_results$z_pred[j] <- as.numeric(fit_try$m$getPars()[2])
    nls_check_results$status[j] <- "ok"

    cat("\nOriginal nls succeeded for t_coma =", tLs, "\n")
    cat("  sCTmax =", nls_check_results$sCTmax_pred[j], "\n")
    cat("  z =", nls_check_results$z_pred[j], "\n")
  }
} # Pretty close to z estimated from the data! 
```


#### Bayesian model adaptation
```{r}

# =========================================================
# Bayesian adaptation of Jorgensen et al. for brown shrimp
# using:
#   - fit_sublethal_tdt
#   - fit_ctmax
# and propagating uncertainty in both directions
# =========================================================

# =========================================================
# 1) Settings
# =========================================================
To <- 17.5
Tc <- 26
t_coma <- c(1, 10, 30, 60, 120, 240, 480)
tLs_plot <- 60
n_iter <- 1000
set.seed(123)

target_ramps_num <- c(0.02, 0.3, 0.5)
target_ramps_chr <- c("0.02", "0.3", "0.5")

# =========================================================
# 2) Helper function: Jorgensen dynamic prediction
# =========================================================
pred_dCTmax_fun <- function(ramprate, tLs, sCTmax, z, To, Tc)
{
  inside <- (log(10) * ramprate * tLs / z) *
    exp((log(10) / z) * (sCTmax - To)) +
    exp((log(10) / z) * (Tc - To))
  
  out <- rep(NA_real_, length(ramprate))
  ok <- is.finite(inside) & inside > 0
  
  out[ok] <- To + (z / log(10)) * log(inside[ok])
  out
}

# =========================================================
# 3) nls helpers
# =========================================================
fit_nls_fixed_z <- function(obs_ctmax, ramprates, tLs, z, To, Tc)
{
  dtmp <- data.frame(
    dCTmax = obs_ctmax,
    ramprate = ramprates
  )
  
  start_vals <- c(max(obs_ctmax, na.rm = TRUE) + 0.25,
                  max(obs_ctmax, na.rm = TRUE) + 0.75,
                  35)
  
  for(s0 in start_vals)
  {
    fit_try <- try(
      nls(
        dCTmax ~ To + (z / log(10)) * log(
          (log(10) * ramprate * tLs / z) *
            exp((log(10) / z) * (sCTmax - To)) +
            exp((log(10) / z) * (Tc - To))
        ),
        start = list(sCTmax = s0),
        data = dtmp,
        control = nls.control(
          maxiter = 1000,
          warnOnly = TRUE
        )
      ),
      silent = TRUE
    )
    
    if(!inherits(fit_try, "try-error"))
    {
      s_est <- as.numeric(coef(fit_try)[["sCTmax"]])
      if(is.finite(s_est))
      {
        return(list(
          converged = TRUE,
          sCTmax = s_est
        ))
      }
    }
  }
  
  list(
    converged = FALSE,
    sCTmax = NA_real_
  )
}

fit_nls_joint <- function(obs_ctmax, ramprates, tLs, To, Tc)
{
  dtmp <- data.frame(
    dCTmax = obs_ctmax,
    ramprate = ramprates
  )
  
  start_grid <- expand.grid(
    sCTmax = c(max(obs_ctmax, na.rm = TRUE) + 0.25, 32, 35),
    z = c(0.7, 1.0, 1.5, 2.5)
  )
  
  for(k in seq_len(nrow(start_grid)))
  {
    fit_try <- try(
      nls(
        dCTmax ~ To + (z / log(10)) * log(
          (log(10) * ramprate * tLs / z) *
            exp((log(10) / z) * (sCTmax - To)) +
            exp((log(10) / z) * (Tc - To))
        ),
        start = list(
          sCTmax = start_grid$sCTmax[k],
          z = start_grid$z[k]
        ),
        data = dtmp,
        control = nls.control(
          maxiter = 1000,
          warnOnly = TRUE
        )
      ),
      silent = TRUE
    )
    
    if(!inherits(fit_try, "try-error"))
    {
      pars <- coef(fit_try)
      s_est <- as.numeric(pars[["sCTmax"]])
      z_est <- as.numeric(pars[["z"]])
      
      if(is.finite(s_est) && is.finite(z_est) && z_est > 0)
      {
        return(list(
          converged = TRUE,
          sCTmax = s_est,
          z = z_est
        ))
      }
    }
  }
  
  list(
    converged = FALSE,
    sCTmax = NA_real_,
    z = NA_real_
  )
}

# =========================================================
# 4) Posterior draws from sublethal TDT model
# model: log10_time_to_death ~ temp_c + ...
# hence z = -1 / slope
# =========================================================
draws_sub <- as.data.frame(posterior::as_draws_df(fit_sublethal_tdt))

if(!("b_temp_c" %in% names(draws_sub)))
{
  stop("Could not find 'b_temp_c' in fit_sublethal_tdt posterior draws.")
}

z_draws_all <- draws_sub %>%
  dplyr::transmute(
    z = -1 / b_temp_c
  ) %>%
  dplyr::filter(
    is.finite(z),
    z > 0
  )

if(nrow(z_draws_all) == 0)
{
  stop("No valid posterior draws of z were found from fit_sublethal_tdt.")
}

# =========================================================
# 5) Posterior TDT predictions directly from fit_sublethal_tdt
# this is the sublethal TDT ribbon for plot 1
# =========================================================
temps_sub_df <- data.frame(
  assay_temp = seq(30.5, 34, by = 0.01)
) %>%
  dplyr::mutate(
    temp_c = assay_temp - temp_mean,
    date_experiment = factor(
      levels(tdt_sub$date_experiment)[1],
      levels = levels(tdt_sub$date_experiment)
    ),
    tank_ID = factor(
      levels(tdt_sub$tank_ID)[1],
      levels = levels(tdt_sub$tank_ID)
    ),
    cup_ID = factor(
      levels(tdt_sub$cup_ID)[1],
      levels = levels(tdt_sub$cup_ID)
    )
  )

lp_sub <- posterior_linpred(
  fit_sublethal_tdt,
  newdata = temps_sub_df,
  re_formula = NA
)

# lp_sub is on log10(minutes) scale
sublethal_tdt_draws <- do.call(
  rbind,
  lapply(seq_len(nrow(lp_sub)), function(i) {
    data.frame(
      iter = i,
      assay_temp = temps_sub_df$assay_temp,
      log10_time = lp_sub[i, ],
      time_min = 10^lp_sub[i, ],
      source = "Sublethal TDT"
    )
  })
)

sublethal_tdt_sum <- sublethal_tdt_draws %>%
  dplyr::group_by(assay_temp, source) %>%
  dplyr::summarise(
    median_log10 = median(log10_time, na.rm = TRUE),
    lower_log10 = quantile(log10_time, 0.025, na.rm = TRUE),
    upper_log10 = quantile(log10_time, 0.975, na.rm = TRUE),
    median_time = median(time_min, na.rm = TRUE),
    lower_time = quantile(time_min, 0.025, na.rm = TRUE),
    upper_time = quantile(time_min, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

# =========================================================
# 6) Posterior expected CTmax from fit_ctmax
# =========================================================
nd_ctmax <- data.frame(
  Ramping = factor(target_ramps_chr, levels = levels(ctmax$Ramping))
)

post_ctmax <- posterior_epred(
  fit_ctmax,
  newdata = nd_ctmax,
  re_formula = NA
)

if(ncol(post_ctmax) != length(target_ramps_num))
{
  stop("Unexpected number of columns in post_ctmax.")
}

post_long <- data.frame(
  Ramping = rep(target_ramps_chr, each = nrow(post_ctmax)),
  CTmax_post = c(
    post_ctmax[, 1],
    post_ctmax[, 2],
    post_ctmax[, 3]
  )
)

post_long$Ramping_num <- as.numeric(post_long$Ramping)

# =========================================================
# 7) Match posterior iterations
# =========================================================
n_iter_use <- min(n_iter, nrow(z_draws_all), nrow(post_ctmax))

z_idx <- sample(seq_len(nrow(z_draws_all)), size = n_iter_use, replace = FALSE)
ctmax_idx <- sample(seq_len(nrow(post_ctmax)), size = n_iter_use, replace = FALSE)

iter_df <- data.frame(
  iter = seq_len(n_iter_use),
  z_sub = z_draws_all$z[z_idx],
  ctmax_002 = post_ctmax[ctmax_idx, 1],
  ctmax_03 = post_ctmax[ctmax_idx, 2],
  ctmax_05 = post_ctmax[ctmax_idx, 3]
)

ramprates_fit <- target_ramps_num

# =========================================================
# 8) Forward branch:
# sublethal TDT -> dynamic CTmax
# fixed z from fit_sublethal_tdt, estimate sCTmax for each t_coma
# =========================================================
forward_fit_list <- vector("list", nrow(iter_df) * length(t_coma))
counter <- 1

for(i in seq_len(nrow(iter_df)))
{
  if(i %% 100 == 0) cat("Forward branch:", i, "of", nrow(iter_df), "\n")
  
  obs_ctmax_i <- c(
    iter_df$ctmax_002[i],
    iter_df$ctmax_03[i],
    iter_df$ctmax_05[i]
  )
  
  for(j in seq_along(t_coma))
  {
    tLs_j <- t_coma[j]
    
    fit_out <- fit_nls_fixed_z(
      obs_ctmax = obs_ctmax_i,
      ramprates = ramprates_fit,
      tLs = tLs_j,
      z = iter_df$z_sub[i],
      To = To,
      Tc = Tc
    )
    
    forward_fit_list[[counter]] <- data.frame(
      iter = iter_df$iter[i],
      t_coma = tLs_j,
      z = iter_df$z_sub[i],
      sCTmax_pred = fit_out$sCTmax,
      converged = fit_out$converged
    )
    
    counter <- counter + 1
  }
}

forward_fits <- do.call(rbind, forward_fit_list) %>%
  dplyr::filter(converged, is.finite(z), is.finite(sCTmax_pred))

if(nrow(forward_fits) == 0)
{
  stop("All forward fits failed.")
}

forward_summary <- forward_fits %>%
  dplyr::group_by(t_coma) %>%
  dplyr::summarise(
    n_iter = dplyr::n(),
    z_median = median(z, na.rm = TRUE),
    z_lower = quantile(z, 0.025, na.rm = TRUE),
    z_upper = quantile(z, 0.975, na.rm = TRUE),
    sCTmax_median = median(sCTmax_pred, na.rm = TRUE),
    sCTmax_lower = quantile(sCTmax_pred, 0.025, na.rm = TRUE),
    sCTmax_upper = quantile(sCTmax_pred, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

print(forward_summary)

# dynamic CTmax ribbon from sublethal TDT branch
ramp_grid <- data.frame(
  ramprate = seq(0.001, 0.70, length.out = 500)
)

forward_plot_fits <- forward_fits %>%
  dplyr::filter(t_coma == tLs_plot)

forward_curve_draws <- do.call(
  rbind,
  lapply(seq_len(nrow(forward_plot_fits)), function(i) {
    data.frame(
      iter = forward_plot_fits$iter[i],
      ramprate = ramp_grid$ramprate,
      CTmax_pred = pred_dCTmax_fun(
        ramprate = ramp_grid$ramprate,
        tLs = tLs_plot,
        sCTmax = forward_plot_fits$sCTmax_pred[i],
        z = forward_plot_fits$z[i],
        To = To,
        Tc = Tc
      ),
      source = "Sublethal TDT"
    )
  })
)

forward_curve_sum <- forward_curve_draws %>%
  dplyr::group_by(ramprate, source) %>%
  dplyr::summarise(
    median = median(CTmax_pred, na.rm = TRUE),
    lower = quantile(CTmax_pred, 0.025, na.rm = TRUE),
    upper = quantile(CTmax_pred, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

# =========================================================
# 9) Inverse branch:
# dynamic CTmax -> TDT
# estimate both z and sCTmax for each t_coma, then fit TDT line
# =========================================================
inverse_fit_list <- vector("list", nrow(iter_df) * length(t_coma))
counter <- 1

for(i in seq_len(nrow(iter_df)))
{
  if(i %% 100 == 0) cat("Inverse branch:", i, "of", nrow(iter_df), "\n")
  
  obs_ctmax_i <- c(
    iter_df$ctmax_002[i],
    iter_df$ctmax_03[i],
    iter_df$ctmax_05[i]
  )
  
  for(j in seq_along(t_coma))
  {
    tLs_j <- t_coma[j]
    
    fit_out <- fit_nls_joint(
      obs_ctmax = obs_ctmax_i,
      ramprates = ramprates_fit,
      tLs = tLs_j,
      To = To,
      Tc = Tc
    )
    
    inverse_fit_list[[counter]] <- data.frame(
      iter = iter_df$iter[i],
      t_coma = tLs_j,
      z_pred = fit_out$z,
      sCTmax_pred = fit_out$sCTmax,
      converged = fit_out$converged
    )
    
    counter <- counter + 1
  }
}

inverse_fits <- do.call(rbind, inverse_fit_list) %>%
  dplyr::filter(converged, is.finite(z_pred), z_pred > 0, is.finite(sCTmax_pred))

if(nrow(inverse_fits) == 0)
{
  stop("All inverse fits failed.")
}

inverse_summary <- inverse_fits %>%
  dplyr::group_by(t_coma) %>%
  dplyr::summarise(
    n_iter = dplyr::n(),
    z_median = median(z_pred, na.rm = TRUE),
    z_lower = quantile(z_pred, 0.025, na.rm = TRUE),
    z_upper = quantile(z_pred, 0.975, na.rm = TRUE),
    sCTmax_median = median(sCTmax_pred, na.rm = TRUE),
    sCTmax_lower = quantile(sCTmax_pred, 0.025, na.rm = TRUE),
    sCTmax_upper = quantile(sCTmax_pred, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

print(inverse_summary)

# Reconstruct TDT line within each posterior draw
inverse_tdt_draws <- do.call(
  rbind,
  lapply(seq_len(n_iter_use), function(i) {
    tmp_i <- inverse_fits %>%
      dplyr::filter(iter == i) %>%
      dplyr::arrange(t_coma)
    
    if(nrow(tmp_i) < 2)
    {
      return(NULL)
    }
    
    tdt_lm <- lm(log10(t_coma) ~ sCTmax_pred, data = tmp_i)
    
    pred_log10 <- predict(
      tdt_lm,
      newdata = data.frame(sCTmax_pred = temps_sub_df$assay_temp)
    )
    
    data.frame(
      iter = i,
      assay_temp = temps_sub_df$assay_temp,
      log10_time = pred_log10,
      time_min = 10^pred_log10,
      source = "Derived from ramping"
    )
  })
)

if(is.null(inverse_tdt_draws) || nrow(inverse_tdt_draws) == 0)
{
  stop("No ramp-derived TDT lines could be reconstructed.")
}

inverse_tdt_sum <- inverse_tdt_draws %>%
  dplyr::group_by(assay_temp, source) %>%
  dplyr::summarise(
    median_log10 = median(log10_time, na.rm = TRUE),
    lower_log10 = quantile(log10_time, 0.025, na.rm = TRUE),
    upper_log10 = quantile(log10_time, 0.975, na.rm = TRUE),
    median_time = median(time_min, na.rm = TRUE),
    lower_time = quantile(time_min, 0.025, na.rm = TRUE),
    upper_time = quantile(time_min, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

# dynamic CTmax ribbon from ramp-derived TDT branch
inverse_plot_fits <- inverse_fits %>%
  dplyr::filter(t_coma == tLs_plot)

inverse_curve_draws <- do.call(
  rbind,
  lapply(seq_len(nrow(inverse_plot_fits)), function(i) {
    data.frame(
      iter = inverse_plot_fits$iter[i],
      ramprate = ramp_grid$ramprate,
      CTmax_pred = pred_dCTmax_fun(
        ramprate = ramp_grid$ramprate,
        tLs = tLs_plot,
        sCTmax = inverse_plot_fits$sCTmax_pred[i],
        z = inverse_plot_fits$z_pred[i],
        To = To,
        Tc = Tc
      ),
      source = "Derived from ramping"
    )
  })
)

inverse_curve_sum <- inverse_curve_draws %>%
  dplyr::group_by(ramprate, source) %>%
  dplyr::summarise(
    median = median(CTmax_pred, na.rm = TRUE),
    lower = quantile(CTmax_pred, 0.025, na.rm = TRUE),
    upper = quantile(CTmax_pred, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

# =========================================================
# 10) Combine for plotting
# =========================================================
tdt_compare_sum <- dplyr::bind_rows(
  sublethal_tdt_sum,
  inverse_tdt_sum
)

ctmax_compare_sum <- dplyr::bind_rows(
  forward_curve_sum,
  inverse_curve_sum
)

ctmax_plot <- ctmax %>%
  dplyr::filter(!is.na(CTmax), !is.na(Ramping)) %>%
  dplyr::mutate(
    Ramping_num = as.numeric(as.character(Ramping))
  )

# =========================================================
# 11) Plot 1: TDT comparison
# =========================================================
p_tdt_compare <- ggplot() +
  geom_ribbon(
    data = tdt_compare_sum %>% dplyr::filter(source == "Sublethal TDT"),
    aes(x = assay_temp, ymin = lower_time, ymax = upper_time),
    fill = "black",
    alpha = 0.14
  ) +
  geom_line(
    data = tdt_compare_sum %>% dplyr::filter(source == "Sublethal TDT"),
    aes(x = assay_temp, y = median_time),
    linewidth = 1.4,
    colour = "black"
  ) +
  geom_ribbon(
    data = tdt_compare_sum %>% dplyr::filter(source == "Derived from ramping"),
    aes(x = assay_temp, ymin = lower_time, ymax = upper_time),
    fill = "grey40",
    alpha = 0.18
  ) +
  geom_line(
    data = tdt_compare_sum %>% dplyr::filter(source == "Derived from ramping"),
    aes(x = assay_temp, y = median_time),
    linewidth = 1.4,
    colour = "grey20",
    linetype = "dashed"
  ) +
  geom_jitter(
    data = tdt_sub,
    aes(x = assay_temp, y = time_to_death),
    width = 0.05,
    height = 0,
    size = 4,
    alpha = 0.75,
    shape = 21,
    col = "white",
    fill = "black"
  ) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(breaks = ybreaks_right, labels = lab_number) +
  coord_cartesian(
    xlim = c(30.5, 34),
    ylim = c(0.2, 480)
  ) +
  labs(
    x = "Assay temperature (°C)",
    y = "Time to death (min)"
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

print(p_tdt_compare)

# =========================================================
# 12) Plot 2: CTmax comparison
# =========================================================
p_ctmax_compare <- ggplot() +
  geom_ribbon(
    data = ctmax_compare_sum %>% dplyr::filter(source == "Sublethal TDT"),
    aes(x = ramprate, ymin = lower, ymax = upper),
    fill = "black",
    alpha = 0.14
  ) +
  geom_line(
    data = ctmax_compare_sum %>% dplyr::filter(source == "Sublethal TDT"),
    aes(x = ramprate, y = median),
    linewidth = 1.4,
    colour = "black"
  ) +
  geom_ribbon(
    data = ctmax_compare_sum %>% dplyr::filter(source == "Derived from ramping"),
    aes(x = ramprate, ymin = lower, ymax = upper),
    fill = "grey40",
    alpha = 0.18
  ) +
  geom_line(
    data = ctmax_compare_sum %>% dplyr::filter(source == "Derived from ramping"),
    aes(x = ramprate, y = median),
    linewidth = 1.4,
    colour = "grey20",
    linetype = "dashed"
  ) +
  ggdist::stat_halfeye(
    data = post_long,
    aes(
      x = Ramping_num,
      y = CTmax_post,
      group = Ramping,
      fill = Ramping,
      colour = Ramping
    ),
    point_interval = ggdist::mean_qi,
    .width = 0.95,
    adjust = 0.9,
    width = 0.04,
    size = 5,
    slab_alpha = 0.3,
    point_size = 4,
    slab_colour = NA,
    justification = -0.01,
    position = position_nudge(x = 0.023)
  ) +
  geom_point(
    data = ctmax_plot,
    aes(
      x = Ramping_num,
      y = CTmax,
      fill = factor(Ramping)
    ),
    position = position_jitter(width = 0.008, height = 0.001),
    alpha = 0.9,
    size = 4,
    stroke = 1,
    col = "white",
    shape = 21
  ) +
  coord_cartesian(
    ylim = c(28, 35),
    xlim = c(0, 0.60)
  ) +
  scale_x_continuous(
    breaks = c(0.02, 0.3, 0.5),
    labels = c("0.02", "0.3", "0.5")
  ) +
  scale_fill_manual(
    values = c(
      "0.02" = "#7e03a8",
      "0.3" = "#cc4778",
      "0.5" = "#f89540"
    ),
    guide = "none"
  ) +
  scale_colour_manual(
    values = c(
      "0.02" = "#7e03a8",
      "0.3" = "#cc4778",
      "0.5" = "#f89540"
    ),
    guide = "none"
  ) +
  labs(
    x = "Ramping rate (°C/min)",
    y = "Critical thermal maximum (°C)"
  ) +
  base_theme +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(fill = NA, colour = "black", linewidth = 2.5)
  )

print(p_ctmax_compare)

# =========================================================
# 13) Save plots
# =========================================================
ggsave(
  file = "fig/TDT_sublethal_vs_ramping_with_uncertainty.png",
  plot = p_tdt_compare,
  width = 9,
  height = 6,
  dpi = 500
)

ggsave(
  file = "fig/CTmax_predictions_sublethal_vs_ramping_with_uncertainty.png",
  plot = p_ctmax_compare,
  width = 9,
  height = 6,
  dpi = 500
)


# =========================
# Compare estimates obtained under both methods
# =========================

# =========================
# Direct static TDT summary
# =========================
static_summary <- draws_sub %>%
  dplyr::transmute(
    slope = b_temp_c,
    z = -1 / b_temp_c
  ) %>%
  dplyr::summarise(
    slope_median = median(slope, na.rm = TRUE),
    slope_lower = quantile(slope, 0.025, na.rm = TRUE),
    slope_upper = quantile(slope, 0.975, na.rm = TRUE),
    z_median = median(z, na.rm = TRUE),
    z_lower = quantile(z, 0.025, na.rm = TRUE),
    z_upper = quantile(z, 0.975, na.rm = TRUE)
  )

# =========================
# Dynamic to static:
# infer TDT slope from CTmax/ramping data
# =========================
dyn_to_static_coef <- lapply(seq_len(n_iter_use), function(i) {
  tmp_i <- inverse_fits %>%
    dplyr::filter(iter == i) %>%
    dplyr::arrange(t_coma)
  
  if(nrow(tmp_i) < 2) return(NULL)
  
  mod <- lm(log10(t_coma) ~ sCTmax_pred, data = tmp_i)
  cf <- coef(mod)
  
  data.frame(
    iter = i,
    slope = unname(cf[2])
  )
}) %>%
  do.call(rbind, .)

dyn_to_static_summary <- dyn_to_static_coef %>%
  dplyr::summarise(
    slope_median = median(slope, na.rm = TRUE),
    slope_lower = quantile(slope, 0.025, na.rm = TRUE),
    slope_upper = quantile(slope, 0.975, na.rm = TRUE)
  )

# =========================
# Final compact table
# =========================
t_keep <- 60
summary_compare <- dplyr::bind_rows(
  data.frame(
    approach = "Static",
    parameter = c("slope", "z"),
    median = c(
      static_summary$slope_median,
      static_summary$z_median
    ),
    lower = c(
      static_summary$slope_lower,
      static_summary$z_lower
    ),
    upper = c(
      static_summary$slope_upper,
      static_summary$z_upper
    )
  ),
  data.frame(
    approach = "Static to dynamic",
    parameter = "sCTmax",
    median = forward_summary$sCTmax_median[forward_summary$t_coma == t_keep],
    lower = forward_summary$sCTmax_lower[forward_summary$t_coma == t_keep],
    upper = forward_summary$sCTmax_upper[forward_summary$t_coma == t_keep]
  ),
  data.frame(
    approach = "Dynamic to static",
    parameter = c("slope", "z", "sCTmax"),
    median = c(
      dyn_to_static_summary$slope_median,
      inverse_summary$z_median[inverse_summary$t_coma == t_keep],
      inverse_summary$sCTmax_median[inverse_summary$t_coma == t_keep]
    ),
    lower = c(
      dyn_to_static_summary$slope_lower,
      inverse_summary$z_lower[inverse_summary$t_coma == t_keep],
      inverse_summary$sCTmax_lower[inverse_summary$t_coma == t_keep]
    ),
    upper = c(
      dyn_to_static_summary$slope_upper,
      inverse_summary$z_upper[inverse_summary$t_coma == t_keep],
      inverse_summary$sCTmax_upper[inverse_summary$t_coma == t_keep]
    )
  )
) %>%
  dplyr::mutate(
    median = round(median, 3),
    lower = round(lower, 3),
    upper = round(upper, 3),
    parameter = dplyr::recode(parameter,
      "sCTmax" = "sCTmax (1 hour)")
  )

print(summary_compare)



################# Compare z #####################

# Static z
z_static <- -1 / draws_sub$b_temp_c

# Dynamic z (inverse)
z_dynamic <- inverse_fits %>%
  dplyr::filter(t_coma == t_keep) %>%
  dplyr::pull(z_pred)

# Match lengths
n_min <- min(length(z_static), length(z_dynamic))

z_static_s <- sample(z_static, n_min)
z_dynamic_s <- sample(z_dynamic, n_min)

# Compute difference
z_diff <- z_static_s - z_dynamic_s

z_test <- data.frame(
  median = median(z_diff),
  lower = quantile(z_diff, 0.025),
  upper = quantile(z_diff, 0.975)
)

print(z_test) # Not statistically different; but lower z on average.
```

# **Comparison with environmental temperatures** 

```{r}
# Load temperature data from the last five years
files <- list.files("data/field_temp", pattern = "\\.csv$", full.names = TRUE)

field_temp <- files %>%
  lapply(read.csv) %>%
  do.call(rbind, .)

# Replace "no data" with NA in all columns
field_temp[field_temp == "no data"] <- NA

# Rename columns
names(field_temp) <- c(
  "timestamp_utc",
  "local_time_cet",
  "salinity_1m_sjoboden",
  "temp_1m_sjoboden",
  "air_temp",
  "sea_level_mm",
  "sea_temp",
  "deep_water_salinity_lab",
  "deep_water_temp_lab",
  "salinity_1m_ctd",
  "surface_water_salinity_lab",
  "surface_water_temp",
  "temp_1m_ctd"
)

names(field_temp)[is.na(names(field_temp))] <- "tmp"

field_temp$timestamp_utc <- as.POSIXct(field_temp$timestamp_utc, tz = "UTC")
field_temp$sea_temp <- as.numeric(field_temp$sea_temp)
field_temp$surface_water_temp <- as.numeric(field_temp$surface_water_temp)

field_temp <- field_temp %>% dplyr::filter(surface_water_temp < 100)

# Year separators
year_lines <- data.frame(
  x = seq(
    as.POSIXct(paste0(format(min(field_temp$timestamp_utc, na.rm = TRUE), "%Y"), "-01-01"), tz = "UTC"),
    as.POSIXct(paste0(format(max(field_temp$timestamp_utc, na.rm = TRUE), "%Y"), "-01-01"), tz = "UTC"),
    by = "year"
  )
)

# Plot temperature time series
ggplot(field_temp, aes(x = timestamp_utc)) +
  geom_line(aes(y = sea_temp, colour = "sea_temp"), linewidth = 0.7) +
  geom_vline(data = year_lines, aes(xintercept = x), linetype = 2, colour = "grey60") +
  labs(
    x = "Date",
    y = "Temperature (°C)",
    colour = NULL
  ) +
  theme_classic(base_size = 16)

max(field_temp$sea_temp, na.rm = T) # 25.1 degrees is the maximum temperature recorded in the past 5 years. 


## ====================================
# Temperatures during the study period
# ====================================4

field_temp <- field_temp %>%
  dplyr::filter(
    format(timestamp_utc, "%Y-%m-%d") >= "2025-06-12" &
    format(timestamp_utc, "%Y-%m-%d") <= "2025-06-30"
  )

field_temp %>% summarise(mean_temp = mean(surface_water_temp), 
                         sd_temp = sd(surface_water_temp),
                         mean_salinity = mean(as.numeric(surface_water_salinity_lab)),
                         sd_salinity = sd(as.numeric(surface_water_salinity_lab)))

```

### **Predicted accumulation of heat injury** 

```{r}
# =========================
# Libraries
# =========================
library(dplyr)
library(posterior)
library(future.apply)

# =========================
# 1) Extract posterior parameter table once
# =========================
extract_4pl_pars <- function(fit_4pl,
                             file_out = "Rdata/fit_4pl_pars.RDS",
                             overwrite = FALSE) {
  
  if(file.exists(file_out) && !overwrite) {
    return(readRDS(file_out))
  }
  
  dr <- posterior::as_draws_df(fit_4pl) %>%
    as.data.frame()
  
  get_par <- function(nm) {
    if(nm %in% names(dr)) {
      dr[[nm]]
    } else {
      rep(0, nrow(dr))
    }
  }
  
  pars <- data.frame(
    .draw = dr$.draw,
    low = plogis(get_par("b_lraw_Intercept")) * 0.49,
    up = 0.51 + plogis(get_par("b_uraw_Intercept")) * 0.49,
    k = exp(get_par("b_logk_Intercept")),
    mid_int = get_par("b_mid_Intercept"),
    mid_temp = get_par("b_mid_temp_c")
  ) %>%
    dplyr::filter(
      is.finite(low),
      is.finite(up),
      is.finite(k),
      is.finite(mid_int),
      is.finite(mid_temp),
      k > 0,
      up > low
    )
  
  saveRDS(pars, file_out)
  pars
}

# =========================
# 2) Core predictor for one scenario
#    summary only by default
# =========================
predict_survival_fluctuating_4pl_fast <- function(temp_df,
                                                  pars,
                                                  temp_mean,
                                                  n_draws = 1000,
                                                  surv_grid = seq(0.99, 0.01, by = -0.05),
                                                  save_draws = FALSE) {
  
  temp_df <- temp_df %>%
    dplyr::arrange(datetime)
  
  if(nrow(temp_df) < 2) {
    stop("temp_df must contain at least 2 rows.")
  }
  
  n_use <- min(n_draws, nrow(pars))
  pars_use <- pars %>%
    dplyr::slice_sample(n = n_use)
  
  dt_h <- as.numeric(diff(temp_df$datetime), units = "hours")
  temp0 <- temp_df$temp[-nrow(temp_df)]
  temp1 <- temp_df$temp[-1]
  tempc0 <- temp0 - temp_mean
  tempc1 <- temp1 - temp_mean
  
  pred_list <- vector("list", nrow(pars_use))
  
  for(i in seq_len(nrow(pars_use))) {
    
    low_i <- pars_use$low[i]
    up_i <- pars_use$up[i]
    k_i <- pars_use$k[i]
    mid0_i <- pars_use$mid_int[i] + pars_use$mid_temp[i] * tempc0
    mid1_i <- pars_use$mid_int[i] + pars_use$mid_temp[i] * tempc1
    
    valid_s <- surv_grid[surv_grid > low_i & surv_grid < up_i]
    
    if(length(valid_s) == 0) {
      next
    }
    
    surv_mat <- matrix(NA_real_,
                       nrow = nrow(temp_df),
                       ncol = length(valid_s))
    
    for(j in seq_along(valid_s)) {
      s_j <- valid_s[j]
      
      log10_t0 <- mid0_i + log((up_i - s_j) / (s_j - low_i)) / k_i
      log10_t1 <- mid1_i + log((up_i - s_j) / (s_j - low_i)) / k_i
      
      t0_h <- 10^log10_t0
      t1_h <- 10^log10_t1
      
      rate0 <- 1 / t0_h
      rate1 <- 1 / t1_h
      
      cum_injury <- c(0, cumsum(dt_h * (rate0 + rate1) / 2))
      surv_mat[, j] <- ifelse(cum_injury < 1, s_j, NA_real_)
    }
    
    pred_survival <- apply(surv_mat, 1, function(x) {
      if(all(is.na(x))) low_i else max(x, na.rm = TRUE)
    })
    
    pred_list[[i]] <- data.frame(
      .draw = pars_use$.draw[i],
      datetime = temp_df$datetime,
      temp = temp_df$temp,
      pred_survival = pred_survival
    )
  }
  
  pred_draws <- dplyr::bind_rows(pred_list)
  
  pred_summary <- pred_draws %>%
    dplyr::group_by(datetime, temp) %>%
    dplyr::summarise(
      survival_median = median(pred_survival, na.rm = TRUE),
      survival_lower = quantile(pred_survival, 0.025, na.rm = TRUE),
      survival_upper = quantile(pred_survival, 0.975, na.rm = TRUE),
      mortality_median = 1 - median(pred_survival, na.rm = TRUE),
      mortality_lower = 1 - quantile(pred_survival, 0.975, na.rm = TRUE),
      mortality_upper = 1 - quantile(pred_survival, 0.025, na.rm = TRUE),
      .groups = "drop"
    )
  
  if(save_draws) {
    list(summary = pred_summary, draws = pred_draws)
  } else {
    list(summary = pred_summary)
  }
}

# =========================
# 3) Wrapper to run multiple scenarios
#    in parallel and save RDS outputs
# =========================
run_fluctuating_scenarios <- function(field_temp,
                                      fit_4pl,
                                      temp_mean,
                                      n_draws = 1000,
                                      surv_grid = seq(0.99, 0.01, by = -0.05),
                                      workers = 4,
                                      overwrite = FALSE,
                                      save_draws = FALSE) {
  
  dir.create("Rdata", showWarnings = FALSE)
  
  pars <- extract_4pl_pars(
    fit_4pl = fit_4pl,
    file_out = "Rdata/fit_4pl_pars.RDS",
    overwrite = overwrite
  )
  
  temp_df <- field_temp %>%
    dplyr::select(datetime = timestamp_utc,
                  temp = surface_water_temp) %>%
    dplyr::filter(!is.na(temp)) %>%
    dplyr::arrange(datetime)
  
  temp_scenarios <- list(
    current = temp_df,
    plus1 = temp_df %>% dplyr::mutate(temp = temp + 1),
    plus2 = temp_df %>% dplyr::mutate(temp = temp + 2),
    plus3 = temp_df %>% dplyr::mutate(temp = temp + 3)
  )
  
  future::plan(future::multisession, workers = workers)
  
  out_list <- future.apply::future_lapply(names(temp_scenarios), function(nm) {
    
    file_out <- paste0("Rdata/fluctuating_", nm, "_", n_draws, "draws.RDS")
    
    if(file.exists(file_out) && !overwrite) {
      return(readRDS(file_out))
    }
    
    out <- predict_survival_fluctuating_4pl_fast(
      temp_df = temp_scenarios[[nm]],
      pars = pars,
      temp_mean = temp_mean,
      n_draws = n_draws,
      surv_grid = surv_grid,
      save_draws = save_draws
    )
    
    out$summary$scenario <- nm
    saveRDS(out, file_out)
    out
  })
  
  future::plan(future::sequential)
  
  pred_all <- dplyr::bind_rows(lapply(out_list, function(x) x$summary))
  
  saveRDS(pred_all, "Rdata/fluctuating_all_summaries.RDS")
  
  summary_end <- pred_all %>%
    dplyr::group_by(scenario) %>%
    dplyr::slice_tail(n = 1) %>%
    dplyr::ungroup()
  
  saveRDS(summary_end, "Rdata/fluctuating_end_summary.RDS")
  
  list(
    summary = pred_all,
    end_summary = summary_end,
    scenario_outputs = out_list
  )
}

# =========================
# 4) Example
# =========================
fluct_out <- run_fluctuating_scenarios(
  field_temp = field_temp,
  fit_4pl = fit_4pl,
  temp_mean = temp_mean,
  n_draws = 1000,
  surv_grid = seq(0.99, 0.01, by = -0.05),
  workers = 4,
  overwrite = FALSE,
  save_draws = FALSE
)

print(fluct_out$end_summary)
```

## **Model incorporating repair**

```{r}
# =========================================================
# 0) Load objects and raw data
# ========================================================

fit_4pl <- readRDS(file = "Rdata/model_lethal_TDT.RDS")

d <- read_excel("data/Data_Shrimp_TDT_CTmax_experiment.xlsx",
                sheet = "LETHAL_TDT")

# Transform data to get counts of individuals that survived/died
d <- d %>%
  mutate(
    survival = 1 - Mortality_after_trial,
    n_total = N_individuals_after_trial,
    n_surv  = round(survival * n_total),
    n_dead  = n_total - n_surv,
    assay_temp = Temperature_assay,
    duration = Duration_exposure_hours,
    logd = log10(duration),
    date_experiment = as.factor(Date),
    tank_ID = as.factor(Tank),
    body_mass = Total_Sample_body_mass_g,
    start_min = as.numeric(format(Starting_time, "%H")) * 60 +
      as.numeric(format(Starting_time, "%M")) +
      as.numeric(format(Starting_time, "%S")) / 60
  )

d$obs <- factor(seq_len(nrow(d)))

temp_mean <- mean(d$assay_temp, na.rm = TRUE)
start_min_mean <- mean(d$start_min, na.rm = TRUE)

d <- d %>%
  mutate(
    temp_c = assay_temp - temp_mean,
    start_min_c = start_min - start_min_mean,
    body_mass_c = body_mass - mean(body_mass, na.rm = TRUE)
  )

# =============================================
# Load field temperatures
# =============================================
files <- list.files("data/field_temp", pattern = "\\.csv$", full.names = TRUE)

field_temp <- files %>%
  lapply(read.csv) %>%
  do.call(rbind, .)

field_temp[field_temp == "no data"] <- NA

names(field_temp) <- c(
  "timestamp_utc",
  "local_time_cet",
  "salinity_1m_sjoboden",
  "temp_1m_sjoboden",
  "air_temp",
  "sea_level_mm",
  "sea_temp",
  "deep_water_salinity_lab",
  "deep_water_temp_lab",
  "salinity_1m_ctd",
  "surface_water_salinity_lab",
  "surface_water_temp",
  "temp_1m_ctd"
)

names(field_temp)[is.na(names(field_temp))] <- "tmp"

field_temp$timestamp_utc <- as.POSIXct(field_temp$timestamp_utc, tz = "UTC")
field_temp$sea_temp <- as.numeric(field_temp$sea_temp)
field_temp$surface_water_temp <- as.numeric(field_temp$surface_water_temp)

field_temp <- field_temp %>%
  dplyr::filter(surface_water_temp < 100)

# =============================================
# Plot theme
# =============================================
base_theme <- theme_classic(base_size = 22) +
  theme(
    axis.title = element_text(size = 24, face = "bold"),
    axis.text  = element_text(size = 20),
    legend.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),
    axis.line = element_line(linewidth = 1),
    axis.ticks = element_line(linewidth = 1),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 2)
  )

# =========================================================
# 1) Extract posterior 4PL parameters once from fit_4pl
# =========================================================
# We only keep the population-level parameters needed to reconstruct:
#   - upper asymptote (up)
#   - lower asymptote (low)
#   - curve steepness (k)
#   - temperature effect on the midpoint (mid)
#
# These posterior draws are the basis of all subsequent uncertainty
# propagation under fluctuating temperatures.
extract_4pl_pars <- function(fit_4pl,
                             file_out = "Rdata/fit_4pl_pars.RDS",
                             overwrite = FALSE) {
  
  if(file.exists(file_out) && !overwrite) {
    return(readRDS(file_out))
  }
  
  dr <- posterior::as_draws_df(fit_4pl) %>%
    as.data.frame()
  
  get_par <- function(nm) {
    if(nm %in% names(dr)) {
      dr[[nm]]
    } else {
      rep(0, nrow(dr))
    }
  }
  
  pars <- data.frame(
    .draw = dr$.draw,
    low = plogis(get_par("b_lraw_Intercept")) * 0.49,
    up = 0.51 + plogis(get_par("b_uraw_Intercept")) * 0.49,
    k = exp(get_par("b_logk_Intercept")),
    mid_int = get_par("b_mid_Intercept"),
    mid_temp = get_par("b_mid_temp_c")
  ) %>%
    dplyr::filter(
      is.finite(low),
      is.finite(up),
      is.finite(k),
      is.finite(mid_int),
      is.finite(mid_temp),
      k > 0,
      up > low,
      low < 0.5,
      up > 0.5
    )
  
  saveRDS(pars, file_out)
  pars
}

pars_4pl <- extract_4pl_pars(
  fit_4pl = fit_4pl,
  file_out = "Rdata/fit_4pl_pars.RDS",
  overwrite = FALSE
)

# =========================================================
# 2) Helper functions linking the fitted 4PL model to a
#    continuous latent injury variable
# =========================================================
# Central idea:
#   - For each posterior draw, the static 4PL model implies a T50 at each T.
#   - We define latent injury dose D such that:
#         D = 1  <=>  the LT50-equivalent cumulative injury dose
#   - Under static conditions at temperature T:
#         D = t / t50(T)
#   - Under fluctuating conditions:
#         D accumulates additively through time, and repair subtracts from D.
#
# We then map D back to survival using the same fitted 4PL curve.
# This gives a continuous mortality prediction without reconstructing a
# separate tolerance landscape from z and CTmax.

# Time to any survival threshold s at temperature T
time_to_surv_threshold <- function(temp,
                                   s,
                                   low,
                                   up,
                                   k,
                                   mid_int,
                                   mid_temp,
                                   temp_mean) {
  
  temp_c <- temp - temp_mean
  mid <- mid_int + mid_temp * temp_c
  log10_t_h <- mid + log((up - s) / (s - low)) / k
  10^log10_t_h
}

# Time to 50% survival at temperature T
time_to_t50 <- function(temp,
                        low,
                        up,
                        k,
                        mid_int,
                        mid_temp,
                        temp_mean) {
  
  time_to_surv_threshold(
    temp = temp,
    s = 0.5,
    low = low,
    up = up,
    k = k,
    mid_int = mid_int,
    mid_temp = mid_temp,
    temp_mean = temp_mean
  )
}

# Map latent injury dose D back to survival using the fitted 4PL shape
# D = 1 corresponds to T50-equivalent injury.
survival_from_dose <- function(dose,
                               low,
                               up,
                               k) {
  
  dose_use <- pmax(dose, 1e-12)
  c0 <- log((up - 0.5) / (0.5 - low)) / k
  
  low + (up - low) / (1 + exp(k * (log10(dose_use) + c0)))
}

# =========================================================
# 3) Diagnostic curve: latent injury dose -> survival
# =========================================================
# This is useful for checking the implied mapping between accumulated
# injury dose and survival, directly from the fitted 4PL model.
get_dose_survival_curve <- function(pars,
                                    dose_seq = 10^seq(-3, 1.2, length.out = 300),
                                    n_draws = 1000,
                                    file_out = "Rdata/dose_survival_curve.RDS",
                                    overwrite = FALSE) {
  
  if(file.exists(file_out) && !overwrite) {
    return(readRDS(file_out))
  }
  
  n_use <- min(n_draws, nrow(pars))
  pars_use <- pars %>%
    dplyr::slice_sample(n = n_use)
  
  out <- do.call(
    rbind,
    lapply(seq_len(nrow(pars_use)), function(i) {
      data.frame(
        .draw = pars_use$.draw[i],
        dose = dose_seq,
        survival = survival_from_dose(
          dose = dose_seq,
          low = pars_use$low[i],
          up = pars_use$up[i],
          k = pars_use$k[i]
        )
      )
    })
  )
  
  out_sum <- out %>%
    dplyr::group_by(dose) %>%
    dplyr::summarise(
      survival_median = median(survival, na.rm = TRUE),
      survival_lower = quantile(survival, 0.025, na.rm = TRUE),
      survival_upper = quantile(survival, 0.975, na.rm = TRUE),
      .groups = "drop"
    )
  
  res <- list(
    draws = out,
    summary = out_sum
  )
  
  saveRDS(res, file_out)
  res
}

dose_surv_curve <- get_dose_survival_curve(
  pars = pars_4pl,
  n_draws = 1000,
  file_out = "Rdata/dose_survival_curve.RDS",
  overwrite = TRUE
)

p_dose_survival <- ggplot(dose_surv_curve$summary,
                          aes(x = dose, y = survival_median)) +
  geom_ribbon(
    aes(ymin = survival_lower, ymax = survival_upper),
    fill = "grey70",
    alpha = 0.3
  ) +
  geom_line(linewidth = 1.5, colour = "black") +
  geom_vline(xintercept = 1, linetype = 2) +
  scale_x_log10() +
  coord_cartesian(ylim = c(0, 1.02)) +
  labs(
    x = "Latent injury dose (LT50-equivalent units)",
    y = "Predicted survival"
  ) +
  base_theme

print(p_dose_survival)

ggsave(
  file = "fig/dose_survival_curve.png",
  plot = p_dose_survival,
  width = 8,
  height = 6,
  dpi = 500
)

# =========================================================
# 3) Derive a damage-rate curve from the fitted 4PL model
# =========================================================
# Here we define DAMAGE in the same diagnostic units used by Arnold et al.:
#
#   damage_rate_pct_min = 100 / time_to_50_in_min
#
# This means:
#   - if T50 = 100 min, damage rate = 1 % LT50-dose per min
#   - if T50 = 10 min,  damage rate = 10 % LT50-dose per min
#   - if T50 = 1 min,   damage rate = 100 % LT50-dose per min
#
# These are diagnostic units for plotting and interpreting net damage.
# The dynamic model later converts this to latent-dose units internally.

get_damage_rate_curve <- function(pars,
                                  temp_mean,
                                  temp_seq = seq(10, 36, by = 0.1),
                                  n_draws = 1000,
                                  file_out = "Rdata/damage_rate_curve_latent_dose.RDS",
                                  overwrite = FALSE) {
  
  if(file.exists(file_out) && !overwrite) {
    return(readRDS(file_out))
  }
  
  n_use <- min(n_draws, nrow(pars))
  pars_use <- pars %>%
    dplyr::slice_sample(n = n_use)
  
  out <- do.call(
    rbind,
    lapply(seq_len(nrow(pars_use)), function(i) {
      
      t50_h <- time_to_t50(
        temp = temp_seq,
        low = pars_use$low[i],
        up = pars_use$up[i],
        k = pars_use$k[i],
        mid_int = pars_use$mid_int[i],
        mid_temp = pars_use$mid_temp[i],
        temp_mean = temp_mean
      )
      
      data.frame(
        .draw = pars_use$.draw[i],
        temp = temp_seq,
        t50_h = t50_h,
        damage_rate_dose_h = 1 / t50_h,
        damage_rate_pct_min = 100 / (60 * t50_h)
      )
    })
  ) %>%
    dplyr::filter(
      is.finite(damage_rate_dose_h),
      damage_rate_dose_h >= 0
    )
  
  out_sum <- out %>%
    dplyr::group_by(temp) %>%
    dplyr::summarise(
      damage_rate_dose_h_median = median(damage_rate_dose_h, na.rm = TRUE),
      damage_rate_dose_h_lower = quantile(damage_rate_dose_h, 0.025, na.rm = TRUE),
      damage_rate_dose_h_upper = quantile(damage_rate_dose_h, 0.975, na.rm = TRUE),
      damage_rate_pct_min_median = median(damage_rate_pct_min, na.rm = TRUE),
      damage_rate_pct_min_lower = quantile(damage_rate_pct_min, 0.025, na.rm = TRUE),
      damage_rate_pct_min_upper = quantile(damage_rate_pct_min, 0.975, na.rm = TRUE),
      .groups = "drop"
    )
  
  res <- list(
    draws = out,
    summary = out_sum
  )
  
  saveRDS(res, file_out)
  res
}

damage_curve <- get_damage_rate_curve(
  pars = pars_4pl,
  temp_mean = temp_mean,
  temp_seq = seq(10, 36, by = 0.1),
  n_draws = 1000,
  file_out = "Rdata/damage_rate_curve_latent_dose.RDS",
  overwrite = TRUE
)

# =========================================================
# 4) Sharpe-Schoolfield repair TPC
# =========================================================
# IMPORTANT:
# This repair function is now defined in the SAME diagnostic units as damage:
#
#   % LT50-dose per min
#
# So its values can be compared directly to damage_rate_pct_min.
#
# The dynamic model later converts these units internally to dose per hour:
#   dose_h = (% per min) / 100 * 60
repair_rate_fun <- function(temp,
                            TA,
                            TAL,
                            TAH,
                            TL,
                            TH,
                            TREF,
                            kdot_ref_pct_min) {
  
  out <- exp(TA * (1 / TREF - 1 / (273.15 + temp))) /
    (1 +
       exp(TAL * (1 / (273.15 + temp) - 1 / TL)) +
       exp(TAH * (1 / TH - 1 / (273.15 + temp)))) *
    kdot_ref_pct_min
  
  out[!is.finite(out)] <- 0
  out[out < 0] <- 0
  out
}

# =========================================================
# 5) Set plausible repair-TPC shape for this system
# =========================================================
# These parameters control the SHAPE of repair vs temperature.
# They do NOT define the absolute magnitude. Magnitude is set below
# via low/moderate/high scenario values.
#
# Here we shift the warm decline of repair lower than before, so that:
#   - repair peaks in the mid/high teens
#   - starts falling strongly by the low 20s
#   - is close to zero by the mid-20s
repair_pars <- list(
  TA = 14065,
  TAL = 50000,
  TAH = 120000,
  TL = 10.5 + 273.15,
  TH = 22.5 + 273.15,
  TREF = 17 + 273.15
)

# =========================================================
# 6) Define none / low / moderate / high repair scenarios
# =========================================================
# Here we follow Arnold et al. directly for the diagnostic repair levels.
# These are expressed as:
#
#   % LT50-dose per min
#
# This is the cleanest way to make the plots comparable to Arnold's.
#
# Note:
# These are scenario values, not empirically fitted repair rates.
kdot_levels_pct_min <- c(
  none = 0,
  low = 0.007,
  moderate = 0.0111,
  high = 0.02
)

kdot_df <- data.frame(
  repair_level = factor(names(kdot_levels_pct_min),
                        levels = c("none", "low", "moderate", "high")),
  kdot_ref_pct_min = as.numeric(kdot_levels_pct_min)
)

# =========================================================
# 7) Plot damage, repair, and net damage over temperature
# =========================================================
# These are the Arnold-style diagnostic plots.
# They are for checking assumptions, not for running the dynamic model.

temp_seq_plot <- seq(10, 36, by = 0.1)

repair_plot_df <- do.call(
  rbind,
  lapply(seq_len(nrow(kdot_df)), function(i) {
    data.frame(
      temp = temp_seq_plot,
      repair_rate_pct_min = repair_rate_fun(
        temp = temp_seq_plot,
        TA = repair_pars$TA,
        TAL = repair_pars$TAL,
        TAH = repair_pars$TAH,
        TL = repair_pars$TL,
        TH = repair_pars$TH,
        TREF = repair_pars$TREF,
        kdot_ref_pct_min = kdot_df$kdot_ref_pct_min[i]
      ),
      repair_level = kdot_df$repair_level[i]
    )
  })
)

damage_plot_df <- damage_curve$summary %>%
  dplyr::transmute(
    temp = temp,
    damage_rate_pct_min_median = damage_rate_pct_min_median,
    damage_rate_pct_min_lower = damage_rate_pct_min_lower,
    damage_rate_pct_min_upper = damage_rate_pct_min_upper
  )

# Damage plot
p_damage <- ggplot(damage_plot_df,
                   aes(x = temp, y = damage_rate_pct_min_median)) +
  geom_ribbon(
    aes(ymin = damage_rate_pct_min_lower, ymax = damage_rate_pct_min_upper),
    fill = "grey70",
    alpha = 0.3
  ) +
  geom_line(linewidth = 1.5, colour = "black") +
  coord_cartesian(
    xlim = c(20, 36),
    ylim = c(0, 50)
  ) +
  labs(
    x = "Temperature (°C)",
    y = expression("Damage rate (% LT50-dose " * min^-1 * ")")
  ) +
  base_theme +
  theme(
    legend.position = "none"
  )

print(p_damage)

ggsave(
  file = "fig/damage_rate_latent_dose.png",
  plot = p_damage,
  width = 8,
  height = 6,
  dpi = 500
)

# Repair plot
p_repair <- ggplot(
  repair_plot_df %>% dplyr::filter(repair_level != "none"),
  aes(x = temp, y = repair_rate_pct_min, colour = repair_level)
) +
  geom_line(linewidth = 1.3) +
  coord_cartesian(
    xlim = c(10, 36),
    ylim = c(0, 0.06)
  ) +
  scale_colour_manual(
    values = c(
      "low" = "#7CAE00",
      "moderate" = "#00BFC4",
      "high" = "#F8766D"
    ),
    labels = c("Low", "Moderate", "High")
  ) +
  labs(
    x = "Temperature (°C)",
    y = expression("Repair rate (% LT50-dose " * min^-1 * ")"),
    colour = NULL
  ) +
  base_theme

print(p_repair)

ggsave(
  file = "fig/repair_rate_latent_dose.png",
  plot = p_repair,
  width = 8,
  height = 6,
  dpi = 500
)

# Net damage plot
net_plot_df <- do.call(
  rbind,
  lapply(levels(kdot_df$repair_level), function(rr) {
    
    repair_tmp <- repair_plot_df %>%
      dplyr::filter(repair_level == rr) %>%
      dplyr::select(temp, repair_rate_pct_min)
    
    out <- damage_plot_df %>%
      dplyr::left_join(repair_tmp, by = "temp") %>%
      dplyr::mutate(
        repair_level = rr,
        net_damage_rate_pct_min = damage_rate_pct_min_median - repair_rate_pct_min
      )
    
    out
  })
)

p_net <- ggplot(
  net_plot_df,
  aes(x = temp, y = net_damage_rate_pct_min, colour = repair_level)
) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_line(linewidth = 1.3) +
  coord_cartesian(
    xlim = c(10, 36),
    ylim = c(-0.06, 0.10)
  ) +
  scale_colour_manual(
    values = c(
      "none" = "#A020F0",
      "low" = "#7CAE00",
      "moderate" = "#00BFC4",
      "high" = "#F8766D"
    ),
    labels = c("None", "Low", "Moderate", "High")
  ) +
  annotate("text", x = 11, y = -0.04,
           label = "Repair\noutweighs\ndamage", hjust = 0) +
  annotate("text", x = 15, y = 0.05,
           label = "Damage\noutweighs\nrepair", hjust = 0) +
  labs(
    x = "Temperature (°C)",
    y = expression("Net damage rate (% LT50-dose " * min^-1 * ")"),
    colour = NULL
  ) +
  base_theme

print(p_net)

ggsave(
  file = "fig/net_damage_rate_latent_dose.png",
  plot = p_net,
  width = 8,
  height = 6,
  dpi = 500
)

# =========================================================
# 8) Core fluctuating-temperature predictor using ONE
#    continuous latent injury variable
# =========================================================
# The dynamic model itself runs on the latent-dose scale.
#
# Therefore, we convert repair from:
#   % LT50-dose per min
# to:
#   latent-dose units per hour
#
# using:
#   repair_rate_dose_h = repair_rate_pct_min / 100 * 60
#
# This makes damage and repair operate on the SAME scale inside the model.
predict_survival_fluctuating_latent_dose <- function(temp_df,
                                                     pars,
                                                     temp_mean,
                                                     n_draws = 1000,
                                                     repair = FALSE,
                                                     TA = NULL,
                                                     TAL = NULL,
                                                     TAH = NULL,
                                                     TL = NULL,
                                                     TH = NULL,
                                                     TREF = NULL,
                                                     kdot_ref_pct_min = 0,
                                                     repair_scales_with_survival = TRUE,
                                                     save_draws = FALSE) {
  
  temp_df <- temp_df %>%
    dplyr::arrange(datetime)
  
  if(nrow(temp_df) < 2) {
    stop("temp_df must contain at least 2 rows.")
  }
  
  n_use <- min(n_draws, nrow(pars))
  pars_use <- pars %>%
    dplyr::slice_sample(n = n_use)
  
  dt_h <- as.numeric(diff(temp_df$datetime), units = "hours")
  temp0 <- temp_df$temp[-nrow(temp_df)]
  temp1 <- temp_df$temp[-1]
  
  pred_list <- vector("list", nrow(pars_use))
  
  for(i in seq_len(nrow(pars_use))) {
    
    low_i <- pars_use$low[i]
    up_i <- pars_use$up[i]
    k_i <- pars_use$k[i]
    mid_int_i <- pars_use$mid_int[i]
    mid_temp_i <- pars_use$mid_temp[i]
    
    # Damage rate on latent-dose scale
    t50_0_h <- time_to_t50(
      temp = temp0,
      low = low_i,
      up = up_i,
      k = k_i,
      mid_int = mid_int_i,
      mid_temp = mid_temp_i,
      temp_mean = temp_mean
    )
    
    t50_1_h <- time_to_t50(
      temp = temp1,
      low = low_i,
      up = up_i,
      k = k_i,
      mid_int = mid_int_i,
      mid_temp = mid_temp_i,
      temp_mean = temp_mean
    )
    
    damage_rate0_dose_h <- 1 / t50_0_h
    damage_rate1_dose_h <- 1 / t50_1_h
    
    # Repair rate on latent-dose scale
    if(repair) {
      repair_rate0_pct_min <- repair_rate_fun(
        temp = temp0,
        TA = TA,
        TAL = TAL,
        TAH = TAH,
        TL = TL,
        TH = TH,
        TREF = TREF,
        kdot_ref_pct_min = kdot_ref_pct_min
      )
      
      repair_rate1_pct_min <- repair_rate_fun(
        temp = temp1,
        TA = TA,
        TAL = TAL,
        TAH = TAH,
        TL = TL,
        TH = TH,
        TREF = TREF,
        kdot_ref_pct_min = kdot_ref_pct_min
      )
      
      repair_rate0_dose_h <- repair_rate0_pct_min / 100 * 60
      repair_rate1_dose_h <- repair_rate1_pct_min / 100 * 60
    } else {
      repair_rate0_dose_h <- rep(0, length(dt_h))
      repair_rate1_dose_h <- rep(0, length(dt_h))
    }
    
    dose <- numeric(length = nrow(temp_df))
    surv <- numeric(length = nrow(temp_df))
    
    dose[1] <- 0
    surv[1] <- up_i
    
    for(tt in seq_along(dt_h)) {
      
      damage_inc <- dt_h[tt] * (damage_rate0_dose_h[tt] + damage_rate1_dose_h[tt]) / 2
      repair_inc_raw <- dt_h[tt] * (repair_rate0_dose_h[tt] + repair_rate1_dose_h[tt]) / 2
      
      if(repair_scales_with_survival) {
        repair_scale <- surv[tt] / up_i
      } else {
        repair_scale <- 1
      }
      
      repair_inc <- repair_inc_raw * repair_scale
      
      dose_new <- max(0, dose[tt] + damage_inc - repair_inc)
      
      surv_potential <- survival_from_dose(
        dose = dose_new,
        low = low_i,
        up = up_i,
        k = k_i
      )
      
      surv_new <- min(surv[tt], surv_potential)
      
      dose[tt + 1] <- dose_new
      surv[tt + 1] <- surv_new
    }
    
    pred_list[[i]] <- data.frame(
      .draw = pars_use$.draw[i],
      datetime = temp_df$datetime,
      temp = temp_df$temp,
      latent_dose = dose,
      pred_survival = surv
    )
  }
  
  pred_draws <- do.call(rbind, pred_list)
  
  pred_summary <- pred_draws %>%
    dplyr::group_by(datetime, temp) %>%
    dplyr::summarise(
      survival_median = median(pred_survival, na.rm = TRUE),
      survival_lower = quantile(pred_survival, 0.025, na.rm = TRUE),
      survival_upper = quantile(pred_survival, 0.975, na.rm = TRUE),
      mortality_median = 1 - median(pred_survival, na.rm = TRUE),
      mortality_lower = 1 - quantile(pred_survival, 0.975, na.rm = TRUE),
      mortality_upper = 1 - quantile(pred_survival, 0.025, na.rm = TRUE),
      latent_dose_median = median(latent_dose, na.rm = TRUE),
      latent_dose_lower = quantile(latent_dose, 0.025, na.rm = TRUE),
      latent_dose_upper = quantile(latent_dose, 0.975, na.rm = TRUE),
      .groups = "drop"
    )
  
  if(save_draws) {
    list(summary = pred_summary, draws = pred_draws)
  } else {
    list(summary = pred_summary)
  }
}

# =========================================================
# 9) Prepare field temperatures and warming scenarios
# =========================================================
temp_df <- field_temp %>%
  dplyr::select(datetime = timestamp_utc,
                temp = surface_water_temp) %>%
  dplyr::filter(!is.na(temp)) %>%
  dplyr::arrange(datetime)

temp_scenarios <- list(
  current = temp_df,
  plus1 = temp_df %>% dplyr::mutate(temp = temp + 1),
  plus2 = temp_df %>% dplyr::mutate(temp = temp + 2),
  plus3 = temp_df %>% dplyr::mutate(temp = temp + 3)
)

# =========================================================
# 10) Run all combinations of scenario x repair level
# =========================================================
run_fluctuating_scenarios_latent_dose <- function(temp_scenarios,
                                                  pars,
                                                  temp_mean,
                                                  repair_pars,
                                                  kdot_levels_pct_min,
                                                  n_draws = 1000,
                                                  workers = 4,
                                                  overwrite = FALSE,
                                                  save_draws = FALSE) {
  
  pars_use <- pars %>%
    dplyr::slice_sample(n = min(n_draws, nrow(pars)))
  
  saveRDS(pars_use, "Rdata/pars_use_latent_dose.RDS")
  
  run_grid <- expand.grid(
    scenario = names(temp_scenarios),
    repair_level = names(kdot_levels_pct_min),
    stringsAsFactors = FALSE
  )
  
  future::plan(future::multisession, workers = workers)
  
  out_list <- future.apply::future_lapply(seq_len(nrow(run_grid)), function(i) {
    
    scn <- run_grid$scenario[i]
    rep_level <- run_grid$repair_level[i]
    kdot_ref_i <- kdot_levels_pct_min[[rep_level]]
    
    file_out <- paste0(
      "Rdata/fluctuating_latentdose_",
      scn,
      "_repair_",
      rep_level,
      "_",
      nrow(pars_use),
      "draws.RDS"
    )
    
    if(file.exists(file_out) && !overwrite) {
      return(readRDS(file_out))
    }
    
    out <- predict_survival_fluctuating_latent_dose(
      temp_df = temp_scenarios[[scn]],
      pars = pars_use,
      temp_mean = temp_mean,
      n_draws = nrow(pars_use),
      repair = rep_level != "none",
      TA = repair_pars$TA,
      TAL = repair_pars$TAL,
      TAH = repair_pars$TAH,
      TL = repair_pars$TL,
      TH = repair_pars$TH,
      TREF = repair_pars$TREF,
      kdot_ref_pct_min = kdot_ref_i,
      repair_scales_with_survival = TRUE,
      save_draws = save_draws
    )
    
    out$summary$scenario <- scn
    out$summary$repair_level <- rep_level
    
    saveRDS(out, file_out)
    out
  })
  
  future::plan(future::sequential)
  
  pred_all <- do.call(rbind, lapply(out_list, function(x) x$summary))
  saveRDS(pred_all, "Rdata/fluctuating_latentdose_all_summaries.RDS")
  
  summary_end <- pred_all %>%
    dplyr::group_by(scenario, repair_level) %>%
    dplyr::slice_tail(n = 1) %>%
    dplyr::ungroup()
  
  saveRDS(summary_end, "Rdata/fluctuating_latentdose_end_summary.RDS")
  
  list(
    summary = pred_all,
    end_summary = summary_end,
    outputs = out_list
  )
}

# =========================================================
# 12) Plot mortality trajectories
# =========================================================
p_mortality <- ggplot(latent_out$summary,
                      aes(x = datetime,
                          y = mortality_median,
                          colour = repair_level,
                          group = interaction(scenario, repair_level))) +
  geom_line(linewidth = 1) +
  facet_wrap(~scenario, ncol = 1) +
  labs(
    x = "Date",
    y = "Predicted mortality",
    colour = "Repair"
  ) +
  base_theme

print(p_mortality)

ggsave(
  file = "fig/fluctuating_mortality_latent_dose.png",
  plot = p_mortality,
  width = 10,
  height = 10,
  dpi = 500
)

# =========================================================
# 13) Plot latent injury dose trajectories
# =========================================================
p_dose <- ggplot(latent_out$summary,
                 aes(x = datetime,
                     y = latent_dose_median,
                     colour = repair_level,
                     group = interaction(scenario, repair_level))) +
  geom_line(linewidth = 1) +
  facet_wrap(~scenario, ncol = 1) +
  labs(
    x = "Date",
    y = "Latent injury dose (LT50-equivalent units)",
    colour = "Repair"
  ) +
  base_theme

print(p_dose)

ggsave(
  file = "fig/fluctuating_latent_dose.png",
  plot = p_dose,
  width = 10,
  height = 10,
  dpi = 500
)

# =========================================================
# 14) Plot survival trajectories
# =========================================================
p_survival <- ggplot(latent_out$summary,
                     aes(x = datetime,
                         y = survival_median,
                         colour = repair_level,
                         group = interaction(scenario, repair_level))) +
  geom_line(linewidth = 1) +
  facet_wrap(~scenario, ncol = 1) +
  labs(
    x = "Date",
    y = "Predicted survival",
    colour = "Repair"
  ) +
  base_theme

print(p_survival)

ggsave(
  file = "fig/fluctuating_survival_latent_dose.png",
  plot = p_survival,
  width = 10,
  height = 10,
  dpi = 500
)
```


# **Figures for publication** 

## **Figure 1** 
```{r, fig.height = 10, fig.width = 10}

```


