---
title: "New case studies (draft): zebrafish and cereal aphids"
subtitle: "A single-condition TDT, and a multi-group comparison fit with the direct CTmax/z parameterisation"
date: 2026-06-22
format:
  html:
    toc: true
    code-fold: show
---

<!-- DRAFT. Data-loading and fitting chunks are `eval: false` until the public
archives are downloaded and their raw columns inspected (see end). Two studies
with public data: zebrafish (Saruhashi 2026, Zenodo 10.5281/zenodo.20075355)
and aphids (Li 2023, Dryad 10.5061/dryad.mcvdnck4j). Snail (Truebano 2018)
deferred — no public deposit. -->

```{r setup}
#| message: false
library(bayesTLS)
library(dplyr)
library(ggplot2)
library(here)
set.seed(123)
models_dir <- here::here("output", "models")
raw_dir    <- here::here("data-raw", "new_case_studies")
```

Two examples show the same package workflow at two levels of ambition. The
**zebrafish** study is the simplest possible case — a single experimental
condition, fit and summarised in three calls. The **aphid** study shows the
payoff of the joint model: rather than fitting a separate curve per group and
comparing point estimates afterwards, we let critical temperature ($CT_{max}$)
and thermal sensitivity ($z$) depend on species **inside one model**, so the
species comparison comes with full posterior uncertainty. In both cases we
cross-check our estimates against the values reported in the original papers —
they need not match exactly, since the joint 4PL fits the whole response surface
rather than reading points off per-temperature regressions.

# 1. Zebrafish larvae (normoxia) — a single-condition TDT

@saruhashi2026 scored survival of zebrafish (*Danio rerio*) larvae across assay
temperatures (38–40 °C) and exposure durations (7.5–240 min). Here we take the
simplest slice — diploid larvae under normoxia — and estimate $z$, $CT_{max}$
and $T_{crit}$, exactly as for a standard lethal-TDT assay.

```{r zf-fit}
#| eval: false
zf <- readr::read_csv(file.path(raw_dir, "saruhashi2026_survival.csv")) |>
  filter(ploidy == "diploid", oxygen == "normoxia") |>
  count(temp, duration, name = "n_total", wt = NULL) |>   # TODO: real counts
  # -> need n_total and n_surv per (temp, duration); reshape from raw alive/dead
  identity()

std_zf <- standardize_data(zf, temp = "temp", duration = "duration",
                           n_total = "n_total", n_surv = "n_surv",
                           duration_unit = "minutes")

# Direct parameterisation: estimate CTmax and z themselves (single condition).
wf_zf <- fit_4pl(std_zf, ctmax = ~ 1, z = ~ 1, t_ref = 60,
                 file = file.path(models_dir, "zf_normoxia"),
                 file_refit = "on_change")

summary(wf_zf)   # brms population-level coefficients + HMC diagnostics
```

```{r zf-tdt}
#| eval: false
extract_tdt(wf_zf, lethal = TRUE)        # z, CTmax(1h), T_crit with 95% CrIs
plot_survival_curves(wf_zf)              # fitted surface over the observed data
```

We then compare our $CT_{max}$ to the value Saruhashi et al. report for this
condition. No field temperature series accompanies this (lab) study, so a
heat-injury projection would require a sourced trace — omitted here.

# 2. Cereal aphids — comparing species with the direct parameterisation

@li2023 measured survival of three cereal-aphid species (*Metopolophium
dirhodum*, *Sitobion avenae*, *Rhopalosiphum padi*) across temperatures and
durations. We compare the **species** at a single age (6 days). Instead of three
separate fits, we model $CT_{max}$ and $z$ as functions of species in one joint
model — the species contrasts then fall straight out of the posterior.

```{r aphid-fit}
#| eval: false
aphids <- readr::read_csv(file.path(raw_dir, "li2023_aphid_survival.csv")) |>
  filter(age == 6, temp >= 30) |>                # 6-day larvae, heat branch
  # -> one row per (species, temp, duration) with n_total and n_surv
  identity()

std_aphid <- standardize_data(aphids, temp = "temp", duration = "duration",
                              n_total = "n_total", n_surv = "n_surv",
                              duration_unit = "minutes")  # keeps `species`

# CTmax and z each depend on species (cell-means coding); up/low/k inherit it.
wf_aphid <- fit_4pl(std_aphid, ctmax = ~ 0 + species, z = ~ 0 + species,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species"),
                    file_refit = "on_change")

summary(wf_aphid)   # per-species CTmaxdev / logz coefficients + diagnostics
```

```{r aphid-species}
#| eval: false
# Per-species z, CTmax and T_crit, straight from the joint fit.
sp <- tls(wf_aphid, by = "species", lethal = TRUE)
sp$summary

# Species contrasts: difference the per-species posterior draws (e.g. R. padi
# minus M. dirhodum for CTmax) and summarise with pMCMC, as in the manuscript.
```

The estimates are then checked two ways: (i) against the $z$ (≈3.2–4.7 °C) and
$CT_{max,1h}$ (≈35–37 °C) Li et al. report, and (ii) against three independent
per-species `fit_4pl()` fits — **only to confirm** the joint model recovers the
same picture; the joint model is what we use for inference.

```{r aphid-confirm}
#| eval: false
# Confirmation only: separate fits should agree with the joint model above.
by_species <- split(std_aphid, std_aphid$species) |>
  lapply(\(d) extract_tdt(fit_4pl(d, ctmax = ~ 1, z = ~ 1, t_ref = 60),
                          lethal = TRUE))
```

Li et al. supply hourly temperature for three China wheat sites; with that trace
we project heat-injury accumulation per species, reproducing their Fig. 2 within
our posterior framework:

```{r aphid-hi}
#| eval: false
# trace = tibble(time, temp) from the Dryad archive (one site/season)
hi <- predict_heat_injury(trace, wf_aphid, target_surv = "relative",
                          trace_unit = "hours")
plot_heat_injury(hi)
```

# Data acquisition / open items

- **Download:** aphids → Dryad `10.5061/dryad.mcvdnck4j`; zebrafish → Zenodo
  `10.5281/zenodo.20075355` (+ code). Inspect each raw schema and reshape to
  `(temp, duration, n_total, n_surv)` cells (plus `species` for aphids); then
  set the chunks above to `eval: true`.
- **Aphid trace:** extract the hourly 3-site series from the Dryad archive and
  archive it under `inst/extdata` (mirroring the existing trace builders).
- **Cold branch (aphids)** and the **ploidy × oxygen** zebrafish comparison are
  natural extensions once the single examples are working.
- **Snail (Truebano 2018):** deferred pending author data request.
