---
title: "New case studies (draft): zebrafish and cereal aphids"
subtitle: "Two multi-group TDT comparisons fit with the direct CTmax/z parameterisation"
date: 2026-06-23
format:
  html:
    embed-resources: true
    toc: true
    code-fold: show
---

<!-- DRAFT. The data are now bundled with the package (`zebrafish_o2`,
`aphid_tdt`) and the verification fits run end-to-end. Fitting chunks are kept
`eval: false` so rendering the tutorial does not refit two brms models; each
`fit_4pl()` call caches to `output/models/` via `file =`, so flip the chunks to
`eval: true` (or knit once) when a rendered, fitted tutorial is wanted.
Provenance: zebrafish = Saruhashi et al. (2026), Zenodo 10.5281/zenodo.20075355,
CC BY 4.0; aphids = Li et al. (2023), Dryad 10.5061/dryad.mcvdnck4j, CC0. -->

```{r setup}
#| message: false
library(bayesTLS)
library(dplyr)
library(ggplot2)
library(here)
set.seed(123)
models_dir <- here::here("output", "models")
```

Both examples fit the same workflow: rather than fitting a separate TDT curve
per group and comparing point estimates afterwards, we let critical temperature
($CT_{max}$) and thermal sensitivity ($z$) depend on a **moderator inside one
joint 4PL**, so the between-group comparison comes with full posterior
uncertainty. The **zebrafish** study compares heat tolerance across an oxygen
gradient; the **aphid** study compares three species. In both we cross-check our
estimates against the original papers — they need not match exactly, since the
joint 4PL fits the whole response surface rather than reading points off
per-temperature regressions.

# 1. Zebrafish larvae across an oxygen gradient

@saruhashi2026 scored survival of zebrafish (*Danio rerio*) larvae across assay
temperatures (26 °C control, 38–40 °C) and exposure durations (3.8–240 min)
under three oxygen treatments — **hypoxia, normoxia, hyperoxia** — to test the
oxygen- and capacity-limited thermal tolerance (OCLTT) prediction that oxygen
availability sets the upper thermal limit. We take the diploid larvae and let
$CT_{max}$ and $z$ depend on the oxygen treatment in one model.

```{r zf-fit}
#| eval: false

data(zebrafish_o2)
zf <-  zebrafish_o2 |>
  filter(ploidy == "diploid")          # focal ploidy; triploids are an extension
str(zf)

std_zf <- standardize_data(zf, temp = "temp", duration = "duration_min",
                           n_total = "n_total", n_surv = "n_surv",
                           duration_unit = "minutes")   # keeps `oxygen`

# Direct parameterisation: CTmax and z each depend on oxygen treatment. A
# single-factor moderator written as treatment coding (~ oxygen) is normalised
# internally to cell-means (~ 0 + oxygen), so either form fits the identical
# model with the same per-group estimates; cell-means just reads most directly
# (one coefficient per treatment).
wf_zf <- fit_4pl(std_zf, ctmax = ~ 0 + oxygen, z = ~ 0 + oxygen, t_ref = 60,
                 file = file.path(models_dir, "zf_oxygen"),
                 file_refit = "on_change", control = list(adapt_delta = 0.98))

summary(wf_zf)   # per-treatment CTmaxdev / logz coefficients + HMC diagnostics
plot(wf_zf)


# Now, lets try a contrast based parameterisation. Same model, different interpretation of coefficients
wf_zf_contrast <- fit_4pl(std_zf, ctmax = ~ 1 + oxygen, z = ~ 1 + oxygen, t_ref = 60,
                 file = file.path(models_dir, "zf_oxygen_contrast"),
                 file_refit = "on_change", control = list(adapt_delta = 0.98))

summary(wf_zf_contrast)
plot(wf_zf_contrast)

# We can extract the posterior from the model. Note that z is back transformed already
z <- extract_tdt(wf_zf_contrast)$z
z_hypox <- z$draws |> filter(oxygen == "hypoxia")
z_hyper <- z$draws |> filter(oxygen == "hyperoxia")
mean(z_hypox$z)
hist(z_hypox$z)
mean(z_hyper$z)

# Now, we can calculate the contrast
diff  <- log(z_hyper$z)- log(z_hypox$z) 
quantile(diff, c(0.025, 0.5, 0.95))
```

```{r zf-tdt}
#| eval: false
# Per-treatment z, CTmax(1h) and T_crit straight from the joint fit. tls() and
# extract_tdt() share one threshold argument, `target_surv`: "relative" (default;
# the per-draw midpoint (low+up)/2), "absolute" (LT50), or a number in (0, 1) for
# an LTx. Lethal endpoints use relative here.
ox <- tls(wf_zf, by = "oxygen", target_surv = "relative", lethal = TRUE)
ox$summary       # tidy long: one row per oxygen treatment x quantity

# Fitted survival surface with the observed counts overlaid.
psc <- predict_survival_curves(wf_zf)
plot_survival_curves(psc, observed = std_zf, log_time = TRUE)

bayes_R2(wf_zf$fit)
```

Our estimates recover the OCLTT gradient @saruhashi2026 report: $CT_{max}$
increases monotonically from hypoxia through normoxia to hyperoxia, i.e. more
oxygen buys more heat tolerance, with the largest uncertainty under hypoxia
(where fish die fast and the high-temperature cells carry little dose-response
information). No field temperature series accompanies this lab study, so a
heat-injury projection would require a sourced trace — omitted here.

# 2. Cereal aphids — comparing species

@li2023 measured survival of three cereal-aphid species (*Metopolophium
dirhodum*, *Sitobion avenae*, *Rhopalosiphum padi*) across temperatures and
durations, at three ages and over both a heat and a cold branch. We compare the
**species** on the heat branch at a single age (6 days). Instead of three
separate fits, we model $CT_{max}$ and $z$ as functions of species in one joint
model — the species contrasts then fall straight out of the posterior.

```{r aphid-fit}
#| eval: false

data(aphid_tdt)
aphids <- aphid_tdt |>
  filter(branch == "heat", age == "6")          # heat branch, 6-day-old aphids

std_aphid <- standardize_data(aphids, temp = "temp", duration = "duration_min",
                              n_total = "n_total", n_surv = "n_surv",
                              duration_unit = "minutes")  # keeps `species`

# CTmax and z each depend on species; up/low/k inherit the same structure.
wf_aphid <- fit_4pl(std_aphid, 
                    ctmax = ~ 0 + species, 
                        z = ~ 0 + species,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species"),
                    file_refit = "on_change")
summary(wf_aphid)
# MOdel is ok, but we see no difference in some parameters so we may wnat to simplfy equation. For example, maybe assume ke is roughly the same for each species     
wf_aphid_nok <- fit_4pl(std_aphid, 
                      ctmax = ~ 0 + species, 
                          z = ~ 0 + species, 
                          k = ~ temp_c,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species_nok"),
                    file_refit = "on_change")
summary(wf_aphid_nok)   # per-species CTmaxdev / logz coefficients + diagnostics

# Now, we just try the data and fits the bare 4pl. Note that it's agnostic to species here so this estimates a marginal 4pl irrespective of species, but this is expected. 
wf_aphid_bare <- fit_4pl(std_aphid,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species_bare"),
                    file_refit = "on_change")
summary(wf_aphid_bare)

wf_aphid_bare_sp <- fit_4pl(std_aphid,
                      low = ~ temp_c + species + species:temp_c,
                       up = ~ temp_c + species + species:temp_c,
                        k = ~ temp_c + species + species:temp_c,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species_bare_sp"),
                    file_refit = "on_change")
summary(wf_aphid_bare_sp)

# This model adds the temp_c x species interaction to mid as well, so z and CTmax
# vary by species (equivalently: fit_4pl(std_aphid, by = "species", ...)). With
# the interaction on ALL four sub-parameters, though, the asymptote/steepness
# temperature-slopes (low/up/k) are only weakly identified by these data —
# survival saturates near 0/1, so there is little information on how those drift
# with temperature within a species. They wander under the default priors and
# mixing collapses (max Rhat ~1.19, min ESS ~17). Fix: regularise THOSE slopes
# toward zero with tighter priors (mid is left untouched, so the per-species
# z/CTmax stay data-driven). This restores healthy mixing (max Rhat ~1.01, min
# ESS ~570, 0 divergences).
pr_type2 <- make_4pl_priors(std_aphid)
is_slope <- function(np) pr_type2$class == "b" & pr_type2$coef == "" & pr_type2$nlpar == np
pr_type2$prior[is_slope("lowraw")] <- "normal(0, 0.25)"   # default normal(0, 0.5)
pr_type2$prior[is_slope("upraw")]  <- "normal(0, 0.25)"   # default normal(0, 0.5)
pr_type2$prior[is_slope("logk")]   <- "normal(0, 0.15)"   # default normal(0, 0.3)

wf_aphid_bare_sp_type2 <- fit_4pl(std_aphid,
                    low = ~ temp_c*species,
                    up = ~ temp_c*species,
                    k = ~ temp_c*species,
                    mid= ~ temp_c*species,
                    t_ref = 60,
                    prior = pr_type2,
                    control = list(adapt_delta = 0.95),
                    file = file.path(models_dir, "aphid_species_bare_sp_type2"),
                    file_refit = "on_change")
summary(wf_aphid_bare_sp_type2)

# Or same model
wf_aphid_bare_sp_type2_by <- fit_4pl(std_aphid,
                    by = "species",
                    t_ref = 60,
                    prior = pr_type2,
                    control = list(adapt_delta = 0.95),
                    file = file.path(models_dir, "aphid_species_bare_sp_type2_by"),
                    file_refit = "on_change")
summary(wf_aphid_bare_sp_type2)

```

```{r aphid-species}
#| eval: false
# Per-species z, CTmax and T_crit from the joint fit.
sp <- tls(wf_aphid, by = "species", target_surv = "relative", lethal = TRUE)
sp$summary          # tidy long: one row per species x quantity (median + 95% CrI)

# Also, absolute, but should be pretty similar. Also, estiate CTmax 4hr which should drop the CTmax 
sp_abs <- tls(wf_aphid, by = "species", target_surv = "absolute", t_ref = 60*4, lethal = TRUE)
sp_abs$summary
# Species contrasts: the full per-draw posteriors live in sp$draws (keyed on
# species and .draw); difference them (e.g. R. padi minus M. dirhodum for CTmax)
# and summarise with pMCMC, as in the manuscript.
```

The estimates are then checked two ways: (i) against the $z$ (≈3.2–4.7 °C) and
$CT_{max,1h}$ (≈35–37 °C) Li et al. report — our joint fit returns
*M. dirhodum* $z\approx4.6$ / $CT_{max}\approx35.3$, *R. padi* $\approx3.5$ /
$37.0$, *S. avenae* $\approx3.5$ / $36.6$, within their reported ranges — and
(ii) against three independent per-species `fit_4pl()` fits, **only to confirm**
the joint model recovers the same picture; the joint model is what we use for
inference.

```{r}
# Fitted survival surface with the observed counts overlaid.
psc_aphid <- predict_survival_curves(wf_aphid)
plot_survival_curves(psc_aphid, observed = std_zf, log_time = TRUE)

bayes_R2(wf_aphid$fit)
```


```{r aphid-confirm}
#| eval: false
# Confirmation only: separate fits should agree with the joint model above.
by_species <- split(std_aphid, std_aphid$species) |>
  lapply(\(d) extract_tdt(fit_4pl(d, ctmax = ~ 1, z = ~ 1, t_ref = 60),
                          lethal = TRUE))
```

# Open items / natural extensions

- **Zebrafish ploidy:** `zebrafish_o2` ships both diploid and triploid larvae;
  adding `ploidy` (or an `oxygen × ploidy` interaction) to the `ctmax`/`z`
  formulas tests the OCLTT delivery-capacity prediction (triploids, with lower
  surface:volume, are predicted to be more oxygen-limited).
- **Aphid cold branch and ages:** `aphid_tdt` carries the cold branch
  (−11 to −3 °C) and ages 2 / 12 alongside the heat / age-6 slice used here;
  the same workflow fits a cold-tolerance ($z$, $CT_{min}$) comparison.
- **Aphid heat-injury projection:** Li et al. project injury under hourly field
  temperatures for three China wheat sites (Wuhan / Xinxiang / Beijing). Those
  series are **not** in the Dryad deposit (only `surv.txt`); sourcing them (the
  authors used the VisualCrossing weather API) would let `predict_heat_injury()`
  reproduce their Fig. 2 within our posterior framework.
- **Snail (Truebano 2018):** deferred pending author data request (no public
  deposit).
