---
title: "New case studies (draft): zebrafish, aphids, leaf PSII, and Drosophila"
subtitle: "Four worked TDT examples — lethal counts, a continuous-proportion sublethal endpoint, and per-group comparisons — each cross-checked by hand against the package functions"
date: 2026-06-23
format:
  html:
    embed-resources: true
    toc: true
    code-fold: show
---

<!-- DRAFT. The data are now bundled with the package (`zebrafish_o2`,
`aphid_tdt`) and the verification fits run end-to-end. Fitting chunks are kept
`eval: false` so rendering the tutorial does not refit two brms models; each
`fit_4pl()` call caches to `output/models/` via `file =`, so flip the chunks to
`eval: true` (or knit once) when a rendered, fitted tutorial is wanted.
Provenance: zebrafish = Saruhashi et al. (2026), Zenodo 10.5281/zenodo.20075355,
CC BY 4.0; aphids = Li et al. (2023), Dryad 10.5061/dryad.mcvdnck4j, CC0. -->

```{r setup}
#| message: false
library(bayesTLS)
library(dplyr)
library(ggplot2)
library(here)
set.seed(123)
models_dir <- here::here("output", "models")
```

The four examples below share one workflow: rather than fitting a separate TDT
curve per group and comparing point estimates afterwards, we fit **one joint
4PL** and read the classical quantities ($z$, $CT_{max}$, and — for lethal
endpoints — $T_{crit}$) straight from its posterior. **Zebrafish** (across an
oxygen gradient) and **aphids** (across three species) put the moderator on
$CT_{max}$/$z$ directly (the *direct* parameterisation); **leaf PSII** is a
continuous-proportion *sublethal* endpoint fit with a Beta likelihood;
**Drosophila** compares the sexes by putting them on the *midpoint*. In every
example we also recompute $z$ and $CT_{max}$ **by hand** from the raw posterior
draws and confirm they reproduce the package functions to machine precision — the
package does no more than the hand calculation, just vectorised over draws and
groups. Where an original paper reports values we cross-check against those too;
they need not match exactly, since the joint 4PL fits the whole response surface
rather than reading points off per-temperature regressions.

# 1. Zebrafish larvae across an oxygen gradient

@saruhashi2026 scored survival of zebrafish (*Danio rerio*) larvae across assay
temperatures (26 °C control, 38–40 °C) and exposure durations (3.8–240 min)
under three oxygen treatments — **hypoxia, normoxia, hyperoxia** — to test the
oxygen- and capacity-limited thermal tolerance (OCLTT) prediction that oxygen
availability sets the upper thermal limit. We take the diploid larvae and let
$CT_{max}$ and $z$ depend on the oxygen treatment in one model.

```{r zf-fit}
#| eval: false

data(zebrafish_o2)
zf <-  zebrafish_o2 |>
  filter(ploidy == "diploid")          # focal ploidy; triploids are an extension
str(zf)

std_zf <- standardize_data(zf, temp = "temp", duration = "duration_min",
                           n_total = "n_total", n_surv = "n_surv",
                           duration_unit = "minutes")   # keeps `oxygen`

# Direct parameterisation: CTmax and z each depend on oxygen treatment. A
# single-factor moderator written as treatment coding (~ oxygen) is normalised
# internally to cell-means (~ 0 + oxygen), so either form fits the identical
# model with the same per-group estimates; cell-means just reads most directly
# (one coefficient per treatment).
wf_zf <- fit_4pl(std_zf, ctmax = ~ 0 + oxygen, z = ~ 0 + oxygen, t_ref = 60,
                 file = file.path(models_dir, "zf_oxygen"),
                 file_refit = "on_change", control = list(adapt_delta = 0.98))

summary(wf_zf)   # per-treatment CTmaxdev / logz coefficients + HMC diagnostics
plot(wf_zf)


# Now, lets try a contrast based parameterisation. Same model, different interpretation of coefficients
wf_zf_contrast <- fit_4pl(std_zf, ctmax = ~ 1 + oxygen, z = ~ 1 + oxygen, t_ref = 60,
                 file = file.path(models_dir, "zf_oxygen_contrast"),
                 file_refit = "on_change", control = list(adapt_delta = 0.98))

summary(wf_zf_contrast)
plot(wf_zf_contrast)

# We can extract the posterior from the model. Note that z is back transformed already
z <- extract_tdt(wf_zf_contrast)$z
z_hypox <- z$draws |> filter(oxygen == "hypoxia")
z_hyper <- z$draws |> filter(oxygen == "hyperoxia")
mean(z_hypox$z)
hist(z_hypox$z)
mean(z_hyper$z)

# Now, we can calculate the contrast
diff  <- log(z_hyper$z)- log(z_hypox$z) 
quantile(diff, c(0.025, 0.5, 0.95))
```

```{r zf-tdt}
#| eval: false
# Per-treatment z, CTmax(1h) and T_crit straight from the joint fit. tls() and
# extract_tdt() share one threshold argument, `target_surv`: "relative" (default;
# the per-draw midpoint (low+up)/2), "absolute" (LT50), or a number in (0, 1) for
# an LTx. Lethal endpoints use relative here.
ox <- tls(wf_zf, by = "oxygen", target_surv = "relative", lethal = TRUE)
ox$summary       # tidy long: one row per oxygen treatment x quantity

# Fitted survival surface with the observed counts overlaid.
psc <- predict_survival_curves(wf_zf)
plot_survival_curves(psc, observed = std_zf, log_time = TRUE)

bayes_R2(wf_zf$fit)
```

```{r zf-handcheck}
#| eval: false
# HAND CHECK (direct parameterisation). With cell-means coding (~ 0 + oxygen) each
# treatment has its own logz and CTmaxdev coefficient, and by construction
#   z = exp(logz)            CTmax(at the fit's t_ref) = T_bar + CTmaxdev.
# That is exactly what extract_tdt()/tls() compute, so the by-hand numbers below
# reproduce `ox$summary` to machine precision.
pd   <- brms::as_draws_df(get_brmsfit(wf_zf))
Tbar <- wf_zf$meta$temp_mean
for (lvl in c("hypoxia", "normoxia", "hyperoxia")) {
  z_hand     <- exp(pd[[paste0("b_logz_oxygen", lvl)]])
  ctmax_hand <- Tbar + pd[[paste0("b_CTmaxdev_oxygen", lvl)]]
  cat(sprintf("%-9s  z = %s   CTmax(1h) = %s\n", lvl,
      paste(round(quantile(z_hand,     c(.025, .5, .975)), 2), collapse = " / "),
      paste(round(quantile(ctmax_hand, c(.025, .5, .975)), 2), collapse = " / ")))
}
```

Our estimates recover the OCLTT gradient @saruhashi2026 report: $CT_{max}$
increases monotonically from hypoxia through normoxia to hyperoxia, i.e. more
oxygen buys more heat tolerance, with the largest uncertainty under hypoxia
(where fish die fast and the high-temperature cells carry little dose-response
information). No field temperature series accompanies this lab study, so a
heat-injury projection would require a sourced trace — omitted here.

# 2. Cereal aphids — comparing species

@li2023 measured survival of three cereal-aphid species (*Metopolophium
dirhodum*, *Sitobion avenae*, *Rhopalosiphum padi*) across temperatures and
durations, at three ages and over both a heat and a cold branch. We compare the
**species** on the heat branch at a single age (6 days). Instead of three
separate fits, we model $CT_{max}$ and $z$ as functions of species in one joint
model — the species contrasts then fall straight out of the posterior.

```{r aphid-fit}
#| eval: false

data(aphid_tdt)
aphids <- aphid_tdt |>
  filter(branch == "heat", age == "6")          # heat branch, 6-day-old aphids

std_aphid <- standardize_data(aphids, temp = "temp", duration = "duration_min",
                              n_total = "n_total", n_surv = "n_surv",
                              duration_unit = "minutes")  # keeps `species`

# CTmax and z each depend on species; up/low/k inherit the same structure.
wf_aphid <- fit_4pl(std_aphid, 
                    ctmax = ~ 0 + species, 
                        z = ~ 0 + species,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species"),
                    file_refit = "on_change")
summary(wf_aphid)
# MOdel is ok, but we see no difference in some parameters so we may wnat to simplfy equation. For example, maybe assume ke is roughly the same for each species     
wf_aphid_nok <- fit_4pl(std_aphid, 
                      ctmax = ~ 0 + species, 
                          z = ~ 0 + species, 
                          k = ~ temp_c,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species_nok"),
                    file_refit = "on_change")
summary(wf_aphid_nok)   # per-species CTmaxdev / logz coefficients + diagnostics

# Now, we just try the data and fits the bare 4pl. Note that it's agnostic to species here so this estimates a marginal 4pl irrespective of species, but this is expected. 
wf_aphid_bare <- fit_4pl(std_aphid,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species_bare"),
                    file_refit = "on_change")
summary(wf_aphid_bare)

wf_aphid_bare_sp <- fit_4pl(std_aphid,
                      low = ~ temp_c + species + species:temp_c,
                       up = ~ temp_c + species + species:temp_c,
                        k = ~ temp_c + species + species:temp_c,
                    t_ref = 60,
                    file = file.path(models_dir, "aphid_species_bare_sp"),
                    file_refit = "on_change")
summary(wf_aphid_bare_sp)

# This model adds the temp_c x species interaction to mid as well, so z and CTmax
# vary by species (equivalently: fit_4pl(std_aphid, by = "species", ...)). With
# the interaction on ALL four sub-parameters, though, the asymptote/steepness
# temperature-slopes (low/up/k) are only weakly identified by these data —
# survival saturates near 0/1, so there is little information on how those drift
# with temperature within a species. They wander under the default priors and
# mixing collapses (max Rhat ~1.19, min ESS ~17). Fix: regularise THOSE slopes
# toward zero with tighter priors (mid is left untouched, so the per-species
# z/CTmax stay data-driven). This restores healthy mixing (max Rhat ~1.01, min
# ESS ~570, 0 divergences).
pr_type2 <- make_4pl_priors(std_aphid)
is_slope <- function(np) pr_type2$class == "b" & pr_type2$coef == "" & pr_type2$nlpar == np
pr_type2$prior[is_slope("lowraw")] <- "normal(0, 0.25)"   # default normal(0, 0.5)
pr_type2$prior[is_slope("upraw")]  <- "normal(0, 0.25)"   # default normal(0, 0.5)
pr_type2$prior[is_slope("logk")]   <- "normal(0, 0.15)"   # default normal(0, 0.3)

wf_aphid_bare_sp_type2 <- fit_4pl(std_aphid,
                    low = ~ temp_c*species,
                    up = ~ temp_c*species,
                    k = ~ temp_c*species,
                    mid= ~ temp_c*species,
                    t_ref = 60,
                    prior = pr_type2,
                    control = list(adapt_delta = 0.95),
                    file = file.path(models_dir, "aphid_species_bare_sp_type2"),
                    file_refit = "on_change")
summary(wf_aphid_bare_sp_type2)

# Or same model
wf_aphid_bare_sp_type2_by <- fit_4pl(std_aphid,
                    by = "species",
                    t_ref = 60,
                    prior = pr_type2,
                    control = list(adapt_delta = 0.95),
                    file = file.path(models_dir, "aphid_species_bare_sp_type2_by"),
                    file_refit = "on_change")
summary(wf_aphid_bare_sp_type2)

```

```{r aphid-species}
#| eval: false
# Per-species z, CTmax and T_crit from the joint fit.
sp <- tls(wf_aphid, by = "species", target_surv = "relative", lethal = TRUE)
sp$summary          # tidy long: one row per species x quantity (median + 95% CrI)

# Also, absolute, but should be pretty similar. Also, estiate CTmax 4hr which should drop the CTmax 
sp_abs <- tls(wf_aphid, by = "species", target_surv = "absolute", t_ref = 60*4, lethal = TRUE)
sp_abs$summary
# Species contrasts: the full per-draw posteriors live in sp$draws (keyed on
# species and .draw); difference them (e.g. R. padi minus M. dirhodum for CTmax)
# and summarise with pMCMC, as in the manuscript.
```

```{r aphid-handcheck}
#| eval: false
# HAND CHECK (direct, cell-means ~ 0 + species): z = exp(logz) and
# CTmax(1h) = T_bar + CTmaxdev per species, straight from the coefficients —
# reproduces sp$summary above exactly.
pd   <- brms::as_draws_df(get_brmsfit(wf_aphid))
Tbar <- wf_aphid$meta$temp_mean
for (s in c("M_dirhodum", "R_padi", "S_avenae")) {
  z_hand     <- exp(pd[[paste0("b_logz_species", s)]])
  ctmax_hand <- Tbar + pd[[paste0("b_CTmaxdev_species", s)]]
  cat(sprintf("%-11s z = %s   CTmax(1h) = %s\n", s,
      paste(round(quantile(z_hand,     c(.025, .5, .975)), 2), collapse = " / "),
      paste(round(quantile(ctmax_hand, c(.025, .5, .975)), 2), collapse = " / ")))
}
```

The estimates are then checked two ways: (i) against the $z$ (≈3.2–4.7 °C) and
$CT_{max,1h}$ (≈35–37 °C) Li et al. report — our joint fit returns
*M. dirhodum* $z\approx4.6$ / $CT_{max}\approx35.3$, *R. padi* $\approx3.5$ /
$37.0$, *S. avenae* $\approx3.5$ / $36.6$, within their reported ranges — and
(ii) against three independent per-species `fit_4pl()` fits, **only to confirm**
the joint model recovers the same picture; the joint model is what we use for
inference.

```{r aphid-survival}
#| eval: false
# Fitted survival surface with the observed counts overlaid.
psc_aphid <- predict_survival_curves(wf_aphid)
plot_survival_curves(psc_aphid, observed = std_aphid, log_time = TRUE)

bayes_R2(wf_aphid$fit)
```


```{r aphid-confirm}
#| eval: false
# Confirmation only: separate fits should agree with the joint model above.
by_species <- split(std_aphid, std_aphid$species) |>
  lapply(\(d) extract_tdt(fit_4pl(d, ctmax = ~ 1, z = ~ 1, t_ref = 60),
                          lethal = TRUE))
```

# 3. Snow-gum leaf PSII — a continuous-proportion sublethal endpoint

Retained photosystem-II function — post-exposure $F_v/F_m$ divided by its
pre-exposure value — is a **continuous proportion in $[0, 1]$ with no count
denominator**, so the 4PL uses a **Beta** likelihood rather than beta-binomial.
Loss of PSII function is **sublethal**, so we report $z$ and $CT_{max}$ but not
$T_{crit}$. Otherwise nothing changes: the same posterior-midpoint extraction
gives $z$ and $CT_{max}$. The data (snow gum, *Eucalyptus pauciflora*) ship with
the package (`?snowgum_psii`).

```{r psii-fit}
#| eval: false

data(snowgum_psii)

# `proportion =` (not n_total/n_surv) flags a continuous response: standardize_data()
# records response_type = "proportion" and clamps to (0.001, 0.999) for the Beta
# density. Random intercepts on mid soak up day / glasshouse-room batch structure.
leaf_std <- standardize_data(snowgum_psii, temp = "Temp", duration = "Time",
                             proportion = "fvfm_prop",
                             random_effects = c("Day", "G_Room"),
                             duration_unit = "minutes")

# Beta family is auto-picked from response_type; named here for clarity. Single
# group (no moderator): one marginal PSII TDT curve.
wf_leaf <- fit_4pl(leaf_std,
                   family = brms::Beta(link = "identity"),
                   random_effects = c("Day", "G_Room"),
                   t_ref = 60,
                   control = list(adapt_delta = 0.99, max_treedepth = 15),
                   file = file.path(models_dir, "fit_leaf_function_4pl"),
                   file_refit = "on_change")
summary(wf_leaf)
extr_wf_leaf <- tls(wf_leaf)
extr_wf_leaf$summary
# Now add random effects cirect to ctmax and z, Cis will change...but we can't really model random effects they onlu have two levels need a rethink
wf_leaf2 <- fit_4pl(leaf_std,
                    ctmax = ~1 + (1|Day) + (1|G_Room), 
                        z = ~1 + (1|Day) + (1|G_Room),
                   family = brms::Beta(link = "identity"),
                   t_ref = 60,
                   control = list(adapt_delta = 0.99, max_treedepth = 15),
                   file = file.path(models_dir, "fit_leaf_function_4pl_ct_z"),
                   file_refit = "on_change")
summary(wf_leaf2)
extr_wf_leaf2 <- tls(wf_leaf2)
extr_wf_leaf2$summary


# Have a look at the funtinal plots

psc_plant <- predict_survival_curves(wf_leaf2)
plot_survival_curves(psc_plant, observed = leaf_std, log_time = TRUE, response_label = "Photosystem II Function")

bayes_R2(wf_leaf2$fit)
bayes_R2(wf_leaf$fit)


```

```{r psii-extract}
#| eval: false
# Package: z and CTmax(1h) at the relative midpoint threshold. Sublethal -> lethal = FALSE.
et_leaf <- extract_tdt(wf_leaf, target_surv = "relative", t_ref = 60, lethal = FALSE)
et_leaf$z$summary       # z (°C)
et_leaf$CTmax$summary   # CTmax at 1 h (°C)

# HAND CHECK. At the relative threshold log10 LT(T) = mid(T) = b_mid_Intercept +
# b_mid_temp_c * temp_c, so
#   z = -1 / b_mid_temp_c        CTmax = T_bar + (log10(60) - b_mid_Intercept) / b_mid_temp_c.
# This IS what extract_tdt() does internally (it reads mid via posterior_linpred),
# so the two agree to machine precision.
pd   <- brms::as_draws_df(get_brmsfit(wf_leaf))
Tbar <- wf_leaf$meta$temp_mean
z_hand     <- -1 / pd$b_mid_temp_c
ctmax_hand <- Tbar + (log10(60) - pd$b_mid_Intercept) / pd$b_mid_temp_c
round(quantile(z_hand,     c(0.025, 0.5, 0.975)), 2)   # == et_leaf$z$summary
round(quantile(ctmax_hand, c(0.025, 0.5, 0.975)), 2)   # == et_leaf$CTmax$summary
```

PSII function falls steeply with temperature: $z \approx 5.1$ °C (95% CrI
4.5–5.8) and a relative $CT_{max}$ at 1 h $\approx 39.9$ °C (35.5–43.7; wide,
because the relative midpoint extrapolates to 1 h for this sublethal plateau).
The by-hand $z$ and $CT_{max}$ reproduce `extract_tdt()` exactly, confirming the
Beta path feeds the same posterior-midpoint machinery as the count families.

# 4. Vinegar fly (*Drosophila suzukii*) — comparing sexes on the midpoint

@orsted_suzukii_2024 measured mortality (plus sublethal knockdown and fertility)
in *D. suzukii* across temperature and exposure duration, for both sexes. The raw
data are **per individual**; we aggregate to dead/total counts per
temperature × duration × sex and let the **midpoint depend on sex** — so $z$ and
$CT_{max}$ differ by sex while the asymptotes and steepness are shared (the lean,
well-identified model). The new `mid =` argument fits this in one call, replacing
a hand-built `brms::bf()`.

```{r dros-fit}
#| eval: false

data(dsuzukii)

# Per-individual records -> dead/total counts per cell.
mort <- dsuzukii |>
  group_by(temp, time, sex) |>
  summarise(n_total = n(), n_dead = sum(dead), .groups = "drop")

std_sex <- standardize_data(mort, temp = "temp", duration = "time",
                            n_total = "n_total", n_dead = "n_dead",
                            duration_unit = "minutes")

# `mid = ~ sex * temp_c`: sex on the midpoint only -> per-sex z and CTmax, pooled
# asymptotes / steepness. fit_4pl records `sex` as a moderator, so the readers
# below auto-group. (Same model the supplement builds by hand with bf().)
wf_dros <- fit_4pl(std_sex, mid = ~ sex * temp_c, t_ref = 60 * 4,
                   control = list(adapt_delta = 0.95),
                   file = file.path(models_dir, "fit_dsuzukii_mort_sex_mid"),
                   file_refit = "on_change")
summary(wf_dros)
```

```{r dros-extract}
#| eval: false
# Package: per-sex z and CTmax at 4 h (Ørsted's reference exposure). Mortality is
# lethal -> T_crit available. tls(wf_dros, by = "sex") gives the same.
et_dros <- extract_tdt(wf_dros, by = "sex", target_surv = "relative",
                       t_ref = 60 * 4, lethal = TRUE)
et_dros$z$summary       # one row per sex
et_dros$CTmax$summary

# HAND CHECK. Treatment coding makes Female the reference. The midpoint slope is
# b_mid_temp_c (F), + b_mid_sexM:temp_c (M); the intercept b_mid_Intercept (F),
# + b_mid_sexM (M). Then z = -1/slope and
# CTmax = T_bar + (log10(4 h) - intercept)/slope — exactly extract_tdt()'s formula.
pd   <- brms::as_draws_df(get_brmsfit(wf_dros))
Tbar <- wf_dros$meta$temp_mean
slope_F <- pd$b_mid_temp_c
slope_M <- pd$b_mid_temp_c + pd[["b_mid_sexM:temp_c"]]
int_F   <- pd$b_mid_Intercept
int_M   <- pd$b_mid_Intercept + pd$b_mid_sexM
hand <- list(z_F     = -1 / slope_F,
             z_M     = -1 / slope_M,
             CTmax_F = Tbar + (log10(60 * 4) - int_F) / slope_F,
             CTmax_M = Tbar + (log10(60 * 4) - int_M) / slope_M)
round(sapply(hand, quantile, c(0.025, 0.5, 0.975)), 2)   # matches et_dros per sex
```

Our joint fit returns $z \approx 3.06$ °C (F) / $3.21$ °C (M) and
$CT_{max, 4h} \approx 35.2$ °C for both sexes — matching @orsted_suzukii_2024
(their Table 1: $z = 3.03$ F / $3.28$ M, $CT_{max, 4h} \approx 35.2$). The
interaction `mid_sexM:temp_c` is small, so the sex difference in $z$ is modest.
The by-hand $z$ and $CT_{max}$ reproduce `extract_tdt()` exactly, and switching to
the absolute LT50 threshold is one argument: `tls(wf_dros, by = "sex",
target_surv = "absolute", t_ref = 60 * 4)`.

# Open items / natural extensions

- **Zebrafish ploidy:** `zebrafish_o2` ships both diploid and triploid larvae;
  adding `ploidy` (or an `oxygen × ploidy` interaction) to the `ctmax`/`z`
  formulas tests the OCLTT delivery-capacity prediction (triploids, with lower
  surface:volume, are predicted to be more oxygen-limited).
- **Aphid cold branch and ages:** `aphid_tdt` carries the cold branch
  (−11 to −3 °C) and ages 2 / 12 alongside the heat / age-6 slice used here;
  the same workflow fits a cold-tolerance ($z$, $CT_{min}$) comparison.
- **Aphid heat-injury projection:** Li et al. project injury under hourly field
  temperatures for three China wheat sites (Wuhan / Xinxiang / Beijing). Those
  series are **not** in the Dryad deposit (only `surv.txt`); sourcing them (the
  authors used the VisualCrossing weather API) would let `predict_heat_injury()`
  reproduce their Fig. 2 within our posterior framework.
- **Drosophila sublethal endpoints:** `dsuzukii` also carries time-to-coma
  (`t_coma`, right-censored knockdown) and fertility (`prod`). The knockdown
  endpoint needs a censored-likelihood fit (flies still awake at the end of an
  exposure are right-censored); the supplement's Case Study 4 works both through.
- **Snail (Truebano 2018):** deferred pending author data request (no public
  deposit).
