---
title: "A hierarchical Bayesian 4PL recovers the classical thermal death time framework and resolves its statistical pathologies"
shorttitle: "Bayesian 4PL for thermal tolerance"
author:
  - name: "Daniel Noble"
    email: daniel.wa.noble@gmail.com
    corresponding: true
    affiliations:
      - id: anu
        name: Division of Ecology and Evolution, Research School of Biology, Australian National University
        city: Canberra
        country: Australia
date: today
abstract: |
  1. *Background.* The thermal death time (TDT) framework is dominant in ectotherm thermal-tolerance research, but its standard two-step pipeline — point-estimate LT50 extraction at each assay temperature followed by ordinary least squares regression of $\log_{10}\text{LT50}$ on temperature — rests on statistical assumptions that are routinely violated and that bias every downstream quantity (z, CTmax, heat-injury predictions).

  2. *Approach.* We develop a single-stage hierarchical Bayesian four-parameter logistic (4PL) survival model fit directly to raw binomial count data, with identifiably-parameterised free asymptotes, observation-level overdispersion, and random effects for every real level of experimental nesting. We prove algebraically that every classical TDT output — slope, *z*-value, CTmax at any reference time, and the heat-injury (HI) integral — is a one-line posterior summary of this fit, with correctly-calibrated credible intervals.

  3. *Demonstration.* We apply the framework to brown shrimp (*Crangon crangon*) thermal-tolerance data spanning three endpoints (lethal mortality, sublethal time-to-death, and dynamic CTmax at three ramping rates), reconcile the endpoints through the @jorgensen_unifying_2021 ramping equation under full posterior propagation, and project mortality risk under measured Baltic field temperatures and warming scenarios using a damage-accumulation model with a Sharpe–Schoolfield repair kernel.

  4. *Implications.* The classical and Bayesian frameworks are equivalent in the limit, but the hierarchical 4PL delivers properly propagated uncertainty, accommodates censoring and overdispersion by construction, and unlocks outputs the two-step pipeline cannot provide (full survival surface, arbitrary mortality percentiles, decomposable between-group contrasts, dynamic-regime predictions). We provide reusable R helpers and recommend the hierarchical 4PL as the default analysis for new TDT data.
keywords:
  - thermal death time
  - tolerance landscape
  - hierarchical Bayesian model
  - 4-parameter logistic
  - heat injury
  - CTmax
  - ectotherm
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    fig-cap-location: bottom
  docx:
    reference-doc: ../bib/templates/template.docx
    toc: false
    number-sections: false
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
editor_options:
  chunk_output_type: console
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
#| label: setup

# Project-wide objects loaded for inline reporting throughout the manuscript.
# Every number, table, and figure is produced from code — no magic numbers.

library(here)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(brms)
library(posterior)
library(flextable)

set.seed(123)

# Shared figure theme: keep all manuscript figures visually consistent.
# Define once here; load via source() into every figure chunk.
# source(here::here("R", "theme_ms.R"))

# Cached fits (produced by R/bayesian_TDT_shrimp.qmd or equivalent scripts).
# Fits are saved as .rds in output/models/ — never re-fit at render time.
# fit_4pl <- readRDS(here::here("output", "models", "fit_lethal_TDT_4pl.rds"))
# fit_sub <- readRDS(here::here("output", "models", "fit_sublethal_TDT.rds"))
# fit_ctmax <- readRDS(here::here("output", "models", "fit_ctmax.rds"))
```

# Introduction

<!--
The introduction should:
- Establish that thermal-tolerance limits are central to forecasting climate-change impacts on ectotherms (cite @sunday_global_2011, @deutsch_latitude_2008, @sunday_redistribution_2012, @pinsky_marine_2019).
- Introduce the TDT / tolerance landscape framework (cite @rezende_landscapes_2014, @jorgensen_unifying_2021).
- State the statistical pathologies of the standard two-step pipeline at a level that is honest but not polemical — point readers to the supplement for the full catalogue.
- Position our contribution against the closest precedents (cite @vajedsamiei_ttl_2024, @rezende_drosophila_mortality_2020, @villeneuve_dynamic_2024, @orsted_finding_2022) — see notes/2026-04-21-tdt-literature-deep-dive.qmd for the framing.
- State the three contributions clearly:
    (i) a single-stage hierarchical Bayesian 4PL with identifiable parameterisation;
    (ii) an algebraic proof that classical TDT outputs are posterior summaries of the joint fit, with credible intervals;
    (iii) a worked, multi-endpoint empirical demonstration that recovers, reconciles, and extends the classical outputs.
-->

## The thermal death time framework

## Statistical liabilities of the two-step pipeline

## Aims and contributions

# The hierarchical Bayesian 4PL

<!--
This is the methodological core of the paper. It should be readable as a
self-contained methods section; the supplement carries the full derivations
and software listings.
-->

## Model specification

<!--
- 4PL on raw binomial counts: n_surv | trials(n_total) ~ Binomial(p)
- p = low + (up - low) / (1 + exp(k*(log10 t - mid)))
- Identifiable parameterisation:
    low = inv_logit(lraw) * 0.49
    up  = 0.51 + inv_logit(uraw) * 0.49
    k   = exp(logk)
- mid ~ temp_c + group covariates + nested random effects
- Observation-level random effect for overdispersion (cite @harrison_olre_2014).
-->

## Priors

## Random-effects structure

## Censoring, boundary observations, and overdispersion

# Equivalence with the classical pipeline

<!--
Algebraic statement: solving p(t; T) = 0.5 for t gives a closed-form
log10 LT50(T) that is linear in T with slope = beta1, and the classical
TDT slope, z, and CTmax fall out as one-line posterior summaries.

Refer to §10 of tdt_problems.qmd for the full derivation; reproduce the
key identities in the manuscript and put the rest in the supplement.

Helpers extract_tdt_params(), LT50_T_curve(), heat_injury(),
time_to_mortality() should be cited and listed in the supplement.
-->

## Closed-form mapping from 4PL parameters to z and CTmax

## Heat-injury integral as a posterior quantity

## What the 4PL delivers that the two-step cannot

# Worked example: brown shrimp thermal tolerance

<!--
Multi-endpoint demonstration on Crangon crangon.
Source pipeline: R/bayesian_TDT_shrimp.qmd.
-->

## Data and experimental design

```{r data-summary}
#| label: tbl-data-summary
#| tbl-cap: "Sample sizes and experimental design across the three thermal-tolerance endpoints (lethal TDT, sublethal time-to-death, dynamic CTmax)."
#| eval: false

# Render a flextable summarising n trials, n individuals, temperature range,
# and exposure range for each of the three endpoints.
```

## Lethal TDT — hierarchical 4PL on counts

```{r fig-survival-curves}
#| label: fig-survival-curves
#| fig-cap: "Posterior survival curves at each lethal-TDT assay temperature, with 95% credible bands and observed survival proportions."
#| eval: false
```

## Sublethal time-to-death

## Dynamic CTmax across ramping rates

## Reconciling the three endpoints via the Jørgensen ramping equation

```{r fig-tdt-ctmax-reconciliation}
#| label: fig-tdt-ctmax-reconciliation
#| fig-cap: "Forward (TDT → dynamic CTmax) and inverse (CTmax → TDT) reconciliation under full posterior propagation."
#| eval: false
```

## Field-temperature heat-injury projections under warming scenarios

```{r fig-hi-warming}
#| label: fig-hi-warming
#| fig-cap: "Posterior cumulative heat-injury and survival under measured Baltic surface-temperature traces and +1, +2, +3 °C warming scenarios, with and without a Sharpe–Schoolfield repair kernel."
#| eval: false
```

# Discussion

## What changes for the field if the hierarchical 4PL becomes the default

## Limits of the present analysis

<!--
Honest discussion of:
- Repair kernel parameters fixed, not fit
- dCTmax reconciliation uses nls() per draw rather than a joint Stan likelihood
- Arrhenius (1/T-Kelvin) sensitivity check belongs in the supplement
- Endpoint heterogeneity is not eliminated, only made visible
-->

## Recommendations for primary studies and meta-analyses

# Conclusions

# Acknowledgements

# Author contributions {.unnumbered}

DN conceived the project, performed the analyses, and wrote the manuscript.

# Data and code availability {.unnumbered}

All data, analysis scripts, and the rendered manuscript are available at <https://github.com/...>. The hierarchical 4PL is implemented in [`brms`](https://paulbuerkner.com/brms/) [@burkner_brms_2017] using [`cmdstanr`](https://mc-stan.org/cmdstanr/) as the Stan backend.

# Conflict of interest {.unnumbered}

The authors declare no competing interests.

# References {.unnumbered}

::: {#refs}
:::
