---
title: "A flexible modelling framework for estimating thermal sensitivity across life"
shorttitle: "Bayesian models of thermal tolerance"
author:
  - name: "Daniel W.A. Noble"
    email: daniel.noble@anu.edu.au
    corresponding: true
    affiliations:
      - ref: anu
  - name: "Pieter A. Arnold"
    affiliations:
      - ref: anu
  - name: "Patrice Pottier"
    email: patrice.pottier@bioenv.gu.se
    affiliations:
      - ref: gu
affiliations:
  - id: anu
    name: Division of Ecology and Evolution, Research School of Biology, Australian National University
    city: Canberra
    country: Australia
  - id: gu
    name: Department of Biological and Environmental Sciences, University of Gothenburg
    city: Gothenburg
    country: Sweden
abstract: |
  1.  The thermal load sensitivity (TLS) framework is the gold standard for estimating and predicting thermal tolerance in nature, but its standard two-step pipeline — point-estimate LT50 extraction at each assay temperature followed by ordinary least squares regression of $\log_{10}\text{LT50}$ on temperature — rests on statistical assumptions that are routinely violated and that bias every downstream quantity (z, CTmax, heat-injury predictions).

  2. We describe a flexible Bayesian, single-stage four-parameter logistic (4PL) model for modelling the impact of temperature and time on survival and fertility. We demonstrate how such an approach can recover all the classical TDT output typical of a two-stage modelling framework — including thermal sensitivity ($z$, slopes), critical thermal limit ($CT_{max_{1hr}}$) and the heat-injury (HI) integral, which allows for one to predict thermal injury in nature under realistic exposure temperatures. Given its fully Bayesian implementation, it allows for error propagation to account for uncertainty at all stages of modelling as well as improved inferences.

  3. We applied our new modelling framework to data from brown shrimp (*Crangon crangon*) and XX to demonstrate the equivalencies and benefits of such an approach. 

  4. We show that the classical and Bayesian frameworks are equivalent, but that the hierarchical 4PL delivers properly propagated uncertainty, can account for complex hierarchical data (i.e., random effects), can accommodate censoring and overdispersion, , and unlocks outputs the two-step pipeline cannot provide (full survival surface, arbitrary mortality percentiles, decomposable between-group contrasts, dynamic-regime predictions). We provide reusable R helpers and recommend the hierarchical 4PL as the default analysis for new TDT data.
keywords:
  - thermal death time
  - tolerance landscape
  - hierarchical Bayesian model
  - 4-parameter logistic
  - heat injury
  - CTmax
  - ectotherm
format:
  html:
    toc: false
    toc-depth: 3
    number-sections: false
    embed-resources: true
    fig-cap-location: bottom
  docx:
    reference-doc: ../bib/templates/template.docx
    toc: false
    number-sections: false
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
editor_options:
  chunk_output_type: console
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
#| label: setup

# Project-wide objects loaded for inline reporting throughout the manuscript.
# Every number, table, and figure is produced from code — no magic numbers.

library(here)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(brms)
library(posterior)
library(flextable)

set.seed(123)

# Shared figure theme: keep all manuscript figures visually consistent.
# Define once here; load via source() into every figure chunk.
# source(here::here("R", "theme_ms.R"))

# Cached fits (produced by R/bayesian_TDT_shrimp.qmd or equivalent scripts).
# Fits are saved as .rds in output/models/ — never re-fit at render time.
# fit_4pl <- readRDS(here::here("output", "models", "fit_lethal_TDT_4pl.rds"))
# fit_sub <- readRDS(here::here("output", "models", "fit_sublethal_TDT.rds"))
# fit_ctmax <- readRDS(here::here("output", "models", "fit_ctmax.rds"))
```

# Introduction

Climate change is driving more frequent and extreme heat events globally, compromising organism fitness [@snook_fertility_2026; @arnold_thermal_2025; @rezende_drosophila_mortality_2020], with cascading consequences for population persistence and biodiversity [@smale_marine_2019; @wiens_localextinctions_2016]. Quantifying differences in thermal tolerance across species is now central to forecasting climate-change impacts on organisms in nature [@sunday_global_2011; @deutsch_latitude_2008; @sunday_redistribution_2012; @pinsky_marine_2019]. Static measures of heat tolerance, such as $CT_{max}$, have long been used to establish differences in vulnerability among taxa and predicting climate change risk [@pottier_amphibians_2025; @sunday_global_2011; @deutsch_latitude_2008; @sunday_redistribution_2012; @pinsky_marine_2019]. While useful, static measures do not capture how the negative effects of extreme heat on fitness depend on both heat intensity and duration, limiting our ability to make realistic predictions under more dynamic thermal environments [@rezende_landscapes_2014; @orsted_finding_2022; @jorgensen_unifying_2021].


The Thermal Load Sensitivity (TLS) framework (or equivalently Thermal Death Time framework) [@rezende_landscapes_2014; @orsted_finding_2022; @arnold_thermal_2025; @jorgensen_nature_2022] resolves these limitations by jointly modelling the effects of heat intensity and duration on fitness (e.g., survival, fertility, reproduction). This matters because physiological damage accumulates in a dose-dependent manner — physiological damage accumulates faster at higher temperatures and over longer exposures [@orsted_finding_2022; @rezende_landscapes_2014; @arnold_thermal_2025]. Organisms experiencing fluctuating thermal environments in nature can therefore accumulate damage to physiological systems that results in sublethal injury that compounds across successive heat exposures if not fully repaired [@arnold_thermal_2025; @buckley_damage_2025; @jorgensen_unifying_2021]. Such processes can lead to delayed responses to extreme heat events.

Thermal load sensitivity modelling, while powerful, traditionally involves a two-step approach [@orsted_finding_2022; @rezende_drosophila_mortality_2020]. First, experiments will typically be done that manipulate a range of temperatures and times (5 temperatures x 5 durations). The proportion of live/dead (or fertile/infertile, if sub-lethal) is then modelled as a function of log transformed time (in minutes or hours) using dose-response curves fit across all temperatures (usually at each temperature treatment independently). A threshold response, e.g., $LT_{50}$ or $LT_{80}$, is then derived from these curves. Second, the log transformed time taken to reach the threshold (e.g., $LT_{50}$) and temperature are regressed against each other using a linear model and thermal sensitivity (i.e., the slope of this regression, *z*) and critical thermal maxima (i.e., the intercept of the regression, $CT_{max_{time}}$) are then derived [@rezende_landscapes_2014; @orsted_finding_2022]. $CT_{max_{time}}$ is usually computed at 1 min or 1 hour duration. Thermal sensitivity and $CT_{max_{1min}}$ are than used to quantify Heat Injury (HI) (e.g., @jorgensen_drosophila_2019; @villeneuve_dynamic_2024) in response to realistic temperatures in nature. Heat injury measures the cumulative percentage or probability until reaching a given threshold (e.g., $LT_{50}$), at which point we would predict 50% of the population to have died. Recent work has emphasized the importance of repair in heat injury dynamics [@arnold_thermal_2025; @buckley_damage_2025] with modelling approaches developed to capture these damage-repair dynamics.

While the two-step approach to modelling thermal sensitivity is intuitive and simple in practice it has numerous practical and statistical limitations. First, the two-step pipeline propagates uncertainty incorrectly. The $LT_{50}$ estimates are estimates with non-negligible sampling variance. Treating them as fixed inputs to the regression discards that variance and miscalibrates the standard errors on the slope ($z$) and intercept ($CT_{max}$) – a known issue in the statistical literature [@carroll_measurement_2006; @burke_ipd_2017]. Second, group comparisons (between species, sexes, or life stages) inherit the miscalibrated uncertainty of the first stage. As such, differences in $z$ between groups cannot be attributed to differences in the upper asymptote, lower asymptote, or curve steepness of the dose-response curve, because all three are collapsed into a single $LT_{50}$ before the comparison is made. Third, the $LT_{50}$-extraction step collapses the within-replicate count, foreclosing hierarchical decomposition of variance across replicates, plates, populations, sexes, and life stages that could affect parameters of dose-response curves [@harrison_olre_2014; @harrison_mixed_2018; @nakagawa_santos_2012]. Fourth, the intrinsically hierarchical structure of (e.g., animals within batches and assay days) is generally ignored which can lead to pseudoreplicaton, resulting in anticonservative standard errors that affect inference. Finally, propagating uncertainty from a two-step fit into downstream predictions — such as cumulative heat injury — is rarely attempted in practice understating the credible range of extreme heat impacts in nature. 

Here, we introduce a well known dose-response model, the four-parameter logistic function [@ritz_drc_2015], and discuss how it can be used to overcome many of the existing challenges of the two-step thermal load sensitivity models. We show how all the classical thermal load sensitivity parameters (*z*, $CT_{max}$) can be derived from this model, while also easily allowing for predictions of heat injury accumulation.  We develop these models within a Bayesian framework which allows for better uncertainty calibration and error propagation with complex hierarchical data structures.

# A non-linear Bayesian hierarchical model for characterizing thermal load sensitivity

Impacts of time on survival or fertility generally follow a dose-response pattern where longer exposure durations exponentially increase mortality. Mortality and fertility are often measured as proportions or counts (e.g., proportion died or number of infertile pairs) [@orsted_finding_2022; @rezende_drosophila_mortality_2020; @orsted_suzukii_2024] with such processes understood to follow a well-known binomial (or Bernoulli) sampling process [@warton_arcsine_2011; @bolker2009]. However, other variables types can be used and modelled with alternative sampling distributions. 

In this section, we describe a flexible four-parameter logistic function that can be used to model binomial data in a dose-dependent manner. Dose-response models are well known models that can be used to fit the impacts of exposure duration on survival probability for a given temperature [@ritz_drc_2015], and indeed are often used in the context of TLS models [@orsted_suzukii_2024]. When fit in a Bayesian framework they provide substantial modelling flexibility with ease in error propagation to downstream analyses and derived variables.

## Model specification

Let $y_{ij}$ denote the number of survivors out of $n_{ij}$ individuals in trial $j$ exposed for time $t_{ij}$ (minutes) at assay temperature $T_{ij}$ (°C). We model the counts directly with a binomial likelihood (but a beta-binomial likelihood function can also be used and controls for overdispersion):

$$
y_{ij} \mid n_{ij}, p_{ij} \;\sim\; \text{Binomial}(n_{ij},\, p_{ij}),
$$ {#eq-likelihood}

with survival probability $p_{ij}$ given by a four-parameter logistic (4PL) function of $\log_{10}$-exposure time,

$$
p_{ij} \;=\; \ell \;+\; \frac{u - \ell}{1 + \exp\!\bigl(k\,(\log_{10} t_{ij} - mid_{ij})\bigr)}.
$$ {#eq-4pl}

where $\ell$ is the lower survival probability asymptote, $u$ is the upper survival probability asymptote, and $k$ is the slope defining the change in survival probability on the $\log_{10} t$ axis. Larger $k$ gives a sharper decline in survival probability with $\log_{10} t$. $mid_{ij}$ is the value of $\log_{10} t$ at which $p$ falls to the midpoint between $\ell$ and $u$. When asymptotes are near 0 and 1, $mid_{ij}$ coincides with $\log_{10}\text{LT50}$ at trial $j$. While other logistic forms can be used (e.g., two or 3 parameter logistics), @eq-4pl is the most flexible model that can account for different asymptotic survival probabilities. 

The key parameters of @eq-4pl (i.e., $\ell$,$u$, $k$, $mid_{ij}$), can be fit as a non-linear function within a likelihood or Bayesian framework, which estimates their correlation and allows for fixed or random effects to be estimated to explain variation in these parameters. Such an approach allows for *a priori* comparisons (e.g., do sexes or species differ in midpoints?), with correct inference (e.g., do species differ significantly from each other?) and allows for variance decomposition to help identify key sources of variance in the data (e.g., estimating variances explained across individuals, species, trails etc.). Such models are most flexibly fit within a Bayesian framework with intuitive interpretation of these parameters and a highly flexible path for error propagation. We show below how such a model can be used to derive all thermal load sensitivity parameters of interest to ecophysiologists. 

# Deriving key thermal load sensitivity parameters 

The TLS framework is elegant in that it derives metrics with clear biological interpretation that can be compared across populations and species (e.g., *z*, $CT_{max_{1hr}}$). Here, we show that we can derived these same metrics directly from our logistic function. 

As indicated above, parameters can be fit to model changes in the mean of any of the four parameters in @eq-4pl. We can easily derive thermal sensitivity (*z*) and $CT_{max_{1hr}}$ by modelling the effect of temperature on midpoints as follows:

$$
\text{mid}_{ij} = \beta_0 + \beta_1\,(T_{ij} - \bar{T}),
$$ {#eq-mid-temp}

where $\text{mid}_{ij}$ is the midpoint ($\log_{10}$-time at 50% survival between $\ell$ and $u$) for trial $j$ at assay temperature $T_{ij}$ (°C); $\beta_0$ is the intercept giving the midpoint at the grand mean temperature $\bar{T}$; $\beta_1$ is the slope describing how the midpoint changes per °C increase in $T_{ij}$; and $\bar{T}$ is the grand mean of assay temperatures used to mean-centre $T_{ij}$, which decorrelates $\beta_0$ from $\beta_1$ and stabilises sampling [@schielzeth_centering_2010]. Stratum-specific (e.g., species, sex, population) intercepts and slopes are introduced as fixed covariates and/or random effects on $\beta_0$ and $\beta_1$ in the full hierarchical model (see @eq-linpred below); the derivations that follow apply unchanged to each stratum's posterior.

Setting $p_{ij} = 0.5$ in @eq-4pl, isolating the exponential, and taking natural logs gives $\log_{10}\text{LT50}(T) = \text{mid}(T) + (1/k)\log[(u - 0.5)/(0.5 - \ell)]$. Substituting @eq-mid-temp and grouping all terms that do not depend on $T$ into an intercept $\alpha$ recovers the classical TDT linear relation:

$$
\log_{10} \text{LT50}(T) \;=\; \underbrace{\Bigl[\beta_0 + \tfrac{1}{k}\log\!\tfrac{u - 0.5}{0.5 - \ell} - \beta_1\bar{T}\Bigr]}_{\alpha} \;+\; \beta_1\,T,
$$ {#eq-lt50}

From @eq-lt50, the classical TLS quantities fall out directly: thermal sensitivity is $z = -1/\beta_1$, and the critical thermal limit at any reference time $t_\text{ref}$ is $CT_{max}(t_\text{ref}) = (\log_{10} t_\text{ref} - \alpha)/\beta_1$ (with $t_\text{ref} = 60$ min for $CT_{max_{1hr}}$). The slope $\beta_1$ — and hence $z$ — is robust to estimated asymptotes because the asymmetry correction $\tfrac{1}{k}\log\tfrac{u - 0.5}{0.5 - \ell}$ does not depend on $T$. Instead, it shifts $\log_{10}\text{LT50}$ vertically without changing the slope. In contrast, threshold summaries are not robust in this way; both $CT_{max}$ at any $t_\text{ref}$ and LT50 at any given $T$ need to be corrected following @eq-lt50 when partial mortality or sparse data do not estimate the asymptotes to be around 0 and 1.

# Predicting heat injury under dynamic 


# Application of Bayesian dose-response models to the TLS framework
We used a four-parameter logistic function to model the impact of temperature and duration (hours/minutes) on survival probability thermal sensitivity, estimate $CT_{max1h}$

To break the well-known likelihood ridges among $(\ell, u, k)$ in unconstrained 4PL fits [@bolker_ecologicalmodels_2008; @gelman_priors_2006], we estimate the identifiable parameterisation

$$
\ell = 0.49\,\text{logit}^{-1}(\tilde{\ell}), \qquad
u   = 0.51 + 0.49\,\text{logit}^{-1}(\tilde{u}), \qquad
k   = \exp(\kappa),
$$ {#eq-reparam}

with $\tilde{\ell},\tilde{u},\kappa \in \mathbb{R}$. This constrains $\ell \in [0, 0.49]$ and $u \in [0.51, 1]$ into disjoint intervals so that $u > \ell$ by construction, the curve direction is unambiguous, and $k$ is positive on the natural scale.

The inflection point $m_{ij}$ is modelled as a linear function of (mean-centred) assay temperature, with additional fixed covariates and nested random effects:

$$
m_{ij} \;=\; \beta_0 \;+\; \beta_1\,(T_{ij} - \bar T) \;+\; \mathbf{x}_{ij}^{\top}\boldsymbol{\gamma} \;+\; b_{j} \;+\; \varepsilon_{ij},
$$ {#eq-linpred}

where $\bar T$ is the grand mean of assay temperatures (centring stabilises sampling and decorrelates $\beta_0$ from $\beta_1$), $\mathbf{x}_{ij}$ holds any additional fixed covariates with coefficients $\boldsymbol{\gamma}$, $b_{j}$ collects the nested random effects and $\varepsilon_{ij} \sim \mathcal{N}(0, \sigma_\varepsilon^2)$ is an observation-level random effect that absorbs the binomial overdispersion characteristic of replicated count assays [@harrison_olre_2014]. However, we could also remove the observation-level random effect from the model and instead fit a beta-binomial error distribution which estimates a dispersion parameter ($\phi$) more naturally.

Together, @eq-likelihood–@eq-linpred define the hierarchical Bayesian 4PL fit directly to the raw binomial counts. Every classical TDT quantity — slope, $z$, CTmax at any reference time, and the heat-injury integral — is recovered as a one-line posterior summary of this single joint fit, as derived in §[Equivalence with the classical pipeline].

# Case Study 1: Thermal sensitivity estimation across life stages in shrimp




## Priors

## Random-effects structure

## Censoring, boundary observations, and overdispersion

# Equivalence with the classical pipeline

<!--
Algebraic statement: solving p(t; T) = 0.5 for t gives a closed-form
log10 LT50(T) that is linear in T with slope = beta1, and the classical
TDT slope, z, and CTmax fall out as one-line posterior summaries.

Refer to §10 of tdt_problems.qmd for the full derivation; reproduce the
key identities in the manuscript and put the rest in the supplement.

Helpers extract_tdt_params(), LT50_T_curve(), heat_injury(),
time_to_mortality() should be cited and listed in the supplement.
-->

## Closed-form mapping from 4PL parameters to z and CTmax

## Heat-injury integral as a posterior quantity

## What the 4PL delivers that the two-step cannot

# Worked example: brown shrimp thermal tolerance

<!--
Multi-endpoint demonstration on Crangon crangon.
Source pipeline: R/bayesian_TDT_shrimp.qmd.
-->

## Data and experimental design

```{r data-summary}
#| label: tbl-data-summary
#| tbl-cap: "Sample sizes and experimental design across the three thermal-tolerance endpoints (lethal TDT, sublethal time-to-death, dynamic CTmax)."
#| eval: false

# Render a flextable summarising n trials, n individuals, temperature range,
# and exposure range for each of the three endpoints.
```

## Lethal TDT — hierarchical 4PL on counts

```{r fig-survival-curves}
#| label: fig-survival-curves
#| fig-cap: "Posterior survival curves at each lethal-TDT assay temperature, with 95% credible bands and observed survival proportions."
#| eval: false
```

## Sublethal time-to-death

## Dynamic CTmax across ramping rates

## Reconciling the three endpoints via the Jørgensen ramping equation

```{r fig-tdt-ctmax-reconciliation}
#| label: fig-tdt-ctmax-reconciliation
#| fig-cap: "Forward (TDT → dynamic CTmax) and inverse (CTmax → TDT) reconciliation under full posterior propagation."
#| eval: false
```

## Field-temperature heat-injury projections under warming scenarios

```{r fig-hi-warming}
#| label: fig-hi-warming
#| fig-cap: "Posterior cumulative heat-injury and survival under measured Baltic surface-temperature traces and +1, +2, +3 °C warming scenarios, with and without a Sharpe–Schoolfield repair kernel."
#| eval: false
```

# Discussion

## What changes for the field if the hierarchical 4PL becomes the default

## Limits of the present analysis

<!--
Honest discussion of:
- Repair kernel parameters fixed, not fit
- dCTmax reconciliation uses nls() per draw rather than a joint Stan likelihood
- Arrhenius (1/T-Kelvin) sensitivity check belongs in the supplement
- Endpoint heterogeneity is not eliminated, only made visible
-->

## Recommendations for primary studies and meta-analyses

# Conclusions

# Acknowledgements

# Author contributions {.unnumbered}

DN conceived the project, performed the analyses, and wrote the manuscript.

# Data and code availability {.unnumbered}

All data, analysis scripts, and the rendered manuscript are available at <https://github.com/...>. The hierarchical 4PL is implemented in [`brms`](https://paulbuerkner.com/brms/) [@burkner_brms_2017] using [`cmdstanr`](https://mc-stan.org/cmdstanr/) as the Stan backend.

# Conflict of interest {.unnumbered}

The authors declare no competing interests.

# References {.unnumbered}

::: {#refs}
:::
