---
title: "A flexible modelling framework for estimating thermal tolerance and sensitivity"
shorttitle: "Bayesian models of thermal tolerance"
authors:
  - name: "Daniel W.A. Noble"
    email: daniel.noble@anu.edu.au
    corresponding: true
    affiliations:
      - ref: anu
  - name: "Pieter A. Arnold"
    affiliations:
      - ref: anu
  - name: "Patrice Pottier"
    email: patrice.pottier@bioenv.gu.se
    affiliations:
      - ref: gu
      - ref: unsw
affiliations:
  - id: anu
    name: Division of Ecology and Evolution, Research School of Biology, Australian National University
    city: Canberra
    country: Australia
  - id: gu
    name: Department of Biological and Environmental Sciences, University of Gothenburg
    city: Gothenburg
    country: Sweden
  - id: unsw
    name: Evolution & Ecology Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales
    city: Sydney
    state: New South Wales
    country: Australia
abstract: |
  The thermal load sensitivity (TLS) and thermal death time (TDT) frameworks are gold standard approaches for estimating thermal tolerance and predict organismal vulnerability under natural thermal regimes. However, these frameworks are usually implemented through a two-step pipeline. Median times to lethal or sublethal endpoints ($t_{50}$) are first estimated independently across a range of assay temperatures, and these $\log_{10}(t_{50})$ estimates are then regressed against temperature. This two-step approach violates key statistical assumptions that are likely to bias the mean and variance in derived quantities such as thermal sensitivity ($z$), critical thermal limits ($CT_{max}$), heat-injury predictions, and survival probabilities.

  We present a flexible Bayesian four-parameter logistic (4PL) model that estimates survival or sublethal responses as a joint function of temperature and exposure duration. We demonstrate how such an approach recovers all the classical outputs of the TLS and TDT two-stage modelling frameworks (i.e., $z$, $CT_{max}$, heat-injury integral, survival probabilities), while propagating uncertainty through all derived quantities.  Using simulated data and empirical case studies in brown shrimp (*Crangon crangon*), zebrafish (*Danio rerio*), and XXX (*XXX*), we demonstrate the benefits of this new modelling framework. 

  We show that the classical and Bayesian frameworks are equivalent, but that the Bayesian approach delivers properly propagated uncertainty, can account for complex hierarchical data (i.e., random effects), can accommodate censoring and overdispersion, and unlocks outputs the two-step pipeline cannot provide, including survival landscape surfaces, between-group contrasts, and inference about the variation in the shape of thermal dose-response curves among groups. We provide a new *R* package ($bayesTLS$) and recommend the bayesian four-parameter logistic model as the default approach for analysing TDT and TLS data, particularly when the goal is to make accurate predictions of survival under natural thermal regimes.
keywords:
  - thermal death time
  - tolerance landscape
  - hierarchical Bayesian model
  - 4-parameter logistic
  - heat injury
  - CT_max
  - ectotherm
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    number-sections: false
    embed-resources: true
    fig-cap-location: bottom
  docx:
    reference-doc: ../bib/templates/template.docx
    toc: false
    number-sections: false
  pdf:
    documentclass: article
    geometry:
      - margin=1in
    fontsize: 11pt
    toc: false
    number-sections: false
    fig-pos: "H"
    colorlinks: true
    link-citations: true
    keep-tex: true
    include-in-header:
      text: |
        % tinytable LaTeX dependencies. See supplement.qmd for rationale.
        \usepackage{tabularray}
        \usepackage{float}
        \usepackage{graphicx}
        \usepackage[normalem]{ulem}
        \providecommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
        \providecommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
        % Author block: Quarto's default \author{names \and …} gives
        % names only. We override with authblk to pin affiliations from
        % YAML to the title page.
        \usepackage[affil-it]{authblk}
        \renewcommand\Authsep{, }
        \renewcommand\Authand{, }
        \renewcommand\Authands{, }
        \renewcommand\Affilfont{\itshape\small}
        \let\OldAuthor\author
        \renewcommand{\author}[1]{}
        \AtBeginDocument{%
          \OldAuthor[1]{Daniel W.A. Noble\thanks{Corresponding author: \texttt{daniel.noble@anu.edu.au}}}%
          \OldAuthor[1]{Pieter A. Arnold}%
          \OldAuthor[2,3]{Patrice Pottier}%
          \affil[1]{Division of Ecology and Evolution, Research School of Biology, Australian National University, Canberra, Australia}%
          \affil[2]{Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden}%
          \affil[3]{Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, New South Wales, Australia}%
        }
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
editor_options:
  chunk_output_type: console
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

::: {.content-visible when-format="docx"}
**Daniel W.A. Noble**^1,\*^, **Pieter A. Arnold**^1^, **Patrice Pottier**^2,3^

^1^Division of Ecology and Evolution, Research School of Biology, Australian National University, Canberra, Australia.
^2^Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden.
^3^Evolution & Ecology Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, New South Wales, Australia.
^\*^Corresponding author: daniel.noble@anu.edu.au

**Abstract**

The thermal load sensitivity (TLS) and thermal death time (TDT) frameworks are gold standard approaches for estimating thermal tolerance and predict organismal vulnerability under natural thermal regimes. However, these frameworks are usually implemented through a two-step pipeline. Median times to lethal or sublethal endpoints ($t_{50}$) are first estimated independently across a range of assay temperatures, and these $\log_{10}(t_{50})$ estimates are then regressed against temperature. This two-step approach violates key statistical assumptions that are likely to bias the mean and variance in derived quantities such as thermal sensitivity ($z$), critical thermal limits ($CT_{max}$), heat-injury predictions, and survival probabilities.

We present a flexible Bayesian four-parameter logistic (4PL) model that estimates survival or sublethal responses as a joint function of temperature and exposure duration. We demonstrate how such an approach recovers all the classical outputs of the TLS and TDT two-stage modelling frameworks (i.e., $z$, $CT_{max}$, heat-injury integral, survival probabilities), while propagating uncertainty through all derived quantities.  Using simulated data and empirical case studies in brown shrimp (*Crangon crangon*), zebrafish (*Danio rerio*), and XXX (*XXX*), we demonstrate the benefits of this new modelling framework. 

We show that the classical and Bayesian frameworks are equivalent, but that the Bayesian approach delivers properly propagated uncertainty, can account for complex hierarchical data (i.e., random effects), can accommodate censoring and overdispersion, and unlocks outputs the two-step pipeline cannot provide, including survival landscape surfaces, between-group contrasts, and inference about the variation in the shape of thermal dose-response curves among groups. We provide a new *R* package ($bayesTLS$) and recommend the bayesian four-parameter logistic model as the default approach for analysing TDT and TLS data, particularly when the goal is to make accurate predictions of survival under natural thermal regimes.

**Keywords:** thermal-death-time; thermal tolerance landscape; multilevel model; 4-parameter logistic; heat injury; heat tolerance; critical thermal maximum; ectotherm
:::

```{r setup}
#| label: setup
#| cache: false

# Project-wide objects loaded for inline reporting throughout the manuscript.
# Every number, table, and figure is produced from code — no magic numbers.

library(here)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(brms)
library(posterior)
library(tinytable)

# Force tinytable to register its LaTeX preamble dependencies on every
# render. See supplement.qmd setup-supp for explanation.
invisible(tinytable::tt(data.frame(x = 1)))

set.seed(123)

# Shared figure theme: keep all manuscript figures visually consistent.
# Define once here; load via source() into every figure chunk.
# source(here::here("R", "theme_ms.R"))

# Cached fits (produced by R/bayesian_TDT_shrimp.qmd or equivalent scripts).
# Fits are saved as .rds in output/models/ — never re-fit at render time.
# fit_4pl <- readRDS(here::here("output", "models", "fit_lethal_TDT_4pl.rds"))
# fit_sub <- readRDS(here::here("output", "models", "fit_sublethal_TDT.rds"))
# fit_ctmax <- readRDS(here::here("output", "models", "fit_ctmax.rds"))
```

# Introduction

Climate change is increasing the frequency, intensity, and duration of extreme heat events globally, with cascading consequences for organismal fitness, population persistence, and biodiversity[@snook_fertility_2026; @arnold_thermal_2025; @rezende_drosophila_mortality_2020; @smale_marine_2019; @wiens_localextinctions_2016]. Quantifying variation in thermal tolerance within and among species is therefore central to forecasting biological responses to climate change in nature [@sunday_global_2011; @deutsch_latitude_2008; @sunday_redistribution_2012; @pinsky_marine_2019; @jorgensen_nature_2022]. Threshold-based measures of heat tolerance, such as the critical thermal maximum ($CT_{max}$), have long been used to establish differences in climate vulnerability among taxa and predicting vulnerability to climate change [@pottier_amphibians_2025; @sunday_global_2011; @deutsch_latitude_2008; @sunday_redistribution_2012; @pinsky_marine_2019]. While useful, threshold-based metrics do not capture how the biological effects of extreme heat on fitness depend on both the duration and intensity of heat stress, limiting our ability to make realistic predictions of the impacts of climate change in nature [@rezende_landscapes_2014; @orsted_finding_2022; @jorgensen_unifying_2021].

The Thermal Death Time (TDT) framework address this limitation by quantifying how the intensity and duration of heat exposure jointly determine survival responses to heat stress [@rezende_landscapes_2014; @orsted_finding_2022; @jorgensen_nature_2022; @jorgensen_drosophila_2019]. Recently, this framework was expanded to include sublethal (e.g., fertility, heat coma) responses to heat stress, providing a flexible framework to broadly explain variation in biological responses to heat stress [@rezende_landscapes_2014; @arnold_thermal_2025; @noble2026]. These frameworks reflect the idea that heat injury accumulates in a dose-dependent manner above a certain threshold ($T_{crit}$)), and injury accumulates more rapidly at higher temperatures and over longer exposures. As a result, organisms may eventually reach a thermal load that leads to the incapacitation of their physiological functions [@orsted_finding_2022; @rezende_landscapes_2014; @arnold_thermal_2025; @jorgensen_drosophila_2019] if repair mechanisms cannot outpace the injury accumulation [@arnold_thermal_2025; @buckley_damage_2025; @jorgensen_unifying_2021]. By linking biological responses to the intensity and duration of heat stress, the TLS and TDT frameworks therefore allow for more realistic predictions of the impacts of  fluctuating thermal regimes  [@rezende_landscapes_2014; @orsted_finding_2022; @jorgensen_unifying_2021], and have shown promise in predicting observed mortality rates in nature [@rezende_drosophila_mortality_2020; @zamora2026predicting].

Despite their conceptual strengths, TLS analyses traditionally involves a two-step modelling approach [@orsted_finding_2022; @rezende_drosophila_mortality_2020; @truebano2018; @jorgensen_drosophila_2019; @li2023]. First, organisms are exposed to a set of assay temperatures for different durations, and the proportion of individuals that are dead, infertile, knocked down, or otherwise affected is modelled as a function of exposure duration. This is typically done by fitting independent dose-response curves at each assay temperature. From these curves, the exposure duration required to reach a specified response threshold is estimated, most often $t_{50}$, the time required to reach 50% mortality, infertility, knockdown, or  other endpoints. Second, the resulting $\log_{10}(t_{50})$ estimates are regressed against assay temperature. Thermal sensitivity (i.e., the inverse of the slope of this regression, $z$) and static critical thermal maximum (i.e., the intercept of the regression, $sCT_{max_{time}}$) are then derived from this linear model [@rezende_landscapes_2014; @orsted_finding_2022]. Thermal sensitivity ($z$) is interpreted as the temperature increase required to reduce $t_{50}$ by an order of magnitude, while $sCT_{max_{time}}$ is usually arbitrarily computed for a 1 min [@rezende_landscapes_2014], 1 hour [@jorgensen_drosophila_2019], or 4 hour [@orsted_suzukii_2024] duration of heat stress exposure. These quantities can then be used to estimate cumulative Heat Injury (HI) under realistic temperature-time series, where injury accumulates as a function of the exposure duration relative to the predicted tolerance time at each temperature [@jorgensen_drosophila_2019; @villeneuve_dynamic_2024]. Recent extensions have further emphasized that injury accumulation may also be partly offset by repair or acclimation mechanisms, motivating the emergence of new models that incorporate damage-repair dynamics [@arnold_thermal_2025; @buckley_damage_2025].

The two-step pipeline is intuitive and easy to implement, yet it has several practical and statistical limitations. First, the two-step pipeline does not typically allow proper uncertainty propagation. The $t_{50}$ estimates derived in the first step have non-negligible sampling variance, but they are commonly treated as fixed point estimates in the second-step regression. Neglecting sampling variance can bias or miscalibrate inference on ($z$) and ($CT_{max}$), and lead to inflated $R^2$ values - a known issue in the statistical literature [@carroll_measurement_2006; @burke_ipd_2017]. Second, estimating $t_{50}$ from dose-response models requires empirical data spanning the full mortality range (i.e., 0 to 100% mortality) to anchor the upper and lower asymptotes of the dose-response curve. When data are sparse or do not span the full mortality range, $t_{50}$ estimates can be incorrectly or impossible to estimate, which is then propagated to bias and miscalibrate downstream estimates of $z$ and $CT_{max}$. Third, group comparisons (between species, sexes, or life stages) inherit this miscalibrated uncertainty. Reducing each dose-response curve to a fixed $t_{50}$ estimate makes it difficult to determine if differences in thermal tolerance and sensitivity between groups can be attributed to differences in the upper asymptote, lower asymptote, or curve steepness of the dose-response curve. Fourth, fitting separate dose-response models for each assay temperature and subsequently regressing $t_{50}$ estimates against temperature assumes that these estimates are independent, and typically ignores the hierarchical structure of experimental data (e.g., individuals nested within vials, plates, tanks, clutches, batches, populations, or experimental days). Ignoring this nesting structure prevents the decomposition of variance among nesting groups, and can lead to pseudoreplication and anticonservative uncertainty estimates, particularly when replicate-level variation affects the shape or position of the dose-response curve [@harrison_olre_2014; @harrison_mixed_2018; @nakagawa_santos_2012]. Finally, the two-step pipeline does not allow downstream predictions of heat injury and survival under fluctuating thermal regimes with uncertainty, which is key to inferring the confidence in our predictions of the impacts of extreme heat impacts. 

Here, we present a single-stage alternative to the classical two-step pipeline: a hierarchical Bayesian four-parameter logistic model fitted jointly to all temperature-by-duration response data. In this framework, survival or sublethal response probabilities are estimated directly as a function of exposure duration and assay temperature, while allowing the parameters of the dose-response curve to vary among groups and hierarchical levels. The four-parameter logistic model provides a flexible sigmoidal description of thermal dose responses, with interpretable parameters for the lower asymptote, upper asymptote, curve steepness, and midpoint [@ritz_drc_2015; @ritz_ecotox_2010; @nielsen_nlmm_2004; @rudemo_random_1989; @ritz_streibig_2008]. We focus on binomial and beta-binomial outcomes because count data dominate TLS studies, but the same modelling logic can be extended to other response types using appropriate likelihoods, including continuous performance traits, or latency data. Similarly, although we use the four-parameter log-logistic form for its flexibility, simpler nested models (3PL, with one asymptote fixed; 2PL, with both fixed at 0 and 1) or alternative monotonic dose-response functions (e.g., Weibull, log-normal) can be implemented within the same general framework [@ritz_drc_2015; @ritz_streibig_2008].

 We show how classical TLS quantities, such as thermal sensitivity ($z$), thermal tolerance ($CT_{max}$), heat injury accumulation, survival under fluctuating thermal regimes, and their uncertainty, can be derived from this single Bayesian four-parameter logistic model. We implement these tools in the R package $bayesTLS$ and recommend this new modelling approach as the default approach for analysing organisma; responses to varyind thermal loads. Using simulations and empirical case studies in brown shrimp (*Crangon crangon*), zebrafish (*Danio rerio*), and XXX (XXX), we demonstrate how to implement, validate, and expand this new modelling approach. 
 

# A non-linear Bayesian hierarchical model for characterizing thermal load sensitivity

## Model specification

Assuming we have mortality data for different times and temperatures, let $y_{ij}$ denote the number of survivors out of $n_{ij}$ individuals in trial $j$ exposed for time $t_{ij}$ (e.g., minutes) at assay temperature $T_{ij}$ (°C). We can model the counts directly with a binomial likelihood (but a beta-binomial likelihood function can also be used and controls for overdispersion):

$$
y_{ij} \mid n_{ij}, p_{ij} \;\sim\; \text{Binomial}(n_{ij},\, p_{ij}),
$$ {#eq-likelihood}

with survival probability $p_{ij}$ given by a four-parameter logistic (4PL) function of $\log_{10}$-exposure time,

$$
p_{ij} \;=\; \text{low} \;+\; \frac{\text{up} - \text{low}}{1 + \exp\!\bigl(k\,(\log_{10} t_{ij} - mid_{ij})\bigr)}.
$$ {#eq-4pl}

where $\text{low}$ is the lower survival probability asymptote, $\text{up}$ is the upper survival probability asymptote, and $k$ is the slope defining the change in survival probability on the $\log_{10} t$ axis. Larger $k$ gives a sharper decline in survival probability with $\log_{10} t$. $mid_{ij}$ is the value of $\log_{10} t$ at which $p$ falls to the midpoint between $\text{low}$ and $\text{up}$. When asymptotes are near 0 and 1, $mid_{ij}$ coincides with $\log_{10}t_{50}$ at trial $j$. While other logistic forms can be used (e.g., two or 3 parameter logistics), @eq-4pl is the most flexible model that can account for different asymptotic survival probabilities. 

The key parameters of @eq-4pl (i.e., $\text{low}$, $\text{up}$, $k$, $mid_{ij}$), can be fit as a non-linear function within a likelihood or Bayesian framework, which estimates their correlation and allows for fixed or random effects to be estimated to explain variation in these parameters. Such an approach allows for *a priori* comparisons (e.g., do sexes or species differ in midpoints?), with correct inference (e.g., do species differ significantly from each other?) and allows for variance decomposition to help identify key sources of variance in the data (e.g., estimating variances explained across individuals, species, trials etc.). Such models are most flexibly fit within a Bayesian framework with intuitive interpretation of these parameters and a highly flexible path for error propagation. We show below how such a model can be used to derive all thermal load sensitivity parameters of interest to ecophysiologists. 

# Deriving key thermal load sensitivity parameters 

The TLS framework is elegant in that it derives metrics with clear biological interpretation that can be compared across populations and species (e.g., *z*, $CT_{max_{1hr}}$). Here, we show that we can derived these same metrics directly from @eq-4pl. 

As indicated above, parameters can be fit to model changes in the mean of any of the four parameters in @eq-4pl. We can easily derive thermal sensitivity (*z*) and $CT_{max_{1hr}}$ by modelling the effect of temperature on midpoints as follows:

$$
\text{mid}_{ij} = \beta_0 + \beta_1\,(T_{ij} - \bar{T}),
$$ {#eq-mid-temp}

where $\text{mid}_{ij}$ is the midpoint ($\log_{10}$-time at 50% survival between $\text{low}$ and $\text{up}$) for trial $j$ at assay temperature $T_{ij}$ (°C); $\beta_0$ is the intercept giving the midpoint at the grand mean temperature $\bar{T}$; $\beta_1$ is the slope describing how the midpoint changes per °C increase in $T_{ij}$; and $\bar{T}$ is the grand mean of assay temperatures used to mean-centre $T_{ij}$, which decorrelates $\beta_0$ from $\beta_1$ and improves parameter estimation and interpretation [@schielzeth_centering_2010]. 

Setting $p_{ij} = 0.5$ in @eq-4pl, isolating the exponential, and taking natural logs gives $\log_{10}t_{50}(T) = \text{mid}(T) + (1/k)\log[(\text{up} - 0.5)/(0.5 - \text{low})]$. Substituting into @eq-mid-temp and grouping all terms that do not depend on $T$ into an intercept, $\alpha$, recovers the classical TDT linear relation:

$$
\log_{10} t_{50}(T) \;=\; \underbrace{\left[\beta_0 + \tfrac{1}{k}\log\!\tfrac{\text{up} - 0.5}{0.5 - \text{low}} - \beta_1\bar{T}\right]}_{\alpha} \;+\; \beta_1\,T,
$$ {#eq-t50}

Note that @eq-t50 subtracts $\beta_1\bar{T}$ in $\alpha$ because we mean-centred $T$. With this subtraction, the intercept estimates $\log_{10}t_{50}$ at the average temperature ($\bar{T}$); without it, the intercept estimates $\log_{10}t_{50}$ at $T = 0$°C. From @eq-t50, the classical TLS quantities fall out directly. Thermal sensitivity is $z = -1/\beta_1$, and the critical thermal limit at any reference time $t_\text{ref}$ is $CT_{max}(t_\text{ref}) = (\log_{10} t_\text{ref} - \alpha)/\beta_1$ (with $t_\text{ref} = 60$ min for $CT_{max_{1hr}}$). Each of these quantities can be estimated separately for different strata (e.g., sex, population, species) by including the relevant fixed covariates and/or random effects in the linear predictor for $\text{mid}_{ij}$ — and, if appropriate, in the linear predictors for $\text{low}$, $\text{up}$, and $k$, each of which is modelled separately in the joint Bayesian fit (see @eq-linpred below).


The same derivation generalises beyond the median: the Bayesian dose-response fit gives access to *any* mortality percentile, not just $t_{50}$ [@rezende_drosophila_mortality_2020; @jorgensen_unifying_2021]. Setting $p_{ij} = x/100$ in @eq-4pl and following the same algebra used for $t_{50}$, the posterior samples of $(\text{low}, \text{up}, k, m(T))$ deliver the time at which $x\%$ of the population reaches the assayed endpoint:

$$
\log_{10} t_x(T) \;=\; m(T) \;-\; \frac{1}{k}\,\ln\!\frac{\text{up} - x/100}{x/100 - \text{low}},
$$ {#eq-LTx}

Varying $x$ in @eq-LTx then yields any threshold of interest. For instance, $t_{10}$ marks the onset of early mortality — useful for risk management — while $t_{90}$ is closer to a population-collapse threshold. As with $t_{50}$, every prediction inherits the joint posterior, so uncertainty bands can be reported alongside the population-level median.

Importantly, the slope $\beta_1$ — and hence $z$ — is robust to estimated asymptotes because the asymmetry correction $\tfrac{1}{k}\log\tfrac{\text{up} - 0.5}{0.5 - \text{low}}$ does not depend on $T$. Instead, it shifts $\log_{10}t_{50}$ vertically without changing the slope. In contrast, threshold summaries are not robust in this way; when partial mortality or sparse data do not estimate the asymptotes to be around 0 and 1, the asymmetry correction must be retained — added to the naive $\log_{10}t_{50}$ (@eq-t50-corrected) and subtracted, after division by the slope (to convert the vertical shift in $\log_{10}t_{50}$ into the corresponding shift in temperature), from the naive $CT_{max}$ (@eq-ctmax-corrected):

$$
\log_{10}t_{50}(T) \;=\; \underbrace{\beta_0 + \beta_1(T - \bar{T})}_{\text{naive estimate}} \;+\; \underbrace{\tfrac{1}{k}\log\!\tfrac{\text{up} - 0.5}{0.5 - \text{low}}}_{\text{correction}},
$$ {#eq-t50-corrected}

$$
CT_{max}(t_\text{ref}) \;=\; \underbrace{\frac{\log_{10}t_\text{ref} - \beta_0 + \beta_1\bar{T}}{\beta_1}}_{\text{naive estimate}} \;-\; \underbrace{\tfrac{1}{\beta_1}\cdot\tfrac{1}{k}\log\!\tfrac{\text{up} - 0.5}{0.5 - \text{low}}}_{\text{correction}}.
$$ {#eq-ctmax-corrected}

In practice, however, these corrections may not be needed so long as the experimental design includes treatments that span the full mortality range which anchors $\text{low}$ and $\text{up}$ tightly near 1 and 0 and makes the correction term effectively zero. When this is the case, $\log_{10}t_{50}$ and $CT_{max}$ can be read directly from the naive forms in @eq-t50-corrected and @eq-ctmax-corrected. Regardless, we recommend using the correction factor as a safeguard. However, computing both the naive and corrected estimates and comparing them is a useful sensitivity check for whether the correction is needed in any given dataset.

# Predicting heat injury and survival under dynamic thermal environments

A particularly powerful aspect of TLS models are their ability to translate dynamic temperature times series' from nature to predict heat injury (HI) accumulation and thus population survival probability [@jorgensen_unifying_2021; @orsted_suzukii_2024; @li2023; @orsted_finding_2022; @rezende_drosophila_mortality_2020]. Heat injury accumulates as the time spent at a given temperature divided by the median lethal time at that temperature, integrated across the temperature trajectory $T(i)$ as follows:

$$
HI(t) \;=\; 100 \int_0^t 10^{\,\big(T(i)\,-\,CT_{max_{1\text{hr}}}\big)/z} \,di,
$$ {#eq-hi}

where $z$ is the thermal sensitivity and $CT_{max_{1\text{hr}}}$ is the static temperature at which 50% of the population reaches the assayed endpoint after one hour of exposure. When $HI(t) = 100\%$, the cumulative dose equals one $t_{50}$ at the 1-hour reference, predicting 50% mortality (or 50% loss of the assayed sublethal endpoint, e.g.\ fertility; @orsted_suzukii_2024). Injury accumulates only when $T(i)$ exceeds a critical temperature $T_c$ below which damage is repaired as fast as it accrues [@jorgensen_unifying_2021; @orsted_finding_2022]. Field temperature records are typically available at hourly resolution. @eq-hi can then be evaluated by summing $100 \cdot 10^{(T_i - CT_{max_{1\text{hr}}})/z}$ across each hourly interval whose temperature $T_i$ exceeds $T_c$, following the implementations in @jorgensen_unifying_2021, @orsted_suzukii_2024, and @li2023. To propagate uncertainty, $z$ and $CT_{max_{1\text{hr}}}$, calculated from our Bayesian dose-response model can be used to calculate $HI(t)$ through time by substituting samples from the posterior distribution for each parameter into @eq-hi. 

A more natural prediction to make may instead be the population-level survival probability over time. This has the advantage of allowing for a direct connection to population dynamical models which typically use survival, growth and fecundity as vital rates for population-level predictions [@noble2026]. To predict survival, we can calculate the *effective exposure time* at a chosen reference temperature $T_{\text{ref}}$. More effective time is accumulated at hot temperatures and less at cool ones, in proportion to the scaling factor $10^{(T(i) - T_{\text{ref}})/z}$,

$$
\tau_{\text{eff}}(t) \;=\; \int_0^t 10^{\,(T(i) - T_{\text{ref}})/z}\,di,
$$ {#eq-tau-eff}

We can then evaluate the 4PL surface once at the cumulative effective exposure to obtain the predicted population-level survival fraction at every point in the trajectory:

$$
S_{\text{pred}}(t) \;=\; S\!\big(T_{\text{ref}},\, \tau_{\text{eff}}(t)\big).
$$ {#eq-S-pred}

Here $S(T, \tau)$ is the survival surface fitted by the dose-response model, so $S_{\text{pred}}(t)$ is the survival probability for an organism held at *constant* $T_{\text{ref}}$ after exposure duration $\tau_{\text{eff}}(t)$. 

# Application of Bayesian dose-response models to case studies

We applied the 4-parameter (4PL) Bayesian logistic model to four case studies covering a broad range of ectothermic groups. For all case studies, we show how much a modelling framework can retrieve classical parameters of the TLS framework such as $z$ and $CT_{max1h}$ using both lethal and sublethal data. We also show how to derive an estimate of $T_{crit}$ from sublethal data (e.g., time to heat knockdown). 

In brown shrimp (*Crangon crangon*), we first show how the 4PL model can be used to generate survival curves, thermal-death-time relationships, and thermal tolerance landscapes with complete uncertainty propagation. We also show how the 4PL can be used to predict survival under dynamic thermal regimes. We then also show how to apply this model to sublethal data (time to knockdown), based on the proportion of live and knocked down individuals across time and temperature. 

In zebrafish (*Danio rerio*), we show how the 4PL can be used to estimate all parameters and tolerance landscapes across three distinct life stages: early embryos, late embryos, and larvae. We show how to define models to derive group-specific parameters and compare posterior distributions to make inference about differences in thermal sensitivity and tolerance across life stages. 

In XXX... 


To break the well-known likelihood ridges among $(\text{low}, u, k)$ in unconstrained 4PL fits [@bolker_ecologicalmodels_2008; @gelman_priors_2006], we estimate the identifiable parameterisation

$$
\text{low} = 0.49\,\text{logit}^{-1}(\widetilde{\text{low}}), \qquad
\text{up}   = 0.51 + 0.49\,\text{logit}^{-1}(\widetilde{\text{up}}), \qquad
k   = \exp(\kappa),
$$ {#eq-reparam}

with $\widetilde{\text{low}},\widetilde{\text{up}},\kappa \in \mathbb{R}$. This constrains $\text{low} \in [0, 0.49]$ and $\text{up} \in [0.51, 1]$ into disjoint intervals so that $u > \text{low}$ by construction, the curve direction is unambiguous, and $k$ is positive on the natural scale.

The inflection point $m_{ij}$ is modelled as a linear function of (mean-centred) assay temperature, with additional fixed covariates and nested random effects:

$$
m_{ij} \;=\; \beta_0 \;+\; \beta_1\,(T_{ij} - \bar T) \;+\; \mathbf{x}_{ij}^{\top}\boldsymbol{\gamma} \;+\; b_{j} \;+\; \varepsilon_{ij},
$$ {#eq-linpred}

where $\bar T$ is the grand mean of assay temperatures (centring stabilises sampling and decorrelates $\beta_0$ from $\beta_1$), $\mathbf{x}_{ij}$ holds any additional fixed covariates with coefficients $\boldsymbol{\gamma}$, $b_{j}$ collects the nested random effects and $\varepsilon_{ij} \sim \mathcal{N}(0, \sigma_\varepsilon^2)$ is an observation-level random effect that absorbs the binomial overdispersion characteristic of replicated count assays [@harrison_olre_2014]. However, we could also remove the observation-level random effect from the model and instead fit a beta-binomial error distribution which estimates a dispersion parameter ($\phi$) more naturally.

Together, @eq-likelihood–@eq-linpred define the hierarchical Bayesian 4PL fit directly to the raw binomial counts. Every classical TDT quantity — slope, $z$, $CT_{max}$ at any reference time, and the heat-injury integral — is recovered as a one-line posterior summary of this single joint fit, as derived in §[Equivalence with the classical pipeline].







## What the 4PL delivers that the two-step cannot

# Worked example: brown shrimp thermal tolerance

<!--
Multi-endpoint demonstration on Crangon crangon.
Source pipeline: R/bayesian_TDT_shrimp.qmd.
-->

## Data and experimental design

<!-- Placeholder: a flextable summarising n trials, n individuals, temperature
     range, and exposure range for each of the three endpoints will be added
     here once the case-study analyses are complete. The chunk is omitted
     while empty so it doesn't produce a dangling cross-reference. -->


## Lethal TDT — hierarchical 4PL on counts

```{r fig-survival-curves}
#| label: fig-survival-curves
#| fig-cap: "Posterior survival curves at each lethal-TDT assay temperature, with 95% credible bands and observed survival proportions."
#| eval: false
```

## Sublethal time-to-death

## Dynamic $CT_{max}$ across ramping rates

## Reconciling the three endpoints via the Jørgensen ramping equation

```{r fig-tdt-ctmax-reconciliation}
#| label: fig-tdt-ctmax-reconciliation
#| fig-cap: "Forward (TDT → dynamic $CT_{max}$) and inverse ($CT_{max}$ → TDT) reconciliation under full posterior propagation."
#| eval: false
```

## Field-temperature heat-injury projections under warming scenarios

```{r fig-hi-warming}
#| label: fig-hi-warming
#| fig-cap: "Posterior cumulative heat-injury and survival under measured Baltic surface-temperature traces and +1, +2, +3 °C warming scenarios, with and without a Sharpe–Schoolfield repair kernel."
#| eval: false
```

# Discussion

## What changes for the field if the hierarchical 4PL becomes the default

## Limits of the present analysis

<!--
Honest discussion of:
- Repair kernel parameters fixed, not fit
- d$CT_{max}$ reconciliation uses nls() per draw rather than a joint Stan likelihood
- Arrhenius (1/T-Kelvin) sensitivity check belongs in the supplement
- Endpoint heterogeneity is not eliminated, only made visible
-->

## Recommendations for primary studies and meta-analyses

# Conclusions

# Acknowledgements

# Author contributions {.unnumbered}



# Data and code availability {.unnumbered}

All data, analysis scripts, and the rendered manuscript and supplement are available at https://github.com/daniel1noble/bayesTLS.

# Conflict of interest {.unnumbered}

The authors declare no competing interests.

# References {.unnumbered}

::: {#refs}
:::
