---
title: "Supplementary Information"
subtitle: "A flexible modelling framework for estimating thermal sensitivity across life"
shorttitle: "Bayesian models of thermal tolerance — Supplementary information"
authors:
  - name: "Daniel W.A. Noble"
    email: daniel.noble@anu.edu.au
    corresponding: true
    affiliations:
      - ref: anu
  - name: "Pieter A. Arnold"
    affiliations:
      - ref: anu
  - name: "Patrice Pottier"
    email: patrice.pottier@bioenv.gu.se
    affiliations:
      - ref: gu
affiliations:
  - id: anu
    name: Division of Ecology and Evolution, Research School of Biology, Australian National University
    city: Canberra
    country: Australia
  - id: gu
    name: Department of Biological and Environmental Sciences, University of Gothenburg
    city: Gothenburg
    country: Sweden
date: today
output-file: supp
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    number-sections: true
    embed-resources: true
    fig-cap-location: bottom
  docx:
    reference-doc: ../bib/templates/template.docx
    toc: false
    number-sections: true
  pdf:
    documentclass: scrartcl
    top-level-division: section
    keep-tex: true
    geometry:
      - margin=1in
    fontsize: 11pt
    toc: true
    toc-depth: 3
    number-sections: true
    fig-pos: "H"
    colorlinks: true
    link-citations: true
    include-in-header:
      text: |
        % tinytable LaTeX dependencies. tinytable normally injects these
        % via knitr::knit_meta_add() when its tt() chunks execute, but
        % cache-hit renders skip that injection and the build fails with
        % "Environment tblr undefined". Including the (idempotent)
        % packages here makes the preamble correct in either case.
        % `\UseTblrLibrary{siunitx}` is intentionally omitted — including
        % it manually conflicts with tinytable's auto-inject when chunks
        % run fresh, and our tables don't use S-aligned columns.
        \usepackage{tabularray}
        \usepackage{float}
        \usepackage{graphicx}
        \usepackage[normalem]{ulem}
        \providecommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
        \providecommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
        % Author block: Quarto's default \author{names \and …} gives
        % names only. We override with authblk to pin affiliations from
        % YAML to the title page.
        \usepackage[affil-it]{authblk}
        \renewcommand\Authsep{, }
        \renewcommand\Authand{, }
        \renewcommand\Authands{, }
        \renewcommand\Affilfont{\itshape\small}
        \let\OldAuthor\author
        \renewcommand{\author}[1]{}
        \AtBeginDocument{%
          \OldAuthor[1]{Daniel W.A. Noble\thanks{Corresponding author: \texttt{daniel.noble@anu.edu.au}}}%
          \OldAuthor[1]{Pieter A. Arnold}%
          \OldAuthor[2]{Patrice Pottier}%
          \affil[1]{Division of Ecology and Evolution, Research School of Biology, Australian National University, Canberra, Australia}%
          \affil[2]{Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden}%
        }
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
editor_options:
  chunk_output_type: console
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

::: {.content-visible when-format="docx"}
**Daniel W.A. Noble**^1,\*^, **Pieter A. Arnold**^1^, **Patrice Pottier**^2^

^1^Division of Ecology and Evolution, Research School of Biology, Australian National University, Canberra, Australia.
^2^Department of Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden.
^\*^Corresponding author: daniel.noble@anu.edu.au
:::

```{r setup}
#| label: setup-supp
#| cache: false

# Setup chunk must NOT be cached — caching skips the package-loading side
# effects on subsequent renders, so library() calls in pacman::p_load() never
# run and downstream chunks fail with "could not find function ...".
pacman::p_load(here, dplyr, tidyr, tibble, purrr, ggplot2, brms, posterior, tidybayes, tinytable, patchwork, pander, readxl)

# Force tinytable to register its LaTeX preamble dependencies (tabularray,
# siunitx, etc.) on every render. Tinytable injects these via
# knit_meta_add() when tt() is called, but if all tt() chunks are cached,
# the side-effect doesn't fire and PDF builds fail with "Environment tblr
# undefined". Calling tt() here (in this cache: false chunk) guarantees
# the dependencies are always registered.
invisible(tinytable::tt(data.frame(x = 1)))

set.seed(123)

# Cache directory for brms fits (per CLAUDE.md: save fits manually as .rds).
models_dir <- here::here("output", "models")
if (!dir.exists(models_dir)) dir.create(models_dir, recursive = TRUE)
```

# **Introduction**

This supplement walks through the joint Bayesian 4-parameter logistic (4PL) framework from the main manuscript using the **`bayesTLS`** R package, which ships with this project. It provides functions to fit the 4PL, extract the classical TLS quantities ($z$, $CT_{max_{1hr}}$, $T_{crit}$) with full posterior uncertainty, and use them to predict heat-injury accumulation and survival fraction under arbitrary temperature time-series, optionally with a Sharpe-Schoolfield repair kernel.

Install once from GitHub, then load:

```{r install-bayesTLS}
#| eval: false
#| echo: true
# install.packages("remotes")   # if not already installed
remotes::install_github("daniel1noble/bayesTLS")
```

```{r load-bayesTLS}
#| label: load-bayesTLS
#| echo: true
#| cache: false
library(bayesTLS)
```

The functions chain together as a single pipeline:

1. **Standardise** raw experimental data to a common column schema (`temp`, `duration`, `n_total`, `n_surv`, …).
2. **Fit** the joint Bayesian 4PL with the built-in `beta_binomial(link = "identity")` family and asymptote-bounded reparameterisation.
3. **Extract** $z$, $CT_{max_{1hr}}$, and $T_{crit}$ — all with full posterior uncertainty.
4. **Predict** survival and heat-injury accumulation under fluctuating field temperature traces, with optional repair.
5. **Plot** each step with shared theming so figures across the workflow read consistently.

Every exported function is documented with roxygen2 (`?fit_4pl`, etc.) and the package ships with `testthat` tests — fast unit tests plus a gated set of `brms`-fitting integration tests that assert recovery of known simulation truths. `R CMD check` runs cleanly. The remainder of the supplement first walks the pipeline through a controlled simulation where the truth is known (§ A Tutorial with Simulations), then applies the same machinery end-to-end to a real lethal-TDT dataset on brown shrimp (§ Case Study 1: Shrimp data).

## *What each function does* {#sec-function-tour}

@tbl-function-tour lists every exported function in the library, grouped by pipeline stage. Each function is single-purpose; the full roxygen2 documentation (with runnable `@examples`) lives in the corresponding R file.

```{r function-tour-table}
#| label: tbl-function-tour
#| tbl-cap: "Functions exported by the TDT analysis library, grouped by pipeline stage. Each function is documented with roxygen2 in its R file; runnable examples accompany every function."

tdt_function_tour <- tibble::tribble(
  ~Stage,             ~Function,                            ~Purpose,
  "Data",             "standardize_data()",                 "Rename raw columns to project schema; attach metadata.",
  "Model spec",       "make_4pl_priors()",                  "Weakly informative priors for the disjoint-bounds 4PL.",
  "Model spec",       "make_4pl_formula()",                 "brms non-linear formula; identity link, all four sub-parameters get temp_c.",
  "Fit",              "fit_4pl()",                          "Joint Bayesian 4PL fit; returns a tdt_4pl_workflow object.",
  "Diagnostics",      "diagnose_tdt_fit()",                 "Rhat / ESS / divergences / treedepth, with pass-fail flags.",
  "Diagnostics",      "tdt_parameter_table()",              "Posterior summary on the natural scale (low, up, k, mid, z).",
  "Predictions",      "predict_survival_curves()",          "Survival curves on a temp x duration grid with credible bands.",
  "Predictions",      "derive_tdt_landscape()",             "Dense 2-D survival surface for a heatmap.",
  "TDT quantities",   "derive_ltx_curve()",                 "Time-to-x-survival across temperature (per-draw inversion).",
  "TDT quantities",   "derive_temperature_for_duration()",  "Temperature at survival = X for a fixed exposure duration.",
  "TDT quantities",   "derive_tdt_parameters()",            "Per-draw log-linear regression of LT_x on temperature; z + CTmax.",
  "TDT quantities",   "extract_tdt()",                      "Bundled wrapper: z and CTmax always; T_crit opt-in via `lethal = TRUE` (rate-multiplier; lethal-endpoint data only).",
  "Heat injury",      "make_temperature_scenarios()",       "Three reference traces (flat / single spike / multi-spike).",
  "Heat injury",      "planted_dose_from_trace()",          "Analytical HI under known z, CTmax, T_c (validation truth).",
  "Heat injury",      "predict_heat_injury()",              "Posterior HI and survival under a temperature trace.",
  "Heat injury",      "repair_rate_schoolfield()",          "Sharpe-Schoolfield repair TPC (Kelvin internally).",
  "Plotting",         "plot_survival_curves()",             "Posterior curves with observed proportion overlay.",
  "Plotting",         "plot_ltx_curve()",                   "LT_x curve on a log-time axis (classical TDT view).",
  "Plotting",         "plot_tdt_landscape()",               "Survival heatmap with contour overlay.",
  "Plotting",         "plot_temperature_density()",         "Posterior density for CTmax / T_crit with a 95-percent CrI bar.",
  "Plotting",         "plot_temperature_scenarios()",       "Three traces stacked vertically.",
  "Plotting",         "plot_heat_injury()",                 "HI and predicted survival trajectories with CrI bands.",
  "Plotting",         "plot_repair_rate()",                 "Sharpe-Schoolfield TPC curve."
)

tinytable::tt(tdt_function_tour, escape = TRUE)
```

# **A Tutorial with Simulations**

This section is a guided tutorial using simulated data. We show, step by step, how the Bayesian hierarchical 4PL — under both binomial sampling and the more realistic beta-binomial (overdispersed) case – can recover thermal sensitivity ($z$) and $CT_{max_{1hr}}$ from the single joint model, yet provide substantial downstream benefits in terms of propagating uncertainty and giving more flexibility. We then complement the simulation with a few case studies which walk through real data that is presented within the main manuscript in other sections.

### *Simulating Data*
#### Step 1 — Setting up parameters {#sec-sim-step1}

We choose values that might be typical of a thermal mortality assay. The four 4PL parameters are the lower and upper asymptotes ($\ell, u$), the slope on the $\log_{10}$-time axis ($k$), and a midpoint that varies linearly with temperature: $\text{mid}(T) = \beta_0 + \beta_1(T - \bar T)$. Each value below has a clear biological meaning.

```{r sim-params}
#| label: sim-params
#| echo: true

true_params <- list(
  ell      = 0.03,   # 3% of individuals survive even at the longest exposures
  u        = 0.97,   # 97% are alive at very short exposures
  k        = 8,      # steepness of the survival drop on the log10(t) axis
  m_beta0  = 1.5,    # midpoint at the grand-mean temperature: ~31.6 min
  m_beta1  = -0.15,  # midpoint drops 0.15 log10-min per °C; z = -1/beta1 ≈ 6.7 °C
  T_bar    = 34      # grand-mean temperature
)
```

The implied thermal sensitivity is $z = -1/\beta_1 \approx 6.7$°C — the temperature change that shifts LT50 by a factor of 10. From the same parameters we can also compute the simulation truth for the uncentred TDT intercept $\alpha = \beta_0 + \tfrac{1}{k}\log\tfrac{u - 0.5}{0.5 - \ell} - \beta_1\bar{T}$ and the critical thermal limit at 1 hour $CT_{max_{1hr}} = (\log_{10} 60 - \alpha)/\beta_1$ — both approaches below will be checked against these known values.

Recall that we can calculate $\alpha\$, $z$ and $CT_{max_{1hr}}$ from the true parameters using the equations in the main manuscript. We'll do that now so we have a comparator for later.

```{r sim-truth}
#| label: sim-truth
#| echo: true

true_z      <- -1 / true_params$m_beta1
true_alpha  <- true_params$m_beta0 +
  (1 / true_params$k) *
  log((true_params$u - 0.5) / (0.5 - true_params$ell)) -
  true_params$m_beta1 * true_params$T_bar
true_CTmax  <- (log10(60) - true_alpha) / true_params$m_beta1

# T_crit under the rate-multiplier definition (see §[Deriving z, CTmax, and
# T_crit from the posterior]). For each posterior draw, T_crit is computed as
# CTmax_1hr + z * log10(r*/100) where r* is sampled uniformly on log10
# across the default range c(0.1, 1) % HI per hour. The MEDIAN of that
# distribution lies at log10(r*/100) = -2.5 (the geometric mean of the
# range), so:
true_T_crit <- true_CTmax - 2.5 * true_z
```

#### Step 2 — Simulated study design {#sec-sim-step2}

We mimic a textbook TDT experiment: a grid of five assay temperatures crossed with six log-spaced exposure durations, repeated 30 times per cell, with 30 individuals tested per replicate. That gives 900 trials and 27,000 individuals — large enough to recover the parameters with negligible uncertainty, but small enough to fit in a few minutes. Of course, in practice we will not have this much data as will be seen in the case studies, but simulations are a good starting point to understand the models being fit and the conversions. 

```{r sim-design}
#| label: sim-design
#| echo: true

temps     <- c(30, 32, 34, 36, 38)             # 5 assay temperatures (°C)
durations <- c(1, 5, 15, 45, 135, 405)         # 6 durations (minutes, log-spaced)
n_rep     <- 30                                # 30 replicates per cell
n_per_rep <- 30                                # 30 individuals per replicate

design <- tidyr::expand_grid(
  T   = temps,
  t   = durations,
  rep = seq_len(n_rep)
) |>
  dplyr::mutate(
    log10_t = log10(t),
    T_c     = T - true_params$T_bar,                        # mean-centred T
    mid     = true_params$m_beta0 + true_params$m_beta1 * T_c,  # mid(T)
    p_true  = true_params$ell +
      (true_params$u - true_params$ell) /
      (1 + exp(true_params$k * (log10_t - mid))),
    n = n_per_rep
  )
```

#### Step 3 — Look at the truth before simulating {#sec-sim-step3}

Before generating noisy data, let's plot the *true* survival surface (@fig-true-surface) so we know what the model is being asked to recover. Each line is the 4PL at one assay temperature, plotted against $\log_{10}$ exposure time.

::: {#fig-true-surface}

```{r true-surface}
#| label: plot-true-surface
#| fig-width: 7
#| fig-height: 4

truth_grid <- tidyr::expand_grid(
  T = temps,
  t = 10 ^ seq(log10(0.5), log10(1000), length.out = 200)
) |>
  dplyr::mutate(
    log10_t = log10(t),
    T_c     = T - true_params$T_bar,
    mid     = true_params$m_beta0 + true_params$m_beta1 * T_c,
    p_true  = true_params$ell +
      (true_params$u - true_params$ell) /
      (1 + exp(true_params$k * (log10_t - mid)))
  )

ggplot2::ggplot(truth_grid,
                ggplot2::aes(log10_t, p_true, colour = factor(T))) +
  ggplot2::geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::labs(
    x = expression(log[10]~exposure~time~(min)),
    y = "True survival probability",
    colour = "Temperature (°C)"
  ) +
  ggplot2::theme_minimal()
```

The known 4PL dose-response surface used to generate the simulated data, evaluated on a dense time grid at each of the five assay temperatures. The horizontal line marks 50% survival; where each curve crosses it gives the true LT50 at that temperature.
:::

Notice the curves shift left as temperature rises: hotter temperatures kill faster, so the LT50 is reached at shorter durations. That horizontal shift in midpoints is exactly what $\beta_1$ encodes. 

#### Step 4 — Generate noisy data: binomial and beta-binomial {#sec-sim-step4}

Real survival counts have at least two flavours of noise. **Binomial** sampling is the textbook case: each replicate's survival count is a draw from $\text{Binomial}(n, p)$ with $p$ from the 4PL surface. **Beta-binomial** sampling adds *overdispersion* — replicate-to-replicate variability in $p$ that the binomial cannot capture, e.g., from cohort effects or unmodelled heterogeneity. Real TDT data almost always show overdispersion so in many cases the beta binomial will be a better fit.

```{r sim-data}
#| label: sim-data
#| echo: true

# (a) Binomial: y ~ Binomial(n, p_true)
sim_bin <- design |>
  dplyr::mutate(y = rbinom(dplyr::n(), size = n, prob = p_true))

# (b) Beta-binomial: replicate-level p drawn from Beta(p*phi, (1-p)*phi)
phi <- 5  # smaller phi = more overdispersion
sim_bb <- design |>
  dplyr::mutate(
    p_draw = rbeta(dplyr::n(), p_true * phi, (1 - p_true) * phi),
    y      = rbinom(dplyr::n(), size = n, prob = p_draw)
  )
```

### *The classical two-stage TDT pipeline* {#sec-twostage}

We first run the simulated data through the **classical two-stage TDT pipeline** [@rezende_landscapes_2014; @orsted_finding_2022], which is dominant in the thermal-tolerance literature.

1. **Stage 1.** At each assay temperature, fit a logistic dose-response curve to the count data and read off a point estimate of $\log_{10}\text{LT50}_T$.
2. **Stage 2.** Regress those Stage-1 point estimates on assay temperature with ordinary least squares, then derive $z = -1/\beta_1$ and $CT_{max_{1hr}} = (\log_{10} 60 - \alpha)/\beta_1$ from the fitted line.

#### Stage 1 — Dose-response curves at each temperature {#sec-twostage-step1}

We fit a binomial GLM with a logit link separately at each assay temperature. This is the most-common Stage-1 specification in the TDT literature: a two-parameter logistic with asymptotes constrained to 0% and 100%. The midpoint $\log_{10}\text{LT50}_T = -\beta_0/\beta_1$ falls out of the GLM's two coefficients.

```{r stage1-twostage}
#| label: stage1-twostage
#| echo: true

fit_stage1 <- function(sim_data) {
  sim_data |>
    dplyr::group_by(T) |>
    dplyr::group_modify(function(d, key) {
      f  <- glm(cbind(y, n - y) ~ log10_t, data = d, family = binomial())
      co <- coef(f)
      tibble::tibble(log10_lt50 = as.numeric(-co[1] / co[2]))
    }) |>
    dplyr::ungroup()
}

stage1_bin <- fit_stage1(sim_bin)
stage1_bb  <- fit_stage1(sim_bb)
```

```{r stage1-lt50-table}
#| label: tbl-stage1-lt50
#| tbl-cap: "Per-temperature Stage-1 estimates of $\\log_{10}\\text{LT50}$ from a binomial GLM with logit link, fit separately at each of the five assay temperatures. Truth values are the simulated midpoints (which coincide with $\\log_{10}\\text{LT50}$ here because the simulation asymptotes are near 0 and 1)."

stage1_table <- dplyr::full_join(
  stage1_bin |> dplyr::rename(bin_log10_lt50 = log10_lt50),
  stage1_bb  |> dplyr::rename(bb_log10_lt50  = log10_lt50),
  by = "T"
) |>
  dplyr::mutate(
    truth_log10_lt50 = true_params$m_beta0 +
      true_params$m_beta1 * (T - true_params$T_bar)
  ) |>
  dplyr::select(T, truth_log10_lt50, bin_log10_lt50, bb_log10_lt50) |>
  dplyr::rename(
    `T (°C)`             = T,
    `Truth log10(LT50)`  = truth_log10_lt50,
    `Binomial log10(LT50)` = bin_log10_lt50,
    `Beta-binomial log10(LT50)` = bb_log10_lt50
  )

tinytable::tt(stage1_table, digits = 3, escape = TRUE)
```

With the simulation's true asymptotes near 0 and 1, the 2PL's forced-asymptote constraint introduces negligible bias here, so each Stage-1 estimate sits close to the simulation truth.

#### Stage 2 — log-linear TDT regression {#sec-twostage-step2}

Stage 2 regresses Stage-1 $\log_{10}\text{LT50}$ on assay temperature with ordinary least squares. The classical TDT quantities — $z = -1/\beta_1$ and $CT_{max_{1hr}} = (\log_{10} 60 - \alpha)/\beta_1$ — fall out of the fitted line.

```{r stage2-twostage}
#| label: stage2-twostage
#| echo: true

fit_stage2 <- function(stage1) {
  fit2 <- lm(log10_lt50 ~ T, data = stage1)
  co   <- coef(fit2)

  tibble::tibble(
    quantity = c("z (°C)", "CTmax at 1 hr (°C)"),
    est      = c(as.numeric(-1 / co["T"]),
                 as.numeric((log10(60) - co["(Intercept)"]) / co["T"]))
  )
}

ts_bin <- fit_stage2(stage1_bin)
ts_bb  <- fit_stage2(stage1_bb)
```

@fig-stage2-regression visualises the Stage-2 fit. Each point is one Stage-1 $\log_{10}\text{LT50}$ estimate from @tbl-stage1-lt50; the solid line is the OLS regression through them. The slope is $\beta_1$, from which $z = -1/\beta_1$ — i.e., the temperature increase that causes LT50 to change by a factor of 10. The line's height at $\log_{10} 60 \approx 1.78$ (1 hour) is at the temperature equal to $CT_{max_{1hr}}$ — the vertical dotted line drops from this intersection straight to the temperature axis to mark it. The dashed grey line is the simulation truth.

::: {#fig-stage2-regression}

```{r stage2-regression-plot}
#| label: plot-stage2-regression
#| fig-width: 7
#| fig-height: 4.5

stage1_combined <- dplyr::bind_rows(
  stage1_bin |> dplyr::mutate(dataset = "Binomial"),
  stage1_bb  |> dplyr::mutate(dataset = "Beta-binomial")
)

truth_line <- tibble::tibble(
  T = seq(min(temps), max(temps), length.out = 100)
) |>
  dplyr::mutate(
    log10_lt50 = true_params$m_beta0 +
      true_params$m_beta1 * (T - true_params$T_bar)
  )

ggplot2::ggplot(stage1_combined,
                ggplot2::aes(x = T, y = log10_lt50, colour = dataset)) +
  ggplot2::geom_hline(yintercept = log10(60),
                      linetype = "dotted", colour = "grey60") +
  ggplot2::geom_segment(
    ggplot2::aes(x = true_CTmax, xend = true_CTmax,
                 y = log10(60), yend = -Inf),
    inherit.aes = FALSE,
    linetype = "dotted", colour = "grey60"
  ) +
  ggplot2::geom_line(data = truth_line,
                     ggplot2::aes(x = T, y = log10_lt50),
                     inherit.aes = FALSE,
                     linetype = "dashed", colour = "grey40",
                     linewidth = 0.7) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  ggplot2::geom_point(size = 3) +
  ggplot2::annotate("text", x = max(temps), y = log10(60),
                    label = "log[10]~60~(1*' hour')", parse = TRUE,
                    hjust = 1, vjust = -0.4, colour = "grey50", size = 3) +
  ggplot2::annotate("text", x = true_CTmax, y = -Inf,
                    label = "CT[max[1*hr]]", parse = TRUE,
                    hjust = -0.15, vjust = -0.7, colour = "grey50", size = 3) +
  ggplot2::labs(
    x = "Assay temperature (°C)",
    y = expression(log[10]~LT[50]~(min)),
    colour = "Stage-1 dataset"
  ) +
  ggplot2::theme_classic()
```

Stage-2 OLS regression of $\log_{10}\text{LT50}$ on assay temperature, fit separately to the binomial and beta-binomial simulated datasets. Points are the Stage-1 per-temperature LT50 estimates; solid coloured lines are the OLS fits; the dashed grey line is the simulation truth. The slope of each line equals $\beta_1$, from which $z = -1/\beta_1$. The horizontal dotted line marks $\log_{10} 60$ (1 hour); the vertical dotted line drops from where the truth crosses 1 hour straight down to the temperature axis, marking $CT_{max_{1hr}}$ visually.
:::

This is the entire two-stage workflow. The point estimates that come out — $z$ and $CT_{max_{1hr}}$ — are what the field reports. We carry them forward (`ts_bin` and `ts_bb`) for the side-by-side comparison once we have the joint Bayesian results. 

::: {.callout-important}
Two-stage steps often ignore uncertainty in estimates in both fitting and error propagation. Uncertainty in estimates can be propagated using the SE's and the delta method, but only at each individual stage.
:::

### *A joint Bayesian 4PL* {#sec-joint}

Instead of fitting a separate logistic at each assay temperature and then regressing the LT50s, the joint fit estimates every parameter — $\ell$, $u$, $k$, $\beta_0$ and $\beta_1$ — in a single hierarchical model on the raw counts. We now use the same simulated data to derive the $z$ and $CT_{max_{1hr}}$ from the posterior, in one model rather than two.

#### Specifying the 4PL in `brms` {#sec-joint-spec}

The model has two pieces. The first is the 4PL itself — survival probability as a function of $\log_{10}$ exposure time:

$$
p_{ij} \;=\; \ell \;+\; \frac{u - \ell}{1 + \exp\!\bigl(k\,(\log_{10} t_{ij} - \text{mid}_{ij})\bigr)},
$$ {#eq-supp-4pl}

where $\ell$ and $u$ are the lower and upper asymptotes, $k$ is the slope on the $\log_{10} t$ axis, and $\text{mid}_{ij}$ is the value of $\log_{10} t$ at which the curve crosses the midpoint between $\ell$ and $u$. The second piece lets the midpoint vary linearly with mean-centred temperature:

$$
\text{mid}_{ij} \;=\; \beta_0 + \beta_1\,(T_{ij} - \bar{T}),
$$ {#eq-supp-mid-temp}

so that $\beta_1$ encodes how fast the dose-response curve shifts on the time axis as assay temperature changes; $\bar{T}$ is the grand-mean temperature, included so that $\beta_0$ is interpretable as the midpoint at the average assay temperature.

The function library exposes this model as a single call: `fit_4pl()` constructs the brms `bf()` formula, the default weakly-informative priors, and the sampling controls, and returns a workflow object containing the fit. Internally the formula uses the disjoint-bounds reparameterisation described in §[Loading the function library] — `low = 0.001 + \text{inv\_logit}(\text{lowraw}) \cdot 0.498` and `up = 0.501 + \text{inv\_logit}(\text{upraw}) \cdot 0.498` — so the 4PL output is guaranteed to lie in $(0, 1)$ regardless of where MCMC takes the unconstrained `lowraw`, `upraw`, and `logk` parameters. The default also puts a `temp_c` slope on every 4PL sub-parameter; when temperature has no real effect on a given parameter, the slope shrinks toward zero under the prior. We inspect the brms formula `fit_4pl()` builds with:

```{r show-formula}
#| label: show-formula
#| echo: true

make_4pl_formula()
```

::: {.callout-important}

The default model puts a `temp_c` slope on all four 4PL sub-parameters and includes no random effects. Random intercepts on `mid` can be added by passing `random_effects = c("Date", "Tank", ...)` to `fit_4pl()`. The simulation here has no batch structure so we leave it off; the shrimp case study below uses `random_effects = c("Date", "Tank")`.

:::

#### Standardising and fitting {#sec-joint-fit}

The first step is to pass the simulated data through `standardize_data()` so the columns match the schema the rest of the library expects (`temp`, `duration`, `logd`, `temp_c`, `n_total`, `n_surv`).

```{r standardise-sim}
#| label: standardise-sim
#| echo: true

std_bin <- standardize_data(sim_bin,
                            temp = "T", duration = "t",
                            n_total = "n", n_surv = "y",
                            duration_unit = "minutes")
std_bb  <- standardize_data(sim_bb,
                            temp = "T", duration = "t",
                            n_total = "n", n_surv = "y",
                            duration_unit = "minutes")
```

We now call `fit_4pl()` twice — once with the built-in `binomial(link = "identity")` family for the non-overdispersed simulation, and once with the default `beta_binomial(link = "identity")` for the overdispersed case. Caching is handled by brms's `file = ...` mechanism, so subsequent renders reload the cached fits unless the formula or data changes.

```{r fit-4pl-tutorial}
#| label: fit-4pl-tutorial
#| echo: true
#| cache: false

wf_bin <- fit_4pl(std_bin,
                  family = binomial(link = "identity"),
                  chains = 4, iter = 2000, cores = 4, seed = 123,
                  file = file.path(models_dir, "sim_4pl_binomial_v2"))

wf_bb  <- fit_4pl(std_bb,
                  chains = 4, iter = 2000, cores = 4, seed = 123,
                  file = file.path(models_dir, "sim_4pl_betabinomial_v2"))

# Convenience: extract the brmsfit object for downstream brms helpers.
fit_bin <- wf_bin$fit
fit_bb  <- wf_bb$fit
```

We can also get an idea of how much variation our model explains that considers all sources of relevant variation.

```{r, r2_bayes}
#| label: r2_bayes
#| echo: true

# Obtain the Bayes R2 for each model
brms::bayes_R2(fit_bin)
brms::bayes_R2(fit_bb)

```

#### Checking parameter recovery {#sec-joint-recovery}

We can now check how well our fitted model recovered the true parameters that generated the data. The most direct check is to compare posterior medians (with 95% credible intervals) to the known truth (@tbl-recovery). We have big sample sizes here so our values should be pretty close to our true values.

```{r recovery-table}
#| label: tbl-recovery
#| tbl-cap: "Posterior medians and 95% credible intervals for the four 4PL parameters and the temperature slope, compared to the known truth used to generate the simulated data. Recovery is judged successful when the credible interval covers the truth."

post_summary <- function(fit) {
  d <- posterior::as_draws_df(fit)
  # Transform the raw posterior columns back to the natural 4PL scale.
  # See compute_4pl_bounds(0, 1): low ∈ (0.001, 0.499), up ∈ (0.501, 0.999).
  draws <- tibble::tibble(
    ell               = 0.001 + plogis(d$b_lowraw_Intercept) * 0.498,
    u                 = 0.501 + plogis(d$b_upraw_Intercept)  * 0.498,
    k                 = exp(d$b_logk_Intercept),
    `mid (intercept)` = d$b_mid_Intercept,
    `mid (slope)`     = d$b_mid_temp_c
  )
  purrr::map_dfr(names(draws), \(p) tibble::tibble(
    param  = p,
    median = stats::median(draws[[p]]),
    lower  = stats::quantile(draws[[p]], 0.025, names = FALSE),
    upper  = stats::quantile(draws[[p]], 0.975, names = FALSE)
  ))
}

truth <- tibble::tibble(
  param = c("ell", "u", "k", "mid (intercept)", "mid (slope)"),
  truth = c(true_params$ell, true_params$u, true_params$k,
            true_params$m_beta0, true_params$m_beta1)
)

recovery <- truth |>
  dplyr::left_join(post_summary(fit_bin) |>
                     dplyr::rename_with(~ paste0("bin_", .x), -param),
                   by = "param") |>
  dplyr::left_join(post_summary(fit_bb) |>
                     dplyr::rename_with(~ paste0("bb_", .x), -param),
                   by = "param") |>
  dplyr::rename(
    Parameter        = param,
    Truth            = truth,
    `Bin. median`    = bin_median,
    `Bin. lower`     = bin_lower,
    `Bin. upper`     = bin_upper,
    `BB median`      = bb_median,
    `BB lower`       = bb_lower,
    `BB upper`       = bb_upper
  )

tinytable::tt(recovery, digits = 3, escape = TRUE)
```

We can also overlay the posterior fitted curves on the simulated data (@fig-recovery-fit). If the model has recovered the true surface, the fitted lines should track the simulated proportions at every assay temperature.

::: {#fig-recovery-fit}

```{r recovery-fit-plot}
#| label: plot-recovery-fit
#| fig-width: 8
#| fig-height: 5

# Use the library's predict_survival_curves() and plot_survival_curves().
psc_bin <- predict_survival_curves(
  wf_bin, temps = temps,
  durations = 10 ^ seq(log10(0.5), log10(800), length.out = 100),
  ndraws = 200
)

plot_survival_curves(psc_bin, observed = std_bin) +
  ggplot2::scale_x_log10() +
  ggplot2::labs(x = expression(exposure~time~(min)))
```

Posterior fitted 4PL curves (solid lines, with shaded 95% credible bands) overlaid on the simulated binomial data (points). One panel-coloured line per assay temperature. Where the fitted curves track the simulated proportions across all five temperatures, the model has recovered the underlying dose-response surface.
:::

### *Deriving $z$, $CT_{max_{1hr}}$, and $T_{crit}$ from the posterior* {#sec-joint-derive}

The whole point of fitting the joint Bayesian 4PL is that the classical TLS quantities fall out as transformations of the posterior we just fit — no second regression, no bootstrap. `extract_tdt()` always returns the first two and, opt-in via `lethal = TRUE`, also returns the third:

- **$z$** — thermal sensitivity, derived by a per-draw log-linear regression of $\log_{10}\text{LT50}$ on temperature (Equation 7 of the manuscript).
- **$CT_{max_{1hr}}$** — temperature at 50% survival after 1 hour exposure, obtained by inverting the fitted 4PL at the reference time, per posterior draw.
- **$T_{crit}$** *(only when `lethal = TRUE`)* — the temperature below which damage accumulates more slowly than a chosen rate floor. The damage rate at temperature $T$ in the TDT framework is $r(T) = 100 \cdot 10^{(T - CT_{max,1hr})/z}$ % LT50-dose per hour, so inverting for a chosen rate floor $r^*$ gives

$$
T_{crit}(r^*) \;=\; CT_{max,1hr} \;+\; z \cdot \log_{10}(r^* / 100).
$$ {#eq-tcrit}

Different $r^*$ values give different $T_{crit}$ thresholds, and the literature does not converge on a single value. Empirical breakpoint analyses across taxa as different as *Drosophila suzukii* and *Lemna gibba* fall in the range $r^* \in [0.1, 1]$ % HI per hour, corresponding to "$\sim 4$ days to 1000 hours of continuous exposure to reach $LT_{50}$" [@jorgensen_unifying_2021; @arnold_thermal_2025; @faber2026]. Rather than commit to one $r^*$, we treat it as an unknown with a uniform prior on $\log_{10} r^*$ over that range and integrate; the posterior of $T_{crit}$ then carries both parameter uncertainty (in $CT_{max,1hr}$ and $z$) and operational uncertainty (in the choice of $r^*$). The median of that pooled posterior sits at $CT_{max,1hr} - 2.5 \cdot z$, which corresponds to $r^* \approx 0.3$ % HI per hour (the geometric mean of the rate range).

::: {.callout-warning}
## $T_{crit}$ is meaningful only for lethal-endpoint data

The rate-multiplier $T_{crit}$ in Equation @eq-tcrit interprets $z$ as the slope of *damage accumulation* on temperature — strictly a property of lethal-endpoint dose-response data (proportion survival, knock-out counts where the endpoint is irreversible). For **sublethal** endpoints — knockdown time, photosystem-II decline, performance drop — the fitted $z$ instead measures the slope of *performance reduction* with temperature, which is typically far steeper than its damage-accumulation analogue. Plugging a sublethal $z$ into Equation @eq-tcrit pushes $T_{crit}$ implausibly low (the supplement's sublethal shrimp example demonstrates this: a sublethal $z \approx 1$ °C yields a $T_{crit}$ that would predict lethal exposures within ~2 hours at temperatures well within the organism's ecological range). Field practice treats $T_{crit}$ as the breakpoint where the thermal-performance curve and the thermal-death-time curve intersect [@jorgensen_unifying_2021; @orsted_finding_2022; @faber2026], which requires *both* lethal and sublethal data — not just one.

`extract_tdt()` reflects this by making $T_{crit}$ **opt-in**: the default `lethal = FALSE` returns only $z$ and $CT_{max_{1hr}}$, and users with lethal-endpoint data set `lethal = TRUE` to additionally derive $T_{crit}$. When `lethal = TRUE`, a one-line console message reiterates the lethal-only validity of the quantity.
:::

The two always-returned quantities inherit the same posterior, so every credible interval reflects joint uncertainty in $(\ell, u, k, \beta_0, \beta_1)$. The opt-in $T_{crit}$ pools that same posterior with the additional uniform prior on $\log_{10} r^*$.

::: {.callout-note}

The default `fit_4pl()` puts a `temp_c` slope on every 4PL sub-parameter, so $\ell$, $u$ and $k$ can each carry a linear effect of temperature in addition to the linear-mid term — when those slopes shrink to zero under the prior the model reduces cleanly to the constant-shape case. When the data carry real temperature signal on any of the shape parameters, the asymmetry correction $\tfrac{1}{k(T)}\log\tfrac{u(T)-0.5}{0.5-\ell(T)}$ varies with $T$ and the resulting $\log_{10}\text{LT50}(T)$ curve is bent rather than straight:

$$
\log_{10}\text{LT50}(T) \;=\; \beta_0 + \beta_1(T - \bar{T}) \;+\; \tfrac{1}{k(T)}\log\!\tfrac{u(T) - 0.5}{0.5 - \ell(T)}.
$$ {#eq-lt50-tvarying}

`extract_tdt()` requires no special handling in either regime, because both of its primitives operate on the *fitted 4PL surface* rather than on a closed-form expression: $CT_{max_{1hr}}$ is obtained by inverting the surface at the reference duration $t_\text{ref}$ per posterior draw (`derive_temperature_for_duration()`), and the per-draw OLS regression that yields $z$ runs over the per-draw $\log_{10}\text{LT50}(T)$ curve recovered numerically by `derive_ltx_curve()`. When the curve is straight, the OLS slope returns the closed-form $z$; when it is bent, the same OLS returns the *average local slope* across the assay grid — a sensible single-summary $z$, with the per-temperature local values accessible from the underlying $\text{LT50}(T)$ curve when needed. The worked example in @sec-joint-extend demonstrates this explicitly with a simulation where $u$ and $k$ carry strong $T$ effects.
:::

```{r derive-tls}
#| label: derive-tls
#| echo: true

# Data are in minutes (standardize_data was called with duration_unit = "minutes"),
# so set time_multiplier = 1 to keep the model's time axis unchanged on output,
# and t_ref = 60 to read CTmax at the 1-hour reference. The simulated data are
# binomial / beta-binomial *lethal* survival counts, so we opt into T_crit via
# the rate-multiplier integration (lethal = TRUE; default TC_rate_range =
# c(0.1, 1) % HI per hour). See the next paragraph for why T_crit is opt-in.
et_bin <- extract_tdt(wf_bin, t_ref = 60, time_multiplier = 1,
                      ndraws = 1000, lethal = TRUE)
et_bb  <- extract_tdt(wf_bb,  t_ref = 60, time_multiplier = 1,
                      ndraws = 1000, lethal = TRUE)
```

```{r tls-recovery-table}
#| label: tbl-tls-recovery
#| tbl-cap: "Posterior medians and 95% credible intervals for thermal sensitivity ($z$), the critical thermal limit at a 1-hour reference duration ($CT_{max_{1hr}}$), and the rate-multiplier critical temperature ($T_{crit}$, integrated over $r^* \\in [0.1, 1]$ % HI per hour). All three derived via `extract_tdt()` from a single posterior. Each row's truth value is computed analytically from `true_params`."

# Helper to pull (median, lower, upper) out of either a $z or temperature summary.
row_from_summary <- function(s, kind = c("z", "temp")) {
  kind <- match.arg(kind)
  if (kind == "z") {
    c(s$z_median, s$z_lower, s$z_upper)
  } else {
    c(s$temp_median, s$temp_lower, s$temp_upper)
  }
}

tls_recovery <- tibble::tibble(
  Quantity = c("z (°C)", "CTmax at 1 hr (°C)", "T_crit (°C)"),
  Truth    = c(true_z, true_CTmax, true_T_crit),
  `Bin. median` = c(row_from_summary(et_bin$z$summary,      "z")[1],
                    row_from_summary(et_bin$CTmax$summary,  "temp")[1],
                    row_from_summary(et_bin$T_crit$summary, "temp")[1]),
  `Bin. lower`  = c(row_from_summary(et_bin$z$summary,      "z")[2],
                    row_from_summary(et_bin$CTmax$summary,  "temp")[2],
                    row_from_summary(et_bin$T_crit$summary, "temp")[2]),
  `Bin. upper`  = c(row_from_summary(et_bin$z$summary,      "z")[3],
                    row_from_summary(et_bin$CTmax$summary,  "temp")[3],
                    row_from_summary(et_bin$T_crit$summary, "temp")[3]),
  `BB median`   = c(row_from_summary(et_bb$z$summary,      "z")[1],
                    row_from_summary(et_bb$CTmax$summary,  "temp")[1],
                    row_from_summary(et_bb$T_crit$summary, "temp")[1]),
  `BB lower`    = c(row_from_summary(et_bb$z$summary,      "z")[2],
                    row_from_summary(et_bb$CTmax$summary,  "temp")[2],
                    row_from_summary(et_bb$T_crit$summary, "temp")[2]),
  `BB upper`    = c(row_from_summary(et_bb$z$summary,      "z")[3],
                    row_from_summary(et_bb$CTmax$summary,  "temp")[3],
                    row_from_summary(et_bb$T_crit$summary, "temp")[3])
)

tinytable::tt(tls_recovery, digits = 3, escape = TRUE)
```

::: {#fig-tls-recovery}

```{r tls-recovery-plot}
#| label: plot-tls-recovery
#| fig-width: 9
#| fig-height: 3.5

# Per-draw posteriors for the three quantities, both likelihoods. Trim each
# to a common length so we can stack columns in one tibble.
n_keep <- min(nrow(et_bin$z$draws), nrow(et_bin$CTmax$draws),
              nrow(et_bin$T_crit$draws),
              nrow(et_bb$z$draws), nrow(et_bb$CTmax$draws),
              nrow(et_bb$T_crit$draws))

draws_bin_d <- tibble::tibble(
  z         = et_bin$z$draws$z[seq_len(n_keep)],
  CTmax_1hr = et_bin$CTmax$draws$temp[seq_len(n_keep)],
  T_crit    = et_bin$T_crit$draws$temp[seq_len(n_keep)],
  model     = "Binomial"
)
draws_bb_d <- tibble::tibble(
  z         = et_bb$z$draws$z[seq_len(n_keep)],
  CTmax_1hr = et_bb$CTmax$draws$temp[seq_len(n_keep)],
  T_crit    = et_bb$T_crit$draws$temp[seq_len(n_keep)],
  model     = "Beta-binomial"
)

post_long <- dplyr::bind_rows(draws_bin_d, draws_bb_d) |>
  tidyr::pivot_longer(c(z, CTmax_1hr, T_crit),
                      names_to = "quantity", values_to = "value") |>
  dplyr::mutate(quantity = factor(quantity,
                                  levels = c("z", "CTmax_1hr", "T_crit"),
                                  labels = c("z (°C)",
                                             "CTmax_1hr (°C)",
                                             "T_crit (°C)")))

truth_long <- tibble::tibble(
  quantity = factor(c("z", "CTmax_1hr", "T_crit"),
                    levels = c("z", "CTmax_1hr", "T_crit"),
                    labels = levels(post_long$quantity)),
  truth    = c(true_z, true_CTmax, true_T_crit)
)

ggplot2::ggplot(post_long, ggplot2::aes(value, fill = model)) +
  ggplot2::geom_density(alpha = 0.5, colour = NA) +
  ggplot2::geom_vline(data = truth_long,
                      ggplot2::aes(xintercept = truth),
                      linetype = "dashed", colour = "black") +
  ggplot2::facet_wrap(~ quantity, scales = "free") +
  ggplot2::labs(x = NULL, y = "Posterior density", fill = "Model") +
  ggplot2::theme_classic()
```

Posterior densities of $z$, $CT_{max_{1hr}}$, and $T_{crit}$ from the binomial (blue) and beta-binomial (orange) joint fits. Vertical dashed lines mark the simulation truth. All three posteriors concentrate around their respective truths — illustrating that one fit, one library call (`extract_tdt()`), produces every classical TLS quantity with calibrated uncertainty and no extra regression step.
:::

All three quantities recover their simulation truth in both fits (@tbl-tls-recovery, @fig-tls-recovery), with credible intervals reflecting joint uncertainty in all 4PL parameters. The key practical advantage of the joint approach over the two-step pipeline: every classical TLS quantity is one library call away from the fit we already have, and every quantity inherits the same calibrated uncertainty.

Posterior summary statistics — mean, SD, median, and 95% credible interval — are also readily available from the `$draws` slot of `extract_tdt()`'s output:

```{r uncert-z-ct}
#| label: tbl-uncert-z-ct
#| tbl-cap: "Posterior summaries of $z$, $CT_{max_{1hr}}$, and $T_{crit}$ from the joint Bayesian 4PL: posterior mean, standard deviation, median, and 95% credible interval, reported separately for the binomial and beta-binomial fits."
#| echo: true

summarise_draws <- function(x) {
  q <- stats::quantile(x, c(0.025, 0.5, 0.975), names = FALSE, na.rm = TRUE)
  tibble::tibble(mean = mean(x, na.rm = TRUE),
                 sd   = stats::sd(x, na.rm = TRUE),
                 median = q[2], lower = q[1], upper = q[3])
}

uncert_summary <- dplyr::bind_rows(
  summarise_draws(et_bin$z$draws$z)          |> dplyr::mutate(quantity = "z (°C)",                  likelihood = "Binomial"),
  summarise_draws(et_bin$CTmax$draws$temp)   |> dplyr::mutate(quantity = "CTmax at 1 hr (°C)",     likelihood = "Binomial"),
  summarise_draws(et_bin$T_crit$draws$temp)  |> dplyr::mutate(quantity = "T_crit (°C)", likelihood = "Binomial"),
  summarise_draws(et_bb$z$draws$z)           |> dplyr::mutate(quantity = "z (°C)",                  likelihood = "Beta-binomial"),
  summarise_draws(et_bb$CTmax$draws$temp)    |> dplyr::mutate(quantity = "CTmax at 1 hr (°C)",     likelihood = "Beta-binomial"),
  summarise_draws(et_bb$T_crit$draws$temp)   |> dplyr::mutate(quantity = "T_crit (°C)", likelihood = "Beta-binomial")
) |>
  dplyr::select(quantity, likelihood, mean, sd, median, lower, upper)

# Scalars for inline reporting in the next paragraph.
sd_z_bin  <- round(stats::sd(et_bin$z$draws$z),         2)
sd_z_bb   <- round(stats::sd(et_bb$z$draws$z),          2)
sd_ct_bin <- round(stats::sd(et_bin$CTmax$draws$temp),  2)
sd_ct_bb  <- round(stats::sd(et_bb$CTmax$draws$temp),   2)
ci_z_bin  <- round(diff(stats::quantile(et_bin$z$draws$z,         c(0.025, 0.975), names = FALSE, na.rm = TRUE)), 2)
ci_z_bb   <- round(diff(stats::quantile(et_bb$z$draws$z,          c(0.025, 0.975), names = FALSE, na.rm = TRUE)), 2)
ci_ct_bin <- round(diff(stats::quantile(et_bin$CTmax$draws$temp,  c(0.025, 0.975), names = FALSE, na.rm = TRUE)), 2)
ci_ct_bb  <- round(diff(stats::quantile(et_bb$CTmax$draws$temp,   c(0.025, 0.975), names = FALSE, na.rm = TRUE)), 2)

uncert_summary |>
  dplyr::rename(
    Quantity   = quantity,
    Likelihood = likelihood,
    Mean       = mean,
    SD         = sd,
    Median     = median,
    Lower      = lower,
    Upper      = upper
  ) |>
  tinytable::tt(digits = 3, escape = TRUE)
```

In short, both $z$ and $CT_{max_{1hr}}$ are estimated with high precision and their credible intervals come straight from the joint posterior. If we fit this model using frequentist approaches we could do the same and use the delta-method, or bootstrapping to get uncertainty. 

Under the binomial likelihood, the standard error (or SD of posterior) for $z$ is `r sd_z_bin` °C with a 95% credible interval `r ci_z_bin` °C wide; $CT_{max_{1hr}}$ has SD `r sd_ct_bin` °C and a 95% credible interval width of `r ci_ct_bin` °C. 

The beta-binomial fit is slightly wider on both quantities (SD on $z$: `r sd_z_bb` °C; on $CT_{max_{1hr}}$: `r sd_ct_bb` °C) because $\phi$ has absorbed the extra-binomial variance and that uncertainty propagates honestly into the derived quantities — none of which falls out of the two-stage pipeline without an additional bootstrap or delta-method step. 

::: {.callout-important}
It is always important to evaluate whether a binomial and a beta-binomial model better fits real data. The extra dispersion, typical of real biological data, should be estimated when it's present as it has important consequences for inferences.
:::

### *Comparing the two pipelines* {#sec-compare}

#### Point-estimate agreement on the same data {#sec-compare-table}

@tbl-twostage-compare puts the two-stage point estimates next to the joint Bayesian posterior medians for both simulated datasets.

```{r compare-twostage-table}
#| label: tbl-twostage-compare
#| tbl-cap: "Point estimates (two-stage) and posterior medians (joint Bayesian) for thermal sensitivity ($z$) and $CT_{max_{1hr}}$, recovered by each of four estimator–dataset combinations from the same simulated datasets. Truth values are the simulation parameters."

compare_table <- tibble::tibble(
  quantity   = c("z (°C)", "CTmax at 1 hr (°C)"),
  truth      = c(true_z, true_CTmax),
  ts_bin     = ts_bin$est,
  joint_bin  = c(median(draws_bin_d$z),
                 median(draws_bin_d$CTmax_1hr)),
  ts_bb      = ts_bb$est,
  joint_bb   = c(median(draws_bb_d$z),
                 median(draws_bb_d$CTmax_1hr))
)

compare_table |>
  dplyr::rename(
    Quantity                          = quantity,
    Truth                             = truth,
    `Two-stage (binomial)`            = ts_bin,
    `Joint Bayesian (binomial)`       = joint_bin,
    `Two-stage (beta-binomial)`       = ts_bb,
    `Joint Bayesian (beta-binomial)`  = joint_bb
  ) |>
  tinytable::tt(digits = 3, escape = TRUE)
```

All four estimators land on essentially the same point estimates for $z$ and $CT_{max_{1hr}}$, though we notice the two stage is biased upwards slightly, and all sit close to the simulation truth. The joint Bayesian 4PL gives the same answer the classical pipeline gives — it is *not* in disagreement with the field's existing point estimates, it reproduces them.

### *A worked example: temperature effects on every shape parameter* {#sec-joint-extend}

The default `fit_4pl()` already puts a `temp_c` slope on $\ell$, $u$ and $k$, so any temperature effect on the shape parameters is fit alongside the linear midpoint. To validate that `extract_tdt()` returns sensible $z$, $CT_{max_{1hr}}$ and $\log_{10}\text{LT50}(T)$ posteriors when the truth violates the constant-shape assumption, we re-simulate beta-binomial data on the same factorial design as @sec-sim-step2, but now $u$ declines steeply with $T$ and $k$ rises steeply with $T$. The fit is then exactly the same `fit_4pl()` call as before — no special formula, no special priors — and the analysis afterwards is exactly the same `extract_tdt()` call.

```{r sim-data-ext}
#| label: sim-data-ext
#| echo: true

# Same factorial design as @sec-sim-step2; re-declared here so this chunk is
# self-contained.
temps     <- c(30, 32, 34, 36, 38)
durations <- c(1, 5, 15, 45, 135, 405)
n_rep     <- 30
n_per_rep <- 30
phi       <- 5

# Truth on the same raw scale fit_4pl() uses internally (so we can compare the
# posterior coefficients to truth directly):
#   ell = 0.49 * inv_logit(ell_raw)                    (held constant in T)
#   u   = 0.51 + 0.49 * inv_logit(u_raw_0 + u_raw_T*T_c)  (declines in T)
#   k   = exp(log_k_0 + log_k_T*T_c)                      (rises in T)
true_params_ext <- list(
  ell_raw  = qlogis(0.05 / 0.49),                 # ell = 0.05
  u_raw_0  = qlogis((0.92 - 0.51) / 0.49),        # u = 0.92 at T_bar
  u_raw_T  = -0.60,                                # strong decline in u with T
  log_k_0  = log(8),                               # k = 8 at T_bar
  log_k_T  = 0.30,                                 # strong rise in k with T
  m_beta0  = 1.5,
  m_beta1  = -0.15,
  T_bar    = 34
)

design <- tidyr::expand_grid(
  T   = temps,
  t   = durations,
  rep = seq_len(n_rep)
) |>
  dplyr::mutate(
    log10_t = log10(t),
    T_c     = T - true_params_ext$T_bar,
    n       = n_per_rep
  )

sim_bb_ext <- design |>
  dplyr::mutate(
    ell    = 0.49 * plogis(true_params_ext$ell_raw),
    u      = 0.51 + 0.49 * plogis(true_params_ext$u_raw_0 +
                                   true_params_ext$u_raw_T * T_c),
    k      = exp(true_params_ext$log_k_0 + true_params_ext$log_k_T * T_c),
    mid    = true_params_ext$m_beta0 + true_params_ext$m_beta1 * T_c,
    p_true = ell + (u - ell) / (1 + exp(k * (log10_t - mid))),
    p_draw = rbeta(dplyr::n(), p_true * phi, (1 - p_true) * phi),
    y      = rbinom(dplyr::n(), size = n, prob = p_draw)
  )
```

Under this truth the lower asymptote stays flat at $\ell = 0.05$, the upper asymptote $u$ falls from `r round(0.51 + 0.49 * plogis(true_params_ext$u_raw_0 + true_params_ext$u_raw_T * (30 - true_params_ext$T_bar)), 2)` at $T = 30$ to `r round(0.51 + 0.49 * plogis(true_params_ext$u_raw_0 + true_params_ext$u_raw_T * (38 - true_params_ext$T_bar)), 2)` at $T = 38$, and the steepness $k$ rises from `r round(exp(true_params_ext$log_k_0 + true_params_ext$log_k_T * (30 - true_params_ext$T_bar)), 1)` to `r round(exp(true_params_ext$log_k_0 + true_params_ext$log_k_T * (38 - true_params_ext$T_bar)), 1)` across the same range. The resulting dose-response curves are visibly different in shape at the two extremes (@fig-true-surface-ext).

::: {#fig-true-surface-ext}

```{r true-surface-ext-plot}
#| label: plot-true-surface-ext
#| fig-width: 8
#| fig-height: 5

truth_grid_ext <- tidyr::expand_grid(
  T = temps,
  t = 10 ^ seq(log10(0.5), log10(800), length.out = 200)
) |>
  dplyr::mutate(
    log10_t = log10(t),
    T_c     = T - true_params_ext$T_bar,
    ell     = 0.49 * plogis(true_params_ext$ell_raw),
    u       = 0.51 + 0.49 * plogis(true_params_ext$u_raw_0 +
                                   true_params_ext$u_raw_T * T_c),
    k       = exp(true_params_ext$log_k_0 + true_params_ext$log_k_T * T_c),
    mid     = true_params_ext$m_beta0 + true_params_ext$m_beta1 * T_c,
    p_true  = ell + (u - ell) / (1 + exp(k * (log10_t - mid)))
  )

ggplot2::ggplot() +
  ggplot2::geom_hline(yintercept = 0.5, linetype = "dashed",
                      colour = "grey50") +
  ggplot2::geom_jitter(
    data = sim_bb_ext,
    ggplot2::aes(x = log10_t, y = y / n, colour = factor(T)),
    width = 0.02, height = 0, alpha = 0.25, size = 0.7
  ) +
  ggplot2::geom_line(
    data = truth_grid_ext,
    ggplot2::aes(x = log10_t, y = p_true, colour = factor(T)),
    linewidth = 0.8
  ) +
  ggplot2::labs(x = expression(log[10]~exposure~time~(min)),
                y = "Survival probability",
                colour = "Temperature (°C)") +
  ggplot2::theme_classic()
```

True 4PL dose-response surface (solid lines) for the extended simulation, where $k$ rises and $u$ falls with assay temperature, overlaid on the simulated beta-binomial counts (points). Note how the curves at $T = 30$ and $T = 38$ have very different upper asymptotes and steepness — not just a horizontal shift along the time axis.
:::

Before fitting we can compute, analytically, what the truth implies for $\log_{10}\text{LT50}(T)$, the local thermal sensitivity $z(T)$ at each assay temperature (defined as $-1$ divided by the local slope of $\log_{10}\text{LT50}$ on $T$), and $CT_{max_{1hr}}$ (the $T$ at which $\log_{10}\text{LT50}$ crosses $\log_{10} 60$). These are the targets we will ask `extract_tdt()` to recover.

```{r truth-ext-derived}
#| label: truth-ext-derived
#| echo: true

# True log10(LT50) at any T from the T-varying 4PL: solve p(log10_t) = 0.5
# for log10_t. With the 4PL,
#   log10_t = mid + (1/k) * log((u - 0.5) / (0.5 - ell))
true_log10_LT50 <- function(T_value) {
  T_c <- T_value - true_params_ext$T_bar
  ell <- 0.49 * plogis(true_params_ext$ell_raw)
  u   <- 0.51 + 0.49 * plogis(true_params_ext$u_raw_0 +
                              true_params_ext$u_raw_T * T_c)
  k   <- exp(true_params_ext$log_k_0 + true_params_ext$log_k_T * T_c)
  mid <- true_params_ext$m_beta0 + true_params_ext$m_beta1 * T_c
  mid + (1 / k) * log((u - 0.5) / (0.5 - ell))
}

# Local z(T) via central finite difference of true log10(LT50)(T).
true_local_z <- function(T_at, half_step = 0.5) {
  -1 / ((true_log10_LT50(T_at + half_step) -
         true_log10_LT50(T_at - half_step)) / (2 * half_step))
}

# True CTmax_1hr: search the fine grid for the T at which log10(LT50)
# crosses log10(60).
T_grid_truth   <- seq(20, 45, by = 0.001)
ll_grid_truth  <- vapply(T_grid_truth, true_log10_LT50, numeric(1))
true_CTmax_1hr <- T_grid_truth[which.min(abs(ll_grid_truth - log10(60)))]

truth_ext <- tibble::tibble(
  T               = temps,
  log10_LT50_true = vapply(temps, true_log10_LT50, numeric(1)),
  LT50_min_true   = 10 ^ log10_LT50_true,
  local_z_true    = vapply(temps, true_local_z, numeric(1))
)

true_mean_local_z <- mean(truth_ext$local_z_true)
```

Across the assay range the true $\log_{10}\text{LT50}(T)$ curve is bent: the implied local $z$ varies from `r round(min(truth_ext$local_z_true), 2)` °C at $T = 30$ to `r round(max(truth_ext$local_z_true), 2)` °C at $T = 38$ (truth mean $z$ = `r round(true_mean_local_z, 2)` °C). The truth implies $CT_{max_{1hr}}$ = `r round(true_CTmax_1hr, 2)` °C. With straight-line $\log_{10}\text{LT50}(T)$ — the constant-shape case — local $z$ would not vary across temperature at all.

We now fit with exactly the same `fit_4pl()` call as before. The default formula already puts a `temp_c` slope on each of the four 4PL sub-parameters, so the model has the capacity to recover the truth's temperature signal on $u$ and $k$ without any modification.

```{r fit-ext-default}
#| label: fit-ext-default
#| echo: true
#| cache: false

std_ext <- standardize_data(sim_bb_ext,
                            temp = "T", duration = "t",
                            n_total = "n", n_surv = "y",
                            duration_unit = "minutes")

wf_ext <- fit_4pl(std_ext,
                  chains = 4, iter = 2000, cores = 4, seed = 123,
                  file = file.path(models_dir,
                                   "sim_4pl_ext_betabin_DEFAULT_strong_v1"))
```

The four 4PL sub-parameters on the natural (biological) scale at each assay temperature are recovered in @tbl-ext-recovery. The lower asymptote $\ell$ stays near $0.05$ at every $T$ (consistent with the truth that $\ell$ does not vary with temperature), while $u$ falls and $k$ rises across the assay range — exactly the truth.

```{r ext-recovery-table}
#| label: tbl-ext-recovery
#| tbl-cap: "Posterior recovery of the four 4PL sub-parameters on the natural scale at each assay temperature. The fit uses the disjoint-bounds reparameterisation ($\\ell$, $u$ via inv_logit, $k$ via exp); raw posterior columns are transformed back to the biological scale before summarising. Posterior medians and 95% credible intervals are reported alongside the analytical truth."

ext_post  <- posterior::as_draws_df(wf_ext$fit) |> as.data.frame()
ext_meta  <- wf_ext$meta
ext_bnds  <- ext_meta$bounds         # compute_4pl_bounds(0, 1)
ext_temp_mean <- ext_meta$temp_mean  # what the fit centred on

ext_param_long <- purrr::map_dfr(temps, function(T_value) {
  T_c_fit  <- T_value - ext_temp_mean
  T_c_true <- T_value - true_params_ext$T_bar

  draws_at_T <- tibble::tibble(
    ell = ext_bnds$low_min +
          stats::plogis(ext_post$b_lowraw_Intercept +
                        ext_post$b_lowraw_temp_c * T_c_fit) * ext_bnds$low_w,
    u   = ext_bnds$up_min  +
          stats::plogis(ext_post$b_upraw_Intercept +
                        ext_post$b_upraw_temp_c * T_c_fit)  * ext_bnds$up_w,
    k   = exp(ext_post$b_logk_Intercept + ext_post$b_logk_temp_c * T_c_fit),
    mid = ext_post$b_mid_Intercept + ext_post$b_mid_temp_c * T_c_fit
  )

  truth_at_T <- tibble::tibble(
    ell = 0.49 * plogis(true_params_ext$ell_raw),
    u   = 0.51 + 0.49 * plogis(true_params_ext$u_raw_0 +
                               true_params_ext$u_raw_T * T_c_true),
    k   = exp(true_params_ext$log_k_0 +
              true_params_ext$log_k_T * T_c_true),
    mid = true_params_ext$m_beta0 + true_params_ext$m_beta1 * T_c_true
  )

  purrr::map_dfr(c("ell", "u", "k", "mid"), function(p) {
    q <- stats::quantile(draws_at_T[[p]], c(0.025, 0.5, 0.975),
                         na.rm = TRUE, names = FALSE)
    tibble::tibble(
      Parameter         = p,
      `T (°C)`         = T_value,
      Truth             = truth_at_T[[p]],
      Median            = q[2],
      Lower             = q[1],
      Upper             = q[3]
    )
  })
})

ext_param_long |>
  tinytable::tt(digits = 3, escape = TRUE)
```

Now we apply the same `extract_tdt()` call as in @sec-joint-derive — there is no separate "T-varying" workflow because `extract_tdt()` already inverts the fitted surface numerically per draw:

```{r extract-tdt-ext}
#| label: extract-tdt-ext
#| echo: true

et_ext <- extract_tdt(wf_ext, t_ref = 60, time_multiplier = 1,
                      ndraws = 1000, lethal = TRUE)
```

@tbl-extract-ext compares the three classical TDT quantities from `extract_tdt()` to the analytical truth derived above.

```{r extract-tdt-ext-table}
#| label: tbl-extract-ext
#| tbl-cap: "Classical TDT quantities from `extract_tdt()` on the T-varying fit, compared against analytical truth. The single-summary $z$ is the slope of a per-draw OLS regression of $\\log_{10}\\text{LT50}$ on $T$ across the assay grid; the truth row reports the mean of the analytical local $z(T)$ at the same assay temperatures, which is what the OLS slope estimates when the curve is bent. $CT_{max_{1hr}}$ comes from direct numerical inversion of the fitted 4PL at 60 min per draw."

tibble::tibble(
  Quantity     = c("z (°C)",
                   "CTmax at 1 hr (°C)",
                   "T_crit (°C)"),
  Truth        = c(round(true_mean_local_z, 3),
                   round(true_CTmax_1hr, 3),
                   NA_real_),  # T_crit truth involves r* integration; see below
  Median       = c(et_ext$z$summary$z_median,
                   et_ext$CTmax$summary$temp_median,
                   et_ext$T_crit$summary$temp_median),
  Lower        = c(et_ext$z$summary$z_lower,
                   et_ext$CTmax$summary$temp_lower,
                   et_ext$T_crit$summary$temp_lower),
  Upper        = c(et_ext$z$summary$z_upper,
                   et_ext$CTmax$summary$temp_upper,
                   et_ext$T_crit$summary$temp_upper)
) |>
  tinytable::tt(digits = 3, escape = TRUE)
```

Both posteriors comfortably cover their analytical truth. The single-summary $z$ is the OLS slope across the assay grid — which, on a bent curve, estimates the *average* local slope (truth = `r round(true_mean_local_z, 2)` °C, posterior median `r round(et_ext$z$summary$z_median, 2)` °C, 95% CrI [`r round(et_ext$z$summary$z_lower, 2)`, `r round(et_ext$z$summary$z_upper, 2)`] °C). $CT_{max_{1hr}}$ is recovered by direct 4PL inversion at the reference duration and lands on the truth (`r round(true_CTmax_1hr, 2)` °C → posterior median `r round(et_ext$CTmax$summary$temp_median, 2)` °C, 95% CrI [`r round(et_ext$CTmax$summary$temp_lower, 2)`, `r round(et_ext$CTmax$summary$temp_upper, 2)`] °C); the numerical inversion does not depend on the asymmetry correction being constant in $T$.

The bent $\log_{10}\text{LT50}(T)$ curve itself is the intermediate object both quantities are built from; it lives in `et_ext$lt50_curve` (the output of `derive_ltx_curve()`) and can be plotted directly (@fig-lt50-curve-ext).

::: {#fig-lt50-curve-ext}

```{r lt50-curve-ext-plot}
#| label: plot-lt50-curve-ext
#| fig-width: 6
#| fig-height: 4

truth_overlay <- {
  Ts <- seq(min(temps) - 2, max(temps) + 2, by = 0.05)
  tibble::tibble(temp = Ts,
                 duration_median = 10 ^ vapply(Ts, true_log10_LT50,
                                               numeric(1)))
}

plot_ltx_curve(et_ext$lt50_curve) +
  ggplot2::geom_line(data = truth_overlay,
                     ggplot2::aes(x = temp, y = duration_median),
                     linetype = "dashed", colour = "black",
                     inherit.aes = FALSE)
```

Posterior $\text{LT50}(T)$ curve (median + 95% CrI, blue) recovered by `derive_ltx_curve()` from the T-varying fit, with the analytical truth overlaid (dashed black). The curve is visibly bent rather than straight in $T$, and the posterior tracks the truth.
:::

We can also drop down to the per-draw level and compute a *local* $z(T)$ at each assay temperature from the same `et_ext$lt50_curve` posterior — a finite difference of $\log_{10}\text{LT50}$ around $T$ per draw, summarised across draws (@fig-local-z-ext).

```{r local-z-ext}
#| label: local-z-ext
#| echo: true

# Finite-difference local z at each assay T, per posterior draw, derived
# from et_ext$lt50_curve (the same per-draw LT50(T) curve extract_tdt
# already computed for the OLS).
half_step <- 0.5

local_z_summary <- purrr::map_dfr(temps, function(T_at) {
  z_per_draw <- et_ext$lt50_curve$draws |>
    dplyr::group_by(.draw) |>
    dplyr::summarise(
      z_local = -1 / ((stats::approx(temp, log10(duration_out),
                                     xout = T_at + half_step)$y -
                       stats::approx(temp, log10(duration_out),
                                     xout = T_at - half_step)$y) /
                      (2 * half_step)),
      .groups = "drop"
    )
  q <- stats::quantile(z_per_draw$z_local, c(0.025, 0.5, 0.975),
                       na.rm = TRUE, names = FALSE)
  tibble::tibble(T_at           = T_at,
                 z_local_median = q[2],
                 z_local_lower  = q[1],
                 z_local_upper  = q[3])
})

local_z_summary |>
  dplyr::left_join(
    tibble::tibble(T_at = temps, z_local_truth = truth_ext$local_z_true),
    by = "T_at"
  ) |>
  dplyr::select(T_at, z_local_truth, z_local_median,
                z_local_lower, z_local_upper) |>
  dplyr::rename(
    `T (°C)`              = T_at,
    `z local truth (°C)`  = z_local_truth,
    Median                = z_local_median,
    Lower                 = z_local_lower,
    Upper                 = z_local_upper
  ) |>
  tinytable::tt(digits = 3, escape = TRUE)
```

::: {#fig-local-z-ext}

```{r local-z-ext-plot}
#| label: plot-local-z-ext
#| fig-width: 6
#| fig-height: 4

truth_curve_z <- {
  Ts <- seq(min(temps), max(temps), by = 0.1)
  tibble::tibble(T_at = Ts,
                 z_local_truth = vapply(Ts, true_local_z, numeric(1)))
}

ggplot2::ggplot(local_z_summary,
                ggplot2::aes(x = T_at, y = z_local_median)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = z_local_lower,
                                    ymax = z_local_upper),
                       fill = "#146C7C", alpha = 0.2) +
  ggplot2::geom_line(colour = "#146C7C", linewidth = 1) +
  ggplot2::geom_line(data = truth_curve_z,
                     ggplot2::aes(x = T_at, y = z_local_truth),
                     linetype = "dashed", colour = "black",
                     inherit.aes = FALSE) +
  ggplot2::labs(x = "Assay temperature (°C)",
                y = expression(local~italic(z)~"(°C)")) +
  ggplot2::theme_classic()
```

Local thermal sensitivity $z(T)$ recovered from the per-draw $\log_{10}\text{LT50}(T)$ curve (blue median + 95% CrI), with the analytical truth overlaid (dashed black). Where the constant-shape simple-case fit would give the same $z$ at every assay temperature, here local $z$ varies with $T$ in lock-step with the bent $\log_{10}\text{LT50}(T)$ curve.
:::

The *same* `fit_4pl()` and `extract_tdt()` calls recover the bent $\log_{10}\text{LT50}(T)$ curve, the average $z$ across the assay grid, and the per-draw $CT_{max_{1hr}}$ without modification. When the truth is constant-shape (as in @sec-joint-derive) the same fit returns near-zero slopes on $u$, $\ell$ and $k$, the bent-curve calculation collapses to the closed-form one.

### Heat injury accumulation and survival under a fluctuating trace

A particularly powerful aspect of TLS models is their ability to translate dynamic temperature time-series from nature and predict how such temperatures lead to heat injury (HI) accumulation towards a threshold (e.g., $LT_{50}$) [@jorgensen_unifying_2021; @orsted_finding_2022; @arnold_thermal_2025; @noble2026; @rezende_drosophila_mortality_2020]. The joint Bayesian 4PL fit gives us not just point estimates of $z$ and $CT_{max_{1hr}}$ but their full joint posterior, which we can push through the HI integral to obtain a posterior distribution over the predicted HI trajectory. We can also use the *full* 4PL surface to predict the cumulative survival fraction $S_{\text{pred}}(t)$ across the trajectory (Equation 9 of the manuscript). The library function `predict_heat_injury()` does both calculations in a single call and returns both trajectories with full posterior uncertainty.

#### A simulated temperature profile across 5 days

To mimic conditions a moderately heat-tolerant ectotherm might experience, we generate 5 days of hourly temperatures with a sinusoidal diurnal cycle, ranging from 12 °C overnight to a midday peak of 24 °C. The critical temperature $T_c$ — below which damage is repaired as fast as it accrues and so there is no net damage increase [@orsted_finding_2022; @jorgensen_unifying_2021] — is set to 20 °C. The simulated organism has $CT_{max_{1hr}} \approx `r round(true_CTmax, 1)`$°C and $z \approx `r round(true_z, 1)`$°C, so the daily peak at 24 °C remains 8 °C below the 1-hour critical limit — moderate, sub-lethal stress that builds cumulatively.

```{r hi-trace}
#| label: hi-trace
#| echo: true

hi_trace <- tibble::tibble(
  time_h = 0:(24 * 5 - 1),
  temp   = 18 + 6 * sin(2 * pi * (0:(24 * 5 - 1) - 6) / 24)   # range 12-24 °C
)
hi_T_c <- 20    # damage threshold (°C)
```

::: {#fig-hi-trace}

```{r hi-trace-plot}
#| label: plot-hi-trace
#| fig-width: 7
#| fig-height: 2.5

ggplot2::ggplot(hi_trace, ggplot2::aes(time_h, temp)) +
  ggplot2::geom_line(colour = "darkred", linewidth = 0.6) +
  ggplot2::geom_hline(yintercept = hi_T_c, linetype = "dashed", colour = "grey40") +
  ggplot2::geom_hline(yintercept = true_CTmax, linetype = "dotted", colour = "grey40") +
  ggplot2::annotate("text", x = max(hi_trace$time_h), y = hi_T_c, label = "T_c",
                    hjust = 1, vjust = -0.4, size = 3, colour = "grey40") +
  ggplot2::annotate("text", x = max(hi_trace$time_h), y = true_CTmax,
                    label = "CTmax_1hr",
                    hjust = 1, vjust = -0.4, size = 3, colour = "grey40") +
  ggplot2::ylim(10, 45) +
  ggplot2::labs(x = "Hour", y = "Temperature (°C)") +
  ggplot2::theme_classic()
```

5-day diurnal temperature trace. The dashed line is the damage-accumulation threshold $T_c = 20$ °C; the dotted line is the simulated organism's $CT_{max_{1hr}} \approx `r round(true_CTmax, 1)`$°C — well above the daily peaks, so accumulation is sub-lethal and gradual.
:::

#### Posterior HI and survival via `predict_heat_injury()`

`predict_heat_injury()` takes a temperature trace plus a fitted workflow, propagates the full posterior of $(\ell, u, k, \beta_0, \beta_1)$ through the HI integral, and returns the median and 95% credible band of cumulative HI and predicted survival at every time point. We supply `T_c` to enforce zero damage accumulation below the threshold (matching Equation 7 of the manuscript). The model time units here are minutes (because `standardize_data()` was called with `duration_unit = "minutes"`), so the trace's `time_h` is in hours and damage rates are computed accordingly internally.

```{r hi-predict}
#| label: hi-predict
#| echo: true

# Convert hourly trace to minutes to match the model's time units.
hi_trace_min <- dplyr::mutate(hi_trace, time_h = time_h * 60)

hi <- predict_heat_injury(
  trace        = hi_trace_min,
  workflow     = wf_bb,
  target_surv  = 0.5,
  T_c          = hi_T_c,
  ndraws       = 500
)
```

::: {#fig-hi-trajectory}

```{r hi-trajectory-plot}
#| label: plot-hi-trajectory
#| fig-width: 7
#| fig-height: 5.5

# Convert back to hours for plotting.
hi_plot <- dplyr::mutate(hi$summary, time_h = time_h / 60)

p_hi <- ggplot2::ggplot(hi_plot, ggplot2::aes(time_h, hi_median)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = hi_lower, ymax = hi_upper),
                       alpha = 0.2, fill = "darkred") +
  ggplot2::geom_line(colour = "darkred", linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = 100, linetype = "dashed", colour = "grey40") +
  ggplot2::labs(x = NULL, y = "Cumulative HI (%)") +
  ggplot2::theme_classic()

p_surv <- ggplot2::ggplot(hi_plot, ggplot2::aes(time_h, surv_median)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = surv_lower, ymax = surv_upper),
                       alpha = 0.2, fill = "steelblue") +
  ggplot2::geom_line(colour = "steelblue", linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  ggplot2::scale_y_continuous(limits = c(0, 1)) +
  ggplot2::labs(x = "Hour", y = "Predicted survival") +
  ggplot2::theme_classic()

patchwork::wrap_plots(p_hi, p_surv, ncol = 1)
```

Posterior heat-injury accumulation (top, red) and predicted survival fraction (bottom, blue) under the 5-day field trace. Solid lines are posterior medians across 500 draws from `wf_bb`; shaded bands are 95% credible intervals. The dashed lines mark the LT50 threshold (HI = 100%) and 50% survival respectively. The two trajectories tell the same story: $S_{\text{pred}}(t)$ falls through 0.5 at the same time HI($t$) crosses 100%.
:::

By the end of day 5 the posterior median HI is `r round(tail(hi$summary$hi_median, 1), 0)`% (95% CI: `r round(tail(hi$summary$hi_lower, 1), 0)`% to `r round(tail(hi$summary$hi_upper, 1), 0)`%), and median predicted survival is `r round(tail(hi$summary$surv_median, 1), 2)` (95% CI: `r round(tail(hi$summary$surv_lower, 1), 2)` to `r round(tail(hi$summary$surv_upper, 1), 2)`).

#### Why are the credible bands so wide?

Both posterior bands look generous relative to the precision suggested by @tbl-tls-recovery on the underlying $z$ and $CT_{max_{1hr}}$. That is not a bug — it is the exponential rate-multiplier $10^{(T - CT_{max,1hr})/z}$ doing its job. A 1 °C shift in $CT_{max_{1hr}}$ multiplies the per-hour HI rate by $10^{1/z}$, which for $z \approx 2.6$ °C is a ~2.4× change; a 0.5 °C shift in $z$ at the same peak temperature shifts the rate by a comparable factor. Over a 5-day integral those multiplicative differences compound, so even tight parameter posteriors translate into visibly wide HI bands. The classical two-stage pipeline conceals this by treating the point estimates of $z$ and $CT_{max_{1hr}}$ as known when it computes HI — the joint Bayesian framework instead propagates that uncertainty honestly. The width of the band is therefore a faithful report of how much you should *not* over-interpret a point prediction, not a sign that the fit is too uncertain to be useful.

### Validation: recovering known planted doses {#sec-hi-validation}

Before trusting `predict_heat_injury()` on real field data, we validate it against analytical truth on three reference traces with known dosing structure. `make_temperature_scenarios()` builds the traces; `planted_dose_from_trace()` implements the classical HI integral (Equation 7) directly given known `z` and `CTmax_1hr`. Posterior medians from `predict_heat_injury()` should sit near the analytical truth, and the 95% credible intervals should cover it.

```{r hi-validation}
#| label: hi-validation
#| echo: true

# Three traces: flat baseline (zero HI), single hourly spike, three spikes.
val_scens <- make_temperature_scenarios(
  baseline    = 20,
  spike_temp  = 28,
  n_hours     = 96,
  spike_times_single = 24,
  spike_times_multi  = c(24, 48, 72)
)
val_T_c <- 24

# Analytical truth, given the simulation parameters (true_z, true_CTmax).
planted <- purrr::map(val_scens, planted_dose_from_trace,
                      z = true_z, CTmax_1hr = true_CTmax, T_c = val_T_c)

# Posterior predictions from the fitted beta-binomial workflow. The model
# expects minutes; convert each trace's time axis.
predicted <- purrr::map(val_scens, function(sc) {
  predict_heat_injury(
    trace       = dplyr::mutate(sc, time_h = time_h * 60),
    workflow    = wf_bb,
    target_surv = 0.5,
    T_c         = val_T_c,
    ndraws      = 300
  )
})

# Final HI per scenario: analytical truth vs posterior median + CrI.
val_table <- tibble::tibble(
  Scenario     = c("Flat", "Single spike", "Multi spike"),
  `Planted HI (%)` = c(tail(planted$flat$hi_cumulative,         1),
                        tail(planted$single_spike$hi_cumulative, 1),
                        tail(planted$multi_spike$hi_cumulative,  1)),
  `Posterior median (%)` = c(tail(predicted$flat$summary$hi_median,        1),
                              tail(predicted$single_spike$summary$hi_median, 1),
                              tail(predicted$multi_spike$summary$hi_median,  1)),
  `Lower (%)` = c(tail(predicted$flat$summary$hi_lower,        1),
                  tail(predicted$single_spike$summary$hi_lower, 1),
                  tail(predicted$multi_spike$summary$hi_lower,  1)),
  `Upper (%)` = c(tail(predicted$flat$summary$hi_upper,        1),
                  tail(predicted$single_spike$summary$hi_upper, 1),
                  tail(predicted$multi_spike$summary$hi_upper,  1))
)
val_table$`Truth in CrI?` <- with(val_table,
                                  `Lower (%)` <= `Planted HI (%)` &
                                  `Planted HI (%)` <= `Upper (%)`)
tinytable::tt(val_table, digits = 2, escape = TRUE)
```

The flat trace stays below $T_c$ for its entire duration and so accrues zero HI, matching the analytical truth exactly. The single- and multi-spike traces deliver known doses analytically; `predict_heat_injury()` recovers each with the truth lying inside its credible interval. This sanity-check pattern — three traces with known doses, plus the recovery check — is what each new dataset's heat-injury prediction should pass before being trusted.



# **Case Study 1: Shrimp data** {#sec-shrimp}

We now apply the full workflow described above — `standardize_data()` → `fit_4pl()` → `extract_tdt()` → `predict_heat_injury()` — to lethal-TDT data from brown shrimp (*Crangon crangon*). The data are 148 replicate trials spanning 30–33 °C and 5 min to 6 hours of exposure, with each replicate exposing a tank of individuals to a fixed temperature × duration combination on a specific date. The two grouping factors `Date` and `Tank` enter the model as random intercepts on `mid`, allowing day-to-day batch variation and tank-level cohort variation to be partitioned out of the dose-response surface.

## Data and standardisation {#sec-shrimp-data}

```{r shrimp-load}
#| label: shrimp-load
#| echo: true

shrimp_raw <- readxl::read_excel(
  here::here("data", "Data_Shrimp_TDT_CTmax_experiment.xlsx"),
  sheet = "LETHAL_TDT"
)

shrimp_std <- standardize_data(
  data           = shrimp_raw,
  temp           = "Temperature_assay",
  duration       = "Duration_exposure_hours",
  n_total        = "N_individuals_after_trial",
  mortality      = "Mortality_after_trial",
  random_effects = c("Date", "Tank"),
  duration_unit  = "hours"
)
```

```{r shrimp-summary-table}
#| label: tbl-shrimp-summary
#| tbl-cap: "Summary of the shrimp lethal-TDT dataset after standardisation. Temperature centre is the grand mean used by the model for `temp_c`."

shrimp_summary <- tibble::tibble(
  `n trials`              = nrow(shrimp_std),
  `Temperature range (°C)` = paste(range(shrimp_std$temp), collapse = "–"),
  `Duration range (hr)`   = paste(round(range(shrimp_std$duration), 2),
                                  collapse = "–"),
  `Median individuals/trial` = stats::median(shrimp_std$n_total),
  `Temperature centre (°C)`  = attr(shrimp_std, "tdt_meta")$temp_mean,
  `n Dates`               = length(unique(shrimp_std$Date)),
  `n Tanks`               = length(unique(shrimp_std$Tank))
)
tinytable::tt(shrimp_summary, digits = 2, escape = TRUE)
```

## Fitting the joint Bayesian 4PL {#sec-shrimp-fit}

We fit the default `fit_4pl()` model — beta-binomial likelihood, identity link, the disjoint-bounds asymptote reparameterisation, and `temp_c` slopes on all four 4PL sub-parameters — with random intercepts on `mid` for `Date` and `Tank`. The fit takes a few minutes on first render; subsequent renders reload it from disk because `file_refit = "on_change"`.

```{r shrimp-fit}
#| label: shrimp-fit
#| echo: true
#| cache: false

wf_shrimp <- fit_4pl(
  shrimp_std,
  random_effects = c("Date", "Tank"),
  chains  = 4, iter = 3000, warmup = 1500, cores = 4, seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12),
  file    = file.path(models_dir, "fit_shrimp_lethal_4pl")
)
```

### Sampling diagnostics

```{r shrimp-diagnostics}
#| label: tbl-shrimp-diagnostics
#| tbl-cap: "Sampling diagnostics for the shrimp 4PL fit. Healthy targets: Rhat < 1.01, ESS > 400, zero divergences."

tinytable::tt(diagnose_tdt_fit(wf_shrimp), digits = 3, escape = TRUE)
```

### Posterior parameters on the natural scale

```{r shrimp-parameter-table}
#| label: tbl-shrimp-parameters
#| tbl-cap: "Posterior medians and 95% credible intervals for the shrimp 4PL fit, transformed back to the natural scale. `low` and `up` are the bottom and top asymptotes of the survival curve at the grand-mean temperature; `k` is the slope on $\\log_{10}$ time; `mid` intercept is $\\log_{10}\\text{LT50}$ in hours at the grand-mean temperature; `mid temp_c slope` is $\\beta_1$, from which $z = -1/\\beta_1$."

tinytable::tt(tdt_parameter_table(wf_shrimp), digits = 3, escape = TRUE)
```

## Survival curves with observed overlay {#sec-shrimp-survival-curves}

We predict survival probability across the observed temperature × duration grid and overlay the observed proportions per replicate. The fitted curves should track the observed clouds at every temperature.

```{r shrimp-survival-curves}
#| label: fig-shrimp-survival-curves
#| fig-cap: "Posterior survival curves for brown shrimp at each assayed temperature, with 95% credible bands. Coloured points are observed per-replicate survival proportions. Duration on the log scale; population-level prediction (random effects marginalised)."
#| fig-width: 8
#| fig-height: 5

psc_shrimp <- predict_survival_curves(
  wf_shrimp,
  temps     = sort(unique(shrimp_std$temp)),
  ndraws    = 500
)
plot_survival_curves(psc_shrimp, observed = shrimp_std) +
  ggplot2::labs(x = "Exposure duration (hours)")
```

## TDT landscape {#sec-shrimp-landscape}

A dense 2-D view of survival across the entire (temperature, duration) plane shows the dose-response surface as a heatmap with overlaid contours at 25, 50, and 75% survival.

```{r shrimp-landscape}
#| label: fig-shrimp-landscape
#| fig-cap: "TDT landscape for brown shrimp: posterior median predicted survival across a 120 × 120 grid of assay temperatures × durations. White contours mark 25%, 50%, and 75% survival isolines. Overlaid points are raw observed proportions."
#| fig-width: 8
#| fig-height: 6

lsp_shrimp <- derive_tdt_landscape(wf_shrimp, ndraws = 500)
plot_tdt_landscape(lsp_shrimp, observed = shrimp_std)
```

## Classical TDT quantities: $z$, $CT_{max_{1hr}}$, $T_{crit}$ {#sec-shrimp-extract}

`extract_tdt()` returns $z$ (from per-draw log-linear regression of the LT50 curve) and $CT_{max_{1hr}}$ (4PL inversion at the 60-minute reference for 50% survival) by default; passing `lethal = TRUE` additionally returns $T_{crit}$ (the rate-multiplier damage-threshold, integrated over $r^* \in [0.1, 1]$ % HI per hour — see @eq-tcrit). The shrimp lethal-TDT dataset measures mortality counts, so we opt in to $T_{crit}$ here. Every output inherits the model's posterior, so all three quantities come with calibrated credible intervals.

```{r shrimp-extract}
#| label: shrimp-extract
#| echo: true

et_shrimp <- extract_tdt(
  wf_shrimp,
  t_ref         = 60,            # CTmax defined at 1-hour exposure
  TC_rate_range = c(0.1, 1),     # T_crit integrated over r* in % HI/hr
  ndraws        = 500,
  lethal        = TRUE           # mortality counts ⇒ opt-in to T_crit
)
```

```{r shrimp-extract-table}
#| label: tbl-shrimp-extract
#| tbl-cap: "Classical TDT summaries for brown shrimp, with full posterior uncertainty. $z$ comes from a per-draw log-linear regression of $\\log_{10}\\text{LT50}$ on temperature. $CT_{max_{1hr}}$ is the temperature at 50% survival after 1-hour exposure (4PL inversion per draw). $T_{crit}$ is the temperature at which the TDT damage rate falls below $r^*$ % HI per hour, with $r^*$ sampled uniformly on $\\log_{10}$ across $[0.1, 1]$ — the pooled posterior carries both parameter uncertainty and operational uncertainty in $r^*$."

shrimp_extract <- tibble::tribble(
  ~Quantity,                        ~Median, ~Lower, ~Upper,
  "z (°C per decade of time)",      et_shrimp$z$summary$z_median,
                                    et_shrimp$z$summary$z_lower,
                                    et_shrimp$z$summary$z_upper,
  "CTmax_1hr (°C)",                 et_shrimp$CTmax$summary$temp_median,
                                    et_shrimp$CTmax$summary$temp_lower,
                                    et_shrimp$CTmax$summary$temp_upper,
  "T_crit (°C, r* in [0.1, 1] per hour)",
                                    et_shrimp$T_crit$summary$temp_median,
                                    et_shrimp$T_crit$summary$temp_lower,
                                    et_shrimp$T_crit$summary$temp_upper
)
tinytable::tt(shrimp_extract, digits = 2, escape = TRUE)
```

```{r shrimp-tcrit-ctmax-plot}
#| label: fig-shrimp-tcrit-ctmax
#| fig-cap: "Posterior densities of $CT_{max_{1hr}}$ (left) and $T_{crit}$ (right) for brown shrimp. The horizontal bar below each density is the 95% credible interval; the point marks the posterior median. The 5–6 °C gap between the two quantities represents the sub-lethal window between the onset of detectable injury and the LT50 limit."
#| fig-width: 8
#| fig-height: 3.5

patchwork::wrap_plots(
  plot_temperature_density(et_shrimp$CTmax,
                            x_label = expression(CT[max[1*hr]]~("°C"))),
  plot_temperature_density(et_shrimp$T_crit,
                            x_label = expression(T[crit]~("°C"))),
  ncol = 2
)
```

## Heat injury under reference traces {#sec-shrimp-hi}

We now project the fitted shrimp model through four temperature traces — flat baseline, single 1-hour spike, three 1-hour spikes, and a multi-day diurnal cycle — and predict cumulative heat-injury (HI) and survival under each. The damage threshold $T_c$ passed to `predict_heat_injury()` is the posterior median of $T_{crit}$ from `extract_tdt()` (model-derived, not hand-picked). The single-spike scenario is deliberately calibrated as an analytical sanity check: the spike temperature is set to the posterior median of $CT_{max_{1hr}}$, so the 1-hour spike is — by definition — a 1-hour exposure at the LT50 temperature and should accumulate ~100% LT50-dose (≈ 50% mortality) by the end of the spike hour.

```{r shrimp-hi}
#| label: shrimp-hi
#| echo: true

# Damage threshold = posterior median of T_crit from the fit. Model-derived.
shrimp_T_c       <- et_shrimp$T_crit$summary$temp_median
shrimp_CTmax_med <- et_shrimp$CTmax$summary$temp_median

shrimp_scens <- make_temperature_scenarios(
  baseline   = 17,                  # shrimp natural holding temperature
  spike_temp = shrimp_CTmax_med,    # 1-hr spike at LT50-temperature → 100% HI
  n_hours    = 96,                  # 4 days; matches diurnal length below
  spike_times_single = 24,
  spike_times_multi  = c(24, 48, 72),
  # Diurnal: 4 days designed to mimic a clustered heat wave on a temperate
  # coastal shrimp habitat. Day 1 is cool (peak below T_crit, no injury);
  # the next three days build a heatwave with progressively higher peaks and
  # warmer nights — the canonical "warm days come in runs, with warm nights"
  # pattern of temporal autocorrelation. Calibrated so the two hottest days
  # each deliver roughly 50–60 % LT50-dose, bringing cumulative HI past 150 %
  # and predicted survival down to ~20 % by day 4.
  diurnal_n_days     = 4,
  diurnal_day_peaks  = c(22.0, 27.0, 30.0, 30.5),
  diurnal_night_temp = c(18.0, 19.5, 21.5, 22.0),
  diurnal_peak_fwhm  = 5,           # ~4–5 h near the daily peak
  diurnal_noise_sd   = 0.3,
  diurnal_seed       = 1L
)

# Predict HI + survival under each scenario, without repair.
shrimp_hi <- purrr::map(shrimp_scens, function(sc) {
  predict_heat_injury(
    trace       = sc,
    workflow    = wf_shrimp,
    target_surv = 0.5,
    T_c         = shrimp_T_c,
    ndraws      = 500
  )
})

# Sharpe-Schoolfield repair parameters tuned to shrimp: optimum ~17 °C,
# r_ref = 0.05 → 5% LT50-dose repaired per hour at TREF (≈ 50% HI cleared
# in 10 hours of permissive temperature).
shrimp_repair_pars <- list(
  TA   = 14065,
  TAL  = 50000,
  TAH  = 120000,
  TL   = 10.5 + 273.15,
  TH   = 22.5 + 273.15,
  TREF = 17   + 273.15,
  r_ref = 0.05
)

# Re-run with repair on the two scenarios that produce non-trivial HI:
# the multi-spike and the diurnal trace.
shrimp_hi_multi_rep <- predict_heat_injury(
  trace       = shrimp_scens$multi_spike,
  workflow    = wf_shrimp,
  target_surv = 0.5,
  T_c         = shrimp_T_c,
  ndraws      = 500,
  repair      = TRUE,
  repair_pars = shrimp_repair_pars
)

shrimp_hi_diurnal_rep <- predict_heat_injury(
  trace       = shrimp_scens$diurnal,
  workflow    = wf_shrimp,
  target_surv = 0.5,
  T_c         = shrimp_T_c,
  ndraws      = 500,
  repair      = TRUE,
  repair_pars = shrimp_repair_pars
)
```

```{r shrimp-hi-traces}
#| label: fig-shrimp-hi-traces
#| fig-cap: "The four reference temperature traces used for the shrimp heat-injury prediction, all 96 hours long. Baseline 17 °C is the natural holding temperature; the 1-hour spikes are set to the posterior median $CT_{max_{1hr}}$ (LT50 temperature) so each spike delivers a full LT50 dose by construction. The diurnal scenario simulates a clustered 4-day heatwave: a cool day with a peak below $T_{crit}$, then three progressively hotter days whose nights also warm with the daytime peaks (warm days come in autocorrelated runs in nature). The two hottest days deliver roughly 50–60 % LT50-dose each. The dashed line marks the model-derived damage threshold $T_c = T_{crit}$."
#| fig-width: 7
#| fig-height: 6

plot_temperature_scenarios(shrimp_scens, T_c = shrimp_T_c)
```

```{r shrimp-hi-single-noreapair}
#| label: fig-shrimp-hi-single-norep
#| fig-cap: "Cumulative heat injury (top) and predicted survival (bottom) for brown shrimp under the **single-spike** trace, **no repair**. Because the spike sits at the posterior median $CT_{max_{1hr}}$, the analytical expectation is ~100% LT50-dose by the end of hour 24, which produces ~50% mortality. The posterior median tracks this prediction. Posterior 95% credible band shown."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(shrimp_hi$single_spike)
```

```{r shrimp-hi-multi-noreapair}
#| label: fig-shrimp-hi-multi-norep
#| fig-cap: "Cumulative heat injury (top) and predicted survival (bottom) for brown shrimp under the **multi-spike** trace, **no repair**. Three 1-hour spikes at $CT_{max_{1hr}}$ deliver ~3 × LT50-dose, driving cumulative HI well past 100% and survival to near zero."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(shrimp_hi$multi_spike)
```

```{r shrimp-hi-multi-rep}
#| label: fig-shrimp-hi-multi-rep
#| fig-cap: "Same multi-spike trace, this time **with Sharpe-Schoolfield repair** active (`r_ref = 0.05` doses per hour at 17 °C — ≈ 50% repair in 10 hours of permissive temperature). Repair clears most of the dose between spikes, leaving residual injury but stopping the runaway accumulation seen without repair."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(shrimp_hi_multi_rep)
```

The diurnal scenario (@fig-shrimp-hi-diurnal-norep, @fig-shrimp-hi-diurnal-rep) is the closest to a real coastal record: a clustered 4-day heatwave. Day 1 stays below $T_{crit}$ and accumulates no injury. Days 2–4 form an autocorrelated run of progressively hotter days, with the hot afternoons coupled to warmer nights (a common natural pattern set by the same synoptic conditions that drive the warm days themselves). The no-repair trajectory shows the expected morphology: a flat first day, then three injury-accumulating bouts whose magnitude scales with the daily peak — by the end of day 4 cumulative HI is well past 100 % and predicted survival has dropped to ~20 %. With repair on, the off-peak hours and the cool first day act as recovery phases, partially offsetting the warm-day damage; the final trajectory is set by the net balance between daily accrual at sub-$CT_{max_{1hr}}$ peaks and overnight + cool-day repair.

```{r shrimp-hi-diurnal-noreapair}
#| label: fig-shrimp-hi-diurnal-norep
#| fig-cap: "Cumulative heat injury (top) and predicted survival (bottom) for brown shrimp under the **4-day diurnal heatwave** trace, **no repair**. The cool first day flat-lines below $T_{crit}$; days 2–4 each accrue an injury bout that scales with the daily peak. By the end of day 4 cumulative HI is well past 100 % and predicted survival has fallen to ~20 %. Posterior median + 95 % credible band."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(shrimp_hi$diurnal)
```

```{r shrimp-hi-diurnal-rep}
#| label: fig-shrimp-hi-diurnal-rep
#| fig-cap: "Same 4-day heatwave trace, this time **with Sharpe-Schoolfield repair** active (`r_ref = 0.05` doses per hour at 17 °C). The cool first day and overnight troughs now act as recovery phases. The trajectory is dominated by the net balance between daily injury accrual at sub-$CT_{max_{1hr}}$ peaks and recovery during cool intervals."
#| fig-width: 7
#| fig-height: 5

plot_heat_injury(shrimp_hi_diurnal_rep)
```

```{r shrimp-hi-summary}
#| label: tbl-shrimp-hi-summary
#| tbl-cap: "Final cumulative HI and predicted survival under each scenario for brown shrimp, posterior medians with 95% credible intervals."

final_row <- function(hi) {
  s <- tail(hi$summary, 1)
  tibble::tibble(
    `HI median (%)` = round(s$hi_median, 1),
    `HI lower`      = round(s$hi_lower,  1),
    `HI upper`      = round(s$hi_upper,  1),
    `Survival median`  = round(s$surv_median, 3),
    `Survival lower`   = round(s$surv_lower,  3),
    `Survival upper`   = round(s$surv_upper,  3)
  )
}

shrimp_hi_table <- dplyr::bind_rows(
  final_row(shrimp_hi$flat)         |> dplyr::mutate(Scenario = "Flat",
                                                     Repair = "None"),
  final_row(shrimp_hi$single_spike) |> dplyr::mutate(Scenario = "Single spike",
                                                     Repair = "None"),
  final_row(shrimp_hi$multi_spike)  |> dplyr::mutate(Scenario = "Multi spike",
                                                     Repair = "None"),
  final_row(shrimp_hi_multi_rep)    |> dplyr::mutate(Scenario = "Multi spike",
                                                     Repair = "Sharpe-Schoolfield"),
  final_row(shrimp_hi$diurnal)      |> dplyr::mutate(Scenario = "Diurnal (7 days)",
                                                     Repair = "None"),
  final_row(shrimp_hi_diurnal_rep)  |> dplyr::mutate(Scenario = "Diurnal (7 days)",
                                                     Repair = "Sharpe-Schoolfield")
) |>
  dplyr::select(Scenario, Repair, dplyr::everything())

tinytable::tt(shrimp_hi_table, escape = TRUE)
```

The shrimp case study shows the complete loop: raw counts → fitted model → classical TDT quantities (with credible intervals on every one of them) → heat-injury predictions under fluctuating field traces (with and without repair). Every quantity reported here inherits the joint posterior of the single 4PL fit — there is no second-stage regression, no point-estimate-then-bootstrap, no separate ramping-rate experiment. The same machinery applies to any TDT-style proportion dataset by changing only `standardize_data()`'s column-name arguments.

## Sublethal time-to-knockdown data {#sec-shrimp-sublethal}

So far the brown-shrimp analysis has used the lethal TDT dataset, which records mortality counts at fixed (temperature × duration) cells. The same animals can also be analysed via a **sublethal** endpoint — loss of response to touch — where each cup contributes a single time-to-knockdown observation rather than a binary survival count at a pre-specified duration. This sublethal dataset (shipped in the project's `data/Data_Shrimp_TDT_CTmax_experiment.xlsx`, sheet `SUBLETHAL_TDT`) covers five assay temperatures from 31 to 33 °C, distributed across three experimental dates and six holding tanks; the exact cup counts per temperature are reported in @tbl-shrimp-sub-summary below.

The classical analysis of time-to-event sublethal data is a log-linear regression of $\log_{10}(\text{time to event})$ on temperature, giving $z = -1/\beta_1$ and $CT_{max_{1hr}} = \bar{T} + (\log_{10} 60 - \beta_0)/\beta_1$ directly from the regression coefficients [@jorgensen_unifying_2021; @orsted_finding_2022]. Below we reproduce that classical analysis in a Bayesian framework (with the same hierarchical structure as the joint 4PL above), then re-cast the *same* time-to-event data as a proportion-knocked-down 4PL fit [@orsted_suzukii_2024], and finally compare the two pipelines' TDT lines and posterior summaries.

### Data preparation {#sec-shrimp-sub-data}

The raw sublethal sheet records one row per cup, with the time when the cup's individuals lost response to touch. We parse the Excel time-of-day values, restrict to non-excluded rows, and add the grouping factors used as random intercepts (`date_experiment`, `tank_ID`, `cup_ID`).

```{r shrimp-sublethal-data}
#| label: shrimp-sublethal-data
#| echo: true

sublethal_raw <- readxl::read_excel(
  here::here("data", "Data_Shrimp_TDT_CTmax_experiment.xlsx"),
  sheet = "SUBLETHAL_TDT"
)

# Time_to_death is stored as a fractional-day string (Excel format).
# Starting_time is a POSIXct on 1899-12-31 carrying the time-of-day; we
# extract its time-of-day and difference the two to get exposure duration
# in minutes.
tdt_sub <- sublethal_raw |>
  dplyr::filter(Exclude == "no") |>
  dplyr::mutate(
    assay_temp      = as.numeric(Assay_temperature),
    date_experiment = factor(as.character(as.Date(Date))),
    tank_ID         = factor(Tank),
    cup_ID          = factor(paste0(Trial_ID, "_", Sample)),
    ttd_frac        = as.numeric(Time_to_death),
    start_frac      = (as.numeric(format(Starting_time, "%H")) * 3600 +
                       as.numeric(format(Starting_time, "%M")) * 60 +
                       as.numeric(format(Starting_time, "%S"))) / 86400,
    time_to_event   = (ttd_frac - start_frac) * 1440          # minutes
  ) |>
  dplyr::filter(is.finite(time_to_event), time_to_event > 0) |>
  dplyr::mutate(
    log10_time = log10(time_to_event),
    temp_mean  = mean(assay_temp),
    temp_c     = assay_temp - temp_mean
  )
```

@tbl-shrimp-sub-summary summarises the sublethal panel: five assay temperatures spanning the sublethal-to-lethal transition.

```{r shrimp-sublethal-summary-table}
#| label: tbl-shrimp-sub-summary
#| tbl-cap: "Brown-shrimp sublethal time-to-knockdown panel: number of cups, median exposure time to knockdown, and 95% range, per assay temperature."

tdt_sub |>
  dplyr::group_by(assay_temp) |>
  dplyr::summarise(
    `N cups`      = dplyr::n(),
    `Median (min)` = round(stats::median(time_to_event), 2),
    `Lower 95-pct (min)`  = round(stats::quantile(time_to_event, 0.025), 2),
    `Upper 95-pct (min)`  = round(stats::quantile(time_to_event, 0.975), 2),
    .groups       = "drop"
  ) |>
  dplyr::rename(`Temperature (°C)` = assay_temp) |>
  tinytable::tt(escape = TRUE)
```

### Linear Bayesian time-to-event model {#sec-shrimp-sub-linear}

The classical sublethal TDT analysis fits a hierarchical log-linear regression: $\log_{10}(\text{time to event}) = \beta_0 + \beta_1 (T - \bar{T}) + \text{random effects} + \varepsilon$, with random intercepts on the three experimental grouping factors (`date_experiment`, `tank_ID`, `cup_ID`). The two scalar TDT quantities follow as transformations of the posterior of $(\beta_0, \beta_1)$:

$$
z \;=\; -\frac{1}{\beta_1}, \qquad
CT_{max_{1hr}} \;=\; \bar{T} \;+\; \frac{\log_{10}(60) - \beta_0}{\beta_1}.
$$

```{r shrimp-sublethal-linear-fit}
#| label: shrimp-sublethal-linear-fit
#| echo: true
#| cache: false

priors_sub_lin <- c(
  brms::prior(normal(2, 1.5),    class = "Intercept"),
  brms::prior(normal(-1, 1),     class = "b"),
  brms::prior(exponential(2),    class = "sd"),
  brms::prior(exponential(2),    class = "sigma")
)

fit_sub_lin <- brms::brm(
  brms::bf(log10_time ~ temp_c +
           (1 | date_experiment) + (1 | tank_ID) + (1 | cup_ID)),
  data       = tdt_sub,
  prior      = priors_sub_lin,
  chains     = 4,
  cores      = 4,
  iter       = 2000,                     # 4 chains x 1000 post-warmup = 4000 draws
  warmup     = 1000,
  init       = 0,
  control    = list(adapt_delta = 0.995, max_treedepth = 14),
  backend    = "cmdstanr",
  seed       = 123,
  silent     = 2,                        # hide cmdstanr chain trace
  refresh    = 0,                        # hide per-iteration progress
  file       = file.path(models_dir, "fit_shrimp_sublethal_linear"),
  file_refit = "on_change"
)
```

Posterior summaries on the natural scale follow from the regression coefficients and the mean-centring temperature:

```{r shrimp-sublethal-linear-derive}
#| label: shrimp-sublethal-linear-derive
#| echo: true

post_sub_lin <- posterior::as_draws_df(fit_sub_lin) |> as.data.frame()
temp_mean_sub <- mean(tdt_sub$assay_temp)

z_lin_draws       <- -1 / post_sub_lin$b_temp_c
CTmax_1hr_lin_draws <- temp_mean_sub +
  (log10(60) - post_sub_lin$b_Intercept) / post_sub_lin$b_temp_c

summarise_post <- function(x) {
  q <- stats::quantile(x, c(0.025, 0.5, 0.975), na.rm = TRUE, names = FALSE)
  tibble::tibble(median = q[2], lower = q[1], upper = q[3])
}
```

```{r shrimp-sublethal-linear-table}
#| label: tbl-shrimp-sub-linear
#| tbl-cap: "Posterior summaries from the linear Bayesian sublethal time-to-event model: thermal sensitivity $z$ and the critical temperature at the 1-hour reference duration $CT_{max_{1hr}}$, derived from the regression coefficients."

tibble::tibble(
  Quantity     = c("z (°C)", "CTmax at 1 hr (°C)"),
  Median       = c(stats::median(z_lin_draws),
                   stats::median(CTmax_1hr_lin_draws)),
  Lower        = c(stats::quantile(z_lin_draws, 0.025, names = FALSE),
                   stats::quantile(CTmax_1hr_lin_draws, 0.025, names = FALSE)),
  Upper        = c(stats::quantile(z_lin_draws, 0.975, names = FALSE),
                   stats::quantile(CTmax_1hr_lin_draws, 0.975, names = FALSE))
) |>
  tinytable::tt(digits = 3, escape = TRUE)
```

### Joint 4PL on proportion knocked-down {#sec-shrimp-sub-4pl}

Following @orsted_suzukii_2024, the *same* time-to-knockdown data can be re-cast as proportion-alive observations. At each assay temperature, the fraction of cups that have *not yet* been knocked down by time $t$ defines a cumulative survival curve; that curve is the canonical 4PL input. Concretely, for each $(\text{date} \times \text{tank} \times \text{temperature})$ trial we count cups at a grid of evaluation times — when $t$ is short, most cups are still alive; when $t$ is long, almost none are. The resulting $(n_{\text{total}}, n_{\text{alive}})$ panel is then fed straight into `standardize_data()` and `fit_4pl()`.

```{r shrimp-sublethal-4pl-panel}
#| label: shrimp-sublethal-4pl-panel
#| echo: true

# Evaluation grid: log-spaced, covering the observed knockdown-time range.
ttd_range <- range(tdt_sub$time_to_event, na.rm = TRUE)
eval_times_min <- 10 ^ seq(log10(max(0.2, ttd_range[1] * 0.5)),
                            log10(ttd_range[2] * 1.5),
                            length.out = 18)

# For each (Date, Tank, Temp) trial, count cups still alive at each eval_time.
sub_4pl_panel <- tdt_sub |>
  dplyr::group_by(date_experiment, tank_ID, assay_temp) |>
  dplyr::summarise(n_total = dplyr::n(),
                   ttds    = list(time_to_event),
                   .groups = "drop") |>
  purrr::pmap_dfr(function(date_experiment, tank_ID, assay_temp,
                           n_total, ttds) {
    tibble::tibble(
      date_experiment = date_experiment,
      tank_ID         = tank_ID,
      assay_temp      = assay_temp,
      duration_min    = eval_times_min,
      n_total         = n_total,
      n_alive         = vapply(eval_times_min,
                                function(t) sum(ttds > t), integer(1))
    )
  })

std_sub_4pl <- standardize_data(
  sub_4pl_panel,
  temp           = "assay_temp",
  duration       = "duration_min",
  n_total        = "n_total",
  n_surv         = "n_alive",
  random_effects = c("date_experiment", "tank_ID"),
  duration_unit  = "minutes"
)
```

```{r shrimp-sublethal-4pl-fit}
#| label: shrimp-sublethal-4pl-fit
#| echo: true
#| cache: false

wf_sub_4pl <- fit_4pl(
  std_sub_4pl,
  random_effects = c("date_experiment", "tank_ID"),
  chains  = 4, iter = 2000, cores = 4, seed = 123,    # 4000 post-warmup draws
  silent  = 2, refresh = 0,                            # hide chain trace
  file    = file.path(models_dir, "fit_shrimp_sublethal_4pl"),
  file_refit = "on_change"
)
```

::: {.callout-note}
The 4PL aggregation pools cups within each $(\text{date} \times \text{tank} \times \text{temperature})$ trial, so `cup_ID` is no longer a grouping variable; cup-level dispersion is instead absorbed by the beta-binomial precision parameter $\phi$. The 4PL therefore uses two random intercepts (`date_experiment`, `tank_ID`) — one fewer than the cup-level linear fit above — plus $\phi$ to capture the extra-binomial variance.
:::

```{r shrimp-sublethal-4pl-extract}
#| label: shrimp-sublethal-4pl-extract
#| echo: true

# Sublethal endpoint (knockdown): leave `lethal = FALSE` so that T_crit is
# not derived. The fitted z for knockdown data measures performance-reduction
# slope rather than damage-accumulation slope, and the two are not
# interchangeable — see the warning callout in @sec-joint-derive.
et_sub_4pl <- extract_tdt(
  wf_sub_4pl,
  t_ref           = 60,
  time_multiplier = 1,
  ndraws          = 1000
)
```

```{r shrimp-sublethal-4pl-table}
#| label: tbl-shrimp-sub-4pl
#| tbl-cap: "Posterior summaries from the joint Bayesian 4PL on proportion-knocked-down data: $z$ and $CT_{max_{1hr}}$. $T_{crit}$ is not reported for sublethal endpoints (see @sec-joint-derive)."

tibble::tribble(
  ~Quantity,             ~Median,                              ~Lower,                              ~Upper,
  "z (°C)",              et_sub_4pl$z$summary$z_median,        et_sub_4pl$z$summary$z_lower,        et_sub_4pl$z$summary$z_upper,
  "CTmax at 1 hr (°C)",  et_sub_4pl$CTmax$summary$temp_median, et_sub_4pl$CTmax$summary$temp_lower, et_sub_4pl$CTmax$summary$temp_upper
) |>
  tinytable::tt(digits = 3, escape = TRUE)
```

### Comparing the two sublethal pipelines {#sec-shrimp-sub-compare}

The two pipelines analyse the *same* knockdown observations through different statistical structures: the linear model regresses individual log-times on temperature, the 4PL fits a sigmoidal cumulative-knockdown curve at each temperature and reads off $\log_{10}\text{LT50}$ from each posterior draw. Their posteriors of $z$ and $CT_{max_{1hr}}$ should agree to within the data's information content. @fig-shrimp-sub-compare overlays the two TDT lines (predicted $\log_{10}(\text{time to 50\% knockdown})$ versus assay temperature), and @tbl-shrimp-sub-compare puts the posterior summaries side-by-side.

```{r shrimp-sublethal-compare-data}
#| label: shrimp-sublethal-compare-data
#| echo: true

temp_grid_sub <- seq(min(tdt_sub$assay_temp) - 0.25,
                     max(tdt_sub$assay_temp) + 0.25, by = 0.05)

# Linear model: predicted log10(time) from the population-level posterior
lin_grid <- tibble::tibble(temp_c = temp_grid_sub - temp_mean_sub)
lin_lp   <- brms::posterior_linpred(fit_sub_lin, newdata = lin_grid,
                                     re_formula = NA)
lin_pred <- tibble::tibble(
  assay_temp     = temp_grid_sub,
  log10_time_med = apply(lin_lp, 2, stats::median),
  log10_time_lo  = apply(lin_lp, 2, stats::quantile, 0.025, names = FALSE),
  log10_time_hi  = apply(lin_lp, 2, stats::quantile, 0.975, names = FALSE),
  method         = "Linear Bayesian"
)

# 4PL model: per-draw log10(LT50) curve from extract_tdt
ltx_sub <- derive_ltx_curve(
  wf_sub_4pl,
  temp_grid       = temp_grid_sub,
  target_surv     = 0.5,
  ndraws          = 1000,
  time_multiplier = 1
)

pl_4pl_pred <- ltx_sub$summary |>
  dplyr::transmute(
    assay_temp     = temp,
    log10_time_med = log10(duration_median),
    log10_time_lo  = log10(duration_lower),
    log10_time_hi  = log10(duration_upper),
    method         = "Joint 4PL (proportion knocked-down)"
  )

tdt_lines_sub <- dplyr::bind_rows(lin_pred, pl_4pl_pred) |>
  dplyr::mutate(method = factor(method,
                                levels = c("Linear Bayesian",
                                           "Joint 4PL (proportion knocked-down)")))
```

::: {#fig-shrimp-sub-compare}

```{r shrimp-sublethal-compare-plot}
#| label: plot-shrimp-sub-compare
#| fig-width: 7
#| fig-height: 5

ggplot2::ggplot(tdt_lines_sub,
                ggplot2::aes(x = assay_temp, y = log10_time_med,
                             colour = method, fill = method)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = log10_time_lo,
                                    ymax = log10_time_hi),
                       alpha = 0.18, colour = NA) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_jitter(
    data = tdt_sub,
    ggplot2::aes(x = assay_temp, y = log10_time),
    inherit.aes = FALSE,
    width = 0.05, height = 0,
    alpha = 0.35, size = 1.2, colour = "grey30"
  ) +
  ggplot2::scale_colour_manual(values = c("#2c7fb8", "#d95f02")) +
  ggplot2::scale_fill_manual(values   = c("#2c7fb8", "#d95f02")) +
  ggplot2::labs(x = "Assay temperature (°C)",
                y = expression(log[10]~"(time to 50% knockdown, min)"),
                colour = NULL, fill = NULL) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "top")
```

Posterior $\log_{10}(\text{time to 50\% knockdown})$ versus assay temperature, from the linear Bayesian model (blue) and the joint 4PL on proportion-knocked-down data (orange). Shaded ribbons are 95% credible intervals; grey points are the individual observed knockdown times. The two pipelines analyse the *same* data and produce TDT lines that overlap across the assay range.
:::

```{r shrimp-sublethal-compare-table}
#| label: tbl-shrimp-sub-compare
#| tbl-cap: "Side-by-side posterior summaries of $z$ and $CT_{max_{1hr}}$ from the linear Bayesian time-to-event model and the joint 4PL on proportion-knocked-down data. Both fits use the same sublethal observations and the same hierarchical grouping structure (date and tank; the linear fit additionally carries a per-cup random intercept that is absorbed into `n_total` in the 4PL aggregation)."

compare_sub <- tibble::tibble(
  Quantity = c("z (°C)", "CTmax at 1 hr (°C)"),
  `Linear median`  = c(stats::median(z_lin_draws),
                       stats::median(CTmax_1hr_lin_draws)),
  `Linear lower`   = c(stats::quantile(z_lin_draws, 0.025, names = FALSE),
                       stats::quantile(CTmax_1hr_lin_draws, 0.025, names = FALSE)),
  `Linear upper`   = c(stats::quantile(z_lin_draws, 0.975, names = FALSE),
                       stats::quantile(CTmax_1hr_lin_draws, 0.975, names = FALSE)),
  `4PL median`     = c(et_sub_4pl$z$summary$z_median,
                       et_sub_4pl$CTmax$summary$temp_median),
  `4PL lower`      = c(et_sub_4pl$z$summary$z_lower,
                       et_sub_4pl$CTmax$summary$temp_lower),
  `4PL upper`      = c(et_sub_4pl$z$summary$z_upper,
                       et_sub_4pl$CTmax$summary$temp_upper)
)

tinytable::tt(compare_sub, digits = 3, escape = TRUE)
```

The two sublethal pipelines yield consistent estimates of $z$ and $CT_{max_{1hr}}$ on this dataset (medians within their mutual credible intervals; see @tbl-shrimp-sub-compare). The 4PL on proportion-knocked-down data has the practical advantage that the *same* `extract_tdt()` workflow used for the lethal counts produces a posterior $\log_{10}\text{LT50}(T)$ curve and a posterior $CT_{max_{1hr}}$ — quantities the linear time-to-event model does not produce directly. $T_{crit}$, however, is not derived from the sublethal fit: as discussed in @sec-joint-derive, the rate-multiplier $T_{crit}$ in Equation @eq-tcrit interprets $z$ as a damage-accumulation slope, which the sublethal $z$ is not. The empirical estimate of $T_{crit}$ for shrimp instead comes from the lethal-TDT fit in @sec-shrimp-extract, where the rate-multiplier interpretation is meaningful.

### A caution on cross-endpoint comparison of $T_{crit}$ {#sec-shrimp-tcrit-endpoint-caveat}

The sublethal $T_{crit}$ on this dataset sits ~3.5 °C above the lethal $T_{crit}$ reported in @tbl-shrimp-extract, and their 95 % credible intervals do not overlap. This is *not* a sign that damage starts accumulating at a higher temperature for the sublethal endpoint — biologically that is impossible, since loss of response to touch manifests before death. The discrepancy is mechanical and traces to the rate-multiplier formula $T_{crit} = CT_{max_{1hr}} + z \cdot \log_{10}(r^*/100)$, which inherits a 2.5× sensitivity to $z$ in the centre of the integrated $r^*$ range. On the shrimp panel $CT_{max_{1hr}}$ for the two endpoints is essentially identical (~31.4–31.6 °C), but $z$ is more than halved for the sublethal endpoint ($\approx 1.1$ vs $\approx 2.6$ °C); $2.5 \times (2.6 - 1.1) \approx 3.7$ °C of $T_{crit}$ gap follows from that alone. The lethal and sublethal TDT lines therefore cross at $\approx CT_{max_{1hr}}$ and the steeper sublethal line, extrapolated below the 31–33 °C assay window, predicts times to knockdown that exceed times to death — biologically impossible and a sign that the linearity assumption is breaking down outside the assay range. The rate-multiplier $T_{crit}$ is therefore best read as an **endpoint-conditional operational summary** of where the *fitted TDT line* crosses a low-rate floor, not as the universal physiological threshold of **havro_faber_2023** or @jorgensen_unifying_2021. The same-species $T_{crit}^{(\text{lethal})}$ and $T_{crit}^{(\text{sublethal})}$ should not be compared directly. See [notes/2026-05-13-tcrit-lethal-vs-sublethal.qmd](../notes/2026-05-13-tcrit-lethal-vs-sublethal.qmd) for the TDT-line comparison plot, the full derivation, and a discussion of two cleaner alternatives — an empirical Q$_{10}$-breakpoint of **havro_faber_2023** per posterior draw (requires permissive-range data this panel does not have) and an absolute-damage-units rescaling (requires a species-specific calibration between endpoints).

# **References**

::: {#refs}
:::

# **Session Information**

```{r session-info}
#| label: session-info
#| results: asis

pander::pander(sessionInfo())
```

:::
