---
title: "Supplementary information for: A hierarchical Bayesian 4PL recovers the classical thermal death time framework and resolves its statistical pathologies"
author:
  - name: "Daniel Noble"
    email: daniel.wa.noble@gmail.com
    corresponding: true
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    fig-cap-location: bottom
  docx:
    reference-doc: ../bib/templates/template.docx
    toc: false
    number-sections: true
bibliography: ../bib/tdt_problems.bib
csl: ../bib/ecology-letters.csl
link-citations: true
crossref:
  fig-prefix: "Figure S"
  tbl-prefix: "Table S"
  eq-prefix: "Equation S"
  sec-prefix: "Section S"
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
#| label: setup-supp

library(here)
library(dplyr)
library(ggplot2)
library(brms)
library(posterior)
library(flextable)

set.seed(123)
```

# Statistical liabilities of the two-step TDT pipeline {#sec-pathologies}

<!--
Full version of the §2–§9 catalogue from tdt_problems.qmd, with citations:
  - Generated-regressor problem (@pagan1984, @murphytopel1985)
  - Heteroscedasticity discarded by point-estimate LT50
  - Right-censoring at temperatures where mortality never reached 50 %
  - Arcsine vs logit (@warton_arcsine_2011)
  - Mean-variance mismatch in OLS on percentages
  - Overdispersion (@harrison_olre_2014, @harrison_bb_2015)
  - Boundary observations
  - Pseudoreplication (@hurlbert_1984, @schielzeth_nested_2013, @bolker2009)
  - 4PL identifiability (@bolker_ecologicalmodels_2008, @ritz_drc_2015)
  - Common-slope-across-temperatures
  - Forking paths (@simmons_falsepositive_2011, @gelman_loken_2014)
  - Endpoint heterogeneity (@terblanche_ctmax_2007, @sinclair_predict_2016)
This text exists in long form in tdt_problems.qmd — port it here and
shorten the main-text version accordingly.
-->

# Algebraic equivalence between the 4PL and the classical TDT pipeline {#sec-equivalence}

## Closed-form LT50 from the 4PL

## Closed-form heat-injury integral

## Closed-form dose → mortality mapping for arbitrary mortality fractions

# Reusable R helpers {#sec-helpers}

<!--
Listings (from §10 of tdt_problems.qmd):
- extract_tdt_params()
- LT50_T_curve()
- heat_injury()
- time_to_mortality()
-->

# Posterior-predictive checks and residual diagnostics {#sec-ppc}

```{r ppc-lethal}
#| label: fig-ppc-lethal
#| fig-cap: "Posterior-predictive checks for the lethal-TDT 4PL: observed vs. simulated counts grouped by assay temperature, and randomised-quantile residuals against fitted values."
#| eval: false
```

# Sensitivity analyses {#sec-sensitivity}

## Celsius-linear vs. Arrhenius (1/T-Kelvin) parameterisation of `mid(T)`

## OLRE vs. beta-binomial overdispersion

## Repair-kernel parameter sensitivity

# Pareto-k diagnostics and `reloo()` results {#sec-loo}

# Computing environment {#sec-session}

```{r session-info}
sessionInfo()
```

# References {.unnumbered}

::: {#refs}
:::
